### 2022-10-19 chapter1.tex
Added chapter1.tex and refs.bib to project (finally!)

\documentclass[10pt,letterpaper]{article}

% PACKAGE CONFIGURATIONS
\usepackage{amsmath}

\usepackage{graphicx}
\graphicspath{{./figures/}}

\usepackage[margin=1in]{geometry}
\usepackage[right]{lineno}  % line numbers
%\usepackage{setspace}  % double spacing
%\doublespacing
\usepackage[skip=10pt]{parskip}

\usepackage{hyperref}

\usepackage[doi=false,url=false]{biblatex}
\addbibresource{refs.bib}

\usepackage[labelfont=bf, labelsep=period]{caption}
\renewcommand{\figurename}{Fig}

\begin{document}
\linenumbers

% Title must be 250 characters or less
\begin{flushleft}
{\Large\textbf{Leveraging genomic redundancy to improve inference and alignment of orthologous proteins}}
\newline
\\
Marc Singleton\textsuperscript{1},
Michael Eisen\textsuperscript{1,2,3}
\\
\bigskip
\textbf{1} Biophysics Graduate Group, UC Berkeley, Berkeley, CA, USA
\\
\textbf{2} Department of Molecular and Cell Biology, UC Berkeley, Berkeley, CA, USA
\\
\textbf{3} Howard Hughes Medical Institute, UC Berkeley, Berkeley, CA, USA
\\
\bigskip

* mbeisen@berkeley.edu

\end{flushleft}

% Abstract must be 300 words or less
\section*{Abstract}
abstract text goes here

\section*{Introduction}
Comparative genomics is powerful tool for yielding insights into evolutionary relationships, molecular function, and the forces that drive gene, genome, and population evolution. These methods often rely on the identification of homologous sequences or homologs, that is sequences with common ancestry, since this ensures that differences between sequences reflects variations in evolution from a common point of divergence. In a sense, each branch that descends from a single ancestral population is a different ``experiment'' by nature, as each is free to explore the evolutionary landscape independent of the others. However, many analyses impose the additional condition that the sequences have diverged through speciation events (orthology) rather than duplications (paralogy) or other mechanisms such as horizontal gene transfer \cite{Fitch1970}. The underlying assumption is ``orthologs'' have conserved and equivalent functions whereas ``paralogs'', by virtue of their redundancy, are more likely to diverge \cite{Ohno1970, Nowak1997, Altenhoff2012, Pegueroles2013, Soria2014}.\footnote{For multiple sequences, orthology is usually defined relative to their most recent common ancestor. This technically includes sequences which split by duplication after this point (``in-paralogs''), but excludes sequences which split by duplication before (``out-paralogs'') \cite{Remm2001}. Many current orthology inference pipelines explicitly incorporate steps to detect in-paralogs. However, the resulting orthologs groups can easily be restricted to those without in-paralogs (``single copy orthologs'') for analyses where an assumption of conserved function is necessary.} This relationship between orthology and function is an essential component of modern biological research since it permits the transfer of annotations between biological systems using sequence similarity alone.

Given this importance, methods for inferring orthologous groups of proteins were developed shortly after the first genomes were sequenced in the late 1990s \cite{Fleischmann1995, Goffeau1996, CESC1998}. One early and influential approach was to cluster triangles of hits resulting from homology searches between pairs of genomes \cite{Tatusov1997}. Graph-based approaches have remained popular, and in the intervening years many other researchers have refined this method by implementing various pre- and post-processing steps. Despite these improvements, many databases and pipelines use the same triangle clustering algorithm or other methods which require relatively few hits between sequences to infer an orthologous group, \textit{e.g.} connected components or Markov clustering \cite{Remm2001, Li2003, Jensen2007, Linard2011, Emms2015, Train2017, Cosentino2018}. However, the scale of biological sequence data has changed dramatically. For example, in the last decade, the number of annotated genomes available from NCBI has increased nearly 20-fold and currently exceeds 900 (Fig. \ref{sfig:ncbi}). Though this figure is only a rough proxy of the total number of assemblies available, it will likely continue to grow rapidly in the coming years as many large-scale genome assembly efforts such as i5K, the Bird 10,000 Genomes Project, and the Vertebrate Genomes Project have already yielded results \cite{Thomas2020, Feng2020, Rhie2021}. Thus, the dense taxonomic sampling made possible by these projects poses new challenges and opportunities for the standard methods of orthology inference and alignment, which implicitly assume fewer and more distantly related genomes or fail to fully leverage the redundancy and organization of their annotations.

Thus, in this work we developed a computational pipeline that can robustly infer and align orthologous groups of proteins even when the genomes are highly redundant (Fig. \ref{fig:graphical_overview}). We did not seek to reinvent the wheel, so the overall principle of clustering a graph of hits from homology searches is unchanged, though many details were modified to maximize the detection of highly diverged orthologs while also minimizing the impact of incomplete or incorrect annotations. Furthermore, as modern genome annotation pipelines frequently produce gene models and protein sequences in tandem, we implemented an additional clustering step to organize the resulting orthologous groups of proteins into gene-level units. However, most of our efforts were instead focused on the final step of aligning the orthologous sequences. Though genome annotation pipelines are often proficient at identifying the locus of genes, the quality of their gene models can vary considerably and can contain inaccurate exon boundaries or start codons \cite{Frankish2015, Dunne2018}. As a result, the structure and lengths of the protein sequences that result from these models can be highly heterogeneous, which poses many challenges for their alignment and subsequent analyses.

To develop these methods, we chose a set of 33 assembled and annotated \textit{Drosophila} genomes. All 12 species from the original \textit{Drosophila} 12 Genomes Consortium are included \cite{D12GC2007}. However, all have been re-sequenced and re-annotated, many with long-read sequencing technology, which has resulted in substantial improvements in the contiguity of their assemblies. Despite this and other genome assembly projects of species in the \textit{Drosophila} genus, there is not yet a database of high-quality alignments of orthologous proteins that reflects these improvements in genome assembly and diversity \cite{Miller2018, Kim2021}. Given the \textit{Drosophila} genus spans diverse habitats and over 50 million years of evolution but maintains a conserved life cycle and body plan, such a resource would facilitate a new generation of studies that illuminate the forces that drive protein evolution in unprecedented detail \cite{Wiegmann2011, Obbard2012}.

\begin{figure}[h!]
\includegraphics[width=\textwidth]{graphical_overview.png}
\centering
\caption{\textbf{Graphical summary of homology inference pipeline.}
1) Protein sequences from annotated genomes are collected. The large, open circles represent the annotated genomes, and the small, filled circles represent the protein sequences associated with each annotation. Sequences that share homology are colored with shades of the same hue. 2) Homology searches are conducted between all query-target pairs of genomes. The lines represent a statistically significant alignment between two sequences in the genome pair. 3) The raw output from the homology searches is processed to yield best hits between pairs of sequences. 4) The network of best hits is clustered into self-consistent orthologous groups. The small, filled circles and lines represent the same sequences and best hits from the previous step. 5) Orthologous groups are grouped by their parent genes, which are represented by the two sets of clusters with warm and cool colors, respectively. 6) Representative sequences in each orthologous group are aligned. Some steps are simplified or omitted for clarity.}
\label{fig:graphical_overview}
\end{figure}

\section*{Results}
\subsection*{Input genomes}
All annotated genomes in the genus \textit{Drosophila} available in April 2022 were downloaded from NCBI's RefSeq database. The assemblies annotated by the NCBI eukaryotic genome annotation pipeline must past several quality checks and have supporting transcript evidence, so the annotations are generally highly complete \cite{Thibaud-Nissen2013}. The \textit{D. miranda} annotation was excluded due to its unusual karyotype \cite{Dobzhansky1935}. Other annotations were excluded after preliminary clustering showed a deficient in the number of orthologous groups containing those genomes, indicating their annotations were less complete (data not shown). The \textit{D. melanogaster} annotation was downloaded from FlyBase \cite{Gramates2022}. The complete list of genomes is given in Table \ref{stable:genomes}.

Transcripts with alternate UTRs can yield duplicate protein sequences in the annotation. While these alternative UTRs can encode biologically distinct functions, their identical coding sequences make them indistinguishable at the protein level, and the duplicates were therefore removed. Though not strictly necessary, this greatly reduced the computational burden of later steps.

\subsection*{Extraction of best hits from BLAST output}
The protein sequences in each genome annotation were searched against each other in reciprocal pairs using BLAST, yielding a list of high-scoring segment pairs (HSPs) for each query-target pair \cite{Camacho2009}. HSPs are local alignments, meaning they do not necessarily span the entire lengths of the query and target sequences. Consequently, the search algorithm may return multiple HSPs if statistically significant regions of homology are separated by nonhomologous or poorly conserved regions. Though the most significant HSP is often used to represent all HSPs between a query-target pair, this approach can fail to rank the pairs by their overall significance if their alignments are broken into multiple HSPs. Furthermore, since query-target pairs were later filtered by the amount overlap between their sequences, it can also exclude pairs that pass the overlap threshold even if the most significant HSP alone does not. Thus, HSPs were merged into a single object called a hit. The best hits for each query were then taken as the highest-scoring hits that passed a minimum overlap criterion and were reciprocal between the query and target sequences.

\subsection*{Clustering in orthologous groups}
The best hits between sequences are naturally visualized as a graph where the sequences are nodes and best hits are edges between nodes (Fig. \ref{fig:components}). Some sets of nodes form tight clusters whereas others are more loosely associated or even isolated. Automated genome annotation is not completely accurate, so some poorly supported hits are likely spurious, arising from homologous regions that were falsely annotated as genic in two closely related genomes. Others are possibly novel genes. Regardless of their origin, these hits are not strong candidates for comparative analyses since an orthology relationship is supported by relatively few genome pairs. In contrast, likely orthologs should consistently identify each other as best reciprocal hits across many genome pairs. Thus, orthologous groups can be operationally defined as self-consistent clusters in the hit graph. Since sequence divergence or assembly and annotation errors may prevent a best reciprocal hit between orthologous sequences across all genome pairs, the clustering method should require a high degree of self-consistency without demanding complete consensus.

\begin{figure}[h!]
\includegraphics[width=\textwidth]{components/out/0022-002F.png}
\centering
\caption{\textbf{Selected connected components of hit graph.}
The best hits are shown as a graph where nodes are sequences and edges are best hits between sequences that are colored by their bit score. Two connected components, sets of nodes joined by a sequence of edges, are shown. The sequences in panels A are annotated as potassium channels whereas the sequences in panel B are uncharacterized. Some sets of nodes have a high density of edges, forming distinct clusters. Other nodes are only sparsely connected to their neighbors. Components can contain nodes of both types.}
\label{fig:components}
\end{figure}

The identification of sets of densely connected nodes in graphs is known as community detection in network analysis. There are many community detection algorithms available, but a popular approach in the context of orthology inference is building clusters progressively by identifying nodes that form a triangle with at least two other nodes in the cluster. While this method is robust when clustering smaller or more diverse sets of genomes, the large number of closely related genomes in the input data creates many edges which form triangles but are otherwise poorly supported by the remaining genomes. However, triangle clustering is a special case of the more general \textit{k}-clique percolation algorithm where \textit{k} = 3 (Fig. \ref{fig:percolation}). Higher values of \textit{k} require greater overlap between a candidate node and those already in the cluster and therefore produce tighter clusters at the cost of excluding more speculative orthology relationships. \textit{k} was set to 4 as compromise between these concerns, yielding 22,813 orthologous groups, a plurality of which contained all 33 species (Fig. \ref{sfig:clusters}).

\begin{figure}[h!]
\includegraphics[width=\textwidth]{percolation/percolation.png}
\includegraphics[width=\textwidth]{percolation/out/005F.png}
\centering
\caption{\textbf{Clique percolation algorithm.}
A) A \textit{k}-clique is a set of \textit{k} nodes which are fully connected, i.e., there is an edge between pair of nodes. Cliques are shown for \textit{k} equal to three, four, five, and six. \textit{k} equal to one and two correspond to a single node and two nodes joined by an edge, respectively. B) The clique percolation algorithm detects clusters by “rolling” a clique of some size \textit{k} over the graph. More specifically, a cluster is initiated when a set of nodes which form a \textit{k}-clique is identified. The cluster expands by shifting the \textit{k}-clique to an adjacent \textit{k}-clique that shares \textit{k}-1 nodes in common with the current \textit{k}-clique. A cluster stops expanding when there are no adjacent \textit{k}-cliques, and the algorithm terminates when there are no \textit{k}-cliques which are not part of a cluster. In this example, \textit{k} was set to four. C) A single component clustered by clique percolation with varying values of \textit{k} is shown. Larger values of \textit{k} define tighter clusters. Nodes are colored according to their cluster. If a node belongs to multiple clusters, it uses a blend of those colors.}
\label{fig:percolation}
\end{figure}

\subsection*{Addition of paralogs to orthologous groups}
A weakness of the best reciprocal hits criterion is its exclusion of recently diverged paralogs. Since only the highest scoring hits for each query are included in the graph, a paralog without a corresponding duplicate in the target genome is ignored if it is marginally more diverged than the other copy. This is corrected by adding likely paralogs to the orthologous groups. Briefly, the protein sequences in each genome were searched against themselves. If the bit score for an intra-genome hit exceeded the bit score of any inter-genome hits for the same query, the two sequences were identified as a paralogous pair. Orthologous groups were then supplemented with paralogs by adding the paired sequences for each of the original members of the orthologous group. Most orthologous groups contain no paralogs, and those that do generally have few relative to the original number of sequences in the group (Fig. \ref{sfig:paralogs}).

\subsection*{Grouping orthologous groups by gene}
As genes can have several annotated isoforms, each gene can also be associated with several orthologous groups. However, the orthologous groups are not organized into gene-level units since they were clustered using sequence similarity only. A graph-based approach was therefore used to group orthologous groups with similar sets of parent genes. First, a gene overlap graph was constructed by defining an edge between orthologous groups if the intersection of their associated sets of parent genes is at least 50\% of the smaller of the two. Gene groups were then taken as the connected components of the resulting graph.

\subsection*{Initial alignment and selection of representative sequences}
Since the annotation pipeline incorporates transcriptome data from a variety of sources, its inputs are heterogeneous in sequencing depth, developmental stage, and tissue of origin across different genomes. As a result, some orthologous genes are annotated with more splice isoforms in some genomes than in others. Under certain conditions, the clustering algorithm can merge smaller or looser clusters of these additional splice isoforms with the main cluster, creating a single orthologous group where some genes have multiple associated sequences. However, these additional sequences can complicate downstream comparative analyses that may not easily generalize to genes with multiple associated sequences. Thus, a single representative was chosen for each gene using an alignment-based strategy detailed in the methods section.

\subsection*{Selection of single copy orthologous groups}
The criteria for selecting orthologous groups for further analyses depends on the biological question under investigation. For example, studies of gene duplication will focus on orthologous groups with paralogs in some lineages but not in others. In contrast, analyses which assume functional conservation should restrict the orthologous groups to single copy orthologs since paralogs more frequently undergo functional divergence \cite{Altenhoff2012, Pegueroles2013, Soria2014}. A simple method for identifying such groups is requiring each species to have exactly one associated gene. However, since the probability of at least one missing gene annotation approaches one as the total number of genomes increases, this is too restrictive and fails to leverage the redundancy of closely related genomes. Instead, a set of phylogenetic diversity criteria detailed in Table \ref{stable:diversity_criteria} were applied to ensure the major lineages were represented in downstream analyses. Furthermore, genome-wide analyses should select one orthologous group per each of the previously identified gene groups as to not bias the results towards genes with many distinct groups of isoforms. In summary, orthologous groups failing the phylogenetic diversity criteria were first removed, and the representative for each gene group was chosen as the highest scoring orthologous group when ranked by the number species and the sum of the bit scores associated with each edge. This significantly reduced the number of orthologous groups from 22,813 to 8,566.

\subsection*{Alignment refinement}
An initial alignment was created for each orthologous group as part of the selection of representative sequences. Though the pipeline’s quality control measures ensure a high degree of overall sequence identity between members of an orthologous group, some sequences contain long “poorly supported” segments which are not aligned to most or any other sequences in the alignment. Since most common multiple sequence alignment algorithms assume the sequences are largely homologous, these segments are sometimes “over-aligned” by forcing them into alignment where chance sequence similarities occur. Typically, these segments remain contiguous, so the alignments alternate between short runs of columns with few or no gaps and large gap-rich regions. More rarely, when long poorly supported segments are adjacent to a long gap in the same sequence, the two are interlaced, yielding long gaps interrupted by short segments of spurious alignment.

The aligner MAFFT has a mode for addressing over-alignment with a parameter, \textit{a\textsubscript{max}}, that adjusts the strength of the correction \cite{Katoh2013, Katoh2016}. \textit{a\textsubscript{max}} varies between 0 and 1, but at values above 0.8, the correction severely degrades the alignment of conserved regions. Thus, the sequences were aligned with a conservative value of 0.4. Though some cases of over-alignment remained, most conserved regions were reliably aligned with few to no errors. To correct the remaining errors in alignments of the single copy orthologs, a statistical model of each alignment called a profile HMM was constructed from the well-aligned columns \cite{Eddy2009}. The original sequences were then aligned to these profiles. By defining highly conserved “anchor” regions, this approach largely prevents the alignment of chance sequence similarities in long poorly supported segments (Fig. \ref{fig:realignment}). Furthermore, since profile HMMs explicitly model the beginnings and ends of alignments, they can fully exclude these segments from the termini.

\begin{figure}[h!]
\includegraphics[width=\textwidth]{realignment/out/23D9_A.png}
\includegraphics[width=\textwidth]{realignment/out/23D9_B.png}
\centering
\caption{\textbf{Alignment with long poorly supported segments.}
The alignments of representative sequences in orthologous group 23D9 are shown before (panel A) and after (panel B) refinement. The refinement process removes long poorly supported segments at the termini. However, if residues are aligned to consensus columns before and after a long poorly segment, it remains in the alignment.}
\label{fig:realignment}
\end{figure}

\subsection*{Alignment curation}
Although the refinement process corrects most cases of over-alignment, the alignment may still contain regions whose aligned segments have poor or inconsistent support. For example, long poorly supported segments in internal regions were not removed from the alignment since they are bounded by at least one consensus column to the left and right. Additionally, some regions have a significant fraction of sequences with strongly supported segments, but the observed gap pattern is discordant with the expected phylogenetic relationships. Since they are present in so few sequences, the former segments are likely artifactual, resulting from errors during assembly or annotation. (Biological explanations such as alternative splice sites, frameshift mutations, or transposition events are also possible, however.) In contrast, the high sequence identity and clear boundaries of the segments in the latter regions suggest they are conserved but skipped exons. Given the heterogeneous sourcing of the transcript evidence, these sequences containing these segments are likely splice isoforms specific to certain tissues or developmental stage.

% To investigate these hypotheses, the genome sequences and supporting transcript evidence for several segments and regions were examined in detail: 0254, 26CF, 202C, 188E, 2E9F, 34B7, 0C31.

Since the segments in these regions are likely the result of incorrect or incomplete annotations rather than meaningful biological variation, maintaining them in the alignments would propagate spurious homologies to subsequent analyses. This is a common issue in alignments generated by automated pipelines, so downstream analyses often focus on the strongly supported regions by removing or “trimming” columns below some threshold number of gaps or sequence identity \cite{Castresana2000, CapellaGutierrez2009}. This approach, however, is inadequate if the taxonomic sampling is dense, as a single indel event along a lineage containing many species can increase the number of gaps above the threshold. Moreover, as this method does not incorporate any spatial information, it can rapidly alternate between trimming and preserving columns. Thus, it can severely disrupt any analyses which are sensitive to the spatial organization of an alignment.

Phylogenetic HMMs (phylo-HMMs) are statistical models that incorporate phylogenetic and spatial information to calculate the probability that each observation in a sequence was generated by one of several “hidden states.” Since they can evaluate both the probability of a gap pattern in a column given the known phylogenetic relationships and the local context, a phylo-HMM was used to segment the alignment into contiguous regions with different patterns of gaps. This phylo-HMM was composed of four hidden states, {1A, 1B, 2, 3}, roughly corresponding to highly conserved regions with few to no gaps, divergent regions, regions with a stable gap pattern discordant with the expected phylogenetic relationships, and regions with poorly supported segments. After the model was trained on manually labeled examples, it was used to assign a label to the columns in each alignment (Fig. \ref{fig:hmm}). Columns assigned to states 1A and 1B are the regions of interest for downstream analyses since the gaps generally follow the expected pattern given the phylogenetic tree. In contrast, columns assigned to states 2 and 3 largely corresponded to the long poorly supported segments and phylogenetically discordant regions discussed previously and were therefore removed from the alignments. Fewer than 10\% of alignments were trimmed, but when regions were trimmed, they were largely inferred as state 3 \ref{sfig:hmm_trim}.

\begin{figure}[h!]
\includegraphics[width=\textwidth]{hmm/hmm_architecture.png}
\includegraphics[width=\textwidth]{hmm/out/2252.png}
\centering
\caption{\textbf{HMM emission architecture and a decoded alignment.}
A) A single ancestral sequence evolves and splits into distinct sequences over time. The tree shows the history of these splits. White and colored boxes indicate gap and non-gap symbols in the alignment, respectively. These sequences are then extracted from genome assemblies by annotation. Annotation can erroneously exclude or include segments from a sequence. In this case, the three N-terminal residues of the fourth sequence were excluded. The sequences are coded into binary symbols where white and grey boxes indicate gap and non-gap symbols, respectively. The emission distributions consist of two components derived from these encoded sequences. The first component models the gap pattern with a Markov process. This Markov process is in turn composed of two subprocesses where the first is a phylogenetic process, and the second is a jump process. These subprocess roughly correspond to changes caused by evolution and annotation, respectively. The second component models the propensity for those patterns to remain constant by counting the number of symbols that remain constant between columns. c\textsubscript{0} and c\textsubscript{1} indicate the first and second columns in the alignment, respectively. The parameters associated with each emission component are shown to the right. Each state has a different set of parameters that yield its characteristic gap pattern and gap stickiness. B) The alignment of the representative sequences in orthologous group 2252 is decoded using the trained HMM.}
\label{fig:hmm}
\end{figure}

Though the phylo-HMM removed phylogenetically incongruent ``insertions'', some sequences still contained extensive segments of uninterrupted gaps. These segments are easily identified in regions which are otherwise highly conserved, so they are also likely the result of incorrect or incomplete annotations. Unfortunately, they can also span more divergent regions, which complicates a simple rule- or threshold-based approach to identifying them. Thus, another phylo-HMM was trained to label each position in a sequence as generated by either a ``missing'' or ``not missing'' state. In this case, however, the aligned sequences were processed individually and not as aligned columns. As the previous phylo-HMM already ensured each column has sufficient support, these labels can instead be used to exclude sequences from downstream analyses depending on the amount of tolerated overlap with the regions of interest.

\section*{Discussion}
The orthologous groups and alignments yielded by this pipeline are a valuable resource for comparative studies of gene birth/death processes and protein evolution at the level of both entire proteome and specific gene families in the \textit{Drosophila} genus. To assist these efforts, consensus phylogenetic trees were inferred from meta-alignments sampled from alignments of single copy orthologous groups (Fig. \ref{fig:trees}). Although this work focused on single copy orthologs, other studies may require different subsets of orthologous groups that demand other pre-processing and alignment strategies. Thus, the orthologous groups and the initial alignments with and without non-representative sequences are included as supplemental data.

\begin{figure}[h!]
\includegraphics[width=\textwidth]{trees/out/trees1.png}
\centering
\caption{\textbf{Phylogenetic tree of species.}
Phylogenetic trees were fit to meta-alignments sampled from alignments created by the orthology inference pipeline with the additional genome of \textit{Scaptodrosophila lebanonensis} (sleb) as an outgroup. The trees fit to each meta-alignment were then combined into a single consensus tree. Values at nodes are bootstrap percentages. A) The LG model was fit to meta-alignments directly sampled from the original protein alignments B). The GTR model was fit to meta-alignments sampled from nucleotide alignments which were “reverse translated” from the original protein alignments and their corresponding coding sequences.}
\label{fig:trees}
\end{figure}

The trends that permitted this work to substantially improve on previous efforts will quickly render its products obsolete. More \textit{Drosophila} genomes will be assembled and annotated, and the existing annotations will be updated. However, an authoritative and lasting set of orthologous groups and alignments is not the primary goal of this work. It instead serves as a case study in how dense taxonomic sampling and modern genome assembly and annotation pipelines present new opportunities and challenges to the traditional techniques for identifying and aligning orthologous groups.

For example, despite many additional pre-processing steps and other tweaks introduced by later authors, the basic framework of orthology inference by clustering the hit graph has remained largely unchanged in the past twenty years \cite{Remm2001, Li2003, Jensen2007, Linard2011, Emms2015, Train2017, Cosentino2018}. This longevity is a testament to the robustness of the underlying idea that orthologous proteins should consistently identify each other as the most similar pairs between their genomes. Even as the number of genomes and their taxonomic density has increased dramatically, many orthology inference pipelines continue to use triangle clustering, which was originally applied to sets of far fewer and more distantly related genomes \cite{Tatusov1997}. This mismatch in scale increases the chance of propagating annotation errors since only a small number of edges are needed to create or merge clusters. Fortunately, triangle clustering is a special case of the more general \textit{k}-clique percolation algorithm where \textit{k} is a tunable parameter that influences the tightness of a cluster. The optimal value of \textit{k} for a given set of genomes is unclear, but it likely depends on the desired trade-off between sensitivity and specificity. However, percolation theory or simulations may yield insights.

A more subtle difficulty is the annotation of multiple isoforms for a single gene. Though prior pipelines have generally selected the longest isoform as the representative before conducting the orthology searches, we felt this approach would introduce artifacts or other issues during alignment. Instead, as protein sequences are increasingly derived from or linked to genomic sequences, we sought to incorporate the full annotations into the orthology inference pipeline. This, however, created two additional complications. First, a single gene could have several associated orthologous groups if its isoforms belonged to different clusters. Second, a single orthologous group could have several isoforms of a single gene if its isoforms were clustered together. In both cases, the presence of multiple sequences for a single gene creates ambiguities over which is the “primary” isoform. Since the first occurs at the level of orthologous groups, the orthologous groups were first grouped by the similarity of their parent genes using a graph-based strategy. Afterwards, a single representative is easily chosen as the group with the largest number of distinct species, though other criteria are possible. Since the second occurs within an orthologous group, the sequences were first aligned, and a representative for each gene was chosen as the sequence which was most concordant with the initial alignment.

The second major innovation in this work is its method for the refinement and trimming of alignments. In refinement, a profile HMM is constructed from the well-aligned columns of an initial alignment, and the original sequences are aligned to this profile. This process usually corrects errors caused by the presence of long poorly supported segments. In trimming, a phylo-HMM is used to remove regions and segments which are poorly supported by the phylogenetic consensus. While the combination of these two steps yielded high-quality alignments that are suitable for further analyses, they are an \textit{ad hoc} fix for underlying issues with the gene models and alignment algorithms. The most principled solution is to optimize or supplement the gene models using the initial alignments generated by the orthology inference pipeline, which is possible with tools such as OMGene or OrthoFiller \cite{Dunne2018, Dunne2017}. However, if preserving the original annotations is desired or necessary, the remaining possibility is to correct the alignments as we have done here. In fact, the errors we sought to address broadly stem from shortcomings of current alignment algorithms rather than errors in the sequences themselves. Though the scoring functions of modern multiple sequence alignment algorithms are complex, they are generally derived from models that penalize gaps with a linear or affine cost. As a result, they often interlace gaps with short, aligned segments rather than a single long gap. However, when the sequences are different isoforms of a single gene, their alignment will necessarily contain contiguous exon-sized gaps. The same is true when aligning isoforms of diverged orthologs, though the relationship between their exons may be complex.

The most popular aligners for protein sequences (Clustal Omega, MAFFT, MUSCLE, T-Coffee) do not include splice sites in their alignments, which makes them prone to aligning non-homologous exons \cite{Sievers2017, Katoh2013, Edgar2004, Notredame2000}. Current algorithms can easily be extended to incorporate splice sites by coding them as a new symbol and preventing alignment between splice sites and amino acids, which was recently implemented in the aligner, \textit{Mirage} \cite{Nord2018}. The biggest challenge in practice, however, is mapping a protein sequence to its genomic sequence to identify splice sites in the protein sequence. Although this information can in principle be derived from the GTF annotation files produced by the NCBI pipeline, annotating the splice junctions in the protein sequences themselves would facilitate splice-sensitive alignment.

These improvements would enhance rather than replace the phylo-HMM trimming method developed in this work. The model could easily be extended to include a state that outputs a splice symbol before transitioning to one of the states in the current architecture. This intermediate state would increase the accuracy of state inference since a splice symbol followed by a phylogenetically discordant gap pattern would strongly signal a state 2 region. The association between state 2 and skipped exons can be made explicit by requiring that transitions to and from state 2 first proceed through the splice state. This of course depends on proper labeling of the training data, which would be trivial since the boundaries between exons would be marked by splice site symbols rather than inferred from gap patterns. Unfortunately, this would not allow the phylo-HMM to label extended exon boundaries as state 2 since they would not be bounded by splice symbols to the left and right. Accordingly, the phylo-HMM would need to permit transitions between state 3 and any other state to accommodate more complex splice variants and other annotation errors. Thus, this extended phylo-HMM would combine the strengths of splice-sensitive alignment with the more heuristic approach used here. Since state 2 inferences would necessarily correspond to skipped exons, they would be suitable for analyses of this form of alternative splicing. Though state 3 inferences would not directly correspond to specific biological process, they still have value as spatially and phylogenetically aware labels for trimming poorly supported segments from alignments.

The phylo-HMM could be further enhanced by expanding its emission distribution to include more symbols in the amino acid alphabet. This would allow it to better model observed substitution patterns between specific symbols, for example the high rate of exchange between gaps and glutamine residues caused by polyglutamine tracts. The transitions between amino acids could be parametrized with a published matrix such as LG \cite{Le2008}, but the transitions between amino acids and gaps would be inferred from labeled data. It is unclear if the resulting gain in accuracy would justify the increased computational burden, however.

Though there are benchmarks available for optimizing and comparing methods of orthology inference, the metrics are calculated over a set of reference genomes which are sparsely sampled over a broad taxonomic range, so it is unclear if they are informative for method designed to yield robust inferences when the genomes are highly related \cite{Nevers2022}. Furthermore, the heterogeneity of genome architectures and annotations may require quality assurance methods tailored to each set of genomes. Thus, there is likely no one-size-fits-all approach to orthology inference, and with many other standalone programs available for more standard use cases (Hieranoid, OMA standalone, OrthoFinder, OrthoInspector, Orthologer), we have chosen not to package the code into an end-to-end pipeline \cite{Kaduk2017, Altenhoff2019, Emms2019, Linard2014, Zdobnov2020}. Instead, we have devoted considerable attention to organizing and documenting the code to make it accessible to a newcomer and thereby facilitate the adaptation of specific steps to similar projects as needed.

In contrast, though many HMM packages are available for the Python programming language, we found none were satisfactorily documented or contained tutorials to introduce HMMs and their APIs to a wide audience. We therefore refactored this code into a package available on PyPI and GitHub called Homomorph. The package itself only implements standard HMM algorithms, but the GitHub repository includes tutorials that introduce the API and implement training routines. Similar tutorials for machine learning libraries such as TensorFlow have undoubtedly fueled the application of neural networks across diverse fields, but HMMs are also powerful models that can be more appropriate when the data obey certain statistical or structural constraints. Thus, we hope this package and its accompanying tutorials will serve as an on-ramp to HMMs and spur their greater adoption by non-specialists.

Though databases of orthologous groups such as COGs, Ensembl Compara, EggNOG, OMA, OrthoDB, OrthoInspector, and OrthoMCL will continue to be useful for comparative studies across broad taxonomic ranges, the increasing speed at which high-quality genome assemblies and annotations are produced means no single database can encompass the most complete data \cite{Galperin2020, Herrero2016, HuertaCepas2018, Altenhoff2020, Zdobnov2020, Nevers2018, Chen2006}. Furthermore, since many early comparative genomics studies spanned diverse branches of the tree of life, future research will likely prioritize taxonomic depth over breadth. Thus, custom sets of orthologous groups will grow more and not less common. Despite the challenges these developments pose, they also present new opportunities to bridge the gap between mutational and macroevolutionary processes.

\section*{Materials and methods}
\subsection*{Sequence de-duplication and BLAST search parameters}
The protein sequences for each annotation were de-duplicated by removing any sequences which had already appeared in association with the same gene. Thus, the first accession associated with a sequence and gene pair is the sequence’s representative accession for the gene. BLAST+ 2.13.0 was used for the sequence similarity searches \cite{Camacho2009}. An E-value cutoff of 1 was used for the initial searches. However, this cutoff was lowered to 1E-10 during processing of the BLAST output.

\subsection*{Extraction of HSPs from BLAST output}
To reduce the computational burden of merging HSPs into hits, the BLAST output was filtered to extract HSPs associated with the highest-scoring gene. The HSPs were first grouped by target protein, and the resulting groups were sorted in descending order by the bit score of their highest-scoring HSP. Iterating over the groups, all the HSPs in a group were passed to the next step until the parent gene of the group was not the parent gene of the highest ranked group. This method collects all candidate HSPs for a target gene if the highest-scoring HSP within a group exceeds the highest-scoring HSP of the next best gene. This is in some senses an extension of the best hit criterion where hits are considered at the level of genes rather than proteins. If multiple genes tied for the highest-scoring HSP, the iteration stopped when the parent gene of the current group matched none of these highest-scoring genes.

\subsection*{Merging of HSPs into hits}
HSPs were merged in two stages where the first combined non-overlapping HSPs, and the second combined the remaining HSPs. In the first stage, proceeding from highest to lowest bit score, HSPs were marked as “disjoint” if they do not overlap with any other HSP previously marked as disjoint. Although this greedy strategy does not necessarily yield the highest-scoring set of disjoint HSPs, it prioritized higher scoring HSPs. In the second stage, all disjoint HSPs were marked as “compatible,” and proceeding from highest to lowest bit score the remaining HSPs were marked as compatible if the overlap with any other compatible HSP was no more than 50\% of the length of either. The best hit for each query was chosen as the hit with the highest sum of bit scores from disjoint HSPs. The best hits were then filtered by overlap and reciprocity criteria. The overlap criterion was applied first and requires that 50\% of residues in the query are aligned in compatible HSPs. This excludes false positives from conserved domains embedded in larger non-homologous proteins by ensuring the hits span a sufficient fraction of the query and target sequences. The reciprocity criterion requires each query-target pair has a corresponding hit where the roles are reversed, which ensures there is no ambiguity in which target is the best match for the query.

\subsection*{Clustering by \textit{k}-clique percolation}
\textit{k}-clique percolation was implemented in two steps. In the first, maximal cliques were identified. In the second, a percolation graph is constructed by defining edges between cliques if they have \textit{k}-1 nodes in common. Clusters are the connected components of this second graph. The first step uses the NetworkX implementation of a maximal clique algorithm. The second step uses a modification of the NetworkX implementation of the \textit{k}-clique community algorithm. The NetworkX implementation exhaustively finds all edges in the percolation graph. Since joining a \textit{k}-clique community only requires that a clique has a single edge connecting it to that a community, this approach is needlessly expensive for large graphs. The custom implementation instead uses a progressive approach where each clique is checked against a list of known communities, merging communities as necessary in each step.

The hit graph is sparse, so these algorithms are efficient when applied to its individual connected components. However, some components have a structure with many maximal cliques, which dramatically slows the first or second step of the clique percolation algorithm. Thus, if either step exceeded 90 s, the process timed out, and the simpler \textit{k}-core algorithm was used instead. Out of over 10,000 connected components, only seven timed out, and many of those contained highly dense clusters of histone sequences.

\subsection*{Addition of paralogs to orthologous groups}
The protein sequences for each annotation were searched against themselves with the same settings as for the inter-genome searches. The resulting output was processed identically except the HSPs were not filtered using the best gene criterion. Thus, all HSPs for each query were merged into hits. The best hit for each query and target gene was then chosen as the hit with the highest sum of bit scores from disjoint HSPs. (Grouping by target gene ensured only the highest-scoring isoform was selected.) Query-target pairs whose hits which exceeded the maximum bit score for all inter-genome hits associated with that query and passed the overlap and reciprocity filters were designated as paralogous pairs. The orthologous groups were supplemented with paralogs by adding the paired sequences for each of the original members of the orthologous group.

\subsection*{Initial alignment and selection of representative sequences}
The sequences in each orthologous group were aligned using MAFFT 7.490 with the following settings: --globalpair --maxiterate 1000 --thread 1 --anysymbol --allowshift --unalignlevel 0.4 --leavegappyregion \cite{Katoh2013}. Representative sequences for each gene were then selected by maximum likelihood according to binary profiles constructed from these alignments. First each sequence was coded into gap and non-gap symbols. The sequences were then grouped by gene, and for each group and position if at least one sequence was aligned in the group, the group contributed one count for the non-gap symbol to the profile at that position. Otherwise, the group contributed a count for the gap symbol at that position. To account for the phylogenetic dependencies between sequences, the counts were weighted according to a Gaussian process over the GTR2 consensus tree described in the section on inferring species trees \cite{Altschul1989}. If a species had multiple genes in the orthologous group, the species weight was divided evenly among them. Each coded sequence was then scored according to this profile, and the maximum likelihood sequence for each gene was selected as its representative. By assigning a non-gap count to groups and positions where at least one sequence is aligned, the profile incentivizes the selection of sequences contain the fewest gaps and match the consensus alignment. Unfortunately, this scheme can cause a sequence to score negative infinity if it has a gap at a position where every group has at least one aligned sequence, so the profile was initialized with a pseudocount of 0.005 for the gap and non-gap symbols at each position.

\subsection*{Alignment refinement}
Consensus columns were manually defined as columns in the initial alignments with at least five non-gap symbols using the Stockholm file format. These labeled alignments were used to build a profile HMM using HMMer 3.3.2 \cite{Eddy2009}. No sequence relative weighting was used, and the effective number of sequences was set to three times the number of sequences in the alignment. HMMer uses strong priors to allow the construction of profiles with relatively few aligned sequences. However, in rare cases a sequence with a long poorly supported segment was misaligned, effectively putting it out of register with the profile and creating a poor-quality alignment. Increasing the effective sequence number corrected this error. The original sequences were then aligned to the profile. Unaligned segments at the N- or C- terminus were removed from the alignment. Unaligned segments in internal regions were extracted and aligned separately for each region using MAFFT with the same settings. The resulting alignments were then “stitched” into the profile alignment. This process was unsuccessful for one alignment due to cryptic error in MAFFT, which reduced the number of refined alignments to 8,565.

This method corrects most cases of misalignment since poorly supported segments are typically aligned to fewer than four other sequences. However, it does not remove poorly supported segments from columns with more than five non-gap symbols, and their contribution is sometimes sufficient to reproduce the misalignment when the sequences are aligned to the profile HMM. A lightweight convolutional neural network was therefore trained to recognize poorly supported segments to more completely remove them from the profile HMM. The network consisted of a two-dimensional embedding from a 22-dimensional one-hot-encoded vector (20 amino acids, an ambiguous amino acid, and a gap), two convolutional layers, and a one-dimensional output. The first layer was composed of four filters. Two of these filters had a kernel size of 10 and a dilation of 1. The other two filters had a kernel size of 60 with a dilation of 3. The second layer was composed of two filters with a kernel size of 10 and a dilation of 1 (Fig. \ref{sfig:cnn_architecture}). The network was trained on 88,513 hand-labeled positions across 122 sequences in 46 unique alignments. These examples were largely long segments aligned to few, if any, other sequences (Fig. \ref{sfig:cnn_training}).

The trained network outputs a probability that each position in an aligned sequence is part of a long poorly supported segment. A simple algorithm was used to detect positions associated with high confidence long poorly supported segments even if the output at that position was relatively low. First, “seeds” were identified at positions where the output exceeded 0.75. These seeds were expanded outwards until the output was below 0.05. A seed and its expansion composed a long poorly supported segment. Consensus columns were defined as columns with at least five non-gap symbols, excluding positions labeled as long poorly supported segments. To maintain the contribution of long poorly supported segments to the profile’s models of insertions, long poorly supported segments were replaced with gaps only at positions aligned to consensus columns.

\subsection*{Alignment curation}
The alignments were coded into gap and non-gap symbols to simplify the emission distributions. The ``insertion'' phylo-HMM was composed of the four hidden states described in the main text. The emission distributions for each consisted of two components which modeled the gap pattern and the propensity for those patterns to remain constant (``gap stickiness''), respectively. The first component was a two-state Markov process which was in turn composed of two subprocess. The first was a phylogenetic process on the on the GTR2 consensus tree described in the section on inferring species trees, and the second was jump process at the tips. The second component was a beta-Bernoulli distribution on the number of symbols which were constant between subsequent columns. The ``missing data'' phylo-HMM was composed of two hidden states which were both parameterized with the same two-state, two component Markov process as the insertion phylo-HMM. However, the emission probabilities were calculated as the posterior probability of the observed symbol given the data rather than the probability of the data. Only the alignments of the single copy orthologous groups were curated, so each tip in the species tree would correspond to a single sequence in the alignment.

All HMM algorithms were implemented with custom code which is available as the package Homomorph on PyPI. The insertion phylo-HMM was trained on 39,298 manually labeled columns in 11 alignments, and the deletion phylo-HMM was trained on 58,655 manually labeled positions in 20 sequences in 9 unique alignments (Fig. \ref{sfig:hmm_training}). Because maximum-likelihood estimation of the model parameters yielded posterior decoding curves which toggled between hidden states too rapidly, the models were instead trained discriminatively \cite{Krogh1999}. The difference, briefly, is maximum-likelihood estimation finds the parameters that best reproduce the observed distributions whereas discriminative training finds the parameters that minimize prediction error. Discriminatively trained models typically perform better in practice since real-world data are rarely fully described by the distribution specified by the model.

The posterior distributions over states were computed for both the insertion and missing data phylo-HMMs. To trim the alignments, the posterior distributions from the insertion phylo-HMM were converted to state assignments with the following method. First the probabilities for states 2 and 3 were combined. Since this probability can change rapidly or gradually depending on the local context, a simple cutoff would not necessarily define the boundaries as the columns where the gap pattern changes most abruptly. Instead, a high cutoff of 0.9 was used to define a “seed” region. The boundaries of the seed were then expanded both inwards and outwards to define intervals from which to select boundaries. The outward expansion halted when the probability or its derivative was below 0.05 and 0.001, respectively. The inward expansion halted when the derivative was below 0.02. The left and right boundaries were chosen as the columns in each expansion with the maximum product between the derivative and the number of sequences which changed to or from a gap between columns. By combining where the model’s confidence changed rapidly with the observed change in the gap pattern, this method generally selected reasonable boundaries. In contrast, the posterior probabilities from the missing data phylo-HMM changed sharply, so positions were labeled as the ``missing data'' state if the posterior probability exceeded 0.95. These assignments can be used to filter segments or entire sequences from downstream analyses and are available in the supplementary data.

\subsection*{Inference of species trees}
Since the roots of phylogenetic trees are not identifiable with commonly used time-reversible substitution models, the orthology inference pipeline was repeated with the outgroup species \textit{Scaptodrosophila lebanonensis}. Orthologous groups with one sequence for each species were aligned, and 100 meta-alignments were constructed by randomly sampling 10,000 columns from these 9,435 alignments. (The alignments were not refined with HMMer before sampling.) To determine the effect of invariant columns and gaps, two sampling strategies were used where invariant columns were allowed or disallowed and the maximum fraction of gaps was set at 0, 50, and 100\%. Their combination yielded six different sets of meta-alignments. A tree was fit to each meta-alignment with the LG substitution model and four discrete gamma rate categories using IQ-TREE 1.6.12 \cite{Nguyen2014, Le2008, Yang1994}. If the sampling strategy allowed invariant columns, an invariant rate category was included. The resulting trees from each set were then merged into a majority consensus tree (Fig. \ref{sfig:trees_LG}).

To fit trees using the GTR model of nucleotide substitution, the protein alignments were converted to nucleotide alignments using the corresponding coding sequences in the genome annotations. Some protein sequences are “low quality,” meaning their coding sequences contain frameshifts, premature stop codons, or other errors even though they are strong hits to known protein-coding genes. The NCBI annotation pipeline corrects some of these defects in the protein sequences, which can complicate a simple “reverse translation” of the alignment. After rejecting alignments where the expected translation from a coding sequence differed from its corresponding protein sequence, 3,425 alignments remained. Consensus tree were derived from meta-alignments sampled from these alignments using the approach described previously except the GTR model was used in place of the LG model.

To fit trees using the two-state GTR model of substitution, the protein alignments were first coded into gap or non-gap symbols. As before, 100 meta-alignments were constructed from these coded alignments for each sampling strategy. In this case, only the presence of invariant columns was varied, yielding two sets of meta-alignments. Trees were then fit using the GTR2 model with no rate categories. An invariant category, however, was included if invariant columns were allowed. Since the bootstrap confidences were sometimes lower than 50\%, the resulting trees from each set were then merged into a loose consensus tree to prevent multifurcations (Fig. \ref{sfig:trees_GTR}).

\subsection*{Code and data availability}
The code used to produce the results and analyses is available at \url{https://github.com/marcsingleton/orthology_inference2023}. The following Python libraries were used: matplotlib, NumPy, pandas, SciPy, and TensorFlow \cite{Hunter2007, Harris2020, McKinney2010, Virtanen2020, Abadi2015}. Relevant output files are available in the supporting information. There is no primary data associated with this manuscript. All primary data are available from publicly accessible sources described in their corresponding sections.

\section*{Acknowledgments}
We thank the past and present members of the Eisen lab for helpful discussions and their assistance in revising the manuscript.

\printbibliography

\clearpage

\section*{Supporting information}
\setcounter{figure}{0}
\renewcommand{\thefigure}{S\arabic{figure}}

\begin{figure}[h!]
\includegraphics[width=0.5\textwidth]{ncbi/out/ncbi.png}
\centering
\caption{\textbf{Cumulative number of different eukaryotic genomes annotated by NCBI.}}
\label{sfig:ncbi}
\end{figure}

\begin{figure}[h!]
\includegraphics[width=\textwidth]{clusters/out/clusters.png}
\centering
\caption{\textbf{Statistics of orthologous groups.}
A) Each species is equally represented in orthologous groups (OGs). B) A plurality of orthologous groups contain all species. C) Nearly all proteins are associated with a single orthologous group. D) The number of orthologous groups associated with a species is strongly correlated with the number of unique annotated proteins, which suggests the annotation pipeline generally identifies conserved genes.}
\label{sfig:clusters}
\end{figure}

\begin{figure}[h!]
\includegraphics[width=\textwidth]{paralogs/out/paralogs.png}
\centering
\caption{\textbf{Addition of paralogs to orthologous groups.}
A) Most orthologous groups (OGs) have no in-paralogs. B, D) Of the groups with paralogs, most have fewer than five. C) The in-paralogs are generally only a small fraction of the sequences in an orthologous group.}
\label{sfig:paralogs}
\end{figure}

\begin{figure}[h!]
\includegraphics[width=\textwidth]{cnn_architecture.png}
\centering
\caption{\textbf{CNN architecture.}}
\label{sfig:cnn_architecture}
\end{figure}

\begin{figure}[h!]
\includegraphics[width=\textwidth]{cnn_training/out/cnn_training.png}
\includegraphics[width=\textwidth]{cnn_training/out/23D9_XP_026832050.1.png}
\centering
\caption{\textbf{CNN data and training details.}
A) Most labels were “negative,” which corresponded to positions which were maintained in the alignment. B, C) The training loss and metrics stabilized by the final training epochs. D, E) The parameters for a model of this size are easily visualized. The embedding learns the gap and unknown amino acid symbols are distinct from the others. The convolutional filters learn kernels which correspond to the length of a typical long poorly supported segment. F) An example alignment where the indicated sequence is labeled with the trained CNN. The blue lines are the CNN’s predictions, and the yellow lines are the labels. The red lines are the weights given to each position. The regions where the weights are zero are masked from the training data, which allowed the model to focus on only the most relevant examples. }
\label{sfig:cnn_training}
\end{figure}

\begin{figure}[h!]
\includegraphics[width=\textwidth]{hmm_training/out/hmm_training.png}
\centering
\caption{\textbf{Insertion phylo-HMM data and training details.}
A) Most columns in the training data were labeled as state 1A or 1B. B-F) The model loss and parameters largely stabilized by the final training iteration. Panels C-F correspond to the parameters in the phylogenetic process, the jump process, the pattern stickiness model, and the transition matrix, respectively. The transition matrix plots are the transition rates to the state indicated on the vertical axis and given in log scale. Self transitions are excluded.}
\label{sfig:hmm_training}
\end{figure}

\begin{figure}[h!]
\includegraphics[width=\textwidth]{hmm_trim/out/hmm_trim.png}
\centering
\caption{\textbf{Insertion phylo-HMM trimming details.}
A) Most alignments were not trimmed of any columns. B) Most alignments have fewer than 5 trimmed regions which cumulatively account for less than 10\% of the columns in the original alignment. Alignments with more trimmed columns generally have one one large trimmed region rather than many small ones. C) Most trimmed regions were inferred as state 3.}
\label{sfig:hmm_trim}
\end{figure}

\begin{figure}[h!]
\includegraphics[width=\textwidth]{trees/out/trees2_LG.png}
\centering
\caption{\textbf{Phylogenetic trees created by different sampling strategies under LG model.}}
\label{sfig:trees_LG}
\end{figure}

\begin{figure}[h!]
\includegraphics[width=\textwidth]{trees/out/trees2_GTR.png}
\centering
\caption{\textbf{Phylogenetic trees created by different sampling strategies under GTR model.}}
\label{sfig:trees_GTR}
\end{figure}

\setcounter{table}{0}
\renewcommand{\thetable}{S\arabic{table}}

\begin{table}[h!]
\centering
\caption{\textbf{Genome annotations.}}
\begin{tabular}{|l|l|l|l|l|}
\hline
\textbf{Species}                       & \textbf{Species ID} & \textbf{Taxon ID} & \textbf{Version} & \textbf{Source} \\ \hline
\textit{Drosophila ananassae}          & dana                & 7217              & 102              & NCBI            \\ \hline
\textit{Drosophila biarmipes}          & dbia                & 125945            & 102              & NCBI            \\ \hline
\textit{Drosophila bipectinata}        & dbip                & 42026             & 102              & NCBI            \\ \hline
\textit{Drosophila elegans}            & dele                & 30023             & 102              & NCBI            \\ \hline
\textit{Drosophila erecta}             & dere                & 7220              & 101              & NCBI            \\ \hline
\textit{Drosophila eugracilis}         & deug                & 29029             & 102              & NCBI            \\ \hline
\textit{Drosophila ficusphila}         & dfic                & 30025             & 102              & NCBI            \\ \hline
\textit{Drosophila grimshawi}          & dgri                & 7222              & 103              & NCBI            \\ \hline
\textit{Drosophila guanche}            & dgua                & 7266              & 100              & NCBI            \\ \hline
\textit{Drosophila hydei}              & dhyd                & 7224              & 101              & NCBI            \\ \hline
\textit{Drosophila innubila}           & dinn                & 198719            & 100              & NCBI            \\ \hline
\textit{Drosophila kikkawai}           & dkik                & 30033             & 102              & NCBI            \\ \hline
\textit{Drosophila mauritiana}         & dmau                & 7226              & 100              & NCBI            \\ \hline
\textit{Drosophila melanogaster}       & dmel                & 7227              & FB2022\_02       & FlyBase         \\ \hline
\textit{Drosophila mojavensis}         & dmoj                & 7230              & 102              & NCBI            \\ \hline
\textit{Drosophila navojoa}            & dnav                & 7232              & 101              & NCBI            \\ \hline
\textit{Drosophila novamexicana}       & dnov                & 47314             & 100              & NCBI            \\ \hline
\textit{Drosophila obscura}            & dobs                & 7282              & 101              & NCBI            \\ \hline
\textit{Drosophila persimilis}         & dper                & 7234              & 101              & NCBI            \\ \hline
\textit{Drosophila pseudoobscura}      & dpse                & 7237              & 104              & NCBI            \\ \hline
\textit{Drosophila rhopaloa}           & drho                & 1041015           & 102              & NCBI            \\ \hline
\textit{Drosophila santomea}           & dsan                & 129105            & 101              & NCBI            \\ \hline
\textit{Drosophila sechellia}          & dsec                & 7238              & 101              & NCBI            \\ \hline
\textit{Drosophila serrata}            & dser                & 7274              & 100              & NCBI            \\ \hline
\textit{Drosophila simulans}           & dsim                & 7240              & 103              & NCBI            \\ \hline
\textit{Drosophila subobscura}         & dsob                & 7241              & 100              & NCBI            \\ \hline
\textit{Drosophila subpulchrella}      & dspu                & 1486046           & 100              & NCBI            \\ \hline
\textit{Drosophila suzukii}            & dsuz                & 28584             & 102              & NCBI            \\ \hline
\textit{Drosophila takahashii}         & dtak                & 29030             & 102              & NCBI            \\ \hline
\textit{Drosophila teissieri}          & dtei                & 7243              & 100              & NCBI            \\ \hline
\textit{Drosophila virilis}            & dvir                & 7244              & 103              & NCBI            \\ \hline
\textit{Drosophila willistoni}         & dwil                & 7260              & 102              & NCBI            \\ \hline
\textit{Drosophila yakuba}             & dyak                & 7245              & 102              & NCBI            \\ \hline
\textit{Scaptodrosophila lebanonensis} & sleb                & 7225              & 100              & NCBI            \\ \hline
\end{tabular}
\label{stable:genomes}
\end{table}

\begin{table}[h!]
\centering
\caption{\textbf{Phylogenetic diversity criteria.}}
\begin{tabular}{|l|l|}
\hline
\textbf{Species   IDs} & \textbf{Minimum   number} \\ \hline
dinn,   dgri, dhyd     & 2                         \\ \hline
dnov,   dvir           & 1                         \\ \hline
dmoj,   dnav           & 1                         \\ \hline
dper,   dpse           & 1                         \\ \hline
dgua,   dsob           & 1                         \\ \hline
dana,   dbip           & 1                         \\ \hline
dkik,   dser           & 1                         \\ \hline
dele,   drho           & 1                         \\ \hline
dtak,   dbia           & 1                         \\ \hline
dspu,   dsuz           & 1                         \\ \hline
dere,   dtei           & 1                         \\ \hline
dsan,   dyak           & 1                         \\ \hline
dmel                   & 1                         \\ \hline
dmau,   dsec, dsim     & 1                         \\ \hline
\end{tabular}
\label{stable:diversity_criteria}
\end{table}

\nolinenumbers

\end{document}

### 2022-11-10 chapter1.tex
Removed line numbers and added asterisk to corresponding author
Corrected some minor typos from Colleen and Julia

@@
 \usepackage[margin=1in]{geometry}
-\usepackage[right]{lineno}  % line numbers
 %\usepackage{setspace}  % double spacing
@@
 \begin{document}
-\linenumbers
 
@@
 Marc Singleton\textsuperscript{1},
-Michael Eisen\textsuperscript{1,2,3}
+Michael Eisen\textsuperscript{1,2,3,*}
 \\
@@
 \section*{Introduction}
-Comparative genomics is powerful tool for yielding insights into evolutionary relationships, molecular function, and the forces that drive gene, genome, and population evolution.
-These methods often rely on the identification of homologous sequences or homologs, that is sequences with common ancestry, since this ensures that differences between sequences reflects variations in evolution from a common point of divergence.
+Comparative genomics is a powerful tool for yielding insights into evolutionary relationships, molecular function, and the forces that drive gene, genome, and population evolution.
+These methods often rely on the identification of homologous sequences or homologs, that is sequences with common ancestry, since this ensures that differences between sequences reflect variations in evolution from a common point of divergence.
 In a sense, each branch that descends from a single ancestral population is a different ``experiment'' by nature, as each is free to explore the evolutionary landscape independent of the others.
 However, many analyses impose the additional condition that the sequences have diverged through speciation events (orthology) rather than duplications (paralogy) or other mechanisms such as horizontal gene transfer \cite{Fitch1970}.
-The underlying assumption is ``orthologs'' have conserved and equivalent functions whereas ``paralogs'', by virtue of their redundancy, are more likely to diverge \cite{Ohno1970, Nowak1997, Altenhoff2012, Pegueroles2013, Soria2014}.\footnote{For multiple sequences, orthology is usually defined relative to their most recent common ancestor.
+The underlying assumption is ``orthologs'' have conserved equivalent functions whereas ``paralogs'', by virtue of their redundancy, are more likely to diverge \cite{Ohno1970, Nowak1997, Altenhoff2012, Pegueroles2013, Soria2014}.\footnote{For multiple sequences, orthology is usually defined relative to their most recent common ancestor.
 This technically includes sequences which split by duplication after this point (``in-paralogs''), but excludes sequences which split by duplication before (``out-paralogs'') \cite{Remm2001}.
@@
 In contrast, columns assigned to states 2 and 3 largely corresponded to the long poorly supported segments and phylogenetically discordant regions discussed previously and were therefore removed from the alignments.
-Fewer than 10\% of alignments were trimmed, but when regions were trimmed, they were largely inferred as state 3 \ref{sfig:hmm_trim}.
+Fewer than 10\% of alignments were trimmed, but when regions were trimmed, they were largely inferred as state 3 (Fig.
+\ref{sfig:hmm_trim}).
 
@@
 As a result, they often interlace gaps with short, aligned segments rather than a single long gap.
-However, when the sequences are different isoforms of a single gene, their alignment will necessarily contain contiguous exon-sized gaps.
+However, when the sequences are different isoforms of a single gene, their alignment will necessarily contain contiguous exon sized gaps.
 The same is true when aligning isoforms of diverged orthologs, though the relationship between their exons may be complex.
@@
 
-\nolinenumbers
-
 \end{document}

### 2022-11-10 chapter1.tex
Added non-breaking spaces to citations and references

@@
 In a sense, each branch that descends from a single ancestral population is a different ``experiment'' by nature, as each is free to explore the evolutionary landscape independent of the others.
-However, many analyses impose the additional condition that the sequences have diverged through speciation events (orthology) rather than duplications (paralogy) or other mechanisms such as horizontal gene transfer \cite{Fitch1970}.
-The underlying assumption is ``orthologs'' have conserved equivalent functions whereas ``paralogs'', by virtue of their redundancy, are more likely to diverge \cite{Ohno1970, Nowak1997, Altenhoff2012, Pegueroles2013, Soria2014}.\footnote{For multiple sequences, orthology is usually defined relative to their most recent common ancestor.
+However, many analyses impose the additional condition that the sequences have diverged through speciation events (orthology) rather than duplications (paralogy) or other mechanisms such as horizontal gene transfer~\cite{Fitch1970}.
+The underlying assumption is ``orthologs'' have conserved equivalent functions whereas ``paralogs'', by virtue of their redundancy, are more likely to diverge~\cite{Ohno1970, Nowak1997, Altenhoff2012, Pegueroles2013, Soria2014}.\footnote{For multiple sequences, orthology is usually defined relative to their most recent common ancestor.
 This technically includes sequences which split by duplication after this point (``in-paralogs''), but excludes sequences which split by duplication before (``out-paralogs'') \cite{Remm2001}.
@@
 
-Given this importance, methods for inferring orthologous groups of proteins were developed shortly after the first genomes were sequenced in the late 1990s \cite{Fleischmann1995, Goffeau1996, CESC1998}.
-One early and influential approach was to cluster triangles of hits resulting from homology searches between pairs of genomes \cite{Tatusov1997}.
+Given this importance, methods for inferring orthologous groups of proteins were developed shortly after the first genomes were sequenced in the late 1990s~\cite{Fleischmann1995, Goffeau1996, CESC1998}.
+One early and influential approach was to cluster triangles of hits resulting from homology searches between pairs of genomes~\cite{Tatusov1997}.
 Graph-based approaches have remained popular, and in the intervening years many other researchers have refined this method by implementing various pre- and post-processing steps.
-Despite these improvements, many databases and pipelines use the same triangle clustering algorithm or other methods which require relatively few hits between sequences to infer an orthologous group, \textit{e.g.} connected components or Markov clustering \cite{Remm2001, Li2003, Jensen2007, Linard2011, Emms2015, Train2017, Cosentino2018}.
+Despite these improvements, many databases and pipelines use the same triangle clustering algorithm or other methods which require relatively few hits between sequences to infer an orthologous group, \textit{e.g.} connected components or Markov clustering~\cite{Remm2001, Li2003, Jensen2007, Linard2011, Emms2015, Train2017, Cosentino2018}.
 However, the scale of biological sequence data has changed dramatically.
-For example, in the last decade, the number of annotated genomes available from NCBI has increased nearly 20-fold and currently exceeds 900 (Fig.
-\ref{sfig:ncbi}).
-Though this figure is only a rough proxy of the total number of assemblies available, it will likely continue to grow rapidly in the coming years as many large-scale genome assembly efforts such as i5K, the Bird 10,000 Genomes Project, and the Vertebrate Genomes Project have already yielded results \cite{Thomas2020, Feng2020, Rhie2021}.
+For example, in the last decade, the number of annotated genomes available from NCBI has increased nearly 20-fold and currently exceeds 900 (Fig.~\ref{sfig:ncbi}).
+Though this figure is only a rough proxy of the total number of assemblies available, it will likely continue to grow rapidly in the coming years as many large-scale genome assembly efforts such as i5K, the Bird 10,000 Genomes Project, and the Vertebrate Genomes Project have already yielded results~\cite{Thomas2020, Feng2020, Rhie2021}.
 Thus, the dense taxonomic sampling made possible by these projects poses new challenges and opportunities for the standard methods of orthology inference and alignment, which implicitly assume fewer and more distantly related genomes or fail to fully leverage the redundancy and organization of their annotations.
 
-Thus, in this work we developed a computational pipeline that can robustly infer and align orthologous groups of proteins even when the genomes are highly redundant (Fig.
-\ref{fig:graphical_overview}).
+Thus, in this work we developed a computational pipeline that can robustly infer and align orthologous groups of proteins even when the genomes are highly redundant (Fig.~\ref{fig:graphical_overview}).
 We did not seek to reinvent the wheel, so the overall principle of clustering a graph of hits from homology searches is unchanged, though many details were modified to maximize the detection of highly diverged orthologs while also minimizing the impact of incomplete or incorrect annotations.
@@
 However, most of our efforts were instead focused on the final step of aligning the orthologous sequences.
-Though genome annotation pipelines are often proficient at identifying the locus of genes, the quality of their gene models can vary considerably and can contain inaccurate exon boundaries or start codons \cite{Frankish2015, Dunne2018}.
+Though genome annotation pipelines are often proficient at identifying the locus of genes, the quality of their gene models can vary considerably and can contain inaccurate exon boundaries or start codons~\cite{Frankish2015, Dunne2018}.
 As a result, the structure and lengths of the protein sequences that result from these models can be highly heterogeneous, which poses many challenges for their alignment and subsequent analyses.
@@
 To develop these methods, we chose a set of 33 assembled and annotated \textit{Drosophila} genomes.
-All 12 species from the original \textit{Drosophila} 12 Genomes Consortium are included \cite{D12GC2007}.
+All 12 species from the original \textit{Drosophila} 12 Genomes Consortium are included~\cite{D12GC2007}.
 However, all have been re-sequenced and re-annotated, many with long-read sequencing technology, which has resulted in substantial improvements in the contiguity of their assemblies.
-Despite this and other genome assembly projects of species in the \textit{Drosophila} genus, there is not yet a database of high-quality alignments of orthologous proteins that reflects these improvements in genome assembly and diversity \cite{Miller2018, Kim2021}.
-Given the \textit{Drosophila} genus spans diverse habitats and over 50 million years of evolution but maintains a conserved life cycle and body plan, such a resource would facilitate a new generation of studies that illuminate the forces that drive protein evolution in unprecedented detail \cite{Wiegmann2011, Obbard2012}.
+Despite this and other genome assembly projects of species in the \textit{Drosophila} genus, there is not yet a database of high-quality alignments of orthologous proteins that reflects these improvements in genome assembly and diversity~\cite{Miller2018, Kim2021}.
+Given the \textit{Drosophila} genus spans diverse habitats and over 50 million years of evolution but maintains a conserved life cycle and body plan, such a resource would facilitate a new generation of studies that illuminate the forces that drive protein evolution in unprecedented detail~\cite{Wiegmann2011, Obbard2012}.
 
@@
 All annotated genomes in the genus \textit{Drosophila} available in April 2022 were downloaded from NCBI's RefSeq database.
-The assemblies annotated by the NCBI eukaryotic genome annotation pipeline must past several quality checks and have supporting transcript evidence, so the annotations are generally highly complete \cite{Thibaud-Nissen2013}.
+The assemblies annotated by the NCBI eukaryotic genome annotation pipeline must past several quality checks and have supporting transcript evidence, so the annotations are generally highly complete~\cite{Thibaud-Nissen2013}.
 The \textit{D.
-miranda} annotation was excluded due to its unusual karyotype \cite{Dobzhansky1935}.
+miranda} annotation was excluded due to its unusual karyotype~\cite{Dobzhansky1935}.
 Other annotations were excluded after preliminary clustering showed a deficient in the number of orthologous groups containing those genomes, indicating their annotations were less complete (data not shown).
 The \textit{D.
-melanogaster} annotation was downloaded from FlyBase \cite{Gramates2022}.
-The complete list of genomes is given in Table \ref{stable:genomes}.
+melanogaster} annotation was downloaded from FlyBase~\cite{Gramates2022}.
+The complete list of genomes is given in Table~\ref{stable:genomes}.
 
@@
 \subsection*{Extraction of best hits from BLAST output}
-The protein sequences in each genome annotation were searched against each other in reciprocal pairs using BLAST, yielding a list of high-scoring segment pairs (HSPs) for each query-target pair \cite{Camacho2009}.
+The protein sequences in each genome annotation were searched against each other in reciprocal pairs using BLAST, yielding a list of high-scoring segment pairs (HSPs) for each query-target pair~\cite{Camacho2009}.
 HSPs are local alignments, meaning they do not necessarily span the entire lengths of the query and target sequences.
@@
 \subsection*{Clustering in orthologous groups}
-The best hits between sequences are naturally visualized as a graph where the sequences are nodes and best hits are edges between nodes (Fig.
-\ref{fig:components}).
+The best hits between sequences are naturally visualized as a graph where the sequences are nodes and best hits are edges between nodes (Fig.~\ref{fig:components}).
 Some sets of nodes form tight clusters whereas others are more loosely associated or even isolated.
@@
 While this method is robust when clustering smaller or more diverse sets of genomes, the large number of closely related genomes in the input data creates many edges which form triangles but are otherwise poorly supported by the remaining genomes.
-However, triangle clustering is a special case of the more general \textit{k}-clique percolation algorithm where \textit{k} = 3 (Fig.
-\ref{fig:percolation}).
+However, triangle clustering is a special case of the more general \textit{k}-clique percolation algorithm where \textit{k} = 3 (Fig.~\ref{fig:percolation}).
 Higher values of \textit{k} require greater overlap between a candidate node and those already in the cluster and therefore produce tighter clusters at the cost of excluding more speculative orthology relationships.
-\textit{k} was set to 4 as compromise between these concerns, yielding 22,813 orthologous groups, a plurality of which contained all 33 species (Fig.
-\ref{sfig:clusters}).
+\textit{k} was set to 4 as compromise between these concerns, yielding 22,813 orthologous groups, a plurality of which contained all 33 species (Fig.~\ref{sfig:clusters}).
 
@@
 Orthologous groups were then supplemented with paralogs by adding the paired sequences for each of the original members of the orthologous group.
-Most orthologous groups contain no paralogs, and those that do generally have few relative to the original number of sequences in the group (Fig.
-\ref{sfig:paralogs}).
+Most orthologous groups contain no paralogs, and those that do generally have few relative to the original number of sequences in the group (Fig.~\ref{sfig:paralogs}).
 
@@
 For example, studies of gene duplication will focus on orthologous groups with paralogs in some lineages but not in others.
-In contrast, analyses which assume functional conservation should restrict the orthologous groups to single copy orthologs since paralogs more frequently undergo functional divergence \cite{Altenhoff2012, Pegueroles2013, Soria2014}.
+In contrast, analyses which assume functional conservation should restrict the orthologous groups to single copy orthologs since paralogs more frequently undergo functional divergence~\cite{Altenhoff2012, Pegueroles2013, Soria2014}.
 A simple method for identifying such groups is requiring each species to have exactly one associated gene.
 However, since the probability of at least one missing gene annotation approaches one as the total number of genomes increases, this is too restrictive and fails to leverage the redundancy of closely related genomes.
-Instead, a set of phylogenetic diversity criteria detailed in Table \ref{stable:diversity_criteria} were applied to ensure the major lineages were represented in downstream analyses.
+Instead, a set of phylogenetic diversity criteria detailed in Table~\ref{stable:diversity_criteria} were applied to ensure the major lineages were represented in downstream analyses.
 Furthermore, genome-wide analyses should select one orthologous group per each of the previously identified gene groups as to not bias the results towards genes with many distinct groups of isoforms.
@@
 
-The aligner MAFFT has a mode for addressing over-alignment with a parameter, \textit{a\textsubscript{max}}, that adjusts the strength of the correction \cite{Katoh2013, Katoh2016}.
+The aligner MAFFT has a mode for addressing over-alignment with a parameter, \textit{a\textsubscript{max}}, that adjusts the strength of the correction~\cite{Katoh2013, Katoh2016}.
 \textit{a\textsubscript{max}} varies between 0 and 1, but at values above 0.8, the correction severely degrades the alignment of conserved regions.
@@
 Though some cases of over-alignment remained, most conserved regions were reliably aligned with few to no errors.
-To correct the remaining errors in alignments of the single copy orthologs, a statistical model of each alignment called a profile HMM was constructed from the well-aligned columns \cite{Eddy2009}.
+To correct the remaining errors in alignments of the single copy orthologs, a statistical model of each alignment called a profile HMM was constructed from the well-aligned columns~\cite{Eddy2009}.
 The original sequences were then aligned to these profiles.
-By defining highly conserved “anchor” regions, this approach largely prevents the alignment of chance sequence similarities in long poorly supported segments (Fig.
-\ref{fig:realignment}).
+By defining highly conserved “anchor” regions, this approach largely prevents the alignment of chance sequence similarities in long poorly supported segments (Fig.~\ref{fig:realignment}).
 Furthermore, since profile HMMs explicitly model the beginnings and ends of alignments, they can fully exclude these segments from the termini.
@@
 Since the segments in these regions are likely the result of incorrect or incomplete annotations rather than meaningful biological variation, maintaining them in the alignments would propagate spurious homologies to subsequent analyses.
-This is a common issue in alignments generated by automated pipelines, so downstream analyses often focus on the strongly supported regions by removing or “trimming” columns below some threshold number of gaps or sequence identity \cite{Castresana2000, CapellaGutierrez2009}.
+This is a common issue in alignments generated by automated pipelines, so downstream analyses often focus on the strongly supported regions by removing or “trimming” columns below some threshold number of gaps or sequence identity~\cite{Castresana2000, CapellaGutierrez2009}.
 This approach, however, is inadequate if the taxonomic sampling is dense, as a single indel event along a lineage containing many species can increase the number of gaps above the threshold.
@@
 This phylo-HMM was composed of four hidden states, {1A, 1B, 2, 3}, roughly corresponding to highly conserved regions with few to no gaps, divergent regions, regions with a stable gap pattern discordant with the expected phylogenetic relationships, and regions with poorly supported segments.
-After the model was trained on manually labeled examples, it was used to assign a label to the columns in each alignment (Fig.
-\ref{fig:hmm}).
+After the model was trained on manually labeled examples, it was used to assign a label to the columns in each alignment (Fig.~\ref{fig:hmm}).
 Columns assigned to states 1A and 1B are the regions of interest for downstream analyses since the gaps generally follow the expected pattern given the phylogenetic tree.
 In contrast, columns assigned to states 2 and 3 largely corresponded to the long poorly supported segments and phylogenetically discordant regions discussed previously and were therefore removed from the alignments.
-Fewer than 10\% of alignments were trimmed, but when regions were trimmed, they were largely inferred as state 3 (Fig.
-\ref{sfig:hmm_trim}).
+Fewer than 10\% of alignments were trimmed, but when regions were trimmed, they were largely inferred as state 3 (Fig.~\ref{sfig:hmm_trim}).
 
@@
 The orthologous groups and alignments yielded by this pipeline are a valuable resource for comparative studies of gene birth/death processes and protein evolution at the level of both entire proteome and specific gene families in the \textit{Drosophila} genus.
-To assist these efforts, consensus phylogenetic trees were inferred from meta-alignments sampled from alignments of single copy orthologous groups (Fig.
-\ref{fig:trees}).
+To assist these efforts, consensus phylogenetic trees were inferred from meta-alignments sampled from alignments of single copy orthologous groups (Fig.~\ref{fig:trees}).
 Although this work focused on single copy orthologs, other studies may require different subsets of orthologous groups that demand other pre-processing and alignment strategies.
@@
 
-For example, despite many additional pre-processing steps and other tweaks introduced by later authors, the basic framework of orthology inference by clustering the hit graph has remained largely unchanged in the past twenty years \cite{Remm2001, Li2003, Jensen2007, Linard2011, Emms2015, Train2017, Cosentino2018}.
+For example, despite many additional pre-processing steps and other tweaks introduced by later authors, the basic framework of orthology inference by clustering the hit graph has remained largely unchanged in the past twenty years~\cite{Remm2001, Li2003, Jensen2007, Linard2011, Emms2015, Train2017, Cosentino2018}.
 This longevity is a testament to the robustness of the underlying idea that orthologous proteins should consistently identify each other as the most similar pairs between their genomes.
-Even as the number of genomes and their taxonomic density has increased dramatically, many orthology inference pipelines continue to use triangle clustering, which was originally applied to sets of far fewer and more distantly related genomes \cite{Tatusov1997}.
+Even as the number of genomes and their taxonomic density has increased dramatically, many orthology inference pipelines continue to use triangle clustering, which was originally applied to sets of far fewer and more distantly related genomes~\cite{Tatusov1997}.
 This mismatch in scale increases the chance of propagating annotation errors since only a small number of edges are needed to create or merge clusters.
@@
 While the combination of these two steps yielded high-quality alignments that are suitable for further analyses, they are an \textit{ad hoc} fix for underlying issues with the gene models and alignment algorithms.
-The most principled solution is to optimize or supplement the gene models using the initial alignments generated by the orthology inference pipeline, which is possible with tools such as OMGene or OrthoFiller \cite{Dunne2018, Dunne2017}.
+The most principled solution is to optimize or supplement the gene models using the initial alignments generated by the orthology inference pipeline, which is possible with tools such as OMGene or OrthoFiller~\cite{Dunne2018, Dunne2017}.
 However, if preserving the original annotations is desired or necessary, the remaining possibility is to correct the alignments as we have done here.
@@
 
-The most popular aligners for protein sequences (Clustal Omega, MAFFT, MUSCLE, T-Coffee) do not include splice sites in their alignments, which makes them prone to aligning non-homologous exons \cite{Sievers2017, Katoh2013, Edgar2004, Notredame2000}.
-Current algorithms can easily be extended to incorporate splice sites by coding them as a new symbol and preventing alignment between splice sites and amino acids, which was recently implemented in the aligner, \textit{Mirage} \cite{Nord2018}.
+The most popular aligners for protein sequences (Clustal Omega, MAFFT, MUSCLE, T-Coffee) do not include splice sites in their alignments, which makes them prone to aligning non-homologous exons~\cite{Sievers2017, Katoh2013, Edgar2004, Notredame2000}.
+Current algorithms can easily be extended to incorporate splice sites by coding them as a new symbol and preventing alignment between splice sites and amino acids, which was recently implemented in the aligner, \textit{Mirage}~\cite{Nord2018}.
 The biggest challenge in practice, however, is mapping a protein sequence to its genomic sequence to identify splice sites in the protein sequence.
@@
 This would allow it to better model observed substitution patterns between specific symbols, for example the high rate of exchange between gaps and glutamine residues caused by polyglutamine tracts.
-The transitions between amino acids could be parametrized with a published matrix such as LG \cite{Le2008}, but the transitions between amino acids and gaps would be inferred from labeled data.
+The transitions between amino acids could be parametrized with a published matrix such as LG~\cite{Le2008}, but the transitions between amino acids and gaps would be inferred from labeled data.
 It is unclear if the resulting gain in accuracy would justify the increased computational burden, however.
 
-Though there are benchmarks available for optimizing and comparing methods of orthology inference, the metrics are calculated over a set of reference genomes which are sparsely sampled over a broad taxonomic range, so it is unclear if they are informative for method designed to yield robust inferences when the genomes are highly related \cite{Nevers2022}.
+Though there are benchmarks available for optimizing and comparing methods of orthology inference, the metrics are calculated over a set of reference genomes which are sparsely sampled over a broad taxonomic range, so it is unclear if they are informative for method designed to yield robust inferences when the genomes are highly related~\cite{Nevers2022}.
 Furthermore, the heterogeneity of genome architectures and annotations may require quality assurance methods tailored to each set of genomes.
-Thus, there is likely no one-size-fits-all approach to orthology inference, and with many other standalone programs available for more standard use cases (Hieranoid, OMA standalone, OrthoFinder, OrthoInspector, Orthologer), we have chosen not to package the code into an end-to-end pipeline \cite{Kaduk2017, Altenhoff2019, Emms2019, Linard2014, Zdobnov2020}.
+Thus, there is likely no one-size-fits-all approach to orthology inference, and with many other standalone programs available for more standard use cases (Hieranoid, OMA standalone, OrthoFinder, OrthoInspector, Orthologer), we have chosen not to package the code into an end-to-end pipeline~\cite{Kaduk2017, Altenhoff2019, Emms2019, Linard2014, Zdobnov2020}.
 Instead, we have devoted considerable attention to organizing and documenting the code to make it accessible to a newcomer and thereby facilitate the adaptation of specific steps to similar projects as needed.
@@
 
-Though databases of orthologous groups such as COGs, Ensembl Compara, EggNOG, OMA, OrthoDB, OrthoInspector, and OrthoMCL will continue to be useful for comparative studies across broad taxonomic ranges, the increasing speed at which high-quality genome assemblies and annotations are produced means no single database can encompass the most complete data \cite{Galperin2020, Herrero2016, HuertaCepas2018, Altenhoff2020, Zdobnov2020, Nevers2018, Chen2006}.
+Though databases of orthologous groups such as COGs, Ensembl Compara, EggNOG, OMA, OrthoDB, OrthoInspector, and OrthoMCL will continue to be useful for comparative studies across broad taxonomic ranges, the increasing speed at which high-quality genome assemblies and annotations are produced means no single database can encompass the most complete data~\cite{Galperin2020, Herrero2016, HuertaCepas2018, Altenhoff2020, Zdobnov2020, Nevers2018, Chen2006}.
 Furthermore, since many early comparative genomics studies spanned diverse branches of the tree of life, future research will likely prioritize taxonomic depth over breadth.
@@
 Thus, the first accession associated with a sequence and gene pair is the sequence’s representative accession for the gene.
-BLAST+ 2.13.0 was used for the sequence similarity searches \cite{Camacho2009}.
+BLAST+ 2.13.0 was used for the sequence similarity searches~\cite{Camacho2009}.
 An E-value cutoff of 1 was used for the initial searches.
@@
 \subsection*{Initial alignment and selection of representative sequences}
-The sequences in each orthologous group were aligned using MAFFT 7.490 with the following settings: --globalpair --maxiterate 1000 --thread 1 --anysymbol --allowshift --unalignlevel 0.4 --leavegappyregion \cite{Katoh2013}.
+The sequences in each orthologous group were aligned using MAFFT 7.490 with the following settings: --globalpair --maxiterate 1000 --thread 1 --anysymbol --allowshift --unalignlevel 0.4 --leavegappyregion~\cite{Katoh2013}.
 Representative sequences for each gene were then selected by maximum likelihood according to binary profiles constructed from these alignments.
@@
 Otherwise, the group contributed a count for the gap symbol at that position.
-To account for the phylogenetic dependencies between sequences, the counts were weighted according to a Gaussian process over the GTR2 consensus tree described in the section on inferring species trees \cite{Altschul1989}.
+To account for the phylogenetic dependencies between sequences, the counts were weighted according to a Gaussian process over the GTR2 consensus tree described in the section on inferring species trees~\cite{Altschul1989}.
 If a species had multiple genes in the orthologous group, the species weight was divided evenly among them.
@@
 Consensus columns were manually defined as columns in the initial alignments with at least five non-gap symbols using the Stockholm file format.
-These labeled alignments were used to build a profile HMM using HMMer 3.3.2 \cite{Eddy2009}.
+These labeled alignments were used to build a profile HMM using HMMer 3.3.2~\cite{Eddy2009}.
 No sequence relative weighting was used, and the effective number of sequences was set to three times the number of sequences in the alignment.
@@
 The other two filters had a kernel size of 60 with a dilation of 3.
-The second layer was composed of two filters with a kernel size of 10 and a dilation of 1 (Fig.
-\ref{sfig:cnn_architecture}).
+The second layer was composed of two filters with a kernel size of 10 and a dilation of 1 (Fig.~\ref{sfig:cnn_architecture}).
 The network was trained on 88,513 hand-labeled positions across 122 sequences in 46 unique alignments.
-These examples were largely long segments aligned to few, if any, other sequences (Fig.
-\ref{sfig:cnn_training}).
+These examples were largely long segments aligned to few, if any, other sequences (Fig.~\ref{sfig:cnn_training}).
 
@@
 All HMM algorithms were implemented with custom code which is available as the package Homomorph on PyPI.
-The insertion phylo-HMM was trained on 39,298 manually labeled columns in 11 alignments, and the deletion phylo-HMM was trained on 58,655 manually labeled positions in 20 sequences in 9 unique alignments (Fig.
-\ref{sfig:hmm_training}).
-Because maximum-likelihood estimation of the model parameters yielded posterior decoding curves which toggled between hidden states too rapidly, the models were instead trained discriminatively \cite{Krogh1999}.
+The insertion phylo-HMM was trained on 39,298 manually labeled columns in 11 alignments, and the deletion phylo-HMM was trained on 58,655 manually labeled positions in 20 sequences in 9 unique alignments (Fig.~\ref{sfig:hmm_training}).
+Because maximum-likelihood estimation of the model parameters yielded posterior decoding curves which toggled between hidden states too rapidly, the models were instead trained discriminatively~\cite{Krogh1999}.
 The difference, briefly, is maximum-likelihood estimation finds the parameters that best reproduce the observed distributions whereas discriminative training finds the parameters that minimize prediction error.
@@
 Their combination yielded six different sets of meta-alignments.
-A tree was fit to each meta-alignment with the LG substitution model and four discrete gamma rate categories using IQ-TREE 1.6.12 \cite{Nguyen2014, Le2008, Yang1994}.
+A tree was fit to each meta-alignment with the LG substitution model and four discrete gamma rate categories using IQ-TREE 1.6.12~\cite{Nguyen2014, Le2008, Yang1994}.
 If the sampling strategy allowed invariant columns, an invariant rate category was included.
-The resulting trees from each set were then merged into a majority consensus tree (Fig.
-\ref{sfig:trees_LG}).
+The resulting trees from each set were then merged into a majority consensus tree (Fig.~\ref{sfig:trees_LG}).
 
@@
 An invariant category, however, was included if invariant columns were allowed.
-Since the bootstrap confidences were sometimes lower than 50\%, the resulting trees from each set were then merged into a loose consensus tree to prevent multifurcations (Fig.
-\ref{sfig:trees_GTR}).
+Since the bootstrap confidences were sometimes lower than 50\%, the resulting trees from each set were then merged into a loose consensus tree to prevent multifurcations (Fig.~\ref{sfig:trees_GTR}).
 
@@
 The code used to produce the results and analyses is available at \url{https://github.com/marcsingleton/orthology_inference2023}.
-The following Python libraries were used: matplotlib, NumPy, pandas, SciPy, and TensorFlow \cite{Hunter2007, Harris2020, McKinney2010, Virtanen2020, Abadi2015}.
+The following Python libraries were used: matplotlib, NumPy, pandas, SciPy, and TensorFlow~\cite{Hunter2007, Harris2020, McKinney2010, Virtanen2020, Abadi2015}.
 Relevant output files are available in the supporting information.

### 2022-11-11 chapter1.tex
Incorporated various clarifications and re-wordings (many suggested by Xiao-Yong)

@@
 These methods often rely on the identification of homologous sequences or homologs, that is sequences with common ancestry, since this ensures that differences between sequences reflect variations in evolution from a common point of divergence.
-In a sense, each branch that descends from a single ancestral population is a different ``experiment'' by nature, as each is free to explore the evolutionary landscape independent of the others.
 However, many analyses impose the additional condition that the sequences have diverged through speciation events (orthology) rather than duplications (paralogy) or other mechanisms such as horizontal gene transfer~\cite{Fitch1970}.
@@
 
-Thus, in this work we developed a computational pipeline that can robustly infer and align orthologous groups of proteins even when the genomes are highly redundant (Fig.~\ref{fig:graphical_overview}).
-We did not seek to reinvent the wheel, so the overall principle of clustering a graph of hits from homology searches is unchanged, though many details were modified to maximize the detection of highly diverged orthologs while also minimizing the impact of incomplete or incorrect annotations.
-Furthermore, as modern genome annotation pipelines frequently produce gene models and protein sequences in tandem, we implemented an additional clustering step to organize the resulting orthologous groups of proteins into gene-level units.
-However, most of our efforts were instead focused on the final step of aligning the orthologous sequences.
-Though genome annotation pipelines are often proficient at identifying the locus of genes, the quality of their gene models can vary considerably and can contain inaccurate exon boundaries or start codons~\cite{Frankish2015, Dunne2018}.
-As a result, the structure and lengths of the protein sequences that result from these models can be highly heterogeneous, which poses many challenges for their alignment and subsequent analyses.
+In this work we therefore developed a computational pipeline that can robustly infer and align orthologous groups of proteins even when the genomes are highly redundant (Fig.~\ref{fig:graphical_overview}).
+Like many other orthology inference pipelines, our overall approach is based on clustering a graph of hits from homology searches.
+However, we modified many details to maximize the detection of highly diverged orthologs while also minimizing the impact of incomplete or incorrect annotations.
+Furthermore, since modern genome annotation pipelines frequently produce gene models and protein sequences in tandem, we implemented an additional clustering step to organize the resulting orthologous groups of proteins into gene-level units.
+However, most of our efforts were focused on the final step of aligning the orthologous sequences.
+Though genome annotation pipelines are often proficient at identifying the overall locus of genes, the accurate identification of exon boundaries and start codons when transcript evidence is limited remains an ongoing challenge~\cite{Frankish2015, Dunne2018}.
+Consequently, protein sequences derived from annotation pipelines can include non-homologous segments of significant length or exclude highly conserved segments.
+Such heterogeneity in the structure and length of the sequences in an orthologous group poses many challenges for their alignment and subsequent analysis.
+Thus, we implemented several novel quality control and data cleaning steps to correct mis-alignments and identify likely sequencing, assembly, or annotation errors.
 
-To develop these methods, we chose a set of 33 assembled and annotated \textit{Drosophila} genomes.
-All 12 species from the original \textit{Drosophila} 12 Genomes Consortium are included~\cite{D12GC2007}.
-However, all have been re-sequenced and re-annotated, many with long-read sequencing technology, which has resulted in substantial improvements in the contiguity of their assemblies.
-Despite this and other genome assembly projects of species in the \textit{Drosophila} genus, there is not yet a database of high-quality alignments of orthologous proteins that reflects these improvements in genome assembly and diversity~\cite{Miller2018, Kim2021}.
+To develop these methods, we chose a set of 33 assembled and annotated \textit{Drosophila} genomes, which includes all 12 species from the original \textit{Drosophila} 12 Genomes Consortium~\cite{D12GC2007}.
+However, the genomes of these 12 species have been re-sequenced since their first release, which has resulted in substantial improvements in their assemblies and annotations.
+Despite these developments and other several other recent genome assembly projects of species in the \textit{Drosophila} genus, there is not yet a collection of high-quality alignments of orthologous proteins that reflects these improvements in genome assembly and diversity~\cite{Miller2018, Kim2021}.
 Given the \textit{Drosophila} genus spans diverse habitats and over 50 million years of evolution but maintains a conserved life cycle and body plan, such a resource would facilitate a new generation of studies that illuminate the forces that drive protein evolution in unprecedented detail~\cite{Wiegmann2011, Obbard2012}.
@@
 \section*{Results}
-\subsection*{Input genomes}
+\subsection*{Input genomes and pre-processing}
 All annotated genomes in the genus \textit{Drosophila} available in April 2022 were downloaded from NCBI's RefSeq database.
-The assemblies annotated by the NCBI eukaryotic genome annotation pipeline must past several quality checks and have supporting transcript evidence, so the annotations are generally highly complete~\cite{Thibaud-Nissen2013}.
+The assemblies annotated by the NCBI eukaryotic genome annotation pipeline have passed several quality checks and all have supporting transcript evidence, so the annotations are generally highly complete~\cite{Thibaud-Nissen2013}.
 The \textit{D.
 miranda} annotation was excluded due to its unusual karyotype~\cite{Dobzhansky1935}.
-Other annotations were excluded after preliminary clustering showed a deficient in the number of orthologous groups containing those genomes, indicating their annotations were less complete (data not shown).
+Other annotations were excluded after preliminary clustering showed a deficiency in the number of orthologous groups containing those genomes, indicating their annotations were less complete (data not shown).
 The \textit{D.
@@
 The complete list of genomes is given in Table~\ref{stable:genomes}.
-
-Transcripts with alternate UTRs can yield duplicate protein sequences in the annotation.
-While these alternative UTRs can encode biologically distinct functions, their identical coding sequences make them indistinguishable at the protein level, and the duplicates were therefore removed.
-Though not strictly necessary, this greatly reduced the computational burden of later steps.
+Many genes have transcripts that differ only in their UTRs, and as a result there are many duplicate protein sequences in the annotations.
+Though not strictly necessary, we removed the duplicates in our pipeline, which greatly reduced the computational burden of later steps.
 
@@
 HSPs are local alignments, meaning they do not necessarily span the entire lengths of the query and target sequences.
-Consequently, the search algorithm may return multiple HSPs if statistically significant regions of homology are separated by nonhomologous or poorly conserved regions.
+Consequently, the search algorithm may return multiple HSPs for each query-target pair if statistically significant regions of homology are separated by nonhomologous or poorly conserved regions.
 Though the most significant HSP is often used to represent all HSPs between a query-target pair, this approach can fail to rank the pairs by their overall significance if their alignments are broken into multiple HSPs.
@@
 There are many community detection algorithms available, but a popular approach in the context of orthology inference is building clusters progressively by identifying nodes that form a triangle with at least two other nodes in the cluster.
-While this method is robust when clustering smaller or more diverse sets of genomes, the large number of closely related genomes in the input data creates many edges which form triangles but are otherwise poorly supported by the remaining genomes.
+While this method is robust when clustering smaller or more diverse sets of genomes, the large number of closely related genomes in the input data yields many nodes which form triangles but lack dense networks of edges with other nodes.
 However, triangle clustering is a special case of the more general \textit{k}-clique percolation algorithm where \textit{k} = 3 (Fig.~\ref{fig:percolation}).
 Higher values of \textit{k} require greater overlap between a candidate node and those already in the cluster and therefore produce tighter clusters at the cost of excluding more speculative orthology relationships.
-\textit{k} was set to 4 as compromise between these concerns, yielding 22,813 orthologous groups, a plurality of which contained all 33 species (Fig.~\ref{sfig:clusters}).
+We set \textit{k} to 4 as compromise between these concerns, yielding 22,813 orthologous groups, a plurality of which contained all 33 species (Fig.~\ref{sfig:clusters}).
 
@@
 \subsection*{Grouping orthologous groups by gene}
-As genes can have several annotated isoforms, each gene can also be associated with several orthologous groups.
+As genes can have several annotated isoforms, each gene can be associated with several orthologous groups.
 However, the orthologous groups are not organized into gene-level units since they were clustered using sequence similarity only.
@@
 \subsection*{Initial alignment and selection of representative sequences}
-Since the annotation pipeline incorporates transcriptome data from a variety of sources, its inputs are heterogeneous in sequencing depth, developmental stage, and tissue of origin across different genomes.
-As a result, some orthologous genes are annotated with more splice isoforms in some genomes than in others.
-Under certain conditions, the clustering algorithm can merge smaller or looser clusters of these additional splice isoforms with the main cluster, creating a single orthologous group where some genes have multiple associated sequences.
+Since the NCBI annotation pipeline incorporates transcriptome data from a variety of sources, its inputs are heterogeneous in sequencing depth, developmental stage, and tissue of origin across different genomes.
+As a result, some genomes are annotated with different or multiple splice isoforms of a given orthologous gene, which can create complex networks in the resulting hit graph.
+For example, if the genomes are variably annotated with one or both of two distinct isoforms, the resulting graph may contain two clusters connected by a "bridge" formed by the genomes which contain only one of the isoforms.
+If the nodes bridging the two clusters form cliques with themselves and the clusters, the clique percolation algorithm will merge all the nodes into a single orthologous group where some genes have multiple associated sequences.
 However, these additional sequences can complicate downstream comparative analyses that may not easily generalize to genes with multiple associated sequences.
-Thus, a single representative was chosen for each gene using an alignment-based strategy detailed in the methods section.
+Thus, in our pipeline a single representative was chosen for each gene using an alignment-based strategy detailed in the methods section.
 
@@
 \textit{a\textsubscript{max}} varies between 0 and 1, but at values above 0.8, the correction severely degrades the alignment of conserved regions.
-Thus, the sequences were aligned with a conservative value of 0.4.
+Thus, in our pipeline the sequences were aligned with a conservative value of 0.4.
 Though some cases of over-alignment remained, most conserved regions were reliably aligned with few to no errors.
@@
 
-Phylogenetic HMMs (phylo-HMMs) are statistical models that incorporate phylogenetic and spatial information to calculate the probability that each observation in a sequence was generated by one of several “hidden states.” Since they can evaluate both the probability of a gap pattern in a column given the known phylogenetic relationships and the local context, a phylo-HMM was used to segment the alignment into contiguous regions with different patterns of gaps.
+Phylogenetic HMMs (phylo-HMMs) are statistical models that incorporate phylogenetic and spatial information to calculate the probability that each observation in a sequence was generated by one of several “hidden states” \cite{Felsenstein1996}.
+Since they can evaluate both the probability of a gap pattern in a column given the known phylogenetic relationships and the local context, a phylo-HMM was used to segment the alignment into contiguous regions with different patterns of gaps.
 This phylo-HMM was composed of four hidden states, {1A, 1B, 2, 3}, roughly corresponding to highly conserved regions with few to no gaps, divergent regions, regions with a stable gap pattern discordant with the expected phylogenetic relationships, and regions with poorly supported segments.
@@
 Fortunately, triangle clustering is a special case of the more general \textit{k}-clique percolation algorithm where \textit{k} is a tunable parameter that influences the tightness of a cluster.
-The optimal value of \textit{k} for a given set of genomes is unclear, but it likely depends on the desired trade-off between sensitivity and specificity.
+The optimal value of \textit{k} for a given set of genomes is unclear and likely depends on the desired trade-off between sensitivity and specificity.
 However, percolation theory or simulations may yield insights.
 
-A more subtle difficulty is the annotation of multiple isoforms for a single gene.
-Though prior pipelines have generally selected the longest isoform as the representative before conducting the orthology searches, we felt this approach would introduce artifacts or other issues during alignment.
+Another challenge is the annotation of multiple isoforms for a single gene.
+Though prior pipelines have generally selected the longest isoform as the representative before conducting the orthology searches, if the sequences do not share a common intron-exon structure this approach can introduce artifacts or other issues during alignment.
 Instead, as protein sequences are increasingly derived from or linked to genomic sequences, we sought to incorporate the full annotations into the orthology inference pipeline.
@@
 Afterwards, a single representative is easily chosen as the group with the largest number of distinct species, though other criteria are possible.
-Since the second occurs within an orthologous group, the sequences were first aligned, and a representative for each gene was chosen as the sequence which was most concordant with the initial alignment.
+Since the second occurs within an orthologous group, the sequences were first aligned, and a representative for each gene was chosen as the sequence which was most concordant with this initial alignment.
 
@@
 However, it does not remove poorly supported segments from columns with more than five non-gap symbols, and their contribution is sometimes sufficient to reproduce the misalignment when the sequences are aligned to the profile HMM.
-A lightweight convolutional neural network was therefore trained to recognize poorly supported segments to more completely remove them from the profile HMM.
+A lightweight convolutional neural network was therefore trained to recognize poorly supported segments to remove them more completely from the profile HMM.
 The network consisted of a two-dimensional embedding from a 22-dimensional one-hot-encoded vector (20 amino acids, an ambiguous amino acid, and a gap), two convolutional layers, and a one-dimensional output.
@@
 The code used to produce the results and analyses is available at \url{https://github.com/marcsingleton/orthology_inference2023}.
+HMM algorithms were implemented in the standalone package Homomorph which is available at \url{https://github.com/marcsingleton/homomorph} and on the Python Package Index (PyPI).
 The following Python libraries were used: matplotlib, NumPy, pandas, SciPy, and TensorFlow~\cite{Hunter2007, Harris2020, McKinney2010, Virtanen2020, Abadi2015}.

### 2022-11-11 figures/cnn_training/cnn_training.py
Changed cnn_training.py to use intervals for display

@@ -148,9 +148,15 @@ fig.savefig('out/cnn_training.png', dpi=300)
 fig.savefig('out/cnn_training.tiff', dpi=300)
 plt.close()
 
-record_ids = [('07C3', 'XP_032288889.1'), ('23D9', 'XP_026832050.1')]
-for record_id in record_ids:
-    OGid, ppid, profile, seq, labels, weights = records[record_id]
+record_ids = [('010D', 'XP_039479706.2', 465, 945, 50),
+              ('010D', 'XP_017002872.2', 465, 945, 50),
+              ('23D9', 'XP_026832050.1', 625, 1331, 25),
+              ('03E7', 'XP_022223442.2', 698, 1656, 20),
+              ('1E46', 'XP_034479221.1', 2407, 2761, 50),
+              ('2CDB', 'XP_046866895.1', 1763, 2425, 25)]
+for OGid, ppid, start, stop, margin in record_ids:
+    start, stop = start - margin, stop + margin
+    _, _, profile, seq, labels, weights = records[(OGid, ppid)]
     output = tf.squeeze(model([np.expand_dims(profile, 0), np.expand_dims(seq, 0)]))  # Expand and contract dims
 
     msa = []
@@ -160,14 +166,15 @@ for record_id in record_ids:
         msa.append({'ppid': msa_ppid, 'spid': msa_spid, 'seq': seq})
     msa = sorted(msa, key=lambda x: tip_order[x['spid']])
 
-    data = [output, labels, weights]
+    data = [output[start:stop], labels[start:stop], weights[start:stop]]
     msa_labels = [msa_record['ppid'] if msa_record['ppid'] == ppid else '' for msa_record in msa]
-    fig = plot_msa_data([msa_record['seq'] for msa_record in msa], data, figsize=(7.5, 3.5),
+    fig = plot_msa_data([msa_record['seq'][start:stop] for msa_record in msa], data, figsize=(7.5, 3.5),
+                        x_start=start,
                         msa_labels=msa_labels, msa_ticklength=1, msa_tickwidth=0.25, msa_tickpad=1.1, msa_labelsize=5,
                         height_ratio=0.5, hspace=0.2, data_max=1.1, data_min=-0.1, data_labels=['output', 'label', 'weight'],
[…]

### 2022-11-13 chapter1.tex
Added TODO notes and new sections

@@
 
-In this work we therefore developed a computational pipeline that can robustly infer and align orthologous groups of proteins even when the genomes are highly redundant (Fig.~\ref{fig:graphical_overview}).
+In this work we therefore developed a computational pipeline that can robustly infer and align orthologous groups of proteins even when the genomes are highly redundant.
 Like many other orthology inference pipelines, our overall approach is based on clustering a graph of hits from homology searches.
@@
 \section*{Results}
+\subsection*{Pipeline overview}
+% TODO: Add description of pipeline, adapting text from Figure 1 caption
+
 \subsection*{Input genomes and pre-processing}
@@
 \subsection*{Clustering in orthologous groups}
+% TODO: Add description of structure of connected components to motivate clustering algorithm
 The best hits between sequences are naturally visualized as a graph where the sequences are nodes and best hits are edges between nodes (Fig.~\ref{fig:components}).
@@
 
+% TODO: Add more discussion of different clustering methods and their weaknesses (include citations)
 The identification of sets of densely connected nodes in graphs is known as community detection in network analysis.
@@
 
-\section*{Discussion}
-The orthologous groups and alignments yielded by this pipeline are a valuable resource for comparative studies of gene birth/death processes and protein evolution at the level of both entire proteome and specific gene families in the \textit{Drosophila} genus.
-To assist these efforts, consensus phylogenetic trees were inferred from meta-alignments sampled from alignments of single copy orthologous groups (Fig.~\ref{fig:trees}).
-Although this work focused on single copy orthologs, other studies may require different subsets of orthologous groups that demand other pre-processing and alignment strategies.
-Thus, the orthologous groups and the initial alignments with and without non-representative sequences are included as supplemental data.
+\subsection*{Inference of species trees}
+% TODO: Add discussion of creating species tree
 
@@
 \end{figure}
+
+\section*{Discussion}
+% TODO: Re-write to remove species tree and combine with following paragraph; also highlight specific contributions
+The orthologous groups and alignments yielded by this pipeline are a valuable resource for comparative studies of gene birth/death processes and protein evolution at the level of both entire proteomes and specific gene families in the \textit{Drosophila} genus.
+To assist these efforts, consensus phylogenetic trees were inferred from meta-alignments sampled from alignments of single copy orthologous groups (Fig.~\ref{fig:trees}).
+Although this work focused on single copy orthologs, other studies may require different subsets of orthologous groups that demand other pre-processing and alignment strategies.
+Thus, the orthologous groups and the initial alignments with and without non-representative sequences are included as supplemental data.
 

### 2022-11-13 chapter1.tex
Clarified caption on CNN training supplemental figure

@@
 The convolutional filters learn kernels which correspond to the length of a typical long poorly supported segment.
-F) An example alignment where the indicated sequence is labeled with the trained CNN.
+F) The alignment of representative sequences in orthologous group 23D9 where the indicated sequence is labeled with the trained CNN.
 The blue lines are the CNN’s predictions, and the yellow lines are the labels.
 The red lines are the weights given to each position.
-The regions where the weights are zero are masked from the training data, which allowed the model to focus on only the most relevant examples.
-}
+The regions where the weights are zero are masked from the training data, which allowed the model to focus on only the most relevant examples.}
 \label{sfig:cnn_training}

### 2022-11-13 figures/realignment/realignment.py
Changed realignment.py to include multiple examples

@@ -4,35 +4,42 @@ import os
 import re
 
 import matplotlib.pyplot as plt
+import numpy as np
 import skbio
 from src.draw import plot_msa
 from src.utils import read_fasta
 
-OGids = ['2AB6', '07C3', '31B4', '269C', '06D1', '23D9']
-msa_pathA = '../../orthology_inference/analysis/ortho_MSA/get_repseqs/out/'
-msa_pathB = '../../orthology_inference/analysis/ortho_MSA/realign_hmmer/out/mafft/'
+records = [('23D9', 'A', 731, 1158, 704, 1000),
+           ('01E0', 'B', 0, 344, 0, 84),
+           ('07C3', 'C', 0, 484, 0, 211),
+           ('0430', 'D', 0, 476, 0, 126)]
+msa_path1 = '../../orthology_inference/analysis/ortho_MSA/get_repseqs/out/'
+msa_path2 = '../../orthology_inference/analysis/ortho_MSA/realign_hmmer/out/mafft/'
 spid_regex = r'spid=([a-z]+)'
 
-adjust_left = 0.01
-adjust_bottom = 0.075
-adjust_right = 0.915
-adjust_top = 0.925
+adjust_left1 = 0.02
+adjust_left2 = 0.05
+adjust_bottom = 0.1
+adjust_right1 = 0.95
+adjust_right2 = 0.85
+adjust_top = 0.9
 
 hspace = 0.3
 x_labelsize = 5
-legend_position = (0.915, 0.5)
-legend_fontsize = 6
+legend_position = (0.86, 0.5)
+legend_fontsize = 5
 legend_handletextpad = 0.5
 legend_markerscale = 1
 
-fig_width = 7.5
-fig_height = 3
-figsize_effective = (fig_width * (adjust_right - adjust_left), fig_height * (adjust_top - adjust_bottom))
-dpi = 300
+fig_width = 3.75
+fig_height = 2.5
+figsize_effective1 = (fig_width * (adjust_right1 - adjust_left1), fig_height * (adjust_top - adjust_bottom))
+figsize_effective2 = (fig_width * (adjust_right2 - adjust_left2), fig_height * (adjust_top - adjust_bottom))
+dpi = 600
 
-panel_labels = ['A', 'B']
 panel_label_fontsize = 'large'
 panel_label_offset = 0.025
+suptitle_fontsize = 'medium'
 
[…]

### 2022-11-13 chapter1.tex
Realignment figure now uses new version and caption

@@
 \begin{figure}[h!]
-\includegraphics[width=\textwidth]{realignment/out/23D9_A.png}
-\includegraphics[width=\textwidth]{realignment/out/23D9_B.png}
+\includegraphics[width=\textwidth]{realignment/out/merged.png}
 \centering
 \caption{\textbf{Alignment with long poorly supported segments.}
-The alignments of representative sequences in orthologous group 23D9 are shown before (panel A) and after (panel B) refinement.
+The alignments of representative sequences in orthologous groups 23D9 (A), 01E0 (B), and 07C3 (C) are shown before and after refinement.
 The refinement process removes long poorly supported segments at the termini.

### 2022-11-13 figures/trees/trees.py
Fixed bad path to trees

@@ -13,7 +13,7 @@ if not os.path.exists('out/'):
 # FIGURE WITH TWO TREES
 trees = []
 for model_label in ['LG', 'GTR']:
-    tree = skbio.read('../../orthology_inference/analysis/ortho_tree/consensus_LG/out/100R_NI.nwk', 'newick', skbio.TreeNode)
+    tree = skbio.read(f'../../orthology_inference/analysis/ortho_tree/consensus_{model_label}/out/100R_NI.nwk', 'newick', skbio.TreeNode)
     tree.assign_supports()
     for node in tree.traverse():
         if node.support == 1:

### 2022-11-13 chapter1.tex
Moved discussion of trees to results and re-wrote first paragraph of discussion

@@
 \subsection*{Inference of species trees}
-% TODO: Add discussion of creating species tree
+Many phylogenetic methods require a species tree to inform the evolutionary relationships between sequences.
+In fact, the phylo-HMMs discussed previously used a species tree as an input, though we omitted this detail for clarity of exposition.
+Therefore, to support the curation step and other downstream analyses, we sought to infer phylogenetic trees from the aligned sequences.
+However, since the roots of phylogenetic trees are not identifiable with commonly used time-reversible substitution models, we repeated the orthology inference pipeline with the outgroup species \textit{Scaptodrosophila lebanonensis}.
+Afterwards, we inferred phylogenetic trees using the LG model of amino acid substitution from 100 meta-alignments sampled from alignments of single copy orthologous groups.
+We then combined them into a single consensus tree (Fig.~\ref{fig:trees}A).
+To provide a similar tree for the analysis of non-coding regions, we inferred phylogenetic trees using the GTR model of nucleotide substitution from 100 meta-alignments sampled from nucleotide alignments which were "reverse translated" from the protein alignments and their corresponding coding sequences.
+The resulting consensus tree is shown in Figure~\ref{fig:trees}B.
+Both trees have an identical topology, which is consistent with other published phylogenies \cite{D12GC2007, DaLage2007}.
 
@@
 \caption{\textbf{Phylogenetic tree of species.}
-Phylogenetic trees were fit to meta-alignments sampled from alignments created by the orthology inference pipeline with the additional genome of \textit{Scaptodrosophila lebanonensis} (sleb) as an outgroup.
-The trees fit to each meta-alignment were then combined into a single consensus tree.
-Values at nodes are bootstrap percentages.
 A) The LG model was fit to meta-alignments directly sampled from the original protein alignments B).
-The GTR model was fit to meta-alignments sampled from nucleotide alignments which were “reverse translated” from the original protein alignments and their corresponding coding sequences.}
+The GTR model was fit to meta-alignments sampled from "reverse translated" nucleotide alignments.
+Values at nodes are bootstrap percentages.}
 \label{fig:trees}
@@
 \section*{Discussion}
-% TODO: Re-write to remove species tree and combine with following paragraph; also highlight specific contributions
+% TODO: Highlight specific contributions
 The orthologous groups and alignments yielded by this pipeline are a valuable resource for comparative studies of gene birth/death processes and protein evolution at the level of both entire proteomes and specific gene families in the \textit{Drosophila} genus.
-To assist these efforts, consensus phylogenetic trees were inferred from meta-alignments sampled from alignments of single copy orthologous groups (Fig.~\ref{fig:trees}).
+To assist these efforts, the final alignments after curation and the labels from the "missing" phylo-HMM are provided as supplemental data.
 Although this work focused on single copy orthologs, other studies may require different subsets of orthologous groups that demand other pre-processing and alignment strategies.
-Thus, the orthologous groups and the initial alignments with and without non-representative sequences are included as supplemental data.
-
-The trends that permitted this work to substantially improve on previous efforts will quickly render its products obsolete.
-More \textit{Drosophila} genomes will be assembled and annotated, and the existing annotations will be updated.
+Therefore, we have included the orthologous groups and the initial alignments with and without non-representative sequences in the supplemental data as well.
+While we anticipate these resources will remain relevant in the near term, the trends that permitted this work to substantially improve on previous efforts will render them obsolete in the coming years as more \textit{Drosophila} genomes are assembled and annotated.
 However, an authoritative and lasting set of orthologous groups and alignments is not the primary goal of this work.
-It instead serves as a case study in how dense taxonomic sampling and modern genome assembly and annotation pipelines present new opportunities and challenges to the traditional techniques for identifying and aligning orthologous groups.
+Instead, it serves as a case study in how dense taxonomic sampling and modern genome assembly and annotation pipelines present new opportunities and challenges to the traditional techniques for identifying and aligning orthologous groups.
 
@@
 \subsection*{Inference of species trees}
-Since the roots of phylogenetic trees are not identifiable with commonly used time-reversible substitution models, the orthology inference pipeline was repeated with the outgroup species \textit{Scaptodrosophila lebanonensis}.
-Orthologous groups with one sequence for each species were aligned, and 100 meta-alignments were constructed by randomly sampling 10,000 columns from these 9,435 alignments.
+The orthology inference pipeline was first repeated with the outgroup species \textit{Scaptodrosophila lebanonensis}.
+Then orthologous groups with one sequence for each species were aligned, and 100 meta-alignments were constructed by randomly sampling 10,000 columns from these 9,435 alignments.
 (The alignments were not refined with HMMer before sampling.) To determine the effect of invariant columns and gaps, two sampling strategies were used where invariant columns were allowed or disallowed and the maximum fraction of gaps was set at 0, 50, and 100\%.

### 2022-11-13 figures/realignment/realignment.py
Fixed bad interval for panel B

@@ -10,7 +10,7 @@ from src.draw import plot_msa
 from src.utils import read_fasta
 
 records = [('23D9', 'A', 731, 1158, 704, 1000),
-           ('01E0', 'B', 0, 344, 0, 84),
+           ('01E0', 'B', 0, 344, 0, 94),
            ('07C3', 'C', 0, 484, 0, 211),
            ('0430', 'D', 0, 476, 0, 126)]
 msa_path1 = '../../orthology_inference/analysis/ortho_MSA/get_repseqs/out/'

### 2022-11-13 chapter1.tex
Moved some caption text into main text for realignment figure

@@
 The original sequences were then aligned to these profiles.
-By defining highly conserved “anchor” regions, this approach largely prevents the alignment of chance sequence similarities in long poorly supported segments (Fig.~\ref{fig:realignment}).
+By defining highly conserved “anchor” regions, this approach largely prevents the alignment of chance sequence similarities in long poorly supported segments.
 Furthermore, since profile HMMs explicitly model the beginnings and ends of alignments, they can fully exclude these segments from the termini.
+This effect is visible in the three examples shown in Figure \ref{fig:realignment}, where each has long poorly supported segments that are completely removed.
 
@@
 \caption{\textbf{Alignment with long poorly supported segments.}
-The alignments of representative sequences in orthologous groups 23D9 (A), 01E0 (B), and 07C3 (C) are shown before and after refinement.
-The refinement process removes long poorly supported segments at the termini.
-However, if residues are aligned to consensus columns before and after a long poorly segment, it remains in the alignment.}
+The alignments of representative sequences in orthologous groups 23D9 (A), 01E0 (B), and 07C3 (C) are shown before and after refinement.}
 \label{fig:realignment}

### 2022-11-13 chapter1.tex
Moved HMM figure caption into main text

@@
 Since they can evaluate both the probability of a gap pattern in a column given the known phylogenetic relationships and the local context, a phylo-HMM was used to segment the alignment into contiguous regions with different patterns of gaps.
-This phylo-HMM was composed of four hidden states, {1A, 1B, 2, 3}, roughly corresponding to highly conserved regions with few to no gaps, divergent regions, regions with a stable gap pattern discordant with the expected phylogenetic relationships, and regions with poorly supported segments.
-After the model was trained on manually labeled examples, it was used to assign a label to the columns in each alignment (Fig.~\ref{fig:hmm}).
+A fully specified phylo-HMM requires a fixed number of hidden states and a probability distribution for each.
+Thus, we identified four distinct types of regions in the alignments, roughly corresponding to highly conserved regions with few to no gaps, divergent regions, regions with a stable gap pattern discordant with the expected phylogenetic relationships, and regions with poorly supported segments.
+For simplicity, however, we refer to the states that generate each type of region as 1A, 1B, 2, and 3, respectively.
+The probability distributions for each state shared a common structure, which is illustrated in Figure \ref{fig:hmm}A.
+We conceptualized the observed alignments as the superposition of two distinct processes.
+The first is a phylogenetic process which evolves and splits a single ancestral sequence over time according to a tree.
+The second is the annotation process which can erroneously exclude or include segments from a sequence.
+The result is an alignment of annotated sequences which contains evolutionary information obscured by "noise" from the annotation process, shown here by the exclusion of three N-terminal residues in the fourth annotated sequence.
+To simplify modeling this behavior with an HMM, we first coded the sequences into binary symbols.
+The distributions for each state then consisted of two components derived from the encoded sequences.
+The first component models the gap pattern with a Markov process.
+This Markov process is in turn composed of two subprocesses where the first is a phylogenetic process, and the second is a jump process.
+These subprocess roughly correspond to changes caused by evolution and annotation, respectively.
+Because this first component did not fully capture the propensity for the gap patterns to remain constant, we included a second component that models the "gap stickiness" as a beta-binomial random variable by counting the number of symbols that remain constant between columns.
+The parameters associated with each component are shown to the right of Figure \ref{fig:hmm}A.
+Each state has a different set of parameters that yield its characteristic gap pattern and gap stickiness.
+
+After the model was trained on manually labeled examples, it was used to assign a label to the columns in each alignment.
 Columns assigned to states 1A and 1B are the regions of interest for downstream analyses since the gaps generally follow the expected pattern given the phylogenetic tree.
 In contrast, columns assigned to states 2 and 3 largely corresponded to the long poorly supported segments and phylogenetically discordant regions discussed previously and were therefore removed from the alignments.
-Fewer than 10\% of alignments were trimmed, but when regions were trimmed, they were largely inferred as state 3 (Fig.~\ref{sfig:hmm_trim}).
+An example decoded alignment is shown in Figure \ref{fig:hmm})B where the decoded states closely follow the expected patterns.
+Overall, fewer than 10\% of alignments were trimmed of any regions.
+However, when regions were trimmed, they were largely inferred as state 3, meaning the removed regions are generally long poorly supported segments that are aligned to few if any other sequences (Fig.~\ref{sfig:hmm_trim}).
 
@@
 \caption{\textbf{HMM emission architecture and a decoded alignment.}
-A) A single ancestral sequence evolves and splits into distinct sequences over time.
-The tree shows the history of these splits.
-White and colored boxes indicate gap and non-gap symbols in the alignment, respectively.
-These sequences are then extracted from genome assemblies by annotation.
-Annotation can erroneously exclude or include segments from a sequence.
-In this case, the three N-terminal residues of the fourth sequence were excluded.
-The sequences are coded into binary symbols where white and grey boxes indicate gap and non-gap symbols, respectively.
-The emission distributions consist of two components derived from these encoded sequences.
-The first component models the gap pattern with a Markov process.
-This Markov process is in turn composed of two subprocesses where the first is a phylogenetic process, and the second is a jump process.
-These subprocess roughly correspond to changes caused by evolution and annotation, respectively.
-The second component models the propensity for those patterns to remain constant by counting the number of symbols that remain constant between columns.
+A) White and colored boxes indicate gap and non-gap symbols in the biological sequences, respectively.
+White and grey boxes indicate gap and non-gap symbols in the encoded sequences, respectively.
 c\textsubscript{0} and c\textsubscript{1} indicate the first and second columns in the alignment, respectively.
-The parameters associated with each emission component are shown to the right.
-Each state has a different set of parameters that yield its characteristic gap pattern and gap stickiness.
 B) The alignment of the representative sequences in orthologous group 2252 is decoded using the trained HMM.}
@@
 To provide a similar tree for the analysis of non-coding regions, we inferred phylogenetic trees using the GTR model of nucleotide substitution from 100 meta-alignments sampled from nucleotide alignments which were "reverse translated" from the protein alignments and their corresponding coding sequences.
-The resulting consensus tree is shown in Figure~\ref{fig:trees}B.
+The resulting consensus tree is shown in Figure \ref{fig:trees}B.
 Both trees have an identical topology, which is consistent with other published phylogenies \cite{D12GC2007, DaLage2007}.

### 2022-11-13 chapter1.tex
Moved graphical overview caption text into main text

@@
 \centering
-\caption{\textbf{Graphical summary of homology inference pipeline.}
-1) Protein sequences from annotated genomes are collected.
-The large, open circles represent the annotated genomes, and the small, filled circles represent the protein sequences associated with each annotation.
+\caption{\textbf{Graphical overview of homology inference pipeline.}
+In panels A-C, the large, open circles represent the annotated genomes, and the small, filled circles represent the protein sequences associated with each annotation.
 Sequences that share homology are colored with shades of the same hue.
-2) Homology searches are conducted between all query-target pairs of genomes.
-The lines represent a statistically significant alignment between two sequences in the genome pair.
-3) The raw output from the homology searches is processed to yield best hits between pairs of sequences.
-4) The network of best hits is clustered into self-consistent orthologous groups.
-The small, filled circles and lines represent the same sequences and best hits from the previous step.
-5) Orthologous groups are grouped by their parent genes, which are represented by the two sets of clusters with warm and cool colors, respectively.
-6) Representative sequences in each orthologous group are aligned.
-Some steps are simplified or omitted for clarity.}
+In panels D-F, the small, filled circles and lines represent the same sequences and best hits from the previous steps.}
 \label{fig:graphical_overview}
@@
 \subsection*{Pipeline overview}
-% TODO: Add description of pipeline, adapting text from Figure 1 caption
+The orthology inference pipeline is summarized in Figure \ref{fig:graphical_overview}.
+Overall, the approach is similar to other graph-based methods of orthology inference.
+First, protein sequences from annotated genomes are collected (Fig.~\ref{fig:graphical_overview}A), and homology searches are then conducted between all query-target pairs of genomes (Fig.~\ref{fig:graphical_overview}B).
+The raw output from these homology searches is processed to yield best hits between pairs of sequences (Fig.~\ref{fig:graphical_overview}C).
+Next, the network of best hits is clustered into self-consistent orthologous groups (Fig.~\ref{fig:graphical_overview}D).
+Since genes can have multiplied associated isoforms, we then implemented a novel second clustering step where orthologous groups are grouped by their parent genes, which are represented by the two sets of clusters with warm and cool colors, respectively (Fig.~\ref{fig:graphical_overview}E).
+Finally, representative sequences in each orthologous group are aligned (Fig.~\ref{fig:graphical_overview}F).
+In the following sections, we discuss each of these and other steps which were omitted for clarity in greater detail.
 

### 2022-11-14 figures/components/components.py
Added additional plots to components figure

@@ -1,11 +1,13 @@
 """Make figure of components."""
 
 import os
-from math import floor, log10
+from math import ceil, floor, log10
 
 import matplotlib as mpl
 import matplotlib.pyplot as plt
 import networkx as nx
+import pandas as pd
+from matplotlib.figure import SubplotParams
 from matplotlib.gridspec import GridSpec
 from numpy import linspace
 
@@ -17,7 +19,10 @@ component_id_groups = [('000C', '001E'),
                        ('002B', '002F')]
 
 fig_width = 7.5
-fig_height = 3.75
+fig_height = 5.625
+wspace = 0
+hspace = 0
+height_ratios = [2, 1]
 dpi = 300
 
 margin_width = 0
@@ -37,10 +42,18 @@ edge_width2 = 0.75
 cmap_base = mpl.cm.get_cmap('plasma_r', 320)  # 64 additional levels so truncated cmap has 256
 cmap = mpl.colors.ListedColormap(cmap_base(linspace(0.25, 1, 256)))
 cbar_fontsize = 8
-cbar_width = 0.125
-cbar_height = 0.02
-cbar_offset_right = 0.01
+cbar_width = 0.2
+cbar_height = 0.015
+cbar_offset_right = 0.02
 cbar_offset_bottom = 0.075
+
+# Load sequence data
+ppid2data = {}
+with open('../../orthology_inference/analysis/ortho_search/sequence_data/out/sequence_data.tsv') as file:
+    field_names = file.readline().rstrip('\n').split('\t')
+    for line in file:
+        fields = {key: value for key, value in zip(field_names, line.rstrip('\n').split('\t'))}
+        ppid2data[fields['ppid']] = (fields['gnid'], fields['spid'])
 
 # Load graph
 graph = {}
@@ -57,15 +70,23 @@ with open('../../orthology_inference/analysis/ortho_cluster2/connect_hit_graph/o
         fields = {key: value for key, value in zip(field_names, line.rstrip('\n').split('\t'))}
         components[fields['component_id']] = set(fields['ppids'].split(','))
 
+# Calculate component stats
+rows = []
+for component_id, component in components.items():
+    ppidnum = len(component)
+    gnidnum = len(set([ppid2data[ppid][0] for ppid in component]))
+    spidnum = len(set([ppid2data[ppid][1] for ppid in component]))
[…]

### 2022-11-15 figures/components/components.py
Changed one of the component_id_groups

@@ -13,7 +13,7 @@ from numpy import linspace
 
 component_id_groups = [('000C', '001E'),
                        ('0030', '0031'),
-                       ('0022', '002F'),
+                       ('0017', '005F'),
                        ('000F', '0013'),
                        ('0007', '0004'),
                        ('002B', '002F')]

### 2022-11-15 chapter1.tex
Changed minor awkward wording
Re-wrote section on connected components to include more discussion of structure of network

@@
 melanogaster} annotation was downloaded from FlyBase~\cite{Gramates2022}.
-The complete list of genomes is given in Table~\ref{stable:genomes}.
+In total, the input data consists of 33 genomes, which are listed in Table~\ref{stable:genomes}.
 Many genes have transcripts that differ only in their UTRs, and as a result there are many duplicate protein sequences in the annotations.
@@
 \subsection*{Clustering in orthologous groups}
-% TODO: Add description of structure of connected components to motivate clustering algorithm
-The best hits between sequences are naturally visualized as a graph where the sequences are nodes and best hits are edges between nodes (Fig.~\ref{fig:components}).
-Some sets of nodes form tight clusters whereas others are more loosely associated or even isolated.
-Automated genome annotation is not completely accurate, so some poorly supported hits are likely spurious, arising from homologous regions that were falsely annotated as genic in two closely related genomes.
-Others are possibly novel genes.
+The best hits between sequences are naturally visualized as a graph where sequences are nodes and best hits are edges between nodes.
+Two connected components, sets of nodes joined by a sequence of edges, are shown in Figure \ref{fig:components}.
+The sequences in panel A all contain C2H2 zinc fingers whereas the sequences in panel B are members of the Par-1 family of serine/threonine protein kinases.
+In both components, some sets of nodes have a high density of edges, forming distinct clusters, whereas other nodes are only sparsely connected to their neighbors.
+To better understand the structure of these two components, we calculated the number of sequences, unique genes, and unique species in each.
+The first has 385, 346, and 33 sequences, genes, and species, respectively, and the second has 222, 33, and 33 sequences, genes, and species, respectively.
+We then plotted the relationship between the number sequences and unique genes across all components to see if this pattern holds true generally (Fig.~\ref{fig:components}C).
+Two distinct trendlines are apparent.
+The first increases linearly with the number of sequences with a slope of one, indicating each sequence is generally associated with a unique gene.
+The second is constant with an intercept of 33, indicating the number of unique genes quickly saturates at the total number of genomes.
+Thus, there are generally two classes of components.
+The first is composed of many distinct genes, whereas the second is composed of many different isoforms of a single group of genes.
+
+The diffuse networks observed in the first component class are likely the result of a combination of factors, including rapid evolution, gene duplication, and annotation errors.
 Regardless of their origin, these hits are not strong candidates for comparative analyses since an orthology relationship is supported by relatively few genome pairs.
-In contrast, likely orthologs should consistently identify each other as best reciprocal hits across many genome pairs.
+Instead, likely orthologs should consistently identify each other as best reciprocal hits across many genome pairs.
+The same is true of the hits in the second component class.
+Although the genes as a unit form a single orthologous group, sequences with few hits are likely non-conserved or tissue-specific isoforms.
 Thus, orthologous groups can be operationally defined as self-consistent clusters in the hit graph.
-Since sequence divergence or assembly and annotation errors may prevent a best reciprocal hit between orthologous sequences across all genome pairs, the clustering method should require a high degree of self-consistency without demanding complete consensus.
+However, sequence divergence or assembly and annotation errors may prevent a best reciprocal hit between orthologous sequences across all genome pairs.
+In fact, although the most common number of reciprocal hits is 32, one fewer than the total number of genomes, many sequences have fewer (Fig.~\ref{fig:components}D).
+Thus, the clustering method should require a high degree of self-consistency without demanding complete consensus.
 
 \begin{figure}[h!]
-\includegraphics[width=\textwidth]{components/out/0022-002F.png}
+\includegraphics[width=\textwidth]{components/out/0017-005F.png}
 \centering
 \caption{\textbf{Selected connected components of hit graph.}
-The best hits are shown as a graph where nodes are sequences and edges are best hits between sequences that are colored by their bit score.
-Two connected components, sets of nodes joined by a sequence of edges, are shown.
-The sequences in panels A are annotated as potassium channels whereas the sequences in panel B are uncharacterized.
-Some sets of nodes have a high density of edges, forming distinct clusters.
-Other nodes are only sparsely connected to their neighbors.
-Components can contain nodes of both types.}
+Panels A and B are two distinct connected components of the graph formed by the best hits between sequences. Edges are colored by the value of their bit score.}
 \label{fig:components}
@@
 Gene groups were then taken as the connected components of the resulting graph.
+% TODO: More description about results of this step
 
@@
 These segments are easily identified in regions which are otherwise highly conserved, so they are also likely the result of incorrect or incomplete annotations.
-Unfortunately, they can also span more divergent regions, which complicates a simple rule- or threshold-based approach to identifying them.
+Unfortunately, they can also span more divergent regions, which complicates a simple rule- or threshold-based approach for identifying them.
 Thus, another phylo-HMM was trained to label each position in a sequence as generated by either a ``missing'' or ``not missing'' state.

### 2022-11-15 chapter1.tex
Re-wrote various captions and figure references to make them sound more like a journal

@@
 \centering
-\caption{\textbf{Graphical overview of homology inference pipeline.}
-In panels A-C, the large, open circles represent the annotated genomes, and the small, filled circles represent the protein sequences associated with each annotation.
+\caption{\textbf{Overview of homology inference pipeline.}
+\textbf{(A-C)} The large, open circles represent the annotated genomes, and the small, filled circles represent the protein sequences associated with each annotation.
 Sequences that share homology are colored with shades of the same hue.
-In panels D-F, the small, filled circles and lines represent the same sequences and best hits from the previous steps.}
+\textbf{(D-F)} The small, filled circles and lines represent the same sequences and best hits from the previous steps.}
 \label{fig:graphical_overview}
@@
 \subsection*{Pipeline overview}
-The orthology inference pipeline is summarized in Figure \ref{fig:graphical_overview}.
-Overall, the approach is similar to other graph-based methods of orthology inference.
+Our pipeline follows a similar overall approach to other graph-based methods of orthology inference.
 First, protein sequences from annotated genomes are collected (Fig.~\ref{fig:graphical_overview}A), and homology searches are then conducted between all query-target pairs of genomes (Fig.~\ref{fig:graphical_overview}B).
@@
 The best hits between sequences are naturally visualized as a graph where sequences are nodes and best hits are edges between nodes.
-Two connected components, sets of nodes joined by a sequence of edges, are shown in Figure \ref{fig:components}.
-The sequences in panel A all contain C2H2 zinc fingers whereas the sequences in panel B are members of the Par-1 family of serine/threonine protein kinases.
+Two connected components, sets of nodes joined by a sequence of edges, are shown (Fig.~\ref{fig:components}A,B).
+The sequences in the first (Fig.~\ref{fig:components}A) all contain C2H2 zinc fingers, whereas the sequences in the second (Fig.~\ref{fig:components}B) are members of the Par-1 family of serine/threonine protein kinases.
 In both components, some sets of nodes have a high density of edges, forming distinct clusters, whereas other nodes are only sparsely connected to their neighbors.
@@
 \centering
-\caption{\textbf{Selected connected components of hit graph.}
-Panels A and B are two distinct connected components of the graph formed by the best hits between sequences. Edges are colored by the value of their bit score.}
+\caption{\textbf{Selected connected components of hit graph and summary statistics.}
+\textbf{(A-B)} Two distinct connected components of the hit graph.
+Edges are colored by the value of their bit score.
+\textbf{(C)} Hexbin plot of the number of sequences and the number of unique genes in each component.
+\textbf{(D)} Histogram of number edges associated with each sequence, \textit{i.e.} the degree of each node.}
 \label{fig:components}
@@
 \caption{\textbf{Clique percolation algorithm.}
-A) A \textit{k}-clique is a set of \textit{k} nodes which are fully connected, i.e., there is an edge between pair of nodes.
+\textbf{(A)} A \textit{k}-clique is a set of \textit{k} nodes which are fully connected, i.e., there is an edge between pair of nodes.
 Cliques are shown for \textit{k} equal to three, four, five, and six.
 \textit{k} equal to one and two correspond to a single node and two nodes joined by an edge, respectively.
-B) The clique percolation algorithm detects clusters by “rolling” a clique of some size \textit{k} over the graph.
+\textbf{(B)} The clique percolation algorithm detects clusters by “rolling” a clique of some size \textit{k} over the graph.
 More specifically, a cluster is initiated when a set of nodes which form a \textit{k}-clique is identified.
@@
 In this example, \textit{k} was set to four.
-C) A single component clustered by clique percolation with varying values of \textit{k} is shown.
+\textbf{(C)} A single component clustered by clique percolation with varying values of \textit{k} is shown.
 Larger values of \textit{k} define tighter clusters.
@@
 Furthermore, since profile HMMs explicitly model the beginnings and ends of alignments, they can fully exclude these segments from the termini.
-This effect is visible in the three examples shown in Figure \ref{fig:realignment}, where each has long poorly supported segments that are completely removed.
+This effect is visible in the illustrated examples, where each has long poorly supported segments that are completely removed (Fig.~\ref{fig:realignment}).
 
@@
 \caption{\textbf{Alignment with long poorly supported segments.}
-The alignments of representative sequences in orthologous groups 23D9 (A), 01E0 (B), and 07C3 (C) are shown before and after refinement.}
+The alignments of representative sequences in orthologous groups 23D9 \textbf{(A)}, 01E0 \textbf{(B)}, and 07C3 \textbf{(C)} before and after refinement.}
 \label{fig:realignment}
@@
 For simplicity, however, we refer to the states that generate each type of region as 1A, 1B, 2, and 3, respectively.
-The probability distributions for each state shared a common structure, which is illustrated in Figure \ref{fig:hmm}A.
-We conceptualized the observed alignments as the superposition of two distinct processes.
+To model probability distributions for each state, we first conceptualized the observed alignments as the superposition of two distinct processes (Fig.~\ref{fig:hmm}A, left).
 The first is a phylogenetic process which evolves and splits a single ancestral sequence over time according to a tree.
@@
 The result is an alignment of annotated sequences which contains evolutionary information obscured by "noise" from the annotation process, shown here by the exclusion of three N-terminal residues in the fourth annotated sequence.
-To simplify modeling this behavior with an HMM, we first coded the sequences into binary symbols.
-The distributions for each state then consisted of two components derived from the encoded sequences.
+To simplify modeling this behavior with an HMM, we coded the sequences into binary symbols.
+The distributions for each state then consisted of two components derived from the encoded sequences (Fig.~\ref{fig:hmm}A, right).
 The first component models the gap pattern with a Markov process.
@@
 Because this first component did not fully capture the propensity for the gap patterns to remain constant, we included a second component that models the "gap stickiness" as a beta-binomial random variable by counting the number of symbols that remain constant between columns.
-The parameters associated with each component are shown to the right of Figure \ref{fig:hmm}A.
-Each state has a different set of parameters that yield its characteristic gap pattern and gap stickiness.
+Each component is associated with a set of parameters, and the unique parameters for each state yield its characteristic gap pattern and gap stickiness.
 
@@
 In contrast, columns assigned to states 2 and 3 largely corresponded to the long poorly supported segments and phylogenetically discordant regions discussed previously and were therefore removed from the alignments.
-An example decoded alignment is shown in Figure \ref{fig:hmm})B where the decoded states closely follow the expected patterns.
+In the example decoded alignment shown, the decoded states closely follow the expected patterns (Fig.~ref{fig:hmm})B).
 Overall, fewer than 10\% of alignments were trimmed of any regions.
@@
 \caption{\textbf{HMM emission architecture and a decoded alignment.}
-A) White and colored boxes indicate gap and non-gap symbols in the biological sequences, respectively.
+\textbf{(A)} Schematic of theoretical alignment generating process and corresponding probabilistic components in HMM.
+White and colored boxes indicate gap and non-gap symbols in the biological sequences, respectively.
 White and grey boxes indicate gap and non-gap symbols in the encoded sequences, respectively.
 c\textsubscript{0} and c\textsubscript{1} indicate the first and second columns in the alignment, respectively.
-B) The alignment of the representative sequences in orthologous group 2252 is decoded using the trained HMM.}
+The parameters associated with each component are shown to the right.
+\textbf{(B)} The alignment of the representative sequences in orthologous group 2252 decoded using the trained HMM.}
 \label{fig:hmm}
@@
 We then combined them into a single consensus tree (Fig.~\ref{fig:trees}A).
-To provide a similar tree for the analysis of non-coding regions, we inferred phylogenetic trees using the GTR model of nucleotide substitution from 100 meta-alignments sampled from nucleotide alignments which were "reverse translated" from the protein alignments and their corresponding coding sequences.
-The resulting consensus tree is shown in Figure \ref{fig:trees}B.
+To provide a similar tree for the analysis of non-coding regions, we inferred phylogenetic trees using the GTR model of nucleotide substitution from 100 meta-alignments sampled from nucleotide alignments which were "reverse translated" from the protein alignments and their corresponding coding sequences (\ref{fig:trees}B).
 Both trees have an identical topology, which is consistent with other published phylogenies \cite{D12GC2007, DaLage2007}.
@@
 \caption{\textbf{Phylogenetic tree of species.}
-A) The LG model was fit to meta-alignments directly sampled from the original protein alignments B).
-The GTR model was fit to meta-alignments sampled from "reverse translated" nucleotide alignments.
+\textbf{(A)} Consensus tree from LG model fit to meta-alignments directly sampled from the original protein alignments.
+\textbf{(B)} Consensus tree from GTR model fit to meta-alignments sampled from "reverse translated" nucleotide alignments.
 Values at nodes are bootstrap percentages.}
@@
 \caption{\textbf{Statistics of orthologous groups.}
-A) Each species is equally represented in orthologous groups (OGs).
-B) A plurality of orthologous groups contain all species.
-C) Nearly all proteins are associated with a single orthologous group.
-D) The number of orthologous groups associated with a species is strongly correlated with the number of unique annotated proteins, which suggests the annotation pipeline generally identifies conserved genes.}
+\textbf{(A)} Each species is equally represented in orthologous groups (OGs).
+\textbf{(B)} A plurality of orthologous groups contain all species.
+\textbf{(C)} Nearly all proteins are associated with a single orthologous group.
+\textbf{(D)} The number of orthologous groups associated with a species is strongly correlated with the number of unique annotated proteins, which suggests the annotation pipeline generally identifies conserved genes.}
 \label{sfig:clusters}
@@
 \caption{\textbf{Addition of paralogs to orthologous groups.}
-A) Most orthologous groups (OGs) have no in-paralogs.
-B, D) Of the groups with paralogs, most have fewer than five.
-C) The in-paralogs are generally only a small fraction of the sequences in an orthologous group.}
+\textbf{(A)} Most orthologous groups (OGs) have no in-paralogs.
+\textbf{(B, D)} Of the groups with paralogs, most have fewer than five.
+\textbf{(C)} The in-paralogs are generally only a small fraction of the sequences in an orthologous group.}
 \label{sfig:paralogs}
@@
 \caption{\textbf{CNN data and training details.}
-A) Most labels were “negative,” which corresponded to positions which were maintained in the alignment.
-B, C) The training loss and metrics stabilized by the final training epochs.
-D, E) The parameters for a model of this size are easily visualized.
+\textbf{(A)} Most labels were “negative,” which corresponded to positions which were maintained in the alignment.
+\textbf{(B, C)} The training loss and metrics stabilized by the final training epochs.
+\textbf{(D, E)} The parameters for a model of this size are easily visualized.
 The embedding learns the gap and unknown amino acid symbols are distinct from the others.
 The convolutional filters learn kernels which correspond to the length of a typical long poorly supported segment.
-F) The alignment of representative sequences in orthologous group 23D9 where the indicated sequence is labeled with the trained CNN.
+\textbf{(F)} The alignment of representative sequences in orthologous group 23D9 where the indicated sequence is labeled with the trained CNN.
 The blue lines are the CNN’s predictions, and the yellow lines are the labels.
@@
 \caption{\textbf{Insertion phylo-HMM data and training details.}
-A) Most columns in the training data were labeled as state 1A or 1B.
-B-F) The model loss and parameters largely stabilized by the final training iteration.
-Panels C-F correspond to the parameters in the phylogenetic process, the jump process, the pattern stickiness model, and the transition matrix, respectively.
+\textbf{(A)} Most columns in the training data were labeled as state 1A or 1B.
+\textbf{(B)} The model loss stabilized by the final training iteration.
+\textbf{(C-F)} The values of parameters in the phylogenetic process, the jump process, the pattern stickiness model, and the transition matrix, respectively, at each training iteration.
 The transition matrix plots are the transition rates to the state indicated on the vertical axis and given in log scale.
@@
 \caption{\textbf{Insertion phylo-HMM trimming details.}
-A) Most alignments were not trimmed of any columns.
-B) Most alignments have fewer than 5 trimmed regions which cumulatively account for less than 10\% of the columns in the original alignment.
+\textbf{(A)} Most alignments were not trimmed of any columns.
+\textbf{(B)} Most alignments have fewer than 5 trimmed regions which cumulatively account for less than 10\% of the columns in the original alignment.
 Alignments with more trimmed columns generally have one one large trimmed region rather than many small ones.
-C) Most trimmed regions were inferred as state 3.}
+\textbf{(C)} Most trimmed regions were inferred as state 3.}
 \label{sfig:hmm_trim}

### 2022-11-15 chapter1.tex
Added additional details to gene clustering section

@@
 First, a gene overlap graph was constructed by defining an edge between orthologous groups if the intersection of their associated sets of parent genes is at least 50\% of the smaller of the two.
-Gene groups were then taken as the connected components of the resulting graph.
-% TODO: More description about results of this step
+Gene groups were then taken as the connected components of the resulting graph, yielding 14,909 groups.
+This is commensurate with the roughly 15,000 genes in each genome, which suggests this approach has successfully clustered orthologous groups derived from a common set of parent genes.
 

### 2022-11-15 chapter1.tex
Added more details and explanation to clique percolation section

@@
 Graph-based approaches have remained popular, and in the intervening years many other researchers have refined this method by implementing various pre- and post-processing steps.
-Despite these improvements, many databases and pipelines use the same triangle clustering algorithm or other methods which require relatively few hits between sequences to infer an orthologous group, \textit{e.g.} connected components or Markov clustering~\cite{Remm2001, Li2003, Jensen2007, Linard2011, Emms2015, Train2017, Cosentino2018}.
+Despite these improvements, many databases and pipelines use the same triangle clustering algorithm or other methods which require relatively few hits between sequences to infer an orthologous group, \textit{e.g.} connected components or Markov clustering~\cite{Remm2001, Enright2002, Li2003, Jensen2007, Linard2011, Emms2015, Train2017, Cosentino2018}.
 However, the scale of biological sequence data has changed dramatically.
@@
 
-% TODO: Add more discussion of different clustering methods and their weaknesses (include citations)
 The identification of sets of densely connected nodes in graphs is known as community detection in network analysis.
-There are many community detection algorithms available, but a popular approach in the context of orthology inference is building clusters progressively by identifying nodes that form a triangle with at least two other nodes in the cluster.
-While this method is robust when clustering smaller or more diverse sets of genomes, the large number of closely related genomes in the input data yields many nodes which form triangles but lack dense networks of edges with other nodes.
-However, triangle clustering is a special case of the more general \textit{k}-clique percolation algorithm where \textit{k} = 3 (Fig.~\ref{fig:percolation}).
-Higher values of \textit{k} require greater overlap between a candidate node and those already in the cluster and therefore produce tighter clusters at the cost of excluding more speculative orthology relationships.
+While many community detection algorithms are available, only some are commonly used in the context of orthology inference.
+One early method that remains popular is building clusters progressively by identifying nodes that form a triangle with at least two other nodes in the cluster \cite{Tatusov1997, Jensen2007}.
+Other approaches include the MCL algorithm, which clusters graphs by similating stochastic flow, and connected components or other single-linkage criteria \cite{Remm2001, Enright2002, Li2003, Emms2015, Train2017, Cosentino2018}.
+While these methods are robust when clustering hit graphs derived from smaller or more diverse sets of genomes, they are not suitable for the large number of closely related genomes in this work since they require relatively few edges to define a cluster.
+For example, the MCL algorithm and connected components method assign a node to a cluster as long as it has a single edge, and triangle clustering only requires two edges to two adjacent nodes.
+
+However, connected components and triangle clustering are special cases of the more general \textit{k}-clique percolation algorithm where \textit{k} equals two and three, respectively.
+The clique percolation algorithm detects clusters by first identifying cliques, sets of nodes which are fully connected, of a specified size \textit{k} in the graph (Fig.~\ref{fig:percolation}A).
+Clusters are then taken as the connected components of an overlap graph where an edge exists between two cliques if they share \textit{k}-1 nodes in common.
+An intuitive way to visualize this algorithm is by “rolling” a clique of some size \textit{k} over the graph (Fig.~\ref{fig:percolation}B).
+More specifically, a cluster is initiated when a set of nodes which form a \textit{k}-clique is identified.
+The cluster expands by shifting the \textit{k}-clique to an adjacent \textit{k}-clique that shares \textit{k}-1 nodes in common with the current \textit{k}-clique.
+A cluster stops expanding when there are no adjacent \textit{k}-cliques, and the algorithm terminates when there are no \textit{k}-cliques which are not part of a cluster.
+The strength of this algorithm is its ability to exclude sparsely connected nodes from clusters with an easily tunable parameter \textit{k}.
+Higher values of \textit{k} require greater overlap between a candidate node and those already in the cluster and therefore produce tighter clusters at the cost of excluding more speculative orthology relationships (Fig.~\ref{fig:percolation}C).
 We set \textit{k} to 4 as compromise between these concerns, yielding 22,813 orthologous groups, a plurality of which contained all 33 species (Fig.~\ref{sfig:clusters}).
@@
 \caption{\textbf{Clique percolation algorithm.}
-\textbf{(A)} A \textit{k}-clique is a set of \textit{k} nodes which are fully connected, i.e., there is an edge between pair of nodes.
-Cliques are shown for \textit{k} equal to three, four, five, and six.
+\textbf{(A)} Cliques for \textit{k} equal to three, four, five, and six.
 \textit{k} equal to one and two correspond to a single node and two nodes joined by an edge, respectively.
-\textbf{(B)} The clique percolation algorithm detects clusters by “rolling” a clique of some size \textit{k} over the graph.
-More specifically, a cluster is initiated when a set of nodes which form a \textit{k}-clique is identified.
-The cluster expands by shifting the \textit{k}-clique to an adjacent \textit{k}-clique that shares \textit{k}-1 nodes in common with the current \textit{k}-clique.
-A cluster stops expanding when there are no adjacent \textit{k}-cliques, and the algorithm terminates when there are no \textit{k}-cliques which are not part of a cluster.
-In this example, \textit{k} was set to four.
-\textbf{(C)} A single component clustered by clique percolation with varying values of \textit{k} is shown.
-Larger values of \textit{k} define tighter clusters.
+\textbf{(B)} Illustration of clique percolation algorithm.
+\textbf{(C)} A single component clustered by clique percolation with varying values of \textit{k}.
 Nodes are colored according to their cluster.

### 2022-11-16 chapter1.tex
Fixed using bad quotation marks

@@
 As a result, some genomes are annotated with different or multiple splice isoforms of a given orthologous gene, which can create complex networks in the resulting hit graph.
-For example, if the genomes are variably annotated with one or both of two distinct isoforms, the resulting graph may contain two clusters connected by a "bridge" formed by the genomes which contain only one of the isoforms.
+For example, if the genomes are variably annotated with one or both of two distinct isoforms, the resulting graph may contain two clusters connected by a ``bridge'' formed by the genomes which contain only one of the isoforms.
 If the nodes bridging the two clusters form cliques with themselves and the clusters, the clique percolation algorithm will merge all the nodes into a single orthologous group where some genes have multiple associated sequences.
@@
 The second is the annotation process which can erroneously exclude or include segments from a sequence.
-The result is an alignment of annotated sequences which contains evolutionary information obscured by "noise" from the annotation process, shown here by the exclusion of three N-terminal residues in the fourth annotated sequence.
+The result is an alignment of annotated sequences which contains evolutionary information obscured by ``noise'' from the annotation process, shown here by the exclusion of three N-terminal residues in the fourth annotated sequence.
 To simplify modeling this behavior with an HMM, we coded the sequences into binary symbols.
@@
 These subprocess roughly correspond to changes caused by evolution and annotation, respectively.
-Because this first component did not fully capture the propensity for the gap patterns to remain constant, we included a second component that models the "gap stickiness" as a beta-binomial random variable by counting the number of symbols that remain constant between columns.
+Because this first component did not fully capture the propensity for the gap patterns to remain constant, we included a second component that models the ``gap stickiness'' as a beta-binomial random variable by counting the number of symbols that remain constant between columns.
 Each component is associated with a set of parameters, and the unique parameters for each state yield its characteristic gap pattern and gap stickiness.
@@
 We then combined them into a single consensus tree (Fig.~\ref{fig:trees}A).
-To provide a similar tree for the analysis of non-coding regions, we inferred phylogenetic trees using the GTR model of nucleotide substitution from 100 meta-alignments sampled from nucleotide alignments which were "reverse translated" from the protein alignments and their corresponding coding sequences (\ref{fig:trees}B).
+To provide a similar tree for the analysis of non-coding regions, we inferred phylogenetic trees using the GTR model of nucleotide substitution from 100 meta-alignments sampled from nucleotide alignments which were ``reverse translated'' from the protein alignments and their corresponding coding sequences (\ref{fig:trees}B).
 Both trees have an identical topology, which is consistent with other published phylogenies \cite{D12GC2007, DaLage2007}.
@@
 \textbf{(A)} Consensus tree from LG model fit to meta-alignments directly sampled from the original protein alignments.
-\textbf{(B)} Consensus tree from GTR model fit to meta-alignments sampled from "reverse translated" nucleotide alignments.
+\textbf{(B)} Consensus tree from GTR model fit to meta-alignments sampled from ``reverse translated'' nucleotide alignments.
 Values at nodes are bootstrap percentages.}
@@
 The orthologous groups and alignments yielded by this pipeline are a valuable resource for comparative studies of gene birth/death processes and protein evolution at the level of both entire proteomes and specific gene families in the \textit{Drosophila} genus.
-To assist these efforts, the final alignments after curation and the labels from the "missing" phylo-HMM are provided as supplemental data.
+To assist these efforts, the final alignments after curation and the labels from the ``missing'' phylo-HMM are provided as supplemental data.
 Although this work focused on single copy orthologs, other studies may require different subsets of orthologous groups that demand other pre-processing and alignment strategies.

### 2022-11-16 chapter1.tex
Added value of k in percolation figure panel B and fixed comma outside quotation marks

@@
 \textit{k} equal to one and two correspond to a single node and two nodes joined by an edge, respectively.
-\textbf{(B)} Illustration of clique percolation algorithm.
+\textbf{(B)} Illustration of clique percolation algorithm where \textit{k = 4}.
 \textbf{(C)} A single component clustered by clique percolation with varying values of \textit{k}.
@@
 
-Though the phylo-HMM removed phylogenetically incongruent ``insertions'', some sequences still contained extensive segments of uninterrupted gaps.
+Though the phylo-HMM removed phylogenetically incongruent ``insertions,'' some sequences still contained extensive segments of uninterrupted gaps.
 These segments are easily identified in regions which are otherwise highly conserved, so they are also likely the result of incorrect or incomplete annotations.

### 2022-11-17 chapter1.tex
Re-worded clustering algorithm discussion to apply to more generally and contrast previous methods to ours

@@
 \section*{Discussion}
-% TODO: Highlight specific contributions
 The orthologous groups and alignments yielded by this pipeline are a valuable resource for comparative studies of gene birth/death processes and protein evolution at the level of both entire proteomes and specific gene families in the \textit{Drosophila} genus.
@@
 
-For example, despite many additional pre-processing steps and other tweaks introduced by later authors, the basic framework of orthology inference by clustering the hit graph has remained largely unchanged in the past twenty years~\cite{Remm2001, Li2003, Jensen2007, Linard2011, Emms2015, Train2017, Cosentino2018}.
+For example, despite many additional pre-processing steps and other tweaks introduced by later authors, the basic framework of orthology inference by clustering the hit graph has remained largely unchanged in the past twenty years~\cite{Tatusov1997, Remm2001, Li2003, Jensen2007, Linard2011, Emms2015, Train2017, Cosentino2018}.
 This longevity is a testament to the robustness of the underlying idea that orthologous proteins should consistently identify each other as the most similar pairs between their genomes.
-Even as the number of genomes and their taxonomic density has increased dramatically, many orthology inference pipelines continue to use triangle clustering, which was originally applied to sets of far fewer and more distantly related genomes~\cite{Tatusov1997}.
+Even as the number of genomes and their taxonomic density has increased dramatically, many orthology inference pipelines continue to use algorithms which were originally applied to sets of far fewer and more distantly related genomes.
 This mismatch in scale increases the chance of propagating annotation errors since only a small number of edges are needed to create or merge clusters.
-Fortunately, triangle clustering is a special case of the more general \textit{k}-clique percolation algorithm where \textit{k} is a tunable parameter that influences the tightness of a cluster.
+Thus, we instead applied a generalization of the triangle and connected components clustering methods called \textit{k}-clique percolation where \textit{k} is a tunable parameter that influences the tightness of a cluster.
 The optimal value of \textit{k} for a given set of genomes is unclear and likely depends on the desired trade-off between sensitivity and specificity.

### 2022-11-17 chapter1.tex
Fixed some non-breaking spaces by references

@@
 While many community detection algorithms are available, only some are commonly used in the context of orthology inference.
-One early method that remains popular is building clusters progressively by identifying nodes that form a triangle with at least two other nodes in the cluster \cite{Tatusov1997, Jensen2007}.
-Other approaches include the MCL algorithm, which clusters graphs by similating stochastic flow, and connected components or other single-linkage criteria \cite{Remm2001, Enright2002, Li2003, Emms2015, Train2017, Cosentino2018}.
+One early method that remains popular is building clusters progressively by identifying nodes that form a triangle with at least two other nodes in the cluster~\cite{Tatusov1997, Jensen2007}.
+Other approaches include the MCL algorithm, which clusters graphs by similating stochastic flow, and connected components or other single-linkage criteria~\cite{Remm2001, Enright2002, Li2003, Emms2015, Train2017, Cosentino2018}.
 While these methods are robust when clustering hit graphs derived from smaller or more diverse sets of genomes, they are not suitable for the large number of closely related genomes in this work since they require relatively few edges to define a cluster.
@@
 
-Phylogenetic HMMs (phylo-HMMs) are statistical models that incorporate phylogenetic and spatial information to calculate the probability that each observation in a sequence was generated by one of several “hidden states” \cite{Felsenstein1996}.
+Phylogenetic HMMs (phylo-HMMs) are statistical models that incorporate phylogenetic and spatial information to calculate the probability that each observation in a sequence was generated by one of several “hidden states”~\cite{Felsenstein1996}.
 Since they can evaluate both the probability of a gap pattern in a column given the known phylogenetic relationships and the local context, a phylo-HMM was used to segment the alignment into contiguous regions with different patterns of gaps.
@@
 To provide a similar tree for the analysis of non-coding regions, we inferred phylogenetic trees using the GTR model of nucleotide substitution from 100 meta-alignments sampled from nucleotide alignments which were ``reverse translated'' from the protein alignments and their corresponding coding sequences (\ref{fig:trees}B).
-Both trees have an identical topology, which is consistent with other published phylogenies \cite{D12GC2007, DaLage2007}.
+Both trees have an identical topology, which is consistent with other published phylogenies~\cite{D12GC2007, DaLage2007}.
 

### 2022-11-17 chapter1.tex
Added abstract

@@
 \section*{Abstract}
-abstract text goes here
+Identifying protein sequences with common ancestry is a core task in bioinformatics and evolutionary biology.
+However, methods for inferring and aligning such sequences in annotated genomes have not kept pace with the increasing scale and complexity of the available data.
+Thus, in this work we implemented several improvements to the traditional methodology that more fully leverage the redundancy of closely related genomes and the organization of their annotations.
+Two highlights include the application of the more flexible \textit{k}-clique percolation algorithm for identifying clusters of orthologous proteins and the development of a novel technique for removing poorly supported regions of alignments with a phylogenetic HMM.
+In making the latter, we also wrote a fully documented Python package Homomorph that implements standard HMM algorithms and created a set of tutorials to promote its use by a wide audience.
+We applied the resulting pipeline to a set of 33 annotated \textit{Drosophila} genomes, generating 22,813 orthologous groups and 8,565 high-quality alignments.
 

### 2022-11-17 chapter1/chapter1.tex
Modified chapter1 to be a subfile and fixed some minor formatting issues

--- 2022-11-17 chapter1.tex
@@
-\documentclass[10pt,letterpaper]{article}
-
-% PACKAGE CONFIGURATIONS
-\usepackage{amsmath}
-
-\usepackage{graphicx}
-\graphicspath{{./figures/}}
-
-\usepackage[margin=1in]{geometry}
-%\usepackage{setspace}  % double spacing
-%\doublespacing
-\usepackage[skip=10pt]{parskip}
-
-\usepackage{hyperref}
-
-\usepackage[doi=false,url=false]{biblatex}
-\addbibresource{refs.bib}
-
-\usepackage[labelfont=bf, labelsep=period]{caption}
-\renewcommand{\figurename}{Fig}
-
-\begin{document}
-
-% Title must be 250 characters or less
-\begin{flushleft}
-{\Large\textbf{Leveraging genomic redundancy to improve inference and alignment of orthologous proteins}}
-\newline
-\\
-Marc Singleton\textsuperscript{1},
-Michael Eisen\textsuperscript{1,2,3,*}
-\\
-\bigskip
-\textbf{1} Biophysics Graduate Group, UC Berkeley, Berkeley, CA, USA
-\\
-\textbf{2} Department of Molecular and Cell Biology, UC Berkeley, Berkeley, CA, USA
-\\
-\textbf{3} Howard Hughes Medical Institute, UC Berkeley, Berkeley, CA, USA
-\\
-\bigskip
-
-* mbeisen@berkeley.edu
-
-\end{flushleft}
-
 % Abstract must be 300 words or less
-\section*{Abstract}
+\begin{abstract}
 Identifying protein sequences with common ancestry is a core task in bioinformatics and evolutionary biology.
@@
 We applied the resulting pipeline to a set of 33 annotated \textit{Drosophila} genomes, generating 22,813 orthologous groups and 8,565 high-quality alignments.
+\end{abstract}
 
@@
 \subsection*{Initial alignment and selection of representative sequences}
-The sequences in each orthologous group were aligned using MAFFT 7.490 with the following settings: --globalpair --maxiterate 1000 --thread 1 --anysymbol --allowshift --unalignlevel 0.4 --leavegappyregion~\cite{Katoh2013}.
+The sequences in each orthologous group were aligned using MAFFT 7.490 with the following settings: -{}-globalpair -{}-maxiterate 1000 -{}-thread 1 -{}-anysymbol -{}-allowshift -{}-leavegappyregion -{}-unalignlevel 0.4~\cite{Katoh2013}.
 Representative sequences for each gene were then selected by maximum likelihood according to binary profiles constructed from these alignments.
@@
 \end{table}
-
-\end{document}

### 2022-11-17 chapter1/chapter1.tex
Moved chapter1 supporting information to separate file

@@
 We thank the past and present members of the Eisen lab for helpful discussions and their assistance in revising the manuscript.
-
-\printbibliography
-
-\clearpage
-
-\section*{Supporting information}
-\setcounter{figure}{0}
-\renewcommand{\thefigure}{S\arabic{figure}}
-
-\begin{figure}[h!]
-\includegraphics[width=0.5\textwidth]{ncbi/out/ncbi.png}
-\centering
-\caption{\textbf{Cumulative number of different eukaryotic genomes annotated by NCBI.}}
-\label{sfig:ncbi}
-\end{figure}
-
-\begin{figure}[h!]
-\includegraphics[width=\textwidth]{clusters/out/clusters.png}
-\centering
-\caption{\textbf{Statistics of orthologous groups.}
-\textbf{(A)} Each species is equally represented in orthologous groups (OGs).
-\textbf{(B)} A plurality of orthologous groups contain all species.
-\textbf{(C)} Nearly all proteins are associated with a single orthologous group.
-\textbf{(D)} The number of orthologous groups associated with a species is strongly correlated with the number of unique annotated proteins, which suggests the annotation pipeline generally identifies conserved genes.}
-\label{sfig:clusters}
-\end{figure}
-
-\begin{figure}[h!]
-\includegraphics[width=\textwidth]{paralogs/out/paralogs.png}
-\centering
-\caption{\textbf{Addition of paralogs to orthologous groups.}
-\textbf{(A)} Most orthologous groups (OGs) have no in-paralogs.
-\textbf{(B, D)} Of the groups with paralogs, most have fewer than five.
-\textbf{(C)} The in-paralogs are generally only a small fraction of the sequences in an orthologous group.}
-\label{sfig:paralogs}
-\end{figure}
-
-\begin{figure}[h!]
-\includegraphics[width=\textwidth]{cnn_architecture.png}
-\centering
-\caption{\textbf{CNN architecture.}}
-\label{sfig:cnn_architecture}
-\end{figure}
-
-\begin{figure}[h!]
-\includegraphics[width=\textwidth]{cnn_training/out/cnn_training.png}
-\includegraphics[width=\textwidth]{cnn_training/out/23D9_XP_026832050.1.png}
-\centering
-\caption{\textbf{CNN data and training details.}
-\textbf{(A)} Most labels were “negative,” which corresponded to positions which were maintained in the alignment.
-\textbf{(B, C)} The training loss and metrics stabilized by the final training epochs.
-\textbf{(D, E)} The parameters for a model of this size are easily visualized.
-The embedding learns the gap and unknown amino acid symbols are distinct from the others.
-The convolutional filters learn kernels which correspond to the length of a typical long poorly supported segment.
-\textbf{(F)} The alignment of representative sequences in orthologous group 23D9 where the indicated sequence is labeled with the trained CNN.
-The blue lines are the CNN’s predictions, and the yellow lines are the labels.
-The red lines are the weights given to each position.
-The regions where the weights are zero are masked from the training data, which allowed the model to focus on only the most relevant examples.}
-\label{sfig:cnn_training}
-\end{figure}
-
-\begin{figure}[h!]
-\includegraphics[width=\textwidth]{hmm_training/out/hmm_training.png}
-\centering
-\caption{\textbf{Insertion phylo-HMM data and training details.}
-\textbf{(A)} Most columns in the training data were labeled as state 1A or 1B.
-\textbf{(B)} The model loss stabilized by the final training iteration.
-\textbf{(C-F)} The values of parameters in the phylogenetic process, the jump process, the pattern stickiness model, and the transition matrix, respectively, at each training iteration.
-The transition matrix plots are the transition rates to the state indicated on the vertical axis and given in log scale.
-Self transitions are excluded.}
-\label{sfig:hmm_training}
-\end{figure}
-
-\begin{figure}[h!]
-\includegraphics[width=\textwidth]{hmm_trim/out/hmm_trim.png}
-\centering
-\caption{\textbf{Insertion phylo-HMM trimming details.}
-\textbf{(A)} Most alignments were not trimmed of any columns.
-\textbf{(B)} Most alignments have fewer than 5 trimmed regions which cumulatively account for less than 10\% of the columns in the original alignment.
-Alignments with more trimmed columns generally have one one large trimmed region rather than many small ones.
-\textbf{(C)} Most trimmed regions were inferred as state 3.}
-\label{sfig:hmm_trim}
-\end{figure}
-
-\begin{figure}[h!]
-\includegraphics[width=\textwidth]{trees/out/trees2_LG.png}
-\centering
-\caption{\textbf{Phylogenetic trees created by different sampling strategies under LG model.}}
-\label{sfig:trees_LG}
-\end{figure}
-
-\begin{figure}[h!]
-\includegraphics[width=\textwidth]{trees/out/trees2_GTR.png}
-\centering
-\caption{\textbf{Phylogenetic trees created by different sampling strategies under GTR model.}}
-\label{sfig:trees_GTR}
-\end{figure}
-
-\setcounter{table}{0}
-\renewcommand{\thetable}{S\arabic{table}}
-
-\begin{table}[h!]
-\centering
-\caption{\textbf{Genome annotations.}}
-\begin{tabular}{|l|l|l|l|l|}
-\hline
-\textbf{Species}                       & \textbf{Species ID} & \textbf{Taxon ID} & \textbf{Version} & \textbf{Source} \\ \hline
-\textit{Drosophila ananassae}          & dana                & 7217              & 102              & NCBI            \\ \hline
-\textit{Drosophila biarmipes}          & dbia                & 125945            & 102              & NCBI            \\ \hline
-\textit{Drosophila bipectinata}        & dbip                & 42026             & 102              & NCBI            \\ \hline
-\textit{Drosophila elegans}            & dele                & 30023             & 102              & NCBI            \\ \hline
-\textit{Drosophila erecta}             & dere                & 7220              & 101              & NCBI            \\ \hline
-\textit{Drosophila eugracilis}         & deug                & 29029             & 102              & NCBI            \\ \hline
-\textit{Drosophila ficusphila}         & dfic                & 30025             & 102              & NCBI            \\ \hline
-\textit{Drosophila grimshawi}          & dgri                & 7222              & 103              & NCBI            \\ \hline
-\textit{Drosophila guanche}            & dgua                & 7266              & 100              & NCBI            \\ \hline
-\textit{Drosophila hydei}              & dhyd                & 7224              & 101              & NCBI            \\ \hline
-\textit{Drosophila innubila}           & dinn                & 198719            & 100              & NCBI            \\ \hline
-\textit{Drosophila kikkawai}           & dkik                & 30033             & 102              & NCBI            \\ \hline
-\textit{Drosophila mauritiana}         & dmau                & 7226              & 100              & NCBI            \\ \hline
-\textit{Drosophila melanogaster}       & dmel                & 7227              & FB2022\_02       & FlyBase         \\ \hline
-\textit{Drosophila mojavensis}         & dmoj                & 7230              & 102              & NCBI            \\ \hline
-\textit{Drosophila navojoa}            & dnav                & 7232              & 101              & NCBI            \\ \hline
-\textit{Drosophila novamexicana}       & dnov                & 47314             & 100              & NCBI            \\ \hline
-\textit{Drosophila obscura}            & dobs                & 7282              & 101              & NCBI            \\ \hline
-\textit{Drosophila persimilis}         & dper                & 7234              & 101              & NCBI            \\ \hline
-\textit{Drosophila pseudoobscura}      & dpse                & 7237              & 104              & NCBI            \\ \hline
-\textit{Drosophila rhopaloa}           & drho                & 1041015           & 102              & NCBI            \\ \hline
-\textit{Drosophila santomea}           & dsan                & 129105            & 101              & NCBI            \\ \hline
-\textit{Drosophila sechellia}          & dsec                & 7238              & 101              & NCBI            \\ \hline
-\textit{Drosophila serrata}            & dser                & 7274              & 100              & NCBI            \\ \hline
-\textit{Drosophila simulans}           & dsim                & 7240              & 103              & NCBI            \\ \hline
-\textit{Drosophila subobscura}         & dsob                & 7241              & 100              & NCBI            \\ \hline
-\textit{Drosophila subpulchrella}      & dspu                & 1486046           & 100              & NCBI            \\ \hline
-\textit{Drosophila suzukii}            & dsuz                & 28584             & 102              & NCBI            \\ \hline
-\textit{Drosophila takahashii}         & dtak                & 29030             & 102              & NCBI            \\ \hline
-\textit{Drosophila teissieri}          & dtei                & 7243              & 100              & NCBI            \\ \hline
-\textit{Drosophila virilis}            & dvir                & 7244              & 103              & NCBI            \\ \hline
-\textit{Drosophila willistoni}         & dwil                & 7260              & 102              & NCBI            \\ \hline
-\textit{Drosophila yakuba}             & dyak                & 7245              & 102              & NCBI            \\ \hline
-\textit{Scaptodrosophila lebanonensis} & sleb                & 7225              & 100              & NCBI            \\ \hline
-\end{tabular}
-\label{stable:genomes}
-\end{table}
-
-\begin{table}[h!]
-\centering
-\caption{\textbf{Phylogenetic diversity criteria.}}
-\begin{tabular}{|l|l|}
-\hline
-\textbf{Species   IDs} & \textbf{Minimum   number} \\ \hline
-dinn,   dgri, dhyd     & 2                         \\ \hline
-dnov,   dvir           & 1                         \\ \hline
-dmoj,   dnav           & 1                         \\ \hline
-dper,   dpse           & 1                         \\ \hline
-dgua,   dsob           & 1                         \\ \hline
-dana,   dbip           & 1                         \\ \hline
-dkik,   dser           & 1                         \\ \hline
-dele,   drho           & 1                         \\ \hline
-dtak,   dbia           & 1                         \\ \hline
-dspu,   dsuz           & 1                         \\ \hline
-dere,   dtei           & 1                         \\ \hline
-dsan,   dyak           & 1                         \\ \hline
-dmel                   & 1                         \\ \hline
-dmau,   dsec, dsim     & 1                         \\ \hline
-\end{tabular}
-\label{stable:diversity_criteria}
-\end{table}

### 2022-11-17 chapter1/chapter1_supporting.tex
Moved chapter1 supporting information to separate file

--- 2022-11-17 chapter1.tex
@@
-\documentclass[10pt,letterpaper]{article}
-
-% PACKAGE CONFIGURATIONS
-\usepackage{amsmath}
-
-\usepackage{graphicx}
-\graphicspath{{./figures/}}
-
-\usepackage[margin=1in]{geometry}
-%\usepackage{setspace}  % double spacing
-%\doublespacing
-\usepackage[skip=10pt]{parskip}
-
-\usepackage{hyperref}
-
-\usepackage[doi=false,url=false]{biblatex}
-\addbibresource{refs.bib}
-
-\usepackage[labelfont=bf, labelsep=period]{caption}
-\renewcommand{\figurename}{Fig}
-
-\begin{document}
-
-% Title must be 250 characters or less
-\begin{flushleft}
-{\Large\textbf{Leveraging genomic redundancy to improve inference and alignment of orthologous proteins}}
-\newline
-\\
-Marc Singleton\textsuperscript{1},
-Michael Eisen\textsuperscript{1,2,3,*}
-\\
-\bigskip
-\textbf{1} Biophysics Graduate Group, UC Berkeley, Berkeley, CA, USA
-\\
-\textbf{2} Department of Molecular and Cell Biology, UC Berkeley, Berkeley, CA, USA
-\\
-\textbf{3} Howard Hughes Medical Institute, UC Berkeley, Berkeley, CA, USA
-\\
-\bigskip
-
-* mbeisen@berkeley.edu
-
-\end{flushleft}
-
-% Abstract must be 300 words or less
-\section*{Abstract}
-abstract text goes here
-
-\section*{Introduction}
-Comparative genomics is a powerful tool for yielding insights into evolutionary relationships, molecular function, and the forces that drive gene, genome, and population evolution.
-These methods often rely on the identification of homologous sequences or homologs, that is sequences with common ancestry, since this ensures that differences between sequences reflect variations in evolution from a common point of divergence.
-However, many analyses impose the additional condition that the sequences have diverged through speciation events (orthology) rather than duplications (paralogy) or other mechanisms such as horizontal gene transfer~\cite{Fitch1970}.
-The underlying assumption is ``orthologs'' have conserved equivalent functions whereas ``paralogs'', by virtue of their redundancy, are more likely to diverge~\cite{Ohno1970, Nowak1997, Altenhoff2012, Pegueroles2013, Soria2014}.\footnote{For multiple sequences, orthology is usually defined relative to their most recent common ancestor.
-This technically includes sequences which split by duplication after this point (``in-paralogs''), but excludes sequences which split by duplication before (``out-paralogs'') \cite{Remm2001}.
-Many current orthology inference pipelines explicitly incorporate steps to detect in-paralogs.
-However, the resulting orthologs groups can easily be restricted to those without in-paralogs (``single copy orthologs'') for analyses where an assumption of conserved function is necessary.} This relationship between orthology and function is an essential component of modern biological research since it permits the transfer of annotations between biological systems using sequence similarity alone.
-
-Given this importance, methods for inferring orthologous groups of proteins were developed shortly after the first genomes were sequenced in the late 1990s~\cite{Fleischmann1995, Goffeau1996, CESC1998}.
-One early and influential approach was to cluster triangles of hits resulting from homology searches between pairs of genomes~\cite{Tatusov1997}.
-Graph-based approaches have remained popular, and in the intervening years many other researchers have refined this method by implementing various pre- and post-processing steps.
-Despite these improvements, many databases and pipelines use the same triangle clustering algorithm or other methods which require relatively few hits between sequences to infer an orthologous group, \textit{e.g.} connected components or Markov clustering~\cite{Remm2001, Enright2002, Li2003, Jensen2007, Linard2011, Emms2015, Train2017, Cosentino2018}.
-However, the scale of biological sequence data has changed dramatically.
-For example, in the last decade, the number of annotated genomes available from NCBI has increased nearly 20-fold and currently exceeds 900 (Fig.~\ref{sfig:ncbi}).
-Though this figure is only a rough proxy of the total number of assemblies available, it will likely continue to grow rapidly in the coming years as many large-scale genome assembly efforts such as i5K, the Bird 10,000 Genomes Project, and the Vertebrate Genomes Project have already yielded results~\cite{Thomas2020, Feng2020, Rhie2021}.
-Thus, the dense taxonomic sampling made possible by these projects poses new challenges and opportunities for the standard methods of orthology inference and alignment, which implicitly assume fewer and more distantly related genomes or fail to fully leverage the redundancy and organization of their annotations.
-
-In this work we therefore developed a computational pipeline that can robustly infer and align orthologous groups of proteins even when the genomes are highly redundant.
-Like many other orthology inference pipelines, our overall approach is based on clustering a graph of hits from homology searches.
-However, we modified many details to maximize the detection of highly diverged orthologs while also minimizing the impact of incomplete or incorrect annotations.
-Furthermore, since modern genome annotation pipelines frequently produce gene models and protein sequences in tandem, we implemented an additional clustering step to organize the resulting orthologous groups of proteins into gene-level units.
-However, most of our efforts were focused on the final step of aligning the orthologous sequences.
-Though genome annotation pipelines are often proficient at identifying the overall locus of genes, the accurate identification of exon boundaries and start codons when transcript evidence is limited remains an ongoing challenge~\cite{Frankish2015, Dunne2018}.
-Consequently, protein sequences derived from annotation pipelines can include non-homologous segments of significant length or exclude highly conserved segments.
-Such heterogeneity in the structure and length of the sequences in an orthologous group poses many challenges for their alignment and subsequent analysis.
-Thus, we implemented several novel quality control and data cleaning steps to correct mis-alignments and identify likely sequencing, assembly, or annotation errors.
-
-To develop these methods, we chose a set of 33 assembled and annotated \textit{Drosophila} genomes, which includes all 12 species from the original \textit{Drosophila} 12 Genomes Consortium~\cite{D12GC2007}.
-However, the genomes of these 12 species have been re-sequenced since their first release, which has resulted in substantial improvements in their assemblies and annotations.
-Despite these developments and other several other recent genome assembly projects of species in the \textit{Drosophila} genus, there is not yet a collection of high-quality alignments of orthologous proteins that reflects these improvements in genome assembly and diversity~\cite{Miller2018, Kim2021}.
-Given the \textit{Drosophila} genus spans diverse habitats and over 50 million years of evolution but maintains a conserved life cycle and body plan, such a resource would facilitate a new generation of studies that illuminate the forces that drive protein evolution in unprecedented detail~\cite{Wiegmann2011, Obbard2012}.
-
-\begin{figure}[h!]
-\includegraphics[width=\textwidth]{graphical_overview.png}
-\centering
-\caption{\textbf{Overview of homology inference pipeline.}
-\textbf{(A-C)} The large, open circles represent the annotated genomes, and the small, filled circles represent the protein sequences associated with each annotation.
-Sequences that share homology are colored with shades of the same hue.
-\textbf{(D-F)} The small, filled circles and lines represent the same sequences and best hits from the previous steps.}
-\label{fig:graphical_overview}
-\end{figure}
-
-\section*{Results}
-\subsection*{Pipeline overview}
-Our pipeline follows a similar overall approach to other graph-based methods of orthology inference.
-First, protein sequences from annotated genomes are collected (Fig.~\ref{fig:graphical_overview}A), and homology searches are then conducted between all query-target pairs of genomes (Fig.~\ref{fig:graphical_overview}B).
-The raw output from these homology searches is processed to yield best hits between pairs of sequences (Fig.~\ref{fig:graphical_overview}C).
-Next, the network of best hits is clustered into self-consistent orthologous groups (Fig.~\ref{fig:graphical_overview}D).
-Since genes can have multiplied associated isoforms, we then implemented a novel second clustering step where orthologous groups are grouped by their parent genes, which are represented by the two sets of clusters with warm and cool colors, respectively (Fig.~\ref{fig:graphical_overview}E).
-Finally, representative sequences in each orthologous group are aligned (Fig.~\ref{fig:graphical_overview}F).
-In the following sections, we discuss each of these and other steps which were omitted for clarity in greater detail.
-
-\subsection*{Input genomes and pre-processing}
-All annotated genomes in the genus \textit{Drosophila} available in April 2022 were downloaded from NCBI's RefSeq database.
-The assemblies annotated by the NCBI eukaryotic genome annotation pipeline have passed several quality checks and all have supporting transcript evidence, so the annotations are generally highly complete~\cite{Thibaud-Nissen2013}.
-The \textit{D.
-miranda} annotation was excluded due to its unusual karyotype~\cite{Dobzhansky1935}.
-Other annotations were excluded after preliminary clustering showed a deficiency in the number of orthologous groups containing those genomes, indicating their annotations were less complete (data not shown).
-The \textit{D.
-melanogaster} annotation was downloaded from FlyBase~\cite{Gramates2022}.
-In total, the input data consists of 33 genomes, which are listed in Table~\ref{stable:genomes}.
-Many genes have transcripts that differ only in their UTRs, and as a result there are many duplicate protein sequences in the annotations.
-Though not strictly necessary, we removed the duplicates in our pipeline, which greatly reduced the computational burden of later steps.
-
-\subsection*{Extraction of best hits from BLAST output}
-The protein sequences in each genome annotation were searched against each other in reciprocal pairs using BLAST, yielding a list of high-scoring segment pairs (HSPs) for each query-target pair~\cite{Camacho2009}.
-HSPs are local alignments, meaning they do not necessarily span the entire lengths of the query and target sequences.
-Consequently, the search algorithm may return multiple HSPs for each query-target pair if statistically significant regions of homology are separated by nonhomologous or poorly conserved regions.
-Though the most significant HSP is often used to represent all HSPs between a query-target pair, this approach can fail to rank the pairs by their overall significance if their alignments are broken into multiple HSPs.
-Furthermore, since query-target pairs were later filtered by the amount overlap between their sequences, it can also exclude pairs that pass the overlap threshold even if the most significant HSP alone does not.
-Thus, HSPs were merged into a single object called a hit.
-The best hits for each query were then taken as the highest-scoring hits that passed a minimum overlap criterion and were reciprocal between the query and target sequences.
-
-\subsection*{Clustering in orthologous groups}
-The best hits between sequences are naturally visualized as a graph where sequences are nodes and best hits are edges between nodes.
-Two connected components, sets of nodes joined by a sequence of edges, are shown (Fig.~\ref{fig:components}A,B).
-The sequences in the first (Fig.~\ref{fig:components}A) all contain C2H2 zinc fingers, whereas the sequences in the second (Fig.~\ref{fig:components}B) are members of the Par-1 family of serine/threonine protein kinases.
-In both components, some sets of nodes have a high density of edges, forming distinct clusters, whereas other nodes are only sparsely connected to their neighbors.
-To better understand the structure of these two components, we calculated the number of sequences, unique genes, and unique species in each.
-The first has 385, 346, and 33 sequences, genes, and species, respectively, and the second has 222, 33, and 33 sequences, genes, and species, respectively.
-We then plotted the relationship between the number sequences and unique genes across all components to see if this pattern holds true generally (Fig.~\ref{fig:components}C).
-Two distinct trendlines are apparent.
-The first increases linearly with the number of sequences with a slope of one, indicating each sequence is generally associated with a unique gene.
-The second is constant with an intercept of 33, indicating the number of unique genes quickly saturates at the total number of genomes.
-Thus, there are generally two classes of components.
-The first is composed of many distinct genes, whereas the second is composed of many different isoforms of a single group of genes.
-
-The diffuse networks observed in the first component class are likely the result of a combination of factors, including rapid evolution, gene duplication, and annotation errors.
-Regardless of their origin, these hits are not strong candidates for comparative analyses since an orthology relationship is supported by relatively few genome pairs.
-Instead, likely orthologs should consistently identify each other as best reciprocal hits across many genome pairs.
-The same is true of the hits in the second component class.
-Although the genes as a unit form a single orthologous group, sequences with few hits are likely non-conserved or tissue-specific isoforms.
-Thus, orthologous groups can be operationally defined as self-consistent clusters in the hit graph.
-However, sequence divergence or assembly and annotation errors may prevent a best reciprocal hit between orthologous sequences across all genome pairs.
-In fact, although the most common number of reciprocal hits is 32, one fewer than the total number of genomes, many sequences have fewer (Fig.~\ref{fig:components}D).
-Thus, the clustering method should require a high degree of self-consistency without demanding complete consensus.
-
-\begin{figure}[h!]
-\includegraphics[width=\textwidth]{components/out/0017-005F.png}
-\centering
-\caption{\textbf{Selected connected components of hit graph and summary statistics.}
-\textbf{(A-B)} Two distinct connected components of the hit graph.
-Edges are colored by the value of their bit score.
-\textbf{(C)} Hexbin plot of the number of sequences and the number of unique genes in each component.
-\textbf{(D)} Histogram of number edges associated with each sequence, \textit{i.e.} the degree of each node.}
-\label{fig:components}
-\end{figure}
-
-The identification of sets of densely connected nodes in graphs is known as community detection in network analysis.
-While many community detection algorithms are available, only some are commonly used in the context of orthology inference.
-One early method that remains popular is building clusters progressively by identifying nodes that form a triangle with at least two other nodes in the cluster~\cite{Tatusov1997, Jensen2007}.
-Other approaches include the MCL algorithm, which clusters graphs by similating stochastic flow, and connected components or other single-linkage criteria~\cite{Remm2001, Enright2002, Li2003, Emms2015, Train2017, Cosentino2018}.
-While these methods are robust when clustering hit graphs derived from smaller or more diverse sets of genomes, they are not suitable for the large number of closely related genomes in this work since they require relatively few edges to define a cluster.
-For example, the MCL algorithm and connected components method assign a node to a cluster as long as it has a single edge, and triangle clustering only requires two edges to two adjacent nodes.
-
-However, connected components and triangle clustering are special cases of the more general \textit{k}-clique percolation algorithm where \textit{k} equals two and three, respectively.
-The clique percolation algorithm detects clusters by first identifying cliques, sets of nodes which are fully connected, of a specified size \textit{k} in the graph (Fig.~\ref{fig:percolation}A).
-Clusters are then taken as the connected components of an overlap graph where an edge exists between two cliques if they share \textit{k}-1 nodes in common.
-An intuitive way to visualize this algorithm is by “rolling” a clique of some size \textit{k} over the graph (Fig.~\ref{fig:percolation}B).
-More specifically, a cluster is initiated when a set of nodes which form a \textit{k}-clique is identified.
-The cluster expands by shifting the \textit{k}-clique to an adjacent \textit{k}-clique that shares \textit{k}-1 nodes in common with the current \textit{k}-clique.
-A cluster stops expanding when there are no adjacent \textit{k}-cliques, and the algorithm terminates when there are no \textit{k}-cliques which are not part of a cluster.
-The strength of this algorithm is its ability to exclude sparsely connected nodes from clusters with an easily tunable parameter \textit{k}.
-Higher values of \textit{k} require greater overlap between a candidate node and those already in the cluster and therefore produce tighter clusters at the cost of excluding more speculative orthology relationships (Fig.~\ref{fig:percolation}C).
-We set \textit{k} to 4 as compromise between these concerns, yielding 22,813 orthologous groups, a plurality of which contained all 33 species (Fig.~\ref{sfig:clusters}).
-
-\begin{figure}[h!]
-\includegraphics[width=\textwidth]{percolation/percolation.png}
-\includegraphics[width=\textwidth]{percolation/out/005F.png}
-\centering
-\caption{\textbf{Clique percolation algorithm.}
-\textbf{(A)} Cliques for \textit{k} equal to three, four, five, and six.
-\textit{k} equal to one and two correspond to a single node and two nodes joined by an edge, respectively.
-\textbf{(B)} Illustration of clique percolation algorithm where \textit{k = 4}.
-\textbf{(C)} A single component clustered by clique percolation with varying values of \textit{k}.
-Nodes are colored according to their cluster.
-If a node belongs to multiple clusters, it uses a blend of those colors.}
-\label{fig:percolation}
-\end{figure}
-
-\subsection*{Addition of paralogs to orthologous groups}
-A weakness of the best reciprocal hits criterion is its exclusion of recently diverged paralogs.
-Since only the highest scoring hits for each query are included in the graph, a paralog without a corresponding duplicate in the target genome is ignored if it is marginally more diverged than the other copy.
-This is corrected by adding likely paralogs to the orthologous groups.
-Briefly, the protein sequences in each genome were searched against themselves.
-If the bit score for an intra-genome hit exceeded the bit score of any inter-genome hits for the same query, the two sequences were identified as a paralogous pair.
-Orthologous groups were then supplemented with paralogs by adding the paired sequences for each of the original members of the orthologous group.
-Most orthologous groups contain no paralogs, and those that do generally have few relative to the original number of sequences in the group (Fig.~\ref{sfig:paralogs}).
-
-\subsection*{Grouping orthologous groups by gene}
-As genes can have several annotated isoforms, each gene can be associated with several orthologous groups.
-However, the orthologous groups are not organized into gene-level units since they were clustered using sequence similarity only.
-A graph-based approach was therefore used to group orthologous groups with similar sets of parent genes.
-First, a gene overlap graph was constructed by defining an edge between orthologous groups if the intersection of their associated sets of parent genes is at least 50\% of the smaller of the two.
-Gene groups were then taken as the connected components of the resulting graph, yielding 14,909 groups.
-This is commensurate with the roughly 15,000 genes in each genome, which suggests this approach has successfully clustered orthologous groups derived from a common set of parent genes.
-
-\subsection*{Initial alignment and selection of representative sequences}
-Since the NCBI annotation pipeline incorporates transcriptome data from a variety of sources, its inputs are heterogeneous in sequencing depth, developmental stage, and tissue of origin across different genomes.
-As a result, some genomes are annotated with different or multiple splice isoforms of a given orthologous gene, which can create complex networks in the resulting hit graph.
-For example, if the genomes are variably annotated with one or both of two distinct isoforms, the resulting graph may contain two clusters connected by a ``bridge'' formed by the genomes which contain only one of the isoforms.
-If the nodes bridging the two clusters form cliques with themselves and the clusters, the clique percolation algorithm will merge all the nodes into a single orthologous group where some genes have multiple associated sequences.
-However, these additional sequences can complicate downstream comparative analyses that may not easily generalize to genes with multiple associated sequences.
-Thus, in our pipeline a single representative was chosen for each gene using an alignment-based strategy detailed in the methods section.
-
-\subsection*{Selection of single copy orthologous groups}
-The criteria for selecting orthologous groups for further analyses depends on the biological question under investigation.
-For example, studies of gene duplication will focus on orthologous groups with paralogs in some lineages but not in others.
-In contrast, analyses which assume functional conservation should restrict the orthologous groups to single copy orthologs since paralogs more frequently undergo functional divergence~\cite{Altenhoff2012, Pegueroles2013, Soria2014}.
-A simple method for identifying such groups is requiring each species to have exactly one associated gene.
-However, since the probability of at least one missing gene annotation approaches one as the total number of genomes increases, this is too restrictive and fails to leverage the redundancy of closely related genomes.
-Instead, a set of phylogenetic diversity criteria detailed in Table~\ref{stable:diversity_criteria} were applied to ensure the major lineages were represented in downstream analyses.
-Furthermore, genome-wide analyses should select one orthologous group per each of the previously identified gene groups as to not bias the results towards genes with many distinct groups of isoforms.
-In summary, orthologous groups failing the phylogenetic diversity criteria were first removed, and the representative for each gene group was chosen as the highest scoring orthologous group when ranked by the number species and the sum of the bit scores associated with each edge.
-This significantly reduced the number of orthologous groups from 22,813 to 8,566.
-
-\subsection*{Alignment refinement}
-An initial alignment was created for each orthologous group as part of the selection of representative sequences.
-Though the pipeline’s quality control measures ensure a high degree of overall sequence identity between members of an orthologous group, some sequences contain long “poorly supported” segments which are not aligned to most or any other sequences in the alignment.
-Since most common multiple sequence alignment algorithms assume the sequences are largely homologous, these segments are sometimes “over-aligned” by forcing them into alignment where chance sequence similarities occur.
-Typically, these segments remain contiguous, so the alignments alternate between short runs of columns with few or no gaps and large gap-rich regions.
-More rarely, when long poorly supported segments are adjacent to a long gap in the same sequence, the two are interlaced, yielding long gaps interrupted by short segments of spurious alignment.
-
-The aligner MAFFT has a mode for addressing over-alignment with a parameter, \textit{a\textsubscript{max}}, that adjusts the strength of the correction~\cite{Katoh2013, Katoh2016}.
-\textit{a\textsubscript{max}} varies between 0 and 1, but at values above 0.8, the correction severely degrades the alignment of conserved regions.
-Thus, in our pipeline the sequences were aligned with a conservative value of 0.4.
-Though some cases of over-alignment remained, most conserved regions were reliably aligned with few to no errors.
-To correct the remaining errors in alignments of the single copy orthologs, a statistical model of each alignment called a profile HMM was constructed from the well-aligned columns~\cite{Eddy2009}.
-The original sequences were then aligned to these profiles.
-By defining highly conserved “anchor” regions, this approach largely prevents the alignment of chance sequence similarities in long poorly supported segments.
-Furthermore, since profile HMMs explicitly model the beginnings and ends of alignments, they can fully exclude these segments from the termini.
-This effect is visible in the illustrated examples, where each has long poorly supported segments that are completely removed (Fig.~\ref{fig:realignment}).
-
-\begin{figure}[h!]
-\includegraphics[width=\textwidth]{realignment/out/merged.png}
-\centering
-\caption{\textbf{Alignment with long poorly supported segments.}
-The alignments of representative sequences in orthologous groups 23D9 \textbf{(A)}, 01E0 \textbf{(B)}, and 07C3 \textbf{(C)} before and after refinement.}
-\label{fig:realignment}
-\end{figure}
-
-\subsection*{Alignment curation}
-Although the refinement process corrects most cases of over-alignment, the alignment may still contain regions whose aligned segments have poor or inconsistent support.
-For example, long poorly supported segments in internal regions were not removed from the alignment since they are bounded by at least one consensus column to the left and right.
-Additionally, some regions have a significant fraction of sequences with strongly supported segments, but the observed gap pattern is discordant with the expected phylogenetic relationships.
-Since they are present in so few sequences, the former segments are likely artifactual, resulting from errors during assembly or annotation.
-(Biological explanations such as alternative splice sites, frameshift mutations, or transposition events are also possible, however.) In contrast, the high sequence identity and clear boundaries of the segments in the latter regions suggest they are conserved but skipped exons.
-Given the heterogeneous sourcing of the transcript evidence, these sequences containing these segments are likely splice isoforms specific to certain tissues or developmental stage.
-
-% To investigate these hypotheses, the genome sequences and supporting transcript evidence for several segments and regions were examined in detail: 0254, 26CF, 202C, 188E, 2E9F, 34B7, 0C31.
-
-Since the segments in these regions are likely the result of incorrect or incomplete annotations rather than meaningful biological variation, maintaining them in the alignments would propagate spurious homologies to subsequent analyses.
-This is a common issue in alignments generated by automated pipelines, so downstream analyses often focus on the strongly supported regions by removing or “trimming” columns below some threshold number of gaps or sequence identity~\cite{Castresana2000, CapellaGutierrez2009}.
-This approach, however, is inadequate if the taxonomic sampling is dense, as a single indel event along a lineage containing many species can increase the number of gaps above the threshold.
-Moreover, as this method does not incorporate any spatial information, it can rapidly alternate between trimming and preserving columns.
-Thus, it can severely disrupt any analyses which are sensitive to the spatial organization of an alignment.
-
-Phylogenetic HMMs (phylo-HMMs) are statistical models that incorporate phylogenetic and spatial information to calculate the probability that each observation in a sequence was generated by one of several “hidden states”~\cite{Felsenstein1996}.
-Since they can evaluate both the probability of a gap pattern in a column given the known phylogenetic relationships and the local context, a phylo-HMM was used to segment the alignment into contiguous regions with different patterns of gaps.
-A fully specified phylo-HMM requires a fixed number of hidden states and a probability distribution for each.
-Thus, we identified four distinct types of regions in the alignments, roughly corresponding to highly conserved regions with few to no gaps, divergent regions, regions with a stable gap pattern discordant with the expected phylogenetic relationships, and regions with poorly supported segments.
-For simplicity, however, we refer to the states that generate each type of region as 1A, 1B, 2, and 3, respectively.
-To model probability distributions for each state, we first conceptualized the observed alignments as the superposition of two distinct processes (Fig.~\ref{fig:hmm}A, left).
-The first is a phylogenetic process which evolves and splits a single ancestral sequence over time according to a tree.
-The second is the annotation process which can erroneously exclude or include segments from a sequence.
-The result is an alignment of annotated sequences which contains evolutionary information obscured by ``noise'' from the annotation process, shown here by the exclusion of three N-terminal residues in the fourth annotated sequence.
-To simplify modeling this behavior with an HMM, we coded the sequences into binary symbols.
-The distributions for each state then consisted of two components derived from the encoded sequences (Fig.~\ref{fig:hmm}A, right).
-The first component models the gap pattern with a Markov process.
-This Markov process is in turn composed of two subprocesses where the first is a phylogenetic process, and the second is a jump process.
-These subprocess roughly correspond to changes caused by evolution and annotation, respectively.
-Because this first component did not fully capture the propensity for the gap patterns to remain constant, we included a second component that models the ``gap stickiness'' as a beta-binomial random variable by counting the number of symbols that remain constant between columns.
-Each component is associated with a set of parameters, and the unique parameters for each state yield its characteristic gap pattern and gap stickiness.
-
-After the model was trained on manually labeled examples, it was used to assign a label to the columns in each alignment.
-Columns assigned to states 1A and 1B are the regions of interest for downstream analyses since the gaps generally follow the expected pattern given the phylogenetic tree.
-In contrast, columns assigned to states 2 and 3 largely corresponded to the long poorly supported segments and phylogenetically discordant regions discussed previously and were therefore removed from the alignments.
-In the example decoded alignment shown, the decoded states closely follow the expected patterns (Fig.~ref{fig:hmm})B).
-Overall, fewer than 10\% of alignments were trimmed of any regions.
-However, when regions were trimmed, they were largely inferred as state 3, meaning the removed regions are generally long poorly supported segments that are aligned to few if any other sequences (Fig.~\ref{sfig:hmm_trim}).
-
-\begin{figure}[h!]
-\includegraphics[width=\textwidth]{hmm/hmm_architecture.png}
-\includegraphics[width=\textwidth]{hmm/out/2252.png}
-\centering
-\caption{\textbf{HMM emission architecture and a decoded alignment.}
-\textbf{(A)} Schematic of theoretical alignment generating process and corresponding probabilistic components in HMM.
-White and colored boxes indicate gap and non-gap symbols in the biological sequences, respectively.
-White and grey boxes indicate gap and non-gap symbols in the encoded sequences, respectively.
-c\textsubscript{0} and c\textsubscript{1} indicate the first and second columns in the alignment, respectively.
-The parameters associated with each component are shown to the right.
-\textbf{(B)} The alignment of the representative sequences in orthologous group 2252 decoded using the trained HMM.}
-\label{fig:hmm}
-\end{figure}
-
-Though the phylo-HMM removed phylogenetically incongruent ``insertions,'' some sequences still contained extensive segments of uninterrupted gaps.
-These segments are easily identified in regions which are otherwise highly conserved, so they are also likely the result of incorrect or incomplete annotations.
-Unfortunately, they can also span more divergent regions, which complicates a simple rule- or threshold-based approach for identifying them.
-Thus, another phylo-HMM was trained to label each position in a sequence as generated by either a ``missing'' or ``not missing'' state.
-In this case, however, the aligned sequences were processed individually and not as aligned columns.
-As the previous phylo-HMM already ensured each column has sufficient support, these labels can instead be used to exclude sequences from downstream analyses depending on the amount of tolerated overlap with the regions of interest.
-
-\subsection*{Inference of species trees}
-Many phylogenetic methods require a species tree to inform the evolutionary relationships between sequences.
-In fact, the phylo-HMMs discussed previously used a species tree as an input, though we omitted this detail for clarity of exposition.
-Therefore, to support the curation step and other downstream analyses, we sought to infer phylogenetic trees from the aligned sequences.
-However, since the roots of phylogenetic trees are not identifiable with commonly used time-reversible substitution models, we repeated the orthology inference pipeline with the outgroup species \textit{Scaptodrosophila lebanonensis}.
-Afterwards, we inferred phylogenetic trees using the LG model of amino acid substitution from 100 meta-alignments sampled from alignments of single copy orthologous groups.
-We then combined them into a single consensus tree (Fig.~\ref{fig:trees}A).
-To provide a similar tree for the analysis of non-coding regions, we inferred phylogenetic trees using the GTR model of nucleotide substitution from 100 meta-alignments sampled from nucleotide alignments which were ``reverse translated'' from the protein alignments and their corresponding coding sequences (\ref{fig:trees}B).
-Both trees have an identical topology, which is consistent with other published phylogenies~\cite{D12GC2007, DaLage2007}.
-
-\begin{figure}[h!]
-\includegraphics[width=\textwidth]{trees/out/trees1.png}
-\centering
-\caption{\textbf{Phylogenetic tree of species.}
-\textbf{(A)} Consensus tree from LG model fit to meta-alignments directly sampled from the original protein alignments.
-\textbf{(B)} Consensus tree from GTR model fit to meta-alignments sampled from ``reverse translated'' nucleotide alignments.
-Values at nodes are bootstrap percentages.}
-\label{fig:trees}
-\end{figure}
-
-\section*{Discussion}
-The orthologous groups and alignments yielded by this pipeline are a valuable resource for comparative studies of gene birth/death processes and protein evolution at the level of both entire proteomes and specific gene families in the \textit{Drosophila} genus.
-To assist these efforts, the final alignments after curation and the labels from the ``missing'' phylo-HMM are provided as supplemental data.
-Although this work focused on single copy orthologs, other studies may require different subsets of orthologous groups that demand other pre-processing and alignment strategies.
-Therefore, we have included the orthologous groups and the initial alignments with and without non-representative sequences in the supplemental data as well.
-While we anticipate these resources will remain relevant in the near term, the trends that permitted this work to substantially improve on previous efforts will render them obsolete in the coming years as more \textit{Drosophila} genomes are assembled and annotated.
-However, an authoritative and lasting set of orthologous groups and alignments is not the primary goal of this work.
-Instead, it serves as a case study in how dense taxonomic sampling and modern genome assembly and annotation pipelines present new opportunities and challenges to the traditional techniques for identifying and aligning orthologous groups.
-
-For example, despite many additional pre-processing steps and other tweaks introduced by later authors, the basic framework of orthology inference by clustering the hit graph has remained largely unchanged in the past twenty years~\cite{Tatusov1997, Remm2001, Li2003, Jensen2007, Linard2011, Emms2015, Train2017, Cosentino2018}.
-This longevity is a testament to the robustness of the underlying idea that orthologous proteins should consistently identify each other as the most similar pairs between their genomes.
-Even as the number of genomes and their taxonomic density has increased dramatically, many orthology inference pipelines continue to use algorithms which were originally applied to sets of far fewer and more distantly related genomes.
-This mismatch in scale increases the chance of propagating annotation errors since only a small number of edges are needed to create or merge clusters.
-Thus, we instead applied a generalization of the triangle and connected components clustering methods called \textit{k}-clique percolation where \textit{k} is a tunable parameter that influences the tightness of a cluster.
-The optimal value of \textit{k} for a given set of genomes is unclear and likely depends on the desired trade-off between sensitivity and specificity.
-However, percolation theory or simulations may yield insights.
-
-Another challenge is the annotation of multiple isoforms for a single gene.
-Though prior pipelines have generally selected the longest isoform as the representative before conducting the orthology searches, if the sequences do not share a common intron-exon structure this approach can introduce artifacts or other issues during alignment.
-Instead, as protein sequences are increasingly derived from or linked to genomic sequences, we sought to incorporate the full annotations into the orthology inference pipeline.
-This, however, created two additional complications.
-First, a single gene could have several associated orthologous groups if its isoforms belonged to different clusters.
-Second, a single orthologous group could have several isoforms of a single gene if its isoforms were clustered together.
-In both cases, the presence of multiple sequences for a single gene creates ambiguities over which is the “primary” isoform.
-Since the first occurs at the level of orthologous groups, the orthologous groups were first grouped by the similarity of their parent genes using a graph-based strategy.
-Afterwards, a single representative is easily chosen as the group with the largest number of distinct species, though other criteria are possible.
-Since the second occurs within an orthologous group, the sequences were first aligned, and a representative for each gene was chosen as the sequence which was most concordant with this initial alignment.
-
-The second major innovation in this work is its method for the refinement and trimming of alignments.
-In refinement, a profile HMM is constructed from the well-aligned columns of an initial alignment, and the original sequences are aligned to this profile.
-This process usually corrects errors caused by the presence of long poorly supported segments.
-In trimming, a phylo-HMM is used to remove regions and segments which are poorly supported by the phylogenetic consensus.
-While the combination of these two steps yielded high-quality alignments that are suitable for further analyses, they are an \textit{ad hoc} fix for underlying issues with the gene models and alignment algorithms.
-The most principled solution is to optimize or supplement the gene models using the initial alignments generated by the orthology inference pipeline, which is possible with tools such as OMGene or OrthoFiller~\cite{Dunne2018, Dunne2017}.
-However, if preserving the original annotations is desired or necessary, the remaining possibility is to correct the alignments as we have done here.
-In fact, the errors we sought to address broadly stem from shortcomings of current alignment algorithms rather than errors in the sequences themselves.
-Though the scoring functions of modern multiple sequence alignment algorithms are complex, they are generally derived from models that penalize gaps with a linear or affine cost.
-As a result, they often interlace gaps with short, aligned segments rather than a single long gap.
-However, when the sequences are different isoforms of a single gene, their alignment will necessarily contain contiguous exon sized gaps.
-The same is true when aligning isoforms of diverged orthologs, though the relationship between their exons may be complex.
-
-The most popular aligners for protein sequences (Clustal Omega, MAFFT, MUSCLE, T-Coffee) do not include splice sites in their alignments, which makes them prone to aligning non-homologous exons~\cite{Sievers2017, Katoh2013, Edgar2004, Notredame2000}.
-Current algorithms can easily be extended to incorporate splice sites by coding them as a new symbol and preventing alignment between splice sites and amino acids, which was recently implemented in the aligner, \textit{Mirage}~\cite{Nord2018}.
-The biggest challenge in practice, however, is mapping a protein sequence to its genomic sequence to identify splice sites in the protein sequence.
-Although this information can in principle be derived from the GTF annotation files produced by the NCBI pipeline, annotating the splice junctions in the protein sequences themselves would facilitate splice-sensitive alignment.
-
-These improvements would enhance rather than replace the phylo-HMM trimming method developed in this work.
-The model could easily be extended to include a state that outputs a splice symbol before transitioning to one of the states in the current architecture.
-This intermediate state would increase the accuracy of state inference since a splice symbol followed by a phylogenetically discordant gap pattern would strongly signal a state 2 region.
-The association between state 2 and skipped exons can be made explicit by requiring that transitions to and from state 2 first proceed through the splice state.
-This of course depends on proper labeling of the training data, which would be trivial since the boundaries between exons would be marked by splice site symbols rather than inferred from gap patterns.
-Unfortunately, this would not allow the phylo-HMM to label extended exon boundaries as state 2 since they would not be bounded by splice symbols to the left and right.
-Accordingly, the phylo-HMM would need to permit transitions between state 3 and any other state to accommodate more complex splice variants and other annotation errors.
-Thus, this extended phylo-HMM would combine the strengths of splice-sensitive alignment with the more heuristic approach used here.
-Since state 2 inferences would necessarily correspond to skipped exons, they would be suitable for analyses of this form of alternative splicing.
-Though state 3 inferences would not directly correspond to specific biological process, they still have value as spatially and phylogenetically aware labels for trimming poorly supported segments from alignments.
-
-The phylo-HMM could be further enhanced by expanding its emission distribution to include more symbols in the amino acid alphabet.
-This would allow it to better model observed substitution patterns between specific symbols, for example the high rate of exchange between gaps and glutamine residues caused by polyglutamine tracts.
-The transitions between amino acids could be parametrized with a published matrix such as LG~\cite{Le2008}, but the transitions between amino acids and gaps would be inferred from labeled data.
-It is unclear if the resulting gain in accuracy would justify the increased computational burden, however.
-
-Though there are benchmarks available for optimizing and comparing methods of orthology inference, the metrics are calculated over a set of reference genomes which are sparsely sampled over a broad taxonomic range, so it is unclear if they are informative for method designed to yield robust inferences when the genomes are highly related~\cite{Nevers2022}.
-Furthermore, the heterogeneity of genome architectures and annotations may require quality assurance methods tailored to each set of genomes.
-Thus, there is likely no one-size-fits-all approach to orthology inference, and with many other standalone programs available for more standard use cases (Hieranoid, OMA standalone, OrthoFinder, OrthoInspector, Orthologer), we have chosen not to package the code into an end-to-end pipeline~\cite{Kaduk2017, Altenhoff2019, Emms2019, Linard2014, Zdobnov2020}.
-Instead, we have devoted considerable attention to organizing and documenting the code to make it accessible to a newcomer and thereby facilitate the adaptation of specific steps to similar projects as needed.
-
-In contrast, though many HMM packages are available for the Python programming language, we found none were satisfactorily documented or contained tutorials to introduce HMMs and their APIs to a wide audience.
-We therefore refactored this code into a package available on PyPI and GitHub called Homomorph.
-The package itself only implements standard HMM algorithms, but the GitHub repository includes tutorials that introduce the API and implement training routines.
-Similar tutorials for machine learning libraries such as TensorFlow have undoubtedly fueled the application of neural networks across diverse fields, but HMMs are also powerful models that can be more appropriate when the data obey certain statistical or structural constraints.
-Thus, we hope this package and its accompanying tutorials will serve as an on-ramp to HMMs and spur their greater adoption by non-specialists.
-
-Though databases of orthologous groups such as COGs, Ensembl Compara, EggNOG, OMA, OrthoDB, OrthoInspector, and OrthoMCL will continue to be useful for comparative studies across broad taxonomic ranges, the increasing speed at which high-quality genome assemblies and annotations are produced means no single database can encompass the most complete data~\cite{Galperin2020, Herrero2016, HuertaCepas2018, Altenhoff2020, Zdobnov2020, Nevers2018, Chen2006}.
-Furthermore, since many early comparative genomics studies spanned diverse branches of the tree of life, future research will likely prioritize taxonomic depth over breadth.
-Thus, custom sets of orthologous groups will grow more and not less common.
-Despite the challenges these developments pose, they also present new opportunities to bridge the gap between mutational and macroevolutionary processes.
-
-\section*{Materials and methods}
-\subsection*{Sequence de-duplication and BLAST search parameters}
-The protein sequences for each annotation were de-duplicated by removing any sequences which had already appeared in association with the same gene.
-Thus, the first accession associated with a sequence and gene pair is the sequence’s representative accession for the gene.
-BLAST+ 2.13.0 was used for the sequence similarity searches~\cite{Camacho2009}.
-An E-value cutoff of 1 was used for the initial searches.
-However, this cutoff was lowered to 1E-10 during processing of the BLAST output.
-
-\subsection*{Extraction of HSPs from BLAST output}
-To reduce the computational burden of merging HSPs into hits, the BLAST output was filtered to extract HSPs associated with the highest-scoring gene.
-The HSPs were first grouped by target protein, and the resulting groups were sorted in descending order by the bit score of their highest-scoring HSP.
-Iterating over the groups, all the HSPs in a group were passed to the next step until the parent gene of the group was not the parent gene of the highest ranked group.
-This method collects all candidate HSPs for a target gene if the highest-scoring HSP within a group exceeds the highest-scoring HSP of the next best gene.
-This is in some senses an extension of the best hit criterion where hits are considered at the level of genes rather than proteins.
-If multiple genes tied for the highest-scoring HSP, the iteration stopped when the parent gene of the current group matched none of these highest-scoring genes.
-
-\subsection*{Merging of HSPs into hits}
-HSPs were merged in two stages where the first combined non-overlapping HSPs, and the second combined the remaining HSPs.
-In the first stage, proceeding from highest to lowest bit score, HSPs were marked as “disjoint” if they do not overlap with any other HSP previously marked as disjoint.
-Although this greedy strategy does not necessarily yield the highest-scoring set of disjoint HSPs, it prioritized higher scoring HSPs.
-In the second stage, all disjoint HSPs were marked as “compatible,” and proceeding from highest to lowest bit score the remaining HSPs were marked as compatible if the overlap with any other compatible HSP was no more than 50\% of the length of either.
-The best hit for each query was chosen as the hit with the highest sum of bit scores from disjoint HSPs.
-The best hits were then filtered by overlap and reciprocity criteria.
-The overlap criterion was applied first and requires that 50\% of residues in the query are aligned in compatible HSPs.
-This excludes false positives from conserved domains embedded in larger non-homologous proteins by ensuring the hits span a sufficient fraction of the query and target sequences.
-The reciprocity criterion requires each query-target pair has a corresponding hit where the roles are reversed, which ensures there is no ambiguity in which target is the best match for the query.
-
-\subsection*{Clustering by \textit{k}-clique percolation}
-\textit{k}-clique percolation was implemented in two steps.
-In the first, maximal cliques were identified.
-In the second, a percolation graph is constructed by defining edges between cliques if they have \textit{k}-1 nodes in common.
-Clusters are the connected components of this second graph.
-The first step uses the NetworkX implementation of a maximal clique algorithm.
-The second step uses a modification of the NetworkX implementation of the \textit{k}-clique community algorithm.
-The NetworkX implementation exhaustively finds all edges in the percolation graph.
-Since joining a \textit{k}-clique community only requires that a clique has a single edge connecting it to that a community, this approach is needlessly expensive for large graphs.
-The custom implementation instead uses a progressive approach where each clique is checked against a list of known communities, merging communities as necessary in each step.
-
-The hit graph is sparse, so these algorithms are efficient when applied to its individual connected components.
-However, some components have a structure with many maximal cliques, which dramatically slows the first or second step of the clique percolation algorithm.
-Thus, if either step exceeded 90 s, the process timed out, and the simpler \textit{k}-core algorithm was used instead.
-Out of over 10,000 connected components, only seven timed out, and many of those contained highly dense clusters of histone sequences.
-
-\subsection*{Addition of paralogs to orthologous groups}
-The protein sequences for each annotation were searched against themselves with the same settings as for the inter-genome searches.
-The resulting output was processed identically except the HSPs were not filtered using the best gene criterion.
-Thus, all HSPs for each query were merged into hits.
-The best hit for each query and target gene was then chosen as the hit with the highest sum of bit scores from disjoint HSPs.
-(Grouping by target gene ensured only the highest-scoring isoform was selected.) Query-target pairs whose hits which exceeded the maximum bit score for all inter-genome hits associated with that query and passed the overlap and reciprocity filters were designated as paralogous pairs.
-The orthologous groups were supplemented with paralogs by adding the paired sequences for each of the original members of the orthologous group.
-
-\subsection*{Initial alignment and selection of representative sequences}
-The sequences in each orthologous group were aligned using MAFFT 7.490 with the following settings: --globalpair --maxiterate 1000 --thread 1 --anysymbol --allowshift --unalignlevel 0.4 --leavegappyregion~\cite{Katoh2013}.
-Representative sequences for each gene were then selected by maximum likelihood according to binary profiles constructed from these alignments.
-First each sequence was coded into gap and non-gap symbols.
-The sequences were then grouped by gene, and for each group and position if at least one sequence was aligned in the group, the group contributed one count for the non-gap symbol to the profile at that position.
-Otherwise, the group contributed a count for the gap symbol at that position.
-To account for the phylogenetic dependencies between sequences, the counts were weighted according to a Gaussian process over the GTR2 consensus tree described in the section on inferring species trees~\cite{Altschul1989}.
-If a species had multiple genes in the orthologous group, the species weight was divided evenly among them.
-Each coded sequence was then scored according to this profile, and the maximum likelihood sequence for each gene was selected as its representative.
-By assigning a non-gap count to groups and positions where at least one sequence is aligned, the profile incentivizes the selection of sequences contain the fewest gaps and match the consensus alignment.
-Unfortunately, this scheme can cause a sequence to score negative infinity if it has a gap at a position where every group has at least one aligned sequence, so the profile was initialized with a pseudocount of 0.005 for the gap and non-gap symbols at each position.
-
-\subsection*{Alignment refinement}
-Consensus columns were manually defined as columns in the initial alignments with at least five non-gap symbols using the Stockholm file format.
-These labeled alignments were used to build a profile HMM using HMMer 3.3.2~\cite{Eddy2009}.
-No sequence relative weighting was used, and the effective number of sequences was set to three times the number of sequences in the alignment.
-HMMer uses strong priors to allow the construction of profiles with relatively few aligned sequences.
-However, in rare cases a sequence with a long poorly supported segment was misaligned, effectively putting it out of register with the profile and creating a poor-quality alignment.
-Increasing the effective sequence number corrected this error.
-The original sequences were then aligned to the profile.
-Unaligned segments at the N- or C- terminus were removed from the alignment.
-Unaligned segments in internal regions were extracted and aligned separately for each region using MAFFT with the same settings.
-The resulting alignments were then “stitched” into the profile alignment.
-This process was unsuccessful for one alignment due to cryptic error in MAFFT, which reduced the number of refined alignments to 8,565.
-
-This method corrects most cases of misalignment since poorly supported segments are typically aligned to fewer than four other sequences.
-However, it does not remove poorly supported segments from columns with more than five non-gap symbols, and their contribution is sometimes sufficient to reproduce the misalignment when the sequences are aligned to the profile HMM.
-A lightweight convolutional neural network was therefore trained to recognize poorly supported segments to remove them more completely from the profile HMM.
-The network consisted of a two-dimensional embedding from a 22-dimensional one-hot-encoded vector (20 amino acids, an ambiguous amino acid, and a gap), two convolutional layers, and a one-dimensional output.
-The first layer was composed of four filters.
-Two of these filters had a kernel size of 10 and a dilation of 1.
-The other two filters had a kernel size of 60 with a dilation of 3.
-The second layer was composed of two filters with a kernel size of 10 and a dilation of 1 (Fig.~\ref{sfig:cnn_architecture}).
-The network was trained on 88,513 hand-labeled positions across 122 sequences in 46 unique alignments.
-These examples were largely long segments aligned to few, if any, other sequences (Fig.~\ref{sfig:cnn_training}).
-
-The trained network outputs a probability that each position in an aligned sequence is part of a long poorly supported segment.
-A simple algorithm was used to detect positions associated with high confidence long poorly supported segments even if the output at that position was relatively low.
-First, “seeds” were identified at positions where the output exceeded 0.75.
-These seeds were expanded outwards until the output was below 0.05.
-A seed and its expansion composed a long poorly supported segment.
-Consensus columns were defined as columns with at least five non-gap symbols, excluding positions labeled as long poorly supported segments.
-To maintain the contribution of long poorly supported segments to the profile’s models of insertions, long poorly supported segments were replaced with gaps only at positions aligned to consensus columns.
-
-\subsection*{Alignment curation}
-The alignments were coded into gap and non-gap symbols to simplify the emission distributions.
-The ``insertion'' phylo-HMM was composed of the four hidden states described in the main text.
-The emission distributions for each consisted of two components which modeled the gap pattern and the propensity for those patterns to remain constant (``gap stickiness''), respectively.
-The first component was a two-state Markov process which was in turn composed of two subprocess.
-The first was a phylogenetic process on the on the GTR2 consensus tree described in the section on inferring species trees, and the second was jump process at the tips.
-The second component was a beta-Bernoulli distribution on the number of symbols which were constant between subsequent columns.
-The ``missing data'' phylo-HMM was composed of two hidden states which were both parameterized with the same two-state, two component Markov process as the insertion phylo-HMM.
-However, the emission probabilities were calculated as the posterior probability of the observed symbol given the data rather than the probability of the data.
-Only the alignments of the single copy orthologous groups were curated, so each tip in the species tree would correspond to a single sequence in the alignment.
-
-All HMM algorithms were implemented with custom code which is available as the package Homomorph on PyPI.
-The insertion phylo-HMM was trained on 39,298 manually labeled columns in 11 alignments, and the deletion phylo-HMM was trained on 58,655 manually labeled positions in 20 sequences in 9 unique alignments (Fig.~\ref{sfig:hmm_training}).
-Because maximum-likelihood estimation of the model parameters yielded posterior decoding curves which toggled between hidden states too rapidly, the models were instead trained discriminatively~\cite{Krogh1999}.
-The difference, briefly, is maximum-likelihood estimation finds the parameters that best reproduce the observed distributions whereas discriminative training finds the parameters that minimize prediction error.
-Discriminatively trained models typically perform better in practice since real-world data are rarely fully described by the distribution specified by the model.
-
-The posterior distributions over states were computed for both the insertion and missing data phylo-HMMs.
-To trim the alignments, the posterior distributions from the insertion phylo-HMM were converted to state assignments with the following method.
-First the probabilities for states 2 and 3 were combined.
-Since this probability can change rapidly or gradually depending on the local context, a simple cutoff would not necessarily define the boundaries as the columns where the gap pattern changes most abruptly.
-Instead, a high cutoff of 0.9 was used to define a “seed” region.
-The boundaries of the seed were then expanded both inwards and outwards to define intervals from which to select boundaries.
-The outward expansion halted when the probability or its derivative was below 0.05 and 0.001, respectively.
-The inward expansion halted when the derivative was below 0.02.
-The left and right boundaries were chosen as the columns in each expansion with the maximum product between the derivative and the number of sequences which changed to or from a gap between columns.
-By combining where the model’s confidence changed rapidly with the observed change in the gap pattern, this method generally selected reasonable boundaries.
-In contrast, the posterior probabilities from the missing data phylo-HMM changed sharply, so positions were labeled as the ``missing data'' state if the posterior probability exceeded 0.95.
-These assignments can be used to filter segments or entire sequences from downstream analyses and are available in the supplementary data.
-
-\subsection*{Inference of species trees}
-The orthology inference pipeline was first repeated with the outgroup species \textit{Scaptodrosophila lebanonensis}.
-Then orthologous groups with one sequence for each species were aligned, and 100 meta-alignments were constructed by randomly sampling 10,000 columns from these 9,435 alignments.
-(The alignments were not refined with HMMer before sampling.) To determine the effect of invariant columns and gaps, two sampling strategies were used where invariant columns were allowed or disallowed and the maximum fraction of gaps was set at 0, 50, and 100\%.
-Their combination yielded six different sets of meta-alignments.
-A tree was fit to each meta-alignment with the LG substitution model and four discrete gamma rate categories using IQ-TREE 1.6.12~\cite{Nguyen2014, Le2008, Yang1994}.
-If the sampling strategy allowed invariant columns, an invariant rate category was included.
-The resulting trees from each set were then merged into a majority consensus tree (Fig.~\ref{sfig:trees_LG}).
-
-To fit trees using the GTR model of nucleotide substitution, the protein alignments were converted to nucleotide alignments using the corresponding coding sequences in the genome annotations.
-Some protein sequences are “low quality,” meaning their coding sequences contain frameshifts, premature stop codons, or other errors even though they are strong hits to known protein-coding genes.
-The NCBI annotation pipeline corrects some of these defects in the protein sequences, which can complicate a simple “reverse translation” of the alignment.
-After rejecting alignments where the expected translation from a coding sequence differed from its corresponding protein sequence, 3,425 alignments remained.
-Consensus tree were derived from meta-alignments sampled from these alignments using the approach described previously except the GTR model was used in place of the LG model.
-
-To fit trees using the two-state GTR model of substitution, the protein alignments were first coded into gap or non-gap symbols.
-As before, 100 meta-alignments were constructed from these coded alignments for each sampling strategy.
-In this case, only the presence of invariant columns was varied, yielding two sets of meta-alignments.
-Trees were then fit using the GTR2 model with no rate categories.
-An invariant category, however, was included if invariant columns were allowed.
-Since the bootstrap confidences were sometimes lower than 50\%, the resulting trees from each set were then merged into a loose consensus tree to prevent multifurcations (Fig.~\ref{sfig:trees_GTR}).
-
-\subsection*{Code and data availability}
-The code used to produce the results and analyses is available at \url{https://github.com/marcsingleton/orthology_inference2023}.
-HMM algorithms were implemented in the standalone package Homomorph which is available at \url{https://github.com/marcsingleton/homomorph} and on the Python Package Index (PyPI).
-The following Python libraries were used: matplotlib, NumPy, pandas, SciPy, and TensorFlow~\cite{Hunter2007, Harris2020, McKinney2010, Virtanen2020, Abadi2015}.
-Relevant output files are available in the supporting information.
-There is no primary data associated with this manuscript.
-All primary data are available from publicly accessible sources described in their corresponding sections.
-
-\section*{Acknowledgments}
-We thank the past and present members of the Eisen lab for helpful discussions and their assistance in revising the manuscript.
-
-\printbibliography
-
-\clearpage
-
 \section*{Supporting information}

### 2022-11-17 chapter1/biorxiv.tex
Added biorxiv.tex

--- 2022-11-17 chapter1.tex
@@
-\documentclass[10pt,letterpaper]{article}
+\documentclass[11pt,letterpaper]{article}
 
@@
 \usepackage{graphicx}
-\graphicspath{{./figures/}}
 
@@
 %\doublespacing
-\usepackage[skip=10pt]{parskip}
+
+\usepackage[skip=10pt]{parskip}  % Remove indents
 
@@
 \usepackage[doi=false,url=false]{biblatex}
-\addbibresource{refs.bib}
+\addbibresource{../refs.bib}
 
@@
 
+% DOCUMENT
 \begin{document}
+\graphicspath{{./figures/}}
 
@@
 
-% Abstract must be 300 words or less
-\section*{Abstract}
-abstract text goes here
-
-\section*{Introduction}
-Comparative genomics is a powerful tool for yielding insights into evolutionary relationships, molecular function, and the forces that drive gene, genome, and population evolution.
-These methods often rely on the identification of homologous sequences or homologs, that is sequences with common ancestry, since this ensures that differences between sequences reflect variations in evolution from a common point of divergence.
-However, many analyses impose the additional condition that the sequences have diverged through speciation events (orthology) rather than duplications (paralogy) or other mechanisms such as horizontal gene transfer~\cite{Fitch1970}.
-The underlying assumption is ``orthologs'' have conserved equivalent functions whereas ``paralogs'', by virtue of their redundancy, are more likely to diverge~\cite{Ohno1970, Nowak1997, Altenhoff2012, Pegueroles2013, Soria2014}.\footnote{For multiple sequences, orthology is usually defined relative to their most recent common ancestor.
-This technically includes sequences which split by duplication after this point (``in-paralogs''), but excludes sequences which split by duplication before (``out-paralogs'') \cite{Remm2001}.
-Many current orthology inference pipelines explicitly incorporate steps to detect in-paralogs.
-However, the resulting orthologs groups can easily be restricted to those without in-paralogs (``single copy orthologs'') for analyses where an assumption of conserved function is necessary.} This relationship between orthology and function is an essential component of modern biological research since it permits the transfer of annotations between biological systems using sequence similarity alone.
-
-Given this importance, methods for inferring orthologous groups of proteins were developed shortly after the first genomes were sequenced in the late 1990s~\cite{Fleischmann1995, Goffeau1996, CESC1998}.
-One early and influential approach was to cluster triangles of hits resulting from homology searches between pairs of genomes~\cite{Tatusov1997}.
-Graph-based approaches have remained popular, and in the intervening years many other researchers have refined this method by implementing various pre- and post-processing steps.
-Despite these improvements, many databases and pipelines use the same triangle clustering algorithm or other methods which require relatively few hits between sequences to infer an orthologous group, \textit{e.g.} connected components or Markov clustering~\cite{Remm2001, Enright2002, Li2003, Jensen2007, Linard2011, Emms2015, Train2017, Cosentino2018}.
-However, the scale of biological sequence data has changed dramatically.
-For example, in the last decade, the number of annotated genomes available from NCBI has increased nearly 20-fold and currently exceeds 900 (Fig.~\ref{sfig:ncbi}).
-Though this figure is only a rough proxy of the total number of assemblies available, it will likely continue to grow rapidly in the coming years as many large-scale genome assembly efforts such as i5K, the Bird 10,000 Genomes Project, and the Vertebrate Genomes Project have already yielded results~\cite{Thomas2020, Feng2020, Rhie2021}.
-Thus, the dense taxonomic sampling made possible by these projects poses new challenges and opportunities for the standard methods of orthology inference and alignment, which implicitly assume fewer and more distantly related genomes or fail to fully leverage the redundancy and organization of their annotations.
-
-In this work we therefore developed a computational pipeline that can robustly infer and align orthologous groups of proteins even when the genomes are highly redundant.
-Like many other orthology inference pipelines, our overall approach is based on clustering a graph of hits from homology searches.
-However, we modified many details to maximize the detection of highly diverged orthologs while also minimizing the impact of incomplete or incorrect annotations.
-Furthermore, since modern genome annotation pipelines frequently produce gene models and protein sequences in tandem, we implemented an additional clustering step to organize the resulting orthologous groups of proteins into gene-level units.
-However, most of our efforts were focused on the final step of aligning the orthologous sequences.
-Though genome annotation pipelines are often proficient at identifying the overall locus of genes, the accurate identification of exon boundaries and start codons when transcript evidence is limited remains an ongoing challenge~\cite{Frankish2015, Dunne2018}.
-Consequently, protein sequences derived from annotation pipelines can include non-homologous segments of significant length or exclude highly conserved segments.
-Such heterogeneity in the structure and length of the sequences in an orthologous group poses many challenges for their alignment and subsequent analysis.
-Thus, we implemented several novel quality control and data cleaning steps to correct mis-alignments and identify likely sequencing, assembly, or annotation errors.
-
-To develop these methods, we chose a set of 33 assembled and annotated \textit{Drosophila} genomes, which includes all 12 species from the original \textit{Drosophila} 12 Genomes Consortium~\cite{D12GC2007}.
-However, the genomes of these 12 species have been re-sequenced since their first release, which has resulted in substantial improvements in their assemblies and annotations.
-Despite these developments and other several other recent genome assembly projects of species in the \textit{Drosophila} genus, there is not yet a collection of high-quality alignments of orthologous proteins that reflects these improvements in genome assembly and diversity~\cite{Miller2018, Kim2021}.
-Given the \textit{Drosophila} genus spans diverse habitats and over 50 million years of evolution but maintains a conserved life cycle and body plan, such a resource would facilitate a new generation of studies that illuminate the forces that drive protein evolution in unprecedented detail~\cite{Wiegmann2011, Obbard2012}.
-
-\begin{figure}[h!]
-\includegraphics[width=\textwidth]{graphical_overview.png}
-\centering
-\caption{\textbf{Overview of homology inference pipeline.}
-\textbf{(A-C)} The large, open circles represent the annotated genomes, and the small, filled circles represent the protein sequences associated with each annotation.
-Sequences that share homology are colored with shades of the same hue.
-\textbf{(D-F)} The small, filled circles and lines represent the same sequences and best hits from the previous steps.}
-\label{fig:graphical_overview}
-\end{figure}
-
-\section*{Results}
-\subsection*{Pipeline overview}
-Our pipeline follows a similar overall approach to other graph-based methods of orthology inference.
-First, protein sequences from annotated genomes are collected (Fig.~\ref{fig:graphical_overview}A), and homology searches are then conducted between all query-target pairs of genomes (Fig.~\ref{fig:graphical_overview}B).
-The raw output from these homology searches is processed to yield best hits between pairs of sequences (Fig.~\ref{fig:graphical_overview}C).
-Next, the network of best hits is clustered into self-consistent orthologous groups (Fig.~\ref{fig:graphical_overview}D).
-Since genes can have multiplied associated isoforms, we then implemented a novel second clustering step where orthologous groups are grouped by their parent genes, which are represented by the two sets of clusters with warm and cool colors, respectively (Fig.~\ref{fig:graphical_overview}E).
-Finally, representative sequences in each orthologous group are aligned (Fig.~\ref{fig:graphical_overview}F).
-In the following sections, we discuss each of these and other steps which were omitted for clarity in greater detail.
-
-\subsection*{Input genomes and pre-processing}
-All annotated genomes in the genus \textit{Drosophila} available in April 2022 were downloaded from NCBI's RefSeq database.
-The assemblies annotated by the NCBI eukaryotic genome annotation pipeline have passed several quality checks and all have supporting transcript evidence, so the annotations are generally highly complete~\cite{Thibaud-Nissen2013}.
-The \textit{D.
-miranda} annotation was excluded due to its unusual karyotype~\cite{Dobzhansky1935}.
-Other annotations were excluded after preliminary clustering showed a deficiency in the number of orthologous groups containing those genomes, indicating their annotations were less complete (data not shown).
-The \textit{D.
-melanogaster} annotation was downloaded from FlyBase~\cite{Gramates2022}.
-In total, the input data consists of 33 genomes, which are listed in Table~\ref{stable:genomes}.
-Many genes have transcripts that differ only in their UTRs, and as a result there are many duplicate protein sequences in the annotations.
-Though not strictly necessary, we removed the duplicates in our pipeline, which greatly reduced the computational burden of later steps.
-
-\subsection*{Extraction of best hits from BLAST output}
-The protein sequences in each genome annotation were searched against each other in reciprocal pairs using BLAST, yielding a list of high-scoring segment pairs (HSPs) for each query-target pair~\cite{Camacho2009}.
-HSPs are local alignments, meaning they do not necessarily span the entire lengths of the query and target sequences.
-Consequently, the search algorithm may return multiple HSPs for each query-target pair if statistically significant regions of homology are separated by nonhomologous or poorly conserved regions.
-Though the most significant HSP is often used to represent all HSPs between a query-target pair, this approach can fail to rank the pairs by their overall significance if their alignments are broken into multiple HSPs.
-Furthermore, since query-target pairs were later filtered by the amount overlap between their sequences, it can also exclude pairs that pass the overlap threshold even if the most significant HSP alone does not.
-Thus, HSPs were merged into a single object called a hit.
-The best hits for each query were then taken as the highest-scoring hits that passed a minimum overlap criterion and were reciprocal between the query and target sequences.
-
-\subsection*{Clustering in orthologous groups}
-The best hits between sequences are naturally visualized as a graph where sequences are nodes and best hits are edges between nodes.
-Two connected components, sets of nodes joined by a sequence of edges, are shown (Fig.~\ref{fig:components}A,B).
-The sequences in the first (Fig.~\ref{fig:components}A) all contain C2H2 zinc fingers, whereas the sequences in the second (Fig.~\ref{fig:components}B) are members of the Par-1 family of serine/threonine protein kinases.
-In both components, some sets of nodes have a high density of edges, forming distinct clusters, whereas other nodes are only sparsely connected to their neighbors.
-To better understand the structure of these two components, we calculated the number of sequences, unique genes, and unique species in each.
-The first has 385, 346, and 33 sequences, genes, and species, respectively, and the second has 222, 33, and 33 sequences, genes, and species, respectively.
-We then plotted the relationship between the number sequences and unique genes across all components to see if this pattern holds true generally (Fig.~\ref{fig:components}C).
-Two distinct trendlines are apparent.
-The first increases linearly with the number of sequences with a slope of one, indicating each sequence is generally associated with a unique gene.
-The second is constant with an intercept of 33, indicating the number of unique genes quickly saturates at the total number of genomes.
-Thus, there are generally two classes of components.
-The first is composed of many distinct genes, whereas the second is composed of many different isoforms of a single group of genes.
-
-The diffuse networks observed in the first component class are likely the result of a combination of factors, including rapid evolution, gene duplication, and annotation errors.
-Regardless of their origin, these hits are not strong candidates for comparative analyses since an orthology relationship is supported by relatively few genome pairs.
-Instead, likely orthologs should consistently identify each other as best reciprocal hits across many genome pairs.
-The same is true of the hits in the second component class.
-Although the genes as a unit form a single orthologous group, sequences with few hits are likely non-conserved or tissue-specific isoforms.
-Thus, orthologous groups can be operationally defined as self-consistent clusters in the hit graph.
-However, sequence divergence or assembly and annotation errors may prevent a best reciprocal hit between orthologous sequences across all genome pairs.
-In fact, although the most common number of reciprocal hits is 32, one fewer than the total number of genomes, many sequences have fewer (Fig.~\ref{fig:components}D).
-Thus, the clustering method should require a high degree of self-consistency without demanding complete consensus.
-
-\begin{figure}[h!]
-\includegraphics[width=\textwidth]{components/out/0017-005F.png}
-\centering
-\caption{\textbf{Selected connected components of hit graph and summary statistics.}
-\textbf{(A-B)} Two distinct connected components of the hit graph.
-Edges are colored by the value of their bit score.
-\textbf{(C)} Hexbin plot of the number of sequences and the number of unique genes in each component.
-\textbf{(D)} Histogram of number edges associated with each sequence, \textit{i.e.} the degree of each node.}
-\label{fig:components}
-\end{figure}
-
-The identification of sets of densely connected nodes in graphs is known as community detection in network analysis.
-While many community detection algorithms are available, only some are commonly used in the context of orthology inference.
-One early method that remains popular is building clusters progressively by identifying nodes that form a triangle with at least two other nodes in the cluster~\cite{Tatusov1997, Jensen2007}.
-Other approaches include the MCL algorithm, which clusters graphs by similating stochastic flow, and connected components or other single-linkage criteria~\cite{Remm2001, Enright2002, Li2003, Emms2015, Train2017, Cosentino2018}.
-While these methods are robust when clustering hit graphs derived from smaller or more diverse sets of genomes, they are not suitable for the large number of closely related genomes in this work since they require relatively few edges to define a cluster.
-For example, the MCL algorithm and connected components method assign a node to a cluster as long as it has a single edge, and triangle clustering only requires two edges to two adjacent nodes.
-
-However, connected components and triangle clustering are special cases of the more general \textit{k}-clique percolation algorithm where \textit{k} equals two and three, respectively.
-The clique percolation algorithm detects clusters by first identifying cliques, sets of nodes which are fully connected, of a specified size \textit{k} in the graph (Fig.~\ref{fig:percolation}A).
-Clusters are then taken as the connected components of an overlap graph where an edge exists between two cliques if they share \textit{k}-1 nodes in common.
-An intuitive way to visualize this algorithm is by “rolling” a clique of some size \textit{k} over the graph (Fig.~\ref{fig:percolation}B).
-More specifically, a cluster is initiated when a set of nodes which form a \textit{k}-clique is identified.
-The cluster expands by shifting the \textit{k}-clique to an adjacent \textit{k}-clique that shares \textit{k}-1 nodes in common with the current \textit{k}-clique.
-A cluster stops expanding when there are no adjacent \textit{k}-cliques, and the algorithm terminates when there are no \textit{k}-cliques which are not part of a cluster.
-The strength of this algorithm is its ability to exclude sparsely connected nodes from clusters with an easily tunable parameter \textit{k}.
-Higher values of \textit{k} require greater overlap between a candidate node and those already in the cluster and therefore produce tighter clusters at the cost of excluding more speculative orthology relationships (Fig.~\ref{fig:percolation}C).
-We set \textit{k} to 4 as compromise between these concerns, yielding 22,813 orthologous groups, a plurality of which contained all 33 species (Fig.~\ref{sfig:clusters}).
-
-\begin{figure}[h!]
-\includegraphics[width=\textwidth]{percolation/percolation.png}
-\includegraphics[width=\textwidth]{percolation/out/005F.png}
-\centering
-\caption{\textbf{Clique percolation algorithm.}
-\textbf{(A)} Cliques for \textit{k} equal to three, four, five, and six.
-\textit{k} equal to one and two correspond to a single node and two nodes joined by an edge, respectively.
-\textbf{(B)} Illustration of clique percolation algorithm where \textit{k = 4}.
-\textbf{(C)} A single component clustered by clique percolation with varying values of \textit{k}.
-Nodes are colored according to their cluster.
-If a node belongs to multiple clusters, it uses a blend of those colors.}
-\label{fig:percolation}
-\end{figure}
-
-\subsection*{Addition of paralogs to orthologous groups}
-A weakness of the best reciprocal hits criterion is its exclusion of recently diverged paralogs.
-Since only the highest scoring hits for each query are included in the graph, a paralog without a corresponding duplicate in the target genome is ignored if it is marginally more diverged than the other copy.
-This is corrected by adding likely paralogs to the orthologous groups.
-Briefly, the protein sequences in each genome were searched against themselves.
-If the bit score for an intra-genome hit exceeded the bit score of any inter-genome hits for the same query, the two sequences were identified as a paralogous pair.
-Orthologous groups were then supplemented with paralogs by adding the paired sequences for each of the original members of the orthologous group.
-Most orthologous groups contain no paralogs, and those that do generally have few relative to the original number of sequences in the group (Fig.~\ref{sfig:paralogs}).
-
-\subsection*{Grouping orthologous groups by gene}
-As genes can have several annotated isoforms, each gene can be associated with several orthologous groups.
-However, the orthologous groups are not organized into gene-level units since they were clustered using sequence similarity only.
-A graph-based approach was therefore used to group orthologous groups with similar sets of parent genes.
-First, a gene overlap graph was constructed by defining an edge between orthologous groups if the intersection of their associated sets of parent genes is at least 50\% of the smaller of the two.
-Gene groups were then taken as the connected components of the resulting graph, yielding 14,909 groups.
-This is commensurate with the roughly 15,000 genes in each genome, which suggests this approach has successfully clustered orthologous groups derived from a common set of parent genes.
-
-\subsection*{Initial alignment and selection of representative sequences}
-Since the NCBI annotation pipeline incorporates transcriptome data from a variety of sources, its inputs are heterogeneous in sequencing depth, developmental stage, and tissue of origin across different genomes.
-As a result, some genomes are annotated with different or multiple splice isoforms of a given orthologous gene, which can create complex networks in the resulting hit graph.
-For example, if the genomes are variably annotated with one or both of two distinct isoforms, the resulting graph may contain two clusters connected by a ``bridge'' formed by the genomes which contain only one of the isoforms.
-If the nodes bridging the two clusters form cliques with themselves and the clusters, the clique percolation algorithm will merge all the nodes into a single orthologous group where some genes have multiple associated sequences.
-However, these additional sequences can complicate downstream comparative analyses that may not easily generalize to genes with multiple associated sequences.
-Thus, in our pipeline a single representative was chosen for each gene using an alignment-based strategy detailed in the methods section.
-
-\subsection*{Selection of single copy orthologous groups}
-The criteria for selecting orthologous groups for further analyses depends on the biological question under investigation.
-For example, studies of gene duplication will focus on orthologous groups with paralogs in some lineages but not in others.
-In contrast, analyses which assume functional conservation should restrict the orthologous groups to single copy orthologs since paralogs more frequently undergo functional divergence~\cite{Altenhoff2012, Pegueroles2013, Soria2014}.
-A simple method for identifying such groups is requiring each species to have exactly one associated gene.
-However, since the probability of at least one missing gene annotation approaches one as the total number of genomes increases, this is too restrictive and fails to leverage the redundancy of closely related genomes.
-Instead, a set of phylogenetic diversity criteria detailed in Table~\ref{stable:diversity_criteria} were applied to ensure the major lineages were represented in downstream analyses.
-Furthermore, genome-wide analyses should select one orthologous group per each of the previously identified gene groups as to not bias the results towards genes with many distinct groups of isoforms.
-In summary, orthologous groups failing the phylogenetic diversity criteria were first removed, and the representative for each gene group was chosen as the highest scoring orthologous group when ranked by the number species and the sum of the bit scores associated with each edge.
-This significantly reduced the number of orthologous groups from 22,813 to 8,566.
-
-\subsection*{Alignment refinement}
-An initial alignment was created for each orthologous group as part of the selection of representative sequences.
-Though the pipeline’s quality control measures ensure a high degree of overall sequence identity between members of an orthologous group, some sequences contain long “poorly supported” segments which are not aligned to most or any other sequences in the alignment.
-Since most common multiple sequence alignment algorithms assume the sequences are largely homologous, these segments are sometimes “over-aligned” by forcing them into alignment where chance sequence similarities occur.
-Typically, these segments remain contiguous, so the alignments alternate between short runs of columns with few or no gaps and large gap-rich regions.
-More rarely, when long poorly supported segments are adjacent to a long gap in the same sequence, the two are interlaced, yielding long gaps interrupted by short segments of spurious alignment.
-
-The aligner MAFFT has a mode for addressing over-alignment with a parameter, \textit{a\textsubscript{max}}, that adjusts the strength of the correction~\cite{Katoh2013, Katoh2016}.
-\textit{a\textsubscript{max}} varies between 0 and 1, but at values above 0.8, the correction severely degrades the alignment of conserved regions.
-Thus, in our pipeline the sequences were aligned with a conservative value of 0.4.
-Though some cases of over-alignment remained, most conserved regions were reliably aligned with few to no errors.
-To correct the remaining errors in alignments of the single copy orthologs, a statistical model of each alignment called a profile HMM was constructed from the well-aligned columns~\cite{Eddy2009}.
-The original sequences were then aligned to these profiles.
-By defining highly conserved “anchor” regions, this approach largely prevents the alignment of chance sequence similarities in long poorly supported segments.
-Furthermore, since profile HMMs explicitly model the beginnings and ends of alignments, they can fully exclude these segments from the termini.
-This effect is visible in the illustrated examples, where each has long poorly supported segments that are completely removed (Fig.~\ref{fig:realignment}).
-
-\begin{figure}[h!]
-\includegraphics[width=\textwidth]{realignment/out/merged.png}
-\centering
-\caption{\textbf{Alignment with long poorly supported segments.}
-The alignments of representative sequences in orthologous groups 23D9 \textbf{(A)}, 01E0 \textbf{(B)}, and 07C3 \textbf{(C)} before and after refinement.}
-\label{fig:realignment}
-\end{figure}
-
-\subsection*{Alignment curation}
-Although the refinement process corrects most cases of over-alignment, the alignment may still contain regions whose aligned segments have poor or inconsistent support.
-For example, long poorly supported segments in internal regions were not removed from the alignment since they are bounded by at least one consensus column to the left and right.
-Additionally, some regions have a significant fraction of sequences with strongly supported segments, but the observed gap pattern is discordant with the expected phylogenetic relationships.
-Since they are present in so few sequences, the former segments are likely artifactual, resulting from errors during assembly or annotation.
-(Biological explanations such as alternative splice sites, frameshift mutations, or transposition events are also possible, however.) In contrast, the high sequence identity and clear boundaries of the segments in the latter regions suggest they are conserved but skipped exons.
-Given the heterogeneous sourcing of the transcript evidence, these sequences containing these segments are likely splice isoforms specific to certain tissues or developmental stage.
-
-% To investigate these hypotheses, the genome sequences and supporting transcript evidence for several segments and regions were examined in detail: 0254, 26CF, 202C, 188E, 2E9F, 34B7, 0C31.
-
-Since the segments in these regions are likely the result of incorrect or incomplete annotations rather than meaningful biological variation, maintaining them in the alignments would propagate spurious homologies to subsequent analyses.
-This is a common issue in alignments generated by automated pipelines, so downstream analyses often focus on the strongly supported regions by removing or “trimming” columns below some threshold number of gaps or sequence identity~\cite{Castresana2000, CapellaGutierrez2009}.
-This approach, however, is inadequate if the taxonomic sampling is dense, as a single indel event along a lineage containing many species can increase the number of gaps above the threshold.
-Moreover, as this method does not incorporate any spatial information, it can rapidly alternate between trimming and preserving columns.
-Thus, it can severely disrupt any analyses which are sensitive to the spatial organization of an alignment.
-
-Phylogenetic HMMs (phylo-HMMs) are statistical models that incorporate phylogenetic and spatial information to calculate the probability that each observation in a sequence was generated by one of several “hidden states”~\cite{Felsenstein1996}.
-Since they can evaluate both the probability of a gap pattern in a column given the known phylogenetic relationships and the local context, a phylo-HMM was used to segment the alignment into contiguous regions with different patterns of gaps.
-A fully specified phylo-HMM requires a fixed number of hidden states and a probability distribution for each.
-Thus, we identified four distinct types of regions in the alignments, roughly corresponding to highly conserved regions with few to no gaps, divergent regions, regions with a stable gap pattern discordant with the expected phylogenetic relationships, and regions with poorly supported segments.
-For simplicity, however, we refer to the states that generate each type of region as 1A, 1B, 2, and 3, respectively.
-To model probability distributions for each state, we first conceptualized the observed alignments as the superposition of two distinct processes (Fig.~\ref{fig:hmm}A, left).
-The first is a phylogenetic process which evolves and splits a single ancestral sequence over time according to a tree.
-The second is the annotation process which can erroneously exclude or include segments from a sequence.
-The result is an alignment of annotated sequences which contains evolutionary information obscured by ``noise'' from the annotation process, shown here by the exclusion of three N-terminal residues in the fourth annotated sequence.
-To simplify modeling this behavior with an HMM, we coded the sequences into binary symbols.
-The distributions for each state then consisted of two components derived from the encoded sequences (Fig.~\ref{fig:hmm}A, right).
-The first component models the gap pattern with a Markov process.
-This Markov process is in turn composed of two subprocesses where the first is a phylogenetic process, and the second is a jump process.
-These subprocess roughly correspond to changes caused by evolution and annotation, respectively.
-Because this first component did not fully capture the propensity for the gap patterns to remain constant, we included a second component that models the ``gap stickiness'' as a beta-binomial random variable by counting the number of symbols that remain constant between columns.
-Each component is associated with a set of parameters, and the unique parameters for each state yield its characteristic gap pattern and gap stickiness.
-
-After the model was trained on manually labeled examples, it was used to assign a label to the columns in each alignment.
-Columns assigned to states 1A and 1B are the regions of interest for downstream analyses since the gaps generally follow the expected pattern given the phylogenetic tree.
-In contrast, columns assigned to states 2 and 3 largely corresponded to the long poorly supported segments and phylogenetically discordant regions discussed previously and were therefore removed from the alignments.
-In the example decoded alignment shown, the decoded states closely follow the expected patterns (Fig.~ref{fig:hmm})B).
-Overall, fewer than 10\% of alignments were trimmed of any regions.
-However, when regions were trimmed, they were largely inferred as state 3, meaning the removed regions are generally long poorly supported segments that are aligned to few if any other sequences (Fig.~\ref{sfig:hmm_trim}).
-
-\begin{figure}[h!]
-\includegraphics[width=\textwidth]{hmm/hmm_architecture.png}
-\includegraphics[width=\textwidth]{hmm/out/2252.png}
-\centering
-\caption{\textbf{HMM emission architecture and a decoded alignment.}
-\textbf{(A)} Schematic of theoretical alignment generating process and corresponding probabilistic components in HMM.
-White and colored boxes indicate gap and non-gap symbols in the biological sequences, respectively.
-White and grey boxes indicate gap and non-gap symbols in the encoded sequences, respectively.
-c\textsubscript{0} and c\textsubscript{1} indicate the first and second columns in the alignment, respectively.
-The parameters associated with each component are shown to the right.
-\textbf{(B)} The alignment of the representative sequences in orthologous group 2252 decoded using the trained HMM.}
-\label{fig:hmm}
-\end{figure}
-
-Though the phylo-HMM removed phylogenetically incongruent ``insertions,'' some sequences still contained extensive segments of uninterrupted gaps.
-These segments are easily identified in regions which are otherwise highly conserved, so they are also likely the result of incorrect or incomplete annotations.
-Unfortunately, they can also span more divergent regions, which complicates a simple rule- or threshold-based approach for identifying them.
-Thus, another phylo-HMM was trained to label each position in a sequence as generated by either a ``missing'' or ``not missing'' state.
-In this case, however, the aligned sequences were processed individually and not as aligned columns.
-As the previous phylo-HMM already ensured each column has sufficient support, these labels can instead be used to exclude sequences from downstream analyses depending on the amount of tolerated overlap with the regions of interest.
-
-\subsection*{Inference of species trees}
-Many phylogenetic methods require a species tree to inform the evolutionary relationships between sequences.
-In fact, the phylo-HMMs discussed previously used a species tree as an input, though we omitted this detail for clarity of exposition.
-Therefore, to support the curation step and other downstream analyses, we sought to infer phylogenetic trees from the aligned sequences.
-However, since the roots of phylogenetic trees are not identifiable with commonly used time-reversible substitution models, we repeated the orthology inference pipeline with the outgroup species \textit{Scaptodrosophila lebanonensis}.
-Afterwards, we inferred phylogenetic trees using the LG model of amino acid substitution from 100 meta-alignments sampled from alignments of single copy orthologous groups.
-We then combined them into a single consensus tree (Fig.~\ref{fig:trees}A).
-To provide a similar tree for the analysis of non-coding regions, we inferred phylogenetic trees using the GTR model of nucleotide substitution from 100 meta-alignments sampled from nucleotide alignments which were ``reverse translated'' from the protein alignments and their corresponding coding sequences (\ref{fig:trees}B).
-Both trees have an identical topology, which is consistent with other published phylogenies~\cite{D12GC2007, DaLage2007}.
-
-\begin{figure}[h!]
-\includegraphics[width=\textwidth]{trees/out/trees1.png}
-\centering
-\caption{\textbf{Phylogenetic tree of species.}
-\textbf{(A)} Consensus tree from LG model fit to meta-alignments directly sampled from the original protein alignments.
-\textbf{(B)} Consensus tree from GTR model fit to meta-alignments sampled from ``reverse translated'' nucleotide alignments.
-Values at nodes are bootstrap percentages.}
-\label{fig:trees}
-\end{figure}
-
-\section*{Discussion}
-The orthologous groups and alignments yielded by this pipeline are a valuable resource for comparative studies of gene birth/death processes and protein evolution at the level of both entire proteomes and specific gene families in the \textit{Drosophila} genus.
-To assist these efforts, the final alignments after curation and the labels from the ``missing'' phylo-HMM are provided as supplemental data.
-Although this work focused on single copy orthologs, other studies may require different subsets of orthologous groups that demand other pre-processing and alignment strategies.
-Therefore, we have included the orthologous groups and the initial alignments with and without non-representative sequences in the supplemental data as well.
-While we anticipate these resources will remain relevant in the near term, the trends that permitted this work to substantially improve on previous efforts will render them obsolete in the coming years as more \textit{Drosophila} genomes are assembled and annotated.
-However, an authoritative and lasting set of orthologous groups and alignments is not the primary goal of this work.
-Instead, it serves as a case study in how dense taxonomic sampling and modern genome assembly and annotation pipelines present new opportunities and challenges to the traditional techniques for identifying and aligning orthologous groups.
-
-For example, despite many additional pre-processing steps and other tweaks introduced by later authors, the basic framework of orthology inference by clustering the hit graph has remained largely unchanged in the past twenty years~\cite{Tatusov1997, Remm2001, Li2003, Jensen2007, Linard2011, Emms2015, Train2017, Cosentino2018}.
-This longevity is a testament to the robustness of the underlying idea that orthologous proteins should consistently identify each other as the most similar pairs between their genomes.
-Even as the number of genomes and their taxonomic density has increased dramatically, many orthology inference pipelines continue to use algorithms which were originally applied to sets of far fewer and more distantly related genomes.
-This mismatch in scale increases the chance of propagating annotation errors since only a small number of edges are needed to create or merge clusters.
-Thus, we instead applied a generalization of the triangle and connected components clustering methods called \textit{k}-clique percolation where \textit{k} is a tunable parameter that influences the tightness of a cluster.
-The optimal value of \textit{k} for a given set of genomes is unclear and likely depends on the desired trade-off between sensitivity and specificity.
-However, percolation theory or simulations may yield insights.
-
-Another challenge is the annotation of multiple isoforms for a single gene.
-Though prior pipelines have generally selected the longest isoform as the representative before conducting the orthology searches, if the sequences do not share a common intron-exon structure this approach can introduce artifacts or other issues during alignment.
-Instead, as protein sequences are increasingly derived from or linked to genomic sequences, we sought to incorporate the full annotations into the orthology inference pipeline.
-This, however, created two additional complications.
-First, a single gene could have several associated orthologous groups if its isoforms belonged to different clusters.
-Second, a single orthologous group could have several isoforms of a single gene if its isoforms were clustered together.
-In both cases, the presence of multiple sequences for a single gene creates ambiguities over which is the “primary” isoform.
-Since the first occurs at the level of orthologous groups, the orthologous groups were first grouped by the similarity of their parent genes using a graph-based strategy.
-Afterwards, a single representative is easily chosen as the group with the largest number of distinct species, though other criteria are possible.
-Since the second occurs within an orthologous group, the sequences were first aligned, and a representative for each gene was chosen as the sequence which was most concordant with this initial alignment.
-
-The second major innovation in this work is its method for the refinement and trimming of alignments.
-In refinement, a profile HMM is constructed from the well-aligned columns of an initial alignment, and the original sequences are aligned to this profile.
-This process usually corrects errors caused by the presence of long poorly supported segments.
-In trimming, a phylo-HMM is used to remove regions and segments which are poorly supported by the phylogenetic consensus.
-While the combination of these two steps yielded high-quality alignments that are suitable for further analyses, they are an \textit{ad hoc} fix for underlying issues with the gene models and alignment algorithms.
-The most principled solution is to optimize or supplement the gene models using the initial alignments generated by the orthology inference pipeline, which is possible with tools such as OMGene or OrthoFiller~\cite{Dunne2018, Dunne2017}.
-However, if preserving the original annotations is desired or necessary, the remaining possibility is to correct the alignments as we have done here.
-In fact, the errors we sought to address broadly stem from shortcomings of current alignment algorithms rather than errors in the sequences themselves.
-Though the scoring functions of modern multiple sequence alignment algorithms are complex, they are generally derived from models that penalize gaps with a linear or affine cost.
-As a result, they often interlace gaps with short, aligned segments rather than a single long gap.
-However, when the sequences are different isoforms of a single gene, their alignment will necessarily contain contiguous exon sized gaps.
-The same is true when aligning isoforms of diverged orthologs, though the relationship between their exons may be complex.
-
-The most popular aligners for protein sequences (Clustal Omega, MAFFT, MUSCLE, T-Coffee) do not include splice sites in their alignments, which makes them prone to aligning non-homologous exons~\cite{Sievers2017, Katoh2013, Edgar2004, Notredame2000}.
-Current algorithms can easily be extended to incorporate splice sites by coding them as a new symbol and preventing alignment between splice sites and amino acids, which was recently implemented in the aligner, \textit{Mirage}~\cite{Nord2018}.
-The biggest challenge in practice, however, is mapping a protein sequence to its genomic sequence to identify splice sites in the protein sequence.
-Although this information can in principle be derived from the GTF annotation files produced by the NCBI pipeline, annotating the splice junctions in the protein sequences themselves would facilitate splice-sensitive alignment.
-
-These improvements would enhance rather than replace the phylo-HMM trimming method developed in this work.
-The model could easily be extended to include a state that outputs a splice symbol before transitioning to one of the states in the current architecture.
-This intermediate state would increase the accuracy of state inference since a splice symbol followed by a phylogenetically discordant gap pattern would strongly signal a state 2 region.
-The association between state 2 and skipped exons can be made explicit by requiring that transitions to and from state 2 first proceed through the splice state.
-This of course depends on proper labeling of the training data, which would be trivial since the boundaries between exons would be marked by splice site symbols rather than inferred from gap patterns.
-Unfortunately, this would not allow the phylo-HMM to label extended exon boundaries as state 2 since they would not be bounded by splice symbols to the left and right.
-Accordingly, the phylo-HMM would need to permit transitions between state 3 and any other state to accommodate more complex splice variants and other annotation errors.
-Thus, this extended phylo-HMM would combine the strengths of splice-sensitive alignment with the more heuristic approach used here.
-Since state 2 inferences would necessarily correspond to skipped exons, they would be suitable for analyses of this form of alternative splicing.
-Though state 3 inferences would not directly correspond to specific biological process, they still have value as spatially and phylogenetically aware labels for trimming poorly supported segments from alignments.
-
-The phylo-HMM could be further enhanced by expanding its emission distribution to include more symbols in the amino acid alphabet.
-This would allow it to better model observed substitution patterns between specific symbols, for example the high rate of exchange between gaps and glutamine residues caused by polyglutamine tracts.
-The transitions between amino acids could be parametrized with a published matrix such as LG~\cite{Le2008}, but the transitions between amino acids and gaps would be inferred from labeled data.
-It is unclear if the resulting gain in accuracy would justify the increased computational burden, however.
-
-Though there are benchmarks available for optimizing and comparing methods of orthology inference, the metrics are calculated over a set of reference genomes which are sparsely sampled over a broad taxonomic range, so it is unclear if they are informative for method designed to yield robust inferences when the genomes are highly related~\cite{Nevers2022}.
-Furthermore, the heterogeneity of genome architectures and annotations may require quality assurance methods tailored to each set of genomes.
-Thus, there is likely no one-size-fits-all approach to orthology inference, and with many other standalone programs available for more standard use cases (Hieranoid, OMA standalone, OrthoFinder, OrthoInspector, Orthologer), we have chosen not to package the code into an end-to-end pipeline~\cite{Kaduk2017, Altenhoff2019, Emms2019, Linard2014, Zdobnov2020}.
-Instead, we have devoted considerable attention to organizing and documenting the code to make it accessible to a newcomer and thereby facilitate the adaptation of specific steps to similar projects as needed.
-
-In contrast, though many HMM packages are available for the Python programming language, we found none were satisfactorily documented or contained tutorials to introduce HMMs and their APIs to a wide audience.
-We therefore refactored this code into a package available on PyPI and GitHub called Homomorph.
-The package itself only implements standard HMM algorithms, but the GitHub repository includes tutorials that introduce the API and implement training routines.
-Similar tutorials for machine learning libraries such as TensorFlow have undoubtedly fueled the application of neural networks across diverse fields, but HMMs are also powerful models that can be more appropriate when the data obey certain statistical or structural constraints.
-Thus, we hope this package and its accompanying tutorials will serve as an on-ramp to HMMs and spur their greater adoption by non-specialists.
-
-Though databases of orthologous groups such as COGs, Ensembl Compara, EggNOG, OMA, OrthoDB, OrthoInspector, and OrthoMCL will continue to be useful for comparative studies across broad taxonomic ranges, the increasing speed at which high-quality genome assemblies and annotations are produced means no single database can encompass the most complete data~\cite{Galperin2020, Herrero2016, HuertaCepas2018, Altenhoff2020, Zdobnov2020, Nevers2018, Chen2006}.
-Furthermore, since many early comparative genomics studies spanned diverse branches of the tree of life, future research will likely prioritize taxonomic depth over breadth.
-Thus, custom sets of orthologous groups will grow more and not less common.
-Despite the challenges these developments pose, they also present new opportunities to bridge the gap between mutational and macroevolutionary processes.
-
-\section*{Materials and methods}
-\subsection*{Sequence de-duplication and BLAST search parameters}
-The protein sequences for each annotation were de-duplicated by removing any sequences which had already appeared in association with the same gene.
-Thus, the first accession associated with a sequence and gene pair is the sequence’s representative accession for the gene.
-BLAST+ 2.13.0 was used for the sequence similarity searches~\cite{Camacho2009}.
-An E-value cutoff of 1 was used for the initial searches.
-However, this cutoff was lowered to 1E-10 during processing of the BLAST output.
-
-\subsection*{Extraction of HSPs from BLAST output}
-To reduce the computational burden of merging HSPs into hits, the BLAST output was filtered to extract HSPs associated with the highest-scoring gene.
-The HSPs were first grouped by target protein, and the resulting groups were sorted in descending order by the bit score of their highest-scoring HSP.
-Iterating over the groups, all the HSPs in a group were passed to the next step until the parent gene of the group was not the parent gene of the highest ranked group.
-This method collects all candidate HSPs for a target gene if the highest-scoring HSP within a group exceeds the highest-scoring HSP of the next best gene.
-This is in some senses an extension of the best hit criterion where hits are considered at the level of genes rather than proteins.
-If multiple genes tied for the highest-scoring HSP, the iteration stopped when the parent gene of the current group matched none of these highest-scoring genes.
-
-\subsection*{Merging of HSPs into hits}
-HSPs were merged in two stages where the first combined non-overlapping HSPs, and the second combined the remaining HSPs.
-In the first stage, proceeding from highest to lowest bit score, HSPs were marked as “disjoint” if they do not overlap with any other HSP previously marked as disjoint.
-Although this greedy strategy does not necessarily yield the highest-scoring set of disjoint HSPs, it prioritized higher scoring HSPs.
-In the second stage, all disjoint HSPs were marked as “compatible,” and proceeding from highest to lowest bit score the remaining HSPs were marked as compatible if the overlap with any other compatible HSP was no more than 50\% of the length of either.
-The best hit for each query was chosen as the hit with the highest sum of bit scores from disjoint HSPs.
-The best hits were then filtered by overlap and reciprocity criteria.
-The overlap criterion was applied first and requires that 50\% of residues in the query are aligned in compatible HSPs.
-This excludes false positives from conserved domains embedded in larger non-homologous proteins by ensuring the hits span a sufficient fraction of the query and target sequences.
-The reciprocity criterion requires each query-target pair has a corresponding hit where the roles are reversed, which ensures there is no ambiguity in which target is the best match for the query.
-
-\subsection*{Clustering by \textit{k}-clique percolation}
-\textit{k}-clique percolation was implemented in two steps.
-In the first, maximal cliques were identified.
-In the second, a percolation graph is constructed by defining edges between cliques if they have \textit{k}-1 nodes in common.
-Clusters are the connected components of this second graph.
-The first step uses the NetworkX implementation of a maximal clique algorithm.
-The second step uses a modification of the NetworkX implementation of the \textit{k}-clique community algorithm.
-The NetworkX implementation exhaustively finds all edges in the percolation graph.
-Since joining a \textit{k}-clique community only requires that a clique has a single edge connecting it to that a community, this approach is needlessly expensive for large graphs.
-The custom implementation instead uses a progressive approach where each clique is checked against a list of known communities, merging communities as necessary in each step.
-
-The hit graph is sparse, so these algorithms are efficient when applied to its individual connected components.
-However, some components have a structure with many maximal cliques, which dramatically slows the first or second step of the clique percolation algorithm.
-Thus, if either step exceeded 90 s, the process timed out, and the simpler \textit{k}-core algorithm was used instead.
-Out of over 10,000 connected components, only seven timed out, and many of those contained highly dense clusters of histone sequences.
-
-\subsection*{Addition of paralogs to orthologous groups}
-The protein sequences for each annotation were searched against themselves with the same settings as for the inter-genome searches.
-The resulting output was processed identically except the HSPs were not filtered using the best gene criterion.
-Thus, all HSPs for each query were merged into hits.
-The best hit for each query and target gene was then chosen as the hit with the highest sum of bit scores from disjoint HSPs.
-(Grouping by target gene ensured only the highest-scoring isoform was selected.) Query-target pairs whose hits which exceeded the maximum bit score for all inter-genome hits associated with that query and passed the overlap and reciprocity filters were designated as paralogous pairs.
-The orthologous groups were supplemented with paralogs by adding the paired sequences for each of the original members of the orthologous group.
-
-\subsection*{Initial alignment and selection of representative sequences}
-The sequences in each orthologous group were aligned using MAFFT 7.490 with the following settings: --globalpair --maxiterate 1000 --thread 1 --anysymbol --allowshift --unalignlevel 0.4 --leavegappyregion~\cite{Katoh2013}.
-Representative sequences for each gene were then selected by maximum likelihood according to binary profiles constructed from these alignments.
-First each sequence was coded into gap and non-gap symbols.
-The sequences were then grouped by gene, and for each group and position if at least one sequence was aligned in the group, the group contributed one count for the non-gap symbol to the profile at that position.
-Otherwise, the group contributed a count for the gap symbol at that position.
-To account for the phylogenetic dependencies between sequences, the counts were weighted according to a Gaussian process over the GTR2 consensus tree described in the section on inferring species trees~\cite{Altschul1989}.
-If a species had multiple genes in the orthologous group, the species weight was divided evenly among them.
-Each coded sequence was then scored according to this profile, and the maximum likelihood sequence for each gene was selected as its representative.
-By assigning a non-gap count to groups and positions where at least one sequence is aligned, the profile incentivizes the selection of sequences contain the fewest gaps and match the consensus alignment.
-Unfortunately, this scheme can cause a sequence to score negative infinity if it has a gap at a position where every group has at least one aligned sequence, so the profile was initialized with a pseudocount of 0.005 for the gap and non-gap symbols at each position.
-
-\subsection*{Alignment refinement}
-Consensus columns were manually defined as columns in the initial alignments with at least five non-gap symbols using the Stockholm file format.
-These labeled alignments were used to build a profile HMM using HMMer 3.3.2~\cite{Eddy2009}.
-No sequence relative weighting was used, and the effective number of sequences was set to three times the number of sequences in the alignment.
-HMMer uses strong priors to allow the construction of profiles with relatively few aligned sequences.
-However, in rare cases a sequence with a long poorly supported segment was misaligned, effectively putting it out of register with the profile and creating a poor-quality alignment.
-Increasing the effective sequence number corrected this error.
-The original sequences were then aligned to the profile.
-Unaligned segments at the N- or C- terminus were removed from the alignment.
-Unaligned segments in internal regions were extracted and aligned separately for each region using MAFFT with the same settings.
-The resulting alignments were then “stitched” into the profile alignment.
-This process was unsuccessful for one alignment due to cryptic error in MAFFT, which reduced the number of refined alignments to 8,565.
-
-This method corrects most cases of misalignment since poorly supported segments are typically aligned to fewer than four other sequences.
-However, it does not remove poorly supported segments from columns with more than five non-gap symbols, and their contribution is sometimes sufficient to reproduce the misalignment when the sequences are aligned to the profile HMM.
-A lightweight convolutional neural network was therefore trained to recognize poorly supported segments to remove them more completely from the profile HMM.
-The network consisted of a two-dimensional embedding from a 22-dimensional one-hot-encoded vector (20 amino acids, an ambiguous amino acid, and a gap), two convolutional layers, and a one-dimensional output.
-The first layer was composed of four filters.
-Two of these filters had a kernel size of 10 and a dilation of 1.
-The other two filters had a kernel size of 60 with a dilation of 3.
-The second layer was composed of two filters with a kernel size of 10 and a dilation of 1 (Fig.~\ref{sfig:cnn_architecture}).
-The network was trained on 88,513 hand-labeled positions across 122 sequences in 46 unique alignments.
-These examples were largely long segments aligned to few, if any, other sequences (Fig.~\ref{sfig:cnn_training}).
-
-The trained network outputs a probability that each position in an aligned sequence is part of a long poorly supported segment.
-A simple algorithm was used to detect positions associated with high confidence long poorly supported segments even if the output at that position was relatively low.
-First, “seeds” were identified at positions where the output exceeded 0.75.
-These seeds were expanded outwards until the output was below 0.05.
-A seed and its expansion composed a long poorly supported segment.
-Consensus columns were defined as columns with at least five non-gap symbols, excluding positions labeled as long poorly supported segments.
-To maintain the contribution of long poorly supported segments to the profile’s models of insertions, long poorly supported segments were replaced with gaps only at positions aligned to consensus columns.
-
-\subsection*{Alignment curation}
-The alignments were coded into gap and non-gap symbols to simplify the emission distributions.
-The ``insertion'' phylo-HMM was composed of the four hidden states described in the main text.
-The emission distributions for each consisted of two components which modeled the gap pattern and the propensity for those patterns to remain constant (``gap stickiness''), respectively.
-The first component was a two-state Markov process which was in turn composed of two subprocess.
-The first was a phylogenetic process on the on the GTR2 consensus tree described in the section on inferring species trees, and the second was jump process at the tips.
-The second component was a beta-Bernoulli distribution on the number of symbols which were constant between subsequent columns.
-The ``missing data'' phylo-HMM was composed of two hidden states which were both parameterized with the same two-state, two component Markov process as the insertion phylo-HMM.
-However, the emission probabilities were calculated as the posterior probability of the observed symbol given the data rather than the probability of the data.
-Only the alignments of the single copy orthologous groups were curated, so each tip in the species tree would correspond to a single sequence in the alignment.
-
-All HMM algorithms were implemented with custom code which is available as the package Homomorph on PyPI.
-The insertion phylo-HMM was trained on 39,298 manually labeled columns in 11 alignments, and the deletion phylo-HMM was trained on 58,655 manually labeled positions in 20 sequences in 9 unique alignments (Fig.~\ref{sfig:hmm_training}).
-Because maximum-likelihood estimation of the model parameters yielded posterior decoding curves which toggled between hidden states too rapidly, the models were instead trained discriminatively~\cite{Krogh1999}.
-The difference, briefly, is maximum-likelihood estimation finds the parameters that best reproduce the observed distributions whereas discriminative training finds the parameters that minimize prediction error.
-Discriminatively trained models typically perform better in practice since real-world data are rarely fully described by the distribution specified by the model.
-
-The posterior distributions over states were computed for both the insertion and missing data phylo-HMMs.
-To trim the alignments, the posterior distributions from the insertion phylo-HMM were converted to state assignments with the following method.
-First the probabilities for states 2 and 3 were combined.
-Since this probability can change rapidly or gradually depending on the local context, a simple cutoff would not necessarily define the boundaries as the columns where the gap pattern changes most abruptly.
-Instead, a high cutoff of 0.9 was used to define a “seed” region.
-The boundaries of the seed were then expanded both inwards and outwards to define intervals from which to select boundaries.
-The outward expansion halted when the probability or its derivative was below 0.05 and 0.001, respectively.
-The inward expansion halted when the derivative was below 0.02.
-The left and right boundaries were chosen as the columns in each expansion with the maximum product between the derivative and the number of sequences which changed to or from a gap between columns.
-By combining where the model’s confidence changed rapidly with the observed change in the gap pattern, this method generally selected reasonable boundaries.
-In contrast, the posterior probabilities from the missing data phylo-HMM changed sharply, so positions were labeled as the ``missing data'' state if the posterior probability exceeded 0.95.
-These assignments can be used to filter segments or entire sequences from downstream analyses and are available in the supplementary data.
-
-\subsection*{Inference of species trees}
-The orthology inference pipeline was first repeated with the outgroup species \textit{Scaptodrosophila lebanonensis}.
-Then orthologous groups with one sequence for each species were aligned, and 100 meta-alignments were constructed by randomly sampling 10,000 columns from these 9,435 alignments.
-(The alignments were not refined with HMMer before sampling.) To determine the effect of invariant columns and gaps, two sampling strategies were used where invariant columns were allowed or disallowed and the maximum fraction of gaps was set at 0, 50, and 100\%.
-Their combination yielded six different sets of meta-alignments.
-A tree was fit to each meta-alignment with the LG substitution model and four discrete gamma rate categories using IQ-TREE 1.6.12~\cite{Nguyen2014, Le2008, Yang1994}.
-If the sampling strategy allowed invariant columns, an invariant rate category was included.
-The resulting trees from each set were then merged into a majority consensus tree (Fig.~\ref{sfig:trees_LG}).
-
-To fit trees using the GTR model of nucleotide substitution, the protein alignments were converted to nucleotide alignments using the corresponding coding sequences in the genome annotations.
-Some protein sequences are “low quality,” meaning their coding sequences contain frameshifts, premature stop codons, or other errors even though they are strong hits to known protein-coding genes.
-The NCBI annotation pipeline corrects some of these defects in the protein sequences, which can complicate a simple “reverse translation” of the alignment.
-After rejecting alignments where the expected translation from a coding sequence differed from its corresponding protein sequence, 3,425 alignments remained.
-Consensus tree were derived from meta-alignments sampled from these alignments using the approach described previously except the GTR model was used in place of the LG model.
-
-To fit trees using the two-state GTR model of substitution, the protein alignments were first coded into gap or non-gap symbols.
-As before, 100 meta-alignments were constructed from these coded alignments for each sampling strategy.
-In this case, only the presence of invariant columns was varied, yielding two sets of meta-alignments.
-Trees were then fit using the GTR2 model with no rate categories.
-An invariant category, however, was included if invariant columns were allowed.
-Since the bootstrap confidences were sometimes lower than 50\%, the resulting trees from each set were then merged into a loose consensus tree to prevent multifurcations (Fig.~\ref{sfig:trees_GTR}).
-
-\subsection*{Code and data availability}
-The code used to produce the results and analyses is available at \url{https://github.com/marcsingleton/orthology_inference2023}.
-HMM algorithms were implemented in the standalone package Homomorph which is available at \url{https://github.com/marcsingleton/homomorph} and on the Python Package Index (PyPI).
-The following Python libraries were used: matplotlib, NumPy, pandas, SciPy, and TensorFlow~\cite{Hunter2007, Harris2020, McKinney2010, Virtanen2020, Abadi2015}.
-Relevant output files are available in the supporting information.
-There is no primary data associated with this manuscript.
-All primary data are available from publicly accessible sources described in their corresponding sections.
-
-\section*{Acknowledgments}
-We thank the past and present members of the Eisen lab for helpful discussions and their assistance in revising the manuscript.
+\input{chapter1.tex}
 
@@
 
-\section*{Supporting information}
-\setcounter{figure}{0}
-\renewcommand{\thefigure}{S\arabic{figure}}
-
-\begin{figure}[h!]
-\includegraphics[width=0.5\textwidth]{ncbi/out/ncbi.png}
-\centering
-\caption{\textbf{Cumulative number of different eukaryotic genomes annotated by NCBI.}}
-\label{sfig:ncbi}
-\end{figure}
-
-\begin{figure}[h!]
-\includegraphics[width=\textwidth]{clusters/out/clusters.png}
-\centering
-\caption{\textbf{Statistics of orthologous groups.}
-\textbf{(A)} Each species is equally represented in orthologous groups (OGs).
-\textbf{(B)} A plurality of orthologous groups contain all species.
-\textbf{(C)} Nearly all proteins are associated with a single orthologous group.
-\textbf{(D)} The number of orthologous groups associated with a species is strongly correlated with the number of unique annotated proteins, which suggests the annotation pipeline generally identifies conserved genes.}
-\label{sfig:clusters}
-\end{figure}
-
-\begin{figure}[h!]
-\includegraphics[width=\textwidth]{paralogs/out/paralogs.png}
-\centering
-\caption{\textbf{Addition of paralogs to orthologous groups.}
-\textbf{(A)} Most orthologous groups (OGs) have no in-paralogs.
-\textbf{(B, D)} Of the groups with paralogs, most have fewer than five.
-\textbf{(C)} The in-paralogs are generally only a small fraction of the sequences in an orthologous group.}
-\label{sfig:paralogs}
-\end{figure}
-
-\begin{figure}[h!]
-\includegraphics[width=\textwidth]{cnn_architecture.png}
-\centering
-\caption{\textbf{CNN architecture.}}
-\label{sfig:cnn_architecture}
-\end{figure}
-
-\begin{figure}[h!]
-\includegraphics[width=\textwidth]{cnn_training/out/cnn_training.png}
-\includegraphics[width=\textwidth]{cnn_training/out/23D9_XP_026832050.1.png}
-\centering
-\caption{\textbf{CNN data and training details.}
-\textbf{(A)} Most labels were “negative,” which corresponded to positions which were maintained in the alignment.
-\textbf{(B, C)} The training loss and metrics stabilized by the final training epochs.
-\textbf{(D, E)} The parameters for a model of this size are easily visualized.
-The embedding learns the gap and unknown amino acid symbols are distinct from the others.
-The convolutional filters learn kernels which correspond to the length of a typical long poorly supported segment.
-\textbf{(F)} The alignment of representative sequences in orthologous group 23D9 where the indicated sequence is labeled with the trained CNN.
-The blue lines are the CNN’s predictions, and the yellow lines are the labels.
-The red lines are the weights given to each position.
-The regions where the weights are zero are masked from the training data, which allowed the model to focus on only the most relevant examples.}
-\label{sfig:cnn_training}
-\end{figure}
-
-\begin{figure}[h!]
-\includegraphics[width=\textwidth]{hmm_training/out/hmm_training.png}
-\centering
-\caption{\textbf{Insertion phylo-HMM data and training details.}
-\textbf{(A)} Most columns in the training data were labeled as state 1A or 1B.
-\textbf{(B)} The model loss stabilized by the final training iteration.
-\textbf{(C-F)} The values of parameters in the phylogenetic process, the jump process, the pattern stickiness model, and the transition matrix, respectively, at each training iteration.
-The transition matrix plots are the transition rates to the state indicated on the vertical axis and given in log scale.
-Self transitions are excluded.}
-\label{sfig:hmm_training}
-\end{figure}
-
-\begin{figure}[h!]
-\includegraphics[width=\textwidth]{hmm_trim/out/hmm_trim.png}
-\centering
-\caption{\textbf{Insertion phylo-HMM trimming details.}
-\textbf{(A)} Most alignments were not trimmed of any columns.
-\textbf{(B)} Most alignments have fewer than 5 trimmed regions which cumulatively account for less than 10\% of the columns in the original alignment.
-Alignments with more trimmed columns generally have one one large trimmed region rather than many small ones.
-\textbf{(C)} Most trimmed regions were inferred as state 3.}
-\label{sfig:hmm_trim}
-\end{figure}
-
-\begin{figure}[h!]
-\includegraphics[width=\textwidth]{trees/out/trees2_LG.png}
-\centering
-\caption{\textbf{Phylogenetic trees created by different sampling strategies under LG model.}}
-\label{sfig:trees_LG}
-\end{figure}
-
-\begin{figure}[h!]
-\includegraphics[width=\textwidth]{trees/out/trees2_GTR.png}
-\centering
-\caption{\textbf{Phylogenetic trees created by different sampling strategies under GTR model.}}
-\label{sfig:trees_GTR}
-\end{figure}
-
-\setcounter{table}{0}
-\renewcommand{\thetable}{S\arabic{table}}
-
-\begin{table}[h!]
-\centering
-\caption{\textbf{Genome annotations.}}
-\begin{tabular}{|l|l|l|l|l|}
-\hline
-\textbf{Species}                       & \textbf{Species ID} & \textbf{Taxon ID} & \textbf{Version} & \textbf{Source} \\ \hline
-\textit{Drosophila ananassae}          & dana                & 7217              & 102              & NCBI            \\ \hline
-\textit{Drosophila biarmipes}          & dbia                & 125945            & 102              & NCBI            \\ \hline
-\textit{Drosophila bipectinata}        & dbip                & 42026             & 102              & NCBI            \\ \hline
-\textit{Drosophila elegans}            & dele                & 30023             & 102              & NCBI            \\ \hline
-\textit{Drosophila erecta}             & dere                & 7220              & 101              & NCBI            \\ \hline
-\textit{Drosophila eugracilis}         & deug                & 29029             & 102              & NCBI            \\ \hline
-\textit{Drosophila ficusphila}         & dfic                & 30025             & 102              & NCBI            \\ \hline
-\textit{Drosophila grimshawi}          & dgri                & 7222              & 103              & NCBI            \\ \hline
-\textit{Drosophila guanche}            & dgua                & 7266              & 100              & NCBI            \\ \hline
-\textit{Drosophila hydei}              & dhyd                & 7224              & 101              & NCBI            \\ \hline
-\textit{Drosophila innubila}           & dinn                & 198719            & 100              & NCBI            \\ \hline
-\textit{Drosophila kikkawai}           & dkik                & 30033             & 102              & NCBI            \\ \hline
-\textit{Drosophila mauritiana}         & dmau                & 7226              & 100              & NCBI            \\ \hline
-\textit{Drosophila melanogaster}       & dmel                & 7227              & FB2022\_02       & FlyBase         \\ \hline
-\textit{Drosophila mojavensis}         & dmoj                & 7230              & 102              & NCBI            \\ \hline
-\textit{Drosophila navojoa}            & dnav                & 7232              & 101              & NCBI            \\ \hline
-\textit{Drosophila novamexicana}       & dnov                & 47314             & 100              & NCBI            \\ \hline
-\textit{Drosophila obscura}            & dobs                & 7282              & 101              & NCBI            \\ \hline
-\textit{Drosophila persimilis}         & dper                & 7234              & 101              & NCBI            \\ \hline
-\textit{Drosophila pseudoobscura}      & dpse                & 7237              & 104              & NCBI            \\ \hline
-\textit{Drosophila rhopaloa}           & drho                & 1041015           & 102              & NCBI            \\ \hline
-\textit{Drosophila santomea}           & dsan                & 129105            & 101              & NCBI            \\ \hline
-\textit{Drosophila sechellia}          & dsec                & 7238              & 101              & NCBI            \\ \hline
-\textit{Drosophila serrata}            & dser                & 7274              & 100              & NCBI            \\ \hline
-\textit{Drosophila simulans}           & dsim                & 7240              & 103              & NCBI            \\ \hline
-\textit{Drosophila subobscura}         & dsob                & 7241              & 100              & NCBI            \\ \hline
-\textit{Drosophila subpulchrella}      & dspu                & 1486046           & 100              & NCBI            \\ \hline
-\textit{Drosophila suzukii}            & dsuz                & 28584             & 102              & NCBI            \\ \hline
-\textit{Drosophila takahashii}         & dtak                & 29030             & 102              & NCBI            \\ \hline
-\textit{Drosophila teissieri}          & dtei                & 7243              & 100              & NCBI            \\ \hline
-\textit{Drosophila virilis}            & dvir                & 7244              & 103              & NCBI            \\ \hline
-\textit{Drosophila willistoni}         & dwil                & 7260              & 102              & NCBI            \\ \hline
-\textit{Drosophila yakuba}             & dyak                & 7245              & 102              & NCBI            \\ \hline
-\textit{Scaptodrosophila lebanonensis} & sleb                & 7225              & 100              & NCBI            \\ \hline
-\end{tabular}
-\label{stable:genomes}
-\end{table}
-
-\begin{table}[h!]
-\centering
-\caption{\textbf{Phylogenetic diversity criteria.}}
-\begin{tabular}{|l|l|}
-\hline
-\textbf{Species   IDs} & \textbf{Minimum   number} \\ \hline
-dinn,   dgri, dhyd     & 2                         \\ \hline
-dnov,   dvir           & 1                         \\ \hline
-dmoj,   dnav           & 1                         \\ \hline
-dper,   dpse           & 1                         \\ \hline
-dgua,   dsob           & 1                         \\ \hline
-dana,   dbip           & 1                         \\ \hline
-dkik,   dser           & 1                         \\ \hline
-dele,   drho           & 1                         \\ \hline
-dtak,   dbia           & 1                         \\ \hline
-dspu,   dsuz           & 1                         \\ \hline
-dere,   dtei           & 1                         \\ \hline
-dsan,   dyak           & 1                         \\ \hline
-dmel                   & 1                         \\ \hline
-dmau,   dsec, dsim     & 1                         \\ \hline
-\end{tabular}
-\label{stable:diversity_criteria}
-\end{table}
+\input{chapter1_supporting.tex}
 

### 2022-11-17 chapter1/chapter1.tex
Wrapped code availability in sloppypar to prevent overfull lines

@@
 \subsection*{Code and data availability}
+\begin{sloppypar}
 The code used to produce the results and analyses is available at \url{https://github.com/marcsingleton/orthology_inference2023}.
@@
 All primary data are available from publicly accessible sources described in their corresponding sections.
+\end{sloppypar}
 

### 2022-11-17 chapter1/chapter1.tex
Moved acknowledgements to biorxiv.tex
Adjusted format of abstract in main and biorxiv

@@
 \begin{abstract}
+\noindent
 Identifying protein sequences with common ancestry is a core task in bioinformatics and evolutionary biology.
@@
 \end{sloppypar}
-
-\section*{Acknowledgments}
-We thank the past and present members of the Eisen lab for helpful discussions and their assistance in revising the manuscript.

### 2022-11-18 chapter1/chapter1.tex
Added discussion of setting local k values

@@
 The optimal value of \textit{k} for a given set of genomes is unclear and likely depends on the desired trade-off between sensitivity and specificity.
-However, percolation theory or simulations may yield insights.
+Furthermore, \textit{k} is not necessarily a global parameter and can instead depend on the properties of each connected component.
+For example, one possibility is to take an entire component as an orthologous group if its number of unique genes and unique species are equal since all the sequences are isoforms of a single set of genes.
+This would effectively set \text{k} equal to one for this component.
+Another approach is to make \textit{k} an decreasing function of the density of edges, so sparser graphs are clustered more permissively.
+However, percolation theory or simulations may yield additional insights.
 

### 2022-11-19 introduction/introduction.tex
Added introduction with outline

======= OUTLINE =======

BACKGROUND
Survery of historical developments that inform dissertation
Write at a high school or early college biology level

I. Proteins are the molecular machines that drive biology
    A. Four major macromolecules and briefly discuss broad roles of each
        1. Nucleic acids
        2. Carbohydrates
        3. Lipids
        4. Protein
    B. Some major examples of proteins
        1. Hemoglobin and sickle cell
        2. Color blindness and opsins
        3. Lactase and lactose intolerance
II. Proteins are linear chains of amino acids with a three-dimensional structure
    A. Background on biochemistry, primary, secondary, and tertiary structure
    B. Structure-function relationships returning to previous hemoglobin example
    C. Structural biology seeks to determine protein structures to understand function
    D. Overview of structural methods including purification steps
        1. X-ray crystallography
        2. NMR
        3. cryo-EM
III. Disordered proteins do not adopt a stable structure
    A. Definition and discussion of protein disorder
        1. Structural heterogeneity, conformational ensembles, floppy
        2. Major sequence characteristics and structural "modes"?
    B. Disordered proteins are challenging to study with traditional methods
    C. Genome sequencing and computational tools revealed disordered proteins are ubiquitous
IV. Disordered proteins perform varied functions
    A. Flexibility confers several advantages
    A. Survey of disordered proteins and their functions
        1. Linkers
        2. Signaling
V. Disordered proteins are enriched in transcriptional machinery
    A. Introduce transcription via the central dogma
    B. Describe transcription factors, cofactors, transcriptional machinery
        1. Examples of different transcription factors
    C. Describe properties of transcription factors
        1. DNA binding and activation domains
        2. Different flavors of activation domains
    D. How transcription factors find their targets to initiate specific genetic programs is one of the major outstanding questions in molecular biology
VI. Disordered proteins are associated with phase separation
    A. Describe phase separation in general
    B. Describe phase separation as related to proteins
    C. Describe functions of phase separation and relationship to subcellular structures
VII. Phase separation via IDRs is a possible mechanism for initiating transcription
    A. Summarize Stadler's papers and others

AIMS
Discuss more recent developments in field and propose driving questions

I. Structure-function relationships of disordered proteins are unclear
    A. Examples of studies that have found specific cases of conservation in IDRs
    B. Alan Moses paper as reference in yeast
II. Use evolutionary analyses to identify conserved properties of disordered regions
    A. Conservation is powerful signal of function
        1. Examples of other studies that applied evolution
    B. Outline overall approach and make correspondence to chapters
        1. Identification of orthologous groups
        2. Application of evolutionary analyses

======= DRAFT =======

Life is a physical phenomenon. Despite the complexity of the cells, tissues, and organs that compose living things, all these processes are governed by the same physical laws that describe how the planets revolve around the sun and the probability that an atom of uranium decays from one moment to another. However, as many systems are too complex to describe with physical models and equations, scientists simplify to achieve levels of abstraction that are more useful. (This is the origin of the common observation that biology is applied chemistry and chemistry is applied physics.) Since life spans a scale from the size of single cell to entire ecosystems, biology likely employs more layers of abstraction than any other scientific discipline. One of the most powerful and widely used frameworks within the life sciences is biochemistry, which characterizes biological processes in terms of their component molecules. A specific focus is four classes of macromolecules called nucleic acids, carbohydrates, lipids, and proteins, all of which are unique to biological systems. Though not all biological molecules are macromolecules and not all biological macromolecules fit neatly in one of these four categories, much of life at the molecular level is understood in terms of their structure and function. Each of the four has a characteristic role. Nucleic acids, i.e. DNA and RNA, are responsible for information storage and transfer. Carbohydrates primarily store energy but can also act as structural components of cells. Lipids are a diverse class of oily molecules which are components of cell membranes, store energy, and transmit signals. Proteins have a diverse range of functions including but not limited to catalyzing reactions, transmitting signals, and providing structure. Proteins have such a diverse range of associated functions, many of which overlap with the other three categories, because proteins are involved in virtually every biological process. Moreover, proteins are highly dynamic. They respond to signals, change shape, and even use the other macromolecules as raw material for their activities. In this way, proteins are like molecular machines.

Macromolecule is a loose term applied to molecules with high relative molecular mass. In practice, it usually refers to one of the classes listed above, among a few other prominent non-biological examples.

### 2022-11-19 introduction/introduction.tex
Polished up some sentences in introduction

@@
 Life is a physical phenomenon.
-Despite the complexity of the cells, tissues, and organs that compose living things, all these processes are governed by the same physical laws that describe how the planets revolve around the sun and the probability that an atom of uranium decays from one moment to another.
-However, as many systems are too complex to describe with physical models and equations, scientists simplify to achieve levels of abstraction that are more useful.
-(This is the origin of the common observation that biology is applied chemistry and chemistry is applied physics.) Since life spans a scale from the size of single cell to entire ecosystems, biology likely employs more layers of abstraction than any other scientific discipline.
+Despite the complexity of living things, all these processes are governed by the same physical laws that describe how the planets revolve around the sun and the probability that an atom of uranium decays from one moment to another.
+However, as many systems are too complex to describe with physical models and equations, scientists simplify these details into levels of abstraction that are more useful.
+(This is the origin of the common observation that biology is applied chemistry and chemistry is applied physics.) Since life spans a scale from a single cell to entire ecosystems, biology likely employs more layers of abstraction than any other scientific discipline.
 One of the most powerful and widely used frameworks within the life sciences is biochemistry, which characterizes biological processes in terms of their component molecules.
@@
 Each of the four has a characteristic role.
-Nucleic acids, i.e.
-DNA and RNA, are responsible for information storage and transfer.
+Nucleic acids, i.e., DNA and RNA, are responsible for information storage and transfer.
 Carbohydrates primarily store energy but can also act as structural components of cells.
 Lipids are a diverse class of oily molecules which are components of cell membranes, store energy, and transmit signals.
-Proteins have a diverse range of functions including but not limited to catalyzing reactions, transmitting signals, and providing structure.
-Proteins have such a diverse range of associated functions, many of which overlap with the other three categories, because proteins are involved in virtually every biological process.
-Moreover, proteins are highly dynamic.
-They respond to signals, change shape, and even use the other macromolecules as raw material for their activities.
-In this way, proteins are like molecular machines.
+Proteins have a range of functions, including catalyzing reactions, transmitting signals, and providing structure.
+Many of these overlap with the functions of the other macromolecule classes because proteins are involved in virtually every biological process.
+However, unlike the others, which are often passive participants, proteins are highly active and dynamic.
+They respond to signals, change shape, and even use the other macromolecules as raw material.
+Proteins are therefore the molecular machines that carry out a cell's activities.
 
@@
 In practice, it usually refers to one of the classes listed above, among a few other prominent non-biological examples.
+
+To illustrate these concepts,

### 2022-11-21 introduction/introduction.tex
Added detailed outline for paragraph on examples of proteins

@@
 Life is a physical phenomenon.
-Despite the complexity of living things, all these processes are governed by the same physical laws that describe how the planets revolve around the sun and the probability that an atom of uranium decays from one moment to another.
-However, as many systems are too complex to describe with physical models and equations, scientists simplify these details into levels of abstraction that are more useful.
-(This is the origin of the common observation that biology is applied chemistry and chemistry is applied physics.) Since life spans a scale from a single cell to entire ecosystems, biology likely employs more layers of abstraction than any other scientific discipline.
-One of the most powerful and widely used frameworks within the life sciences is biochemistry, which characterizes biological processes in terms of their component molecules.
+Despite the complexity of living things, their processes are governed by the same physical laws that describe how the planets revolve around the sun and the probability that an atom of uranium decays from one moment to another.
+However, as many systems are too complex to describe with physical models and equations, scientists simplify their details into levels of abstraction that are more useful.
+(This is the origin of the common observation that biology is applied chemistry and chemistry is applied physics.) For example, a punnettSince life spans a scale from single cells to entire ecosystems, biology likely employs more layers of abstraction than any other scientific discipline.
+One of the most powerful and widely used frameworks within the life sciences is biochemistry, which characterizes biological processes in terms of their component molecules and chemical reactions.
 A specific focus is four classes of macromolecules called nucleic acids, carbohydrates, lipids, and proteins, all of which are unique to biological systems.
@@
 Lipids are a diverse class of oily molecules which are components of cell membranes, store energy, and transmit signals.
-Proteins have a range of functions, including catalyzing reactions, transmitting signals, and providing structure.
+Proteins have a range of functions, including catalyzing reactions, transmitting signals, transporting materials, and providing structure.
 Many of these overlap with the functions of the other macromolecule classes because proteins are involved in virtually every biological process.
 However, unlike the others, which are often passive participants, proteins are highly active and dynamic.
-They respond to signals, change shape, and even use the other macromolecules as raw material.
-Proteins are therefore the molecular machines that carry out a cell's activities.
+They respond to signals, change shape, and often use the other macromolecules as substrates in their activities.
+Proteins are essentially the molecular machines that carry out life's functions.
 
-Macromolecule is a loose term applied to molecules with high relative molecular mass.
+FOOTNOTE: Macromolecule is a loose term applied to molecules with high molecular mass.
 In practice, it usually refers to one of the classes listed above, among a few other prominent non-biological examples.
 
-To illustrate these concepts,
+Some examples will illustrate the central role of proteins more clearly.
+These examples should be accessible to a wide audience, so nothing too abstract. Ideally should relate to human physiology and disease.
+    1. Hemoglobin and sickle cell
+        A. Introduce red blood cells transport oxygen
+        B. Hemoglobin is protein that binds oxygen.
+            i. Physically interacts with a molecule of oxygen
+        C. Sickle cell is disease caused by mutation in hemoglobin protein.
+            i. Cells produce hemoglobin with small mistake which causes them to clump into chains
+            ii. This distorts the shape of red blood cells into a sickle
+            iii. Red blood cells do not flow as easily through blood vessels causing pain and oxygen deprivation
+    2. Photopsins and color blindness
+        A. Photopsins are light sensitive proteins in cells in the retina at the back of the eye
+        B. Three different versions are sensitive to red, green, and blue light
+        C. Color blindness is the result of cells missing one of these proteins
+    3. Lactase and lactose intolerance
+        A. Lactose is a sugar found in milk that requires specific protein, lactase, to digest
+        B. Many humans who can digest milk products in childhood, lose this ability in adulthood suffer symptoms such as bloating and diarrhea after consuming dairy products
+        C. They stop producing lactase, so the undigested sugars are broken down by bacteria in the colon
+
+Despite performing this diverse range of functions, all proteins are made up of a set of twenty simple building blocks called amino acids.
+Each amino acid in a protein is bonded to the next, forming a linear chain like beads on a string.

### 2022-11-21 introduction/introduction.tex
Added paragraph of protein structure and added Punnett square example

@@
 Despite the complexity of living things, their processes are governed by the same physical laws that describe how the planets revolve around the sun and the probability that an atom of uranium decays from one moment to another.
-However, as many systems are too complex to describe with physical models and equations, scientists simplify their details into levels of abstraction that are more useful.
-(This is the origin of the common observation that biology is applied chemistry and chemistry is applied physics.) For example, a punnettSince life spans a scale from single cells to entire ecosystems, biology likely employs more layers of abstraction than any other scientific discipline.
+However, as many systems are too complex to describe with physical models and equations, scientists simplify them into levels of abstraction that are more useful.
+(This is the origin of the common observation that biology is applied chemistry and chemistry is applied physics.) For example, Punnett squares facilitate the prediction of genotypes and phenotypes by distilling the complexities and nuances of diverse reproductive systems into a set of simple rules.
+However, since life spans a scale from single cells to entire ecosystems, biology likely employs more layers of abstraction than any other scientific discipline.
 One of the most powerful and widely used frameworks within the life sciences is biochemistry, which characterizes biological processes in terms of their component molecules and chemical reactions.
@@
 
-Despite performing this diverse range of functions, all proteins are made up of a set of twenty simple building blocks called amino acids.
-Each amino acid in a protein is bonded to the next, forming a linear chain like beads on a string.
+Despite performing this diverse range of functions, all proteins are made from of a set of 20 simple building blocks called amino acids.
+Though each amino acid is chemically unique, they share a common backbone composed of two distinct and complementary receptor and donor sites for chemical bonds.
+Thus, in a protein the amino acids are bonded in a linear chain like beads on a string.
+However, once synthesized, proteins are not tidy rod-shaped molecules.
+Instead, the chain loops and weaves between itself creating a characteristic three-dimensional structure in a process called folding.
+These structures, which are highly stable and characteristic of each protein, are a result of the interactions between the amino acids in the chain and the surrounding medium, which is typically water.
+Because each amino acid has unique geometric and chemical properties that influence the energetics of these interactions, a protein's three-dimensional structure is encoded by the sequence of amino acids that compose it.
+A protein's function is in turn a direct result of its structure.
+For example, the structure of hemoglobin precisely positions its amino acids and a helper molecule called a heme group to create a pocket that can stably but reversibly bind oxygen.
+This allows hemoglobin to carry oxygen throughout the body until it is delivered to its destination.
+
+FOOTNOTE: The term amino acid encompasses any compound that contains an amino and carboxyl group.
+However, proteins are only synthesized from the twenty "canonical" amino acids.
+Another two (selenocysteine and pyrrolysine) are incorporated via a distinct mechanism under rare circumstances and are therefore considered non-standard.

### 2022-11-30 chapter1/chapter1.tex
Added note about scaling of axis for components figure
Updated text with new refinement process
Removed references to CNN in main text
Removed CNN methods and replaced with new methods
Removed references and citations to HMMer
Removed supplementary figures associated with CNN
Corrected a few uses of divergent instead of diverged

@@
 \textbf{(C)} Hexbin plot of the number of sequences and the number of unique genes in each component.
-\textbf{(D)} Histogram of number edges associated with each sequence, \textit{i.e.} the degree of each node.}
+\textbf{(D)} Histogram of number edges associated with each sequence, \textit{i.e.} the degree of each node.
+Only the lower 99th percentile of the distribution is shown.}
 \label{fig:components}
@@
 Thus, in our pipeline a single representative was chosen for each gene using an alignment-based strategy detailed in the methods section.
+Briefly, a statistical profile was created from an alignment of the sequences in each orthologous group, and the representative for each gene was chosen as the sequence which best matched this profile.
 
@@
 \subsection*{Alignment refinement}
-An initial alignment was created for each orthologous group as part of the selection of representative sequences.
-Though the pipeline’s quality control measures ensure a high degree of overall sequence identity between members of an orthologous group, some sequences contain long “poorly supported” segments which are not aligned to most or any other sequences in the alignment.
+Though the pipeline’s quality control measures ensure a high degree of overall sequence identity between members of an orthologous group, some sequences contain long “poorly supported” segments which have no homology to most or any other sequences in the alignment.
 Since most common multiple sequence alignment algorithms assume the sequences are largely homologous, these segments are sometimes “over-aligned” by forcing them into alignment where chance sequence similarities occur.
@@
 The aligner MAFFT has a mode for addressing over-alignment with a parameter, \textit{a\textsubscript{max}}, that adjusts the strength of the correction~\cite{Katoh2013, Katoh2016}.
-\textit{a\textsubscript{max}} varies between 0 and 1, but at values above 0.8, the correction severely degrades the alignment of conserved regions.
-Thus, in our pipeline the sequences were aligned with a conservative value of 0.4.
-Though some cases of over-alignment remained, most conserved regions were reliably aligned with few to no errors.
-To correct the remaining errors in alignments of the single copy orthologs, a statistical model of each alignment called a profile HMM was constructed from the well-aligned columns~\cite{Eddy2009}.
-The original sequences were then aligned to these profiles.
-By defining highly conserved “anchor” regions, this approach largely prevents the alignment of chance sequence similarities in long poorly supported segments.
-Furthermore, since profile HMMs explicitly model the beginnings and ends of alignments, they can fully exclude these segments from the termini.
-This effect is visible in the illustrated examples, where each has long poorly supported segments that are completely removed (Fig.~\ref{fig:realignment}).
+\textit{a\textsubscript{max}} varies between 0 and 1, with higher values yielding a stronger correction.
+While values above 0.8 completely eliminate over-alignment and successfully align highly conserved regions, the alignment as a whole is severely degraded, as even homologous sequences with a small amount of divergence are separated by gaps.
+Thus, in our pipeline orthologous groups were aligned in two stages.
+In the first, the sequences were aligned with a strong correction of 0.7.
+Highly conserved regions were identified, which divided the alignment into a complementary set of diverged regions.
+The sequences in each of these regions were then extracted and aligned separately with a more conservative value for \textit{a\textsubscript{max}} of 0.4.
+The resulting ``sub-alignments'' were ``stitched'' back into their positions in the original alignment.
+By defining highly conserved “anchor” regions, this approach largely prevents the alignment of chance sequence similarities in long poorly supported segments (Fig.~\ref{fig:realignment}).
 
@@
 A fully specified phylo-HMM requires a fixed number of hidden states and a probability distribution for each.
-Thus, we identified four distinct types of regions in the alignments, roughly corresponding to highly conserved regions with few to no gaps, divergent regions, regions with a stable gap pattern discordant with the expected phylogenetic relationships, and regions with poorly supported segments.
+Thus, we identified four distinct types of regions in the alignments, roughly corresponding to highly conserved regions with few to no gaps, diverged regions, regions with a stable gap pattern discordant with the expected phylogenetic relationships, and regions with poorly supported segments.
 For simplicity, however, we refer to the states that generate each type of region as 1A, 1B, 2, and 3, respectively.
@@
 These segments are easily identified in regions which are otherwise highly conserved, so they are also likely the result of incorrect or incomplete annotations.
-Unfortunately, they can also span more divergent regions, which complicates a simple rule- or threshold-based approach for identifying them.
+Unfortunately, they can also span more diverged regions, which complicates a simple rule- or threshold-based approach for identifying them.
 Thus, another phylo-HMM was trained to label each position in a sequence as generated by either a ``missing'' or ``not missing'' state.
@@
 The second major innovation in this work is its method for the refinement and trimming of alignments.
-In refinement, a profile HMM is constructed from the well-aligned columns of an initial alignment, and the original sequences are aligned to this profile.
-This process usually corrects errors caused by the presence of long poorly supported segments.
-In trimming, a phylo-HMM is used to remove regions and segments which are poorly supported by the phylogenetic consensus.
+Since sequences produced from automated annotation pipelines can contain long segments which are not homologous to most or any other sequences in their respective orthologous groups, their alignments may contain over-aligned or poorly supported regions which can introduce artifacts into downstream analyses.
+Thus, in refinement over-alignment is avoided by aligning the sequences in two stages.
+In the first highly conserved regions are aligned using a strong correction for over-alignment, and in the second more diverged regions are aligned with a weaker correction.
+This process usually prevents errors caused by long poorly supported segments without degrading the quality of the alignment.
+In trimming, a phylo-HMM is used to remove regions which are poorly supported by the phylogenetic consensus.
+
 While the combination of these two steps yielded high-quality alignments that are suitable for further analyses, they are an \textit{ad hoc} fix for underlying issues with the gene models and alignment algorithms.
@@
 \subsection*{Alignment refinement}
-Consensus columns were manually defined as columns in the initial alignments with at least five non-gap symbols using the Stockholm file format.
-These labeled alignments were used to build a profile HMM using HMMer 3.3.2~\cite{Eddy2009}.
-No sequence relative weighting was used, and the effective number of sequences was set to three times the number of sequences in the alignment.
-HMMer uses strong priors to allow the construction of profiles with relatively few aligned sequences.
-However, in rare cases a sequence with a long poorly supported segment was misaligned, effectively putting it out of register with the profile and creating a poor-quality alignment.
-Increasing the effective sequence number corrected this error.
-The original sequences were then aligned to the profile.
-Unaligned segments at the N- or C- terminus were removed from the alignment.
-Unaligned segments in internal regions were extracted and aligned separately for each region using MAFFT with the same settings.
-The resulting alignments were then “stitched” into the profile alignment.
-This process was unsuccessful for one alignment due to cryptic error in MAFFT, which reduced the number of refined alignments to 8,565.
-
-This method corrects most cases of misalignment since poorly supported segments are typically aligned to fewer than four other sequences.
-However, it does not remove poorly supported segments from columns with more than five non-gap symbols, and their contribution is sometimes sufficient to reproduce the misalignment when the sequences are aligned to the profile HMM.
-A lightweight convolutional neural network was therefore trained to recognize poorly supported segments to remove them more completely from the profile HMM.
-The network consisted of a two-dimensional embedding from a 22-dimensional one-hot-encoded vector (20 amino acids, an ambiguous amino acid, and a gap), two convolutional layers, and a one-dimensional output.
-The first layer was composed of four filters.
-Two of these filters had a kernel size of 10 and a dilation of 1.
-The other two filters had a kernel size of 60 with a dilation of 3.
-The second layer was composed of two filters with a kernel size of 10 and a dilation of 1 (Fig.~\ref{sfig:cnn_architecture}).
-The network was trained on 88,513 hand-labeled positions across 122 sequences in 46 unique alignments.
-These examples were largely long segments aligned to few, if any, other sequences (Fig.~\ref{sfig:cnn_training}).
-
-The trained network outputs a probability that each position in an aligned sequence is part of a long poorly supported segment.
-A simple algorithm was used to detect positions associated with high confidence long poorly supported segments even if the output at that position was relatively low.
-First, “seeds” were identified at positions where the output exceeded 0.75.
-These seeds were expanded outwards until the output was below 0.05.
-A seed and its expansion composed a long poorly supported segment.
-Consensus columns were defined as columns with at least five non-gap symbols, excluding positions labeled as long poorly supported segments.
-To maintain the contribution of long poorly supported segments to the profile’s models of insertions, long poorly supported segments were replaced with gaps only at positions aligned to consensus columns.
+In the first stage, the representative sequences in the single copy orthologous groups were aligned with the same settings as described in the previous section, except \textit{a\textsubscript{max}} was set to 0.7.
+A binary profile was created from the alignment using Gaussian process sequence weighting as described in the previous section.
+(Because the orthologous groups were single copy and contained only representative sequences, each sequence received the full weight associated with its species.) The binary profile was then converted into a binary mask by identifying where the weighted fraction of non-gap symbols exceeded 0.5.
+The binary mask was closed with a structuring element of size three, and highly conserved regions were identified as the contiguous intervals of this closed mask with a minimum length of 10.
+Diverged regions were taken as the complement of the highly conserved regions.
+For each diverged region, the corresponding segments of the sequences in the initial alignment were extracted and aligned with \textit{a\textsubscript{max}} set to 0.4.
+The resulting sub-alignments were then stitched into the initial alignment.
 
@@
 Then orthologous groups with one sequence for each species were aligned, and 100 meta-alignments were constructed by randomly sampling 10,000 columns from these 9,435 alignments.
-(The alignments were not refined with HMMer before sampling.) To determine the effect of invariant columns and gaps, two sampling strategies were used where invariant columns were allowed or disallowed and the maximum fraction of gaps was set at 0, 50, and 100\%.
+(The alignments were not refined before sampling.) To determine the effect of invariant columns and gaps, two sampling strategies were used where invariant columns were allowed or disallowed and the maximum fraction of gaps was set at 0, 50, and 100\%.
 Their combination yielded six different sets of meta-alignments.

### 2022-11-30 chapter1/chapter1_supporting.tex
Removed supporting figure renaming commands from tex file
Updated text with new refinement process
Removed references to CNN in main text
Removed CNN methods and replaced with new methods
Removed references and citations to HMMer
Removed supplementary figures associated with CNN
Corrected a few uses of divergent instead of diverged

@@
 \section*{Supporting information}
-\setcounter{figure}{0}
-\renewcommand{\thefigure}{S\arabic{figure}}
 
@@
 \label{sfig:paralogs}
-\end{figure}
-
-\begin{figure}[h!]
-\includegraphics[width=\textwidth]{cnn_architecture.png}
-\centering
-\caption{\textbf{CNN architecture.}}
-\label{sfig:cnn_architecture}
-\end{figure}
-
-\begin{figure}[h!]
-\includegraphics[width=\textwidth]{cnn_training/out/cnn_training.png}
-\includegraphics[width=\textwidth]{cnn_training/out/23D9_XP_026832050.1.png}
-\centering
-\caption{\textbf{CNN data and training details.}
-\textbf{(A)} Most labels were “negative,” which corresponded to positions which were maintained in the alignment.
-\textbf{(B, C)} The training loss and metrics stabilized by the final training epochs.
-\textbf{(D, E)} The parameters for a model of this size are easily visualized.
-The embedding learns the gap and unknown amino acid symbols are distinct from the others.
-The convolutional filters learn kernels which correspond to the length of a typical long poorly supported segment.
-\textbf{(F)} The alignment of representative sequences in orthologous group 23D9 where the indicated sequence is labeled with the trained CNN.
-The blue lines are the CNN’s predictions, and the yellow lines are the labels.
-The red lines are the weights given to each position.
-The regions where the weights are zero are masked from the training data, which allowed the model to focus on only the most relevant examples.}
-\label{sfig:cnn_training}
 \end{figure}

### 2022-11-30 chapter1/figures/cnn_training/cnn_training.py
Removed cnn_training.py

@@ -1,180 +1,0 @@
-"""Make figure of CNN training."""
-
-import os
-import re
-
-import matplotlib.pyplot as plt
-import numpy as np
-import pandas as pd
-import skbio
-import tensorflow as tf
-from matplotlib.figure import SubplotParams
-from matplotlib.gridspec import GridSpec
-from src.draw import plot_msa_data, default_sym2color
-from src.utils import read_fasta
-
-alphabet = ['A', 'R', 'N', 'D', 'C', 'Q', 'E', 'G', 'H', 'I', 'L', 'K', 'M', 'F', 'P', 'S', 'T', 'W', 'Y', 'V', 'X', '-']
-sym2idx = {sym: i for i, sym in enumerate(alphabet)}
-ppid_regex = r'ppid=([A-Za-z0-9_.]+)'
-spid_regex = r'spid=([a-z]+)'
-
-tree = skbio.read('../../orthology_inference/analysis/ortho_tree/consensus_LG/out/100R_NI.nwk', 'newick', skbio.TreeNode)
-tip_order = {tip.name: i for i, tip in enumerate(tree.tips())}
-
-# Load labels
-OGid2ppids = {}
-ppid2labels = {}
-with open('../../orthology_inference/analysis/ortho_MSA/realign_cnn/labels.tsv') as file:
-    field_names = file.readline().rstrip('\n').split('\t')
-    for line in file:
-        if line.startswith('#'):
-            continue
-        fields = {key: value for key, value in zip(field_names, line.rstrip('\n').split('\t'))}
-        if fields['active'] == 'False':
-            continue
-        OGid, ppid = fields['OGid'], fields['ppid']
-        try:
-            OGid2ppids[OGid].add(ppid)
-        except KeyError:
-            OGid2ppids[OGid] = {ppid}
-        start, stop, label = int(fields['start']), int(fields['stop']), int(fields['label'])
-        try:
-            ppid2labels[ppid].append((start, stop, label))
-        except KeyError:
-            ppid2labels[ppid] = [(start, stop, label)]
-
-# Create records
-records = {}
-for OGid, ppids in OGid2ppids.items():
-    msa1 = read_fasta(f'../../orthology_inference/analysis/ortho_MSA/get_repseqs/out/{OGid}.afa')
-
-    # Convert alignment to indices and make profile
-    ppid2idx, msa2 = {}, []
[…]

### 2022-11-30 chapter1/chapter1.tex
Made quotation marks consistent with LaTeX style

@@
 Clusters are then taken as the connected components of an overlap graph where an edge exists between two cliques if they share \textit{k}-1 nodes in common.
-An intuitive way to visualize this algorithm is by “rolling” a clique of some size \textit{k} over the graph (Fig.~\ref{fig:percolation}B).
+An intuitive way to visualize this algorithm is by ``rolling'' a clique of some size \textit{k} over the graph (Fig.~\ref{fig:percolation}B).
 More specifically, a cluster is initiated when a set of nodes which form a \textit{k}-clique is identified.
@@
 \subsection*{Alignment refinement}
-Though the pipeline’s quality control measures ensure a high degree of overall sequence identity between members of an orthologous group, some sequences contain long “poorly supported” segments which have no homology to most or any other sequences in the alignment.
-Since most common multiple sequence alignment algorithms assume the sequences are largely homologous, these segments are sometimes “over-aligned” by forcing them into alignment where chance sequence similarities occur.
+Though the pipeline’s quality control measures ensure a high degree of overall sequence identity between members of an orthologous group, some sequences contain long ``poorly supported'' segments which have no homology to most or any other sequences in the alignment.
+Since most common multiple sequence alignment algorithms assume the sequences are largely homologous, these segments are sometimes ``over-aligned'' by forcing them into alignment where chance sequence similarities occur.
 Typically, these segments remain contiguous, so the alignments alternate between short runs of columns with few or no gaps and large gap-rich regions.
@@
 The resulting ``sub-alignments'' were ``stitched'' back into their positions in the original alignment.
-By defining highly conserved “anchor” regions, this approach largely prevents the alignment of chance sequence similarities in long poorly supported segments (Fig.~\ref{fig:realignment}).
+By defining highly conserved ``anchor'' regions, this approach largely prevents the alignment of chance sequence similarities in long poorly supported segments (Fig.~\ref{fig:realignment}).
 
@@
 Since the segments in these regions are likely the result of incorrect or incomplete annotations rather than meaningful biological variation, maintaining them in the alignments would propagate spurious homologies to subsequent analyses.
-This is a common issue in alignments generated by automated pipelines, so downstream analyses often focus on the strongly supported regions by removing or “trimming” columns below some threshold number of gaps or sequence identity~\cite{Castresana2000, CapellaGutierrez2009}.
+This is a common issue in alignments generated by automated pipelines, so downstream analyses often focus on the strongly supported regions by removing or ``trimming'' columns below some threshold number of gaps or sequence identity~\cite{Castresana2000, CapellaGutierrez2009}.
 This approach, however, is inadequate if the taxonomic sampling is dense, as a single indel event along a lineage containing many species can increase the number of gaps above the threshold.
@@
 
-Phylogenetic HMMs (phylo-HMMs) are statistical models that incorporate phylogenetic and spatial information to calculate the probability that each observation in a sequence was generated by one of several “hidden states”~\cite{Felsenstein1996}.
+Phylogenetic HMMs (phylo-HMMs) are statistical models that incorporate phylogenetic and spatial information to calculate the probability that each observation in a sequence was generated by one of several ``hidden states''~\cite{Felsenstein1996}.
 Since they can evaluate both the probability of a gap pattern in a column given the known phylogenetic relationships and the local context, a phylo-HMM was used to segment the alignment into contiguous regions with different patterns of gaps.
@@
 Second, a single orthologous group could have several isoforms of a single gene if its isoforms were clustered together.
-In both cases, the presence of multiple sequences for a single gene creates ambiguities over which is the “primary” isoform.
+In both cases, the presence of multiple sequences for a single gene creates ambiguities over which is the ``primary'' isoform.
 Since the first occurs at the level of orthologous groups, the orthologous groups were first grouped by the similarity of their parent genes using a graph-based strategy.
@@
 HSPs were merged in two stages where the first combined non-overlapping HSPs, and the second combined the remaining HSPs.
-In the first stage, proceeding from highest to lowest bit score, HSPs were marked as “disjoint” if they do not overlap with any other HSP previously marked as disjoint.
+In the first stage, proceeding from highest to lowest bit score, HSPs were marked as ``disjoint'' if they do not overlap with any other HSP previously marked as disjoint.
 Although this greedy strategy does not necessarily yield the highest-scoring set of disjoint HSPs, it prioritized higher scoring HSPs.
-In the second stage, all disjoint HSPs were marked as “compatible,” and proceeding from highest to lowest bit score the remaining HSPs were marked as compatible if the overlap with any other compatible HSP was no more than 50\% of the length of either.
+In the second stage, all disjoint HSPs were marked as ``compatible,'' and proceeding from highest to lowest bit score the remaining HSPs were marked as compatible if the overlap with any other compatible HSP was no more than 50\% of the length of either.
 The best hit for each query was chosen as the hit with the highest sum of bit scores from disjoint HSPs.
@@
 Since this probability can change rapidly or gradually depending on the local context, a simple cutoff would not necessarily define the boundaries as the columns where the gap pattern changes most abruptly.
-Instead, a high cutoff of 0.9 was used to define a “seed” region.
+Instead, a high cutoff of 0.9 was used to define a ``seed'' region.
 The boundaries of the seed were then expanded both inwards and outwards to define intervals from which to select boundaries.
@@
 To fit trees using the GTR model of nucleotide substitution, the protein alignments were converted to nucleotide alignments using the corresponding coding sequences in the genome annotations.
-Some protein sequences are “low quality,” meaning their coding sequences contain frameshifts, premature stop codons, or other errors even though they are strong hits to known protein-coding genes.
-The NCBI annotation pipeline corrects some of these defects in the protein sequences, which can complicate a simple “reverse translation” of the alignment.
+Some protein sequences are ``low quality,'' meaning their coding sequences contain frameshifts, premature stop codons, or other errors even though they are strong hits to known protein-coding genes.
+The NCBI annotation pipeline corrects some of these defects in the protein sequences, which can complicate a simple ``reverse translation'' of the alignment.
 After rejecting alignments where the expected translation from a coding sequence differed from its corresponding protein sequence, 3,425 alignments remained.

### 2022-11-30 chapter1/chapter1.tex
Fixed bad figure reference
Minor wording change in refinement methods
Removed reference to first stage since second stage is not referenced in method

@@
 In contrast, columns assigned to states 2 and 3 largely corresponded to the long poorly supported segments and phylogenetically discordant regions discussed previously and were therefore removed from the alignments.
-In the example decoded alignment shown, the decoded states closely follow the expected patterns (Fig.~ref{fig:hmm})B).
+In the example decoded alignment shown, the decoded states closely follow the expected patterns (Fig.~\ref{fig:hmm}B).
 Overall, fewer than 10\% of alignments were trimmed of any regions.
@@
 \subsection*{Alignment refinement}
-In the first stage, the representative sequences in the single copy orthologous groups were aligned with the same settings as described in the previous section, except \textit{a\textsubscript{max}} was set to 0.7.
+The representative sequences in the single copy orthologous groups were aligned with the same settings as described in the previous section except \textit{a\textsubscript{max}} was set to 0.7.
 A binary profile was created from the alignment using Gaussian process sequence weighting as described in the previous section.

### 2022-11-30 chapter1/chapter1.tex
Updated realignment.py figures to reflect new alignments

@@
 Since most common multiple sequence alignment algorithms assume the sequences are largely homologous, these segments are sometimes ``over-aligned'' by forcing them into alignment where chance sequence similarities occur.
-Typically, these segments remain contiguous, so the alignments alternate between short runs of columns with few or no gaps and large gap-rich regions.
-More rarely, when long poorly supported segments are adjacent to a long gap in the same sequence, the two are interlaced, yielding long gaps interrupted by short segments of spurious alignment.
+Typically, these segments remain contiguous, so the alignments alternate between short runs of columns with few or no gaps and large gap-rich regions (Fig.~\ref{fig:realignment}A-B, left).
+More rarely, when long poorly supported segments are adjacent to a long gap in the same sequence, the two are interlaced, yielding long gaps interrupted by short segments of spurious alignment (Fig.~\ref{fig:realignment}C, left).
 
@@
 The resulting ``sub-alignments'' were ``stitched'' back into their positions in the original alignment.
-By defining highly conserved ``anchor'' regions, this approach largely prevents the alignment of chance sequence similarities in long poorly supported segments (Fig.~\ref{fig:realignment}).
+By defining highly conserved ``anchor'' regions, this approach largely prevents the alignment of chance sequence similarities in long poorly supported segments (Fig.~\ref{fig:realignment}, right).
 
@@
 \caption{\textbf{Alignment with long poorly supported segments.}
-The alignments of representative sequences in orthologous groups 23D9 \textbf{(A)}, 01E0 \textbf{(B)}, and 07C3 \textbf{(C)} before and after refinement.}
+The alignments of representative sequences in orthologous groups 0167 \textbf{(A)}, 2770 \textbf{(B)}, and 23D9 \textbf{(C)} before and after refinement.}
 \label{fig:realignment}

### 2022-11-30 chapter1/figures/realignment/realignment.py

@@ -9,12 +9,12 @@ import skbio
 from src.draw import plot_msa
 from src.utils import read_fasta
 
-records = [('23D9', 'A', 731, 1158, 704, 1000),
-           ('01E0', 'B', 0, 344, 0, 94),
-           ('07C3', 'C', 0, 484, 0, 211),
-           ('0430', 'D', 0, 476, 0, 126)]
+records = [('0167', 'A', 2357, 2714, 2763, 2831),
+           ('2770', 'B', 2440, 2831, 2235, 2743),
+           ('23D9', 'C', 731, 1150, 702, 982),
+           ('35D6', 'D', 2774, 3162, 3003, 3355)]
 msa_path1 = '../../orthology_inference/analysis/ortho_MSA/get_repseqs/out/'
-msa_path2 = '../../orthology_inference/analysis/ortho_MSA/realign_hmmer/out/mafft/'
+msa_path2 = '../../orthology_inference/analysis/ortho_MSA/realign_fastas/out/'
 spid_regex = r'spid=([a-z]+)'
 
 adjust_left1 = 0.02

### 2022-12-05 chapter1/chapter1.tex
Updated alignment trimming methods and fixed inconsistency in number style

@@
 Higher values of \textit{k} require greater overlap between a candidate node and those already in the cluster and therefore produce tighter clusters at the cost of excluding more speculative orthology relationships (Fig.~\ref{fig:percolation}C).
-We set \textit{k} to 4 as compromise between these concerns, yielding 22,813 orthologous groups, a plurality of which contained all 33 species (Fig.~\ref{sfig:clusters}).
+We set \textit{k} to four as compromise between these concerns, yielding 22,813 orthologous groups, a plurality of which contained all 33 species (Fig.~\ref{sfig:clusters}).
 
@@
 The representative sequences in the single copy orthologous groups were aligned with the same settings as described in the previous section except \textit{a\textsubscript{max}} was set to 0.7.
-A binary profile was created from the alignment using Gaussian process sequence weighting as described in the previous section.
+A binary profile was created from the alignment using Gaussian process sequence weighting as described in the section on selecting representative sequences.
 (Because the orthologous groups were single copy and contained only representative sequences, each sequence received the full weight associated with its species.) The binary profile was then converted into a binary mask by identifying where the weighted fraction of non-gap symbols exceeded 0.5.
@@
 
-The posterior distributions over states were computed for both the insertion and missing data phylo-HMMs.
-To trim the alignments, the posterior distributions from the insertion phylo-HMM were converted to state assignments with the following method.
-First the probabilities for states 2 and 3 were combined.
-Since this probability can change rapidly or gradually depending on the local context, a simple cutoff would not necessarily define the boundaries as the columns where the gap pattern changes most abruptly.
-Instead, a high cutoff of 0.9 was used to define a ``seed'' region.
-The boundaries of the seed were then expanded both inwards and outwards to define intervals from which to select boundaries.
-The outward expansion halted when the probability or its derivative was below 0.05 and 0.001, respectively.
-The inward expansion halted when the derivative was below 0.02.
-The left and right boundaries were chosen as the columns in each expansion with the maximum product between the derivative and the number of sequences which changed to or from a gap between columns.
+The posterior distributions over states were calculated for each alignment using the trained insertion phylo-HMM.
+States 2 and 3 have distinct characteristics which require different trimming strategies, so regions with a high probability of state 3 were handled first.
+Because the posterior probability of a state can change rapidly or gradually depending on the local context, a simple cutoff would not necessarily define the boundaries as the columns where the gap pattern changes most abruptly.
+Instead, a high cutoff of 0.95 was used to define a ``seed'' region.
+The left and right endpoints of the seed were then expanded both inwards and outwards to define two intervals from which boundaries were selected.
+The outward expansion halted when the probability or its derivative was below 0.01 and 0.001, respectively.
+The inward expansion halted when the derivative was below 0.001.
+The left and right boundaries were chosen as the columns in each interval with the maximum product between the derivative and the change in the gap profile between columns.
+The gap profile was calculated as the number of gaps in each column using Gaussian process sequence weighting as described in the section on selecting representative sequences.
 By combining where the model’s confidence changed rapidly with the observed change in the gap pattern, this method generally selected reasonable boundaries.
-In contrast, the posterior probabilities from the missing data phylo-HMM changed sharply, so positions were labeled as the ``missing data'' state if the posterior probability exceeded 0.95.
-These assignments can be used to filter segments or entire sequences from downstream analyses and are available in the supplementary data.
+
+Because the long poorly supported segments in state 3 regions were sometimes aligned to highly conserved columns or short segments in other sequences, trimming columns entirely would remove these segments from the other sequences even if they do not qualify as long and poorly supported themselves.
+Thus, regions with high state 3 probabilities were trimmed at the level of individual sequences rather than entire columns with the following method.
+First, the mean number of non-gap symbols in a region was calculated using Gaussian process sequence weighting.
+(The sequences with the five most non-gap symbols were also excluded to not bias the estimate with long poorly supported segments.) The final mean $\mu$ was taken as the minimum of this value and two.
+A cutoff \textit{k}, derived from a geometric model of the number of non-gap symbols and a significance level $\alpha$, was calculated using the following equation $k = \log(\alpha) / \log(1 - p) - 1$ where $p = \frac{1}{\mu + 1}$ and $\alpha = 0.01$.
+Any sequence whose number of non-gap symbols in the region equaled or exceeded this value was trimmed by replacing all non-gap symbols with gaps.
+
+To trim the remaining state 2 regions, the posterior probability of state 2 was added to a modified state 3 probability which was set to zero for any state 3 regions identified in the previous step.
+This ensured that any regions which were intermediate between state 2 and 3 were included.
+Regions were defined from this combined probability using the algorithm described previously except with a high probability cutoff of 0.9 instead of 0.95.
+The posterior probabilities from the missing data phylo-HMM were also converted into state assignments using a similar method.
+However, the initial seeds were defined with a cutoff of 0.95, and the seeds were expanded outward to first non-gap symbol or until the posterior probability of the ``missing data'' state was below 0.5.
+These assignments, which are available in the supplementary data, can be used to filter segments or entire sequences from downstream analyses.
 

### 2022-12-06 introduction/introduction.tex
Added paragraph about structural methods to introduction

@@
 This allows hemoglobin to carry oxygen throughout the body until it is delivered to its destination.
+However, people affected by sickle cell disease have a defect, i.e., a mutation, in the sequence of their hemoglobin proteins which causes it to malfunction.
+Frequently this mutation is a single change where the sixth amino acid in the sequence, glutamate, is substituted for a valine.
+This creates a sticky patch on the surface of hemoglobin, and under low-oxygen conditions normal hemoglobin changes shape to expose a sticky patch on its surface as well.
+The two patches are complementary, which allows hemoglobin proteins to clump together into long, fibrous strands.
+These strands distort the shape of red cells, giving them their characteristic sickle shape.
 
 FOOTNOTE: The term amino acid encompasses any compound that contains an amino and carboxyl group.
-However, proteins are only synthesized from the twenty "canonical" amino acids.
+However, proteins are only synthesized from the twenty ``canonical'' amino acids.
 Another two (selenocysteine and pyrrolysine) are incorporated via a distinct mechanism under rare circumstances and are therefore considered non-standard.
+
+Because sequence dictates structure which in turn dictates function, understanding these relationships in proteins is essential for understanding biological systems as a whole.
+While biologists are generally interested in function since it is the most directly associated with the biological role of a protein, function and biological role are not always easily identified or measured.
+Thus, determining a protein's structure is frequently the first step of detailed study of a protein's function.
+Though in recent years researchers have developed powerful computational tools that can accurately predict structure from sequence alone, historically structures were determined experimentally, and experimental methods remain the gold standard today.
+While many methods can reveal information about the structure of a protein, the most powerful techniques, X-ray crystallography and NMR spectroscopy, can map the spatial coordinates of every atom in a protein.
+However, this resolution requires extremely pure samples of the protein of interest.
+Since proteins are only produced by living systems, preparations begin with a complex mixture consisting of cells or tissue, and the protein of interest is isolated with series of extraction and purification steps.
+Some proteins are only produced in small amounts or degrade easily, so each step may require substantial optimization to achieve a sufficient yield.
+When structural techniques were first developed in the late 1950s, they were so time-consuming that a graduate student could dedicate an entire PhDs to solving a single protein structure.
+Many developments have substantially accelerated the process, but it remains a labor-intensive technique which may require several months of effort.
+However, the result is a powerful map that scientists use to suggest hypotheses and interpret data.
+
+FOOTNOTE: Function and biological role are frequently, though not always, exact synonyms since biological role is often defined more broadly and loosely than function.
+For example, the strict function of hemoglobin is to bind oxygen, an activity it shares with a related protein myoglobin.
+However, the two have different roles.
+Hemoglobin is found in red blood cells where it primarily acts a carrier.
+In contrast, myoglobin is found in muscle cells, where it stores oxygen until needed.
+
+The success of structural methods through the latter half of the 20th century established the view that while proteins were not completely rigid and could adopt a variety of related structures or ``conformations,'' most functional proteins had a single dominant structure.
+While some proteins had regions which returned undefined or highly variable atomic coordinates when analysed with structural techniques, these were viewed as exceptions.
+However, in the late 1990s the first genomes were sequenced, which allowed scientists to analyse the complete set of protein sequences encoded by an organism's DNA for the first time.
+
+When did scientists first notice the prevalence of IDRs?
+What tools did they use to detect it?
+In what organisms?

### 2022-12-07 chapter1/figures/hmm_training/hmm_training.py
Removed unnecessary parameter in GridSpec call

@@ -39,7 +39,7 @@ if not os.path.exists('out/'):
     os.mkdir('out/')
 
 fig = plt.figure(figsize=(7.5, 5.5), subplotpars=SubplotParams(0.3, 0.225, 0.9, 0.9, 0, 0))
-gs = GridSpec(2, 4, height_ratios=[1, 1.5], left=0.5)
+gs = GridSpec(2, 4, height_ratios=[1, 1.5])
 
 # Plot state distribution
 counts = {label: 0 for label in state_labels}

### 2022-12-07 chapter1/figures/hmm_trim/hmm_trim.py
Updated hmm_trim.py with new trimming algorithm

@@ -3,6 +3,7 @@
 import os
 
 import matplotlib.pyplot as plt
+import numpy as np
 import pandas as pd
 from matplotlib.figure import SubplotParams
 from matplotlib.gridspec import GridSpec
@@ -17,55 +18,82 @@ with open('../../orthology_inference/analysis/ortho_MSA/realign_fastas/out/error
         if error_flag1 == 'False' and error_flag2 == 'False':
             OGids.append(OGid)
 
-df = pd.read_table('../../orthology_inference/analysis/ortho_MSA/insertion_trim/out/trim_stats.tsv')
+df1 = pd.read_table('../../orthology_inference/analysis/ortho_MSA/insertion_trim/out/seq_stats.tsv')
+df2 = pd.read_table('../../orthology_inference/analysis/ortho_MSA/insertion_trim/out/region_stats.tsv')
 
-df['length'] = df['stop'] - df['start']
-df['length_ratio'] = df['length'] / df['colnum']
-df['norm2'] = df['posterior2'] / df['length']
-df['norm3'] = df['posterior3'] / df['length']
-groups = df.groupby('OGid')
+df2['length'] = df2['stop'] - df2['start']
+df2['length_ratio'] = df2['length'] / df2['colnum']
+df2['norm2'] = df2['posterior2'] / df2['length']
+df2['norm3'] = df2['posterior3'] / df2['length']
+
+groups1 = df1.groupby('OGid')
+groups2 = df2.groupby('OGid')
 
 if not os.path.exists('out/'):
     os.mkdir('out')
 
-fig = plt.figure(figsize=(7.5, 4.5), subplotpars=SubplotParams(0.075, 0.075, 0.925, 0.925))
-gs = GridSpec(2, 2)
+fig = plt.figure(figsize=(7.5, 7.5), subplotpars=SubplotParams(0.2, 0.2, 0.9, 0.9))
+gs = GridSpec(3, 2)
 
 # Pie chart by presence of trims
-values = [len(set(OGids) - set(df['OGid'])), len(set(df['OGid']))]
-labels = [f'{label}\n{value:,}' for label, value in zip(['w/o trims', 'w/ trims'], values)]
+union = set(df1['OGid']) | set(df2['OGid'])
+intersection = set(df1['OGid']) & set(df2['OGid'])
+OGids1 = set(df1['OGid']) - intersection
+OGids2 = set(df2['OGid']) - intersection
+
[…]

### 2022-12-08 chapter1/figures/hmm_trim/hmm_trim.py
Added additional panels to hmm_trim.py

@@ -32,8 +32,8 @@ groups2 = df2.groupby('OGid')
 if not os.path.exists('out/'):
     os.mkdir('out')
 
-fig = plt.figure(figsize=(7.5, 7.5), subplotpars=SubplotParams(0.2, 0.2, 0.9, 0.9))
-gs = GridSpec(3, 2)
+fig = plt.figure(figsize=(7.5, 8), subplotpars=SubplotParams(0.25, 0.325, 0.9, 0.925))
+gs = GridSpec(4, 2)
 
 # Pie chart by presence of trims
 union = set(df1['OGid']) | set(df2['OGid'])
@@ -50,7 +50,7 @@ ax.pie(values, labels=labels, labeldistance=None, colors=colors)
 ax.legend(loc='center left', bbox_to_anchor=(1, 0.5), fontsize=8)
 subfig.suptitle('A', x=0.025, y=0.975, fontweight='bold')
 
-# Hexbin of posterior2 vs posterior3
+# Hexbin of posterior2 vs posterior3 in region trims
 subfig = fig.add_subfigure(gs[0, 1], facecolor='none')
 ax = subfig.add_subplot()
 hb = ax.hexbin(df2['norm2'], df2['norm3'], bins='log', gridsize=25, mincnt=1, linewidth=0)
@@ -68,32 +68,48 @@ ax.set_xlabel('Number of sequence trims in OG')
 ax.set_ylabel('Number of OGs')
 subfig.suptitle('C', x=0.025, y=0.975, fontweight='bold')
 
-# Distribution of number of removed symbols in OG
-idx = int(np.ceil(len(df1) * 0.95))
-counts = df1['count'].sort_values(ignore_index=True)[:idx].value_counts()
-subfig = fig.add_subfigure(gs[1, 1], facecolor='none')
-ax = subfig.add_subplot()
-ax.bar(counts.index, counts.values, width=1)
-ax.set_xlabel('Number of non-gap symbols in sequence trim')
-ax.set_ylabel('Number of sequence trims')
-subfig.suptitle('D', x=0.025, y=0.975, fontweight='bold')
-
-# Distribution of number of region trims
+# Distribution of number of region trims in OGs
 counts = groups2.size().value_counts()
-subfig = fig.add_subfigure(gs[2, 0], facecolor='none')
+subfig = fig.add_subfigure(gs[1, 1], facecolor='none')
 ax = subfig.add_subplot()
 ax.bar(counts.index, counts.values, width=1)
[…]

### 2022-12-08 chapter1/chapter1_supporting.tex
Updated caption for hmm_trim figure

@@
 \caption{\textbf{Insertion phylo-HMM trimming details.}
-\textbf{(A)} Most alignments were not trimmed of any columns.
-\textbf{(B)} Most alignments have fewer than 5 trimmed regions which cumulatively account for less than 10\% of the columns in the original alignment.
-Alignments with more trimmed columns generally have one one large trimmed region rather than many small ones.
-\textbf{(C)} Most trimmed regions were inferred as state 3.}
+\textbf{(A)} Most alignments were not trimmed.
+Of the alignments with trims, most were trimmed only at the level of sequences.
+\textbf{(B)} Most trimmed regions were inferred primarily as state 2.
+\textbf{(C)} Most alignments with sequence trims have fewer than 10 segments removed.
+\textbf{(D)} Most alignments with region trims have fewer than five regions removed.
+\textbf{(E, G)} The number of non-gap symbols in sequence trims can vary considerably, but for nearly all sequence trims each non-gap symbol in the removed segment is aligned to fewer than five non-gap symbols on average.
+\textbf{(F, H)} The length of region trims can also vary considerably, but generally each region trims accounts for fewer than 10\% of the columns in the original alignment.}
 \label{sfig:hmm_trim}

### 2022-12-09 introduction/introduction.tex
Added footnote commands in introduction.tex

@@
 A specific focus is four classes of macromolecules called nucleic acids, carbohydrates, lipids, and proteins, all of which are unique to biological systems.
-Though not all biological molecules are macromolecules and not all biological macromolecules fit neatly in one of these four categories, much of life at the molecular level is understood in terms of their structure and function.
+Though not all biological molecules are macromolecules\footnote{Macromolecule is a loose term applied to molecules with high molecular mass.
+In practice, it usually refers to one of the classes listed above, among a few other prominent non-biological examples.} and not all biological macromolecules fit neatly in one of these four categories, much of life at the molecular level is understood in terms of their structure and function.
 Each of the four has a characteristic role.
@@
 Proteins are essentially the molecular machines that carry out life's functions.
-
-FOOTNOTE: Macromolecule is a loose term applied to molecules with high molecular mass.
-In practice, it usually refers to one of the classes listed above, among a few other prominent non-biological examples.
 
@@
 
-Despite performing this diverse range of functions, all proteins are made from of a set of 20 simple building blocks called amino acids.
+Despite performing this diverse range of functions, all proteins are made from of a set of 20 simple building blocks called amino acids\footnote{The term amino acid encompasses any compound that contains an amino and carboxyl group.
+However, proteins are only synthesized from the twenty ``canonical'' amino acids.
+Another two (selenocysteine and pyrrolysine) are incorporated via a distinct mechanism under rare circumstances and are therefore considered non-standard.}.
 Though each amino acid is chemically unique, they share a common backbone composed of two distinct and complementary receptor and donor sites for chemical bonds.
@@
 
-FOOTNOTE: The term amino acid encompasses any compound that contains an amino and carboxyl group.
-However, proteins are only synthesized from the twenty ``canonical'' amino acids.
-Another two (selenocysteine and pyrrolysine) are incorporated via a distinct mechanism under rare circumstances and are therefore considered non-standard.
-
 Because sequence dictates structure which in turn dictates function, understanding these relationships in proteins is essential for understanding biological systems as a whole.
-While biologists are generally interested in function since it is the most directly associated with the biological role of a protein, function and biological role are not always easily identified or measured.
+While biologists are generally interested in function since it is the most directly associated with the biological role\footnote{Function and biological role are frequently, though not always, exact synonyms since biological role is often defined more broadly and loosely than function.
+For example, the strict function of hemoglobin is to bind oxygen, an activity it shares with a related protein myoglobin.
+However, the two have different roles.
+Hemoglobin is found in red blood cells where it primarily acts a carrier.
+In contrast, myoglobin is found in muscle cells, where it stores oxygen until needed.} of a protein, function and biological role are not always easily identified or measured.
 Thus, determining a protein's structure is frequently the first step of detailed study of a protein's function.
@@
 
-FOOTNOTE: Function and biological role are frequently, though not always, exact synonyms since biological role is often defined more broadly and loosely than function.
-For example, the strict function of hemoglobin is to bind oxygen, an activity it shares with a related protein myoglobin.
-However, the two have different roles.
-Hemoglobin is found in red blood cells where it primarily acts a carrier.
-In contrast, myoglobin is found in muscle cells, where it stores oxygen until needed.
-
 The success of structural methods through the latter half of the 20th century established the view that while proteins were not completely rigid and could adopt a variety of related structures or ``conformations,'' most functional proteins had a single dominant structure.

### 2022-12-09 chapter2/chapter2.tex
Added templates for remaining chapters

\begin{abstract}
\noindent
Abstract text for this chapter
\end{abstract}

\section*{Introduction}
Here's where my chapter on evolutionary analyses will go!

\section*{Results}

\section*{Discussion}

\section*{Materials and methods}

### 2022-12-09 introduction/introduction.tex
Converted protein outline into text and tweaked a few phrases elsewhere

@@
 Life is a physical phenomenon.
-Despite the complexity of living things, their processes are governed by the same physical laws that describe how the planets revolve around the sun and the probability that an atom of uranium decays from one moment to another.
+Despite the complexity of living things, their processes are governed by the same physical laws that describe planets' motion around the sun and the decay of uranium atoms into lighter elements.
 However, as many systems are too complex to describe with physical models and equations, scientists simplify them into levels of abstraction that are more useful.
@@
 Some examples will illustrate the central role of proteins more clearly.
-These examples should be accessible to a wide audience, so nothing too abstract. Ideally should relate to human physiology and disease.
-    1. Hemoglobin and sickle cell
-        A. Introduce red blood cells transport oxygen
-        B. Hemoglobin is protein that binds oxygen.
-            i. Physically interacts with a molecule of oxygen
-        C. Sickle cell is disease caused by mutation in hemoglobin protein.
-            i. Cells produce hemoglobin with small mistake which causes them to clump into chains
-            ii. This distorts the shape of red blood cells into a sickle
-            iii. Red blood cells do not flow as easily through blood vessels causing pain and oxygen deprivation
-    2. Photopsins and color blindness
-        A. Photopsins are light sensitive proteins in cells in the retina at the back of the eye
-        B. Three different versions are sensitive to red, green, and blue light
-        C. Color blindness is the result of cells missing one of these proteins
-    3. Lactase and lactose intolerance
-        A. Lactose is a sugar found in milk that requires specific protein, lactase, to digest
-        B. Many humans who can digest milk products in childhood, lose this ability in adulthood suffer symptoms such as bloating and diarrhea after consuming dairy products
-        C. They stop producing lactase, so the undigested sugars are broken down by bacteria in the colon
+Blood is a part of the circulatory system which is responsible for transporting nutrients and waste.
+Though blood is a complex mixture, containing a cocktail of cells, proteins, sugars, gases, and ions dissolved in a medium of water, its primary cellular component is red blood cells.
+These cells, which give blood its red color, ferry oxygen from lungs throughout the body.
+While water can dissolve some oxygen, the body requires more oxygen more quickly than is available in the aqueous component of blood alone.
+Thus, red blood are packed with a special protein called hemoglobin, which binds\footnote{In biology, ``binding'' is a slippery word whose exact meaning can varying greatly depending on the context, but it generally means that a molecule physically interacts with another molecular for an extended period.} oxygen.
+Each red blood cell contains as many as 270 million molecules of hemoglobin, each of which can carry up to four oxygen molecules.
+To put this value in perspective, if each molecule of hemoglobin were the size of a tennis ball, the hemoglobin molecules in a single red blood cell would cover a tennis court in over 10 packed layers\footnote{A standard tennis ball is 1.35 in or 0.1125 ft in radius, so the volume of 270 million tennis balls of this size packed as efficiently as possible is given by $2.70 \times 10^6 \frac{4}{3} \pi (0.1125)^3 \frac{3\sqrt{2}}{\pi} = 2.17 \times 10^4 \text{ft^2}$.
+A singles tennis court is 78 ft by 27 ft or 2106 ft\textsuperscript{2}, so dividing the first by the second yields slightly over 10 layers.}.
+Because red blood cells are so dense with hemoglobin, any defect in hemoglobin can dramatically impact the structure of the red blood cells themselves.
+A well-studied example is sickle cell disease where an error in the body's hemoglobin molecules deforms red blood cells into a characteristic sickle shape.
+This prevents them from easily flowing through blood vessels, resulting in pain and oxygen deprivation.
+
+Whereas hemoglobin is an example of a protein mediating transport, proteins are also involved in transmitting signals and catalyzing chemical reactions.
+For example, the back of the eye contains a light-sensitive surface called the retina which is composed of photoreceptor cells.
+These cells respond to light because they produce special proteins called opsins that translate light into chemical and electrical signals which are then interpreted by the brain.
+Humans, and primates broadly, have three types of opsins that mediate our color vision, which are sensitive to red, green, and blue light, respectively.
+Color blindness is the result of cells missing one of these proteins, typically either the red or green opsin.
+In contrast to sickle cell disease, here the condition is caused by a missing rather than a mutated protein.
+However, in other cases, a protein is not missing or mutated, but instead not produced at the right time and place.
+For example, lactose is a sugar found in milk that requires a secific protein, lactase, to metabolize properly.
+Many humans who can digest milk products in childhood, lose this ability in adulthood because they stop producing lactose.
+As a result, lactose in dairy products passes undigested into the colon where it is broken down by bacteria, causing symptoms such as bloating and diarrhea.
 
@@
 This allows hemoglobin to carry oxygen throughout the body until it is delivered to its destination.
-However, people affected by sickle cell disease have a defect, i.e., a mutation, in the sequence of their hemoglobin proteins which causes it to malfunction.
+However, people affected by sickle cell disease have a mutation in the sequence of their hemoglobin proteins which causes it to malfunction.
 Frequently this mutation is a single change where the sixth amino acid in the sequence, glutamate, is substituted for a valine.
@@
 Some proteins are only produced in small amounts or degrade easily, so each step may require substantial optimization to achieve a sufficient yield.
-When structural techniques were first developed in the late 1950s, they were so time-consuming that a graduate student could dedicate an entire PhDs to solving a single protein structure.
+When structural techniques were first developed in the late 1950s, they were so time-consuming that a graduate student could dedicate an entire PhD to solving a single protein structure.
 Many developments have substantially accelerated the process, but it remains a labor-intensive technique which may require several months of effort.

### 2022-12-09 introduction/introduction.tex
Made outline into comments

@@
-======= OUTLINE =======
-
-BACKGROUND
-Survery of historical developments that inform dissertation
-Write at a high school or early college biology level
-
-I. Proteins are the molecular machines that drive biology
-    A. Four major macromolecules and briefly discuss broad roles of each
-        1. Nucleic acids
-        2. Carbohydrates
-        3. Lipids
-        4. Protein
-    B. Some major examples of proteins
-        1. Hemoglobin and sickle cell
-        2. Color blindness and opsins
-        3. Lactase and lactose intolerance
-II. Proteins are linear chains of amino acids with a three-dimensional structure
-    A. Background on biochemistry, primary, secondary, and tertiary structure
-    B. Structure-function relationships returning to previous hemoglobin example
-    C. Structural biology seeks to determine protein structures to understand function
-    D. Overview of structural methods including purification steps
-        1. X-ray crystallography
-        2. NMR
-        3. cryo-EM
-III. Disordered proteins do not adopt a stable structure
-    A. Definition and discussion of protein disorder
-        1. Structural heterogeneity, conformational ensembles, floppy
-        2. Major sequence characteristics and structural "modes"?
-    B. Disordered proteins are challenging to study with traditional methods
-    C. Genome sequencing and computational tools revealed disordered proteins are ubiquitous
-IV. Disordered proteins perform varied functions
-    A. Flexibility confers several advantages
-    A. Survey of disordered proteins and their functions
-        1. Linkers
-        2. Signaling
-V. Disordered proteins are enriched in transcriptional machinery
-    A. Introduce transcription via the central dogma
-    B. Describe transcription factors, cofactors, transcriptional machinery
-        1. Examples of different transcription factors
-    C. Describe properties of transcription factors
-        1. DNA binding and activation domains
-        2. Different flavors of activation domains
-    D. How transcription factors find their targets to initiate specific genetic programs is one of the major outstanding questions in molecular biology
-VI. Disordered proteins are associated with phase separation
-    A. Describe phase separation in general
-    B. Describe phase separation as related to proteins
-    C. Describe functions of phase separation and relationship to subcellular structures
-VII. Phase separation via IDRs is a possible mechanism for initiating transcription
-    A. Summarize Stadler's papers and others
-
-AIMS
-Discuss more recent developments in field and propose driving questions
-
-I. Structure-function relationships of disordered proteins are unclear
-    A. Examples of studies that have found specific cases of conservation in IDRs
-    B. Alan Moses paper as reference in yeast
-II. Use evolutionary analyses to identify conserved properties of disordered regions
-    A. Conservation is powerful signal of function
-        1. Examples of other studies that applied evolution
-    B. Outline overall approach and make correspondence to chapters
-        1. Identification of orthologous groups
-        2. Application of evolutionary analyses
-
-======= DRAFT =======
-
+%======= OUTLINE =======
+%
+%BACKGROUND
+%Survery of historical developments that inform dissertation
+%Write at a high school or early college biology level
+%
+%I. Proteins are the molecular machines that drive biology
+%    A. Four major macromolecules and briefly discuss broad roles of each
+%        1. Nucleic acids
+%        2. Carbohydrates
+%        3. Lipids
+%        4. Protein
+%    B. Some major examples of proteins
+%        1. Hemoglobin and sickle cell
+%        2. Color blindness and opsins
+%        3. Lactase and lactose intolerance
+%II. Proteins are linear chains of amino acids with a three-dimensional structure
+%    A. Background on biochemistry, primary, secondary, and tertiary structure
+%    B. Structure-function relationships returning to previous hemoglobin example
+%    C. Structural biology seeks to determine protein structures to understand function
+%    D. Overview of structural methods including purification steps
+%        1. X-ray crystallography
+%        2. NMR
+%        3. cryo-EM
+%III. Disordered proteins do not adopt a stable structure
+%    A. Definition and discussion of protein disorder
+%        1. Structural heterogeneity, conformational ensembles, floppy
+%        2. Major sequence characteristics and structural "modes"?
+%    B. Disordered proteins are challenging to study with traditional methods
+%    C. Genome sequencing and computational tools revealed disordered proteins are ubiquitous
+%IV. Disordered proteins perform varied functions
+%    A. Flexibility confers several advantages
+%    A. Survey of disordered proteins and their functions
+%        1. Linkers
+%        2. Signaling
+%V. Disordered proteins are enriched in transcriptional machinery
+%    A. Introduce transcription via the central dogma
+%    B. Describe transcription factors, cofactors, transcriptional machinery
+%        1. Examples of different transcription factors
+%    C. Describe properties of transcription factors
+%        1. DNA binding and activation domains
+%        2. Different flavors of activation domains
+%    D. How transcription factors find their targets to initiate specific genetic programs is one of the major outstanding questions in molecular biology
+%VI. Disordered proteins are associated with phase separation
+%    A. Describe phase separation in general
+%    B. Describe phase separation as related to proteins
+%    C. Describe functions of phase separation and relationship to subcellular structures
+%VII. Phase separation via IDRs is a possible mechanism for initiating transcription
+%    A. Summarize Stadler's papers and others
+%
+%AIMS
+%Discuss more recent developments in field and propose driving questions
+%
+%I. Structure-function relationships of disordered proteins are unclear
+%    A. Examples of studies that have found specific cases of conservation in IDRs
+%    B. Alan Moses paper as reference in yeast
+%II. Use evolutionary analyses to identify conserved properties of disordered regions
+%    A. Conservation is powerful signal of function
+%        1. Examples of other studies that applied evolution
+%    B. Outline overall approach and make correspondence to chapters
+%        1. Identification of orthologous groups
+%        2. Application of evolutionary analyses
+%
+%======= DRAFT =======
 Life is a physical phenomenon.

### 2022-12-09 introduction/introduction.tex
Tweaked some sentences in the introduction.tex

@@
 Despite the complexity of living things, their processes are governed by the same physical laws that describe planets' motion around the sun and the decay of uranium atoms into lighter elements.
-However, as many systems are too complex to describe with physical models and equations, scientists simplify them into levels of abstraction that are more useful.
+However, many systems are too complex to describe with physical models and equations, so scientists simplify them into levels of abstraction that are more useful.
 (This is the origin of the common observation that biology is applied chemistry and chemistry is applied physics.) For example, Punnett squares facilitate the prediction of genotypes and phenotypes by distilling the complexities and nuances of diverse reproductive systems into a set of simple rules.
@@
 Each of the four has a characteristic role.
-Nucleic acids, i.e., DNA and RNA, are responsible for information storage and transfer.
+Nucleic acids, \textit{i.e.}, DNA and RNA, are responsible for information storage and transfer.
 Carbohydrates primarily store energy but can also act as structural components of cells.
@@
 While water can dissolve some oxygen, the body requires more oxygen more quickly than is available in the aqueous component of blood alone.
-Thus, red blood are packed with a special protein called hemoglobin, which binds\footnote{In biology, ``binding'' is a slippery word whose exact meaning can varying greatly depending on the context, but it generally means that a molecule physically interacts with another molecular for an extended period.} oxygen.
+Thus, red blood are packed with a special protein called hemoglobin, which binds\footnote{In biology, ``binding'' is a slippery word whose exact meaning can varying greatly depending on the context.
+However, it generally means that a molecule physically interacts with another molecular for an extended period.} oxygen.
 Each red blood cell contains as many as 270 million molecules of hemoglobin, each of which can carry up to four oxygen molecules.
-To put this value in perspective, if each molecule of hemoglobin were the size of a tennis ball, the hemoglobin molecules in a single red blood cell would cover a tennis court in over 10 packed layers\footnote{A standard tennis ball is 1.35 in or 0.1125 ft in radius, so the volume of 270 million tennis balls of this size packed as efficiently as possible is given by $2.70 \times 10^6 \frac{4}{3} \pi (0.1125)^3 \frac{3\sqrt{2}}{\pi} = 2.17 \times 10^4 \text{ft^2}$.
+To put this value in perspective, if each molecule of hemoglobin were the size of a tennis ball, the hemoglobin molecules in a single red blood cell would cover a tennis court in over 10 packed layers\footnote{A standard tennis ball is 1.35 in or 0.1125 ft in radius, so the volume of 270 million tennis balls of this size packed as efficiently as possible is given by $2.70 \times 10^6 \times \frac{4}{3} \pi (0.1125 \text{ ft})^3 \times \frac{3\sqrt{2}}{\pi} = 2.17 \times 10^4 \text{ ft}^3$.
 A singles tennis court is 78 ft by 27 ft or 2106 ft\textsuperscript{2}, so dividing the first by the second yields slightly over 10 layers.}.
@@
 Humans, and primates broadly, have three types of opsins that mediate our color vision, which are sensitive to red, green, and blue light, respectively.
-Color blindness is the result of cells missing one of these proteins, typically either the red or green opsin.
-In contrast to sickle cell disease, here the condition is caused by a missing rather than a mutated protein.
+Color blindness is the result of photoreceptor cells missing one of these proteins, typically either the red or green opsin.
+In contrast to sickle cell disease, this condition is caused by a missing rather than a mutated protein.
 However, in other cases, a protein is not missing or mutated, but instead not produced at the right time and place.
-For example, lactose is a sugar found in milk that requires a secific protein, lactase, to metabolize properly.
+For example, lactose is a sugar found in milk that requires a specific protein, lactase, to metabolize properly.
 Many humans who can digest milk products in childhood, lose this ability in adulthood because they stop producing lactose.
@@
 Despite performing this diverse range of functions, all proteins are made from of a set of 20 simple building blocks called amino acids\footnote{The term amino acid encompasses any compound that contains an amino and carboxyl group.
-However, proteins are only synthesized from the twenty ``canonical'' amino acids.
+However, proteins are only synthesized from the 20 ``canonical'' amino acids.
 Another two (selenocysteine and pyrrolysine) are incorporated via a distinct mechanism under rare circumstances and are therefore considered non-standard.}.
@@
 
-Because sequence dictates structure which in turn dictates function, understanding these relationships in proteins is essential for understanding biological systems as a whole.
+Because sequence dictates structure and structure dictates function, understanding these relationships in proteins is essential for understanding biological systems as a whole.
 While biologists are generally interested in function since it is the most directly associated with the biological role\footnote{Function and biological role are frequently, though not always, exact synonyms since biological role is often defined more broadly and loosely than function.

### 2022-12-09 introduction/introduction.tex
Added cryo-EM as example of structural technique

@@
 Though in recent years researchers have developed powerful computational tools that can accurately predict structure from sequence alone, historically structures were determined experimentally, and experimental methods remain the gold standard today.
-While many methods can reveal information about the structure of a protein, the most powerful techniques, X-ray crystallography and NMR spectroscopy, can map the spatial coordinates of every atom in a protein.
+While many methods can reveal information about the structure of a protein, the most powerful techniques, X-ray crystallography, NMR spectroscopy, and cryogenic electron microscopy (cryo-EM), can map the spatial coordinates of every atom in a protein.
 However, this resolution requires extremely pure samples of the protein of interest.

### 2022-12-09 introduction/introduction.tex
Added discussion of purpose of IDR flexibility and relation to signaling pathways to outline
Tweaked wording about experimental methods remaining gold standard

@@
 %III. Disordered proteins do not adopt a stable structure
-%    A. Definition and discussion of protein disorder
+%    A. History of disordered proteins and initial recognition
+%        1. Success of structural methods
+%    B. Definition and discussion of protein disorder
 %        1. Structural heterogeneity, conformational ensembles, floppy
 %        2. Major sequence characteristics and structural "modes"?
-%    B. Disordered proteins are challenging to study with traditional methods
-%    C. Genome sequencing and computational tools revealed disordered proteins are ubiquitous
+%    C. Disordered proteins are challenging to study with traditional methods
+%    D. Genome sequencing and computational tools revealed disordered proteins are ubiquitous
 %IV. Disordered proteins perform varied functions
 %    A. Flexibility confers several advantages
-%    A. Survey of disordered proteins and their functions
+%        1. Structural flexibility is directly related to functional plasticity
+%        2. Proteins involved in signaling, transport, and catalysis require specific geometries to interact their intended targets efficiently
+%        3. Disordered proteins can interact with multiple targets
+%            i. Environmental signals can fine-tune which interacts are preferred
+%            ii. Proteins, structured and disordered, frequently undergo modifications which change their behavior
+%            iii. A common modification is phosphorylation where a phosphate group is attached to a specific amino acid in the protein
+%                a. Phosphate groups contain three negative charges, so the addition can dramatically affect a protein's folding energetics and induce a change in structure
+%            iv. Thus, phosphorylations commonly toggle proteins between an inactive and active state
+%                a. Proteins are phosphorylated by a class of proteins called kinases, which are in turn activated by interactions with or modifications by other proteins
+%                b. A single stimulus can set off a sequence of events, called a signaling cascade, spanning hundreds of reactions between as many distinct proteins
+%                c.
+Biologists are sometimes highly formulaic with their naming conventions, resulting in proteins like MAP kinase kinase kinase which phosphorylates MAP kinase kinase which phosphorylates MAP kinase which phosphorylates the MAP protein.
+[FOOTNOTE]
+%            v. Because IDRs expose much of their chain to the surrounding environment, they have many sites which are available for modification
+%                a. Thus, they often function as hubs in networks of signaling pathways [CITATION]
+%                b. Some IDRs are roughly like molecular computers which integrate signals from varying sources
+%    B. Survey of disordered proteins and their functions
 %        1. Linkers
@@
 Thus, determining a protein's structure is frequently the first step of detailed study of a protein's function.
-Though in recent years researchers have developed powerful computational tools that can accurately predict structure from sequence alone, historically structures were determined experimentally, and experimental methods remain the gold standard today.
+Though in recent years researchers have developed powerful computational tools that can accurately predict structure from sequence alone, historically structures were determined experimentally, and experimental methods still remain the gold standard.
 While many methods can reveal information about the structure of a protein, the most powerful techniques, X-ray crystallography, NMR spectroscopy, and cryogenic electron microscopy (cryo-EM), can map the spatial coordinates of every atom in a protein.

### 2022-12-14 introduction/introduction.tex
Added paragraph about early disorder examples and tweaked wording in a few other places

@@
 However, it generally means that a molecule physically interacts with another molecular for an extended period.} oxygen.
-Each red blood cell contains as many as 270 million molecules of hemoglobin, each of which can carry up to four oxygen molecules.
+Each red blood cell contains as many as 270 million molecules of hemoglobin, each of which can carry up to four oxygen molecules~\cite{Pierig2008}.
 To put this value in perspective, if each molecule of hemoglobin were the size of a tennis ball, the hemoglobin molecules in a single red blood cell would cover a tennis court in over 10 packed layers\footnote{A standard tennis ball is 1.35 in or 0.1125 ft in radius, so the volume of 270 million tennis balls of this size packed as efficiently as possible is given by $2.70 \times 10^6 \times \frac{4}{3} \pi (0.1125 \text{ ft})^3 \times \frac{3\sqrt{2}}{\pi} = 2.17 \times 10^4 \text{ ft}^3$.
@@
 
-Despite performing this diverse range of functions, all proteins are made from of a set of 20 simple building blocks called amino acids\footnote{The term amino acid encompasses any compound that contains an amino and carboxyl group.
+Despite performing this diverse range of functions, all proteins are made from of a set of 20 simple building blocks called amino acids.\footnote{The term amino acid encompasses any compound that contains an amino and carboxyl group.
 However, proteins are only synthesized from the 20 ``canonical'' amino acids.
-Another two (selenocysteine and pyrrolysine) are incorporated via a distinct mechanism under rare circumstances and are therefore considered non-standard.}.
-Though each amino acid is chemically unique, they share a common backbone composed of two distinct and complementary receptor and donor sites for chemical bonds.
+Another two (selenocysteine and pyrrolysine) are incorporated via a distinct mechanism under rare circumstances and are therefore considered non-standard.} Though each amino acid is chemically unique, they share a common backbone composed of two distinct and complementary receptor and donor sites for chemical bonds.
 Thus, in a protein the amino acids are bonded in a linear chain like beads on a string.
 However, once synthesized, proteins are not tidy rod-shaped molecules.
-Instead, the chain loops and weaves between itself creating a characteristic three-dimensional structure in a process called folding.
+Instead, the chain loops and weaves between itself creating a three-dimensional structure in a process called folding.
 These structures, which are highly stable and characteristic of each protein, are a result of the interactions between the amino acids in the chain and the surrounding medium, which is typically water.
@@
 
-Because sequence dictates structure and structure dictates function, understanding these relationships in proteins is essential for understanding biological systems as a whole.
-While biologists are generally interested in function since it is the most directly associated with the biological role\footnote{Function and biological role are frequently, though not always, exact synonyms since biological role is often defined more broadly and loosely than function.
+Clearly, understanding the relationship between the sequence, structure, and function of proteins is essential for unraveling more complex biological phenomena.
+While biologists are generally most interested in the function of a protein since it is the most directly associated with its biological role\footnote{Function and biological role are frequently, though not always, exact synonyms since biological role is often defined more broadly and loosely than function.
 For example, the strict function of hemoglobin is to bind oxygen, an activity it shares with a related protein myoglobin.
 However, the two have different roles.
-Hemoglobin is found in red blood cells where it primarily acts a carrier.
-In contrast, myoglobin is found in muscle cells, where it stores oxygen until needed.} of a protein, function and biological role are not always easily identified or measured.
-Thus, determining a protein's structure is frequently the first step of detailed study of a protein's function.
+Hemoglobin is found in red blood cells where it acts as a carrier during transport.
+In contrast, myoglobin is found in muscle cells, where it stores oxygen until needed.}, function and biological role are not always easily identified or measured.
+Thus, determining a protein's structure is frequently the first step of more detailed studies of its function.
 Though in recent years researchers have developed powerful computational tools that can accurately predict structure from sequence alone, historically structures were determined experimentally, and experimental methods still remain the gold standard.
@@
 
-The success of structural methods through the latter half of the 20th century established the view that while proteins were not completely rigid and could adopt a variety of related structures or ``conformations,'' most functional proteins had a single dominant structure.
-While some proteins had regions which returned undefined or highly variable atomic coordinates when analysed with structural techniques, these were viewed as exceptions.
-However, in the late 1990s the first genomes were sequenced, which allowed scientists to analyse the complete set of protein sequences encoded by an organism's DNA for the first time.
-
-When did scientists first notice the prevalence of IDRs?
-What tools did they use to detect it?
-In what organisms?
+The success of structural methods at elucidating the molecular details of protein function cemented the view that function largely depends on the presence of a fixed structure.
+While scientists understood proteins were not completely rigid and could adopt a variety of related structures or ``conformations,'' many believed that functional proteins largely had a single dominant structure.
+Despite its strength, exceptions to the structure-function paradigm were known.
+For example, elastin is a protein secreted by cells which allows tissues like skin or blood vessels to repeatedly expand and contract.
+It imparts this elasticity by forming networks of disordered chains that act like molecular springs.
+When a tissue experiences a force, the chains stretch to accommodate it.
+When the force is removed, the chains return to their random orientations, which reduces their end-to-end length and forces the tissue to contract~\cite{Vrhovski1998, Alberts2014}.
+However, as a result of this unique role in providing tissue structure, elastin's disorder was viewed as a specific adaptation rather than a general mode of protein function.
+In other cases, proteins had regions which returned undefined or highly variable atomic coordinates when analysed with structural techniques, but because these regions were often short loops between structured regions, they were seen as ``filler'' which facilitated the structure of the functional portions of the protein.
+By the late 1990s, though, enough exceptions had accumulated that scientists began to recognize that fully and partially disordered proteins were involved in many biological processes~\cite{Wright1999}.

### 2022-12-14 chapter1/chapter1_supporting.tex
Added note about truncated distribution in hmm_trim caption
Renamed hmm_training/ and hmm_trim/ to insertion_training/ and insertion_trim/

@@
 \begin{figure}[h!]
-\includegraphics[width=\textwidth]{hmm_training/out/hmm_training.png}
+\includegraphics[width=\textwidth]{insertion_training/out/insertion_training.png}
 \centering
@@
 \begin{figure}[h!]
-\includegraphics[width=\textwidth]{hmm_trim/out/hmm_trim.png}
+\includegraphics[width=\textwidth]{insertion_trim/out/insertion_trim.png}
 \centering
@@
 \textbf{(E, G)} The number of non-gap symbols in sequence trims can vary considerably, but for nearly all sequence trims each non-gap symbol in the removed segment is aligned to fewer than five non-gap symbols on average.
+Only the lower 95\% of the distribution of the number of non-gap symbols in the sequence trims is shown.
 \textbf{(F, H)} The length of region trims can also vary considerably, but generally each region trims accounts for fewer than 10\% of the columns in the original alignment.}

### 2022-12-14 chapter1/figures/insertion_training/insertion_training.py

@@ -1,4 +1,4 @@
-"""Make figure of HMM training."""
+"""Make figure of insertion HMM training."""
 
 import os
 import json
@@ -118,6 +118,6 @@ subfig.suptitle('F', x=0.025, y=0.975, fontweight='bold')
 handles = [Line2D([], [], label=label, color=color) for label, color in zip(state_labels, state_colors)]
 fig.legend(handles=handles, ncol=len(handles), bbox_to_anchor=(0.5, -0.01), loc='lower center')
 
-fig.savefig('out/hmm_training.png', dpi=300)
-fig.savefig('out/hmm_training.tiff', dpi=300)
+fig.savefig('out/insertion_training.png', dpi=300)
+fig.savefig('out/insertion_training.tiff', dpi=300)
 plt.close()

### 2022-12-14 chapter1/figures/insertion_trim/insertion_trim.py

@@ -1,4 +1,4 @@
-"""Make figure of HMM trimming data."""
+"""Make figure of insertion HMM trimming data."""
 
 import os
 
@@ -111,6 +111,6 @@ ax.set_xlabel('Length ratio of region trim')
 ax.set_ylabel('Number of region trims')
 subfig.suptitle('H', x=0.025, y=0.975, fontweight='bold')
 
-fig.savefig('out/hmm_trim.png', dpi=300)
-fig.savefig('out/hmm_trim.tiff', dpi=300)
+fig.savefig('out/insertion_trim.png', dpi=300)
+fig.savefig('out/insertion_trim.tiff', dpi=300)
 plt.close()

### 2022-12-14 chapter1/figures/insertion_training/insertion_training.py
Centered pie chart in insertion_training.py

@@ -49,7 +49,7 @@ for labels in OGid2labels.values():
 values = [counts[label] for label in state_labels]
 labels = [f'{label}\n{value:,}' for label, value in zip(state_labels, values)]
 subfig = fig.add_subfigure(gs[0, :2])
-ax = subfig.add_subplot()
+ax = subfig.add_axes((0.1, 0.1, 0.8, 0.8))
 ax.pie(values, colors=state_colors, labels=labels, labeldistance=1.3, textprops={'ha': 'center'})
 subfig.suptitle('A', x=0.025, y=0.975, fontweight='bold')
 

### 2022-12-14 chapter1/chapter1.tex
Added TODOs about updating trimming method in main text and methods
Added templates for remaining chapters
Updated figure labels to match file names

@@
-% Abstract must be 300 words or less
 \begin{abstract}
@@
 Overall, fewer than 10\% of alignments were trimmed of any regions.
-However, when regions were trimmed, they were largely inferred as state 3, meaning the removed regions are generally long poorly supported segments that are aligned to few if any other sequences (Fig.~\ref{sfig:hmm_trim}).
+However, when regions were trimmed, they were largely inferred as state 3, meaning the removed regions are generally long poorly supported segments that are aligned to few if any other sequences (Fig.~\ref{sfig:insertion_trim}).
+
+% TODO: Update text for new figure
 
@@
 All HMM algorithms were implemented with custom code which is available as the package Homomorph on PyPI.
-The insertion phylo-HMM was trained on 39,298 manually labeled columns in 11 alignments, and the deletion phylo-HMM was trained on 58,655 manually labeled positions in 20 sequences in 9 unique alignments (Fig.~\ref{sfig:hmm_training}).
+The insertion phylo-HMM was trained on 39,298 manually labeled columns in 11 alignments, and the deletion phylo-HMM was trained on 58,655 manually labeled positions in 20 sequences in 9 unique alignments (Fig.~\ref{sfig:insertion_training}).
 Because maximum-likelihood estimation of the model parameters yielded posterior decoding curves which toggled between hidden states too rapidly, the models were instead trained discriminatively~\cite{Krogh1999}.
@@
 By combining where the model’s confidence changed rapidly with the observed change in the gap pattern, this method generally selected reasonable boundaries.
+
+% TODO: Update training numbers, cutoffs, and add additional state 3 check
 

### 2022-12-14 chapter1/chapter1.tex
Added missing_training figure to supplement

@@
 All HMM algorithms were implemented with custom code which is available as the package Homomorph on PyPI.
-The insertion phylo-HMM was trained on 39,298 manually labeled columns in 11 alignments, and the deletion phylo-HMM was trained on 58,655 manually labeled positions in 20 sequences in 9 unique alignments (Fig.~\ref{sfig:insertion_training}).
+The insertion phylo-HMM was trained on 39,298 manually labeled columns in 11 alignments, and the deletion phylo-HMM was trained on 58,655 manually labeled positions in 20 sequences in 9 unique alignments (Fig.~\ref{sfig:insertion_training}, Fig.~\ref{sfig:missing_training}).
 Because maximum-likelihood estimation of the model parameters yielded posterior decoding curves which toggled between hidden states too rapidly, the models were instead trained discriminatively~\cite{Krogh1999}.

### 2022-12-14 chapter1/chapter1_supporting.tex
Updated figure labels to match file names
Added missing_training figure to supplement

@@
 Self transitions are excluded.}
-\label{sfig:hmm_training}
+\label{sfig:insertion_training}
 \end{figure}
@@
 \textbf{(F, H)} The length of region trims can also vary considerably, but generally each region trims accounts for fewer than 10\% of the columns in the original alignment.}
-\label{sfig:hmm_trim}
+\label{sfig:insertion_trim}
+\end{figure}
+
+\begin{figure}[h!]
+\includegraphics[width=\textwidth]{missing_training/out/missing_training.png}
+\centering
+\caption{\textbf{Missing phylo-HMM data and training details.}
+\textbf{(A)} Most columns in the training data were labeled as state 1, which is referred to as the ``not missing'' state in the main text.
+\textbf{(B)} The model loss stabilized by the final training iteration.
+\textbf{(C-F)} The values of parameters in the phylogenetic process, the jump process, and the transition matrix, respectively, at each training iteration.
+The transition matrix plots are the transition rates to the state indicated on the vertical axis and given in log scale.
+Self transitions are excluded.}
+\label{sfig:missing_training}
 \end{figure}

### 2022-12-14 chapter1/figures/missing_training/missing_training.py

@@ -1,0 +1,116 @@
+"""Make figure of missing HMM training."""
+
+import os
+import json
+
+import matplotlib.pyplot as plt
+from matplotlib.figure import SubplotParams
+from matplotlib.gridspec import GridSpec
+from matplotlib.lines import Line2D
+
+ppid_regex = r'ppid=([A-Za-z0-9_.]+)'
+spid_regex = r'spid=([a-z]+)'
+state_labels = ['1', '2']
+state_colors = ['C0', 'C1']
+
+# Load labels
+ids2labels = {}
+label_set = set()
+with open('../../orthology_inference/analysis/ortho_MSA/missing_hmm/labels.tsv') as file:
+    field_names = file.readline().rstrip('\n').split('\t')
+    for line in file:
+        if line.startswith('#'):
+            continue
+        fields = {key: value for key, value in zip(field_names, line.rstrip('\n').split('\t'))}
+        OGid, ppid, start, stop, label = fields['OGid'], fields['ppid'], int(fields['start']), int(fields['stop']), fields['label']
+        label_set.add(label)
+        try:
+            ids2labels[(OGid, ppid)].append((start, stop, label))
+        except KeyError:
+            ids2labels[(OGid, ppid)] = [(start, stop, label)]
+
+if set(state_labels) != label_set:
+    raise RuntimeError('state_labels is not equal to set of state_labels')
+
+# Load history and model parameters
+with open('../../orthology_inference/analysis/ortho_MSA/missing_hmm/out/history.json') as file:
+    history = json.load(file)
+with open('../../orthology_inference/analysis/ortho_MSA/missing_hmm/out/model.json') as file:
+    model_json = json.load(file)
+
+if not os.path.exists('out/'):
+    os.mkdir('out/')
+
+fig = plt.figure(figsize=(7.5, 5.5), subplotpars=SubplotParams(0.325, 0.225, 0.9, 0.9, 0, 0))
+gs = GridSpec(2, 6, height_ratios=[1, 1.5])
+
+# Plot state distribution
+counts = {label: 0 for label in state_labels}
+for labels in ids2labels.values():
+    for start, stop, label in labels:
+        counts[label] += stop - start
+values = [counts[label] for label in state_labels]
+labels = [f'{label}\n{value:,}' for label, value in zip(state_labels, values)]
+subfig = fig.add_subfigure(gs[0, :3])
[…]

### 2022-12-14 chapter1/chapter1.tex
Updated trimming method text with new training numbers and clearer descriptions

--- 2022-12-14 chapter1/chapter1.tex
@@
 All HMM algorithms were implemented with custom code which is available as the package Homomorph on PyPI.
-The insertion phylo-HMM was trained on 39,298 manually labeled columns in 11 alignments, and the deletion phylo-HMM was trained on 58,655 manually labeled positions in 20 sequences in 9 unique alignments (Fig.~\ref{sfig:insertion_training}).
+The insertion phylo-HMM was trained on 47,387 manually labeled columns in 14 alignments, and the deletion phylo-HMM was trained on 67,001 manually labeled positions in 23 sequences in 11 unique alignments (Fig.~\ref{sfig:insertion_training}, Fig.~\ref{sfig:missing_training}).
 Because maximum-likelihood estimation of the model parameters yielded posterior decoding curves which toggled between hidden states too rapidly, the models were instead trained discriminatively~\cite{Krogh1999}.
@@
 The posterior distributions over states were calculated for each alignment using the trained insertion phylo-HMM.
+Regions with a high probability of state 2 or 3 are candidates for trimming.
+However, because the probability of a state can change rapidly or gradually depending on the local context, a simple cutoff would not necessarily define the boundaries of these regions as the columns where the gap pattern changes most abruptly.
+Instead, the following algorithm was used.
+First, a high cutoff defines a ``seed'' region.
+The left and right endpoints of the seed are then expanded both inwards and outwards to define two intervals from which boundaries are selected.
+The outward expansion halts when the probability or its derivative is below two distinct thresholds, respectively.
+The inward expansion halts when the derivative is below a different threshold.
+The left and right boundaries are chosen as the columns in each interval with the maximum product between the derivative and the change in the gap profile between columns.
+The gap profile is calculated as the number of gaps in each column using Gaussian process sequence weighting as described in the section on selecting representative sequences.
+By combining where the model’s confidence changes rapidly with the observed change in the gap pattern, this method generally selects reasonable boundaries.
+
  States 2 and 3 have distinct characteristics which require different trimming strategies, so regions with a high probability of state 3 were handled first.
-Because the posterior probability of a state can change rapidly or gradually depending on the local context, a simple cutoff would not necessarily define the boundaries as the columns where the gap pattern changes most abruptly.
-Instead, a high cutoff of 0.95 was used to define a ``seed'' region.
-The left and right endpoints of the seed were then expanded both inwards and outwards to define two intervals from which boundaries were selected.
-The outward expansion halted when the probability or its derivative was below 0.01 and 0.001, respectively.
-The inward expansion halted when the derivative was below 0.001.
-The left and right boundaries were chosen as the columns in each interval with the maximum product between the derivative and the change in the gap profile between columns.
-The gap profile was calculated as the number of gaps in each column using Gaussian process sequence weighting as described in the section on selecting representative sequences.
-By combining where the model’s confidence changed rapidly with the observed change in the gap pattern, this method generally selected reasonable boundaries.
-
-% TODO: Update training numbers, cutoffs, and add additional state 3 check
-
 Because the long poorly supported segments in state 3 regions were sometimes aligned to highly conserved columns or short segments in other sequences, trimming columns entirely would remove these segments from the other sequences even if they do not qualify as long and poorly supported themselves.
 Thus, regions with high state 3 probabilities were trimmed at the level of individual sequences rather than entire columns with the following method.
-First, the mean number of non-gap symbols in a region was calculated using Gaussian process sequence weighting.
+First, the probability of state 3 of columns with a gap profile value less than or equal to 0.1 was set to 0 to break long poorly supported segments aligned to highly conserved columns into separate regions.
+Regions were then defined with the previously described algorithm using high and low cutoffs of 0.75 and 0.01, respectively.
+The outer and inner derivative cutoffs were both 0.001.
+The mean number of non-gap symbols in a region was calculated using Gaussian process sequence weighting.
 (The sequences with the five most non-gap symbols were also excluded to not bias the estimate with long poorly supported segments.) The final mean $\mu$ was taken as the minimum of this value and two.
@@
 This ensured that any regions which were intermediate between state 2 and 3 were included.
-Regions were defined from this combined probability using the algorithm described previously except with a high probability cutoff of 0.9 instead of 0.95.
+Regions were defined from this combined probability using the algorithm described previously except with a high probability cutoff of 0.9 instead of 0.75.
 The posterior probabilities from the missing data phylo-HMM were also converted into state assignments using a similar method.
-However, the initial seeds were defined with a cutoff of 0.95, and the seeds were expanded outward to first non-gap symbol or until the posterior probability of the ``missing data'' state was below 0.5.
+However, the initial seeds were defined with a cutoff of 0.75, and the seeds were expanded outward to first non-gap symbol or until the posterior probability of the ``missing data'' state was below 0.05.
 These assignments, which are available in the supplementary data, can be used to filter segments or entire sequences from downstream analyses.

### 2022-12-14 chapter1/figures/insertion_trim/insertion_trim.py
Changed naming of sequence trim to segment trim

@@ -42,7 +42,7 @@ OGids1 = set(df1['OGid']) - intersection
 OGids2 = set(df2['OGid']) - intersection
 
 values = [len(set(OGids)) - len(union), len(OGids1), len(OGids2), len(intersection)]
-labels = [f'{label}\n({value:,})' for label, value in zip(['no trims', 'sequence trims only', 'region trims only', 'sequence and region trims'], values)]
+labels = [f'{label}\n({value:,})' for label, value in zip(['no trims', 'segment trims only', 'region trims only', 'segment and region trims'], values)]
 colors = ['C0', 'C2', 'C1', 'C3']
 subfig = fig.add_subfigure(gs[0, 0], facecolor='none')
 ax = subfig.add_axes((0, 0.1, 0.45, 0.9))
@@ -59,12 +59,12 @@ ax.set_ylabel('Average state 3\nposterior in region trim')
 fig.colorbar(hb)
 subfig.suptitle('B', x=0.025, y=0.975, fontweight='bold')
 
-# Distribution of number of sequence trims in OGs
+# Distribution of number of segment trims in OGs
 counts = groups1.size().value_counts()
 subfig = fig.add_subfigure(gs[1, 0], facecolor='none')
 ax = subfig.add_subplot()
 ax.bar(counts.index, counts.values, width=1)
-ax.set_xlabel('Number of sequence trims in OG')
+ax.set_xlabel('Number of segment trims in OG')
 ax.set_ylabel('Number of OGs')
 subfig.suptitle('C', x=0.025, y=0.975, fontweight='bold')
 
@@ -77,14 +77,14 @@ ax.set_xlabel('Number of region trims in OG')
 ax.set_ylabel('Number of OGs')
 subfig.suptitle('D', x=0.025, y=0.975, fontweight='bold')
 
-# Distribution of number of removed symbols in sequence trims (truncated)
+# Distribution of number of removed symbols in segment trims (truncated)
 idx = int(np.ceil(len(df1) * 0.95))
 counts = df1['sym_count'].sort_values(ignore_index=True)[:idx].value_counts()
 subfig = fig.add_subfigure(gs[2, 0], facecolor='none')
 ax = subfig.add_subplot()
 ax.bar(counts.index, counts.values, width=1)
-ax.set_xlabel('Number of non-gap symbols in sequence trim')
-ax.set_ylabel('Number of\nsequence trims')
+ax.set_xlabel('Number of non-gap symbols in segment trim')
[…]

### 2022-12-14 chapter1/chapter1.tex
Updated text for description of insertion trimming results

@@
 In the example decoded alignment shown, the decoded states closely follow the expected patterns (Fig.~\ref{fig:hmm}B).
-Overall, fewer than 10\% of alignments were trimmed of any regions.
-However, when regions were trimmed, they were largely inferred as state 3, meaning the removed regions are generally long poorly supported segments that are aligned to few if any other sequences (Fig.~\ref{sfig:insertion_trim}).
-
-% TODO: Update text for new figure
+Overall, 29\% of alignments were trimmed of at least one segment or region.
+However, 87\% of the trims were segment trims, meaning the removed segments were largely inferred as state 3 and therefore were likely long poorly supported segments aligned to few if any other sequences (Fig.~\ref{sfig:insertion_trim}).
 

### 2022-12-16 introduction/introduction.tex
Expanded paragraph about recognition of disorder

@@
 %                c.
-Biologists are sometimes highly formulaic with their naming conventions, resulting in proteins like MAP kinase kinase kinase which phosphorylates MAP kinase kinase which phosphorylates MAP kinase which phosphorylates the MAP protein.
+Biologists are sometimes descriptive to a fault, resulting in proteins like MAP kinase kinase kinase which phosphorylates MAP kinase kinase which phosphorylates MAP kinase which phosphorylates the MAP protein.
 [FOOTNOTE]
@@
 When a tissue experiences a force, the chains stretch to accommodate it.
-When the force is removed, the chains return to their random orientations, which reduces their end-to-end length and forces the tissue to contract~\cite{Vrhovski1998, Alberts2014}.
-However, as a result of this unique role in providing tissue structure, elastin's disorder was viewed as a specific adaptation rather than a general mode of protein function.
-In other cases, proteins had regions which returned undefined or highly variable atomic coordinates when analysed with structural techniques, but because these regions were often short loops between structured regions, they were seen as ``filler'' which facilitated the structure of the functional portions of the protein.
-By the late 1990s, though, enough exceptions had accumulated that scientists began to recognize that fully and partially disordered proteins were involved in many biological processes~\cite{Wright1999}.
+When the force is removed, the chains return to their random orientations, which reduces their end-to-end length and forces the tissue to return to its original shape~\cite{Vrhovski1998, Alberts2014}.
+However, as a result of this unique role in providing tissue elasticity, elastin's disorder was viewed as a specific adaptation rather than a general mechanism of protein function.
+In other cases, proteins had regions which returned undefined or highly variable atomic coordinates when analysed with structural techniques, but because these regions were often short loops between structured regions, they were seen as ``linkers'' which facilitated the structure of the functional portions of the protein.
+By the early 2000s, though, enough exceptions had accumulated that scientists began to recognize that fully and partially disordered proteins were involved in many biological processes~\cite{Plaxco1997, Wright1999, Dunker2001}.
+Many of these examples were proteins which folded on binding to their targets, commonly other proteins.
+This mechanism was a departure from the prevailing model of interactions between biological macromolecules which required highly stable and complementary interfaces, like two puzzle pieces fitting together.
+As a result, scientists speculated that disorder was an adaptation that allowed proteins to efficiently relay and regulate signals by enabling interactions with many possible targets.
+Furthermore, the flexibility of disordered protein would permit environmental conditions to easily modulate these interactions.
+In the following years, as the complete genomes of several scientifically important model organisms such as \textit{S.
+cerevisiae} (baker's yeast), \textit{C.
+elegans} (roundworm), and \textit{D.
+melanogaster} (fruit fly) were sequenced for the first time, researchers applied computational methods for predicting disorder to the proteins inferred from their genomes.
+They discovered that disorder is ubiquitous in eukaryotic\footnote{All life belongs to one of three categories, or domains.
+Two, Archaea and Bacteria, are all single-celled organisms with simple cellular structures.
+In contrast, the cells of members of Eukarya, \textit{i.e.}, eukaryotes, are complex and contain substructures called organelles, among other differences.
+All animals, plants, and fungi are eukaryotes, though the domain includes many microorganisms as well.} organisms, with estimates of the fraction of proteins containing disordered segments of greater than 30 residues\footnote{The amino acids that compose the links of a protein chain are conventionally called residues to distinguish them from their related, but chemically distinct, free forms.} ranging between 28 and 63\%~\cite{Dunker2000, Ward2004}.
+For reference, although the lengths of proteins can vary dramatically, a ``typical'' protein contains on the order of a few hundred residues, so these regions can compose a significant fraction of a protein's length.
+Thus, disorder was recognized as a pervasive but poorly understood feature of proteins.
+
+Many structural and functional studies of protein disorder followed in the coming years.

### 2022-12-16 chapter1/figures/insertion_training/insertion_training.py
Added parentheses to numbers in pie chart labels

@@ -47,7 +47,7 @@ for labels in OGid2labels.values():
     for start, stop, label in labels:
         counts[label] += stop - start
 values = [counts[label] for label in state_labels]
-labels = [f'{label}\n{value:,}' for label, value in zip(state_labels, values)]
+labels = [f'{label}\n({value:,})' for label, value in zip(state_labels, values)]
 subfig = fig.add_subfigure(gs[0, :2])
 ax = subfig.add_axes((0.1, 0.1, 0.8, 0.8))
 ax.pie(values, colors=state_colors, labels=labels, labeldistance=1.3, textprops={'ha': 'center'})

### 2022-12-16 chapter1/figures/missing_training/missing_training.py

@@ -50,10 +50,10 @@ for labels in ids2labels.values():
     for start, stop, label in labels:
         counts[label] += stop - start
 values = [counts[label] for label in state_labels]
-labels = [f'{label}\n{value:,}' for label, value in zip(state_labels, values)]
+labels = [f'{label}\n({value:,})' for label, value in zip(state_labels, values)]
 subfig = fig.add_subfigure(gs[0, :3])
 ax = subfig.add_axes((0.1, 0.1, 0.8, 0.8))
-ax.pie(values, colors=state_colors, labels=labels, labeldistance=1.4, textprops={'ha': 'center'})
+ax.pie(values, colors=state_colors, labels=labels, labeldistance=1.45, textprops={'ha': 'center'})
 subfig.suptitle('A', x=0.025, y=0.975, fontweight='bold')
 
 # Plot loss curve

### 2022-12-16 chapter1/chapter1.tex
Updated number of alignments in chapter 1 abstract

@@
 In making the latter, we also wrote a fully documented Python package Homomorph that implements standard HMM algorithms and created a set of tutorials to promote its use by a wide audience.
-We applied the resulting pipeline to a set of 33 annotated \textit{Drosophila} genomes, generating 22,813 orthologous groups and 8,565 high-quality alignments.
+We applied the resulting pipeline to a set of 33 annotated \textit{Drosophila} genomes, generating 22,813 orthologous groups and 8,566 high-quality alignments.
 \end{abstract}

### 2022-12-16 chapter1/figures/missing_trim/missing_trim.py
Added missing_trim/ figure

@@ -1,0 +1,68 @@
+"""Make figure of missing HMM trimming data."""
+
+import os
+
+import matplotlib.pyplot as plt
+import pandas as pd
+from matplotlib.figure import SubplotParams
+from matplotlib.gridspec import GridSpec
+
+# Load OGids
+OGids = []
+with open('../../orthology_inference/analysis/ortho_MSA/realign_fastas/out/errors.tsv') as file:
+    field_names = file.readline().rstrip('\n').split('\t')
+    for line in file:
+        fields = {key: value for key, value in zip(field_names, line.rstrip('\n').split('\t'))}
+        OGid, error_flag1, error_flag2 = fields['OGid'], fields['error_flag1'], fields['error_flag2']
+        if error_flag1 == 'False' and error_flag2 == 'False':
+            OGids.append(OGid)
+
+df = pd.read_table('../../orthology_inference/analysis/ortho_MSA/missing_trim/out/trim_stats.tsv')
+
+df['length'] = df['stop'] - df['start']
+df['length_ratio'] = df['length'] / df['colnum']
+ppid_groups = df.groupby(['OGid', 'ppid'])  # Use combined key in case ppid in multiple OGids
+OGid_groups = df.groupby(['OGid'])
+
+if not os.path.exists('out/'):
+    os.mkdir('out/')
+
+fig = plt.figure(figsize=(7.5, 4), subplotpars=SubplotParams(0.25, 0.325, 0.9, 0.925))
+gs = GridSpec(2, 2)
+
+# Number of OGs with trims
+values = [len(set(OGids)) - df['OGid'].nunique(), df['OGid'].nunique()]
+labels = [f'{label}\n({value:,})' for label, value in zip(['no missing trims', 'missing trims'], values)]
+subfig = fig.add_subfigure(gs[0, 0], facecolor='none')
+ax = subfig.add_axes((0.1, 0.1, 0.8, 0.8))
+ax.pie(values, labels=labels, labeldistance=1.75, textprops={'ha': 'center'})
+subfig.suptitle('A', x=0.025, y=0.975, fontweight='bold')
+
+# Distribution of number sequences with trims in OGs
+counts = OGid_groups['ppid'].nunique().value_counts()
+subfig = fig.add_subfigure(gs[0, 1], facecolor='none')
+ax = subfig.add_subplot()
+ax.bar(counts.index, counts.values, width=1)
[…]

### 2022-12-16 chapter1/chapter1.tex
Added missing_trim figure to supplement and reference in main text

@@
 These segments are easily identified in regions which are otherwise highly conserved, so they are also likely the result of incorrect or incomplete annotations.
-Unfortunately, they can also span more diverged regions, which complicates a simple rule- or threshold-based approach for identifying them.
+However, they can also span more diverged regions, which complicates a simple rule- or threshold-based approach for identifying them.
 Thus, another phylo-HMM was trained to label each position in a sequence as generated by either a ``missing'' or ``not missing'' state.
-In this case, however, the aligned sequences were processed individually and not as aligned columns.
+In this case, though, the aligned sequences were processed individually and not as aligned columns.
 As the previous phylo-HMM already ensured each column has sufficient support, these labels can instead be used to exclude sequences from downstream analyses depending on the amount of tolerated overlap with the regions of interest.
+Overall, 15\% of alignments have at least one sequence with a segment of ``missing'' data (Fig.~\ref{sfig:missing_trim}).
 

### 2022-12-16 chapter1/chapter1_supporting.tex
Added missing_trim figure to supplement and reference in main text

@@
 \label{sfig:missing_training}
+\end{figure}
+
+\begin{figure}[h!]
+\includegraphics[width=\textwidth]{missing_trim/out/missing_trim.png}
+\centering
+\caption{\textbf{Missing phylo-HMM trimming details.}
+\textbf{(A)} Most alignments have no sequences with ``missing'' segments.
+\textbf{(B)} Of the alignments with sequences trimmed of missing segments, a majority have only one trimmed sequence.
+\textbf{(C-D)} The length of missing segments can vary considerably, both in terms of the number of positions as well as its ratio to the number of columns in the alignment.}
+\label{sfig:missing_trim}
 \end{figure}

### 2022-12-16 chapter1/chapter1.tex
Fixed comma outside of quotation marks

@@
 However, many analyses impose the additional condition that the sequences have diverged through speciation events (orthology) rather than duplications (paralogy) or other mechanisms such as horizontal gene transfer~\cite{Fitch1970}.
-The underlying assumption is ``orthologs'' have conserved equivalent functions whereas ``paralogs'', by virtue of their redundancy, are more likely to diverge~\cite{Ohno1970, Nowak1997, Altenhoff2012, Pegueroles2013, Soria2014}.\footnote{For multiple sequences, orthology is usually defined relative to their most recent common ancestor.
+The underlying assumption is ``orthologs'' have conserved equivalent functions whereas ``paralogs,'' by virtue of their redundancy, are more likely to diverge~\cite{Ohno1970, Nowak1997, Altenhoff2012, Pegueroles2013, Soria2014}.\footnote{For multiple sequences, orthology is usually defined relative to their most recent common ancestor.
 This technically includes sequences which split by duplication after this point (``in-paralogs''), but excludes sequences which split by duplication before (``out-paralogs'') \cite{Remm2001}.

### 2022-12-20 introduction/introduction.tex
Added to outline of protein sequence-structure relationships and refined wording on "biological role"

@@
 %        1. Success of structural methods
-%    B. Definition and discussion of protein disorder
-%        1. Structural heterogeneity, conformational ensembles, floppy
-%        2. Major sequence characteristics and structural "modes"?
-%    C. Disordered proteins are challenging to study with traditional methods
-%    D. Genome sequencing and computational tools revealed disordered proteins are ubiquitous
-%IV. Disordered proteins perform varied functions
+%    B. Disordered proteins are challenging to study with traditional methods
+%    C. Genome sequencing and computational tools revealed disordered proteins are ubiquitous
+%IV. Sequence-structure relationships in protein disorder
+%    A. Prompted by these observations, researchers began to clarify the sequence-structure-function relationships that characterize protein disorder
+%        1. They discovered that related to structure and sequence but the details differ
+%    B. Though structured proteins are not completely rigid, they generally have a small number of structures
+%        1. As mentioned previously, even before disorder was recognized as a common feature of proteins, scientists understood proteins were not completely rigid
+%        2. At the molecular level, everything is in constant motion. For example, at room temperature the average water molecule moves at a speed of SPEED.
+%            i. However, collisions and interactions with other water molecules generally prevent them from escaping into the atmosphere
+%        3. Proteins are also in motion, but they are constrained both by the rigid bonds and interactions between amino acid residues in the chain
+%        4. Thus, all proteins flex and breathe. However, these motions are minor variations on the stable structure.
+%        5. In other cases, proteins have multiple structures which are related to different functional states.
+%            i. For example, hemoglobin has two forms, traditionally called T and R.
+%            ii. The T form is hemoglobin's oxygen-free state.
+%            iii. On binding oxygen, its structure changes slightly, with the shift on the scale of a few hydrogen atoms at the points of greatest change.
+%            iv. This changes the orientation between the atoms in hemoglobin that interact with oxygen, allowing it to bind more tightly.
+%            v. Promotes oxygen uptake at other three binding sites, promoting efficiently transport
+%        6. Although proteins have multiple conformations, they are generally stable and the intermediates are not.
+%        7. Thus, proteins with multiple conformations generally exist in discrete states.
+%    C. Protein disorder means their atoms do not have fixed spatial relationships
+%       1. In contrast, disordered proteins are described as ``conformational ensembles''
+%       2. They are constantly sampling conformations which may have large differences in the spatial relationships between residues
+%    D. Protein folding driven by hydrophobic interactions
+%       1. Disordered regions largely characterized by enrichment and depletion in hydrophilic and hydrophobic residues, respectively
+%       2. Hydrophobic residues are ``water-fearing,'' and like oil in water, they tend to self-associate.
+%       3. Hydrophilic residues are ``water-loving,'' and easily interact with water.
+%       4. Hydrophobic interacts are a major force driving protein folding, and many folded proteins have a hydrophobic core
+%    E. Sequence-structure relationships in disordered proteins
+%       1. Although, disordered regions are conformationally heterogeneous, they
+%V. Functional roles of protein disorder
 %    A. Flexibility confers several advantages
@@
 %        2. Signaling
-%V. Disordered proteins are enriched in transcriptional machinery
+%VI. Disordered proteins are enriched in transcriptional machinery
 %    A. Introduce transcription via the central dogma
@@
 %    D. How transcription factors find their targets to initiate specific genetic programs is one of the major outstanding questions in molecular biology
-%VI. Disordered proteins are associated with phase separation
+%VII. Disordered proteins are associated with phase separation
 %    A. Describe phase separation in general
@@
 %    C. Describe functions of phase separation and relationship to subcellular structures
-%VII. Phase separation via IDRs is a possible mechanism for initiating transcription
+%VIII. Phase separation via IDRs is a possible mechanism for initiating transcription
 %    A. Summarize Stadler's papers and others
@@
 %AIMS
-%Discuss more recent developments in field and propose driving questions
-%
 %I. Structure-function relationships of disordered proteins are unclear
@@
 Clearly, understanding the relationship between the sequence, structure, and function of proteins is essential for unraveling more complex biological phenomena.
-While biologists are generally most interested in the function of a protein since it is the most directly associated with its biological role\footnote{Function and biological role are frequently, though not always, exact synonyms since biological role is often defined more broadly and loosely than function.
-For example, the strict function of hemoglobin is to bind oxygen, an activity it shares with a related protein myoglobin.
-However, the two have different roles.
+Though biologists study all three properties of proteins, they are generally most interested in function since it is the most directly related to the biological processes the protein takes part in\footnote{Function generally refers to \textit{molecular} function which is a description of a specific chemical activity possessed by a protein.
+A biological process, however, is the larger ``biological program'' which is accomplished by the action of multiple linked molecular functions.
+For example, the molecular function of hemoglobin is to bind oxygen, an activity it shares with a related protein myoglobin.
+However, the two have different roles in the process of oxygen transport and storage.
 Hemoglobin is found in red blood cells where it acts as a carrier during transport.
-In contrast, myoglobin is found in muscle cells, where it stores oxygen until needed.}, function and biological role are not always easily identified or measured.
-Thus, determining a protein's structure is frequently the first step of more detailed studies of its function.
+In contrast, myoglobin is found in muscle cells, where it stores oxygen until needed.}, function and biological process are not always easily identified or measured.
+Thus, determining a protein's structure is frequently the first step of detailed studies of its function.
 Though in recent years researchers have developed powerful computational tools that can accurately predict structure from sequence alone, historically structures were determined experimentally, and experimental methods still remain the gold standard.
@@
 The success of structural methods at elucidating the molecular details of protein function cemented the view that function largely depends on the presence of a fixed structure.
-While scientists understood proteins were not completely rigid and could adopt a variety of related structures or ``conformations,'' many believed that functional proteins largely had a single dominant structure.
+While scientists understood proteins were not completely rigid and could adopt a variety of related structures, many believed that functional proteins largely had a single dominant structure.
 Despite its strength, exceptions to the structure-function paradigm were known.

### 2023-01-24 chapter1/figures/trees/trees.py
Updated trees.py for updates to plot_tree

@@ -26,8 +26,8 @@ dpi = 300
 
 fig, axs = plt.subplots(1, 2, figsize=(7.5, 4), layout='tight')
 for ax, tree, xlabel, title_label in zip(axs, trees, xlabels, title_labels):
-    plot_tree(tree, ax, tip_fontsize=8, support_labels=True,
-              support_format_spec='.2f', support_fontsize=8.5,
+    plot_tree(tree, ax=ax, tip_fontsize=8,
+              support_labels=True, support_format_spec='.2f', support_fontsize=8.5,
               support_ha='right', support_hoffset=-0.006, support_voffset=-0.006)
     ax.yaxis.set_visible(False)
     ax.spines['left'].set_visible(False)
@@ -59,8 +59,8 @@ for model_label, xlabel in zip(model_labels, xlabels):
 
     fig, axs = plt.subplots(len(ylabels), len(title_labels), figsize=(7.5, 6), layout='constrained')
     for ax, tree in zip(axs.ravel(), trees):
-        plot_tree(tree, ax, tip_fontsize=6, tip_offset=0.002, support_labels=True,
-                  support_format_spec='.2f', support_fontsize=6,
+        plot_tree(tree, ax=ax, linewidth=0.6, tip_fontsize=6,
+                  support_labels=True, support_format_spec='.2f', support_fontsize=6,
                   support_ha='right', support_hoffset=-0.007, support_voffset=-0.007)
         ax.set_yticks([])
         ax.spines['left'].set_visible(False)

### 2023-01-24 chapter1/figures/hmm/hmm.py
Updated hmm.py to match updates to draw.py and phylo.py in src/

@@ -19,10 +19,10 @@ spid_regex = r'spid=([a-z]+)'
 state_labels = ['1A', '1B', '2', '3']
 state_colors = ['C0', 'C3', 'C1', 'C2']
 
-adjust_left = 0.025
-adjust_bottom = 0.06
+adjust_left = 0.1
+adjust_bottom = 0.05
 adjust_right = 0.905
-adjust_top = 0.9
+adjust_top = 0.925
 
 height_ratio = 0.5
 hspace = 0.3
@@ -34,7 +34,7 @@ legend_markerscale = 1
 
 fig_width = 7.5
 fig_height = 3
-dpi = 300
+dpi = 400
 
 panel_label = 'B'
 panel_label_fontsize = 'large'
@@ -70,6 +70,13 @@ for OGid in OGids:
 
     # Load tree and convert to vectors at tips
     tree = tree_template.shear([record['spid'] for record in msa])
+    for node in tree.postorder():  # Ensure tree is ordered as in original
+        if node.is_tip():
+            node.value = tip_order[node.name]
+        else:
+            node.children = sorted(node.children, key=lambda x: x.value)
+            node.value = sum([child.value for child in node.children])
+
     tips = {tip.name: tip for tip in tree.tips()}
     for record in msa:
         spid, seq = record['spid'], record['seq']
@@ -97,8 +104,10 @@ for OGid in OGids:
     data = [fbs[label] for label in state_labels]
 
     fig = plot_msa_data([record['seq'] for record in msa], data,
+                        tree=tree, tree_kwargs={'linewidth': 0.5, 'tip_labels': False, 'xmin_pad': 0.01, 'xmax_pad': 0.025},
+                        height_ratio=height_ratio, hspace=hspace,
                         figsize=(fig_width, fig_height),
-                        height_ratio=height_ratio, hspace=hspace, left=adjust_left, bottom=adjust_bottom, right=adjust_right, top=adjust_top,
+                        left=adjust_left, bottom=adjust_bottom, right=adjust_right, top=adjust_top, anchor=(0, 0.5),
                         data_max=1.05, data_min=-0.05, data_labels=state_labels, data_colors=state_colors,
                         x_labelsize=x_labelsize,
                         msa_legend=True, legend_kwargs={'bbox_to_anchor': legend_position, 'loc': 'center left', 'fontsize': legend_fontsize,

### 2023-01-26 chapter0/chapter0.tex
Renamed introduction/ to chapter0/

--- 2022-12-20 introduction/introduction.tex
@@
 %        3. cryo-EM
-%III. Disordered proteins do not adopt a stable structure
+%III. IDRs do not adopt a stable structure
 %    A. History of disordered proteins and initial recognition
@@
 %    C. Genome sequencing and computational tools revealed disordered proteins are ubiquitous
-%IV. Sequence-structure relationships in protein disorder
+%IV. IDRs have sequence-structure relationships (of a different kind)
 %    A. Prompted by these observations, researchers began to clarify the sequence-structure-function relationships that characterize protein disorder
@@
 %       1. Although, disordered regions are conformationally heterogeneous, they
-%V. Functional roles of protein disorder
+%V. IDRs are functional
 %    A. Flexibility confers several advantages
@@
 %        2. Signaling
-%VI. Disordered proteins are enriched in transcriptional machinery
+%VI. IDRs are enriched in transcriptional machinery
 %    A. Introduce transcription via the central dogma
@@
 %    D. How transcription factors find their targets to initiate specific genetic programs is one of the major outstanding questions in molecular biology
-%VII. Disordered proteins are associated with phase separation
+%VII. IDRs are associated with phase separation
 %    A. Describe phase separation in general
@@
 %    C. Describe functions of phase separation and relationship to subcellular structures
-%VIII. Phase separation via IDRs is a possible mechanism for initiating transcription
+%VIII. IDR-mediated phase separation is a possible mechanism for initiating transcription
 %    A. Summarize Stadler's papers and others
@@
 All animals, plants, and fungi are eukaryotes, though the domain includes many microorganisms as well.} organisms, with estimates of the fraction of proteins containing disordered segments of greater than 30 residues\footnote{The amino acids that compose the links of a protein chain are conventionally called residues to distinguish them from their related, but chemically distinct, free forms.} ranging between 28 and 63\%~\cite{Dunker2000, Ward2004}.
-For reference, although the lengths of proteins can vary dramatically, a ``typical'' protein contains on the order of a few hundred residues, so these regions can compose a significant fraction of a protein's length.
+For reference, though the lengths of proteins can vary dramatically, a ``typical'' protein contains on the order of a few hundred residues, so these regions can compose a significant fraction of a protein's length.
+Because these segments were disordered in their native state, \textit{i.e.} under normal operating conditions, such segments were termed ``intrinsically disordered regions'' (IDRs) to emphasize the disorder was not induced by exposure to chemicals or heat.
+Furthermore, while most proteins were predicted to contain a mixture of structured domains and IDRs, some ``intrinsically disordered proteins'' (IDPs) were entirely or almost entirely disordered.
 Thus, disorder was recognized as a pervasive but poorly understood feature of proteins.
 
-Many structural and functional studies of protein disorder followed in the coming years.
+Many studies investigated various structural and functional aspects of IDRs and IDPs followed in the following years.

### 2023-02-04 chapter0/chapter0.tex
Removed commas after i.e.

@@
 Each of the four has a characteristic role.
-Nucleic acids, \textit{i.e.}, DNA and RNA, are responsible for information storage and transfer.
+Nucleic acids, \textit{i.e.} DNA and RNA, are responsible for information storage and transfer.
 Carbohydrates primarily store energy but can also act as structural components of cells.
@@
 Two, Archaea and Bacteria, are all single-celled organisms with simple cellular structures.
-In contrast, the cells of members of Eukarya, \textit{i.e.}, eukaryotes, are complex and contain substructures called organelles, among other differences.
+In contrast, the cells of members of Eukarya, \textit{i.e.} eukaryotes, are complex and contain substructures called organelles, among other differences.
 All animals, plants, and fungi are eukaryotes, though the domain includes many microorganisms as well.} organisms, with estimates of the fraction of proteins containing disordered segments of greater than 30 residues\footnote{The amino acids that compose the links of a protein chain are conventionally called residues to distinguish them from their related, but chemically distinct, free forms.} ranging between 28 and 63\%~\cite{Dunker2000, Ward2004}.

### 2023-02-07 chapter3/chapter3.tex
Added templates for remaining chapters
Wrote introduction for chapter3

--- 2022-12-09 chapter2/chapter2.tex
@@
 \section*{Introduction}
-Here's where my chapter on evolutionary analyses will go!
+Most code written for data analysis is \textit{ad hoc}, that is, it is intended to accomplish a single task and cannot be used for any other purpose.
+However, often solving one problem entails solving several other problems along the way, and sometimes those intermediate solutions are general enough to be widely useful.
+This happened twice over the course of this work, both times involving fitting probabilistic models to data.
+The first was mixture models, which describe data that contain several subpopulations, each with its own statistical properties, mixed together.
+For example, in exam scores there are often two clusters: the high-scoring group of students who understood the material and the low-scoring group who did not.
+The second was hidden Markov models (HMMs), which like mixture models, describe data with several subpopulations.
+However, in HMMs the data are ordered in a ``time series'' and observations from a given subpopulation tend to occur consecutively.
+These models are frequently used to analyze systems which toggle between different states over time, for example periods of growth and recession in the economy.
 
-\section*{Results}
+Both types of models are well-known and frequently used.
+Despite this, none of the major scientific computing libraries for Python such as SciPy or statistical modules like scikit-learn or statsmodels contained generic implementations of either.
+A major factor in this exclusion is likely the high level of stability and consistency these packages guarantee.
+Mixture models and HMMs are not so much single models but rather recipes for creating models.
+As a result, implementing these models for arbitrary distributions is not trivial, so these packages often sacrifice generality to support training methods for common use cases.
+For example, scikit-learn implements mixture models where all subpopulations are restricted to normal distributions.
+Other more specialized packages like hmm-learn and pomegranate are more flexible, but they still limit their built-in support to a relatively small number of common distributions, including normal, exponential, categorical, and Poisson, among others.
+Furthermore, to support these additional features these packages implement custom APIs which may pose a barrier to users who are already comfortable with the interface of SciPy's statistical module.
 
-\section*{Discussion}
+We therefore sought to create lightweight implementations of mixture models and HMMs which are available in the packages MixMod and Homomorph, respectively.
+A key design goal was to ensure these packages were both immediately accessible but also easily extensible.
+Thus, both packages conform to the SciPy API for specifying distributions, which allows use of its large number of distributions but also permits advanced users to easily implement custom distributions as needed.
+Fitting models parameterized by arbitrary distributions to data remains a significant challenge, however.
+MixMod supports fitting models whose subpopulations are parameterized by roughly a dozen distinct distributions, including several common continuous and discrete distributions.
+Homomorph has no built-in methods for fitting models.
+We instead wrote an extensive tutorial that covers the theory and implementation of several model fitting approaches for specific distributions.
+This tutorial is the centerpiece of a series of tutorials for both MixMod which demonstrate their APIs and common applications through a series of examples.
+Our hope is together these packages and tutorials will bridge the gap between beginner and advanced users by illustrating how complex probabilistic models are constructed from simple pieces.
 
-\section*{Materials and methods}
+The tutorials are available alongside the source code for these packages and will likely continue to undergo revisions and expansions.
+They are, however, adapted from their current form in the following sections.
+Though the tutorials are largely independent of the others, here they are presented in their intended order.
+The first two are brief introductions to the MixMod and Homomorph APIs, respectively.
+The next discusses a generalization of HMMs called autoregressive HMMs (ARHMMs).
+The final tutorial covers methods for training HMMs.
+As each is a self-contained document with an introduction and conclusion, they are presented without further commentary.

### 2023-02-07 chapter3/chapter3.tex
Added abstract for chapter3

@@
 \noindent
-Abstract text for this chapter
+Fitting statistical models to data is often a key step in the scientific method because it can formalize hypotheses and conclusions as unambiguous and testable statements.
+The major scientific computing library for the programming language Python, SciPy, provides ready-to-use implementations of core statistical functions, which allows users with varying levels of expertise to easily apply them to their data.
+However, SciPy does not support two common and powerful types of models called mixture models and hidden Markov models (HMMs).
+Other more specialized packages such as hmm-learn and pomegranate provide implementations for a restricted subset of these models, but they use APIs which are not compatible with SciPy.
+This can pose a barrier to entry for beginners and prevent more advanced users from easily extending these packages' capabilities.
+We therefore created two packages, MixMod and Homomorph, that implement mixture models and HMMs, respectively, and conform to the SciPy API for specifying distributions.
+Each package is fully documented, and we wrote a set of tutorials which both introduce their APIs and illustrate various training techniques through a series of examples.
+These packages are available on the Python Package Index (PyPI) under the names mixmod and homomorph, and the source code is hosted alongside the tutorials on GitHub at \url{https://github.com/marcsingleton/mixmod} and \url{https://github.com/marcsingleton/homomorph}, respectively.
 \end{abstract}

### 2023-02-07 chapter1/chapter1.tex
Added SVG files to repository
Various minor re-wordings and corrections
Change list:
Made use of past tense consistent in methods
Removed unnecessary transitions like "also" and "then," especially in the methods

@@
 Two highlights include the application of the more flexible \textit{k}-clique percolation algorithm for identifying clusters of orthologous proteins and the development of a novel technique for removing poorly supported regions of alignments with a phylogenetic HMM.
-In making the latter, we also wrote a fully documented Python package Homomorph that implements standard HMM algorithms and created a set of tutorials to promote its use by a wide audience.
+In making the latter, we wrote a fully documented Python package Homomorph that implements standard HMM algorithms and created a set of tutorials to promote its use by a wide audience.
 We applied the resulting pipeline to a set of 33 annotated \textit{Drosophila} genomes, generating 22,813 orthologous groups and 8,566 high-quality alignments.
@@
 \begin{figure}[h!]
-\includegraphics[width=\textwidth]{graphical_overview.png}
+\includegraphics[width=\textwidth]{graphical_overview/out/graphical_overview.png}
 \centering
@@
 Our pipeline follows a similar overall approach to other graph-based methods of orthology inference.
-First, protein sequences from annotated genomes are collected (Fig.~\ref{fig:graphical_overview}A), and homology searches are then conducted between all query-target pairs of genomes (Fig.~\ref{fig:graphical_overview}B).
+First, protein sequences from annotated genomes are collected (Fig.~\ref{fig:graphical_overview}A), and homology searches are conducted between all query-target pairs of genomes (Fig.~\ref{fig:graphical_overview}B).
 The raw output from these homology searches is processed to yield best hits between pairs of sequences (Fig.~\ref{fig:graphical_overview}C).
@@
 Highly conserved regions were identified, which divided the alignment into a complementary set of diverged regions.
-The sequences in each of these regions were then extracted and aligned separately with a more conservative value for \textit{a\textsubscript{max}} of 0.4.
+The sequences in each of these regions were extracted and aligned separately with a more conservative value for \textit{a\textsubscript{max}} of 0.4.
 The resulting ``sub-alignments'' were ``stitched'' back into their positions in the original alignment.
@@
 
-Though there are benchmarks available for optimizing and comparing methods of orthology inference, the metrics are calculated over a set of reference genomes which are sparsely sampled over a broad taxonomic range, so it is unclear if they are informative for method designed to yield robust inferences when the genomes are highly related~\cite{Nevers2022}.
+Though benchmarks are available for optimizing and comparing methods of orthology inference, the metrics are calculated over a set of reference genomes which are sparsely sampled over a broad taxonomic range, so it is unclear if they are informative for method designed to yield robust inferences when the genomes are highly related~\cite{Nevers2022}.
 Furthermore, the heterogeneity of genome architectures and annotations may require quality assurance methods tailored to each set of genomes.
@@
 The protein sequences for each annotation were de-duplicated by removing any sequences which had already appeared in association with the same gene.
-Thus, the first accession associated with a sequence and gene pair is the sequence’s representative accession for the gene.
+Thus, the first accession associated with a sequence and gene pair was the sequence’s representative accession for the gene.
 BLAST+ 2.13.0 was used for the sequence similarity searches~\cite{Camacho2009}.
@@
 The HSPs were first grouped by target protein, and the resulting groups were sorted in descending order by the bit score of their highest-scoring HSP.
-Iterating over the groups, all the HSPs in a group were passed to the next step until the parent gene of the group was not the parent gene of the highest ranked group.
-This method collects all candidate HSPs for a target gene if the highest-scoring HSP within a group exceeds the highest-scoring HSP of the next best gene.
+Iterating over the groups, all HSPs in a group were passed to the next step until the parent gene of the group was not the parent gene of the highest ranked group.
+This method collected all candidate HSPs for a target gene if the highest-scoring HSP within a group exceeded the highest-scoring HSP of the next best gene.
 This is in some senses an extension of the best hit criterion where hits are considered at the level of genes rather than proteins.
@@
 HSPs were merged in two stages where the first combined non-overlapping HSPs, and the second combined the remaining HSPs.
-In the first stage, proceeding from highest to lowest bit score, HSPs were marked as ``disjoint'' if they do not overlap with any other HSP previously marked as disjoint.
-Although this greedy strategy does not necessarily yield the highest-scoring set of disjoint HSPs, it prioritized higher scoring HSPs.
+In the first stage, proceeding from highest to lowest bit score, HSPs were marked as ``disjoint'' if they did not overlap with any other HSP previously marked as disjoint.
+Although this greedy strategy did not necessarily yield the highest-scoring set of disjoint HSPs, it prioritized higher scoring HSPs.
 In the second stage, all disjoint HSPs were marked as ``compatible,'' and proceeding from highest to lowest bit score the remaining HSPs were marked as compatible if the overlap with any other compatible HSP was no more than 50\% of the length of either.
 The best hit for each query was chosen as the hit with the highest sum of bit scores from disjoint HSPs.
-The best hits were then filtered by overlap and reciprocity criteria.
-The overlap criterion was applied first and requires that 50\% of residues in the query are aligned in compatible HSPs.
-This excludes false positives from conserved domains embedded in larger non-homologous proteins by ensuring the hits span a sufficient fraction of the query and target sequences.
-The reciprocity criterion requires each query-target pair has a corresponding hit where the roles are reversed, which ensures there is no ambiguity in which target is the best match for the query.
+The best hits were filtered by overlap and reciprocity criteria.
+The overlap criterion was applied first and required that 50\% of residues in the query were aligned in compatible HSPs.
+This excluded false positives from conserved domains embedded in larger non-homologous proteins by ensuring the hits spanned a sufficient fraction of the query and target sequences.
+The reciprocity criterion required each query-target pair had a corresponding hit where the roles were reversed, which ensured there was no ambiguity in which target was the best match for the query.
 
@@
 In the first, maximal cliques were identified.
-In the second, a percolation graph is constructed by defining edges between cliques if they have \textit{k}-1 nodes in common.
-Clusters are the connected components of this second graph.
-The first step uses the NetworkX implementation of a maximal clique algorithm.
-The second step uses a modification of the NetworkX implementation of the \textit{k}-clique community algorithm.
+In the second, a percolation graph was constructed by defining edges between cliques if they had \textit{k}-1 nodes in common.
+Clusters were the connected components of this second graph.
+The first step used the NetworkX implementation of a maximal clique algorithm.
+The second step used a modification of the NetworkX implementation of the \textit{k}-clique community algorithm.
 The NetworkX implementation exhaustively finds all edges in the percolation graph.
-Since joining a \textit{k}-clique community only requires that a clique has a single edge connecting it to that a community, this approach is needlessly expensive for large graphs.
-The custom implementation instead uses a progressive approach where each clique is checked against a list of known communities, merging communities as necessary in each step.
+Since joining a \textit{k}-clique community only requires that a clique has a single edge connecting it to that a community, this approach was needlessly expensive for large graphs.
+The custom implementation instead used a progressive approach where each clique was checked against a list of known communities, merging communities as necessary in each step.
 
-The hit graph is sparse, so these algorithms are efficient when applied to its individual connected components.
-However, some components have a structure with many maximal cliques, which dramatically slows the first or second step of the clique percolation algorithm.
+The hit graph was sparse, so these algorithms were efficient when applied to its individual connected components.
+However, some components had a structure with many maximal cliques, which dramatically slowed the first or second step of the clique percolation algorithm.
 Thus, if either step exceeded 90 s, the process timed out, and the simpler \textit{k}-core algorithm was used instead.
@@
 Thus, all HSPs for each query were merged into hits.
-The best hit for each query and target gene was then chosen as the hit with the highest sum of bit scores from disjoint HSPs.
-(Grouping by target gene ensured only the highest-scoring isoform was selected.) Query-target pairs whose hits which exceeded the maximum bit score for all inter-genome hits associated with that query and passed the overlap and reciprocity filters were designated as paralogous pairs.
+The best hit for each query and target gene was chosen as the hit with the highest sum of bit scores from disjoint HSPs.
+(Grouping by target gene ensured only the highest-scoring isoform was selected.) Query-target pairs whose hits exceeded the maximum bit score for all inter-genome hits associated with that query and passed the overlap and reciprocity filters were designated as paralogous pairs.
 The orthologous groups were supplemented with paralogs by adding the paired sequences for each of the original members of the orthologous group.
@@
 The sequences in each orthologous group were aligned using MAFFT 7.490 with the following settings: -{}-globalpair -{}-maxiterate 1000 -{}-thread 1 -{}-anysymbol -{}-allowshift -{}-leavegappyregion -{}-unalignlevel 0.4~\cite{Katoh2013}.
-Representative sequences for each gene were then selected by maximum likelihood according to binary profiles constructed from these alignments.
+Representative sequences for each gene were selected by maximum likelihood according to binary profiles constructed from these alignments.
 First each sequence was coded into gap and non-gap symbols.
-The sequences were then grouped by gene, and for each group and position if at least one sequence was aligned in the group, the group contributed one count for the non-gap symbol to the profile at that position.
+The sequences were grouped by gene, and for each group and position if at least one sequence was aligned in the group, the group contributed one count for the non-gap symbol to the profile at that position.
 Otherwise, the group contributed a count for the gap symbol at that position.
@@
 If a species had multiple genes in the orthologous group, the species weight was divided evenly among them.
-Each coded sequence was then scored according to this profile, and the maximum likelihood sequence for each gene was selected as its representative.
-By assigning a non-gap count to groups and positions where at least one sequence is aligned, the profile incentivizes the selection of sequences contain the fewest gaps and match the consensus alignment.
-Unfortunately, this scheme can cause a sequence to score negative infinity if it has a gap at a position where every group has at least one aligned sequence, so the profile was initialized with a pseudocount of 0.005 for the gap and non-gap symbols at each position.
+Each coded sequence was scored according to this profile, and the maximum likelihood sequence for each gene was selected as its representative.
+By assigning a non-gap count to groups and positions where at least one sequence was aligned, the profile prioritized the selection of sequences with the fewest gaps that best matched the consensus alignment.
+Since this scheme can cause a sequence to score negative infinity if it has a gap at a position where every group has at least one aligned sequence, the profile was initialized with a pseudocount of 0.005 for the gap and non-gap symbols at each position.
 
@@
 A binary profile was created from the alignment using Gaussian process sequence weighting as described in the section on selecting representative sequences.
-(Because the orthologous groups were single copy and contained only representative sequences, each sequence received the full weight associated with its species.) The binary profile was then converted into a binary mask by identifying where the weighted fraction of non-gap symbols exceeded 0.5.
+(Because the orthologous groups were single copy and contained only representative sequences, each sequence received the full weight associated with its species.) The binary profile was converted into a binary mask by identifying where the weighted fraction of non-gap symbols exceeded 0.5.
 The binary mask was closed with a structuring element of size three, and highly conserved regions were identified as the contiguous intervals of this closed mask with a minimum length of 10.
@@
 For each diverged region, the corresponding segments of the sequences in the initial alignment were extracted and aligned with \textit{a\textsubscript{max}} set to 0.4.
-The resulting sub-alignments were then stitched into the initial alignment.
+The resulting sub-alignments were stitched into the initial alignment.
 
@@
 However, the emission probabilities were calculated as the posterior probability of the observed symbol given the data rather than the probability of the data.
-Only the alignments of the single copy orthologous groups were curated, so each tip in the species tree would correspond to a single sequence in the alignment.
+Only the alignments of the single copy orthologous groups were curated, so each tip in the species tree corresponded to a single sequence in the alignment.
 
@@
 The posterior distributions over states were calculated for each alignment using the trained insertion phylo-HMM.
-Regions with a high probability of state 2 or 3 are candidates for trimming.
-However, because the probability of a state can change rapidly or gradually depending on the local context, a simple cutoff would not necessarily define the boundaries of these regions as the columns where the gap pattern changes most abruptly.
+Regions with a high probability of state 2 or 3 were candidates for trimming.
+However, because the probability of a state can change rapidly or gradually depending on the local context, a simple cutoff would not necessarily define the boundaries of these regions as the columns where the gap pattern changed most abruptly.
 Instead, the following algorithm was used.
-First, a high cutoff defines a ``seed'' region.
-The left and right endpoints of the seed are then expanded both inwards and outwards to define two intervals from which boundaries are selected.
-The outward expansion halts when the probability or its derivative is below two distinct thresholds, respectively.
-The inward expansion halts when the derivative is below a different threshold.
-The left and right boundaries are chosen as the columns in each interval with the maximum product between the derivative and the change in the gap profile between columns.
-The gap profile is calculated as the number of gaps in each column using Gaussian process sequence weighting as described in the section on selecting representative sequences.
-By combining where the model’s confidence changes rapidly with the observed change in the gap pattern, this method generally selects reasonable boundaries.
+First, a high cutoff defined a ``seed'' region.
+The left and right endpoints of the seed were then expanded both inwards and outwards to define two intervals from which boundaries were selected.
+The outward expansion halted when the probability or its derivative was below two distinct thresholds, respectively.
+The inward expansion halted when the derivative was below a different threshold.
+The left and right boundaries were chosen as the columns in each interval with the maximum product between the derivative and the change in the gap profile between columns.
+The gap profile was calculated as the number of gaps in each column using Gaussian process sequence weighting as described in the section on selecting representative sequences.
+By combining where the model’s confidence changed rapidly with the observed change in the gap pattern, this method generally selected reasonable boundaries.
 
- States 2 and 3 have distinct characteristics which require different trimming strategies, so regions with a high probability of state 3 were handled first.
-Because the long poorly supported segments in state 3 regions were sometimes aligned to highly conserved columns or short segments in other sequences, trimming columns entirely would remove these segments from the other sequences even if they do not qualify as long and poorly supported themselves.
-Thus, regions with high state 3 probabilities were trimmed at the level of individual sequences rather than entire columns with the following method.
-First, the probability of state 3 of columns with a gap profile value less than or equal to 0.1 was set to 0 to break long poorly supported segments aligned to highly conserved columns into separate regions.
-Regions were then defined with the previously described algorithm using high and low cutoffs of 0.75 and 0.01, respectively.
+ States 2 and 3 have distinct characteristics which required different trimming strategies.
+Regions with a high probability of state 3 were handled first.
+Because the long poorly supported segments in state 3 regions were sometimes aligned to highly conserved columns or short segments in other sequences, trimming columns entirely would remove these segments from the other sequences even if they would not qualify as long and poorly supported themselves.
+Thus, regions with high state 3 probabilities were trimmed at the level of individual sequences rather than entire columns using the following method.
+First, the probability of state 3 for columns with a gap profile value less than or equal to 0.1 was set to 0 to break long poorly supported segments aligned to highly conserved columns into separate regions.
+Regions were defined with the previously described algorithm using high and low cutoffs of 0.75 and 0.01, respectively.
 The outer and inner derivative cutoffs were both 0.001.
@@
 (The sequences with the five most non-gap symbols were also excluded to not bias the estimate with long poorly supported segments.) The final mean $\mu$ was taken as the minimum of this value and two.
-A cutoff \textit{k}, derived from a geometric model of the number of non-gap symbols and a significance level $\alpha$, was calculated using the following equation $k = \log(\alpha) / \log(1 - p) - 1$ where $p = \frac{1}{\mu + 1}$ and $\alpha = 0.01$.
+A cutoff \textit{k}, derived from a geometric model of the number of non-gap symbols and a significance level $\alpha$, was calculated using the equation $k = \log(\alpha) / \log(1 - p) - 1$ where $p = \frac{1}{\mu + 1}$ and $\alpha = 0.01$.
 Any sequence whose number of non-gap symbols in the region equaled or exceeded this value was trimmed by replacing all non-gap symbols with gaps.
@@
 Regions were defined from this combined probability using the algorithm described previously except with a high probability cutoff of 0.9 instead of 0.75.
-The posterior probabilities from the missing data phylo-HMM were also converted into state assignments using a similar method.
-However, the initial seeds were defined with a cutoff of 0.75, and the seeds were expanded outward to first non-gap symbol or until the posterior probability of the ``missing data'' state was below 0.05.
+The posterior probabilities from the missing data phylo-HMM were converted into state assignments using a similar method.
+However, the initial seeds were defined with a cutoff of 0.75, and the seeds were expanded outward to the first non-gap symbol or until the posterior probability of the ``missing data'' state was below 0.05.
 These assignments, which are available in the supplementary data, can be used to filter segments or entire sequences from downstream analyses.
@@
 The orthology inference pipeline was first repeated with the outgroup species \textit{Scaptodrosophila lebanonensis}.
-Then orthologous groups with one sequence for each species were aligned, and 100 meta-alignments were constructed by randomly sampling 10,000 columns from these 9,435 alignments.
+Orthologous groups with one sequence for each species were aligned, and 100 meta-alignments were constructed by randomly sampling 10,000 columns from these 9,435 alignments.
 (The alignments were not refined before sampling.) To determine the effect of invariant columns and gaps, two sampling strategies were used where invariant columns were allowed or disallowed and the maximum fraction of gaps was set at 0, 50, and 100\%.
@@
 If the sampling strategy allowed invariant columns, an invariant rate category was included.
-The resulting trees from each set were then merged into a majority consensus tree (Fig.~\ref{sfig:trees_LG}).
+The resulting trees from each set were merged into a majority consensus tree (Fig.~\ref{sfig:trees_LG}).
 
 To fit trees using the GTR model of nucleotide substitution, the protein alignments were converted to nucleotide alignments using the corresponding coding sequences in the genome annotations.
-Some protein sequences are ``low quality,'' meaning their coding sequences contain frameshifts, premature stop codons, or other errors even though they are strong hits to known protein-coding genes.
+Some protein sequences were ``low quality,'' meaning their coding sequences contained frameshifts, premature stop codons, or other errors even though they were strong hits to known protein-coding genes.
 The NCBI annotation pipeline corrects some of these defects in the protein sequences, which can complicate a simple ``reverse translation'' of the alignment.
@@
 In this case, only the presence of invariant columns was varied, yielding two sets of meta-alignments.
-Trees were then fit using the GTR2 model with no rate categories.
+Trees were fit using the GTR2 model with no rate categories.
 An invariant category, however, was included if invariant columns were allowed.
-Since the bootstrap confidences were sometimes lower than 50\%, the resulting trees from each set were then merged into a loose consensus tree to prevent multifurcations (Fig.~\ref{sfig:trees_GTR}).
+Since the bootstrap confidences were sometimes lower than 50\%, the resulting trees from each set were merged into a loose consensus tree to prevent multifurcations (Fig.~\ref{sfig:trees_GTR}).
 

### 2023-02-15 chapter1/chapter1.tex
Removed section numbering from chapter1.tex and reorganized imports

@@
 
-\section*{Introduction}
+\section{Introduction}
 Comparative genomics is a powerful tool for yielding insights into evolutionary relationships, molecular function, and the forces that drive gene, genome, and population evolution.
@@
 
-\section*{Results}
-\subsection*{Pipeline overview}
+\section{Results}
+\subsection{Pipeline overview}
 Our pipeline follows a similar overall approach to other graph-based methods of orthology inference.
@@
 
-\subsection*{Input genomes and pre-processing}
+\subsection{Input genomes and pre-processing}
 All annotated genomes in the genus \textit{Drosophila} available in April 2022 were downloaded from NCBI's RefSeq database.
@@
 
-\subsection*{Extraction of best hits from BLAST output}
+\subsection{Extraction of best hits from BLAST output}
 The protein sequences in each genome annotation were searched against each other in reciprocal pairs using BLAST, yielding a list of high-scoring segment pairs (HSPs) for each query-target pair~\cite{Camacho2009}.
@@
 
-\subsection*{Clustering in orthologous groups}
+\subsection{Clustering in orthologous groups}
 The best hits between sequences are naturally visualized as a graph where sequences are nodes and best hits are edges between nodes.
@@
 
-\subsection*{Addition of paralogs to orthologous groups}
+\subsection{Addition of paralogs to orthologous groups}
 A weakness of the best reciprocal hits criterion is its exclusion of recently diverged paralogs.
@@
 
-\subsection*{Grouping orthologous groups by gene}
+\subsection{Grouping orthologous groups by gene}
 As genes can have several annotated isoforms, each gene can be associated with several orthologous groups.
@@
 
-\subsection*{Initial alignment and selection of representative sequences}
+\subsection{Initial alignment and selection of representative sequences}
 Since the NCBI annotation pipeline incorporates transcriptome data from a variety of sources, its inputs are heterogeneous in sequencing depth, developmental stage, and tissue of origin across different genomes.
@@
 
-\subsection*{Selection of single copy orthologous groups}
+\subsection{Selection of single copy orthologous groups}
 The criteria for selecting orthologous groups for further analyses depends on the biological question under investigation.
@@
 
-\subsection*{Alignment refinement}
+\subsection{Alignment refinement}
 Though the pipeline’s quality control measures ensure a high degree of overall sequence identity between members of an orthologous group, some sequences contain long ``poorly supported'' segments which have no homology to most or any other sequences in the alignment.
@@
 
-\subsection*{Alignment curation}
+\subsection{Alignment curation}
 Although the refinement process corrects most cases of over-alignment, the alignment may still contain regions whose aligned segments have poor or inconsistent support.
@@
 
-\subsection*{Inference of species trees}
+\subsection{Inference of species trees}
 Many phylogenetic methods require a species tree to inform the evolutionary relationships between sequences.
@@
 
-\section*{Discussion}
+\section{Discussion}
 The orthologous groups and alignments yielded by this pipeline are a valuable resource for comparative studies of gene birth/death processes and protein evolution at the level of both entire proteomes and specific gene families in the \textit{Drosophila} genus.
@@
 
-\section*{Materials and methods}
-\subsection*{Sequence de-duplication and BLAST search parameters}
+\section{Materials and methods}
+\subsection{Sequence de-duplication and BLAST search parameters}
 The protein sequences for each annotation were de-duplicated by removing any sequences which had already appeared in association with the same gene.
@@
 
-\subsection*{Extraction of HSPs from BLAST output}
+\subsection{Extraction of HSPs from BLAST output}
 To reduce the computational burden of merging HSPs into hits, the BLAST output was filtered to extract HSPs associated with the highest-scoring gene.
@@
 
-\subsection*{Merging of HSPs into hits}
+\subsection{Merging of HSPs into hits}
 HSPs were merged in two stages where the first combined non-overlapping HSPs, and the second combined the remaining HSPs.
@@
 
-\subsection*{Clustering by \textit{k}-clique percolation}
+\subsection{Clustering by \textit{k}-clique percolation}
 \textit{k}-clique percolation was implemented in two steps.
@@
 
-\subsection*{Addition of paralogs to orthologous groups}
+\subsection{Addition of paralogs to orthologous groups}
 The protein sequences for each annotation were searched against themselves with the same settings as for the inter-genome searches.
@@
 
-\subsection*{Initial alignment and selection of representative sequences}
+\subsection{Initial alignment and selection of representative sequences}
 The sequences in each orthologous group were aligned using MAFFT 7.490 with the following settings: -{}-globalpair -{}-maxiterate 1000 -{}-thread 1 -{}-anysymbol -{}-allowshift -{}-leavegappyregion -{}-unalignlevel 0.4~\cite{Katoh2013}.
@@
 
-\subsection*{Alignment refinement}
+\subsection{Alignment refinement}
 The representative sequences in the single copy orthologous groups were aligned with the same settings as described in the previous section except \textit{a\textsubscript{max}} was set to 0.7.
@@
 
-\subsection*{Alignment curation}
+\subsection{Alignment curation}
 The alignments were coded into gap and non-gap symbols to simplify the emission distributions.
@@
 
-\subsection*{Inference of species trees}
+\subsection{Inference of species trees}
 The orthology inference pipeline was first repeated with the outgroup species \textit{Scaptodrosophila lebanonensis}.
@@
 
-\subsection*{Code and data availability}
+\subsection{Code and data availability}
 \begin{sloppypar}

### 2023-02-15 chapter2/chapter2.tex
Updated section formatting in remaining files

@@
 \noindent
-Abstract text for this chapter
+Here's where my abstract will go!
 \end{abstract}
 
-\section*{Introduction}
-Here's where my chapter on evolutionary analyses will go!
+\section{Introduction}
+Here's where my introduction will go!
 
-\section*{Results}
+\section{Results}
+Here's where my results will go!
 
-\section*{Discussion}
+\section{Discussion}
+Here's where my discussion will go!
 
-\section*{Materials and methods}
+\section{Materials and methods}
+Here's where my materials and methods will go!

### 2023-02-15 chapter3/chapter3.tex
Updated section formatting in remaining files

@@
 
-\section*{Introduction}
+\section{Introduction}
 Most code written for data analysis is \textit{ad hoc}, that is, it is intended to accomplish a single task and cannot be used for any other purpose.

### 2023-02-15 main.tex
Updated section formatting in remaining files

\documentclass[11pt,letterpaper,oneside]{book}

% PACKAGE CONFIGURATIONS
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage[labelfont=bf, labelsep=period]{caption}
\renewcommand{\figurename}{Fig}

\usepackage[margin=1in]{geometry}
%\usepackage{setspace}  % double spacing
%\doublespacing
\usepackage[skip=10pt]{parskip}  % Remove indents
\usepackage{titlesec}  % Modify chapter title format
\titleformat{\chapter}
    [display]% Shape (chapter label in separate paragraph)
    {\bfseries\Large}% Format
    {\MakeUppercase{\chaptertitlename} \thechapter}% Label
    {0in}% Sep
    {}% Before code
    []% After code
\setcounter{tocdepth}{1}  % Only number divisions to section

\usepackage{hyperref}
\usepackage[doi=false,url=false]{biblatex}
\addbibresource{refs.bib}

\pagestyle{plain}

% ENVIRONMENTS
\newenvironment{abstract}%  Astract environment
    {% Begin code
    \small%
    \begin{center}%
    \textbf{Abstract}%
    \end{center}%
    \quotation%
    }%
    {\par% End code
    }

% DOCUMENT
\begin{document}

\begin{titlepage}
    \begin{center}
    My long and grand sounding dissertation title\\
    \bigskip
    By\\
    Marc Singleton\\
    \vfill
    A dissertation submitted in partial satisfaction of the\\
    requirements for the degree of\\
    Doctor of Philosophy\\
    in\\
    Biophysics\\
    in the\\
    Graduate Division\\
    of the\\
    University of California, Berkeley\\
    \vfill
    Committee in charge:\\
    Professor Mike Eisen, Chair\\
    Professor Liana Lareau\\
    Professor Susan Marqusee\\
    Professor Priya Moorjani\\
    \bigskip
    Spring 2023\\
    \end{center}
\end{titlepage}

\thispagestyle{empty}
\null  % Necessary to have text or command before \vfill for proper spacing
\vfill
\begin{center}
    \copyright \  2023 Marc Singleton
\end{center}
\clearpage

\setcounter{page}{1}  % Reset page counter after copyright
\begin{center}
    Abstract\\
    \bigskip
    My long and grand sounding dissertation title\\
    by\\
    Marc Singleton\\
    \bigskip
    Doctor of Philosophy in Biophysics\\
    University of California, Berkeley\\
    Professor Mike Eisen, Chair\\
    \bigskip
\end{center}
Here's the text for my abstract!
\clearpage

\setcounter{page}{1}  % Reset page counter after abstract
\pagenumbering{roman}  % Set numbering style to roman for pre-main content

\null  % Necessary to have text or command before \vfill for proper spacing
\vfill
\begin{center}
    To my parents, Amy and David, who were always willing\\
    to indulge another ``experiment'' in their kitchen sink
\end{center}
\vfill
\clearpage

\begin{center}
    \bfseries\Large Acknowledgements
\end{center}
\bigskip
If it takes a village to raise a child, then it must take at least three to raise a scientist. Over the years, I've had countless friends, teachers, and mentors. These relationships have shaped the person I am today, so I'm grateful to all the people who took time out of their busy lives to help me in ways both small and large. However, I want to single out the following people who had an especially important impact on my time in graduate school.\\\\
Thank you to my advisor, Mike, for supporting me as a scientist and a person in equal measure.\\\\
Thank you to my committee, Susan, Liana, and Priya, for giving me the pep talk I needed every year.\\\\
Thank you to my lab, Holli, Stadler, Ciera, Ashley, Victoria, Augusto, and Xiao-Yong, for creating a space where it's okay to ask questions. A special thank you to my bay mates, Jenna and Colleen, for getting me through the hard days with sound advice and a healthy dose of humor.\\\\
Thank you to my undergraduate research mentors, Cathy and Jessica, for teaching me the most exciting science is at the intersection of different fields.\\\\
Thank you to my parents, for everything.\\\\
Thank you to my brothers, Andrew and Daniel, for giving me thick skin in the way only big brothers can.\\\\
Thank you to all the friends I've made in California and the Bay Area, for the much-needed life outside work.\\\\
Finally, thank you to my partner, Joseph, for going on this crazy ride with me. I couldn't have done this without you.
\clearpage

\tableofcontents

\setcounter{chapter}{-1}  % Set chapter counter to 0
\chapter{Introduction}
\graphicspath{{chapter0/figures/}}
\setcounter{page}{1}  % Reset page counter after pre-main content
\pagenumbering{arabic}  % Set numbering style to arabic for main content
\input{chapter0/chapter0.tex}

\chapter{Leveraging genomic redundancy to improve inference and alignment of orthologous proteins}
\graphicspath{{chapter1/figures/}}
\input{chapter1/chapter1.tex}

\chapter{Evolutionary analyses of IDRs reveal conserved features}
\input{chapter2/chapter2.tex}

\chapter{Software tools and tutorials}
\input{chapter3/chapter3.tex}

\printbibliography[heading=bibintoc]

\appendix
\setcounter{figure}{0}
\renewcommand{\thefigure}{\thechapter\arabic{figure}}
\setcounter{table}{0}
\renewcommand{\thetable}{\thechapter\arabic{table}}

\chapter{Supporting information for chapter 1}
\input{chapter1/chapter1_supporting.tex}

\end{document}

### 2023-02-20 chapter3/chapter3.tex
Modified end of introduction to chapter 3 to reflect inclusion of only the training tutorial

@@
 
-The tutorials are available alongside the source code for these packages and will likely continue to undergo revisions and expansions.
-They are, however, adapted from their current form in the following sections.
-Though the tutorials are largely independent of the others, here they are presented in their intended order.
-The first two are brief introductions to the MixMod and Homomorph APIs, respectively.
-The next discusses a generalization of HMMs called autoregressive HMMs (ARHMMs).
-The final tutorial covers methods for training HMMs.
-As each is a self-contained document with an introduction and conclusion, they are presented without further commentary.
+The tutorials are available alongside the source code for these packages on GitHub.
+Though each is largely independent of the others, they follow a logical progression.
+For example, both Mixmod and Homomorph have tutorials which serve as brief introductions to the theory, applications, and APIs for their respective models.
+Homomorph then has two additional tutorials.
+The first discusses a generalization of HMMs called autoregressive HMMs (ARHMMs), and the second covers methods for training HMMs.
+For brevity, only the training tutorial is adapted from its current form in the following section.
+As the tutorial is a self-contained document with an introduction and conclusion, it is presented without further commentary.

### 2023-02-21 main.tex
Added notebook environments for code listings and re-organized packages again

@@
 
-% PACKAGE CONFIGURATIONS
+% PACKAGE AND ENVIRONMENT CONFIGURATIONS
 \usepackage{amsmath}
 \usepackage{graphicx}
-\usepackage[labelfont=bf, labelsep=period]{caption}
-\renewcommand{\figurename}{Fig}
 
+% Code listings
+\usepackage[outputdir=out]{minted}
+\usepackage{tcolorbox}
+\usepackage{lmodern}
+
+% Margins and spacing
 \usepackage[margin=1in]{geometry}
@@
 \usepackage[skip=10pt]{parskip}  % Remove indents
+\pagestyle{plain}
+
+% Chapter and figure formatting
 \usepackage{titlesec}  % Modify chapter title format
-\titleformat{\chapter}
-    [display]% Shape (chapter label in separate paragraph)
+\titleformat{\chapter}[display]% Shape (chapter label in separate paragraph)
                       {\bfseries\Large}% Format
@@
 \setcounter{tocdepth}{1}  % Only number divisions to section
+\usepackage[labelfont=bf, labelsep=period]{caption}
+\renewcommand{\figurename}{Fig}
 
-\usepackage{hyperref}
+% References
 \usepackage[doi=false,url=false]{biblatex}
 \addbibresource{refs.bib}
-
-\pagestyle{plain}
+\usepackage{hyperref}
 
 % ENVIRONMENTS
-\newenvironment{abstract}%  Astract environment
+\newenvironment{abstract}
 {% Begin code
-    \small%
-    \begin{center}%
-    \textbf{Abstract}%
-    \end{center}%
-    \quotation%
-    }%
-    {\par% End code
+ \small
+ \begin{center}
+ \textbf{Abstract}
+ \end{center}
+ \quotation
+}
+{% End code
+}
+
+\definecolor{incolor}{RGB}{51,63,153}
+\definecolor{outcolor}{RGB}{200,78,41}
+\definecolor{bgcolor}{RGB}{247,247,247}
+
+\newlength{\promptwidth}
+\setlength{\promptwidth}{0.5in}
+
+\newcommand{\prompt}[2]
+{\makebox[0pt][r]{\texttt{\color{#2}#1:\ }}\vspace{-\baselineskip}}
+
+\newenvironment
+{NotebookIn}
+{
+ \VerbatimEnvironment
+ \begin{tcolorbox}[sharp corners,
+                   frame empty,
+                   size=fbox,
+                   colback=bgcolor,
+                   left skip=\promptwidth,
+                   title=\prompt{In}{incolor},
+                   attach title to upper]
+ \begin{minted}{python}
+}
+{
+ \end{minted}
+ \end{tcolorbox}
+}
+
+\newenvironment
+{NotebookOut}
+{
+ \begin{tcolorbox}[sharp corners,
+                   frame empty,
+                   size=fbox,
+                   colback=white,
+                   left skip=\promptwidth,
+                   title=\prompt{Out}{outcolor},
+                   attach title to upper]
+ \\ \ttfamily
+}
+{
+ \end{tcolorbox}
+}
+
+\newenvironment
+{NotebookImage}
+{
+ \begin{tcolorbox}[sharp corners,
+                   frame empty,
+                   size=fbox,
+                   colback=white,
+                   left skip=\promptwidth,
+                   title=\prompt{Out}{outcolor},
+                   attach title to upper]
+ \\
+}
+{
+ \end{tcolorbox}
 }
@@
 \chapter{Software tools and tutorials}
+\graphicspath{{chapter3/}}
 \input{chapter3/chapter3.tex}
@@
 \chapter{Supporting information for chapter 1}
+\graphicspath{{chapter1/figures/}}
 \input{chapter1/chapter1_supporting.tex}

### 2023-02-22 chapter3/introduction.tex
Re-named chapter 3

--- 2023-02-20 chapter3/chapter3.tex
@@
 
-\section{Introduction}
+\section{Overview of tools and tutorials}
 Most code written for data analysis is \textit{ad hoc}, that is, it is intended to accomplish a single task and cannot be used for any other purpose.

### 2023-02-23 main.tex
Re-named chapter 3
Modified notebook environments and added new material as inputs
Change list:
Added % to environments to prevent interpretation as returns
Made code listing boxes breakable
Made outlines white because they appeared on broken boxes regardless of other settings
Adjusted indent

@@
 \usepackage{amsmath}
+\DeclareMathOperator*{\argmax}{arg\!\max}
+\DeclareMathOperator*{\argmin}{arg\!\min}
+
 \usepackage{graphicx}
@@
 \usepackage{tcolorbox}
+\tcbuselibrary{breakable}
+\tcbset{colframe=white}
 \usepackage{lmodern}
@@
 \newlength{\promptwidth}
-\setlength{\promptwidth}{0.5in}
+\setlength{\promptwidth}{0.375in}
 
@@
 {NotebookIn}
-{
- \VerbatimEnvironment
+{%
+ \VerbatimEnvironment%
  \begin{tcolorbox}[sharp corners,
                    frame empty,
+                   breakable,
                    size=fbox,
@@
                    title=\prompt{In}{incolor},
-                   attach title to upper]
- \begin{minted}{python}
+                   attach title to upper]%
+ \begin{minted}{python}%
 }
-{
- \end{minted}
- \end{tcolorbox}
+{%
+ \end{minted}%
+ \end{tcolorbox}%
 }
@@
 {NotebookOut}
-{
+{%
+ \VerbatimEnvironment%
+ \begin{tcolorbox}[sharp corners,
+                   frame empty,
+                   breakable,
+                   size=fbox,
+                   colback=white,
+                   left skip=\promptwidth,
+                   title=\prompt{Out}{outcolor},
+                   attach title to upper]%
+ \begin{minted}{text}%
+}
+{%
+ \end{minted}%
+ \end{tcolorbox}%
+}
+
+\newenvironment
+{NotebookImage}
+{%
  \begin{tcolorbox}[sharp corners,
@@
                    title=\prompt{Out}{outcolor},
-                   attach title to upper]
- \\ \ttfamily
+                   attach title to upper]%
+ \\%
 }
-{
- \end{tcolorbox}
-}
-
-\newenvironment
-{NotebookImage}
-{
- \begin{tcolorbox}[sharp corners,
-                   frame empty,
-                   size=fbox,
-                   colback=white,
-                   left skip=\promptwidth,
-                   title=\prompt{Out}{outcolor},
-                   attach title to upper]
- \\
-}
-{
- \end{tcolorbox}
+{%
+ \end{tcolorbox}%
 }
@@
 
-\chapter{Software tools and tutorials}
+\chapter{Tools and tutorials for fitting mixture models and HMMs}
 \graphicspath{{chapter3/}}
-\input{chapter3/chapter3.tex}
+\input{chapter3/introduction.tex}
+\input{chapter3/training_tutorial}
 
@@
 \chapter{Supporting information for chapter 1}
+\label{appendix:a}
 \graphicspath{{chapter1/figures/}}
@@
 
+\chapter{Supporting information for chapter 3}
+\label{appendix:b}
+\setlength{\promptwidth}{0in}  % No indent for appendix code
+\input{chapter3/training_tutorial_supporting.tex}
+
 \end{document}

### 2023-02-23 chapter3/training_tutorial.tex
Added remaining chapter 3 material

\section{HMM training tutorial introduction}
Welcome to the training tutorial! Since Homomorph doesn't have built-in methods for fitting HMMs to data, in this tutorial we'll cover examples of training HMMs where the hidden states are both known and unknown using maximum likelihood estimation and the estimation-maximization algorithm, respectively. We'll then discuss an advanced approach called discriminative training.

The material in sections four through six owes a heavy debt to two sources, respectively. The first is the book \textit{Biological Sequence Analysis}, which is an excellent resource on HMMs and other probabilistic models and their applications~\cite{Durbin1998}. The second is the paper Hidden Neural Networks by Anders Krogh and Søren Riism~\cite{Krogh1999}. Many of the theoretical results for HMMs shown here are adapted or expanded from the derivations outlined in those references, so this tutorial wouldn't have been possible without them.

Clearly we have a lot of ground to cover, so let's get started!

\section{What is training?}

\subsection{Training, informally}

Training and learning are jargon commonly used in machine learning. While convenient as shorthand, they obscure what's actually happening: parameter estimation or, more generally, optimization. Statistical models are specified by some number of parameters, so the name of the game is choosing parameters that fit the data the best. Depending on the model, this can be as simple as taking an average or as complex as needing specialized algorithms and thousands of computing hours. Note that the notion of a ``best'' fit to the data isn't always immediately obvious, and some applications will define it differently. (We'll see at least two distinct definitions in the following sections.) In every case, however, some quantity of interest is defined that is hopefully correlated with the ability to predict real-world outcomes, and then some computational method is used to find parameters that maximize or minimize that quantity.

\subsection{Training, formally}

Now that we understand what training is intuitively, let's briefly discuss how it's defined mathematically. This section will mainly serve to introduce some terminology used throughout this tutorial, but it will also highlight the features that are common to all optimization problems regardless of application or mathematical form.

At the core of every optimization problem is a function whose output is optimized with respect to one or more inputs. This is called the \textit{objective function} or simply the \textit{objective}. This terminology is common when discussing optimization generically, and the kind of extremum (minimum or maximum) is unstated. In machine learning and statistics, however, the objective is often framed as a kind of distance between the model's output and the actual data. Distance has a very specific meaning in mathematics, so instead this quantity is called \textit{loss}. Naturally, then, the goal is to minimize loss. Unlike in many optimization problems where the optimal value of the objective is as important as the inputs that achieve that optimum, when fitting models the loss is often only a means to an end. Though the loss has applications when comparing the fits of different models, the optimized parameters are typically the ultimate goal.

The form of the loss function depends on the exact nature of the problem, so it's difficult to go any further without speaking in overly general terms. Thus, to illustrate these concepts mathematically, we'll use an extremely common loss function, the mean squared error. Let's now define the data, model, and loss precisely. The data, $D$, are composed of $N$ examples which are the indexed ordered pairs $\{(x_i, y_i):1 \le i \le N\}$. Each $x_i$ and $y_i$ is an input and output, respectively. Though these quantities can be vector-valued, we'll keep our discussion general enough to not have to worry about these details. The model is a function, $f$, which accepts an input, $x$, and returns an output, $\hat{y}$. The output is designated with a hat to indicate it was calculated from the model rather than observed. This function also accepts a set of parameters, $\Theta$, which can be tuned to better fit the data. Thus, we write $\hat{y} = f(x, \Theta)$. Finally the loss function, $L$, accepts the data, model, and parameters and returns a measure of the deviation from the data, $L(D, f, \Theta)$. For the mean squared error, this is written as

\begin{align*}
L(D, f, \Theta)
&= \frac{1}{N} \sum_{i=1}^N \left( \hat{y}_i - y_i \right)^2 \\
&= \frac{1}{N} \sum_{i=1}^N \left( f(x_i, \Theta) - y_i \right)^2. \\
\end{align*}

In this case, the loss is expressed as a sum over individual examples. This is extremely common, so losses are frequently written in terms of an individual input and output pair rather than the data as a whole. Furthermore, since the data and model are usually fixed, they are often dropped as arguments, making the loss a function of the parameters alone:

\begin{equation*}
L(D, f, \Theta) = L(\Theta).
\end{equation*}

The optimized or ``learned'' parameters are then written as

\begin{equation*}
\hat{\Theta} = \argmin_\Theta L(\Theta).
\end{equation*}

Finding these optimal parameters is not always straightforward. For simple models, such as one-dimensional linear regression where $f(x, \Theta) = \theta_1 x + \theta_0$, there are closed-form solutions. However, many modern machine learning models offer no such luxury, so other approaches are needed. There are various techniques for such cases depending on the structure of the model. However, all are usually iterative in nature, meaning they gradually decrease the loss in a series of steps. For complex models and large data sets, each step can involve significant computation, which is why these models are often trained on computational clusters with specialized hardware.

\section{Training with known states}

Sometimes the universe is kind and gives us data where the underlying states are known. This is by far the easiest case since all the information is available to us, and as HMMs are probabilistic models, we can rely on the rich theory of mathematical statistics to define optimality and identify the corresponding parameters. Depending on the emission distributions, there are even closed-form solutions for these optimal parameters!

\subsection{Maximum likelihood estimation}

The strength of probabilistic models is that for a given set of parameters, every possible input is associated with a probability or probability density, so objective functions are naturally defined in terms of these quantities. This doesn't answer the question of what probability we should try to optimize, however. There are several common approaches, but one of the most natural is to maximize the probability of the data with respect to the parameters. Let's clarify what this means by writing it mathematically. We call our data, which is a set of indexed observations $\{x_i:1 \le i \le N\}$, $D$ and the probability mass function $f(x, \Theta) = P(X = x | \Theta)$. (The right-hand side translates to ``the probability that the random variable called $X$ assumes the value $x$ given the set of parameters $\Theta$.'' In probability, the random processes that generate observations are distinguished from the observations themselves.) The probability of the data is then written as

\begin{align*}
P(D|\Theta)
&= \prod_{i=1}^N P(X=x_i | \Theta) \\
&= \prod_{i=1}^N f(x_i, \Theta) \\
\end{align*}

since each observation is independent and identically distributed. Sums are easier to work with, so it's common to take the logarithm of both sides and call the result

\begin{equation*}
L(\Theta) = \log P(D|\Theta) = \sum_{i=1}^N \log f(x_i, \Theta)
\end{equation*}

the \textit{log-likelihood function}. The original product then is the \textit{likelihood function}; however, the logarithm doesn't change the position of the extrema, so this distinction is not very important for our purposes.

Before moving forward, let's discuss some nuances of the likelihood function. First, even though it's defined in terms of observations that assume discrete values (notice that $f$ is probability rather than a probability density), the definition is the same for continuous observations. The only difference is the likelihood is no longer considered a probability since it's technically a probability density. However, this not a common interpretation anyway since even for discrete data the probability of any set of observations approaches zero as the number of observations increases. It's instead more useful to compare ratios of likelihoods under different models. That, however, is a topic for another time. Another subtle point is the likelihood function is not the probability that the parameters are correct. Instead a particular choice of parameters is viewed as certain, and then the probability of the data is calculated given that choice. Finally, as with the mean squared loss defined earlier, though the likelihood is technically a function of both the parameters and the data, the data are often dropped as an argument to emphasize that they are typically fixed.

Unlike for a loss function, we want to maximize the log-likelihood since this maximizes the probability of the data under the model. This defines the following set of parameters

\begin{equation*}
\hat{\Theta} = \argmax_\Theta L(\Theta)
\end{equation*}

which are known as \textit{maximum likelihood estimates} (MLEs). Though this objective isn't quite a loss in the sense of a distance from some desired outputs, it's similar in spirit. Since the model should assign common events high probabilities (and accordingly rare events low probabilities since the total probability is constrained to sum to one), the log-likelihood in effect penalizes deviations from the empirical distribution. In fact, simply negating the log-likelihood function converts the maximization into a minimization, making it a kind of loss. Furthermore, for some models algebraic manipulations can reveal an expression which is more readily interpreted as a distance.

Calculus tells us that for differentiable functions local extrema are necessarily where the derivative relative to the input is zero. If $\Theta$ is a set of $N$ parameters, this condition must occur simultaneously for the derivative relative to each. In other words,

\begin{equation*}
\frac{\partial L}{\partial \theta_1} = 0, \quad \ldots, \quad \frac{\partial L}{\partial \theta_N} = 0
\end{equation*}

for each $\theta_i \in \Theta$. In some cases these equations can be solved explicitly, but often numerical techniques are needed. We'll see two examples of such approaches in later sections.

\subsection{Maximum likelihood estimates for HMMs}

\subsubsection{Decomposition into independent products}

Now we'll turn to HMMs and derive the MLEs for labeled data. Here the data, $D$, are again composed of $N$ examples of ordered pairs $\{(x_i, y_i): 1 \le i \le N\}$ where $x_i$ is a sequence of states and $y_i$ is a sequence of emissions, each with length $T_i$. The probability of the data given the parameters is then

\begin{align*}
P(D|\Theta)
&= P(X_1=x_1, Y_1=y_1, \ldots, X_N=x_N, Y_N=y_N|\Theta) \\
&= \prod_{i=1}^N P(X_i=x_i, Y_i=y_i|\Theta). \\
\end{align*}

The joint probability expands into a product of the probabilities of individual examples since each is independent and identically distributed. Let's focus on one pair of state and emission sequences denoted $x$ and $y$, each with length $T$. Using the Markov property of HMMs, we can derive an expression for their joint probability:

\begin{align*}
P(D|\Theta)
=& \ P(X=x, Y=y|\Theta) \\
=& \ P(Y_1=y_1|X_1=x_1,\Theta) P(X_1=x_1|\Theta) \\
 & \times \prod_{t=1}^{T-1} P(Y_{t+1}=y_{t+1}|X_{t+1}=x_{t+1},\Theta) P(X_{t+1}=x_{t+1}|X_t=x_t,\Theta). \\
\end{align*}

(Note that here the subscripts refer to the index within a sequence rather than the index of the example.)

All these capital letters are cluttering this expression, so we'll make a few common substitutions to simplify it. First, we assume there are $S$ states numbered from $1$ to $S$ and write the transition and start probabilities as $P(X_{t+1}=j|X_t=i, \Theta) = a_{ij}$ and $P(X_1=i|\Theta) = a_{0i}$, respectively. Second, we define $e_i(y_t) = P(Y_t=y_t|X_t=i, \Theta)$. We can think of each $e_i$ as a function that accepts an emission $y_t$ and outputs a probability or probability density. Putting all this together, we have

\begin{equation*}
P(X=x, Y=y|\Theta) = e_{x_1}(y_1)a_{0x_1} \prod_{t=1}^{T-1} e_{x_{t+1}}(y_{t+1}) a_{x_t x_{t+1}}.
\end{equation*}

Since log-likelihoods are easier to work with, we take the logarithm of both sides and write the product as a sum of log terms:

\begin{align*}
\log P(X=x, Y=y|\Theta)
&= \log \left( e_{x_1}(y_1)a_{0x_1} \prod_{t=1}^{T-1} e_{x_{t+1}}(y_{t+1}) a_{x_t x_{t+1}} \right) \\
&= \log a_{0x_1} + \sum_{t=1}^{T-1} \log a_{x_t x_{t+1}} + \sum_{t=1}^T \log e_{x_t}(y_t). \\
\end{align*}

This is now a fairly clean expression since at each step we ``pick'' the right transition probability and emission distribution using the state sequence. However, it will be useful both theoretically and computationally to write this expression in terms of the number of times each transition appears in the sequence. To do this, we first need the Kronecker delta function, which is defined as

\begin{equation*}
\delta_{ij} =
\begin{cases}
    0 &\text{if } i \neq j \\
    1 &\text{if } i=j. \\
\end{cases}
\end{equation*}

Then we can write the number of transitions between states $i$ and $j$ as

\begin{equation*}
n_{ij} = \sum_{t=1}^{T-1} \delta_{ix_t} \delta_{jx_{t+1}}.
\end{equation*}

We can define a similar variable that counts the number of times each state starts the state sequence:

\begin{equation*}
n_{0i} = \delta_{ix_1}.
\end{equation*}

For a single example, this might seem unnecessarily complex since one $n_{0i}$ is equal to one and all others are zero. However, this form will be convenient for generalizing to data that contain multiple examples.

Inspection of the previous equation shows we can write it using the quantities we've defined and a similar trick with the Kronecker delta function for the emissions:

\begin{align*}
\log P(X=x, Y=y|\Theta)
&= \log a_{0x_1}+ \sum_{t=1}^{T-1} \log a_{x_t x_{t+1}} + \sum_{t=1}^T \log e_{x_t}(y_t) \\
&= \sum_{i=1}^S n_{0i} \log a_{0i} +
   \sum_{i=1}^S \sum_{j=1}^S n_{ij} \log a_{ij} +
   \sum_{i=1}^S \sum_{t=1}^T \delta_{ix_t} \log e_{x_t}(y_t). \\
\end{align*}

(To ensure this expression is valid for forbidden start states or transitions, we define $0 \log0 = 0$.)

While this may look complicated, it's the same sum of the log probabilities of each start state, transition, and emission. However, when it's written in this form, two things are clear. First, many calculations are expressed as the logarithm of a parameter multiplied by the number of times it appears in the training data. Second, each of these sums is independent of the others, meaning they have no parameters in common. In fact, the transition sum can be broken into independent sums for each initial state. The same is true for the emission sum if none of the emission distributions share parameters. This dramatically simplifies the optimization since we can maximize the probability of the entire expression by maximizing each sum separately. Finally, although we've derived this expression for a single pair of state and emission sequences, the form is identical for the data as a whole. The only differences are the counts are taken over all state sequences and the emission sum is taken over all emission sequences.

\subsubsection{MLEs for categorical distributions}

Now that we've broken the maximization problem into a set of simpler problems, let's review the solutions for each. The optimal parameters for the emissions will depend on their distributions, but the start and transition distributions will always take the form of a single choice from a set of options. This is formally called a \textit{categorical distribution} which itself is a special case of the \textit{multinomial distribution}. Fortunately, the MLEs for categorical distributions have a simple form when they are parameterized directly in terms of the probability of selecting each outcome. The derivation is somewhat involved, so we'll skip to the final result. Using the count variables defined earlier, the MLE for each $a_{ij}$ is given as

\begin{equation*}
\hat{a}_{ij} = \frac{n_{ij}}{\sum_{j=1}^S n_{ij}}.
\end{equation*}

The interpretation is intuitive. Our estimate of the probability of state $i$ transitioning to state $j$ is simply the fraction of times we observe this in the data! One problem with this equation, however, is if we're working with a relatively small amount of data, we may never observe a certain transition and estimate its probability as zero. This means according to the model the transition is impossible, which may be contrary to our hypothesis of the underlying process. In these cases, it's customary to add a small non-negative correction factor, $r_{ij}$, for each pair of states:

\begin{equation*}
\hat{a}_{ij} = \frac{n_{ij} + r_{ij}}{\sum_{j=1}^S n_{ij} + r_{ij}}.
\end{equation*}

These corrections may look like a sloppy fix, but they have a natural Bayesian interpretation as the parameters of a Dirichlet prior on the transition probabilities. What this means in practice is the size of each $r_{ij}$ reflects the prior expectation for the probability of that transition, with larger values indicating more certainty.

\subsection{Examples}

Let's now take what we've learned and apply it to some examples. We'll first write code to estimate the parameters for an HMM with categorical emission distributions since we've already reviewed the MLEs in the previous section. We'll then use the principle of maximum likelihood to derive the estimators for other common emission distributions and write implementations from scratch.

\subsubsection{Categorical emission distribution}

To get started, we'll first import the packages and some plotting settings used throughout this tutorial.

\begin{NotebookIn}
import pprint
import random
from functools import reduce
from itertools import accumulate

import homomorph
import matplotlib.pyplot as plt
import numpy as np
import scipy.stats as stats
from numpy import exp, log
from sklearn.metrics import roc_curve
from utils import fit_CML

legend_kwargs = {'frameon': False,
                 'loc': 'center left',
                 'bbox_to_anchor': (1, 0.5)}
\end{NotebookIn}

Let's create our HMM. Since the purpose of this tutorial is to illustrate training techniques rather than to motivate the applications of HMMs with relevant examples, we'll arbitrarily label the states with numbers and the emissions with letters.

\begin{NotebookIn}
t_dists = {1: {1: 0.95, 2: 0.05},
           2: {1: 0.05, 2: 0.9, 3: 0.05},
           3: {2: 0.35, 3: 0.65}}
e_dists = {1: {'A': 1},
           2: {'A': 0.5, 'B': 0.5},
           3: {'B': 1}}
start_dist = {1: 0.2, 2: 0.5, 3: 0.3}

model = HMM(t_dists=t_dists, e_dists=e_dists, start_dist=start_dist)
model
\end{NotebookIn}

\begin{NotebookOut}
HMM(states={1, 2, 3},
    stop_states=[],
    name='hmm')
\end{NotebookOut}

We'll now generate 10 examples, each with length 200. The simulations are returned as a single sequence of (state, emission) tuples. However, since in the theory section we defined the states and emissions as separate sequences, we'll do a bit of Python magic to put the data in this form.

\begin{NotebookIn}
data = [model.simulate(200, random_state=i) for i in range(10)]
print('Original form:', data[0][:5])
data = [list(zip(*example)) for example in data]
print('New form:', [seq[:5] for seq in data[0]])
\end{NotebookIn}

\begin{NotebookOut}
Original form: [(2, 'A'), (1, 'A'), (1, 'A'), (1, 'A'), (1, 'A')]
New form: [(2, 1, 1, 1, 1), ('A', 'A', 'A', 'A', 'A')]
\end{NotebookOut}

We're now ready to implement the MLEs for the transition probabilities. Before, though, we should discuss selection of the model structure, that is, the number of states and the allowed transitions between those states. Since in this case the data are simulated, we know the structure exactly. However, modeling real-world data is often far more complicated since it's rare to have perfect knowledge of the data generating process even if the states are labeled. Thus, it's tempting to use a ``fully connected'' model that allows transitions between each pair of states and let the model learn what transitions are actually used from the data. That said, constraining the allowed transitions can lead to better models, particularly if the constraints reflect a feature of the underlying process.

For this example, we'll assume that we have some ``domain knowledge'' that permits us to know the states and their allowed transitions. For example, state 2 can transition to every state, but states 1 and 3 can only transition to themselves and state 2. It's also customary to add a small pseudocount to the allowed transitions to ensure they're permitted by our model in the rare chance that we don't observe them in the data. There are a variety of ways to implement this, but in the approach shown below we first instantiate a transition count dictionary with the pseudocounts. We then iterate over the data to add the observed transitions and afterwards normalize the counts by the total number for each initial state to obtain the estimated transition probabilities.

The strength of this approach is it yields a nested dictionary which we can use as an input to the \texttt{HMM} class. We'll actually use the original \texttt{t\_dists} to establish the model structure, but if it weren't available, we could create a similar nested object which encodes the same information. (Unfortunately, it's hard to avoid specifying at least some portions of the model manually unless it's fully-connected or its structure is highly modular.)

\begin{NotebookIn}
# Make transition count dicts and add pseudocounts
t_pseudo = 0.1
t_counts = {}
for state1, t_dist in t_dists.items():
    t_count = {}
    for state2 in t_dist:
        t_count[state2] = t_pseudo
    t_counts[state1] = t_count

# Add observed counts
for example in data:
    xs, ys = example
    state0 = xs[0]
    for state1 in xs[1:]:
        t_counts[state0][state1] += 1
        state0 = state1

# Normalize counts
t_dists_hat = {}
for state1, t_count in t_counts.items():
    t_sum = sum(t_count.values())
    t_dist_hat = {}
    for state2, count in t_count.items():
        t_dist_hat[state2] = count / t_sum
    t_dists_hat[state1] = t_dist_hat
t_dists_hat
\end{NotebookIn}

\begin{NotebookOut}
{1: {1: 0.9548785594639866,
     2: 0.0451214405360134},
 2: {1: 0.05571315102689209,
     2: 0.8965165097015771,
     3: 0.04777033927153069},
 3: {2: 0.2924773022049287,
     3: 0.7075226977950714}}
\end{NotebookOut}

So far so good! The estimated transition probabilities are extremely close to their actual values.

We'll now estimate the emission probabilities. Since the emission distributions are also categorical, the code has nearly the same structure. However, we'll make a small change to illustrate some choices inherent in model selection. Although we assumed we knew the states and allowed transitions perfectly, let's say this isn't true for the emissions. For example, even though we never observe states 1 and 3 emitting an A and B, respectively, we still aren't 100\% convinced that it's impossible. Thus, we'll first gather all the possible emission types from the data and instantiate the emission distributions with a small pseudocount for each type. From there the code is largely the same.

\begin{NotebookIn}
# Collect all possible emissions
e_set = set()
for example in data:
    xs, ys = example
    e_set.update(ys)

# Make emission count dicts and add pseudocounts
e_pseudo = 0.1
e_counts = {}
for state in t_dists:
    e_counts[state] = {emit: e_pseudo for emit in e_set}

# Add observed counts
for example in data:
    xs, ys = example
    for state, emit in zip(xs, ys):
        e_counts[state][emit] += 1

# Normalize counts
e_dists_hat = {}
for state, e_count in e_counts.items():
    e_sum = sum(e_count.values())
    e_dist_hat = {}
    for emit, count in e_count.items():
        e_dist_hat[emit] = count / e_sum
    e_dists_hat[state] = e_dist_hat
e_dists_hat
\end{NotebookIn}

\begin{NotebookOut}
{1: {'A': 0.999896071502806, 'B': 0.00010392849719393058},
 2: {'A': 0.48755937570685365, 'B': 0.5124406242931463},
 3: {'A': 0.000648508430609598, 'B': 0.9993514915693904}}
\end{NotebookOut}

Again the estimates are very close!

Let's finish this out with the start distribution. The overall idea is exactly the same, except now we're working with a single distribution rather than a dictionary of distributions. Technically, we can only estimate the start distribution from the initial state of each example. Since this severely limits the amount of data relative to the transitions, it may be tempting to use the state counts over all time steps instead. However, this quantity will estimate the equilibrium distribution of the underlying Markov process, which is a function of the transition probabilities and independent of the start distribution. However, there may be data-specific reasons to think these quantities are equal, so this is again an example of a model selection decision.

\begin{NotebookIn}
# Make start count dicts and add pseudocounts
start_pseudo = 0.1
start_count = {}
for state in start_dist:
    start_count[state] = start_pseudo

# Add observed counts
for example in data:
    xs, ys = example
    start_count[xs[0]] += 1

# Normalize counts
start_sum = sum(start_count.values())
start_dist_hat = {}
for state, count in start_count.items():
    start_dist_hat[state] = count / start_sum
start_dist_hat
\end{NotebookIn}

\begin{NotebookOut}
{1: 0.10679611650485439, 2: 0.5922330097087379, 3: 0.3009708737864078}
\end{NotebookOut}

We're now ready to combine all these individual parameter estimates to create an estimated model.

\begin{NotebookIn}
model_hat = HMM(t_dists=t_dists_hat,
                e_dists=e_dists_hat,
                start_dist=start_dist_hat)
model_hat
\end{NotebookIn}

\begin{NotebookOut}
HMM(states={1, 2, 3},
    stop_states=[],
    name='hmm')
\end{NotebookOut}

Though the estimates match the parameters closely, to really see how well they compare, let's make some predictions using both. We'll use posterior decoding to obtain a distribution over states at each time step since this will give us a more nuanced picture of how each model interprets the data.

\begin{NotebookIn}
xs, ys = data[0]
fbs = model.forward_backward(ys)
fbs_hat = model_hat.forward_backward(ys)

fig, axs = plt.subplots(4, 1, figsize=(6.4, 9.6), sharex=True)

axs[0].plot(ys)
for state in t_dists:
    axs[1].plot([x == state for x in xs], label=state)
for state, line in sorted(fbs.items()):
    axs[2].plot(line, label=state)
for state, line in sorted(fbs_hat.items()):
    axs[3].plot(line, label=state)
axs[3].set_xlabel('Time step')
axs[0].set_ylabel('Emission')
axs[1].set_ylabel('Label')
axs[2].set_ylabel('Probability')
axs[3].set_ylabel('Probability')
axs[1].legend(title='true states', **legend_kwargs)
axs[2].legend(title='model', **legend_kwargs)
axs[3].legend(title='model_hat', **legend_kwargs);
\end{NotebookIn}

\begin{NotebookImage}
\includegraphics[width=0.75\textwidth]{figures/out/categorical.png}
\end{NotebookImage}

The posterior decoding curves are effectively identical. There are, of course, some minor differences but no outstanding patterns. Generally these models are good at detecting the time steps corresponding to state 1 but much worse at distinguishing between states 2 and 3. We can understand this qualitatively by examining the original model parameters. State 1 is highly ``sticky'' and only emits As, so long runs of As are extremely likely to correspond to state 1. State 3 only emits Bs, but it's fairly likely to switch to state 2. Since state 2 is equally likely to emit an A as a B, it's difficult to know if a B was emitted because the model remained in state 3 or because it switched to state 2. This shows that state inference can be highly variable even if the parameters are known exactly.

\subsubsection{Discrete emission distribution}

For the next example, we'll derive and implement the MLE for the parameter of a \textit{Poisson distribution}. Poisson distributions are commonly used to model count data with no upper bound. The underlying assumptions are the counts represent events that occur with some average rate over time and the number of events in one interval is independent of the number of events in any other non-overlapping interval. These conditions impose few restrictions, so Poisson distributions are used to model a variety of phenomena, ranging from the number of particle decays in a radioactive sample to the number of requests arriving at a web server.

If the events of a Poisson process $X$ occur at an average rate of $\lambda$, then the probability that $x$ events are observed in an interval of length $t$ is given by

\begin{align*}
P(X = x | \lambda, t)
&= f(x, \lambda, t) \\
&= e^{-\lambda t} \frac{(\lambda t)^x}{x!}. \\
\end{align*}

The rate $\lambda$ and length of the interval $t$ always appear as the product $\lambda t$, so Poisson distributions are often parameterized in terms of $\lambda$ only. Confusingly, this is frequently still called a rate even though it's actually a unitless quantity. To match these conventions, we'll also drop $t$ as a parameter, although we'll avoid referring to $\lambda$ as a rate.

We can now write the log-likelihood function explicitly:

\begin{align*}
L(\lambda)
&= \sum_{i=1}^N \log f(x_i, \lambda) \\
&= \sum_{i=1}^N \log \left( e^{-\lambda} \frac{\lambda^x_i}{x_i!} \right) \\
&= \sum_{i=1}^N -\lambda + x_i \log \lambda - \log (x_i!) \\
&= -N\lambda + \sum_{i=1}^N x_i \log \lambda - \log (x_i!). \\
\end{align*}

To find the MLE for $\lambda$, we 1) take the derivative relative to $\lambda$ and 2) solve for $\lambda$ when this expression is zero. Since in the second step we solve for specific values where the derivative is zero, we replace $\lambda$ with $\hat{\lambda}$ to clarify this distinction.

\textbf{Step 1: Differentiate the log-likelihood function}
\begin{align*}
\frac{dL(\lambda)}{d\lambda}
&= -N + \sum_{i=1}^N \frac{x_i}{\lambda} \\
\end{align*}

\textbf{Step 2: Solve for the MLE}
\begin{align*}
0 &= -N + \sum_{i=1}^N \frac{x_i}{\hat{\lambda}} \\
N &= \frac{\sum_{i=1}^N x_i}{\hat{\lambda}} \\
\hat{\lambda} &= \frac{\sum_{i=1}^N x_i}{N} = \bar{x} \\
\end{align*}

Pleasingly, the MLE for $\lambda$ is the average of the observations. Now let's implement this in code using a model with two arbitrary states. This will follow the same format as the previous section, so we'll proceed with little comment.

\begin{NotebookIn}
t_dists = {1: {1: 0.95, 2: 0.05},
           2: {1: 0.25, 2: 0.75}}
e_dists = {1: stats.poisson(3),
           2: stats.poisson(0.5)}
start_dist = {1: 0.5, 2: 0.5}

model = HMM(t_dists=t_dists, e_dists=e_dists, start_dist=start_dist)

data = [model.simulate(200, random_state=i) for i in range(10)]
data = [list(zip(*example)) for example in data]
\end{NotebookIn}

The estimate for each state's $\lambda$ is simply the average of the emissions associated with that state. However, the emissions are separated across multiple examples and not organized by state, so we'll first gather them in a dictionary keyed by state and then take the average.

\begin{NotebookIn}
# Make emission dicts keyed by state
state2emits = {}
for state in t_dists:
    state2emits[state] = []

# Add emissions
for example in data:
    xs, ys = example
    for state, emit in zip(xs, ys):
        state2emits[state].append(emit)

# Average emissions
lambda_hats = {}
for state, emits in state2emits.items():
    lambda_hat = sum(emits) / len(emits)
    lambda_hats[state] = lambda_hat
lambda_hats
\end{NotebookIn}

\begin{NotebookOut}
{1: 2.991077119184194, 2: 0.46635730858468677}
\end{NotebookOut}

Though this form of the parameter estimates is convenient for inspection, discrete emission distributions over an infinite domain are implemented as SciPy random variables in Homomorph.

\begin{NotebookIn}
e_dists_hat = {}
for state, lambda_hat in lambda_hats.items():
    e_dists_hat[state] = stats.poisson(lambda_hat)
\end{NotebookIn}

The estimators for the transition and start distributions are the same, so we can copy those cells from the previous example.

\begin{NotebookIn}
# Make transition count dicts and add pseudocounts
t_pseudo = 0.1
t_counts = {}
for state1, t_dist in t_dists.items():
    t_count = {}
    for state2 in t_dist:
        t_count[state2] = t_pseudo
    t_counts[state1] = t_count

# Add observed counts
for example in data:
    xs, ys = example
    state0 = xs[0]
    for state1 in xs[1:]:
        t_counts[state0][state1] += 1
        state0 = state1

# Normalize counts
t_dists_hat = {}
for state1, t_count in t_counts.items():
    t_sum = sum(t_count.values())
    t_dist_hat = {}
    for state2, count in t_count.items():
        t_dist_hat[state2] = count / t_sum
    t_dists_hat[state1] = t_dist_hat
t_dists_hat
\end{NotebookIn}

\begin{NotebookOut}
{1: {1: 0.9487589559877175, 2: 0.0512410440122825},
 2: {1: 0.19452247191011232, 2: 0.8054775280898876}}
\end{NotebookOut}

\begin{NotebookIn}
# Make start count dicts and add pseudocounts
start_pseudo = 0.1
start_count = {}
for state in start_dist:
    start_count[state] = start_pseudo

# Add observed counts
for example in data:
    xs, ys = example
    start_count[xs[0]] += 1

# Normalize counts
start_sum = sum(start_count.values())
start_dist_hat = {}
for state, count in start_count.items():
    start_dist_hat[state] = count / start_sum
start_dist_hat
\end{NotebookIn}

\begin{NotebookOut}
{1: 0.303921568627451, 2: 0.696078431372549}
\end{NotebookOut}

With the parameter estimates in hand, we can instantiate an estimated model and compare the decoded states to those from the actual model and the true states.

\begin{NotebookIn}
model_hat = HMM(t_dists=t_dists_hat,
                e_dists=e_dists_hat,
                start_dist=start_dist_hat)

xs, ys = data[0]
fbs = model.forward_backward(ys)
fbs_hat = model_hat.forward_backward(ys)

fig, axs = plt.subplots(4, 1, figsize=(6.4, 9.6), sharex=True)

axs[0].plot(ys)
for state in t_dists:
    axs[1].plot([x == state for x in xs], label=state)
for state, line in sorted(fbs.items()):
    axs[2].plot(line, label=state)
for state, line in sorted(fbs_hat.items()):
    axs[3].plot(line, label=state)
axs[3].set_xlabel('Time step')
axs[0].set_ylabel('Emission')
axs[1].set_ylabel('Label')
axs[2].set_ylabel('Probability')
axs[3].set_ylabel('Probability')
axs[1].legend(title='true states', **legend_kwargs)
axs[2].legend(title='model', **legend_kwargs)
axs[3].legend(title='model_hat', **legend_kwargs);
\end{NotebookIn}

\begin{NotebookImage}
\includegraphics[width=0.75\textwidth]{figures/out/poisson.png}
\end{NotebookImage}

\subsubsection{Continuous emission distribution}
The normal distribution, with its iconic bell-shaped density curve, is practically synonymous with statistics. This association is deserved as the normal distribution is the foundation of many significant results and methods in both mathematical and applied statistics. For our purposes, however, we only need to know that the normal distribution is a common and robust model for continuous measurements, so in this example we'll derive the MLEs for its parameters.

As seen in the previous example, once the MLEs are derived, the implementations are often straightforward translations from mathematical symbols to code. In fact, for many common distributions, including the normal distribution, the MLEs have closed-form expressions that can be interpreted in terms of familiar statistical quantities like the mean or variance. Thus, for this example we'll skip the implementations and only present the derivations.

Let's first review the probability density function of a normal distribution:

\begin{equation*}
f(x, \mu, \sigma^2) = \frac{1}{\sqrt{2\pi\sigma^2}} e^{-\frac{\left( x-\mu \right)^2}{2\sigma^2}}.
\end{equation*}

Though the expression may look intimidating, the important take-away is the distribution has two parameters $\mu$ and $\sigma^2$, which are equal to its mean and variance, respectively. In qualitative terms, $\mu$ controls the position of the peak of the curve and $\sigma^2$ controls its width.

We will now substitute the density into the log-likelihood function:

\begin{align*}
L(\mu, \sigma^2)
&= \sum_{i=1}^N \log f(x_i, \mu, \sigma^2) \\
&= \sum_{i=1}^N -\frac{1}{2}\log(2\pi) - \frac{1}{2}\log(\sigma^2) - \frac{\left( x_i-\mu \right)^2}{2\sigma^2} \\
&= -\frac{N}{2}\log(2\pi) - \frac{N}{2}\log(\sigma^2) - \sum_{i=1}^N \frac{\left( x_i-\mu \right)^2}{2\sigma^2}. \\
\end{align*}

As the density function has two parameters, $\mu$ and $\sigma^2$, the log-likelihood is a function of these variables. This changes the next step slightly from the previous example since we have to take the partial derivative relative to each and solve the resulting system of equations. In this case the algebra works out nicely, but sometimes numerical optimization techniques are required.

We'll start with the partial derivative for $\mu$. A subtle point for distributions with multiple parameters is in the second step when the derivative is set to zero, we replace each parameter with its estimated counterpart since we're solving for specific points where the derivatives relative to each parameter are simultaneously zero.

\textbf{Step 1: Differentiate the log-likelihood function}
\begin{align*}
\frac{\partial L}{\partial \mu}
&= \sum_{i=1}^N \frac{\left( x_i- \mu \right)}{\sigma^2} \\
\end{align*}

\textbf{Step 2: Solve for the MLE}
\begin{align*}
0 &= \sum_{i=1}^N \frac{\left( x_i- \hat{\mu} \right)}{\hat{\sigma}^2} \\
N\hat{\mu} &= \sum_{i=1}^N x_i \\
\hat{\mu} &= \frac{\sum_{i=1}^N x_i}{N} = \bar{x} \\
\end{align*}

Now let's find the MLE for $\sigma^2$. Another nuance for the normal distribution in particular is the derivative is taken relative to the variance, $\sigma^2$, and not the standard deviation, $\sigma$. Though this choice does not impact the resulting formula, it both simplifies the calculation and reflects the natural role of the variance as the more ``fundamental'' statistical quantity.

\textbf{Step 1: Differentiate the log-likelihood function}
\begin{align*}
\frac{\partial L}{\partial \sigma^2}
&= -\frac{N}{2\sigma}^2 + \sum_{i=1}^N \frac{\left( x_i- \mu \right)^2}{2\sigma^4} \\
\end{align*}

\textbf{Step 2: Solve for the MLE}
\begin{align*}
0 &= -\frac{N}{2\hat{\sigma}^2} + \sum_{i=1}^N \frac{\left( x_i- \hat{\mu} \right)^2}{2\hat{\sigma}^4} \\
\frac{N}{\hat{\sigma}^2} &= \sum_{i=1}^N \frac{\left( x_i- \hat{\mu} \right)^2}{\hat{\sigma}^4} \\
{\hat{\sigma}^2} &= \frac{\sum_{i=1}^N \left( x_i- \hat{\mu} \right)^2}{N} \\
\end{align*}

As hinted in the introduction for this section, the MLEs for $\mu$ and $\sigma^2$ are simply the mean and variance of the data! Notice the MLE for $\sigma^2$ includes $\hat{\mu}$, so in practice we would calculate $\hat{\mu}$ first and substitute that value in the expression for $\hat{\sigma}^2$. Those with some background in statistics might notice the formula for $\hat{\sigma}^2$ has a factor of $N$ rather than $N-1$ in the denominator. This is no typo. It turns out that while MLEs are optimal in many ways, they are not always unbiased, meaning they can on average be higher or lower than the true value. In this case, the MLE for $\sigma^2$ is low by a factor of exactly $\frac{N-1}{N}$ on average, so some formulas divide by this quantity to remove the bias. This actually makes our estimate more imprecise, so there are some trade-offs involved in using one formula over the other. The good news is the difference is trivial for most data sets of realistic size, so the choice is largely inconsequential.

\section{Training with unknown states}

\subsection{Estimation-maximization, informally}

When we knew the states, we could assign each emission to a log-likelihood function and optimize the parameters of these functions separately. However, when the states are unknown, the problem of estimating parameters is much harder because we can't decompose the log-likelihood into independent terms. (We'll see this formally in the next section.) Not all hope is lost though. Remember that we can estimate parameters when we know the states, and when we know parameters we can estimate state probabilities. While this may seem like a chicken and egg problem, we can use this relationship as an iterative method for estimating parameters when the states are unknown. In practice, it works as the following:

\begin{enumerate}
\item Make an informed guess of the initial parameters.
\item Use the current parameters to estimate posterior state probabilities.
\item Use the posterior state probabilities to improve the parameter estimates.
\item Repeat steps 2 and 3 a fixed number of times or until some convergence criteria is met.
\end{enumerate}

In the context of HMMs, this procedure is known as the \textit{Baum-Welch algorithm}, but it's a special case of a more general technique called \textit{estimation-maximization} (EM). Though it intuitively makes sense, it's not at all clear that it will work in practice. For example, the likelihood function could ping-pong up and down without ever settling down to a single value. Fortunately, each iteration is guaranteed to improve the likelihood. Unfortunately, the sequence may converge to a local rather than a global maximum, \textit{i.e.} a good answer but not necessarily the best. Thus, it's common to run the algorithm multiple times with different initial parameters and choose the final parameters as those with the largest likelihood.

Hopefully this discussion has given a conceptual overview of the Baum-Welch algorithm. Clearly we've skipped many details, so in the next section we'll implement it from scratch for categorical distributions. Then in the following sections, we'll introduce the EM algorithm formally and use it to derive the update equations for a normal distribution.

\subsection{Implementing Baum-Welch for categorical distributions}

The steps in the Baum-Welch algorithm as presented above should at this point make sense except step 3. How exactly do we use posterior decoding to improve the parameter estimates? For a generic distribution, we'll need the EM formalism to derive the update equations properly, but for a categorical distribution we can intuit our way to the answer. We'll start with the transition distributions since we've already seen their MLEs, but then we'll derive a similar result for the emission distributions. Recall that the MLEs for the parameters of a transition distribution are written in terms of the number of observed transitions between states $i$ and $j$, $n_{ij}$:

\begin{equation*}
\hat{a}_{ij} = \frac{n_{ij}}{\sum_{j=1}^S n_{ij}}.
\end{equation*}

Unfortunately, we don't have access to these counts since the states are unknown. However, what if we replaced these counts with how often we thought they happened under our current parameter estimates? For example, if between time $t$ and $t+1$, we calculate there's a 75\% chance the process remained in state 1 and a 25\% chance it transitioned to state 2, that's effectively a 0.75 count towards $n_{11}$ and a 0.25 count towards $n_{12}$. Although we originally defined the $n_{ij}$ variables in terms of whole numbers, the equations still yield valid estimates with fractional counts.

Now the name of the game is to calculate these probabilities over all time steps. Formally we're looking for:

% In alignment environments, the \left and \right commands must be balanced on each line and on the same side of &
% To split lines with these commands, use invisible brackets with \left. and \right. to balance each line
% Aligning to the right of = can cause kerning problems, so explicit spaces are used
\begin{align*}
n_{ij}
=& \ \sum_{t=1}^{T-1} P(X_t=i, X_{t+1}=j|Y=y, \Theta) \\
=& \ \frac{1}{P(Y=y|\Theta)}
   \sum_{t=1}^{T-1} P(X_t=i, X_{t+1}=j, Y=y|\Theta) \\
=& \ \frac{1}{P(Y=y|\Theta)}
   \sum_{t=1}^{T-1} \Bigl[ P(X_t=i, Y_1=y_1, \ldots, Y_t=y_t|\Theta) \Bigr. \\
 & \times \Bigl. P(X_{t+1}=j, Y_{t+1}=y_{t+1}, \ldots, Y_T=y_T|X_t=i, Y_1=y_1, \ldots, Y_t=y_t, \Theta) \Bigr] \\
=& \ \frac{1}{P(Y=y|\Theta)}
   \sum_{t=1}^{T-1} \Bigl[ P(X_t=i, Y_1=y_1, \ldots, Y_t=y_t|\Theta) \Bigr. \\
 & \times \Bigl. P(X_{t+1}=j, Y_{t+1}=y_{t+1}, \ldots, Y_T=y_T|X_t=i, \Theta) \Bigr]. \\
\end{align*}

The first term is the forward variable evaluated at time $t$ and state $i$, which we'll denote by $f_i(t)$. We haven't formally introduced this quantity in this tutorial, but we can easily obtain these values with the \texttt{forward} method of an \texttt{HMM} instance. Incidentally, $P(Y=y|\Theta)=\sum_i f_i(T)$ since this sums the probability of the entire emission sequence over all possible final states. Let's now focus on the second term in the sum and define the following events to clean up the notation:

\begin{itemize}
\item $A: Y_{t+2}=y_{t+2}, \ldots, Y_T=y_t$
\item $B: Y_{t+1}=y_{t+1}$
\item $C: X_{t+1}=j$
\item $D: X_t=i$.
\end{itemize}

Thus, we have

\begin{align*}
P(X_{t+1}=j, Y_{t+1}=y_{t+1}, \ldots, Y_T=y_T|X_t=i, \Theta)
&= P(A, B, C|D, \Theta) \\
&= P(A|B, C, D, \Theta) \cdot P(B|C, D, \Theta) \cdot P(C|D, \Theta) \\
&= P(A|C, \Theta) \cdot P(B|C, \Theta) \cdot P(C|D, \Theta). \\
\end{align*}

From right to left, the third term is $a_{ij}$, the second term is $e_j(y_{t+1})$, and the first term is the backward variable evaluated at time $t+1$ and state $j$, which we'll denote by $b_j(t+1)$. Analogous to the forward variable, the backward variable is available via the \texttt{backward} method of an \texttt{HMM} instance.

Putting everything together we have

\begin{equation*}
n_{ij} = \frac{1}{P(Y=y|\Theta)} \sum_{t=1}^{T-1} f_i(t)a_{ij}e_j(y_{t+1})b_j(t+1).
\end{equation*}

Now let's tackle the emission distributions. When the states are known, the MLEs are the same, but instead of counting transitions between states, we count emissions. (The emissions essentially take on the role of the ``target'' state $j$.) When the states are unknown, we replace these counts with how much we think each state is responsible for each emission. In other words, if at time $t$ we observe emission $i$ and we calculate an 90\% chance of state 1 and a 10\% chance of state 2, we assign a 0.9 count towards $n_{1i}$ and a 0.1 count towards $n_{2i}$. Written mathematically,

\begin{align*}
n_{ij}
&= \sum_{t: y_t=j} P(X_t=i|Y=y, \Theta) \\
&= \frac{1}{P(Y=y|\Theta)}
   \sum_{t: y_t=j} P(X_t=i, Y=y|\Theta) \\
&= \frac{1}{P(Y=y|\Theta)}
   \sum_{t: y_t=j} P(Y_1=y_1, \ldots, Y_t=y_t, X_t=i|\Theta)
                   P(Y_{t+1}=y_{t+1}, \ldots, Y_T=y_T| X_t=i, \Theta) \\
&= \frac{1}{P(Y=y|\Theta)}
   \sum_{t: y_t=j} f_i(t)b_i(t). \\
\end{align*}

With these results, we can implement the Baum-Welch algorithm. For continuity, we'll use the same example from the MLE section.

\begin{NotebookIn}
t_dists = {1: {1: 0.95, 2: 0.05},
           2: {1: 0.05, 2: 0.9, 3: 0.05},
           3: {2: 0.35, 3: 0.65}}
e_dists = {1: {'A': 1},
           2: {'A': 0.5, 'B': 0.5},
           3: {'B': 1}}
start_dist = {1: 0.2, 2: 0.5, 3: 0.3}

model = HMM(t_dists=t_dists, e_dists=e_dists, start_dist=start_dist)

data = [model.simulate(200, random_state=i) for i in range(10)]
data = [list(zip(*example)) for example in data]
\end{NotebookIn}

To start the algorithm, we first need a model structure and initial estimates for its parameters. To keep the Baum-Welch approach on equal footing with the previous example, we'll assume we know the disallowed transitions but not the disallowed emissions. For the initial estimates, we'll use uniform random values to keep the code simple, but more sophisticated approaches can use specific distributions for each parameter. Though it's good practice to use the best result from several different random initializations, if the initial parameters are hard coded, the states must be different from each other in some way. If the states are all identical, there's no way for the model to ``break symmetry,'' and all the updates will yield parameters that are the same as the initial ones.

\begin{NotebookIn}
random.seed(1)

# Make transition count dicts and add pseudocounts
t_counts = {}
for state1, t_dist in t_dists.items():
    t_count = {}
    for state2 in t_dist:
        t_count[state2] = random.random()
    t_counts[state1] = t_count

# Normalize counts
t_dists_hat = {}
for state1, t_count in t_counts.items():
    t_sum = sum(t_count.values())
    t_dist_hat = {}
    for state2, count in t_count.items():
        t_dist_hat[state2] = count / t_sum
    t_dists_hat[state1] = t_dist_hat
t_dists_hat
\end{NotebookIn}

\begin{NotebookOut}
{1: {1: 0.13685528663315571,
     2: 0.8631447133668443},
 2: {1: 0.5043817911017634,
     2: 0.16844258615115162,
     3: 0.3271756227470851},
 3: {2: 0.4082259386891194,
     3: 0.5917740613108805}}
\end{NotebookOut}

\begin{NotebookIn}
random.seed(2)

# Collect all possible emissions
e_set = set()
for example in data:
    xs, ys = example
    e_set.update(ys)

# Make emission count dicts and add pseudocounts
e_counts = {}
for state in t_dists:
    e_counts[state] = {emit: random.random() for emit in e_set}

# Normalize counts
e_dists_hat = {}
for state, e_count in e_counts.items():
    e_sum = sum(e_count.values())
    e_dist_hat = {}
    for emit, count in e_count.items():
        e_dist_hat[emit] = count / e_sum
    e_dists_hat[state] = e_dist_hat
e_dists_hat
\end{NotebookIn}

\begin{NotebookOut}
{1: {'A': 0.5021552995618842, 'B': 0.49784470043811585},
 2: {'A': 0.39987288219487493, 'B': 0.6001271178051251},
 3: {'A': 0.531667470841593, 'B': 0.4683325291584069}}
\end{NotebookOut}

\begin{NotebookIn}
random.seed(3)

# Make start count dicts and add pseudocounts
start_count = {}
for state in start_dist:
    start_count[state] = random.random()

# Normalize counts
start_sum = sum(start_count.values())
start_dist_hat = {}
for state, count in start_count.items():
    start_dist_hat[state] = count / start_sum
start_dist_hat
\end{NotebookIn}

\begin{NotebookOut}
{1: 0.2065397992628295, 2: 0.47236009956297115, 3: 0.32110010117419935}
\end{NotebookOut}

Let's now code the main loop, beginning with the definition of the stopping conditions. We can operationally define convergence as when the improvement in the log-likelihood is below some threshold, \texttt{epsilon}. In practice, this may take too long, so we'll also define a maximum number of iterations, \texttt{maxiter}. What follows, then, is a fairly straightforward implementation of the equations we derived previously. There are two nuanced points, however. First, to preserve the model structure, the counts are always taken over the allowed transitions or emissions. (In theory, disallowed transitions or emissions should always have a probability of zero, but floating point errors may yield unexpected results.) Second, the forward and backward variables are scaled to sum to one at each time step for numerical stability. Thus, the true value at time $t$ is a product of the raw value and all scaling factors up to and including time $t$. (Since the backward variable is calculated recursively from the final instead of the first observation, the product is taken from $t$ to $T$.) A side effect of this representation is $P(Y=y|\Theta)$ is simply the product of all scaling factors.

For brevity, the input code is given in Appendix~\ref{appendix:b}, and only the final values are shown below.

\begin{NotebookOut}
FINAL VALUES
log-likelihood: -906.2061302148018
delta log-likelihood: 0.009596164686740849
t_dists: {1: {1: 0.19048033686643145,
              2: 0.8095196631335685},
          2: {1: 0.730991899350023,
              2: 0.168079808479332,
              3: 0.10092829217064504},
          3: {2: 0.04584548747372808,
              3: 0.9541545125262719}}
e_dists: {1: {'A': 0.28936393517521436,
              'B': 0.7106360648247857},
          2: {'A': 0.48150408200102607,
              'B': 0.5184959179989739},
          3: {'A': 0.9999943519811173,
              'B': 5.648018882737235e-06}}
start_dist: {1: 0.7875017247391405,
             2: 0.005518967674210817,
             3: 0.20697930758664865}
\end{NotebookOut}

Based on the results, the algorithm seems to have merged states 2 and 3 into a single state that emits a mixture of A and B. State 1 was largely estimated correctly, but it's labeled as state 3 in the estimated model. There are two factors which may explain this result. First, the initial values were chosen poorly. We generated them randomly, so we were potentially unlucky, and there was a ``bad'' local maximum near those initial values. Additionally, since we didn't prime the initial values with any information that state 1 only emits A and state 3 only emits B, the algorithm exchanged them. For both these reasons, it's best practice to use multiple initializations that are random but still encode the expected behavior for each state. Try playing around with different random seeds or initialization schemes to see if the fit improves!

The second reason why the algorithm merged states 2 and 3 is state 2 is inherently difficult to fit. Under the current parameters, state 2 is just as likely to emit an A as a B, which maximizes the uncertainty associated with its emissions. (Compare this to states 1 or 3 whose emissions we know with certainty.) While this example is highly artificial, it illustrates that although we can define an HMM however we like, its emission distribution may be well-described by multiple sets of parameters, \textit{i.e.}, the likelihood surface is broad and flat or has multiple peaks.

\subsection{Estimation-maximization, formally}

Though what we've done so far makes sense intuitively, let's look at what's happening more formally in order to generalize this approach to other emission distributions. Recall that in maximum likelihood estimation, we find parameters $\hat{\Theta}$ that maximize the logarithm of probability of the data, $D$:

\begin{align*}
\hat{\Theta}
&= \argmax_\Theta L(\Theta) \\
&= \argmax_\Theta \log P(D|\Theta). \\
\end{align*}

In the HMM setting with known states, the data are composed of $N$ examples of ordered pairs $\{(x_i, y_i): 1 \le i \le N\}$ where $x_i$ is a sequence of states and $y_i$ is a sequence of emissions, each with length $T_i$. For simplicity, however, we will only consider a single example $(x, y)$ where $x_t$ and $y_t$ indicate the $t$th state and emission in those sequences, respectively. The log-likelihood function is then written as $L(\Theta) = \log P(X=x, Y=y|\Theta)$. So far this is all review from the previous section. What about the case when we only know the emissions, $y$? Then the log-likelihood is $L(\Theta) = \log P(Y=y|\Theta)$. Though we can calculate this quantity directly with the forward algorithm and thus in principle apply a number of optimization algorithms, we instead will take a probabilistic approach and write the log-likelihood function as a sum over all possible state sequences:

\begin{align*}
L(\Theta)
&= \log P(Y=y|\Theta) \\
&= \log \sum_x P(Y=y, X=x|\Theta). \\
\end{align*}

The only problem is that for an emission sequence of length $T$ and a model with $S$ hidden states, there are $S^T$ possible state sequences. This sum therefore has an exponential number of terms and is intractable for any data of realistic size.

Thus, this expression hasn't gotten us anywhere yet. However, it's possible to derive a related but more tractable quantity whose improvements lower bound improvements for the log-likelihood. This quantity, commonly denoted as $Q(\Theta|\Theta_n)$, is defined as the expectation of the log-likelihood function with respect to the conditional distribution of the states, $X$, given the emissions, $y$:

\begin{align*}
Q(\Theta|\Theta_n)
&= E_{X|Y=y,\Theta_n} \log P(Y=y, X|\Theta) \\
&= \sum_x P(X=x|Y=y, \Theta_n) \log P(Y=y, X=x|\Theta). \\
\end{align*}

This is a confusing expression, especially since the notation for conditional expectation is dense. However, in the second line the expectation is written explicitly, showing it is simply a sum of the joint log-likelihoods of the states and emissions given the parameters where each is weighted by the quantity $P(X=x|Y=y, \Theta_n)$. Notice that $P(X=x|Y=y, \Theta_n)$ is constant for a given $\Theta_n$, so $Q(\Theta|\Theta_n)$ is function only of the log-likelihood terms.

Next we define the new estimates of the parameters as

\begin{equation*}
\Theta_{n+1} = \argmax_\Theta Q(\Theta|\Theta_n).
\end{equation*}

These two steps are where the name expectation-maximization is derived since the first is the calculation of an expectation, and the second is a maximization of that expectation. It may seem like we haven't done much to simplify the problem since we're still dealing with sums over all state sequences. However, by moving the sum outside the logarithm, we've greatly simplified the maximization step as we will see in the next section.

By repeating this process, we can iteratively improve our estimates of the parameters. As mentioned earlier, these improvements in $Q(\Theta|\Theta_n)$ lower bound improvements in the log-likelihood. We won't show the derivation here, but formally this means

\begin{equation*}
\log P(Y|\Theta) - \log P(Y|\Theta_n) \ge Q(\Theta|\Theta_n) - Q(\Theta_n|\Theta_n).
\end{equation*}

Since in each step $Q(\Theta|\Theta_n)$ is maximized, the right quantity is always non-negative. Thus, the log-likelihood of the emissions is improved by at least that much as well. Written mathematically,

\begin{equation*}
Q(\Theta_{n+1}|\Theta_n) - Q(\Theta_n|\Theta_n) \ge 0
\implies
\log P(Y|\Theta_{n+1}) - \log P(Y|\Theta_n) \ge 0.
\end{equation*}

\subsection{Deriving the update equations for a normal distribution}

We'll now apply this theory to the HMM context. First we'll show not only that $Q(\Theta|\Theta_n)$ is tractable to calculate but also that it decomposes into independent terms. Then we'll derive the update equations for normal emission distributions. Let's get started!

We first expand the expression for $Q(\Theta|\Theta_n)$ using the probabilistic structure of an HMM:

\begin{align*}
Q(\Theta|\Theta_n)
=& \sum_x P(X=x|Y=y, \Theta_n) \log P(Y=y, X=x|\Theta) \\
=& \sum_x P(X=x|Y=y, \Theta_n)
          \log \biggl[P(Y_1=y_1|X_1=x_1, \Theta)P(X_1=x_1|\Theta) \biggr. \\
                     & \biggl. \prod_{t=1}^{T-1} P(Y_{t+1}=y_{t+1}|X_{t+1}=x_{t+1}, \Theta) P(X_{t+1}=x_{t+1}|X_t=x_t, \Theta)
               \biggr] \\
=& \sum_x P(X=x|Y=y, \Theta_n)
          \biggl[\log P(X_1=x_1|\Theta) \biggr. \\
                & + \biggl. \sum_{t=1}^T \log P(Y_t=y_t|X_t=x_t, \Theta)
                + \sum_{t=1}^{T-1} \log P(X_{t+1}=x_{t+1}|X_t=x_t, \Theta)
          \biggr] \\
=& \sum_i P(X_1=i|Y=y, \Theta_n) \log P(X_1=i|\Theta) \\
 & + \sum_i \sum_{t=1}^T P(X_t=i|Y=y, \Theta_n) \log P(Y_t=y_t|X_t=i, \Theta) \\
 & + \sum_{i, j} \sum_{t=1}^{T-1} P(X_t=i, X_{t+1}=j|Y=y, \Theta_n) \log P(X_{t+1}=j|X_t=i, \Theta). \\
\end{align*}

In the final step, the outer sum was distributed over each term and exchanged with the inner sum. This step is shown explicitly for the second term below:

\begin{align*}
& \sum_x P(X=x|Y=y, \Theta_n) \sum_{t=1}^T \log P(Y_t=y_t|X_t=x_t, \Theta) \\
&= \sum_x \sum_{t=1}^T P(X=x|Y=y, \Theta_n) \log P(Y_t=y_t|X_t=x_t, \Theta) \\
&= \sum_i \sum_{x \setminus x_t} \sum_{t=1}^T P(X=x|Y=y, \Theta_n) \log P(Y_t=y_t|X_t=i, \Theta) \\
&= \sum_i \sum_{t=1}^T \log P(Y_t=y_t|X_t=i, \Theta) \sum_{x \setminus x_t} P(X=x|Y=y, \Theta_n) \\
&= \sum_i \sum_{t=1}^T P(X_t=i|Y=y, \Theta_n) \log P(Y_t=y_t|X_t=i, \Theta). \\
\end{align*}

(The final line used that the probability of a state sequence with a fixed state $X_t=i$ is the sum of the probabilities of all state sequences that include that fixed state.)

Though moving the sum over all state sequences outside the logarithm seemed like a minor mathematical detail, it, in combination with the probabilistic structure of an HMM, makes its computation tractable. Another consequence is that if all transitions and emissions are governed by different parameters, we can maximize the entire expression by optimizing the corresponding sum for each state or pair of states individually. The fractional ``counts'' $P(X_t=i|Y=y, \Theta)$ and $P(X_t=i, X_{t+1}=j|Y=y, \Theta)$ defined in the previous section are even the coefficients of the log-likelihood terms, which is directly related to their appearance in the update equations.

Now that we've demonstrated that each term can be optimized separately, we'll derive the update equations for a normal emission distribution. We begin by writing the term we wish to optimize for a fixed state $i$, which we call $q(\Theta)$:

\begin{equation*}
q(\Theta) = \sum_{t=1}^T P(X_t=i|Y=y, \Theta_n) \log P(Y_t=y_t|X_t=i, \Theta).
\end{equation*}

We'll write $w_{it} = P(X_t=i|Y=y, \Theta_n)$ to simplify the notation and to emphasize this quantity is fixed during the maximization step. We'll also substitute the expression for a normal density with mean $\mu_i$ and variance $\sigma_i^2$:

\begin{align*}
q(\mu_i, \sigma_i^2)
&= \sum_{t=1}^T P(X_t=i|Y=y, \Theta_n) \log P(Y_t=y_t|X_t=i, \Theta) \\
&= \sum_{t=1}^T w_{it} \log \left(
                                  \frac{1}{\sqrt{2\pi\sigma^2}} e^{-\frac{\left( y-\mu \right)^2}{2\sigma^2}}
                             \right) \\
&= \sum_{t=1}^T w_{it} \left(
                             -\frac{1}{2}\log \sigma^2
                             -\frac{1}{2}\log 2\pi
                             -\frac{\left( y_t - \mu_i \right)^2}{2\sigma^2}
                       \right). \\
\end{align*}

This quantity should look extremely familiar from the section on deriving the MLEs for a normal distribution! However, let's bring it home and take some derivatives. We'll start with the parameter for the mean.

\textbf{Step 1: Differentiate the log-likelihood function}
\begin{equation*}
\frac{\partial q}{\partial \mu_i} = \sum_t^T w_{it} \frac{(y_t-\mu_i)}{\sigma_i^2}
\end{equation*}

\textbf{Step 2: Solve for the MLE}
\begin{align*}
0 &= \sum_{t=1}^T w_{it} \frac{(y_t-\hat{\mu}_i)}{\hat{\sigma}_i^2} \\
0 &= \sum_{t=1}^T w_{it} (y_t-\hat{\mu}_i) \\
\sum_{t=1}^T w_{it} \hat{\mu}_i &= \sum_{t=1}^T w_{it} y_t \\
\hat{\mu}_i &= \frac{\sum_{t=1}^T w_{it} y_t}{\sum_{t=1}^T w_{it}} \\
\end{align*}

The resulting formula is very similar to the corresponding MLE derived in the previous section. However, here each observation is multiplied by a weight, and the sum is divided by the total weight rather than the number of observations. (In fact, setting the weight for each observation to one recovers the original formula!) Since the weights for each observation must sum to one, we've essentially partitioned each observation across all states where the amount that an observation contributes to a state is given by our confidence that it was generated by that state.

Let's now derive the update equation for the variance.

\textbf{Step 1: Differentiate the log-likelihood function}
\begin{equation*}
\frac{\partial q}{\partial \sigma_i^2} = \sum_{t=1}^T w_{it}
                                         \left( -\frac{1}{2 \sigma_i^2}
                                                + \frac{\left( y_t - \mu_i \right)^2}{2 \sigma_i^4}
                                         \right)
\end{equation*}

\textbf{Step 2: Solve for the MLE}
\begin{align*}
0 &= \sum_{t=1}^T w_{it}
     \left( -\frac{1}{2 \hat{\sigma}_i^2}
            + \frac{\left( y_t - \hat{\mu}_i \right)^2}{2 \hat{\sigma}_i^4}
     \right) \\
\frac{1}{\hat{\sigma}_i^2} \sum_{t=1}^T w_{it} &= \frac{1}{\hat{\sigma}_i^4}
                                                  \sum_{t=1}^T w_{it} \left( y_t - \hat{\mu}_i \right)^2 \\
\hat{\sigma}_i^2 &= \frac{\sum_{t=1}^T w_{it} \left( y_t - \hat{\mu}_i \right)^2}
                         {\sum_{t=1}^T w_{it}} \\
\end{align*}

Again, the update equation has the same form as the corresponding MLE except each term in the sum is weighted. This is a common occurrence when using the EM algorithm and is a consequence of the construction of the objective function as a weighted sum of log-likelihood functions. Thus, when the MLEs are interpretable as quantities like counts or averages, the corresponding update equations are often ``weighted'' versions of those estimators. Of course, it's not always obvious how the weights will come into play, especially for MLEs which don't have an obvious interpretation, so it's good idea to solve the equations or look up a reference rather than guessing.

\section{Discriminative training}

In this next section, we'll cover discriminative training, which is sometimes called conditional maximum likelihood (CML) estimation. Technically the discriminative training method falls under the heading of training with known states since it requires labeled data. However, there are enough differences in the objective function and the optimization algorithm that it deserves its own treatment. As before, the material is largely divided between a section that motivates the problem and develops the theory and a section that translates that theory into code.

\subsection{Discriminative training theory}

Previously, we saw that maximum likelihood estimation and the EM algorithm often obtained parameter estimates which were reasonably close to their true values. However, what if we don't really care about the parameters, and we're instead more interested in using an HMM to predict the hidden states? What if the model we've chosen for our data is wrong, or, even worse, the data are so messy that no tractable statistical model could ever hope to describe it?

The bad news is all these things are frequently true when working with HMMs. While sometimes the parameters are inherently meaningful, more often than not HMMs are a tool to label data. In these cases, it's more important to accurately predict the states than to estimate the underlying parameters. If the data are actually described by the model, this distinction isn't as relevant because with enough data the parameters that yield the most accurate predictions will converge to the true parameters. However, real-world data are almost never fully described by statistical models, and in these cases the parameters that best fit the data are often not those that best predict the states. Furthermore, the phrase ``with enough data'' is actually a sneaky way of saying ``as the number of observations approaches infinity,'' so for all practical purposes the two parameter sets are solving distinct problems.

Let's clarify these distinctions by writing them formally. Recall that in maximum likelihood estimation for HMMs with known states, we solve the following optimization:

\begin{align*}
\hat{\Theta}_{ML}
&= \argmax_\Theta L(\Theta) \\
&= \argmax_\Theta \log P(X=x, Y=y|\Theta) \\
\end{align*}

where $\Theta$ is the set of model parameters and $x$ and $y$ are sequences of states and emissions, respectively, with lengths $T$. In general, the optimization is conducted over a set of paired state and emission sequences, but since this won't change the key features of the following discussion, for simplicity we'll consider only a single state-emission sequence pair.

In contrast, conditional maximum likelihood solves the following optimization:

\begin{align*}
\hat{\Theta}_{CML}
&= \argmax_\Theta \log P(X=x|Y=y, \Theta) \\
&= \argmax_\Theta \log \frac{P(X=x, Y=y, \Theta)}{P(Y=y|\Theta)} \\
&= \argmax_\Theta \log P(X=x, Y=y, \Theta) - \log P(Y=y|\Theta). \\
\end{align*}

It's clear the ML and CML objectives are closely related, so let's use this relationship to gain a deeper understanding of how CML parameter estimates differ from their ML counterparts. To ease the following discussion, we'll make the substitutions

\begin{equation*}
\log P(X=x, Y=y|\Theta) = L_c(\Theta)
\end{equation*}

and

\begin{equation*}
\log P(Y=y|\Theta) = L_f(\Theta).
\end{equation*}

$L_c$ is the joint log-likelihood of the state and emission sequences, so the term is ``clamped'' by the true states. $L_f$ is the log-likelihood of the emission sequences only, so the term is ``free'' to account for all possible state sequences. To maximize the difference between $L_c$ and $L_f$, it seems the optimized parameters should make $L_c$ as large as possible and $L_f$ as small as possible. However, $L_f \ge L_c$ for any choice of $\Theta$ because $L_f$ includes the contributions of all possible state sequences. Thus, the best we can do is to make $L_c$ as close as possible to $L_f$, and this only happens if we minimize the contribution of state sequences which are not the true sequence. In other words, conditional maximum likelihood is maximum likelihood with an added penalty for parameter choices that favor alternate state sequences.

In the maximum likelihood case, we were able to decompose the optimization into independent products under some fairly general assumptions because when the state sequence was known the likelihood was a single product of transition and emission probabilities. Here, however, the free term, $L_f$, forces us to consider all possible state sequences. As noted in the previous section, the number of state sequences grows exponentially with length, so even if we derived a closed-form solution, we would still have to contend with an intractable number of terms. We can't apply the EM algorithm either because this optimization doesn't depend on any hidden variables.

We'll instead take a route distinct from all the approaches seen so far by using a method called gradient descent. Gradient descent is the workhorse for many modern machine learning techniques, so there are numerous online resources which explain it in detail. However, the basic idea is since the derivative of a function relative to some input is a measure of its response to a change in that input, we can use that information to iteratively improve an initial guess until we reach a minimum. For functions of two variables, the process is often visualized as progressively climbing down a mountain by taking small steps in the direction of steepest descent. The power of this technique is that it's much easier to take derivatives than it is to identify their zeros, so gradient descent can be applied to almost any optimization problem.

Thus, we need to compute derivatives of the conditional likelihood function. Because gradient descent is for finding minima, however, we redefine our maximization as a minimization by multiplying by negative one:

\begin{align*}
\hat{\Theta}_{CML}
&= \argmin_\Theta -\log P(X=x|Y=y, \Theta) \\
&= \argmin_\Theta \log P(Y=y|\Theta) - \log P(X=x, Y=y|\Theta) \\
&= \argmin_\Theta L_f(\Theta) - L_c(\Theta) \\
&= \argmin_\Theta L(\Theta). \\
\end{align*}

It's a small change, but when we write the update equations, we can use the conventional form instead of having to flip the sign of the derivatives.

We're now ready to take derivatives. We'll start with $L_f$ for a generic parameter $\theta \in \Theta$:

\begin{align*}
\frac{\partial L_f}{\partial \theta}
&= \frac{1}{P(Y=y|\Theta)} \frac{\partial P(Y=y|\Theta)}{\partial \theta} \\
&= \sum_x \frac{1}{P(Y=y|\Theta)} \frac{\partial P(X=x, Y=y|\Theta)}{\partial \theta} \\
&= \sum_x \frac{P(X=x, Y=y|\Theta)}{P(Y=y|\Theta)} \frac{\partial \log P(X=x, Y=y|\Theta)}{\partial \theta} \\
&= \sum_x P(X=x|Y=y, \Theta) \frac{\partial \log P(X=x, Y=y|\Theta)}{\partial \theta}. \\
\end{align*}

This expression is almost identical in structure to the function $Q(\Theta|\Theta_n)$ used in the EM algorithm. We can therefore use a similar series of tricks, namely expanding the log of a product as a sum of log terms and exchanging the order of summation, to obtain a tractable expression. To reduce the clutter, we'll also use the shorthand defined in the section on maximum likelihood estimation where $P(X_{t+1}=j|X_t=i) = a_{ij}$ and $e_i(y_t) = P(Y_t=y_t|X_t=i, \Theta)$. After all the algebraic dust settles, we obtain the equation

\begin{equation*}
\frac{\partial L_f}{\partial \theta} = \sum_{t,i} \frac{n_i(t)}{e_i(y_t)} \frac{\partial e_i(y_t)}{\partial \theta} + \sum_{t,i,j} \frac{n_{ij}(t)}{a_{ij}} \frac{\partial a_{ij}}{\partial \theta}.
\end{equation*}

The variables $n_i(t)$ and $n_{ij}(t)$ are defined as $P(X_t=i|Y=y, \Theta)$ and $P(X_t=i, X_{t+1}=j|Y=y, \Theta)$. Though this may look like new notation, they are nearly identical to quantities defined in the section on the EM algorithm and can be efficiently calculated with the forward and backward variables.

A similar computation shows the derivative of the clamped log-likelihood is

\begin{equation*}
\frac{\partial L_c}{\partial \theta} = \sum_{t,i} \frac{m_i(t)}{e_i(y_t)} \frac{\partial e_i(y_t)}{\partial \theta} + \sum_{t,i,j} \frac{m_{ij}(t)}{a_{ij}} \frac{\partial a_{ij}}{\partial \theta}.
\end{equation*}

$m_i(t)$ and $m_{ij}(t)$ are defined like $n_i(t)$ and $n_{ij}(t)$. However, because the clamped log-likelihood is taken only over the true state sequence, they are in practice indicators of the state or transition at each time step. This means they are one if a state $i$ or transition from state $i$ to state $j$ occurred at time $t$ and zero otherwise.

HMMs are typically parameterized directly in terms of their transition probabilities. In this case, the derivative of the loss function simplifies to

\begin{equation*}
\frac{\partial L}{\partial a_{ij}} = -\frac{m_{ij} - n_{ij}}{a_{ij}}
\end{equation*}

where $n_{ij}$ and $m_{ij}$ are sums over all time steps of the previously defined quantities. While this is the derivative we're seeking, there is one issue. For a generic parameter, gradient descent updates are written in the form

\begin{equation*}
\theta_{t+1} = \theta_t - \eta \frac{\partial L}{\partial \theta_t}
\end{equation*}

where $\eta$ is the \textit{learning rate}, which controls the size of the steps down the mountain. When $\eta$ is large, we make bold steps at the cost of sometimes going uphill if the loss landscape is bumpy. By making $\eta$ small, we can ensure we always go downhill but potentially at a glacial pace. Our problem is this update equation does not incorporate the constraints $a_{ij} \ge 0$ and $\sum_j a_{ij} = 1$. In other words, there's nothing stopping a gradient descent update from making a transition probability greater than one or, even worse, negative. If that happens, we've moved outside the parameter space of our model, so there's no guarantee the gradient descent updates would remain meaningful for our optimization even if every mathematical operation were defined.

This is a fundamental shortcoming of derivative-based methods, and there are many variants of gradient descent, \textit{e.g.} projected gradient descent, which address constrained optimization. Fortunately, in this case the constraints allow us to still use vanilla gradient descent via a clever variable transformation. We instead define

\begin{equation*}
a_{ij} = \frac{e^{z_{ij}}}{\sum_{j^\prime} e^{z_{ij^\prime}}}
\end{equation*}

and perform gradient descent on the auxiliary variables $z_{ij}$. (Note that the symbol $e$ without any subscripts refers to the constant and not an emission distribution.) The derivative relative to $z_{ij}$ is then given by

\begin{equation*}
\frac{\partial L}{z_{ij}} = -\left[
                                   m_{ij} - n_{ij}
                                   - a_{ij} \sum_{j^\prime} \left( m_{ij^\prime} - n_{ij^\prime} \right)
                             \right].
\end{equation*}

Further manipulations can yield an update equation that eliminates the auxiliary variables. However, in our implementation we'll work with the auxiliary variables directly, so this is sufficient for our needs.

\subsection{Deriving the update equations for normal distributions}

This is about as far as we can go without specifying the model further, so let's now consider the case where the emission distributions are normal and each state is governed by different parameters. If $f_i$ is the probability density function for state $i$ with mean $\mu_i$ and variance $\sigma_i^2$, then its derivative relative to $\mu_i$ is

\begin{align*}
\frac{\partial f_i(y)}{\partial \mu_i}
&= \frac{1}{\sqrt{2\pi\sigma_i^2}}
   e^{-\frac{\left( y-\mu_i \right)^2}{2\sigma_i^2}}
   \frac{y - \mu_i}{\sigma_i^2} \\
&= f_i(y) \frac{y - \mu_i}{\sigma_i^2}, \\
\end{align*}

and the derivative relative to $\sigma_i^2$ is

\begin{align*}
\frac{\partial f_i(y)}{\partial \sigma_i^2}
&= \frac{1}{\sqrt{2\pi}} \cdot -\frac{1}{2}
   \left( \sigma_i^2 \right)^{-\frac{3}{2}}
   e^{-\frac{\left( y-\mu_i \right)^2}{2\sigma_i^2}}
   +
   \frac{1}{\sqrt{2\pi}}
   \left( \sigma_i^2 \right)^{-\frac{1}{2}}
   e^{-\frac{\left( y-\mu_i \right)^2}{2\sigma_i^2}}
   \cdot -\frac{\left( y-\mu_i \right)^2}{2}
   \cdot -\left( \sigma_i^2 \right)^{-2} \\
&= \frac{f_i(y)}{2} \left[ \left( -\sigma_i^2 \right)^{-1}
                           + \left( y - \mu_i \right)^2 \left( \sigma_i^2 \right)^{-2}
                    \right]. \\
\end{align*}

Because the variance is non-negative, we define

\begin{equation*}
\sigma_i^2 = e^{z_i}
\end{equation*}

and perform gradient descent on the auxiliary variable $z_i$. The new derivative is then easily calculated as

\begin{equation*}
\frac{\partial f_i(y)}{\partial z_i} = \frac{\partial f_i(y)}{\partial \sigma_i^2}
                                        \frac{\partial \sigma_i^2}{\partial z_i}
\end{equation*}

using the chain rule.

The final step is to put all these pieces together by substituting the above expressions into the derivatives of the free and clamped log-likelihoods and taking their difference. For $\mu_i$, this is

\begin{equation*}
\frac{\partial L}{\partial \mu_i}
= -\left[
         \sum_t \left( m_i(t) - n_i(t) \right)
                \left( \frac{y_t - \mu_i}{\sigma_i^2} \right)
   \right].
\end{equation*}

\subsection{Labeling data with a misspecified model}

We're finally ready to see discriminative training in action! To illustrate the differences between maximum likelihood and conditional maximum likelihood, we'll examine a case of model misspecification. Specifically we'll fit normal distributions to data where the emissions of one state follow a gamma distribution. The gamma distribution is highly flexible, so its parameter \texttt{a} can dramatically impact the shape of its density curve. However, for many values of \texttt{a} the gamma distribution is roughly normal with a right skew. Thus, the goal of this example is to understand the impact of this skew on the parameter estimates and accuracy of state inference for each training method.

Let's start by creating the components of an HMM with two states and plotting the density functions of their emission distributions.

\begin{NotebookIn}
t_dists = {1: {1: 0.95, 2: 0.05},
           2: {1: 0.1, 2: 0.9}}
e_dists = {1: stats.norm(loc=-2.5, scale=25**0.5),  # Scale is std dev
           2: stats.gamma(a=1.5, scale=16**0.5)}
start_dist = {1: 0.5, 2: 0.5}

xs = np.linspace(-25, 25, 250)
ys1 = e_dists[1].pdf(xs)
ys2 = e_dists[2].pdf(xs)

plt.plot(xs, ys1, label='1')
plt.plot(xs, ys2, label='2')
plt.ylabel('Density')
plt.legend(**legend_kwargs);
\end{NotebookIn}

\begin{NotebookImage}
\includegraphics[width=0.75\textwidth]{figures/out/norm-gamma_pdf.png}
\end{NotebookImage}

Though the densities are clearly distinct, they do have a non-trivial overlap.

Let's generate some data and look at a single example to get a sense of the kinds of sequences this HMM produces.

\begin{NotebookIn}
model = HMM(t_dists=t_dists, e_dists=e_dists, start_dist=start_dist)

data = [model.simulate(200, random_state=i) for i in range(10)]
data = [list(zip(*example)) for example in data]

xs, ys = data[0]
lines = {}
for state in t_dists:
    lines[state] = [x == state for x in xs]

fig, axs = plt.subplots(2, 1, sharex=True)
axs[0].plot(ys)
for state, line in sorted(lines.items()):
    axs[1].plot(line, label=state)
axs[1].set_xlabel('Time step')
axs[0].set_ylabel('Emission')
axs[1].set_ylabel('Label')
axs[1].legend(**legend_kwargs);
\end{NotebookIn}

\begin{NotebookImage}
\includegraphics[width=0.75\textwidth]{figures/out/norm-gamma_trace.png}
\end{NotebookImage}

It's apparent there are alternating high and low intervals. However, as a result of the overlap between the emission distributions, the boundaries are not always clearcut.

We'll now fit models with normally distributed emissions using both training methods, beginning with maximum likelihood. The following code implements the same maximum likelihood estimators derived in previous sections. However, because this is an advanced example, we'll do the calculations with NumPy arrays. Because NumPy arrays vectorize operations, this is generally more efficient and cleaner than iterating over each state-emission pair in a for loop.

\begin{NotebookIn}
xstack = np.stack([xs for xs, ys in data])
ystack = np.stack([ys for xs, ys in data])

# Make estimated transition distributions
t_dists_ML = {}
for state1, t_dist in t_dists.items():
    t_dist_ML = {}
    x1 = (xstack[:, :-1] == state1)
    x1_sum = x1.sum()
    for state2 in t_dist:
        x2 = (xstack[:, 1:] == state2)
        x12_sum = (x1 & x2).sum()
        t_dist_ML[state2] = x12_sum / x1_sum
    t_dists_ML[state1] = t_dist_ML
print('ML ESTIMATED T_DISTS')
for state, t_dist_ML in t_dists_ML.items():
    print(f'{state}: {t_dist_ML}')
print()

# Make estimated emission distributions
e_params_ML = {}
for state in e_dists:
    xs = xstack.ravel() == state
    ys = ystack.ravel()[xs]
    loc = ys.mean()
    scale = ys.var()
    e_params_ML[state] = {'mu': loc, 'sigma2': scale}
print('ML ESTIMATED E_PARAMS')
for state, e_param_ML in e_params_ML.items():
    print(f'{state}: {e_param_ML}')
print()

# Make estimated start distribution
start_dist_ML = {}
for state in start_dist:
    start_dist_ML[state] = (xstack[:, 0] == state).sum() / xstack.shape[0]
print('ML ESTIMATED START_DIST')
print(start_dist_ML)
\end{NotebookIn}

\begin{NotebookOut}
ML ESTIMATED T_DISTS
1: {1: 0.9470684039087948, 2: 0.052931596091205214}
2: {1: 0.08923884514435695, 2: 0.910761154855643}

ML ESTIMATED E_PARAMS
1: {'mu': -2.708260094140367, 'sigma2': 24.878695602046125}
2: {'mu': 5.875899028606219, 'sigma2': 22.49724725215567}

ML ESTIMATED START_DIST
{1: 0.3, 2: 0.7}
\end{NotebookOut}

Let's now tackle conditional maximum likelihood training, and after we'll compare the two. Because the implementation is somewhat lengthy, it's included as the function \texttt{fit\_CML} in the accompanying \texttt{utils} module. We only have to supply all the pieces. First, each parameter requires an initial estimate. While it's best practice to inject some randomness into these initial values and choose the best result after several runs, for simplicity, we'll use the ML estimates directly. Because \texttt{t\_dists} and \texttt{start\_dist} will always take the form of a nested dictionary and dictionary, respectively, \texttt{fit\_CML} accepts these objects as is. The emission distributions are more complex, however, because we need to provide the 1) parameter names and their initial estimates, 2) the emission probability or density functions, 3) their derivatives relative to each parameter, 4) functions which transform any constrained parameters to their auxiliary variables, and 5) corresponding functions which reverse the transformation. Accordingly, each of these is supplied as a separate argument in the form of simple or nested dictionaries.

Fortunately, the ML parameter estimates are already in the proper form, so we can go directly to the emission density functions. The hardest part here is choosing our parameters and their names since they must be consistent throughout all functions we define. We'll use the names \texttt{mu} and \texttt{sigma2}, so they correspond to the equations derived in the previous section. These parameters, however, differ slightly from those used by the SciPy statistics module, so we'll create a thin wrapper around its implementation of the normal density function.

\begin{NotebookIn}
def norm_pdf(y, mu, sigma2):
    return stats.norm.pdf(y, loc=mu, scale=sigma2**0.5)
\end{NotebookIn}

After defining our new density function, we simply map the states to it in a dictionary.

\begin{NotebookIn}
e_funcs = {1: norm_pdf,
           2: norm_pdf}
\end{NotebookIn}

The derivatives require more explanation. As we saw before, the derivative of the conditional log-likelihood function involves a sum over a function that includes $n_i$ and  $m_i$. However, the terms involving those variables are the same for any emission distribution whose parameters are not shared between states, so it would be redundant to include them every time. Instead, in \texttt{fit\_CML} the derivative relative to the auxiliary variable $z$ takes the form

\begin{equation*}
\frac{\partial L}{\partial z}
= -\left[
         \sum_t \left( m_i(t) - n_i(t) \right)
                f(y_t)
   \right],
\end{equation*}

and we only provide the function $f$ as the ``derivative.'' Note the derivative is relative to an unconstrained auxiliary variable. For unconstrained parameters, there is no difference, but the derivatives of constrained parameters must be relative to their unconstrained counterparts. These functions must also accept the emissions $y$ as their first argument and the parameters by their names as subsequent arguments. Finally, because all parameters and their auxiliary variables are passed into the derivative functions, they must allow a variable number of keyword arguments with the \texttt{**kwargs} syntax to ``catch'' any unused arguments.

Though these requirements may sound complex, in practice they are fairly intuitive. For example, our derivatives are direct translations of the expressions we found earlier.

\begin{NotebookIn}
def norm_prime_mu(y, mu, sigma2, **kwargs):
    return (y - mu) / sigma2

def norm_prime_sigma2(y, mu, sigma2, **kwargs):
    term1 = - 1 / sigma2
    term2 = (y - mu) ** 2 / sigma2 ** 2
    return 0.5 * (term1 + term2) * sigma2
\end{NotebookIn}

These functions are then mapped to their corresponding state and parameter combination using a nested dictionary.

\begin{NotebookIn}
e_primes = {1: {'mu': norm_prime_mu, 'sigma2': norm_prime_sigma2},
            2: {'mu': norm_prime_mu, 'sigma2': norm_prime_sigma2}}
\end{NotebookIn}

The two sets of variable transformation functions are defined and structured similarly. The auxiliary variables are named with the convention that the corresponding auxiliary variable for parameter \texttt{param} is called \texttt{param\_aux}.

\begin{NotebookIn}
def mu2aux(mu, **kwargs):
    return mu

def aux2mu(mu_aux, **kwargs):
    return mu_aux

def sigma22aux(sigma2, **kwargs):
    return log(sigma2)

def aux2sigma2(sigma2_aux, **kwargs):
    return exp(sigma2_aux)

e_param2aux = {1: {'mu': mu2aux, 'sigma2': sigma22aux},
               2: {'mu': mu2aux, 'sigma2': sigma22aux}}
e_aux2param = {1: {'mu': aux2mu, 'sigma2': aux2sigma2},
               2: {'mu': aux2mu, 'sigma2': aux2sigma2}}
\end{NotebookIn}

We're finally ready to fit our model to the data! All we have to do now is pass our arguments into \texttt{fit\_CML}. There are a few more optional parameters, but their meanings should be clear from the previous discussion.

\begin{NotebookIn}
params = fit_CML(data,
                 t_dists=t_dists_ML,
                 e_params=e_params_ML, e_funcs=e_funcs, e_primes=e_primes,
                 e_param2aux=e_param2aux, e_aux2param=e_aux2param,
                 start_dist=start_dist_ML, eta=0.3, maxiter=500, verbose=False)
t_dists_CML, e_params_CML, start_dist_CML = params

print('CML ESTIMATED T_DISTS')
for state, t_dist_CML in t_dists_CML.items():
    print(f'{state}: {t_dist_CML}')
print()
print('CML ESTIMATED E_PARAMS')
for state, e_param_CML in e_params_CML.items():
    print(f'{state}: {e_param_CML}')
print()
print('CML ESTIMATED START_DIST')
print(start_dist_CML)
\end{NotebookIn}

\begin{NotebookOut}
CML ESTIMATED T_DISTS
1: {1: 0.9485741440805543, 2: 0.05142585591944567}
2: {1: 0.09604341135218855, 2: 0.9039565886478114}

CML ESTIMATED E_PARAMS
1: {'mu': -2.7344728682990804, 'sigma2': 23.630241342073358}
2: {'mu': 5.990857687710255, 'sigma2': 12.528992301575078}

CML ESTIMATED START_DIST
{1: 0.014065968700277301, 2: 0.9859340312997227}
\end{NotebookOut}

The parameter estimates don't appear to have changed too drastically, but let's plot the density functions with the true parameters against the two estimated densities to get a better sense of how they differ.

\begin{NotebookIn}
fig, axs = plt.subplots(2, 1, figsize=(6.4, 7.2))
xs = np.linspace(-25, 25, 250)

ys1 = e_dists[1].pdf(xs)
ys2 = e_dists[2].pdf(xs)
for ax in axs:
    ax.plot(xs, ys1, label='true state 1')
    ax.plot(xs, ys2, label='true state 2')
    ax.set_ylabel('Density')

ys1 = stats.norm.pdf(xs,
                     loc=e_params_ML[1]['mu'],
                     scale=e_params_ML[1]['sigma2']**0.5)
ys2 = stats.norm.pdf(xs,
                     loc=e_params_ML[2]['mu'],
                     scale=e_params_ML[2]['sigma2']**0.5)
axs[0].plot(xs, ys1, label='ML state 1')
axs[0].plot(xs, ys2, label='ML state 2')
axs[0].legend(**legend_kwargs)

ys1 = stats.norm.pdf(xs,
                     loc=e_params_CML[1]['mu'],
                     scale=e_params_CML[1]['sigma2']**0.5)
ys2 = stats.norm.pdf(xs,
                     loc=e_params_CML[2]['mu'],
                     scale=e_params_CML[2]['sigma2']**0.5)
axs[1].plot(xs, ys1, label='CML state 1')
axs[1].plot(xs, ys2, label='CML state 2')
axs[1].legend(**legend_kwargs);
\end{NotebookIn}

\begin{NotebookImage}
\includegraphics[width=0.75\textwidth]{figures/out/norm-gamma_hat_pdf.png}
\end{NotebookImage}

The change is most obvious for state 2 whose curve is much more peaked. However, even the state 1 curve has shifted slightly as well. These differences can be easily understood in terms of the objective functions. Maximum likelihood estimation chooses parameters that best describe the data, so the distribution is wider to account for the skew. In contrast, conditional maximum likelihood estimation chooses parameters that maximize prediction accuracy, so the distribution is narrower to better capture the decision boundary.

However, has all this work actually yielded an improvement in the accuracy of state inference? We can check by using the estimated models to calculate the probability of each state at each time step. Since these probabilities range continuously from zero to one, they are in a sense soft labels. Accuracy calculations, though, require hard state assignments. There are any number of methods to convert the probabilities into states, but the easiest is using a threshold of 0.5. This is a natural choice because we have no reason to favor one state over the other, but in general the threshold should reflect the costs of incorrect assignments for each state. For example, credit card companies err on the side of caution when alerting customers to potential fraud because the cost of confirming a valid transaction is much smaller than refunding a fraudulent one.

\begin{NotebookIn}
e_dists_ML = {}
for state, e_param_ML in e_params_ML.items():
    e_dists_ML[state] = stats.norm(loc=e_param_ML['mu'],
                                   scale=e_param_ML['sigma2']**0.5)
e_dists_CML = {}
for state, e_param_CML in e_params_CML.items():
    e_dists_CML[state] = stats.norm(loc=e_param_CML['mu'],
                                    scale=e_param_CML['sigma2']**0.5)

model_ML = HMM(t_dists=t_dists_ML,
               e_dists=e_dists_ML,
               start_dist=start_dist_ML)
model_CML = HMM(t_dists=t_dists_CML,
                e_dists=e_dists_CML,
                start_dist=start_dist_CML)

xstack_ML = []
xstack_CML = []
for example in data:
    _, ys = example
    fbs_ML = model_ML.forward_backward(ys)
    fbs_CML = model_CML.forward_backward(ys)

    xstack_ML.append(fbs_ML[1])
    xstack_CML.append(fbs_CML[1])
xstack_ML = np.stack(xstack_ML)
xstack_CML = np.stack(xstack_CML)

threshold = 0.5
accuracy_ML = ((xstack_ML >= threshold) == (xstack == 1)).sum()
accuracy_CML = ((xstack_CML >= threshold) == (xstack == 1)).sum()
print('ML accuracy:', accuracy_ML / xstack.size)
print('CML accuracy:', accuracy_CML / xstack.size)
\end{NotebookIn}

\begin{NotebookOut}
ML accuracy: 0.948
CML accuracy: 0.947
\end{NotebookOut}

The CML model actually has a slight decrease in accuracy. However, we're not seeing the full picture by using a single threshold, and we should instead measure the models' responses as the threshold is varied continuously over the full range of the data. A common implementation of this idea is the \textit{receiver operating characteristic} (ROC) curve which plots the true positive rate against the false positive rate at varying thresholds. It essentially measures the trade-off between the increases in true and false positives as the threshold is decreased. Thus, the curve of a model with good separation between classes should rise steeply and rapidly level off.

Let's plot the ROC curves for the two models.

\begin{NotebookIn}
fpr_ML, tpr_ML, _ = roc_curve((xstack == 1).ravel(), xstack_ML.ravel())
fpr_CML, tpr_CML, _ = roc_curve((xstack == 1).ravel(), xstack_CML.ravel())

fig, ax = plt.subplots()
ax.plot(fpr_ML, tpr_ML, label='ML')
ax.plot(fpr_CML, tpr_CML, label='CML')
ax.set_xlabel('False positive rate')
ax.set_ylabel('True positive rate')
ax.legend(**legend_kwargs)

idx_ML = fpr_ML <= 0.1
idx_CML = fpr_CML <= 0.1
axins = ax.inset_axes([0.25, 0.1, 0.7, 0.5])
axins.plot(fpr_ML[idx_ML], tpr_ML[idx_ML], label='ML')
axins.plot(fpr_CML[idx_CML], tpr_CML[idx_CML], label='CML')
axins.set_xticklabels([])
axins.set_yticklabels([])
ax.indicate_inset_zoom(axins, edgecolor='black');
\end{NotebookIn}

\begin{NotebookImage}
\includegraphics[width=0.75\textwidth]{figures/out/norm-gamma_ROC.png}
\end{NotebookImage}

A small effect but an effect nonetheless! Because in this case the ML model was already close to the true model, the improvement in the discriminatively trained model was fairly marginal. However, the greater the mismatch between the data and a model, the greater the effect on state inference, so for real data the difference between the two approaches can be substantial.

\section{Conclusion}

We have at last reached the end of the training tutorial! It's been quite a tour through several fundamental concepts in statistics and machine learning. While many of these ideas seem complex, as shown throughout this tutorial, they can often be implemented in only a few dozen lines of code. The previous examples have covered some of the most widely used distributions in statistics, but we've only scratched the surface of what is possible with probabilistic modeling. Fortunately, once the theoretical heavy lifting is out of the way, supporting a new distribution is often a relatively straightforward task of writing and solving a set of equations. Hopefully, then, these examples will inspire you to derive and implement your own MLEs and EM or gradient update equations if the need arises. Good luck!

### 2023-02-23 chapter3/training_tutorial_supporting.tex
Added remaining chapter 3 material

\begin{center}
\textbf{Listing B1: Implementation of Baum-Welch for categorical distributions}
\end{center}

\begin{NotebookIn}
epsilon = 0.01
maxiter = 100

ll0 = None
model_hat = HMM(t_dists=t_dists_hat,
                e_dists=e_dists_hat,
                start_dist=start_dist_hat)
for numiter in range(maxiter):
    # Initialize count dictionaries
    ps = []
    t_counts = {state1: {state2: 0 for state2 in t_dist}
                for state1, t_dist in t_dists_hat.items()}
    e_counts = {state: {emit: 0 for emit in e_dist}
                for state, e_dist in e_dists_hat.items()}
    start_count = {state: 0 for state in start_dist_hat}

    # Get counts across all examples
    for example in data:
        xs, ys = example
        fs, ss_f = model_hat.forward(ys)
        bs, ss_b = model_hat.backward(ys)

        p = reduce(lambda x, y: x+y, map(log, ss_f))
        ss_f = list(accumulate(map(log, ss_f)))
        ss_b = list(accumulate(map(log, ss_b[::-1])))[::-1]
        ps.append(p)

        # t_counts
        for t in range(len(ys)-1):
            for state1, t_count in t_counts.items():
                for state2 in t_count:
                    term1 = fs[state1][t] * t_dists_hat[state1][state2]
                    term2 = e_dists_hat[state2][ys[t+1]] * bs[state2][t+1]
                    count = term1 * term2
                    t_count[state2] += count * exp(ss_f[t] + ss_b[t+1] - p)

        # e_counts
        for t in range(len(ys)):
            for state, e_count in e_counts.items():
                if ys[t] in e_count:
                    count = fs[state][t]*bs[state][t]
                    e_count[ys[t]] += count * exp(ss_f[t] + ss_b[t] - p)

        # start_count
        for state in start_count:
            count = fs[state][0]*bs[state][0]
            start_count[state] += count * exp(ss_f[0] + ss_b[0] - p)

    # Format parameters for display
    t_string = pprint.pformat(t_dists_hat)
    t_string = t_string.replace('\n', '\n' + len('t_dists: ')*' ')
    e_string = pprint.pformat(e_dists_hat)
    e_string = e_string.replace('\n', '\n' + len('e_dists: ')*' ')
    start_string = pprint.pformat(start_dist_hat)

    # Check stop condition
    # Don't want to repeat calculations,
    # so ith iterate checks previous update
    # For example, 0th iterate shows initial parameters,
    # and 1st iterate shows results of first update
    ll = sum(ps)
    if ll0 is not None and abs(ll - ll0) < epsilon:
        print(f'FINAL VALUES')
        print('log-likelihood:', ll)
        print('delta log-likelihood:', ll-ll0 if ll0 is not None else None)
        print('t_dists:', t_string)
        print('e_dists:', e_string)
        print('start_dist:', start_string)
        break

    # Print results
    print(f'ITERATION {numiter}')
    print('log-likelihood:', ll)
    print('delta log-likelihood:', ll-ll0 if ll0 is not None else None)
    print('t_dists:', t_string)
    print('e_dists:', e_string)
    print('start_dist:', start_string)
    print()

    # Normalize all counts and update model
    t_dists_hat = {}
    for state1, t_count in t_counts.items():
        t_sum = sum(t_count.values())
        t_dist_hat = {}
        for state2, count in t_count.items():
            t_dist_hat[state2] = count / t_sum
        t_dists_hat[state1] = t_dist_hat

    e_dists_hat = {}
    for state, e_count in e_counts.items():
        e_sum = sum(e_count.values())
        e_dist_hat = {}
        for emit, count in e_count.items():
            e_dist_hat[emit] = count / e_sum
        e_dists_hat[state] = e_dist_hat

    start_sum = sum(start_count.values())
    start_dist_hat = {}
    for state, count in start_count.items():
        start_dist_hat[state] = count / start_sum

    ll0 = ll
    model_hat = HMM(t_dists=t_dists_hat,
                    e_dists=e_dists_hat,
                    start_dist=start_dist_hat)
\end{NotebookIn}

### 2023-02-23 chapter3/figures/figures.py

@@ -1,0 +1,672 @@
+"""Make all figures from training tutorial."""
+
+# In[1]:
+import os
+import pprint
+import random
+from functools import reduce
+from itertools import accumulate
+
+import matplotlib.pyplot as plt
+import numpy as np
+import scipy.stats as stats
+from homomorph import HMM
+from numpy import exp, log
+from sklearn.metrics import roc_curve
+from utils import fit_CML
+
+dpi = 300
+legend_kwargs = {'frameon': False, 'loc': 'center left', 'bbox_to_anchor': (1, 0.5)}
+
+if not os.path.exists('out/'):
+    os.mkdir('out/')
+
+
+# In[2]:
+t_dists = {1: {1: 0.95, 2: 0.05},
+           2: {1: 0.05, 2: 0.9, 3: 0.05},
+           3: {2: 0.35, 3: 0.65}}
+e_dists = {1: {'A': 1},
+           2: {'A': 0.5, 'B': 0.5},
+           3: {'B': 1}}
+start_dist = {1: 0.2, 2: 0.5, 3: 0.3}
+
+model = HMM(t_dists=t_dists, e_dists=e_dists, start_dist=start_dist)
+
+
+# In[3]:
+data = [model.simulate(200, random_state=i) for i in range(10)]  # Use a different random seed for each example
+data = [list(zip(*example)) for example in data]
+
+
+# In[4]:
+# Make transition count dicts and add pseudocounts
+t_pseudo = 0.1
+t_counts = {}
+for state1, t_dist in t_dists.items():
+    t_count = {}
+    for state2 in t_dist:
+        t_count[state2] = t_pseudo
+    t_counts[state1] = t_count
+
+# Add observed counts
+for example in data:
+    xs, ys = example
+    state0 = xs[0]
+    for state1 in xs[1:]:
+        t_counts[state0][state1] += 1
+        state0 = state1
+
+# Normalize counts
+t_dists_hat = {}
+for state1, t_count in t_counts.items():
+    t_sum = sum(t_count.values())
+    t_dist_hat = {}
+    for state2, count in t_count.items():
+        t_dist_hat[state2] = count / t_sum
+    t_dists_hat[state1] = t_dist_hat
+
+
+# In[5]:
+# Collect all possible emissions
+e_set = set()
+for example in data:
+    xs, ys = example
+    e_set.update(ys)
+
+# Make emission count dicts and add pseudocounts
+e_pseudo = 0.1
+e_counts = {}
+for state in t_dists:
[…]

### 2023-02-23 chapter3/figures/utils.py

@@ -1,0 +1,266 @@
+"""Utility functions for training tutorial."""
+
+import pprint
+from copy import deepcopy
+from functools import reduce
+from itertools import accumulate
+
+import homomorph
+import numpy as np
+from numpy import exp, log
+
+
+class ArrayRV:
+    """Class that allows HMM class to interface with a pre-computed probability array.
+
+    The rvs method is only defined so the HMM recognizes it as a proper
+    random variable. Since random variates are not needed in this script, the
+    body of the method is left blank.
+    """
+    def __init__(self, array):
+        self.array = array
+
+    def pmf(self, x):
+        return self.array[x]
+
+    def rvs(self, random_state=None):
+        pass
+
+
+class HMM(homomorph.HMM):
+    """A class that modifies the default HMM to accept pre-calculated intermediates.
+
+    Discriminative training requires decoding of both states and transitions.
+    Both of these calculations rely on the forward and backward variables.
+    Since these are the most computationally intensive steps in decoding and
+    both are used in state and transition decoding, this modified class accepts
+    the forward and backward variables as arguments rather than calculating
+    them from scratch each time.
+    """
+    def forward_backward1(self, emits, fs, ss_f, bs, ss_b):
+        """Posterior decoding of states."""
+        p = reduce(lambda x, y: x+y, map(log, ss_f))
+        ss_f = list(accumulate(map(log, ss_f)))
+        ss_b = list(accumulate(map(log, ss_b[::-1])))[::-1]
+
+        fbs = {state: [] for state in self.states}
+        for i in range(len(emits)):
+            for state, fb in fbs.items():
+                fbs[state].append(fs[state][i]*bs[state][i]*exp(ss_f[i]+ss_b[i]-p))
+
+        return fbs
+
+    def forward_backward2(self, emits, fs, ss_f, bs, ss_b):
+        """Posterior decoding of transitions."""
+        p = reduce(lambda x, y: x+y, map(log, ss_f))
+        ss_f = list(accumulate(map(log, ss_f)))
+        ss_b = list(accumulate(map(log, ss_b[::-1])))[::-1]
+
+        pairs = [(state0, state1) for state0, t_dist in self.t_dists.items() for state1 in t_dist]
+        fbs = {pair: [] for pair in pairs}
+        for i in range(len(emits)-1):
+            for (state0, state1), fb in fbs.items():
[…]

### 2023-02-24 chapter0/chapter0.tex
Updated section formatting in remaining files
Added discussion of structure to introduction

@@
 %======= DRAFT =======
+\section{Background}
 Life is a physical phenomenon.
@@
 
-Many studies investigated various structural and functional aspects of IDRs and IDPs followed in the following years.
+Over the following years, many studies investigated the structural and functional properties of IDRs.
+They found that although IDRs still have sequence-structure-function relationships, they play by a very different set of rules.
+These differences manifest at all three levels but are at first most easily understood in terms of structure.
+Strictly speaking, a protein's structure refers its three-dimensional arrangement of atoms.
+Thus, ``structured'' regions in proteins, called domains\footnote{Though there are various overlapping definitions, domains typically refer to independently folding regions of a protein.
+Domains are also described as discrete functional or evolutionary elements since proteins may contain several domains which are connected by unstructured ``linker'' sequences.}, typically adopt a small number of structures.
+This does not imply structured domains are completely rigid, though.
+At the molecular level, everything is in constant motion.
+For example, at room temperature an average water molecule moves at over 500 meters per second!
+However, the aqueous environment is so dense that it will collide with something after moving only a fraction of its own length.
+Structured domains are buffeted by these collisions with water and other molecules, but they are constrained by the rigid bonds and interactions between amino acid residues in the chain.
+Thus, while structured domains can ``flex'' and ``breathe,'' these motions are minor variations on their overall structure.
+
+In some cases, structured domains have multiple structures, or ``conformations,''  which are related to different functional states.
+For example, hemoglobin has two forms, traditionally called the T and R states.
+The T state is hemoglobin's oxygen-free form, but on binding oxygen its structure shifts to the R state.
+The change is small, differing only by a few hydrogen atoms at the points of greatest change.
+However, this enough to re-orient the atoms that interact with oxygen, allowing it to bind more tightly and promoting oxygen uptake at the other three binding sites.
+Once the red blood cells reach their destination, other physiological factors favor the adoption of the T state, which coordinates the release of all four oxygen atoms.
+Thus, when structured domains have multiple conformations, they are generally discrete forms without stable intermediates.
+
+In contrast, IDRs have no fixed spatial relationship between their atoms and are sometimes described as ``conformational ensembles,'' \textit{i.e.} a collection of related structures.
+
+\section{Aims}
+
+This is where my aims will go!

### 2023-02-27 chapter0/chapter0.tex
Removed tennis court analogy and replaced with statistic on volume

@@
 Each red blood cell contains as many as 270 million molecules of hemoglobin, each of which can carry up to four oxygen molecules~\cite{Pierig2008}.
-To put this value in perspective, if each molecule of hemoglobin were the size of a tennis ball, the hemoglobin molecules in a single red blood cell would cover a tennis court in over 10 packed layers\footnote{A standard tennis ball is 1.35 in or 0.1125 ft in radius, so the volume of 270 million tennis balls of this size packed as efficiently as possible is given by $2.70 \times 10^6 \times \frac{4}{3} \pi (0.1125 \text{ ft})^3 \times \frac{3\sqrt{2}}{\pi} = 2.17 \times 10^4 \text{ ft}^3$.
-A singles tennis court is 78 ft by 27 ft or 2106 ft\textsuperscript{2}, so dividing the first by the second yields slightly over 10 layers.}.
-Because red blood cells are so dense with hemoglobin, any defect in hemoglobin can dramatically impact the structure of the red blood cells themselves.
+Because red blood cells are so dense with hemoglobin, composing roughly 35\% of their total volume, any defect in hemoglobin can dramatically impact the structure of the red blood cells themselves~\cite{Kanias2009}.
 A well-studied example is sickle cell disease where an error in the body's hemoglobin molecules deforms red blood cells into a characteristic sickle shape.

### 2023-02-27 chapter0/chapter0.tex
Added footnote to water speed

@@
 At the molecular level, everything is in constant motion.
-For example, at room temperature an average water molecule moves at over 500 meters per second!
-However, the aqueous environment is so dense that it will collide with something after moving only a fraction of its own length.
+For example, at room temperature an average water molecule moves at over 500 meters per second!\footnote{This value was derived using the Maxwell–Boltzmann distribution, which is a physical model of the speeds of particles in an ideal gas.
+Clearly, water is not a gas at room temperature, so it should be considered a rough approximation.} However, the aqueous environment is so dense that it will collide with something after moving only a fraction of its own length.
 Structured domains are buffeted by these collisions with water and other molecules, but they are constrained by the rigid bonds and interactions between amino acid residues in the chain.

### 2023-03-08 chapter0/chapter0.tex
Clarified wording around water speed analogy

@@
 For example, at room temperature an average water molecule moves at over 500 meters per second!\footnote{This value was derived using the Maxwell–Boltzmann distribution, which is a physical model of the speeds of particles in an ideal gas.
-Clearly, water is not a gas at room temperature, so it should be considered a rough approximation.} However, the aqueous environment is so dense that it will collide with something after moving only a fraction of its own length.
-Structured domains are buffeted by these collisions with water and other molecules, but they are constrained by the rigid bonds and interactions between amino acid residues in the chain.
+Clearly, water is not a gas at room temperature, so it should be considered a rough approximation.} However, liquid water is so dense that it will collide with something after moving only a fraction of its own length.
+Likewise, in the cellular environment structured domains are buffeted by collisions with water and other molecules, but they are constrained by the rigid bonds and interactions between amino acid residues in the chain.
 Thus, while structured domains can ``flex'' and ``breathe,'' these motions are minor variations on their overall structure.

### 2023-03-09 chapter0/chapter0.tex
Clarified relationship between not having defined structured and the word disorder

@@
 However, as a result of this unique role in providing tissue elasticity, elastin's disorder was viewed as a specific adaptation rather than a general mechanism of protein function.
-In other cases, proteins had regions which returned undefined or highly variable atomic coordinates when analysed with structural techniques, but because these regions were often short loops between structured regions, they were seen as ``linkers'' which facilitated the structure of the functional portions of the protein.
+In other cases, proteins had regions which returned undefined or highly variable atomic coordinates when analysed with structural techniques, indicating they lacked defined structures and were ``disordered.'' Because these segments were often short loops between structured regions, they were seen as ``linkers'' which facilitated the structure of the functional portions of proteins.
 By the early 2000s, though, enough exceptions had accumulated that scientists began to recognize that fully and partially disordered proteins were involved in many biological processes~\cite{Plaxco1997, Wright1999, Dunker2001}.

### 2023-03-09 chapter0/chapter0.tex
Replaced "structured domains" with the more specific "folded domains"

@@
 Strictly speaking, a protein's structure refers its three-dimensional arrangement of atoms.
-Thus, ``structured'' regions in proteins, called domains\footnote{Though there are various overlapping definitions, domains typically refer to independently folding regions of a protein.
-Domains are also described as discrete functional or evolutionary elements since proteins may contain several domains which are connected by unstructured ``linker'' sequences.}, typically adopt a small number of structures.
-This does not imply structured domains are completely rigid, though.
+Thus, ``structured'' regions in proteins, often called domains\footnote{Though there are various overlapping definitions, domains typically refer to independently folding regions of a protein.
+Domains are also described as discrete functional or evolutionary elements since proteins may contain several domains which are connected by unstructured ``linker'' sequences.}, typically fold into a small number of related structures.
+This does not imply these folded domains are completely rigid, though.
 At the molecular level, everything is in constant motion.
@@
 Clearly, water is not a gas at room temperature, so it should be considered a rough approximation.} However, liquid water is so dense that it will collide with something after moving only a fraction of its own length.
-Likewise, in the cellular environment structured domains are buffeted by collisions with water and other molecules, but they are constrained by the rigid bonds and interactions between amino acid residues in the chain.
-Thus, while structured domains can ``flex'' and ``breathe,'' these motions are minor variations on their overall structure.
+Likewise, in the cellular environment folded domains are buffeted by collisions with water and other molecules, but they are constrained by the rigid bonds and interactions between amino acid residues in the chain.
+Thus, while folded domains can ``flex'' and ``breathe,'' these motions are minor variations on their overall structure.
 

### 2023-03-09 chapter0/chapter0.tex
Added proteasome example

@@
 
-In some cases, structured domains have multiple structures, or ``conformations,''  which are related to different functional states.
+In some cases, folded domains have multiple structures, or ``conformations,'' which are related to different functional states.
 For example, hemoglobin has two forms, traditionally called the T and R states.
 The T state is hemoglobin's oxygen-free form, but on binding oxygen its structure shifts to the R state.
-The change is small, differing only by a few hydrogen atoms at the points of greatest change.
+The change is small, differing at most by only a few hydrogen atoms.
 However, this enough to re-orient the atoms that interact with oxygen, allowing it to bind more tightly and promoting oxygen uptake at the other three binding sites.
 Once the red blood cells reach their destination, other physiological factors favor the adoption of the T state, which coordinates the release of all four oxygen atoms.
-Thus, when structured domains have multiple conformations, they are generally discrete forms without stable intermediates.
+In the other cases, conformational changes can dramatically re-organize a protein's structure.
+For example, the 26S proteasome is a complex of proteins responsible for degrading other proteins.
+Its structure is highly complex and consists of over three dozen distinct protein ``subunits,'' which are in turn organized into three subcomplexes: a lid, a base, and a core~\cite{Finley2016, Bard2018}.
+The functions of these subcomplexes are roughly analogous to the parts of a paper shredder.
+The lid is like the outer shell because it regulates access to the ``motor'' in the base and the ``blades'' in the core that pull in and degrade the protein, respectively.\footnote{As with many analogies, this comparison to a paper shredder simplifies several structural and functional details of the proteasome.
+For example, in a paper shredder the motor powers the blades which both pull in and shred the paper.
+In the proteasome, however, these are distinct steps.
+The motor subunits in the base first physically interact with the protein to simultaneously unfold and pull it into the base.
+A different set of subunits in the base then break the exposed chemical bonds between amino acid residues in the chain.} Like a paper shredder, the motor is only engaged when a protein is correctly positioned in the lid.
+Unlike a paper shredder, however, the motor is activated by a conformational change rather than a physical switch.
+When the tail of a protein marked for degradation is inserted in the motor, the lid shifts by nearly forty hydrogen atoms to align the motor with the pore that leads into the core subcomplex.
+Despite the scale of this re-arrangement, it occurs over the span of only half a second~\cite{Bard2019}.
+Thus, when folded domains have multiple conformations, they are generally discrete forms without stable intermediates.
+
+In contrast, IDRs have no fixed spatial relationship between their atoms and are sometimes described as ``conformational ensembles,'' \textit{i.e.} heterogeneous collections of structures.
+Though folded domains can undergo conformational changes with dramatic differences in the distan, they are usually triggered by specific events, and the intermediate structures are transitory, whereas IDRs populate a continuum of structures with large variations in the distances and orientations between residues.
+I'm usually in
 

### 2023-03-11 chapter0/chapter0.tex
Added section on sequence-structure relationships of IDRs

@@
 
-In contrast, IDRs have no fixed spatial relationship between their atoms and are sometimes described as ``conformational ensembles,'' \textit{i.e.} heterogeneous collections of structures.
-Though folded domains can undergo conformational changes with dramatic differences in the distan, they are usually triggered by specific events, and the intermediate structures are transitory, whereas IDRs populate a continuum of structures with large variations in the distances and orientations between residues.
-I'm usually in
+In contrast, IDRs have no fixed spatial relationship between their atoms, so their structures are sometimes described as ``conformational ensembles,'' \textit{i.e.} collections of conformations where the distances and orientations between residues can vary considerably.
+Furthermore, IDRs populate a continuum of structural states over time whereas when folded domains undergo dramatic conformational changes, they are usually triggered by specific environmental signals or chemical modifications, and the intermediate structures are transitory.
+Despite the diversity of conformations available to IDRs, they can be broadly grouped into one of several qualitative descriptions which range from extended ``coils'' to more compact ``globules.'' The specific conformational profile favored by a given IDR, however, is dictated by its local composition of amino acid residues.
+Because the chemical properties of each amino acid in the protein alphabet are dictated its specific arrangement of atoms, each has a unique impact on a protein's structure.
+However, to simplify discussion and analysis amino acids are often compared by quantitative factors like size or charge.
+One of the most useful scales for describing an amino acid's overall effect on protein structure is hydrophobicity, which measures a molecule's tendency to associate with water.
+Molecules which attract water are hydrophilic (``water loving''), and molecules which repel water are hydrophobic (``water fearing'').
+Hydrophilic molecules, like sugar or alcohol, easily dissolve in water whereas hydrophobic molecules, like fats and oils, remain separate.
+Since the cellular environment is largely water, the hydrophobic residues in proteins tend to aggregate into a ``hydrophobic core.'' This ``hydrophobic collapse'' is a major driving force in the early stages of protein folding, so a protein's relative number of hydrophilic and hydrophobic residues is a key determinant of whether it is folded or disordered.
+
+Unsurprisingly, disordered regions are characterized by a relative depletion of hydrophobic residues.
+However, there is no simple formula which accurately predicts disorder in a protein using only the hydrophobicity values of its constituent residues.
+Other factors, such as the number and distribution of charged residues, also impact a sequence's predisposition for disorder~\cite{vanderLee2014, Das2015}.
+For example, sequences with high number of either positively or negatively charged residues, called polyelectrolytes, tend to form stiff rods because the like charges repel each other.
+However, if a sequence contains a high number of positively and negatively charged residues in roughly equal proportion, the sequence is called a strong polyampholyte, and its conformational profile depends on the distribution of those charged residues.
+If the two classes of residues are segregated into separate blocks of like charges, they attract and form hairpins.
+However, if they are evenly distributed, the attractive and repulsive forces balance on average, and the sequence generally assumes expanded coil-like conformations.
+High numbers of polar amino acids, which are hydrophilic but not charged, are associated with semi-compact globules.
+Though their ``side chains,'' the portions which give each amino acid its unique identity, are hydrophilic, their interactions with water are not sufficient to overcome the tendency of the hydrophobic ``backbone,'' composed the donor and receptor sites common to all amino acids, to self-associate.
+Thus, like folded domains, the structures of disordered proteins are dictated by their sequences.
+However, because IDRs do not make stable contacts between specific residues in their chains, multiple sequences can generally correspond to a single conformational profile.
 

### 2023-03-11 chapter0/chapter0.tex
Added aims to outline
Changed example of physical process to EM because it's more relevant to biology than decay

@@
 %       2. They are constantly sampling conformations which may have large differences in the spatial relationships between residues
-%    D. Protein folding driven by hydrophobic interactions
+%       3. The conformations of disordered regions may adopt more extended conformations or collapsed conformations
+%       3. The composition of a disordered region dictates its conformational profile. However, to describe
+%       4. As mentioned previously, proteins are linear chains of amino acids, and the amino alphabet contains 20 letters
+%       5. Though each amino acid is chemically unique, many of their properties can be understood in terms of
 %       1. Disordered regions largely characterized by enrichment and depletion in hydrophilic and hydrophobic residues, respectively
@@
 %    A. Flexibility confers several advantages
-%        1. Structural flexibility is directly related to functional plasticity
+%        1. The structural flexibility of IDRs is directly related their functional plasticity
 %        2. Proteins involved in signaling, transport, and catalysis require specific geometries to interact their intended targets efficiently
@@
 %AIMS
-%I. Structure-function relationships of disordered proteins are unclear
-%    A. Examples of studies that have found specific cases of conservation in IDRs
-%    B. Alan Moses paper as reference in yeast
+%I.
+Despite significant increases in the accuracy of computational methods which predict disorder from sequence alone, the relationship between sequence and function in IDRs remains poorly understood, and functions are not readily associated with arbitrary IDRs.
+%    A. In contrast, folded domains can identified by characteristic sequences of amino acids
+%        1. Because structured domains make specific contacts between residues, their sequences of amino acids are generally conserved or evolve slowly
+%        2. Thus, for structured domains their specific sequence of amino acids constitute a unique signature
+%        3. Traditional bioinformatics methods use these signatures to detect similarities between proteins
+%        4. Structural and functional information is transferred between proteins with similar sequences
+%        5. In practice, however, this is extremely difficult. Fully mapping the relationship between sequences and structure was an active area of research for decades
+%            a.
+Recent developments in machine learning have largely resolved this question in most cases, but still required an advanced computational model trained on nearly two hundred thousand experimentally determined structures.
+%    B. Disordered proteins, however, challenge this sequence-dependent model of protein structure and function.
+%    X. Though there are exceptions, many IDRs evolve extremely rapidly and are not identifiable by the specific sequence of amino acids
+%    C. However, growing evidence that IDRs evolve under a different set of constraints
+%        0. Rather than preserving a specific sequence in order to preserve contacts and structure, they preserve distributed features such as composition, length, charge, or the presence of motifs
+%        1. Because many sequences can yield an IDR with a specific conformational profile or local arrangements, they are not constrained to preserve the precise order of amino acid residues
+%    D. Examples of studies that have found specific cases of conservation in IDRs
+%        1. Earlier specific examples of feature conservation in proteins
+%        2.
+However, a recent study from Zarin et al was the first to demonstrate of widespread of feature conservation in IDRs across an entire proteome, that is the set of all proteins found in an organism
+%    E.
+The analyses used simulations and evolutionary comparisons across a set of yeast genomes, a foundational model organism in molecular biology research.
+However, no subsequent studies have investigated similar patterns of conservations, or any patterns of conservation at all, characterize in IDRs in other systems.
+%        0. No conservation global patterns of conservation would suggest such preserved conservation are unique to yeast
+%        1. Patterns of conservation but are distinct from yeast would indicate global patterns
+%        2. Similar patterns of conservation with similar functional associations would indicate a universal taxonomy of IDRs
 %II. Use evolutionary analyses to identify conserved properties of disordered regions
-%    A. Conservation is powerful signal of function
-%        1. Examples of other studies that applied evolution
-%    B. Outline overall approach and make correspondence to chapters
-%        1. Identification of orthologous groups
-%        2. Application of evolutionary analyses
-%
+%    A.
+Though modern molecular genetic and gene editing techniques it is in principle possible to test the relationship between distributed features and the structure and function of specific proteins in living systems directly, the vast number of combinations make this approach infeasible in practice.
+%    B. Instead life constantly explores the space of allowed proteins through evolutionary changes
+%    C. Properties which are unimportant for a protein's function, or at a larger scale, an organism's survival offer no benefit for their maintenance and are gradually degraded and ultimately lost
+%    A. Thus, conservation is powerful signal of function, which are illuminated by evolutionary comparisons between related sequences.
+%    X.
+Such comparisons, however, require the identification of IDRs with common ancestry that perform ``equivalent'' functions across many distinct organisms.
+Given the difficulties with identifying equivalent IDRs, this may seem like a chicken and egg problem.
+Fortunately IDRs are frequently associated with more conserved folded domains.
+Thus, identifying evolutionarily related sequences by their overall sequence signatures and ``aligning'' can in turn identify IDRs with common ancestry.
+Thus the first step of these analyses is the identification of such proteins with common ancestry, called orthologs.
+Since the rise full genome sequencing in the late 1990s, researchers have developed computational methods for identifying orthologs and aligning them.
+While these methods are generally effective, they are conducted by automated computational pipelines and therefore prone to errors, especially when highly divergent disordered.
+Fortunately over the past five years, advances in DNA sequencing technology has yielded a dramatic increase in the number of genomes available for analysis.
+However, conventional methods have not fully leveraged this increase in information to improve the process of identifying and aligning orthologous sequences.
+Thus, in my first chapter I developed new methods for leveraging this information and apply it them to a set of Drosophila genomes.
+Drosophila is a
+% In my second chapter, I use the alignments generated in the first chapter to analyze IDRs with a variety of evolutionary models.
+Finally, in my third chapter I discuss several general purpose computational tools I developed in the process.
+% X. Finally in the third chapter
 %======= DRAFT =======
@@
 Life is a physical phenomenon.
-Despite the complexity of living things, their processes are governed by the same physical laws that describe planets' motion around the sun and the decay of uranium atoms into lighter elements.
+Despite the complexity of living things, their processes are governed by the same physical laws that describe planets' motion around the sun and the propagation of electromagnetic waves through space.
 However, many systems are too complex to describe with physical models and equations, so scientists simplify them into levels of abstraction that are more useful.

### 2023-03-11 chapter0/chapter0.tex
Added stem for function paragraph and discussion of MoRFs and fuzzy complexes

@@
 
-In contrast, IDRs have no fixed spatial relationship between their atoms and are sometimes described as ``conformational ensembles,'' \textit{i.e.} a collection of related structures.
+The structural diversity of IDRs is directly related to their functional plasticity.
+
+Although the previous discussion has presented of disorder and structure as distinct phenomona , disorder is a spectrum
+MorFs
+folding on binding
+fuzzy complexes
 

### 2023-03-12 chapter0/chapter0.tex
Tweaked wording in outline
Split large paragraph on recognition of disorder and added citation to early example of conformational heterogeneity

@@
 %I.
-Despite significant increases in the accuracy of computational methods which predict disorder from sequence alone, the relationship between sequence and function in IDRs remains poorly understood, and functions are not readily associated with arbitrary IDRs.
+Despite significant advances in predicting disorder from sequence alone and relating those predictions to specific conformational profiles, the relationship between sequence and function in IDRs remains poorly understood.
+Thus, functions are not readily associated with arbitrary IDRs.
 %    A. In contrast, folded domains can identified by characteristic sequences of amino acids
@@
 %        4. Structural and functional information is transferred between proteins with similar sequences
-%        5. In practice, however, this is extremely difficult. Fully mapping the relationship between sequences and structure was an active area of research for decades
+%        5. In practice, however, this is extremely difficult. Fully mapping the relationship between sequence and structure of folded domains was an active area of research for decades
 %            a.
-Recent developments in machine learning have largely resolved this question in most cases, but still required an advanced computational model trained on nearly two hundred thousand experimentally determined structures.
+Recent developments in machine learning have in many senses solved this problem, but still required an advanced computational model trained on nearly two hundred thousand experimentally determined structures.
 %    B. Disordered proteins, however, challenge this sequence-dependent model of protein structure and function.
@@
 The success of structural methods at elucidating the molecular details of protein function cemented the view that function largely depends on the presence of a fixed structure.
-While scientists understood proteins were not completely rigid and could adopt a variety of related structures, many believed that functional proteins largely had a single dominant structure.
+While scientists understood proteins were not completely rigid and could adopt a variety of related structures, many believed that functional proteins largely had a single dominant structure~\cite{Karush1950}.
 Despite its strength, exceptions to the structure-function paradigm were known.
@@
 Furthermore, the flexibility of disordered protein would permit environmental conditions to easily modulate these interactions.
+
 In the following years, as the complete genomes of several scientifically important model organisms such as \textit{S.
@@
 
-Over the following years, many studies investigated the structural and functional properties of IDRs.
+Many studies investigated the structural and functional properties of IDRs over the following years.
 They found that although IDRs still have sequence-structure-function relationships, they play by a very different set of rules.

### 2023-03-12 chapter0/chapter0.tex
Replaced conformational profile with conformational class

@@
 Furthermore, IDRs populate a continuum of structural states over time whereas when folded domains undergo dramatic conformational changes, they are usually triggered by specific environmental signals or chemical modifications, and the intermediate structures are transitory.
-Despite the diversity of conformations available to IDRs, they can be broadly grouped into one of several qualitative descriptions which range from extended ``coils'' to more compact ``globules.'' The specific conformational profile favored by a given IDR, however, is dictated by its local composition of amino acid residues.
+Despite the diversity of conformations available to IDRs, they can be broadly grouped into one of several qualitative descriptions which range from extended ``coils'' to more compact ``globules.'' The specific conformational class of a given IDR, however, is dictated by its local composition of amino acid residues.
 Because the chemical properties of each amino acid in the protein alphabet are dictated its specific arrangement of atoms, each has a unique impact on a protein's structure.
@@
 For example, sequences with high number of either positively or negatively charged residues, called polyelectrolytes, tend to form stiff rods because the like charges repel each other.
-However, if a sequence contains a high number of positively and negatively charged residues in roughly equal proportion, the sequence is called a strong polyampholyte, and its conformational profile depends on the distribution of those charged residues.
+However, if a sequence contains a high number of positively and negatively charged residues in roughly equal proportion, the sequence is called a strong polyampholyte, and its conformational class depends on the distribution of those charged residues.
 If the two classes of residues are segregated into separate blocks of like charges, they attract and form hairpins.
@@
 Thus, like folded domains, the structures of disordered proteins are dictated by their sequences.
-However, because IDRs do not make stable contacts between specific residues in their chains, multiple sequences can generally correspond to a single conformational profile.
+However, because IDRs do not make stable contacts between specific residues in their chains, multiple sequences can generally correspond to a single conformational class.
 

### 2023-03-13 chapter0/chapter0.tex
Added initial aims paragraphs

@@
 %======= DRAFT =======
+
+\begin{abstract}
+\noindent
+Here's where my abstract will go!
+\end{abstract}
+
 \section{Background}
@@
 
-This is where my aims will go!
+Despite recent advances in identifying IDRs and their conformational ensembles from their sequences alone, the relationship between the sequence and function remains poorly understood.
+In contrast, predictions of structure and function are readily available for many folded domains.
+Because they make specific contacts between residues, the sequences of folded domains are generally conserved or evolve slowly.
+Thus, the specific sequence of a folded domain constitutes its unique ``signature.'' Traditional bioinformatics techniques use these signatures to detect similar protein sequences and transfer structural and functional information between them~\cite{Camacho2009, Eddy2009, Mistry2020}.
+Though this approach is simple in principle, it is extremely difficult to perfect in practice, and fully mapping the relationship between the sequence and structure of folded domains alone was an active area of research for decades.
+Part of the challenge is the available data is extremely sparse relative to the sheer number of possible protein sequences.
+Even for a moderately sized protein of 100 amino acid residues, there are $20^{100} \approx 1.3 \times 10^{130}$ possible sequences.
+Only in the past few years have researchers in many senses solved this problem by using advanced techniques from machine learning to leverage the information encoded in nearly two hundred thousand experimentally determined structures~\cite{Jumper2021}.
+
+IDRs, however, challenge this sequence-dependent model of protein structure and function.
+Because they do not make stable contacts between residues which establish a fixed structure, IDRs are not generally constrained to maintain a specific sequence of residues.
+Thus, while there are exceptions, many IDRs evolve extremely rapidly, and related IDRs are therefore not easily identified by their sequence.
+There is growing evidence, though, that IDRs evolve under a different set of constraints.
+Because the composition and patterning of residues in an IDR dictates its conformational class, many distinct sequences can yield a similar set of conformational ensembles.
+Furthermore, because binding motifs and modification sites in IDRs are usually fewer than ten residues, their interaction interfaces are compact and can occur in multiple positions without compromising function~\cite{Tompa2014}.
+Thus, rather than conserving specific sequences, IDRs conserve distributed ``molecular features'' associated with those sequences.
+By the mid 2010s several studies had demonstrated evidence of such constraint in the flexibility, chemical composition, net charge, or charge distribution of IDRs~\cite{Daughdrill2007, Moesa2012, Zarin2017, Beh2012}.
+While these earlier studies were generally restricted to specific features or proteins, in 2019 Zarin \textit{et al.} demonstrated conservation of various features in a comprehensive set of IDR-associated properties across IDRs in the entire yeast proteome~\cite{Zarin2019}.
+By comparing the observed values of these features against those generated under a simulated model of evolution, they clustered IDRs by their ``evolutionary signatures,'' \texit{i.e.} patterns of conserved features.
+Furthermore, they identified specific biological functions associated with these groups, which for the first time provided a global view of the relationship between sequence and function in IDRs.

### 2023-03-13 chapter0/chapter0.tex
Added aims paragraph about implications in new organism

@@
 Furthermore, they identified specific biological functions associated with these groups, which for the first time provided a global view of the relationship between sequence and function in IDRs.
+
+These analyses were conducted using a set of IDRs identified in various species of yeast, which is a widely used model organism in molecular biology research.
+However, no known subsequent studies have determined if similar patterns of conservation are found in the IDRs of other systems.
+As another foundational model organism with abundant genomic information across many evolutionary lineages, the fruit fly, \textit{Drosophila melanogaster}, is a natural choice for subsequent investigation.
+Furthermore, given its complex multicellular development process and shared signaling pathways with humans, the findings of such a study would significantly advance our understanding of the role of IDRs in human health and disease.
+The concordance of these results with the previously identified IDR clusters would also have profound implications for the broader mechanisms of IDR evolution.
+For example, the absence of global patterns of evolutionary signatures across IDRs in \textit{Drosophila} would suggest are property of IDRs which is unique to yeast.
+In contrast, the identification of clusters with similarity to those in yeast would indicate the existence of a taxonomy of IDRs which is conserved across the tree of life.

### 2023-03-14 chapter0/chapter0.tex
Added references for genomic information in aims section

@@
 However, no known subsequent studies have determined if similar patterns of conservation are found in the IDRs of other systems.
-As another foundational model organism with abundant genomic information across many evolutionary lineages, the fruit fly, \textit{Drosophila melanogaster}, is a natural choice for subsequent investigation.
+As another foundational model organism with abundant genomic information across many evolutionary lineages, the fruit fly, \textit{Drosophila melanogaster}, is a natural choice for subsequent investigation~\cite{Yang2018, Miller2018, Kim2021}.
 Furthermore, given its complex multicellular development process and shared signaling pathways with humans, the findings of such a study would significantly advance our understanding of the role of IDRs in human health and disease.

### 2023-03-14 chapter0/chapter0.tex
Added paragraph about comparative genomics to aims

@@
 In contrast, the identification of clusters with similarity to those in yeast would indicate the existence of a taxonomy of IDRs which is conserved across the tree of life.
+
+Though modern genetic engineering techniques enable the direct manipulation of DNA sequences in living systems, gene editing remains a lengthy and work-intensive process in fruit flies.
+Therefore, experimentally testing the vast number of sequences needed to fully map the relationship between distributed features and function in IDRs is infeasible.
+The comparative genomics approach instead leverages the work done by nature to identify evolutionarily conserved features and generate specific hypotheses to guide experiments~\cite{Hardison2003}.
+As life is constantly exploring the space of allowed proteins through evolutionary change, features which are unimportant for a protein's function, or at a larger scale, an organism's survival offer no benefit for their maintenance and are therefore gradually degraded and lost.
+Thus, conservation in related sequences is powerful signal of function.

### 2023-03-14 chapter1/chapter1.tex
Added citation to transcriptomics of original 12 Drosophila genomes

@@
 To develop these methods, we chose a set of 33 assembled and annotated \textit{Drosophila} genomes, which includes all 12 species from the original \textit{Drosophila} 12 Genomes Consortium~\cite{D12GC2007}.
-However, the genomes of these 12 species have been re-sequenced since their first release, which has resulted in substantial improvements in their assemblies and annotations.
-Despite these developments and other several other recent genome assembly projects of species in the \textit{Drosophila} genus, there is not yet a collection of high-quality alignments of orthologous proteins that reflects these improvements in genome assembly and diversity~\cite{Miller2018, Kim2021}.
+However, the genomes of these 12 species have been re-sequenced since their first release, which has resulted in substantial improvements in their assemblies and annotations~\cite{Yang2018, Miller2018}.
+Despite these developments and other several other recent genome assembly projects of species in the \textit{Drosophila} genus, there is not yet a collection of high-quality alignments of orthologous proteins that reflects these improvements in genome assembly and diversity~\cite{Kim2021}.
 Given the \textit{Drosophila} genus spans diverse habitats and over 50 million years of evolution but maintains a conserved life cycle and body plan, such a resource would facilitate a new generation of studies that illuminate the forces that drive protein evolution in unprecedented detail~\cite{Wiegmann2011, Obbard2012}.

### 2023-03-14 chapter0/chapter0.tex
Fixed some typos in paragraph about implications of aims

@@
 The concordance of these results with the previously identified IDR clusters would also have profound implications for the broader mechanisms of IDR evolution.
-For example, the absence of global patterns of evolutionary signatures across IDRs in \textit{Drosophila} would suggest are property of IDRs which is unique to yeast.
-In contrast, the identification of clusters with similarity to those in yeast would indicate the existence of a taxonomy of IDRs which is conserved across the tree of life.
+For example, the absence of global patterns of evolutionary signatures across IDRs in \textit{Drosophila} would suggest they are property of IDRs which is unique to yeast.
+In contrast, the identification of clusters similar to those in yeast would indicate the existence of a taxonomy of IDRs which is conserved across the tree of life.
 

### 2023-03-14 chapter0/chapter0.tex
Added final paragraph to introduction

@@
 Thus, conservation in related sequences is powerful signal of function.
+
+These comparisons, however, require the identification of IDRs with common ancestry that perform ``equivalent'' functions across many distinct organisms.
+Given the difficulties with identifying similar IDRs by their sequences, this may seem like a chicken and egg problem.
+Fortunately, IDRs are frequently associated with more conserved folded domains.
+Thus, identifying evolutionarily related proteins by their overall sequence signatures and ``aligning'' them will in turn identify the equivalent IDRs in those sequences.
+The first step of an evolutionary analysis of IDRs, then, is the identification of proteins with common ancestry, called orthologs.
+Since the first genomes were sequenced in the late 1990s, researchers have developed techniques for identifying and aligning orthologs~\cite{Fleischmann1995, Goffeau1996, CESC1998, Tatusov1997}.
+While these methods are generally effective, they are conducted by automated computational pipelines and prone to errors when processing the highly divergent sequences that characterize many IDRs.
+The evolutionary relationships between the genomes of closely related species generally make such mistakes easier to identify, and fortunately over the past five years advances in DNA sequencing technology have yielded dramatic increases in the number of sequenced genomes in the \textit{Drosophila} genus.
+However, because the existing methods for ortholog identification were designed for fewer or more distantly related genomes, they do not fully leveraged the available genomic redundancy to minimize errors.
+Thus, in the first chapter I develop a novel method for identifying orthologs which addresses this shortcoming and apply it to 33 \textit{Drosophila} genomes to generate a set of aligned orthologs.
+In the second chapter I then identify rapidly evolving IDRs in these ``alignments'' and analyse them with a variety of evolutionary models to detect patterns of conservation.
+Finally, in the third chapter I discuss several software tools and tutorials for fitting statistical models to data, which were created while pursuing the previous aims.

### 2023-03-15 chapter0/chapter0.tex
Added aims overview figure

@@
 
+\begin{figure}[h!]
+\includegraphics[width=\textwidth]{out/aims_overview.png}
+\centering
+\caption{\textbf{Graphical overview of aims.}
+\textbf{(A)} The large open circles represent the genomes of different \textit{Drosophila} species, and the small, filled circles represent protein sequences in those genomes.
+Orthologs are colored with shades of the same hue.
+\textbf{(B-D)} Each horizontal line represents a sequence in panel A, and together each set of lines creates an ``alignment.'' The lines are broken where ``gaps'' are inserted to align equivalent segments.
+The grey segments denote regions of the alignment which are not part of subsequent analyses.
+\textbf{(E)} Each aligned IDR is scored on four features, where the ``strength'' of that feature is indicated by the height and color of its bar.
+\textbf{(F)} Scoring many IDRs on these features yields three distinct clusters.}
+\label{fig:aims_overview}
+\end{figure}
+
 These comparisons, however, require the identification of IDRs with common ancestry that perform ``equivalent'' functions across many distinct organisms.
@@
 However, because the existing methods for ortholog identification were designed for fewer or more distantly related genomes, they do not fully leveraged the available genomic redundancy to minimize errors.
-Thus, in the first chapter I develop a novel method for identifying orthologs which addresses this shortcoming and apply it to 33 \textit{Drosophila} genomes to generate a set of aligned orthologs.
-In the second chapter I then identify rapidly evolving IDRs in these ``alignments'' and analyse them with a variety of evolutionary models to detect patterns of conservation.
+Thus, in the first chapter I develop a novel method for identifying orthologs which addresses this shortcoming and apply it to 33 \textit{Drosophila} genomes to generate a set of aligned orthologs (Fig.~\ref{fig:aims_overview}A-B).
+In the second chapter I then identify rapidly evolving IDRs in these ``alignments'' and analyse them with a variety of evolutionary models to detect patterns of conservation (Fig.~\ref{fig:aims_overview}C-F).
 Finally, in the third chapter I discuss several software tools and tutorials for fitting statistical models to data, which were created while pursuing the previous aims.

### 2023-03-15 chapter0/chapter0.tex
Changed transitory to more common synonym transient

@@
 In contrast, IDRs have no fixed spatial relationship between their atoms, so their structures are sometimes described as ``conformational ensembles,'' \textit{i.e.} collections of conformations where the distances and orientations between residues can vary considerably.
-Furthermore, IDRs populate a continuum of structural states over time whereas when folded domains undergo dramatic conformational changes, they are usually triggered by specific environmental signals or chemical modifications, and the intermediate structures are transitory.
+Furthermore, IDRs populate a continuum of structural states over time whereas when folded domains undergo dramatic conformational changes, they are usually triggered by specific environmental signals or chemical modifications, and the intermediate structures are transient.
 Despite the diversity of conformations available to IDRs, they can be broadly grouped into one of several qualitative descriptions which range from extended ``coils'' to more compact ``globules.'' The specific conformational class of a given IDR, however, is dictated by its local composition of amino acid residues.

### 2023-03-15 chapter1/figures/hmm/hmm.py
Renamed src/ to src1/ to allow srcs for different chapters

@@ -8,9 +8,9 @@ import homomorph
 import matplotlib.pyplot as plt
 import numpy as np
 import skbio
-from src.draw import plot_msa_data
-from src.ortho_MSA import phylo
-from src.utils import read_fasta
+from src1.draw import plot_msa_data
+from src1.ortho_MSA import phylo
+from src1.utils import read_fasta
 
 OGids = ['2252', '2A57', '360E']
 msa_path = '../../orthology_inference/analysis/ortho_MSA/realign_fastas/out/'

### 2023-03-15 chapter1/figures/realignment/realignment.py

@@ -6,8 +6,8 @@ import re
 import matplotlib.pyplot as plt
 import numpy as np
 import skbio
-from src.draw import plot_msa
-from src.utils import read_fasta
+from src1.draw import plot_msa
+from src1.utils import read_fasta
 
 records = [('0167', 'A', 2357, 2714, 2763, 2831),
            ('2770', 'B', 2440, 2831, 2235, 2743),

### 2023-03-15 chapter1/figures/trees/trees.py

@@ -5,7 +5,7 @@ import os
 import matplotlib.pyplot as plt
 import skbio
 from matplotlib import rcParams
-from src.draw import plot_tree
+from src1.draw import plot_tree
 
 if not os.path.exists('out/'):
     os.mkdir('out/')

### 2023-03-15 chapter3/figures/figures.py
Removed chapter3/figures utils.py and linked to homomorph copy
The homomorph copy is essentially identical except the deltas in the print statements are the actual symbols.

@@ -13,7 +13,7 @@ import scipy.stats as stats
 from homomorph import HMM
 from numpy import exp, log
 from sklearn.metrics import roc_curve
-from utils import fit_CML
+from src3.utils import fit_CML
 
 dpi = 300
 legend_kwargs = {'frameon': False, 'loc': 'center left', 'bbox_to_anchor': (1, 0.5)}

### 2023-03-15 chapter3/figures/utils.py

@@ -1,266 +1,0 @@
-"""Utility functions for training tutorial."""
-
-import pprint
-from copy import deepcopy
-from functools import reduce
-from itertools import accumulate
-
-import homomorph
-import numpy as np
-from numpy import exp, log
-
-
-class ArrayRV:
-    """Class that allows HMM class to interface with a pre-computed probability array.
-
-    The rvs method is only defined so the HMM recognizes it as a proper
-    random variable. Since random variates are not needed in this script, the
-    body of the method is left blank.
-    """
-    def __init__(self, array):
-        self.array = array
-
-    def pmf(self, x):
-        return self.array[x]
-
-    def rvs(self, random_state=None):
-        pass
-
-
-class HMM(homomorph.HMM):
-    """A class that modifies the default HMM to accept pre-calculated intermediates.
-
-    Discriminative training requires decoding of both states and transitions.
-    Both of these calculations rely on the forward and backward variables.
-    Since these are the most computationally intensive steps in decoding and
-    both are used in state and transition decoding, this modified class accepts
-    the forward and backward variables as arguments rather than calculating
-    them from scratch each time.
-    """
-    def forward_backward1(self, emits, fs, ss_f, bs, ss_b):
-        """Posterior decoding of states."""
-        p = reduce(lambda x, y: x+y, map(log, ss_f))
-        ss_f = list(accumulate(map(log, ss_f)))
-        ss_b = list(accumulate(map(log, ss_b[::-1])))[::-1]
-
-        fbs = {state: [] for state in self.states}
-        for i in range(len(emits)):
-            for state, fb in fbs.items():
-                fbs[state].append(fs[state][i]*bs[state][i]*exp(ss_f[i]+ss_b[i]-p))
-
-        return fbs
-
-    def forward_backward2(self, emits, fs, ss_f, bs, ss_b):
-        """Posterior decoding of transitions."""
-        p = reduce(lambda x, y: x+y, map(log, ss_f))
-        ss_f = list(accumulate(map(log, ss_f)))
-        ss_b = list(accumulate(map(log, ss_b[::-1])))[::-1]
-
-        pairs = [(state0, state1) for state0, t_dist in self.t_dists.items() for state1 in t_dist]
-        fbs = {pair: [] for pair in pairs}
-        for i in range(len(emits)-1):
-            for (state0, state1), fb in fbs.items():
[…]

### 2023-03-15 chapter1/chapter1.tex
Moved exported PNGs of some SVGs to out/ folders and added SVGs
Removed Tensorflow as citation in chapter 1

@@
 \begin{figure}[h!]
-\includegraphics[width=\textwidth]{percolation/percolation.png}
+\includegraphics[width=\textwidth]{percolation/out/percolation.png}
 \includegraphics[width=\textwidth]{percolation/out/005F.png}
@@
 \begin{figure}[h!]
-\includegraphics[width=\textwidth]{hmm/hmm_architecture.png}
+\includegraphics[width=\textwidth]{hmm/out/hmm_architecture.png}
 \includegraphics[width=\textwidth]{hmm/out/2252.png}
@@
 HMM algorithms were implemented in the standalone package Homomorph which is available at \url{https://github.com/marcsingleton/homomorph} and on the Python Package Index (PyPI).
-The following Python libraries were used: matplotlib, NumPy, pandas, SciPy, and TensorFlow~\cite{Hunter2007, Harris2020, McKinney2010, Virtanen2020, Abadi2015}.
+The following Python libraries were used: matplotlib, NumPy, pandas, and SciPy~\cite{Hunter2007, Harris2020, McKinney2010, Virtanen2020}.
 Relevant output files are available in the supporting information.

### 2023-03-16 chapter0/chapter0.tex
Tweaked wording to remove phrase "structured domains" as it was not formally introduced yet

@@
 Because these segments were disordered in their native state, \textit{i.e.} under normal operating conditions, such segments were termed ``intrinsically disordered regions'' (IDRs) to emphasize the disorder was not induced by exposure to chemicals or heat.
-Furthermore, while most proteins were predicted to contain a mixture of structured domains and IDRs, some ``intrinsically disordered proteins'' (IDPs) were entirely or almost entirely disordered.
+Furthermore, while most proteins were predicted to contain a mixture of structure and disorder, some ``intrinsically disordered proteins'' (IDPs) were entirely or almost entirely disordered.
 Thus, disorder was recognized as a pervasive but poorly understood feature of proteins.

### 2023-03-17 chapter0/chapter0.tex
Added abstract

@@
 \noindent
-Here's where my abstract will go!
+Living systems are governed by the interactions between large collections of atoms called macromolecules.
+The most important class of these macromolecules are proteins, which are the molecular machines that carry out a cell's processes.
+Though proteins are linear chains of simpler building blocks called amino acids, many proteins accomplish their functions by ``folding'' into well-defined three-dimensional structures.
+This stability makes folded proteins easier to study with various biochemical and bioinformatics techniques, so for many years scientists believed fixed structures were necessary for protein function.
+By the early 2000s, however, evidence had accumulated that unstructured or ``intrinsically disordered'' regions (IDRs) were ubiquitous in proteins.
+Furthermore, despite rapidly fluctuating through a range of configurations, these regions were essential for many cellular processes such as signaling and regulation.
+Although our understanding of the structure and function of IDRs has grown significantly over the past two decades, predicting their functions from their sequences of amino acids remains a significant challenge.
+Because IDRs are structurally unconstrained, their sequences evolve rapidly and are therefore not amenable to traditional bioinformatics analyses which depend on the precise order of amino acids to make comparisons with known proteins.
+There is increasing evidence, though, that IDRs conserve distributed features such as their chemical composition or net charge, and a recent study clustered IDRs with similar patterns of conserved features into groups with distinct functions.
+This study, however, was restricted to IDRs in a set of yeast genomes, so it is unclear if these global relationships between conserved features and function are unique to yeast or a general property of IDR evolution.
+Thus, in this work I conduct a similar set of evolutionary analyses of IDRs in the genomes of 33 different species of fruit flies, which another essential model organism in molecular biology and genetics research.
+The analysis is divided into two major sections.
+In the first, sets of proteins with common ancestry are identified in these genomes, and in the second IDRs in these sets of proteins are analysed with a variety of evolutionary models to detect patterns of conservation.
 \end{abstract}

### 2023-03-17 chapter0/chapter0.tex
Trimmed abstract of some supporting details

@@
 Though proteins are linear chains of simpler building blocks called amino acids, many proteins accomplish their functions by ``folding'' into well-defined three-dimensional structures.
-This stability makes folded proteins easier to study with various biochemical and bioinformatics techniques, so for many years scientists believed fixed structures were necessary for protein function.
-By the early 2000s, however, evidence had accumulated that unstructured or ``intrinsically disordered'' regions (IDRs) were ubiquitous in proteins.
-Furthermore, despite rapidly fluctuating through a range of configurations, these regions were essential for many cellular processes such as signaling and regulation.
+For many years scientists believed fixed structures were necessary for protein function, but by the early 2000s, however, evidence had accumulated that unstructured or ``intrinsically disordered'' regions (IDRs) were ubiquitous in proteins.
+Furthermore, these regions were essential for many cellular processes such as signaling and regulation.
 Although our understanding of the structure and function of IDRs has grown significantly over the past two decades, predicting their functions from their sequences of amino acids remains a significant challenge.
@@
 This study, however, was restricted to IDRs in a set of yeast genomes, so it is unclear if these global relationships between conserved features and function are unique to yeast or a general property of IDR evolution.
-Thus, in this work I conduct a similar set of evolutionary analyses of IDRs in the genomes of 33 different species of fruit flies, which another essential model organism in molecular biology and genetics research.
-The analysis is divided into two major sections.
-In the first, sets of proteins with common ancestry are identified in these genomes, and in the second IDRs in these sets of proteins are analysed with a variety of evolutionary models to detect patterns of conservation.
+Thus, in this work I conduct a similar set of evolutionary analyses of IDRs in the genomes of 33 different species of fruit flies.
 \end{abstract}

### 2023-03-17 chapter0/chapter0.tex
Added paragraph on SLiMs

@@
 
-The structural diversity of IDRs is directly related to their functional plasticity.
+The structural diversity of IDRs is directly related to their functional plasticity, and as IDRs are not confined to one conformation, they can interact with and bind to many possible partners.
+Often these partners are other proteins, but they can also be other macromolecules like DNA or even small molecules and ions.
+As a result of this adaptability, IDRs are enriched in proteins involved in cell signaling and regulation.
+Because cells are highly compartmentalized, they have a dizzying array of mechanisms to relay messages.\footnote{The most fundamental compartment is the cell, which roughly separates inside from outside and life from non-life.
+However, the cells of more complex organisms called eukaryotes contain additional subcompartments called organelles.
+Compartmentalization is essential in living systems because it confines and concentrates biochemical reactions that would be harmful if they occurred at the wrong place or time.
+However, it also introduces many complications because information in the form of physical or chemical signals cannot travel freely.} Many are mediated by the interaction of one protein with another, which in turn creates a change in the state of the system that propagates the signal.
+Often these state changes are chemical alterations made by one protein to another, which are termed ``post-translational modifications'' (PTMs).\footnote{A protein's amino acid sequence is encoded in a cell's DNA, and the process of reading that information to create a protein (from an intermediate molecule called RNA) is translation.
+Therefore, any modifications to a protein after its initial synthesis are post-translational.} One of the most common modifications is phosphorylation where a phosphate group is attached to a specific amino acid in the protein.
+Phosphate groups contain three negative charges, so their addition can dramatically affect a protein's structural energetics and induce a conformational change.
+Thus, phosphorylation often plays a key role in toggling proteins between inactive and active states.
+In general, though, PTMs encompass a variety of chemical modifications with similarly diverse impacts on a protein's behavior.
+Because IDRs are generally exposed to their environment, they are frequent targets of PTMs.
+However, PTMs do not occur haphazardly in proteins but are instead targeted to binding sites created by sequential patterns of residues called short linear motifs (SLiMs).
+SLiMs are a general mechanism for mediating interactions with proteins, so while many SLiMs are targets of PTMs, others simply recruit binding partners.
+In contrast to the highly structured interfaces that characterize folded domains, however, SLiMs are usually no more than ten residues.
+Thus, their interactions are relatively weak and therefore highly transient.
+As their flexibility makes SLiMs easily accessible to binding partners, they are highly enriched in IDRs.
 

### 2023-03-17 chapter0/chapter0.tex
Added some unorganized sentences to finish SLiMs paragraph and begin TF paragraph

@@
 
-Although the previous discussion has presented of disorder and structure as distinct phenomona , disorder is a spectrum
+SLiMs can also occur combinatorially.
+Furthermore, SLiMs are easily tuned because PTMs or evolution can change their properties by modifying only a few residues.
+This dynamism and modularity is essential in signalling as it allows the efficient propagation and modulation of information.
+Thus, they often function as hubs in networks of signaling pathways [CITATION]
+Some IDRs are roughly like molecular computers which integrate signals from varying sources
+As IDRs are highly flexible, they can bind to many partners, sometimes simultaneously, and therefore often act as hubs in these signaling networks.
+
+Some of the most prominent examples of IDRs in cell signaling and regulation are the activation domains of transcription factors.
+Each cell expresses a unique complement of proteins that in large part determines its identity.
+
+Although the previous discussion has presented of disorder and structure as distinct phenomona, disorder is a spectrum
 MorFs

### 2023-03-17 main.tex
Added \newpage command to ensure page counter is not set for TOC
Changed figure command to use chapter.fignum style
Added chapter 2 appendix
Added title

@@
 \usepackage[labelfont=bf, labelsep=period]{caption}
-\renewcommand{\figurename}{Fig}
+\renewcommand{\thefigure}{\thechapter.\arabic{figure}}
 
@@
 \begin{center}
-    My long and grand sounding dissertation title\\
+    Finding structure in disorder: Evolutionary analyses of disordered proteins in \textit{Drosophila}\\
     \bigskip
@@
     \bigskip
-    My long and grand sounding dissertation title\\
+    Finding structure in disorder: Evolutionary analyses of disordered proteins in \textit{Drosophila}\\
     by\\
@@
 \tableofcontents
+\newpage  % End page so numbering is set properly
+\setcounter{chapter}{-1}  % Set chapter counter to 0
+\setcounter{page}{1}  % Reset page counter after pre-main content
+\pagenumbering{arabic}  % Set numbering style to arabic for main content
 
-\setcounter{chapter}{-1}  % Set chapter counter to 0
 \chapter{Introduction}
 \graphicspath{{chapter0/figures/}}
-\setcounter{page}{1}  % Reset page counter after pre-main content
-\pagenumbering{arabic}  % Set numbering style to arabic for main content
 \input{chapter0/chapter0.tex}
@@
 
+\chapter{Supporting information for chapter 2}
+\label{appendix:b}
+\graphicspath{{chapter2/figures/}}
+\input{chapter2/chapter2_supporting.tex}
+
 \chapter{Supporting information for chapter 3}
-\label{appendix:b}
+\label{appendix:c}
 \setlength{\promptwidth}{0in}  % No indent for appendix code

### 2023-03-17 chapter2/chapter2.tex
Added some sections to materials and methods of chapter 2

@@
 \section{Materials and methods}
-Here's where my materials and methods will go!
+
+\subsection{Alignment and tree generation}
+
+\subsection{IDR prediction and filtering}
+
+\subsection{Fitting substitution models}
+
+\subsection{Brownian motion and Ornstein-Uhlenbeck analyses}
+
+\subsection{Simulation analyses}
+
+\subsection{Code and data availability}
+The code used to produce the results and analyses is available at \url{https://github.com/marcsingleton/IDR_evolution2023}.
+The following Python libraries were used: matplotlib, NumPy, pandas, and SciPy~\cite{Hunter2007, Harris2020, McKinney2010, Virtanen2020}.
+Relevant output files are available in the supporting information.
+There is no primary data associated with this manuscript.
+All primary data are available from publicly accessible sources described in their corresponding sections.

### 2023-03-17 chapter0/chapter0.tex
Moved parenthetical sentence about "applied sciences" into footnote

@@
 Despite the complexity of living things, their processes are governed by the same physical laws that describe planets' motion around the sun and the propagation of electromagnetic waves through space.
-However, many systems are too complex to describe with physical models and equations, so scientists simplify them into levels of abstraction that are more useful.
-(This is the origin of the common observation that biology is applied chemistry and chemistry is applied physics.) For example, Punnett squares facilitate the prediction of genotypes and phenotypes by distilling the complexities and nuances of diverse reproductive systems into a set of simple rules.
+However, many systems are too complex to describe with physical models and equations, so scientists simplify them into levels of abstraction that are more useful.\footnote{This is the origin of the common observation that biology is applied chemistry and chemistry is applied physics.} For example, Punnett squares facilitate the prediction of genotypes and phenotypes by distilling the complexities and nuances of diverse reproductive systems into a set of simple rules.
 However, since life spans a scale from single cells to entire ecosystems, biology likely employs more layers of abstraction than any other scientific discipline.

### 2023-03-17 chapter0/chapter0.tex
Finished SLiM paragraph!

@@
 In contrast to the highly structured interfaces that characterize folded domains, however, SLiMs are usually no more than ten residues.
-Thus, their interactions are relatively weak and therefore highly transient.
-As their flexibility makes SLiMs easily accessible to binding partners, they are highly enriched in IDRs.
-
-SLiMs can also occur combinatorially.
-Furthermore, SLiMs are easily tuned because PTMs or evolution can change their properties by modifying only a few residues.
-This dynamism and modularity is essential in signalling as it allows the efficient propagation and modulation of information.
-Thus, they often function as hubs in networks of signaling pathways [CITATION]
-Some IDRs are roughly like molecular computers which integrate signals from varying sources
-As IDRs are highly flexible, they can bind to many partners, sometimes simultaneously, and therefore often act as hubs in these signaling networks.
+Thus, their interactions are relatively weak and highly transient.
+As their flexibility makes SLiMs easily accessible, they are also enriched in IDRs, which in turn allows them to bind to many partners, sometimes simultaneously.
+IDRs therefore often act as hubs in complex regulatory networks by propagating signals from diverse sources or organizing binding partners into higher-order structures.
+Furthermore, these signals and interactions are easily tuned because PTMs can modulate their meanings and strengths, respectively, by modifying only a few residues.
+Thus, IDRs are like ``molecular computers'' that integrate complex data and respond accordingly to execute different genetic ``programs.''
 

### 2023-03-17 chapter0/chapter0.tex
Removed most of outline and moved remaining sections to their positions in the text
Tweaked some wording, mostly removing unnecessary uses of "set of"

@@
-%======= OUTLINE =======
-%
-%BACKGROUND
-%Survery of historical developments that inform dissertation
-%Write at a high school or early college biology level
-%
-%I. Proteins are the molecular machines that drive biology
-%    A. Four major macromolecules and briefly discuss broad roles of each
-%        1. Nucleic acids
-%        2. Carbohydrates
-%        3. Lipids
-%        4. Protein
-%    B. Some major examples of proteins
-%        1. Hemoglobin and sickle cell
-%        2. Color blindness and opsins
-%        3. Lactase and lactose intolerance
-%II. Proteins are linear chains of amino acids with a three-dimensional structure
-%    A. Background on biochemistry, primary, secondary, and tertiary structure
-%    B. Structure-function relationships returning to previous hemoglobin example
-%    C. Structural biology seeks to determine protein structures to understand function
-%    D. Overview of structural methods including purification steps
-%        1. X-ray crystallography
-%        2. NMR
-%        3. cryo-EM
-%III. IDRs do not adopt a stable structure
-%    A. History of disordered proteins and initial recognition
-%        1. Success of structural methods
-%    B. Disordered proteins are challenging to study with traditional methods
-%    C. Genome sequencing and computational tools revealed disordered proteins are ubiquitous
-%IV. IDRs have sequence-structure relationships (of a different kind)
-%    A. Prompted by these observations, researchers began to clarify the sequence-structure-function relationships that characterize protein disorder
-%        1. They discovered that related to structure and sequence but the details differ
-%    B. Though structured proteins are not completely rigid, they generally have a small number of structures
-%        1. As mentioned previously, even before disorder was recognized as a common feature of proteins, scientists understood proteins were not completely rigid
-%        2. At the molecular level, everything is in constant motion. For example, at room temperature the average water molecule moves at a speed of SPEED.
-%            i. However, collisions and interactions with other water molecules generally prevent them from escaping into the atmosphere
-%        3. Proteins are also in motion, but they are constrained both by the rigid bonds and interactions between amino acid residues in the chain
-%        4. Thus, all proteins flex and breathe. However, these motions are minor variations on the stable structure.
-%        5. In other cases, proteins have multiple structures which are related to different functional states.
-%            i. For example, hemoglobin has two forms, traditionally called T and R.
-%            ii. The T form is hemoglobin's oxygen-free state.
-%            iii. On binding oxygen, its structure changes slightly, with the shift on the scale of a few hydrogen atoms at the points of greatest change.
-%            iv. This changes the orientation between the atoms in hemoglobin that interact with oxygen, allowing it to bind more tightly.
-%            v. Promotes oxygen uptake at other three binding sites, promoting efficiently transport
-%        6. Although proteins have multiple conformations, they are generally stable and the intermediates are not.
-%        7. Thus, proteins with multiple conformations generally exist in discrete states.
-%    C. Protein disorder means their atoms do not have fixed spatial relationships
-%       1. In contrast, disordered proteins are described as ``conformational ensembles''
-%       2. They are constantly sampling conformations which may have large differences in the spatial relationships between residues
-%       3. The conformations of disordered regions may adopt more extended conformations or collapsed conformations
-%       3. The composition of a disordered region dictates its conformational profile. However, to describe
-%       4. As mentioned previously, proteins are linear chains of amino acids, and the amino alphabet contains 20 letters
-%       5. Though each amino acid is chemically unique, many of their properties can be understood in terms of
-%       1. Disordered regions largely characterized by enrichment and depletion in hydrophilic and hydrophobic residues, respectively
-%       2. Hydrophobic residues are ``water-fearing,'' and like oil in water, they tend to self-associate.
-%       3. Hydrophilic residues are ``water-loving,'' and easily interact with water.
-%       4. Hydrophobic interacts are a major force driving protein folding, and many folded proteins have a hydrophobic core
-%    E. Sequence-structure relationships in disordered proteins
-%       1. Although, disordered regions are conformationally heterogeneous, they
-%V. IDRs are functional
-%    A. Flexibility confers several advantages
-%        1. The structural flexibility of IDRs is directly related their functional plasticity
-%        2. Proteins involved in signaling, transport, and catalysis require specific geometries to interact their intended targets efficiently
-%        3. Disordered proteins can interact with multiple targets
-%            i. Environmental signals can fine-tune which interacts are preferred
-%            ii. Proteins, structured and disordered, frequently undergo modifications which change their behavior
-%            iii. A common modification is phosphorylation where a phosphate group is attached to a specific amino acid in the protein
-%                a. Phosphate groups contain three negative charges, so the addition can dramatically affect a protein's folding energetics and induce a change in structure
-%            iv. Thus, phosphorylations commonly toggle proteins between an inactive and active state
-%                a. Proteins are phosphorylated by a class of proteins called kinases, which are in turn activated by interactions with or modifications by other proteins
-%                b. A single stimulus can set off a sequence of events, called a signaling cascade, spanning hundreds of reactions between as many distinct proteins
-%                c.
-Biologists are sometimes descriptive to a fault, resulting in proteins like MAP kinase kinase kinase which phosphorylates MAP kinase kinase which phosphorylates MAP kinase which phosphorylates the MAP protein.
-[FOOTNOTE]
-%            v. Because IDRs expose much of their chain to the surrounding environment, they have many sites which are available for modification
-%                a. Thus, they often function as hubs in networks of signaling pathways [CITATION]
-%                b. Some IDRs are roughly like molecular computers which integrate signals from varying sources
-%    B. Survey of disordered proteins and their functions
-%        1. Linkers
-%        2. Signaling
-%VI. IDRs are enriched in transcriptional machinery
-%    A. Introduce transcription via the central dogma
-%    B. Describe transcription factors, cofactors, transcriptional machinery
-%        1. Examples of different transcription factors
-%    C. Describe properties of transcription factors
-%        1. DNA binding and activation domains
-%        2. Different flavors of activation domains
-%    D. How transcription factors find their targets to initiate specific genetic programs is one of the major outstanding questions in molecular biology
-%VII. IDRs are associated with phase separation
-%    A. Describe phase separation in general
-%    B. Describe phase separation as related to proteins
-%    C. Describe functions of phase separation and relationship to subcellular structures
-%VIII. IDR-mediated phase separation is a possible mechanism for initiating transcription
-%    A. Summarize Stadler's papers and others
-%
-%AIMS
-%I.
-Despite significant advances in predicting disorder from sequence alone and relating those predictions to specific conformational profiles, the relationship between sequence and function in IDRs remains poorly understood.
-Thus, functions are not readily associated with arbitrary IDRs.
-%    A. In contrast, folded domains can identified by characteristic sequences of amino acids
-%        1. Because structured domains make specific contacts between residues, their sequences of amino acids are generally conserved or evolve slowly
-%        2. Thus, for structured domains their specific sequence of amino acids constitute a unique signature
-%        3. Traditional bioinformatics methods use these signatures to detect similarities between proteins
-%        4. Structural and functional information is transferred between proteins with similar sequences
-%        5. In practice, however, this is extremely difficult. Fully mapping the relationship between sequence and structure of folded domains was an active area of research for decades
-%            a.
-Recent developments in machine learning have in many senses solved this problem, but still required an advanced computational model trained on nearly two hundred thousand experimentally determined structures.
-%    B. Disordered proteins, however, challenge this sequence-dependent model of protein structure and function.
-%    X. Though there are exceptions, many IDRs evolve extremely rapidly and are not identifiable by the specific sequence of amino acids
-%    C. However, growing evidence that IDRs evolve under a different set of constraints
-%        0. Rather than preserving a specific sequence in order to preserve contacts and structure, they preserve distributed features such as composition, length, charge, or the presence of motifs
-%        1. Because many sequences can yield an IDR with a specific conformational profile or local arrangements, they are not constrained to preserve the precise order of amino acid residues
-%    D. Examples of studies that have found specific cases of conservation in IDRs
-%        1. Earlier specific examples of feature conservation in proteins
-%        2.
-However, a recent study from Zarin et al was the first to demonstrate of widespread of feature conservation in IDRs across an entire proteome, that is the set of all proteins found in an organism
-%    E.
-The analyses used simulations and evolutionary comparisons across a set of yeast genomes, a foundational model organism in molecular biology research.
-However, no subsequent studies have investigated similar patterns of conservations, or any patterns of conservation at all, characterize in IDRs in other systems.
-%        0. No conservation global patterns of conservation would suggest such preserved conservation are unique to yeast
-%        1. Patterns of conservation but are distinct from yeast would indicate global patterns
-%        2. Similar patterns of conservation with similar functional associations would indicate a universal taxonomy of IDRs
-%II. Use evolutionary analyses to identify conserved properties of disordered regions
-%    A.
-Though modern molecular genetic and gene editing techniques it is in principle possible to test the relationship between distributed features and the structure and function of specific proteins in living systems directly, the vast number of combinations make this approach infeasible in practice.
-%    B. Instead life constantly explores the space of allowed proteins through evolutionary changes
-%    C. Properties which are unimportant for a protein's function, or at a larger scale, an organism's survival offer no benefit for their maintenance and are gradually degraded and ultimately lost
-%    A. Thus, conservation is powerful signal of function, which are illuminated by evolutionary comparisons between related sequences.
-%    X.
-Such comparisons, however, require the identification of IDRs with common ancestry that perform ``equivalent'' functions across many distinct organisms.
-Given the difficulties with identifying equivalent IDRs, this may seem like a chicken and egg problem.
-Fortunately IDRs are frequently associated with more conserved folded domains.
-Thus, identifying evolutionarily related sequences by their overall sequence signatures and ``aligning'' can in turn identify IDRs with common ancestry.
-Thus the first step of these analyses is the identification of such proteins with common ancestry, called orthologs.
-Since the rise full genome sequencing in the late 1990s, researchers have developed computational methods for identifying orthologs and aligning them.
-While these methods are generally effective, they are conducted by automated computational pipelines and therefore prone to errors, especially when highly divergent disordered.
-Fortunately over the past five years, advances in DNA sequencing technology has yielded a dramatic increase in the number of genomes available for analysis.
-However, conventional methods have not fully leveraged this increase in information to improve the process of identifying and aligning orthologous sequences.
-Thus, in my first chapter I developed new methods for leveraging this information and apply it them to a set of Drosophila genomes.
-Drosophila is a
-% In my second chapter, I use the alignments generated in the first chapter to analyze IDRs with a variety of evolutionary models.
-Finally, in my third chapter I discuss several general purpose computational tools I developed in the process.
-% X. Finally in the third chapter
-%======= DRAFT =======
-
 \begin{abstract}
@@
 This study, however, was restricted to IDRs in a set of yeast genomes, so it is unclear if these global relationships between conserved features and function are unique to yeast or a general property of IDR evolution.
-Thus, in this work I conduct a similar set of evolutionary analyses of IDRs in the genomes of 33 different species of fruit flies.
+Thus, in this work I conduct a series of evolutionary analyses of IDRs in the genomes of 33 different species of fruit flies to detect patterns of conservation.
 \end{abstract}
@@
 In the proteasome, however, these are distinct steps.
-The motor subunits in the base first physically interact with the protein to simultaneously unfold and pull it into the base.
-A different set of subunits in the base then break the exposed chemical bonds between amino acid residues in the chain.} Like a paper shredder, the motor is only engaged when a protein is correctly positioned in the lid.
+The motor subunits in the base first physically interact with the protein to simultaneously unfold and pull it into the core.
+Different subunits in the core then break the exposed chemical bonds between amino acid residues in the chain.} Like a paper shredder, the motor is only engaged when a protein is correctly positioned in the lid.
 Unlike a paper shredder, however, the motor is activated by a conformational change rather than a physical switch.
@@
 
-Although the previous discussion has presented of disorder and structure as distinct phenomona, disorder is a spectrum
-MorFs
-folding on binding
-fuzzy complexes
+% IDRs are enriched in transcriptional machinery
+%    A. Introduce transcription via the central dogma
+%    B. Describe transcription factors, cofactors, transcriptional machinery
+%        1. Examples of different transcription factors
+%    C. Describe properties of transcription factors
+%        1. DNA binding and activation domains
+%        2. Different flavors of activation domains
+%    D. How transcription factors find their targets to initiate specific genetic programs is one of the major outstanding questions in molecular biology
+
+% Disorder is spectrum
+%    A. MorFs, folding on binding, fuzzy complexes
 
@@
 There is growing evidence, though, that IDRs evolve under a different set of constraints.
-Because the composition and patterning of residues in an IDR dictates its conformational class, many distinct sequences can yield a similar set of conformational ensembles.
+Because the composition and patterning of residues in an IDR dictates its conformational class, many distinct sequences can yield similar conformational ensembles.
 Furthermore, because binding motifs and modification sites in IDRs are usually fewer than ten residues, their interaction interfaces are compact and can occur in multiple positions without compromising function~\cite{Tompa2014}.
@@
 By the mid 2010s several studies had demonstrated evidence of such constraint in the flexibility, chemical composition, net charge, or charge distribution of IDRs~\cite{Daughdrill2007, Moesa2012, Zarin2017, Beh2012}.
-While these earlier studies were generally restricted to specific features or proteins, in 2019 Zarin \textit{et al.} demonstrated conservation of various features in a comprehensive set of IDR-associated properties across IDRs in the entire yeast proteome~\cite{Zarin2019}.
+While these earlier studies were generally restricted to specific features or proteins, in 2019 Zarin \textit{et al.} demonstrated conservation of various features from a comprehensive set of IDR-associated properties in IDRs across the entire yeast proteome~\cite{Zarin2019}.
 By comparing the observed values of these features against those generated under a simulated model of evolution, they clustered IDRs by their ``evolutionary signatures,'' \texit{i.e.} patterns of conserved features.
@@
 
-These analyses were conducted using a set of IDRs identified in various species of yeast, which is a widely used model organism in molecular biology research.
+These analyses were conducted using IDRs identified in various species of yeast, which is a widely used model organism in molecular biology research.
 However, no known subsequent studies have determined if similar patterns of conservation are found in the IDRs of other systems.

### 2023-03-18 chapter0/chapter0.tex
Removed scare quotes around real scientific words in chapter 0

@@
 The most important class of these macromolecules are proteins, which are the molecular machines that carry out a cell's processes.
-Though proteins are linear chains of simpler building blocks called amino acids, many proteins accomplish their functions by ``folding'' into well-defined three-dimensional structures.
-For many years scientists believed fixed structures were necessary for protein function, but by the early 2000s, however, evidence had accumulated that unstructured or ``intrinsically disordered'' regions (IDRs) were ubiquitous in proteins.
+Though proteins are linear chains of simpler building blocks called amino acids, many proteins accomplish their functions by folding into well-defined three-dimensional structures.
+For many years scientists believed fixed structures were necessary for protein function, but by the early 2000s, however, evidence had accumulated that unstructured or intrinsically disordered regions (IDRs) were ubiquitous in proteins.
 Furthermore, these regions were essential for many cellular processes such as signaling and regulation.
@@
 While water can dissolve some oxygen, the body requires more oxygen more quickly than is available in the aqueous component of blood alone.
-Thus, red blood are packed with a special protein called hemoglobin, which binds\footnote{In biology, ``binding'' is a slippery word whose exact meaning can varying greatly depending on the context.
+Thus, red blood are packed with a special protein called hemoglobin, which binds\footnote{In biology, binding is a slippery word whose exact meaning can varying greatly depending on the context.
 However, it generally means that a molecule physically interacts with another molecular for an extended period.} oxygen.
@@
 However, as a result of this unique role in providing tissue elasticity, elastin's disorder was viewed as a specific adaptation rather than a general mechanism of protein function.
-In other cases, proteins had regions which returned undefined or highly variable atomic coordinates when analysed with structural techniques, indicating they lacked defined structures and were ``disordered.'' Because these segments were often short loops between structured regions, they were seen as ``linkers'' which facilitated the structure of the functional portions of proteins.
+In other cases, proteins had regions which returned undefined or highly variable atomic coordinates when analysed with structural techniques, indicating they lacked defined structures and were disordered.
+Because these segments were often short loops between structured regions, they were seen as linkers which facilitated the structure of the functional portions of proteins.
 By the early 2000s, though, enough exceptions had accumulated that scientists began to recognize that fully and partially disordered proteins were involved in many biological processes~\cite{Plaxco1997, Wright1999, Dunker2001}.
@@
 All animals, plants, and fungi are eukaryotes, though the domain includes many microorganisms as well.} organisms, with estimates of the fraction of proteins containing disordered segments of greater than 30 residues\footnote{The amino acids that compose the links of a protein chain are conventionally called residues to distinguish them from their related, but chemically distinct, free forms.} ranging between 28 and 63\%~\cite{Dunker2000, Ward2004}.
-For reference, though the lengths of proteins can vary dramatically, a ``typical'' protein contains on the order of a few hundred residues, so these regions can compose a significant fraction of a protein's length.
-Because these segments were disordered in their native state, \textit{i.e.} under normal operating conditions, such segments were termed ``intrinsically disordered regions'' (IDRs) to emphasize the disorder was not induced by exposure to chemicals or heat.
-Furthermore, while most proteins were predicted to contain a mixture of structure and disorder, some ``intrinsically disordered proteins'' (IDPs) were entirely or almost entirely disordered.
+For reference, though the lengths of proteins can vary dramatically, a typical protein contains on the order of a few hundred residues, so these regions can compose a significant fraction of a protein's length.
+Because these segments were disordered in their native state, \textit{i.e.} under normal operating conditions, such segments were termed intrinsically disordered regions (IDRs) to emphasize the disorder was not induced by exposure to chemicals or heat.
+Furthermore, while most proteins were predicted to contain a mixture of structure and disorder, some intrinsically disordered proteins (IDPs) were entirely or almost entirely disordered.
 Thus, disorder was recognized as a pervasive but poorly understood feature of proteins.
@@
 Strictly speaking, a protein's structure refers its three-dimensional arrangement of atoms.
-Thus, ``structured'' regions in proteins, often called domains\footnote{Though there are various overlapping definitions, domains typically refer to independently folding regions of a protein.
-Domains are also described as discrete functional or evolutionary elements since proteins may contain several domains which are connected by unstructured ``linker'' sequences.}, typically fold into a small number of related structures.
+Thus, structured regions in proteins, often called domains\footnote{Though there are various overlapping definitions, domains typically refer to independently folding regions of a protein.
+Domains are also described as discrete functional or evolutionary elements since proteins may contain several domains which are connected by unstructured linker sequences.}, typically fold into a small number of related structures.
 This does not imply these folded domains are completely rigid, though.
@@
 
-In some cases, folded domains have multiple structures, or ``conformations,'' which are related to different functional states.
+In some cases, folded domains have multiple structures, or conformations, which are related to different functional states.
 For example, hemoglobin has two forms, traditionally called the T and R states.
@@
 For example, the 26S proteasome is a complex of proteins responsible for degrading other proteins.
-Its structure is highly complex and consists of over three dozen distinct protein ``subunits,'' which are in turn organized into three subcomplexes: a lid, a base, and a core~\cite{Finley2016, Bard2018}.
+Its structure is highly complex and consists of over three dozen distinct protein subunits, which are in turn organized into three subcomplexes: a lid, a base, and a core~\cite{Finley2016, Bard2018}.
 The functions of these subcomplexes are roughly analogous to the parts of a paper shredder.
@@
 
-In contrast, IDRs have no fixed spatial relationship between their atoms, so their structures are sometimes described as ``conformational ensembles,'' \textit{i.e.} collections of conformations where the distances and orientations between residues can vary considerably.
+In contrast, IDRs have no fixed spatial relationship between their atoms, so their structures are sometimes described as conformational ensembles, \textit{i.e.} collections of conformations where the distances and orientations between residues can vary considerably.
 Furthermore, IDRs populate a continuum of structural states over time whereas when folded domains undergo dramatic conformational changes, they are usually triggered by specific environmental signals or chemical modifications, and the intermediate structures are transient.
-Despite the diversity of conformations available to IDRs, they can be broadly grouped into one of several qualitative descriptions which range from extended ``coils'' to more compact ``globules.'' The specific conformational class of a given IDR, however, is dictated by its local composition of amino acid residues.
+Despite the diversity of conformations available to IDRs, they can be broadly grouped into one of several qualitative descriptions which range from extended coils to more compact globules.
+The specific conformational class of a given IDR, however, is dictated by its local composition of amino acid residues.
 Because the chemical properties of each amino acid in the protein alphabet are dictated its specific arrangement of atoms, each has a unique impact on a protein's structure.
@@
 One of the most useful scales for describing an amino acid's overall effect on protein structure is hydrophobicity, which measures a molecule's tendency to associate with water.
-Molecules which attract water are hydrophilic (``water loving''), and molecules which repel water are hydrophobic (``water fearing'').
+Molecules which attract water are hydrophilic (water loving), and molecules which repel water are hydrophobic (water fearing).
 Hydrophilic molecules, like sugar or alcohol, easily dissolve in water whereas hydrophobic molecules, like fats and oils, remain separate.
-Since the cellular environment is largely water, the hydrophobic residues in proteins tend to aggregate into a ``hydrophobic core.'' This ``hydrophobic collapse'' is a major driving force in the early stages of protein folding, so a protein's relative number of hydrophilic and hydrophobic residues is a key determinant of whether it is folded or disordered.
+Since the cellular environment is largely water, the hydrophobic residues in proteins tend to aggregate into a hydrophobic core.
+This hydrophobic collapse is a major driving force in the early stages of protein folding, so a protein's relative number of hydrophilic and hydrophobic residues is a key determinant of whether it is folded or disordered.
 
@@
 High numbers of polar amino acids, which are hydrophilic but not charged, are associated with semi-compact globules.
-Though their ``side chains,'' the portions which give each amino acid its unique identity, are hydrophilic, their interactions with water are not sufficient to overcome the tendency of the hydrophobic ``backbone,'' composed the donor and receptor sites common to all amino acids, to self-associate.
+Though their side chains, the portions which give each amino acid its unique identity, are hydrophilic, their interactions with water are not sufficient to overcome the tendency of the hydrophobic backbone, composed the donor and receptor sites common to all amino acids, to self-associate.
 Thus, like folded domains, the structures of disordered proteins are dictated by their sequences.
@@
 However, it also introduces many complications because information in the form of physical or chemical signals cannot travel freely.} Many are mediated by the interaction of one protein with another, which in turn creates a change in the state of the system that propagates the signal.
-Often these state changes are chemical alterations made by one protein to another, which are termed ``post-translational modifications'' (PTMs).\footnote{A protein's amino acid sequence is encoded in a cell's DNA, and the process of reading that information to create a protein (from an intermediate molecule called RNA) is translation.
+Often these state changes are chemical alterations made by one protein to another, which are termed post-translational modifications (PTMs).\footnote{A protein's amino acid sequence is encoded in a cell's DNA, and the process of reading that information to create a protein (from an intermediate molecule called RNA) is translation.
 Therefore, any modifications to a protein after its initial synthesis are post-translational.} One of the most common modifications is phosphorylation where a phosphate group is attached to a specific amino acid in the protein.
@@
 Orthologs are colored with shades of the same hue.
-\textbf{(B-D)} Each horizontal line represents a sequence in panel A, and together each set of lines creates an ``alignment.'' The lines are broken where ``gaps'' are inserted to align equivalent segments.
+\textbf{(B-D)} Each horizontal line represents a sequence in panel A, and together each set of lines creates an alignment.
+The lines are broken where gaps are inserted to align equivalent segments.
 The grey segments denote regions of the alignment which are not part of subsequent analyses.
-\textbf{(E)} Each aligned IDR is scored on four features, where the ``strength'' of that feature is indicated by the height and color of its bar.
+\textbf{(E)} Each aligned IDR is scored on four features, where the strength of that feature is indicated by the height and color of its bar.
 \textbf{(F)} Scoring many IDRs on these features yields three distinct clusters.}
@@
 
-These comparisons, however, require the identification of IDRs with common ancestry that perform ``equivalent'' functions across many distinct organisms.
+These comparisons, however, require the identification of IDRs with common ancestry that perform equivalent functions across many distinct organisms.
 Given the difficulties with identifying similar IDRs by their sequences, this may seem like a chicken and egg problem.
 Fortunately, IDRs are frequently associated with more conserved folded domains.
-Thus, identifying evolutionarily related proteins by their overall sequence signatures and ``aligning'' them will in turn identify the equivalent IDRs in those sequences.
+Thus, identifying evolutionarily related proteins by their overall sequence signatures and aligning them will in turn identify the equivalent IDRs in those sequences.
 The first step of an evolutionary analysis of IDRs, then, is the identification of proteins with common ancestry, called orthologs.
@@
 Thus, in the first chapter I develop a novel method for identifying orthologs which addresses this shortcoming and apply it to 33 \textit{Drosophila} genomes to generate a set of aligned orthologs (Fig.~\ref{fig:aims_overview}A-B).
-In the second chapter I then identify rapidly evolving IDRs in these ``alignments'' and analyse them with a variety of evolutionary models to detect patterns of conservation (Fig.~\ref{fig:aims_overview}C-F).
+In the second chapter I then identify rapidly evolving IDRs in these alignments and analyse them with a variety of evolutionary models to detect patterns of conservation (Fig.~\ref{fig:aims_overview}C-F).
 Finally, in the third chapter I discuss several software tools and tutorials for fitting statistical models to data, which were created while pursuing the previous aims.

### 2023-03-18 chapter1/chapter1.tex
Removed scare quotes around real scientific words in chapter 1

@@
 However, many analyses impose the additional condition that the sequences have diverged through speciation events (orthology) rather than duplications (paralogy) or other mechanisms such as horizontal gene transfer~\cite{Fitch1970}.
-The underlying assumption is ``orthologs'' have conserved equivalent functions whereas ``paralogs,'' by virtue of their redundancy, are more likely to diverge~\cite{Ohno1970, Nowak1997, Altenhoff2012, Pegueroles2013, Soria2014}.\footnote{For multiple sequences, orthology is usually defined relative to their most recent common ancestor.
-This technically includes sequences which split by duplication after this point (``in-paralogs''), but excludes sequences which split by duplication before (``out-paralogs'') \cite{Remm2001}.
+The underlying assumption is orthologs have conserved equivalent functions whereas paralogs, by virtue of their redundancy, are more likely to diverge~\cite{Ohno1970, Nowak1997, Altenhoff2012, Pegueroles2013, Soria2014}.\footnote{For multiple sequences, orthology is usually defined relative to their most recent common ancestor.
+This technically includes sequences which split by duplication after this point (in-paralogs), but excludes sequences which split by duplication before (out-paralogs) \cite{Remm2001}.
 Many current orthology inference pipelines explicitly incorporate steps to detect in-paralogs.
-However, the resulting orthologs groups can easily be restricted to those without in-paralogs (``single copy orthologs'') for analyses where an assumption of conserved function is necessary.} This relationship between orthology and function is an essential component of modern biological research since it permits the transfer of annotations between biological systems using sequence similarity alone.
+However, the resulting orthologs groups can easily be restricted to those without in-paralogs (single copy orthologs) for analyses where an assumption of conserved function is necessary.} This relationship between orthology and function is an essential component of modern biological research since it permits the transfer of annotations between biological systems using sequence similarity alone.
 
@@
 
-Phylogenetic HMMs (phylo-HMMs) are statistical models that incorporate phylogenetic and spatial information to calculate the probability that each observation in a sequence was generated by one of several ``hidden states''~\cite{Felsenstein1996}.
+Phylogenetic HMMs (phylo-HMMs) are statistical models that incorporate phylogenetic and spatial information to calculate the probability that each observation in a sequence was generated by one of several hidden states~\cite{Felsenstein1996}.
 Since they can evaluate both the probability of a gap pattern in a column given the known phylogenetic relationships and the local context, a phylo-HMM was used to segment the alignment into contiguous regions with different patterns of gaps.
@@
 The second is the annotation process which can erroneously exclude or include segments from a sequence.
-The result is an alignment of annotated sequences which contains evolutionary information obscured by ``noise'' from the annotation process, shown here by the exclusion of three N-terminal residues in the fourth annotated sequence.
+The result is an alignment of annotated sequences which contains evolutionary information obscured by noise from the annotation process, shown here by the exclusion of three N-terminal residues in the fourth annotated sequence.
 To simplify modeling this behavior with an HMM, we coded the sequences into binary symbols.
@@
 
-Though the phylo-HMM removed phylogenetically incongruent ``insertions,'' some sequences still contained extensive segments of uninterrupted gaps.
+Though the phylo-HMM removed phylogenetically incongruent insertions, some sequences still contained extensive segments of uninterrupted gaps.
 These segments are easily identified in regions which are otherwise highly conserved, so they are also likely the result of incorrect or incomplete annotations.

### 2023-03-18 chapter3/training_tutorial.tex
Added chapter 2 appendix
Removed scare quotes around real scientific words in chapter 3

@@
 
-The optimized or ``learned'' parameters are then written as
+The optimized or learned parameters are then written as
 
@@
 
-This is now a fairly clean expression since at each step we ``pick'' the right transition probability and emission distribution using the state sequence.
+This is now a fairly clean expression since at each step we pick the right transition probability and emission distribution using the state sequence.
 However, it will be useful both theoretically and computationally to write this expression in terms of the number of times each transition appears in the sequence.
@@
 However, modeling real-world data is often far more complicated since it's rare to have perfect knowledge of the data generating process even if the states are labeled.
-Thus, it's tempting to use a ``fully connected'' model that allows transitions between each pair of states and let the model learn what transitions are actually used from the data.
+Thus, it's tempting to use a fully connected model that allows transitions between each pair of states and let the model learn what transitions are actually used from the data.
 That said, constraining the allowed transitions can lead to better models, particularly if the constraints reflect a feature of the underlying process.
 
-For this example, we'll assume that we have some ``domain knowledge'' that permits us to know the states and their allowed transitions.
+For this example, we'll assume that we have some domain knowledge that permits us to know the states and their allowed transitions.
 For example, state 2 can transition to every state, but states 1 and 3 can only transition to themselves and state 2.
@@
 Another nuance for the normal distribution in particular is the derivative is taken relative to the variance, $\sigma^2$, and not the standard deviation, $\sigma$.
-Though this choice does not impact the resulting formula, it both simplifies the calculation and reflects the natural role of the variance as the more ``fundamental'' statistical quantity.
+Though this choice does not impact the resulting formula, it both simplifies the calculation and reflects the natural role of the variance as the more fundamental statistical quantity.
 
@@
 When the states are known, the MLEs are the same, but instead of counting transitions between states, we count emissions.
-(The emissions essentially take on the role of the ``target'' state $j$.) When the states are unknown, we replace these counts with how much we think each state is responsible for each emission.
+(The emissions essentially take on the role of the target state $j$.) When the states are unknown, we replace these counts with how much we think each state is responsible for each emission.
 In other words, if at time $t$ we observe emission $i$ and we calculate an 90\% chance of state 1 and a 10\% chance of state 2, we assign a 0.9 count towards $n_{1i}$ and a 0.1 count towards $n_{2i}$.
@@
 Though it's good practice to use the best result from several different random initializations, if the initial parameters are hard coded, the states must be different from each other in some way.
-If the states are all identical, there's no way for the model to ``break symmetry,'' and all the updates will yield parameters that are the same as the initial ones.
+If the states are all identical, there's no way for the model to break symmetry, and all the updates will yield parameters that are the same as the initial ones.
 
@@
 
-For brevity, the input code is given in Appendix~\ref{appendix:b}, and only the final values are shown below.
+For brevity, the input code is given in Appendix~\ref{appendix:c}, and only the final values are shown below.
 
@@
 This is a common occurrence when using the EM algorithm and is a consequence of the construction of the objective function as a weighted sum of log-likelihood functions.
-Thus, when the MLEs are interpretable as quantities like counts or averages, the corresponding update equations are often ``weighted'' versions of those estimators.
+Thus, when the MLEs are interpretable as quantities like counts or averages, the corresponding update equations are often weighted versions of those estimators.
 Of course, it's not always obvious how the weights will come into play, especially for MLEs which don't have an obvious interpretation, so it's good idea to solve the equations or look up a reference rather than guessing.

### 2023-03-18 chapter0/chapter0.tex
Added citations to SLiMs paragraph

@@
 SLiMs are a general mechanism for mediating interactions with proteins, so while many SLiMs are targets of PTMs, others simply recruit binding partners.
-In contrast to the highly structured interfaces that characterize folded domains, however, SLiMs are usually no more than ten residues.
+In contrast to the highly structured interfaces that characterize folded domains, however, SLiMs are usually no more than ten residues~\cite{Tompa2014}.
 Thus, their interactions are relatively weak and highly transient.
 As their flexibility makes SLiMs easily accessible, they are also enriched in IDRs, which in turn allows them to bind to many partners, sometimes simultaneously.
-IDRs therefore often act as hubs in complex regulatory networks by propagating signals from diverse sources or organizing binding partners into higher-order structures.
+IDRs therefore often act as hubs in complex regulatory networks by propagating signals from diverse sources or organizing binding partners into higher-order structures~\cite{Dunker2005, Wright2014}.
 Furthermore, these signals and interactions are easily tuned because PTMs can modulate their meanings and strengths, respectively, by modifying only a few residues.

### 2023-03-18 chapter0/chapter0.tex
Tweaked wording in domains of life footnote to make it less redundant

@@
 In contrast, the cells of members of Eukarya, \textit{i.e.} eukaryotes, are complex and contain substructures called organelles, among other differences.
-All animals, plants, and fungi are eukaryotes, though the domain includes many microorganisms as well.} organisms, with estimates of the fraction of proteins containing disordered segments of greater than 30 residues\footnote{The amino acids that compose the links of a protein chain are conventionally called residues to distinguish them from their related, but chemically distinct, free forms.} ranging between 28 and 63\%~\cite{Dunker2000, Ward2004}.
+Animals, plants, and fungi are eukaryotes, though the domain includes many microorganisms as well.} organisms, with estimates of the fraction of proteins containing disordered segments of greater than 30 residues\footnote{The amino acids that compose the links of a protein chain are conventionally called residues to distinguish them from their related, but chemically distinct, free forms.} ranging between 28 and 63\%~\cite{Dunker2000, Ward2004}.
 For reference, though the lengths of proteins can vary dramatically, a typical protein contains on the order of a few hundred residues, so these regions can compose a significant fraction of a protein's length.

### 2023-03-20 main.tex
Removed parents names from dedication

@@
 \begin{center}
-    To my parents, Amy and David, who were always willing\\
-    to indulge another ``experiment'' in their kitchen sink
+    To my parents, who were always willing to indulge\\
+    another ``experiment'' in their kitchen sink
 \end{center}
@@
 Thank you to my undergraduate research mentors, Cathy and Jessica, for teaching me the most exciting science is at the intersection of different fields.\\\\
-Thank you to my parents, for everything.\\\\
+Thank you to my parents, Amy and David, for everything.\\\\
 Thank you to my brothers, Andrew and Daniel, for giving me thick skin in the way only big brothers can.\\\\

### 2023-03-20 chapter0/chapter0.tex
Replaced dramatic with large to sound less sensational

@@
 In contrast, IDRs have no fixed spatial relationship between their atoms, so their structures are sometimes described as conformational ensembles, \textit{i.e.} collections of conformations where the distances and orientations between residues can vary considerably.
-Furthermore, IDRs populate a continuum of structural states over time whereas when folded domains undergo dramatic conformational changes, they are usually triggered by specific environmental signals or chemical modifications, and the intermediate structures are transient.
+Furthermore, IDRs populate a continuum of structural states over time whereas when folded domains undergo large conformational changes, they are usually triggered by specific environmental signals or chemical modifications, and the intermediate structures are transient.
 Despite the diversity of conformations available to IDRs, they can be broadly grouped into one of several qualitative descriptions which range from extended coils to more compact globules.

### 2023-03-20 chapter0/chapter0.tex
Added phrase referencing conformational class to better lead into subsequent discussion

@@
 However, there is no simple formula which accurately predicts disorder in a protein using only the hydrophobicity values of its constituent residues.
-Other factors, such as the number and distribution of charged residues, also impact a sequence's predisposition for disorder~\cite{vanderLee2014, Das2015}.
+Other factors, such as the number and distribution of charged residues, also impact a sequence's predisposition for disorder and its resulting conformational class~\cite{vanderLee2014, Das2015}.
 For example, sequences with high number of either positively or negatively charged residues, called polyelectrolytes, tend to form stiff rods because the like charges repel each other.

### 2023-03-21 chapter0/chapter0.tex
Removed reference to regulation which better reflects phrasing in subsequent paragraph
Added paragraphs on IDRs in transcription

@@
 Often these partners are other proteins, but they can also be other macromolecules like DNA or even small molecules and ions.
-As a result of this adaptability, IDRs are enriched in proteins involved in cell signaling and regulation.
+As a result of this adaptability, IDRs are enriched in proteins involved in cell signaling.
 Because cells are highly compartmentalized, they have a dizzying array of mechanisms to relay messages.\footnote{The most fundamental compartment is the cell, which roughly separates inside from outside and life from non-life.
@@
 
-Some of the most prominent examples of IDRs in cell signaling and regulation are the activation domains of transcription factors.
-Each cell expresses a unique complement of proteins that in large part determines its identity.
+Some of the most prominent examples of IDRs in cell signaling are found in the proteins that regulate the creation of other proteins.
+A protein's amino acid sequence is encoded in DNA in a unit called a gene, and since all cells in an organism share the same genome, they can in principle synthesize any protein encoded in it.\footnote{The DNA sequences of different cells in an organism are not always strictly identical.
+For example, spontaneous mutations can create changes that range from single-letter substitutions to large-scale rearrangements.
+In other cases, such as during the creation of sex or immune cells, a cell's DNA is intentionally modified to generate genetic diversity.} However, different cell types instead express unique complements of proteins that in large part determine their identities.
+This selective conversion of the information stored in DNA into proteins is called gene expression and is a highly regulated process that varies in space and time, \textit{i.e.} between different cells and within a single cell.
+Though gene expression involves dozens, if not hundreds, of distinct biochemical reactions, most occur as steps within two major processes: transcription and translation.
+In transcription, the sequence of a protein encoded the cell's DNA is transcribed into an intermediate molecule called messenger RNA (mRNA).
+mRNA is also a kind of nucleic acid, and is chemically closely related to DNA.
+However, as a much smaller molecule containing only the information needed to synthesize a protein, mRNA is more easily transported and manipulated in subsequent steps.
+Thus, if a gene stored as DNA is a cell's master record of a protein sequence, mRNA is its working copy.
+In translation, a vast molecular machine called the ribosome then synthesizes a protein by translating the information encoded in a strand of mRNA into a sequence of amino acids.
 
-% IDRs are enriched in transcriptional machinery
-%    A. Introduce transcription via the central dogma
-%    B. Describe transcription factors, cofactors, transcriptional machinery
-%        1. Examples of different transcription factors
-%    C. Describe properties of transcription factors
-%        1. DNA binding and activation domains
-%        2. Different flavors of activation domains
-%    D. How transcription factors find their targets to initiate specific genetic programs is one of the major outstanding questions in molecular biology
-
-% Disorder is spectrum
-%    A. MorFs, folding on binding, fuzzy complexes
+Both transcription and translation are tightly regulated processes, but because cells employ numerous interlocking mechanisms to control access to DNA and the creation of mRNA, transcription is often the major regulatory checkpoint in gene expression.
+A key step in this process is the recruitment of the protein complex which transcribes DNA into mRNA to the beginning of a gene's sequence, \textit{i.e.} its transcription start site (TSS).
+The placement of this complex, called RNA polymerase, is dictated by regulatory sequences encoded in the DNA, and while some regulatory sequences are ubiquitous, like those that mark a gene's TSS, many are gene-specific.
+Thus, the unique collection of regulatory sequences associated with a gene determines many aspects of its expression.
+RNA polymerase does not directly contact these gene-specific regulatory elements, however.
+Instead, proteins called transcription factors directly bind these sequences and interact with other components of the transcriptional machinery which in turn recruit RNA polymerase to the TSS.
+Accordingly, a typical transcription factor has two parts: a DNA-binding domain and an activation domain.
+DNA-binding domains are usually structured because they make stable contacts with specific regulatory sequences.
+In contrast, activations domains interact with the transcriptional machinery and are highly enriched in IDRs.
+Decades of research have identified many of the proteins involved in these interactions as well as some common properties of their sequences.
+Despite this progress, however, the precise mechanisms by which transcription factors locate their DNA targets and recruit the transcriptional machinery to initiate expression of specific genes remains an unsolved problem in molecular biology.
+As disordered regions play a key role in this process, studies which broadly investigate the relationship between the sequences, structures, and functions of IDRs therefore have significant implications for our understanding of gene regulation.
 

### 2023-03-21 chapter0/chapter0.tex
Fixed a variety of typos and instances of awkward or unclear wording

@@
 Though proteins are linear chains of simpler building blocks called amino acids, many proteins accomplish their functions by folding into well-defined three-dimensional structures.
-For many years scientists believed fixed structures were necessary for protein function, but by the early 2000s, however, evidence had accumulated that unstructured or intrinsically disordered regions (IDRs) were ubiquitous in proteins.
+For many years scientists believed fixed structures were necessary for protein function, but by the early 2000s evidence had accumulated that unstructured or intrinsically disordered regions (IDRs) were ubiquitous in proteins.
 Furthermore, these regions were essential for many cellular processes such as signaling and regulation.
@@
 Life is a physical phenomenon.
-Despite the complexity of living things, their processes are governed by the same physical laws that describe planets' motion around the sun and the propagation of electromagnetic waves through space.
-However, many systems are too complex to describe with physical models and equations, so scientists simplify them into levels of abstraction that are more useful.\footnote{This is the origin of the common observation that biology is applied chemistry and chemistry is applied physics.} For example, Punnett squares facilitate the prediction of genotypes and phenotypes by distilling the complexities and nuances of diverse reproductive systems into a set of simple rules.
+Despite the complexity of living things, their processes are governed by the same physical laws that describe the planets' motion around the sun and the propagation of electromagnetic waves through space.
+However, many systems are too complex to describe with physical models and equations, so scientists simplify them into levels of abstraction that are more useful.\footnote{This is the origin of the observation that biology is applied chemistry and chemistry is applied physics.} For example, Punnett squares facilitate the prediction of genotypes and phenotypes by distilling the complexities and nuances of diverse reproductive systems into a set of simple rules.
 However, since life spans a scale from single cells to entire ecosystems, biology likely employs more layers of abstraction than any other scientific discipline.
@@
 A specific focus is four classes of macromolecules called nucleic acids, carbohydrates, lipids, and proteins, all of which are unique to biological systems.
-Though not all biological molecules are macromolecules\footnote{Macromolecule is a loose term applied to molecules with high molecular mass.
+Though not all biological molecules are macromolecules\footnote{Macromolecule is a loose term applied to molecules with high molecular masses.
 In practice, it usually refers to one of the classes listed above, among a few other prominent non-biological examples.} and not all biological macromolecules fit neatly in one of these four categories, much of life at the molecular level is understood in terms of their structure and function.
@@
 Some examples will illustrate the central role of proteins more clearly.
-Blood is a part of the circulatory system which is responsible for transporting nutrients and waste.
+Blood is a part of the circulatory system, which is responsible for transporting nutrients and waste.
 Though blood is a complex mixture, containing a cocktail of cells, proteins, sugars, gases, and ions dissolved in a medium of water, its primary cellular component is red blood cells.
@@
 While water can dissolve some oxygen, the body requires more oxygen more quickly than is available in the aqueous component of blood alone.
-Thus, red blood are packed with a special protein called hemoglobin, which binds\footnote{In biology, binding is a slippery word whose exact meaning can varying greatly depending on the context.
+Thus, red blood cells are packed with a special protein called hemoglobin, which binds\footnote{In biology, binding is a slippery word whose exact meaning can varying greatly depending on the context.
 However, it generally means that a molecule physically interacts with another molecular for an extended period.} oxygen.
@@
 These cells respond to light because they produce special proteins called opsins that translate light into chemical and electrical signals which are then interpreted by the brain.
-Humans, and primates broadly, have three types of opsins that mediate our color vision, which are sensitive to red, green, and blue light, respectively.
+Humans, and primates broadly, have three types of opsins, which are sensitive to red, green, and blue light, respectively, that mediate our color vision.
 Color blindness is the result of photoreceptor cells missing one of these proteins, typically either the red or green opsin.
 In contrast to sickle cell disease, this condition is caused by a missing rather than a mutated protein.
-However, in other cases, a protein is not missing or mutated, but instead not produced at the right time and place.
+In other cases, however, a protein is not missing or mutated, but instead not produced at the right time and place.
 For example, lactose is a sugar found in milk that requires a specific protein, lactase, to metabolize properly.
-Many humans who can digest milk products in childhood, lose this ability in adulthood because they stop producing lactose.
+Many humans who can digest milk products in childhood lose this ability in adulthood because they stop producing lactose.
 As a result, lactose in dairy products passes undigested into the colon where it is broken down by bacteria, causing symptoms such as bloating and diarrhea.
@@
 However, people affected by sickle cell disease have a mutation in the sequence of their hemoglobin proteins which causes it to malfunction.
-Frequently this mutation is a single change where the sixth amino acid in the sequence, glutamate, is substituted for a valine.
+Frequently this mutation is a single change where the sixth amino acid in the sequence, glutamate, is substituted for valine.
 This creates a sticky patch on the surface of hemoglobin, and under low-oxygen conditions normal hemoglobin changes shape to expose a sticky patch on its surface as well.
@@
 Clearly, understanding the relationship between the sequence, structure, and function of proteins is essential for unraveling more complex biological phenomena.
-Though biologists study all three properties of proteins, they are generally most interested in function since it is the most directly related to the biological processes the protein takes part in\footnote{Function generally refers to \textit{molecular} function which is a description of a specific chemical activity possessed by a protein.
+Though biologists study all three properties of proteins, they are generally most interested in function since it is the most directly related to the biological processes the protein takes part in.\footnote{Function generally refers to \textit{molecular} function which is a description of a specific chemical activity possessed by a protein.
 A biological process, however, is the larger ``biological program'' which is accomplished by the action of multiple linked molecular functions.
@@
 Hemoglobin is found in red blood cells where it acts as a carrier during transport.
-In contrast, myoglobin is found in muscle cells, where it stores oxygen until needed.}, function and biological process are not always easily identified or measured.
+In contrast, myoglobin is found in muscle cells, where it stores oxygen until needed.} However, functions and biological processes are not always easily identified or measured.
 Thus, determining a protein's structure is frequently the first step of detailed studies of its function.
@@
 However, this resolution requires extremely pure samples of the protein of interest.
-Since proteins are only produced by living systems, preparations begin with a complex mixture consisting of cells or tissue, and the protein of interest is isolated with series of extraction and purification steps.
+Since proteins are only produced by living systems, preparations begin with a complex mixture consisting of cells or tissue, and the protein of interest is isolated through a series of extraction and purification steps.
 Some proteins are only produced in small amounts or degrade easily, so each step may require substantial optimization to achieve a sufficient yield.
@@
 
-The success of structural methods at elucidating the molecular details of protein function cemented the view that function largely depends on the presence of a fixed structure.
+The success of structural methods at elucidating the molecular details of protein function cemented the view that function depends on the presence of a fixed structure.
 While scientists understood proteins were not completely rigid and could adopt a variety of related structures, many believed that functional proteins largely had a single dominant structure~\cite{Karush1950}.
-Despite its strength, exceptions to the structure-function paradigm were known.
+Despite its strength, exceptions to this structure-function paradigm were known.
 For example, elastin is a protein secreted by cells which allows tissues like skin or blood vessels to repeatedly expand and contract.
@@
 By the early 2000s, though, enough exceptions had accumulated that scientists began to recognize that fully and partially disordered proteins were involved in many biological processes~\cite{Plaxco1997, Wright1999, Dunker2001}.
-Many of these examples were proteins which folded on binding to their targets, commonly other proteins.
+Many examples were proteins which folded on binding to their targets, commonly other proteins.
 This mechanism was a departure from the prevailing model of interactions between biological macromolecules which required highly stable and complementary interfaces, like two puzzle pieces fitting together.
 As a result, scientists speculated that disorder was an adaptation that allowed proteins to efficiently relay and regulate signals by enabling interactions with many possible targets.
-Furthermore, the flexibility of disordered protein would permit environmental conditions to easily modulate these interactions.
+Furthermore, the flexibility of disordered proteins would permit environmental conditions to easily modulate these interactions.
 
@@
 In contrast, the cells of members of Eukarya, \textit{i.e.} eukaryotes, are complex and contain substructures called organelles, among other differences.
-Animals, plants, and fungi are eukaryotes, though the domain includes many microorganisms as well.} organisms, with estimates of the fraction of proteins containing disordered segments of greater than 30 residues\footnote{The amino acids that compose the links of a protein chain are conventionally called residues to distinguish them from their related, but chemically distinct, free forms.} ranging between 28 and 63\%~\cite{Dunker2000, Ward2004}.
+Animals, plants, and fungi are eukaryotes, but the domain includes many microorganisms as well.} organisms, with estimates of the fraction of proteins containing disordered segments of greater than 30 residues\footnote{The amino acids that compose the links of a protein chain are conventionally called residues to distinguish them from their related, but chemically distinct, free forms.} ranging between 28 and 63\%~\cite{Dunker2000, Ward2004}.
 For reference, though the lengths of proteins can vary dramatically, a typical protein contains on the order of a few hundred residues, so these regions can compose a significant fraction of a protein's length.
-Because these segments were disordered in their native state, \textit{i.e.} under normal operating conditions, such segments were termed intrinsically disordered regions (IDRs) to emphasize the disorder was not induced by exposure to chemicals or heat.
+Because these segments were disordered in their native state, \textit{i.e.} under normal operating conditions, they were termed intrinsically disordered regions (IDRs) to emphasize the disorder was not induced by exposure to chemicals or heat.
 Furthermore, while most proteins were predicted to contain a mixture of structure and disorder, some intrinsically disordered proteins (IDPs) were entirely or almost entirely disordered.
@@
 In the proteasome, however, these are distinct steps.
-The motor subunits in the base first physically interact with the protein to simultaneously unfold and pull it into the core.
+The motor subunits in the base first physically interact with the target protein to simultaneously unfold and pull it into the core.
 Different subunits in the core then break the exposed chemical bonds between amino acid residues in the chain.} Like a paper shredder, the motor is only engaged when a protein is correctly positioned in the lid.
@@
 In contrast, IDRs have no fixed spatial relationship between their atoms, so their structures are sometimes described as conformational ensembles, \textit{i.e.} collections of conformations where the distances and orientations between residues can vary considerably.
-Furthermore, IDRs populate a continuum of structural states over time whereas when folded domains undergo large conformational changes, they are usually triggered by specific environmental signals or chemical modifications, and the intermediate structures are transient.
+Furthermore, IDRs populate a continuum of structural states over time, whereas when folded domains undergo large conformational changes, they are usually triggered by specific environmental signals or chemical modifications, and the intermediate structures are transient.
 Despite the diversity of conformations available to IDRs, they can be broadly grouped into one of several qualitative descriptions which range from extended coils to more compact globules.
@@
 Molecules which attract water are hydrophilic (water loving), and molecules which repel water are hydrophobic (water fearing).
-Hydrophilic molecules, like sugar or alcohol, easily dissolve in water whereas hydrophobic molecules, like fats and oils, remain separate.
+Hydrophilic molecules, like sugar or alcohol, easily dissolve in water, whereas hydrophobic molecules, like fats and oils, remain separate.
 Since the cellular environment is largely water, the hydrophobic residues in proteins tend to aggregate into a hydrophobic core.
@@
 Compartmentalization is essential in living systems because it confines and concentrates biochemical reactions that would be harmful if they occurred at the wrong place or time.
-However, it also introduces many complications because information in the form of physical or chemical signals cannot travel freely.} Many are mediated by the interaction of one protein with another, which in turn creates a change in the state of the system that propagates the signal.
+However, it also introduces many complications because information in the form of physical or chemical signals cannot travel freely.} Many are mediated by interactions between proteins, which in turn create a change in the state of the system that propagates the signal.
 Often these state changes are chemical alterations made by one protein to another, which are termed post-translational modifications (PTMs).\footnote{A protein's amino acid sequence is encoded in a cell's DNA, and the process of reading that information to create a protein (from an intermediate molecule called RNA) is translation.
@@
 SLiMs are a general mechanism for mediating interactions with proteins, so while many SLiMs are targets of PTMs, others simply recruit binding partners.
-In contrast to the highly structured interfaces that characterize folded domains, however, SLiMs are usually no more than ten residues~\cite{Tompa2014}.
+In contrast to the highly structured interfaces that characterize interactions with folded domains, however, SLiMs are usually no more than ten residues~\cite{Tompa2014}.
 Thus, their interactions are relatively weak and highly transient.
@@
 Despite this progress, however, the precise mechanisms by which transcription factors locate their DNA targets and recruit the transcriptional machinery to initiate expression of specific genes remains an unsolved problem in molecular biology.
-As disordered regions play a key role in this process, studies which broadly investigate the relationship between the sequences, structures, and functions of IDRs therefore have significant implications for our understanding of gene regulation.
+As disordered regions play a key role in this process, studies which broadly investigate the relationship between the sequence, structure, and function of IDRs therefore have significant implications for our understanding of gene regulation.
 
@@
 Because the composition and patterning of residues in an IDR dictates its conformational class, many distinct sequences can yield similar conformational ensembles.
-Furthermore, because binding motifs and modification sites in IDRs are usually fewer than ten residues, their interaction interfaces are compact and can occur in multiple positions without compromising function~\cite{Tompa2014}.
-Thus, rather than conserving specific sequences, IDRs conserve distributed ``molecular features'' associated with those sequences.
+Furthermore, because modification and binding sites in IDRs are usually fewer than ten residues, their interaction interfaces are compact and can occur in multiple positions without compromising function~\cite{Tompa2014}.
+Thus, rather than conserving specific sequences, IDRs are hypothesized to conserve distributed ``molecular features'' associated with those sequences.
 By the mid 2010s several studies had demonstrated evidence of such constraint in the flexibility, chemical composition, net charge, or charge distribution of IDRs~\cite{Daughdrill2007, Moesa2012, Zarin2017, Beh2012}.
 While these earlier studies were generally restricted to specific features or proteins, in 2019 Zarin \textit{et al.} demonstrated conservation of various features from a comprehensive set of IDR-associated properties in IDRs across the entire yeast proteome~\cite{Zarin2019}.
-By comparing the observed values of these features against those generated under a simulated model of evolution, they clustered IDRs by their ``evolutionary signatures,'' \texit{i.e.} patterns of conserved features.
+By comparing the observed values of these features against those generated under a simulated model of evolution, they clustered IDRs by their ``evolutionary signatures,'' \textit{i.e.} patterns of conserved features.
 Furthermore, they identified specific biological functions associated with these groups, which for the first time provided a global view of the relationship between sequence and function in IDRs.
@@
 The comparative genomics approach instead leverages the work done by nature to identify evolutionarily conserved features and generate specific hypotheses to guide experiments~\cite{Hardison2003}.
-As life is constantly exploring the space of allowed proteins through evolutionary change, features which are unimportant for a protein's function, or at a larger scale, an organism's survival offer no benefit for their maintenance and are therefore gradually degraded and lost.
+As life is constantly exploring the space of allowed proteins through evolutionary change, features which are unimportant for a protein's function or, at a larger scale, an organism's survival offer no benefit for their maintenance and are therefore gradually degraded and lost.
 Thus, conservation in related sequences is powerful signal of function.
@@
 The evolutionary relationships between the genomes of closely related species generally make such mistakes easier to identify, and fortunately over the past five years advances in DNA sequencing technology have yielded dramatic increases in the number of sequenced genomes in the \textit{Drosophila} genus.
-However, because the existing methods for ortholog identification were designed for fewer or more distantly related genomes, they do not fully leveraged the available genomic redundancy to minimize errors.
+However, because the existing methods for ortholog identification were designed for fewer or more distantly related genomes, they do not fully leverage such genomic redundancy to minimize errors.
 Thus, in the first chapter I develop a novel method for identifying orthologs which addresses this shortcoming and apply it to 33 \textit{Drosophila} genomes to generate a set of aligned orthologs (Fig.~\ref{fig:aims_overview}A-B).

### 2023-03-21 chapter0/chapter0.tex
Moved footnotes to end of clauses or sentences

@@
 A specific focus is four classes of macromolecules called nucleic acids, carbohydrates, lipids, and proteins, all of which are unique to biological systems.
-Though not all biological molecules are macromolecules\footnote{Macromolecule is a loose term applied to molecules with high molecular masses.
-In practice, it usually refers to one of the classes listed above, among a few other prominent non-biological examples.} and not all biological macromolecules fit neatly in one of these four categories, much of life at the molecular level is understood in terms of their structure and function.
-Each of the four has a characteristic role.
+Though not all biological molecules are macromolecules and not all biological macromolecules fit neatly in one of these four categories, much of life at the molecular level is understood in terms of their structure and function.\footnote{Macromolecule is a loose term applied to molecules with high molecular masses.
+In practice, it usually refers to one of the classes listed above, among a few other prominent non-biological examples.} Each of the four has a characteristic role.
 Nucleic acids, \textit{i.e.} DNA and RNA, are responsible for information storage and transfer.
@@
 While water can dissolve some oxygen, the body requires more oxygen more quickly than is available in the aqueous component of blood alone.
-Thus, red blood cells are packed with a special protein called hemoglobin, which binds\footnote{In biology, binding is a slippery word whose exact meaning can varying greatly depending on the context.
-However, it generally means that a molecule physically interacts with another molecular for an extended period.} oxygen.
-Each red blood cell contains as many as 270 million molecules of hemoglobin, each of which can carry up to four oxygen molecules~\cite{Pierig2008}.
+Thus, red blood cells are packed with a special protein called hemoglobin, which binds oxygen.\footnote{In biology, binding is a slippery word whose exact meaning can varying greatly depending on the context.
+However, it generally means that a molecule physically interacts with another molecular for an extended period.} Each red blood cell contains as many as 270 million molecules of hemoglobin, each of which can carry up to four oxygen molecules~\cite{Pierig2008}.
 Because red blood cells are so dense with hemoglobin, composing roughly 35\% of their total volume, any defect in hemoglobin can dramatically impact the structure of the red blood cells themselves~\cite{Kanias2009}.
@@
 melanogaster} (fruit fly) were sequenced for the first time, researchers applied computational methods for predicting disorder to the proteins inferred from their genomes.
-They discovered that disorder is ubiquitous in eukaryotic\footnote{All life belongs to one of three categories, or domains.
+They discovered that disorder is ubiquitous in eukaryotic organisms,\footnote{All life belongs to one of three categories, or domains.
 Two, Archaea and Bacteria, are all single-celled organisms with simple cellular structures.
 In contrast, the cells of members of Eukarya, \textit{i.e.} eukaryotes, are complex and contain substructures called organelles, among other differences.
-Animals, plants, and fungi are eukaryotes, but the domain includes many microorganisms as well.} organisms, with estimates of the fraction of proteins containing disordered segments of greater than 30 residues\footnote{The amino acids that compose the links of a protein chain are conventionally called residues to distinguish them from their related, but chemically distinct, free forms.} ranging between 28 and 63\%~\cite{Dunker2000, Ward2004}.
-For reference, though the lengths of proteins can vary dramatically, a typical protein contains on the order of a few hundred residues, so these regions can compose a significant fraction of a protein's length.
+Animals, plants, and fungi are eukaryotes, but the domain includes many microorganisms as well.} with estimates of the fraction of proteins containing disordered segments of greater than 30 residues ranging between 28 and 63\%~\cite{Dunker2000, Ward2004}.\footnote{The amino acids that compose the links of a protein chain are conventionally called residues to distinguish them from their related, but chemically distinct, free forms.} For reference, though the lengths of proteins can vary dramatically, a typical protein contains on the order of a few hundred residues, so these regions can compose a significant fraction of a protein's length.
 Because these segments were disordered in their native state, \textit{i.e.} under normal operating conditions, they were termed intrinsically disordered regions (IDRs) to emphasize the disorder was not induced by exposure to chemicals or heat.
@@
 Strictly speaking, a protein's structure refers its three-dimensional arrangement of atoms.
-Thus, structured regions in proteins, often called domains\footnote{Though there are various overlapping definitions, domains typically refer to independently folding regions of a protein.
-Domains are also described as discrete functional or evolutionary elements since proteins may contain several domains which are connected by unstructured linker sequences.}, typically fold into a small number of related structures.
-This does not imply these folded domains are completely rigid, though.
+Thus, structured regions in proteins, often called domains, typically fold into a small number of related structures.\footnote{Though there are various overlapping definitions, domains typically refer to independently folding regions of a protein.
+Domains are also described as discrete functional or evolutionary elements since proteins may contain several domains which are connected by unstructured linker sequences.} This does not imply these folded domains are completely rigid, though.
 At the molecular level, everything is in constant motion.

### 2023-03-21 chapter0/chapter0.tex
Added citations for studies of activation domains

@@
 DNA-binding domains are usually structured because they make stable contacts with specific regulatory sequences.
-In contrast, activations domains interact with the transcriptional machinery and are highly enriched in IDRs.
-Decades of research have identified many of the proteins involved in these interactions as well as some common properties of their sequences.
+In contrast, activations domains interact with the transcriptional machinery and are highly enriched in IDRs~\cite{Liu2006}.
+Decades of research have identified many of the proteins involved in these interactions as well as some common properties of their sequences~\cite{Ma1987, Sigler1988, Gerber1994, Arnold2018, Staller2018, Sanborn2021}.
 Despite this progress, however, the precise mechanisms by which transcription factors locate their DNA targets and recruit the transcriptional machinery to initiate expression of specific genes remains an unsolved problem in molecular biology.

### 2023-03-22 chapter0/chapter0.tex
Added more citations covering FlyBase, conformational classes, disorder prediction, and AD sequence features

@@
 
-Unsurprisingly, disordered regions are characterized by a relative depletion of hydrophobic residues.
+Unsurprisingly, disordered regions are characterized by a relative depletion of hydrophobic residues~\cite{Uversky2000}.
 However, there is no simple formula which accurately predicts disorder in a protein using only the hydrophobicity values of its constituent residues.
-Other factors, such as the number and distribution of charged residues, also impact a sequence's predisposition for disorder and its resulting conformational class~\cite{vanderLee2014, Das2015}.
+Other factors, such as the number and distribution of charged residues, also impact a sequence's predisposition for disorder and its resulting conformational class~\cite{Mao2010, vanderLee2014, Das2015}.
 For example, sequences with high number of either positively or negatively charged residues, called polyelectrolytes, tend to form stiff rods because the like charges repel each other.
@@
 In contrast, activations domains interact with the transcriptional machinery and are highly enriched in IDRs~\cite{Liu2006}.
-Decades of research have identified many of the proteins involved in these interactions as well as some common properties of their sequences~\cite{Ma1987, Sigler1988, Gerber1994, Arnold2018, Staller2018, Sanborn2021}.
+Decades of research have identified many of the proteins involved in these interactions as well as some common properties of their sequences~\cite{Ma1987, Sigler1988, Gerber1994, Arnold2018, Staller2018, Erijman2020, Sanborn2021}.
 Despite this progress, however, the precise mechanisms by which transcription factors locate their DNA targets and recruit the transcriptional machinery to initiate expression of specific genes remains an unsolved problem in molecular biology.
@@
 
-Despite recent advances in identifying IDRs and their conformational ensembles from their sequences alone, the relationship between the sequence and function remains poorly understood.
+Despite recent advances in identifying IDRs and their conformational ensembles from their sequences alone, the relationship between the sequence and function remains poorly understood~\cite{Mao2010, vanderLee2014, Das2015, Necci2021}.
 In contrast, predictions of structure and function are readily available for many folded domains.
@@
 However, no known subsequent studies have determined if similar patterns of conservation are found in the IDRs of other systems.
-As another foundational model organism with abundant genomic information across many evolutionary lineages, the fruit fly, \textit{Drosophila melanogaster}, is a natural choice for subsequent investigation~\cite{Yang2018, Miller2018, Kim2021}.
+As another foundational model organism with abundant genomic information across many evolutionary lineages, the fruit fly, \textit{Drosophila melanogaster}, is a natural choice for subsequent investigation~\cite{Yang2018, Miller2018, Kim2021, Gramates2022}.
 Furthermore, given its complex multicellular development process and shared signaling pathways with humans, the findings of such a study would significantly advance our understanding of the role of IDRs in human health and disease.

### 2023-03-30 chapter1/figures/hmm/hmm.py
Removed MSA path as parameter since other file paths are hardcoded

@@ -13,7 +13,6 @@ from src1.ortho_MSA import phylo
 from src1.utils import read_fasta
 
 OGids = ['2252', '2A57', '360E']
-msa_path = '../../orthology_inference/analysis/ortho_MSA/realign_fastas/out/'
 spid_regex = r'spid=([a-z]+)'
 
 state_labels = ['1A', '1B', '2', '3']
@@ -53,7 +52,7 @@ if not os.path.exists('out/'):
 for OGid in OGids:
     # Load MSA
     msa = []
-    for header, seq in read_fasta(f'{msa_path}/{OGid}.afa'):
+    for header, seq in read_fasta(f'../../orthology_inference/analysis/ortho_MSA/realign_fastas/out/{OGid}.afa'):
         spid = re.search(spid_regex, header).group(1)
         msa.append({'spid': spid, 'seq': seq})
     msa = sorted(msa, key=lambda x: tip_order[x['spid']])

### 2023-03-30 chapter1/figures/realignment/realignment.py

@@ -13,8 +13,6 @@ records = [('0167', 'A', 2357, 2714, 2763, 2831),
            ('2770', 'B', 2440, 2831, 2235, 2743),
            ('23D9', 'C', 731, 1150, 702, 982),
            ('35D6', 'D', 2774, 3162, 3003, 3355)]
-msa_path1 = '../../orthology_inference/analysis/ortho_MSA/get_repseqs/out/'
-msa_path2 = '../../orthology_inference/analysis/ortho_MSA/realign_fastas/out/'
 spid_regex = r'spid=([a-z]+)'
 
 adjust_left1 = 0.02
@@ -46,8 +44,8 @@ if not os.path.exists('out/'):
     os.mkdir('out/')
 
 for OGid, panel_label, start1, stop1, start2, stop2 in records:
-        path1 = f'{msa_path1}/{OGid}.afa'
-        path2 = f'{msa_path2}/{OGid}.afa'
+        path1 = f'../../orthology_inference/analysis/ortho_MSA/get_repseqs/out/{OGid}.afa'
+        path2 = f'../../orthology_inference/analysis/ortho_MSA/realign_fastas/out/{OGid}.afa'
 
         msa1 = []
         for header, seq in read_fasta(path1):

### 2023-03-30 chapter1/figures/realignment/realignment.py
Updated realignment.py to plot all panels on single figure

@@ -4,7 +4,6 @@ import os
 import re
 
 import matplotlib.pyplot as plt
-import numpy as np
 import skbio
 from src1.draw import plot_msa
 from src1.utils import read_fasta
@@ -29,8 +28,8 @@ legend_fontsize = 5
 legend_handletextpad = 0.5
 legend_markerscale = 1
 
-fig_width = 3.75
-fig_height = 2.5
+panel_width = 7.5
+panel_height = 2.5
 dpi = 600
 
 panel_label_fontsize = 'large'
@@ -43,53 +42,44 @@ tip_order = {tip.name: i for i, tip in enumerate(tree_order.tips())}
 if not os.path.exists('out/'):
     os.mkdir('out/')
 
-for OGid, panel_label, start1, stop1, start2, stop2 in records:
-        path1 = f'../../orthology_inference/analysis/ortho_MSA/get_repseqs/out/{OGid}.afa'
-        path2 = f'../../orthology_inference/analysis/ortho_MSA/realign_fastas/out/{OGid}.afa'
+records = records[:-1]  # Exclude final alignment
+fig = plt.figure(figsize=(panel_width, len(records) * panel_height))
+gs = plt.GridSpec(len(records), 2)
+for row_idx, record in enumerate(records):
+    OGid, panel_label, start1, stop1, start2, stop2 = record
 
-        msa1 = []
-        for header, seq in read_fasta(path1):
-            spid = re.search(spid_regex, header).group(1)
-            msa1.append({'spid': spid, 'seq': seq})
-        msa1 = sorted(msa1, key=lambda x: tip_order[x['spid']])
+    msa1 = []
+    for header, seq in read_fasta(f'../../orthology_inference/analysis/ortho_MSA/get_repseqs/out/{OGid}.afa'):
+        spid = re.search(spid_regex, header).group(1)
+        msa1.append({'spid': spid, 'seq': seq})
+    msa1 = sorted(msa1, key=lambda x: tip_order[x['spid']])
 
-        msa2 = []
-        for header, seq in read_fasta(path2):
-            spid = re.search(spid_regex, header).group(1)
-            msa2.append({'spid': spid, 'seq': seq})
-        msa2 = sorted(msa2, key=lambda x: tip_order[x['spid']])
+    msa2 = []
+    for header, seq in read_fasta(f'../../orthology_inference/analysis/ortho_MSA/realign_fastas/out/{OGid}.afa'):
[…]

### 2023-03-30 chapter1/figures/clusters/clusters.py
Removed unnecessary import of GridSpec

@@ -5,7 +5,6 @@ import os
 import matplotlib.pyplot as plt
 import pandas as pd
 from matplotlib.figure import SubplotParams
-from matplotlib.gridspec import GridSpec
 
 # Load genomes
 spids = set()
@@ -45,7 +44,7 @@ if not os.path.exists('out/'):
     os.mkdir('out/')
 
 fig = plt.figure(figsize=(7.5, 4.5), subplotpars=SubplotParams(0.25, 0.2, 1, 0.9))
-gs = GridSpec(2, 2)
+gs = plt.GridSpec(2, 2)
 
 counts_OGid = OGs[['spid', 'OGid']].drop_duplicates()['spid'].value_counts().sort_index()
 labels, height = zip(*counts_OGid.items())

### 2023-03-30 chapter1/figures/components/components.py

@@ -8,7 +8,6 @@ import matplotlib.pyplot as plt
 import networkx as nx
 import pandas as pd
 from matplotlib.figure import SubplotParams
-from matplotlib.gridspec import GridSpec
 from numpy import linspace
 
 component_id_groups = [('000C', '001E'),
@@ -84,7 +83,7 @@ if not os.path.exists('out/'):
 
 for component_ids in component_id_groups:
     fig = plt.figure(figsize=(fig_width, fig_height), subplotpars=SubplotParams(0.25, 0.25, 0.9, 0.9, 0, 0))
-    gs = GridSpec(2, len(component_ids), wspace=wspace, hspace=hspace, height_ratios=height_ratios)
+    gs = plt.GridSpec(2, len(component_ids), wspace=wspace, hspace=hspace, height_ratios=height_ratios)
     panel_width = fig_width / (len(component_ids) + (len(component_ids) - 1) * wspace)
     panel_height = fig_height * height_ratios[0] / (sum(height_ratios) + hspace)
 

### 2023-03-30 chapter1/figures/insertion_training/insertion_training.py

@@ -5,7 +5,6 @@ import json
 
 import matplotlib.pyplot as plt
 from matplotlib.figure import SubplotParams
-from matplotlib.gridspec import GridSpec
 from matplotlib.lines import Line2D
 
 spid_regex = r'spid=([a-z]+)'
@@ -39,7 +38,7 @@ if not os.path.exists('out/'):
     os.mkdir('out/')
 
 fig = plt.figure(figsize=(7.5, 5.5), subplotpars=SubplotParams(0.3, 0.225, 0.9, 0.9, 0, 0))
-gs = GridSpec(2, 4, height_ratios=[1, 1.5])
+gs = plt.GridSpec(2, 4, height_ratios=[1, 1.5])
 
 # Plot state distribution
 counts = {label: 0 for label in state_labels}

### 2023-03-30 chapter1/figures/insertion_trim/insertion_trim.py

@@ -6,7 +6,6 @@ import matplotlib.pyplot as plt
 import numpy as np
 import pandas as pd
 from matplotlib.figure import SubplotParams
-from matplotlib.gridspec import GridSpec
 
 # Load OGids
 OGids = []
@@ -33,7 +32,7 @@ if not os.path.exists('out/'):
     os.mkdir('out')
 
 fig = plt.figure(figsize=(7.5, 8), subplotpars=SubplotParams(0.25, 0.325, 0.9, 0.925))
-gs = GridSpec(4, 2)
+gs = plt.GridSpec(4, 2)
 
 # Pie chart by presence of trims
 union = set(df1['OGid']) | set(df2['OGid'])

### 2023-03-30 chapter1/figures/missing_training/missing_training.py

@@ -5,7 +5,6 @@ import json
 
 import matplotlib.pyplot as plt
 from matplotlib.figure import SubplotParams
-from matplotlib.gridspec import GridSpec
 from matplotlib.lines import Line2D
 
 ppid_regex = r'ppid=([A-Za-z0-9_.]+)'
@@ -42,7 +41,7 @@ if not os.path.exists('out/'):
     os.mkdir('out/')
 
 fig = plt.figure(figsize=(7.5, 5.5), subplotpars=SubplotParams(0.325, 0.225, 0.9, 0.9, 0, 0))
-gs = GridSpec(2, 6, height_ratios=[1, 1.5])
+gs = plt.GridSpec(2, 6, height_ratios=[1, 1.5])
 
 # Plot state distribution
 counts = {label: 0 for label in state_labels}

### 2023-03-30 chapter1/figures/missing_trim/missing_trim.py

@@ -5,7 +5,6 @@ import os
 import matplotlib.pyplot as plt
 import pandas as pd
 from matplotlib.figure import SubplotParams
-from matplotlib.gridspec import GridSpec
 
 # Load OGids
 OGids = []
@@ -28,7 +27,7 @@ if not os.path.exists('out/'):
     os.mkdir('out/')
 
 fig = plt.figure(figsize=(7.5, 4), subplotpars=SubplotParams(0.25, 0.325, 0.9, 0.925))
-gs = GridSpec(2, 2)
+gs = plt.GridSpec(2, 2)
 
 # Number of OGs with trims
 values = [len(set(OGids)) - df['OGid'].nunique(), df['OGid'].nunique()]

### 2023-03-30 chapter1/figures/paralogs/paralogs.py

@@ -5,7 +5,6 @@ import os
 import matplotlib.pyplot as plt
 import pandas as pd
 from matplotlib.figure import SubplotParams
-from matplotlib.gridspec import GridSpec
 
 # Load OGs
 OGs1 = {}
@@ -36,7 +35,7 @@ if not os.path.exists('out/'):
     os.mkdir('out/')
 
 fig = plt.figure(figsize=(7.5, 4.5), subplotpars=SubplotParams(0.15, 0.275, 0.85, 0.9, 0, 0))
-gs = GridSpec(2, 2)
+gs = plt.GridSpec(2, 2)
 
 counts = (df['delta'] == 0).value_counts()
 labels = [('w/o in-paralogs' if idx else 'w/ in-paralogs') + f'\n({value:,})' for idx, value in zip(counts.index, counts.values)]

### 2023-03-30 chapter1/figures/clusters/clusters.py
Removed obscure subplotparams argument from figure and replaced with explicitly setting axes

@@ -4,7 +4,6 @@ import os
 
 import matplotlib.pyplot as plt
 import pandas as pd
-from matplotlib.figure import SubplotParams
 
 # Load genomes
 spids = set()
@@ -43,14 +42,15 @@ gnidnum = OGs['gnid'].nunique()
 if not os.path.exists('out/'):
     os.mkdir('out/')
 
-fig = plt.figure(figsize=(7.5, 4.5), subplotpars=SubplotParams(0.25, 0.2, 1, 0.9))
+fig = plt.figure(figsize=(7.5, 4.5))
 gs = plt.GridSpec(2, 2)
+rect = (0.25, 0.2, 0.7, 0.7)
 
 counts_OGid = OGs[['spid', 'OGid']].drop_duplicates()['spid'].value_counts().sort_index()
 labels, height = zip(*counts_OGid.items())
 xs = list(range(len(labels)))
 subfig = fig.add_subfigure(gs[0, 0], facecolor='none')
-ax = subfig.add_subplot()
+ax = subfig.add_axes(rect)
 ax.bar(xs, height)
 ax.set_xticks(xs, labels, rotation=60, fontsize=6)
 ax.set_xlabel('Species')
@@ -60,7 +60,7 @@ subfig.suptitle('A', x=0.025, y=0.975, fontweight='bold')
 counts = groups['spid'].nunique().value_counts()
 xs, height = zip(*counts.items())
 subfig = fig.add_subfigure(gs[0, 1], facecolor='none')
-ax = subfig.add_subplot()
+ax = subfig.add_axes(rect)
 ax.bar(xs, height)
 ax.set_xlabel('Number of species in OG')
 ax.set_ylabel('Number of OGs')
@@ -69,7 +69,7 @@ subfig.suptitle('B', x=0.025, y=0.975, fontweight='bold')
 counts = OGs.groupby('ppid')['OGid'].nunique().value_counts()
 xs, height = zip(*counts.items())
 subfig = fig.add_subfigure(gs[1, 0], facecolor='none')
-ax = subfig.add_subplot()
+ax = subfig.add_axes(rect)
 ax.bar(counts.index, counts.values, width=1)
 ax.set_xlabel('Number of OGs associated with protein')
 ax.set_ylabel('Number of proteins')
@@ -77,7 +77,7 @@ subfig.suptitle('C', x=0.025, y=0.975, fontweight='bold')
 
[…]

### 2023-03-30 chapter1/figures/components/components.py

@@ -7,7 +7,6 @@ import matplotlib as mpl
 import matplotlib.pyplot as plt
 import networkx as nx
 import pandas as pd
-from matplotlib.figure import SubplotParams
 from numpy import linspace
 
 component_id_groups = [('000C', '001E'),
@@ -19,9 +18,8 @@ component_id_groups = [('000C', '001E'),
 
 fig_width = 7.5
 fig_height = 5.625
-wspace = 0
-hspace = 0
 height_ratios = [2, 1]
+rect = (0.25, 0.25, 0.65, 0.65)
 dpi = 300
 
 margin_width = 0
@@ -82,10 +80,10 @@ if not os.path.exists('out/'):
     os.mkdir('out/')
 
 for component_ids in component_id_groups:
-    fig = plt.figure(figsize=(fig_width, fig_height), subplotpars=SubplotParams(0.25, 0.25, 0.9, 0.9, 0, 0))
-    gs = plt.GridSpec(2, len(component_ids), wspace=wspace, hspace=hspace, height_ratios=height_ratios)
-    panel_width = fig_width / (len(component_ids) + (len(component_ids) - 1) * wspace)
-    panel_height = fig_height * height_ratios[0] / (sum(height_ratios) + hspace)
+    fig = plt.figure(figsize=(fig_width, fig_height))
+    gs = plt.GridSpec(2, len(component_ids), height_ratios=height_ratios)
+    panel_width = fig_width / len(component_ids)
+    panel_height = fig_height * height_ratios[0] / sum(height_ratios)
 
     for i, (component_id, panel_label) in enumerate(zip(component_ids, panel_labels)):
         # Create graph
@@ -150,7 +148,7 @@ for component_ids in component_id_groups:
         ax.axis('off')
 
     subfig = fig.add_subfigure(gs[1, 0])
-    ax = subfig.add_subplot(aspect='equal')
+    ax = subfig.add_axes(rect, aspect='equal')
     poly = ax.hexbin(df['ppidnum'], df['gnidnum'], bins='log', gridsize=30, mincnt=1, linewidth=0)
     ax.set_xlabel('Number of proteins in component')
     ax.set_ylabel('Number of unique\ngenes in component')
@@ -165,7 +163,7 @@ for component_ids in component_id_groups:
         counts[degree] = counts.get(degree, 0) + 1
 
     subfig = fig.add_subfigure(gs[1, 1])
-    ax = subfig.add_subplot()
[…]

### 2023-03-30 chapter1/figures/insertion_training/insertion_training.py

@@ -4,7 +4,6 @@ import os
 import json
 
 import matplotlib.pyplot as plt
-from matplotlib.figure import SubplotParams
 from matplotlib.lines import Line2D
 
 spid_regex = r'spid=([a-z]+)'
@@ -37,8 +36,9 @@ with open('../../orthology_inference/analysis/ortho_MSA/insertion_hmm/out/model.
 if not os.path.exists('out/'):
     os.mkdir('out/')
 
-fig = plt.figure(figsize=(7.5, 5.5), subplotpars=SubplotParams(0.3, 0.225, 0.9, 0.9, 0, 0))
+fig = plt.figure(figsize=(7.5, 5.5))
 gs = plt.GridSpec(2, 4, height_ratios=[1, 1.5])
+gridspec_kw = {'left': 0.3, 'right': 0.9, 'top': 0.9, 'bottom': 0.225, 'hspace': 0.15}
 
 # Plot state distribution
 counts = {label: 0 for label in state_labels}
@@ -56,7 +56,7 @@ subfig.suptitle('A', x=0.025, y=0.975, fontweight='bold')
 xs = [record['iter_num'] for record in history]
 ys = [record['ll'] for record in history]
 subfig = fig.add_subfigure(gs[0, 2:])
-ax = subfig.add_subplot()
+ax = subfig.add_axes((0.3, 0.225, 0.6, 0.675))
 ax.plot(xs, ys)
 ax.set_xlabel('Iteration')
 ax.set_ylabel('Conditional\nlog-likelihood')
@@ -65,7 +65,7 @@ subfig.suptitle('B', x=0.025, y=0.975, fontweight='bold')
 # Plot model parameters
 params = ['pi', 'q0', 'q1']
 subfig = fig.add_subfigure(gs[1, 0], facecolor='none')
-axs = subfig.subplots(len(params), 1, sharex=True, gridspec_kw={'hspace': 0.15})
+axs = subfig.subplots(len(params), 1, sharex=True, gridspec_kw=gridspec_kw)
 for label, color in zip(state_labels, state_colors):
     for ax, param in zip(axs, params):
         xs = [record['iter_num'] for record in history]
@@ -77,7 +77,7 @@ subfig.suptitle('C', x=0.025, y=0.975, fontweight='bold')
 
 params = ['p0', 'p1']
 subfig = fig.add_subfigure(gs[1, 1], facecolor='none')
-axs = subfig.subplots(len(params), 1, sharex=True, gridspec_kw={'hspace': 0.15})
[…]

### 2023-03-30 chapter1/figures/insertion_trim/insertion_trim.py

@@ -5,7 +5,6 @@ import os
 import matplotlib.pyplot as plt
 import numpy as np
 import pandas as pd
-from matplotlib.figure import SubplotParams
 
 # Load OGids
 OGids = []
@@ -31,8 +30,9 @@ groups2 = df2.groupby('OGid')
 if not os.path.exists('out/'):
     os.mkdir('out')
 
-fig = plt.figure(figsize=(7.5, 8), subplotpars=SubplotParams(0.25, 0.325, 0.9, 0.925))
+fig = plt.figure(figsize=(7.5, 8))
 gs = plt.GridSpec(4, 2)
+rect = (0.25, 0.325, 0.65, 0.6)
 
 # Pie chart by presence of trims
 union = set(df1['OGid']) | set(df2['OGid'])
@@ -51,7 +51,7 @@ subfig.suptitle('A', x=0.025, y=0.975, fontweight='bold')
 
 # Hexbin of posterior2 vs posterior3 in region trims
 subfig = fig.add_subfigure(gs[0, 1], facecolor='none')
-ax = subfig.add_subplot()
+ax = subfig.add_axes(rect)
 hb = ax.hexbin(df2['norm2'], df2['norm3'], bins='log', gridsize=25, mincnt=1, linewidth=0)
 ax.set_xlabel('Average state 2\nposterior in region trim')
 ax.set_ylabel('Average state 3\nposterior in region trim')
@@ -61,7 +61,7 @@ subfig.suptitle('B', x=0.025, y=0.975, fontweight='bold')
 # Distribution of number of segment trims in OGs
 counts = groups1.size().value_counts()
 subfig = fig.add_subfigure(gs[1, 0], facecolor='none')
-ax = subfig.add_subplot()
+ax = subfig.add_axes(rect)
 ax.bar(counts.index, counts.values, width=1)
 ax.set_xlabel('Number of segment trims in OG')
 ax.set_ylabel('Number of OGs')
@@ -70,7 +70,7 @@ subfig.suptitle('C', x=0.025, y=0.975, fontweight='bold')
 # Distribution of number of region trims in OGs
 counts = groups2.size().value_counts()
 subfig = fig.add_subfigure(gs[1, 1], facecolor='none')
-ax = subfig.add_subplot()
+ax = subfig.add_axes(rect)
 ax.bar(counts.index, counts.values, width=1)
 ax.set_xlabel('Number of region trims in OG')
 ax.set_ylabel('Number of OGs')
[…]

### 2023-03-30 chapter1/figures/missing_training/missing_training.py

@@ -4,7 +4,6 @@ import os
 import json
 
 import matplotlib.pyplot as plt
-from matplotlib.figure import SubplotParams
 from matplotlib.lines import Line2D
 
 ppid_regex = r'ppid=([A-Za-z0-9_.]+)'
@@ -40,8 +39,9 @@ with open('../../orthology_inference/analysis/ortho_MSA/missing_hmm/out/model.js
 if not os.path.exists('out/'):
     os.mkdir('out/')
 
-fig = plt.figure(figsize=(7.5, 5.5), subplotpars=SubplotParams(0.325, 0.225, 0.9, 0.9, 0, 0))
+fig = plt.figure(figsize=(7.5, 5.5))
 gs = plt.GridSpec(2, 6, height_ratios=[1, 1.5])
+gridspec_kw = {'left': 0.325, 'right': 0.9, 'top': 0.9, 'bottom': 0.225, 'hspace': 0.15}
 
 # Plot state distribution
 counts = {label: 0 for label in state_labels}
@@ -59,7 +59,7 @@ subfig.suptitle('A', x=0.025, y=0.975, fontweight='bold')
 xs = [record['iter_num'] for record in history]
 ys = [record['ll'] for record in history]
 subfig = fig.add_subfigure(gs[0, 3:])
-ax = subfig.add_subplot()
+ax = subfig.add_axes((0.325, 0.225, 0.575, 0.675))
 ax.plot(xs, ys)
 ax.set_xlabel('Iteration')
 ax.set_ylabel('Conditional\nlog-likelihood')
@@ -68,7 +68,7 @@ subfig.suptitle('B', x=0.025, y=0.975, fontweight='bold')
 # Plot model parameters
 params = ['pi', 'q0', 'q1']
 subfig = fig.add_subfigure(gs[1, :2], facecolor='none')
-axs = subfig.subplots(len(params), 1, sharex=True, gridspec_kw={'hspace': 0.15})
+axs = subfig.subplots(len(params), 1, sharex=True, gridspec_kw=gridspec_kw)
 for label, color in zip(state_labels, state_colors):
     for ax, param in zip(axs, params):
         xs = [record['iter_num'] for record in history]
@@ -81,7 +81,7 @@ subfig.suptitle('C', x=0.025, y=0.975, fontweight='bold')
 
 params = ['p0', 'p1']
 subfig = fig.add_subfigure(gs[1, 2:4], facecolor='none')
[…]

### 2023-03-30 chapter1/figures/missing_trim/missing_trim.py

@@ -4,7 +4,6 @@ import os
 
 import matplotlib.pyplot as plt
 import pandas as pd
-from matplotlib.figure import SubplotParams
 
 # Load OGids
 OGids = []
@@ -26,8 +25,9 @@ OGid_groups = df.groupby(['OGid'])
 if not os.path.exists('out/'):
     os.mkdir('out/')
 
-fig = plt.figure(figsize=(7.5, 4), subplotpars=SubplotParams(0.25, 0.325, 0.9, 0.925))
+fig = plt.figure(figsize=(7.5, 4))
 gs = plt.GridSpec(2, 2)
+rect = (0.25, 0.325, 0.65, 0.6)
 
 # Number of OGs with trims
 values = [len(set(OGids)) - df['OGid'].nunique(), df['OGid'].nunique()]
@@ -40,7 +40,7 @@ subfig.suptitle('A', x=0.025, y=0.975, fontweight='bold')
 # Distribution of number sequences with trims in OGs
 counts = OGid_groups['ppid'].nunique().value_counts()
 subfig = fig.add_subfigure(gs[0, 1], facecolor='none')
-ax = subfig.add_subplot()
+ax = subfig.add_axes(rect)
 ax.bar(counts.index, counts.values, width=1)
 ax.set_xlabel('Number of unique sequences with trims in OG')
 ax.set_ylabel('Number of OGs')
@@ -48,7 +48,7 @@ subfig.suptitle('B', x=0.025, y=0.975, fontweight='bold')
 
 # Distribution of length of trims
 subfig = fig.add_subfigure(gs[1, 0], facecolor='none')
-ax = subfig.add_subplot()
+ax = subfig.add_axes(rect)
 ax.hist(df['length'], bins=100)
 ax.set_xlabel('Length of trim')
 ax.set_ylabel('Number of trims')
@@ -56,7 +56,7 @@ subfig.suptitle('C', x=0.025, y=0.975, fontweight='bold')
 
 # Distribution of length ratio of trims
 subfig = fig.add_subfigure(gs[1, 1], facecolor='none')
-ax = subfig.add_subplot()
+ax = subfig.add_axes(rect)
 ax.hist(df['length_ratio'], bins=50)
 ax.set_xlabel('Length ratio of trim')
 ax.set_ylabel('Number of trims')

### 2023-03-30 chapter1/figures/paralogs/paralogs.py

@@ -4,7 +4,6 @@ import os
 
 import matplotlib.pyplot as plt
 import pandas as pd
-from matplotlib.figure import SubplotParams
 
 # Load OGs
 OGs1 = {}
@@ -34,18 +33,19 @@ df = pd.DataFrame(rows)
 if not os.path.exists('out/'):
     os.mkdir('out/')
 
-fig = plt.figure(figsize=(7.5, 4.5), subplotpars=SubplotParams(0.15, 0.275, 0.85, 0.9, 0, 0))
+fig = plt.figure(figsize=(7.5, 4.5))
 gs = plt.GridSpec(2, 2)
+rect = (0.15, 0.275, 0.7, 0.625)
 
 counts = (df['delta'] == 0).value_counts()
 labels = [('w/o in-paralogs' if idx else 'w/ in-paralogs') + f'\n({value:,})' for idx, value in zip(counts.index, counts.values)]
 subfig = fig.add_subfigure(gs[0, 0])
-ax = subfig.add_subplot()
+ax = subfig.add_axes(rect)
 ax.pie(counts.values, labels=labels, labeldistance=2, textprops={'ha': 'center'})
 subfig.suptitle('A', x=0.025, y=0.975, fontweight='bold')
 
 subfig = fig.add_subfigure(gs[1, 0])
-ax = subfig.add_subplot()
+ax = subfig.add_axes(rect)
 hb = ax.hexbin(df['ppidnum1'], df['ppidnum2'], bins='log', gridsize=25, mincnt=1, linewidth=0)
 ax.set_xlabel('Number of proteins\nin OG w/o in-paralogs')
 ax.set_ylabel('Number of proteins\nin OG w/ in-paralogs')
@@ -55,14 +55,14 @@ ax.set_aspect(1)
 
 counts = df.loc[df['delta'] != 0, 'delta'].value_counts()
 subfig = fig.add_subfigure(gs[0, 1])
-ax = subfig.add_subplot()
+ax = subfig.add_axes(rect)
 ax.bar(counts.index, counts.values, width=1)
 ax.set_xlabel('Number of in-paralogs in OG')
 ax.set_ylabel('Number of OGs')
 subfig.suptitle('B', x=0.025, y=0.975, fontweight='bold')
 
 subfig = fig.add_subfigure(gs[1, 1])
-ax = subfig.add_subplot()
+ax = subfig.add_axes(rect)
 ax.bar(counts.index, counts.values, width=1)
[…]

### 2023-04-04 chapter2/chapter2.tex
Added new methods section for definition of features
Added citation for scikit-learn

@@
 
+\subsection{Definition and calculation of features}
+
 \subsection{Brownian motion and Ornstein-Uhlenbeck analyses}
@@
 The code used to produce the results and analyses is available at \url{https://github.com/marcsingleton/IDR_evolution2023}.
-The following Python libraries were used: matplotlib, NumPy, pandas, and SciPy~\cite{Hunter2007, Harris2020, McKinney2010, Virtanen2020}.
+The following Python libraries were used: matplotlib, NumPy, pandas, SciPy, and scikit-learn~\cite{Hunter2007, Harris2020, McKinney2010, Virtanen2020, Pedregosa2011}.
 Relevant output files are available in the supporting information.

### 2023-04-04 chapter2/chapter2.tex
Added methods heading for GO term analyses
Added text for feature definition and supplement tables

@@
 \subsection{Definition and calculation of features}
+Features were calculated as in Zarin \textit{et al.} with the following modifications~\cite{Zarin2019}.
+The regular expression used for polar residue fraction was [QNSTCH], which, in contrast to the original study, excludes glycine residues.
+Additionally, length, expressed in log scale, was replaced with a feature proportional radius of gyration for an excluded-volume polymer~\cite{Flory1949}.
+Because the radii of gyration of chemically denatured proteins closely match the values expected for equivalent random coils, we felt this feature would better capture the relationship between an IDR's length and its biophysical properties~\cite{Kohn2004}.
+Finally, several motifs from ELM were replaced with their metazoan counterparts or updated versions of the same entries~\cite{Kumar2021}.
+These differences are noted in the supplementary data.
+Kappa, omega, SCD, hydropathy, PPII propensity, and Wootton-Federhen sequence complexity were calculated with localCIDER 0.1.19~\cite{Holehouse2017}.
+Isoelectric point was calculated with the Python package isoelectric, which is available on PyPI or at \url{https://isoelectric.org/}~\cite{Kozlowski2016}.
+Otherwise, features were calculated with custom code.
+A full list of features and their definitions is given in Table~\ref{stable:features} and Table~\ref{stable:regexes}.
 
@@
 \subsection{Simulation analyses}
+
+\subsection{GO term analyses}
 

### 2023-04-04 chapter2/chapter2_supporting.tex
Added text for feature definition and supplement tables

Here's where my chapter 2 appendix will go!

\begin{landscape}
\footnotesize
\begin{longtable}{|l|l|l|l|l|l|}
\caption{\textbf{Features and their definitions.}}
\label{stable:features}
% Start first head
\hline
\textbf{Feature ID}    & \textbf{Feature name}                                                            & \textbf{Group name}                                                & \textbf{Source} & \textbf{Description}                                                                                                             & \begin{tabular}[c]{@{}l@{}}\textbf{Changes from}\\\textbf{Zarin \textit{et al.}}\end{tabular}
\endfirsthead

% Start head
\multicolumn{6}{l}
{\textbf{\tablename\ \thetable} (continued)} \\
\hline
\textbf{Feature ID}    & \textbf{Feature name}                                                            & \textbf{Group name}                                                & \textbf{Source} & \textbf{Description}                                                                                                             & \begin{tabular}[c]{@{}l@{}}\textbf{Changes from}\\\textbf{Zarin \textit{et al.}}\end{tabular}
\endhead

% Start foot
\multicolumn{6}{|c|}{Continued on next page} \\
\hline
\endfoot

% Start last foot
\endlastfoot

\hline
fraction\_S            & S fraction                                                                       & \begin{tabular}[c]{@{}l@{}}amino acid\\content\end{tabular}        &                 &                                                                                                                                  &                                                                                                \\
\hline
fraction\_P            & P fraction                                                                       & \begin{tabular}[c]{@{}l@{}}amino acid\\content\end{tabular}        &                 &                                                                                                                                  &                                                                                                \\
\hline
fraction\_T            & T fraction                                                                       & \begin{tabular}[c]{@{}l@{}}amino acid\\content\end{tabular}        &                 &                                                                                                                                  &                                                                                                \\
\hline
fraction\_A            & A fraction                                                                       & \begin{tabular}[c]{@{}l@{}}amino acid\\content\end{tabular}        &                 &                                                                                                                                  &                                                                                                \\
\hline
fraction\_H            & H fraction                                                                       & \begin{tabular}[c]{@{}l@{}}amino acid\\content\end{tabular}        &                 &                                                                                                                                  &                                                                                                \\
\hline
fraction\_Q            & Q fraction                                                                       & \begin{tabular}[c]{@{}l@{}}amino acid\\content\end{tabular}        &                 &                                                                                                                                  &                                                                                                \\
\hline
fraction\_N            & N fraction                                                                       & \begin{tabular}[c]{@{}l@{}}amino acid\\content\end{tabular}        &                 &                                                                                                                                  &                                                                                                \\
\hline
fraction\_G            & G fraction                                                                       & \begin{tabular}[c]{@{}l@{}}amino acid\\content\end{tabular}        &                 &                                                                                                                                  &                                                                                                \\
\hline
FCR                    & \begin{tabular}[c]{@{}l@{}}fraction charged\\residues\end{tabular}               & \begin{tabular}[c]{@{}l@{}}charge\\properties\end{tabular}         &                 & basic residue fraction + acidic residue fraction                                                                                 &                                                                                                \\
\hline
NCPR                   & \begin{tabular}[c]{@{}l@{}}net charge\\per residue\end{tabular}                  & \begin{tabular}[c]{@{}l@{}}charge\\properties\end{tabular}         &                 & basic residue fraction - acidic residue fraction                                                                                 &                                                                                                \\
\hline
net\_charge            & net charge                                                                       & \begin{tabular}[c]{@{}l@{}}charge\\properties\end{tabular}         &                 & \#{[}RK] - \#{[}DE]                                                                                                              &                                                                                                \\
\hline
net\_charge\_P         & \begin{tabular}[c]{@{}l@{}}net charge with\\phosphorylation\end{tabular}         & \begin{tabular}[c]{@{}l@{}}charge\\properties\end{tabular}         &                 & \begin{tabular}[c]{@{}l@{}}net charge including phosphorylation of [ST]P\\consensus sites with -1.5 charge per site\end{tabular} &                                                                                                \\
\hline
kappa                  & kappa                                                                            & \begin{tabular}[c]{@{}l@{}}charge\\properties\end{tabular}         & localcider      & \begin{tabular}[c]{@{}l@{}}measure of separation between positively\\and negatively charged residues\end{tabular}                &                                                                                                \\
\hline
omega                  & omega                                                                            & \begin{tabular}[c]{@{}l@{}}charge\\properties\end{tabular}         & localcider      & \begin{tabular}[c]{@{}l@{}}measure of separation between charged\\residues or prolines and all other residues\end{tabular}       &                                                                                                \\
\hline
SCD                    & \begin{tabular}[c]{@{}l@{}}sequence charge\\decoration\end{tabular}              & \begin{tabular}[c]{@{}l@{}}charge\\properties\end{tabular}         & localcider      & \begin{tabular}[c]{@{}l@{}}measure of separation between positively\\and negatively charged residues\end{tabular}                &                                                                                                \\
\hline
RK\_ratio              & R/K ratio                                                                        & \begin{tabular}[c]{@{}l@{}}charge\\properties\end{tabular}         &                 & \begin{tabular}[c]{@{}l@{}}adjusted ratio of arginine to\\lysine residues: (\#R + 1)/(\#K + 1)\end{tabular}                      &                                                                                                \\
\hline
ED\_ratio              & E/D ratio                                                                        & \begin{tabular}[c]{@{}l@{}}charge\\properties\end{tabular}         &                 & \begin{tabular}[c]{@{}l@{}}adjusted ratio of glutamic acid to\\aspartic acid residues: (\#E + 1)/(\#D + 1)\end{tabular}          &                                                                                                \\
\hline
fraction\_acidic       & \begin{tabular}[c]{@{}l@{}}acidic\\residue fraction~\end{tabular}                & \begin{tabular}[c]{@{}l@{}}physiochemical\\properties\end{tabular} &                 &                                                                                                                                  &                                                                                                \\
\hline
fraction\_basic        & \begin{tabular}[c]{@{}l@{}}basic\\residue fraction\end{tabular}                  & \begin{tabular}[c]{@{}l@{}}physiochemical\\properties\end{tabular} &                 &                                                                                                                                  &                                                                                                \\
\hline
fraction\_aliphatic    & \begin{tabular}[c]{@{}l@{}}aliphatic\\residue fraction\end{tabular}              & \begin{tabular}[c]{@{}l@{}}physiochemical\\properties\end{tabular} &                 &                                                                                                                                  &                                                                                                \\
\hline
fraction\_polar        & \begin{tabular}[c]{@{}l@{}}polar\\residue fraction\end{tabular}                  & \begin{tabular}[c]{@{}l@{}}physiochemical\\properties\end{tabular} &                 &                                                                                                                                  & removed glycine                                                                                \\
\hline
fraction\_chainexp     & \begin{tabular}[c]{@{}l@{}}chain-expanding\\residue fraction\end{tabular}        & \begin{tabular}[c]{@{}l@{}}physiochemical\\properties\end{tabular} &                 &                                                                                                                                  &                                                                                                \\
\hline
fraction\_aromatic     & \begin{tabular}[c]{@{}l@{}}aromatic\\residue fraction\end{tabular}               & \begin{tabular}[c]{@{}l@{}}physiochemical\\properties\end{tabular} &                 &                                                                                                                                  &                                                                                                \\
\hline
fraction\_disorder     & \begin{tabular}[c]{@{}l@{}}disorder-promoting\\residue fraction\end{tabular}     & \begin{tabular}[c]{@{}l@{}}physiochemical\\properties\end{tabular} &                 &                                                                                                                                  &                                                                                                \\
\hline
radius\_gyration       & radius of gyration                                                               & \begin{tabular}[c]{@{}l@{}}physiochemical\\properties\end{tabular} &                 & number of residues to the 0.6 power                                                                                              & \begin{tabular}[c]{@{}l@{}}substituted\\for length\end{tabular}                                \\
\hline
hydropathy             & hydropathy                                                                       & \begin{tabular}[c]{@{}l@{}}physiochemical\\properties\end{tabular} & localcider      & normalized Kyte-Doolittle scale                                                                                                  &                                                                                                \\
\hline
isopoint               & isoelectric point                                                                & \begin{tabular}[c]{@{}l@{}}physiochemical\\properties\end{tabular} & isoelectric     & pH where charge of peptide is neutral                                                                                            &                                                                                                \\
\hline
PPII\_propensity       & PPII propensity                                                                  & \begin{tabular}[c]{@{}l@{}}physiochemical\\properties\end{tabular} & localcider      & \begin{tabular}[c]{@{}l@{}}propensity for proline to form\\left-handed helices\end{tabular}                                      &                                                                                                \\
\hline
repeat\_Q              & Q repeat fraction                                                                & \begin{tabular}[c]{@{}l@{}}repeats and\\complexity\end{tabular}    &                 & fraction 2 or more consecutive Q                                                                                                 &                                                                                                \\
\hline
repeat\_N              & N repeat fraction                                                                & \begin{tabular}[c]{@{}l@{}}repeats and\\complexity\end{tabular}    &                 & fraction 2 or more consecutive N                                                                                                 &                                                                                                \\
\hline
repeat\_S              & S repeat fraction                                                                & \begin{tabular}[c]{@{}l@{}}repeats and\\complexity\end{tabular}    &                 & fraction 2 or more consecutive S                                                                                                 &                                                                                                \\
\hline
repeat\_G              & G repeat fraction                                                                & \begin{tabular}[c]{@{}l@{}}repeats and\\complexity\end{tabular}    &                 & fraction 2 or more consectuive G                                                                                                 &                                                                                                \\
\hline
repeat\_E              & E repeat fraction                                                                & \begin{tabular}[c]{@{}l@{}}repeats and\\complexity\end{tabular}    &                 & fraction 2 or more consecutive E                                                                                                 &                                                                                                \\
\hline
repeat\_D              & D repeat fraction                                                                & \begin{tabular}[c]{@{}l@{}}repeats and\\complexity\end{tabular}    &                 & fraction 2 or more consecutive D                                                                                                 &                                                                                                \\
\hline
repeat\_K              & K repeat fraction                                                                & \begin{tabular}[c]{@{}l@{}}repeats and\\complexity\end{tabular}    &                 & fraction 2 or more consecutive K                                                                                                 &                                                                                                \\
\hline
repeat\_R              & R repeat fraction                                                                & \begin{tabular}[c]{@{}l@{}}repeats and\\complexity\end{tabular}    &                 & fraction 2 or more consecutive R                                                                                                 &                                                                                                \\
\hline
repeat\_P              & P repeat fraction                                                                & \begin{tabular}[c]{@{}l@{}}repeats and\\complexity\end{tabular}    &                 & fraction 2 or more consecutive P                                                                                                 &                                                                                                \\
\hline
repeat\_QN             & {[}QN] repeat fraction                                                           & \begin{tabular}[c]{@{}l@{}}repeats and\\complexity\end{tabular}    &                 & fraction 2 or more consecutive [QN]                                                                                              &                                                                                                \\
\hline
repeat\_RG             & {[}RG] repeat fraction                                                           & \begin{tabular}[c]{@{}l@{}}repeats and\\complexity\end{tabular}    &                 & fraction 2 or more consecutive [RG]                                                                                              &                                                                                                \\
\hline
repeat\_FG             & {[}FG] repeat fraction                                                           & \begin{tabular}[c]{@{}l@{}}repeats and\\complexity\end{tabular}    &                 & fraction 2 or more consecutive [FG]                                                                                              &                                                                                                \\
\hline
repeat\_SG             & {[}SG] repeat fraction                                                           & \begin{tabular}[c]{@{}l@{}}repeats and\\complexity\end{tabular}    &                 & fraction 2 or more consecutive [SG]                                                                                              &                                                                                                \\
\hline
repeat\_SR             & {[}SR] repeat fraction                                                           & \begin{tabular}[c]{@{}l@{}}repeats and\\complexity\end{tabular}    &                 & fraction 2 or more consecutive [SR]                                                                                              &                                                                                                \\
\hline
repeat\_AP             & {[}KAP] repeat fraction                                                          & \begin{tabular}[c]{@{}l@{}}repeats and\\complexity\end{tabular}    &                 & fraction 2 or more consecutive [KAP]                                                                                             &                                                                                                \\
\hline
repeat\_TS             & {[}PTS] repeat fraction                                                          & \begin{tabular}[c]{@{}l@{}}repeats and\\complexity\end{tabular}    &                 & fraction 2 or more consecutive [PTS]                                                                                             &                                                                                                \\
\hline
wf\_complexity         & \begin{tabular}[c]{@{}l@{}}Wootton-Federhen\\sequence complexity\end{tabular}    & \begin{tabular}[c]{@{}l@{}}repeats and\\complexity\end{tabular}    & localcider      & \begin{tabular}[c]{@{}l@{}}complexity based on SEG algorithm:\\blob length=IDR length, step size=1\end{tabular}                  &                                                                                                \\
\hline
CLV\_Separin\_Metazoa  & separase cleavage motif                                                          & motifs                                                             &                 &                                                                                                                                  & \begin{tabular}[c]{@{}l@{}}Metazoa motif\\instead of fungi\end{tabular}                        \\
\hline
DEG\_APCC\_KENBOX\_2   & \begin{tabular}[c]{@{}l@{}}APCC-binding\\destruction motif\end{tabular}          & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
\hline
DEG\_APCC\_TPR\_1      & APCC-TPR-docking motif                                                           & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
\hline
DOC\_CKS1\_1           & Cks1 ligand                                                                      & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
\hline
DOC\_MAPK\_DCC\_7      & MAPK docking motif                                                               & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
\hline
DOC\_MAPK\_gen\_1      & MAPK docking motif                                                               & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
\hline
DOC\_MAPK\_HePTP\_8    & MAPK docking motif                                                               & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
\hline
DOC\_PP1\_RVXF\_1      & PP1-docking motif RVXF                                                           & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
\hline
DOC\_PP2B\_PxIxI\_1    & \begin{tabular}[c]{@{}l@{}}calcineurin (PP2B)-\\docking motif PxIxI\end{tabular} & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
\hline
LIG\_APCC\_Cbox\_1     & \begin{tabular}[c]{@{}l@{}}APC/C\_Apc2-docking\\motif\end{tabular}               & motifs                                                             &                 &                                                                                                                                  & \begin{tabular}[c]{@{}l@{}}Metazoa motif\\instead of fungi\end{tabular}                        \\
\hline
LIG\_AP\_GAE\_1        & \begin{tabular}[c]{@{}l@{}}gamma-adaptin ear\\interaction motif\end{tabular}     & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
\hline
LIG\_CaM\_IQ\_9        & \begin{tabular}[c]{@{}l@{}}helical calmodulin\\binding motif\end{tabular}        & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
\hline
LIG\_EH\_1             & EH ligand                                                                        & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
\hline
LIG\_eIF4E\_1          & eIF4E binding motif                                                              & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
\hline
LIG\_GLEBS\_BUB3\_1    & GLEBS motif                                                                      & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
\hline
LIG\_LIR\_Gen\_1       & \begin{tabular}[c]{@{}l@{}}Atg8 protein\\family ligands\end{tabular}             & motifs                                                             &                 &                                                                                                                                  & \begin{tabular}[c]{@{}l@{}}same ELM entry,\\but updated regex\end{tabular}                     \\
\hline
LIG\_PCNA\_PIPBox\_1   & PCNA binding PIP box                                                             & motifs                                                             &                 &                                                                                                                                  & \begin{tabular}[c]{@{}l@{}}same ELM entry,\\but updated regex\end{tabular}                     \\
\hline
LIG\_SUMO\_SIM\_par\_1 & \begin{tabular}[c]{@{}l@{}}SUMO interaction\\site\end{tabular}                   & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
\hline
MOD\_CDK\_SPxK\_1      & \begin{tabular}[c]{@{}l@{}}CDK phosphorylation\\site\end{tabular}                & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
\hline
MOD\_LATS\_1           & \begin{tabular}[c]{@{}l@{}}LATS kinase\\phosphorylation motif\end{tabular}       & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
\hline
MOD\_SUMO\_for\_1      & sumoylation site                                                                 & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
\hline
TRG\_ER\_FFAT\_1       & FFAT motif                                                                       & motifs                                                             &                 &                                                                                                                                  & \begin{tabular}[c]{@{}l@{}}same ELM entry,\\but updated regex\end{tabular}                     \\
\hline
TRG\_Golgi\_diPhe\_1   & ER export signals                                                                & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
\hline
TRG\_NLS\_MonoExtN\_4  & \begin{tabular}[c]{@{}l@{}}classical nuclear\\localization signals\end{tabular}  & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
\hline
MOD\_CDK\_STP          & \begin{tabular}[c]{@{}l@{}}CDK phosphorylation\\motif\end{tabular}               & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
\hline
MOD\_MEC1              & \begin{tabular}[c]{@{}l@{}}Mec1 phosphorylation\\motif\end{tabular}              & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
\hline
MOD\_PRK1              & \begin{tabular}[c]{@{}l@{}}Prk1 phosphorylation\\motif\end{tabular}              & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
\hline
MOD\_IPL1              & \begin{tabular}[c]{@{}l@{}}Ipl1 phosphorylation\\motif\end{tabular}              & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
\hline
MOD\_PKA               & \begin{tabular}[c]{@{}l@{}}Pka phosphorylation\\motif\end{tabular}               & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
\hline
MOD\_CKII              & \begin{tabular}[c]{@{}l@{}}Ckii phosphorylation\\motif\end{tabular}              & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
\hline
MOD\_IME2              & \begin{tabular}[c]{@{}l@{}}Ime2 phosphorylation\\motif\end{tabular}              & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
\hline
DOC\_PRO               & proline rich motif                                                               & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
\hline
TRG\_ER\_HDEL          & ER localization motif                                                            & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
\hline
TRG\_MITOCHONDRIA      & \begin{tabular}[c]{@{}l@{}}mitochondrial\\localization motif\end{tabular}        & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
\hline
MOD\_ISOMERASE         & \begin{tabular}[c]{@{}l@{}}disulfide isomerase\\motif\end{tabular}               & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
\hline
TRG\_FG                & \begin{tabular}[c]{@{}l@{}}FG nucleoporin\\motif\end{tabular}                    & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
\hline
INT\_RGG               & RGG motif                                                                        & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
\hline
\end{longtable}
\end{landscape}

\begin{landscape}
\footnotesize
\begin{longtable}{|l|l|}
\caption{\textbf{Feature regular expresions.}}
\label{stable:regexes}
% Start first head
\hline
\textbf{Feature ID}    & \textbf{Regular expression}
\endfirsthead

% Start head
\multicolumn{2}{l}
{\textbf{\tablename\ \thetable} (continued)} \\
\hline
\textbf{Feature ID}    & \textbf{Regular expression}
\endhead

% Start foot
\multicolumn{2}{|c|}{Continued on next page} \\
\hline
\endfoot

% Start last foot
\endlastfoot

\hline
fraction\_S            & S                                                                                                                          \\
\hline
fraction\_P            & P                                                                                                                          \\
\hline
fraction\_T            & T                                                                                                                          \\
\hline
fraction\_A            & A                                                                                                                          \\
\hline
fraction\_H            & H                                                                                                                          \\
\hline
fraction\_Q            & Q                                                                                                                          \\
\hline
fraction\_N            & N                                                                                                                          \\
\hline
fraction\_G            & G                                                                                                                          \\
\hline
fraction\_acidic       & {[}DE]                                                                                                                     \\
\hline
fraction\_basic        & {[}RK]                                                                                                                     \\
\hline
fraction\_aliphatic    & {[}ALMIV]                                                                                                                  \\
\hline
fraction\_polar        & {[}QNSTCH]                                                                                                                 \\
\hline
fraction\_chainexp     & {[}EDRKP]                                                                                                                  \\
\hline
fraction\_aromatic     & {[}FYW]                                                                                                                    \\
\hline
fraction\_disorder     & {[}TAGRDHQKSEP]                                                                                                            \\
\hline
repeat\_Q              & Q\{2,\}                                                                                                                    \\
\hline
repeat\_N              & N\{2,\}                                                                                                                    \\
\hline
repeat\_S              & S\{2,\}                                                                                                                    \\
\hline
repeat\_G              & G\{2,\}                                                                                                                    \\
\hline
repeat\_E              & E\{2,\}                                                                                                                    \\
\hline
repeat\_D              & D\{2,\}                                                                                                                    \\
\hline
repeat\_K              & K\{2,\}                                                                                                                    \\
\hline
repeat\_R              & R\{2,\}                                                                                                                    \\
\hline
repeat\_P              & P\{2,\}                                                                                                                    \\
\hline
repeat\_QN             & {[}QN]\{2,\}                                                                                                               \\
\hline
repeat\_RG             & {[}RG]\{2,\}                                                                                                               \\
\hline
repeat\_FG             & {[}FG]\{2,\}                                                                                                               \\
\hline
repeat\_SG             & {[}SG]\{2,\}                                                                                                               \\
\hline
repeat\_SR             & {[}SR]\{2,\}                                                                                                               \\
\hline
repeat\_AP             & {[}KAP]\{2,\}                                                                                                              \\
\hline
repeat\_TS             & {[}PTS]\{2,\}                                                                                                              \\
\hline
CLV\_Separin\_Metazoa  & E[IMPVL][MLVP]R.                                                                                                           \\
\hline
DEG\_APCC\_KENBOX\_2   & .KEN.                                                                                                                      \\
\hline
DEG\_APCC\_TPR\_1      & .[ILM]R                                                                                                                    \\
\hline
DOC\_CKS1\_1           & {[}MPVLIFWYQ].(T)P..                                                                                                       \\
\hline
DOC\_MAPK\_DCC\_7      & {[}RK].\{2,4\}{[}LIVP]P.[LIV].[LIVMF]\textbar{}{[}RK].\{2,4\}{[}LIVP].P[LIV].[LIVMF]                                       \\
\hline
DOC\_MAPK\_gen\_1      & {[}KR]\{0,2\}{[}KR].\{0,2\}{[}KR].\{2,4\}{[}ILVM].[ILVF]                                                                   \\
\hline
DOC\_MAPK\_HePTP\_8    & ([LIV][\^{}P][\^{}P][RK]....[LIVMP].[LIV].[LIVMF])\textbar{}([LIV][\^{}P][\^{}P][RK][RK]G.\{4,7\}{[}LIVMP].[LIV].[LIVMF])  \\
\hline
DOC\_PP1\_RVXF\_1      & ..[RK].\{0,1\}{[}VIL][\^{}P][FW].                                                                                          \\
\hline
DOC\_PP2B\_PxIxI\_1    & .P[\^{}P]I[\^{}P][IV][\^{}P]                                                                                               \\
\hline
LIG\_APCC\_Cbox\_1     & {[}DE]R[YFH][ILFVM][PAG].R                                                                                                 \\
\hline
LIG\_AP\_GAE\_1        & {[}DE][DES][DEGAS]F[SGAD][DEAP][LVIMFD]                                                                                    \\
\hline
LIG\_CaM\_IQ\_9        & {[}ACLIVTM][\^{}P][\^{}P][ILVMFCT]Q[\^{}P][\^{}P][\^{}P][RK][\^{}P]\{4,5\}{[}RKQ][\^{}P][\^{}P]                            \\
\hline
LIG\_EH\_1             & .NPF.                                                                                                                      \\
\hline
LIG\_eIF4E\_1          & Y....L[VILMF]                                                                                                              \\
\hline
LIG\_GLEBS\_BUB3\_1    & {[}EN][FYLW][NSQ].EE[ILMVF][\^{}P][LIVMFA]                                                                                 \\
\hline
LIG\_LIR\_Gen\_1       & {[}EDST].\{0,2\}{[}WFY][\^{}RKPGWFY][\^{}PG][ILVFM]((.\{0,4\}{[}PLAFIVMY])\textbar{}(\$)\textbar{}(.\{0,3\}{[}ED]))        \\
\hline
LIG\_PCNA\_PIPBox\_1   & {[}QM].[\^{}FHWY][LIVM][\^{}P][\^{}PFWYMLIV](([FYHL][FYW])\textbar{}([FYH][FYWL]))..                                       \\
\hline
LIG\_SUMO\_SIM\_par\_1 & {[}DEST]\{0,5\}.[VILPTM][VIL][DESTVILMA][VIL].\{0,1\}{[}DEST]\{1,10\}                                                      \\
\hline
MOD\_CDK\_SPxK\_1      & ...([ST])P.[KR]                                                                                                            \\
\hline
MOD\_LATS\_1           & H.[KR]..([ST])[\^{}P]                                                                                                      \\
\hline
MOD\_SUMO\_for\_1      & {[}VILMAFP](K).E                                                                                                           \\
\hline
TRG\_ER\_FFAT\_1       & {[}EDS].\{0,4\}{[}ED][FY][FYKREM][DE][AC].\{1,2\}{[}EDST]                                                                  \\
\hline
TRG\_Golgi\_diPhe\_1   & Q.\{6,6\}FF.\{6,7\}                                                                                                        \\
\hline
TRG\_NLS\_MonoExtN\_4  & (([PKR].\{0,1\}{[}\^{}DE])\textbar{}([PKR]))((K[RK])\textbar{}(RK))(([\^{}DE][KR])\textbar{}([KR][\^{}DE]))[\^{}DE]        \\
\hline
MOD\_CDK\_STP          & {[}ST]P                                                                                                                    \\
\hline
MOD\_MEC1              & {[}ST]Q                                                                                                                    \\
\hline
MOD\_PRK1              & {[}LIVM]....TG                                                                                                             \\
\hline
MOD\_IPL1              & {[}RK].[ST][LIV]                                                                                                           \\
\hline
MOD\_PKA               & R[RK].S                                                                                                                    \\
\hline
MOD\_CKII              & {[}ST][DE].[DE]                                                                                                            \\
\hline
MOD\_IME2              & RP.[ST]                                                                                                                    \\
\hline
DOC\_PRO               & P..P                                                                                                                       \\
\hline
TRG\_ER\_HDEL          & HDEL                                                                                                                       \\
\hline
TRG\_MITOCHONDRIA      & {[}MR]L[RK]                                                                                                                \\
\hline
MOD\_ISOMERASE         & C..C                                                                                                                       \\
\hline
TRG\_FG                & F.FG\textbar{}GLFG                                                                                                         \\
\hline
INT\_RGG               & RGG\textbar{}RG                                                                                                            \\
\hline
\end{longtable}
\end{landscape}

### 2023-04-04 chapter2/chapter2_supporting.tex
Added citation link to table header

@@
 \hline
-\textbf{Feature ID}    & \textbf{Feature name}                                                            & \textbf{Group name}                                                & \textbf{Source} & \textbf{Description}                                                                                                             & \begin{tabular}[c]{@{}l@{}}\textbf{Changes from}\\\textbf{Zarin \textit{et al.}}\end{tabular}
+\textbf{Feature ID}    & \textbf{Feature name}                                                            & \textbf{Group name}                                                & \textbf{Source} & \textbf{Description}                                                                                                             & \begin{tabular}[c]{@{}l@{}}\textbf{Changes from}\\\textbf{Zarin \textit{et al.}}~\cite{Zarin2019}\end{tabular}
 \endfirsthead
@@
 \hline
-\textbf{Feature ID}    & \textbf{Feature name}                                                            & \textbf{Group name}                                                & \textbf{Source} & \textbf{Description}                                                                                                             & \begin{tabular}[c]{@{}l@{}}\textbf{Changes from}\\\textbf{Zarin \textit{et al.}}\end{tabular}
+\textbf{Feature ID}    & \textbf{Feature name}                                                            & \textbf{Group name}                                                & \textbf{Source} & \textbf{Description}                                                                                                             & \begin{tabular}[c]{@{}l@{}}\textbf{Changes from}\\\textbf{Zarin \textit{et al.}~\cite{Zarin2019}}\end{tabular}
 \endhead

### 2023-04-04 chapter1/chapter1.tex
Fixed bad path in to new realignment figure
Added clarification to Gaussian process sequence weighting
Fixed calling missing data phylo-HMM as deletion phylo-HMM

@@
 \begin{figure}[h!]
-\includegraphics[width=\textwidth]{realignment/out/merged.png}
+\includegraphics[width=\textwidth]{realignment/out/realignment.png}
 \centering
@@
 All HMM algorithms were implemented with custom code which is available as the package Homomorph on PyPI.
-The insertion phylo-HMM was trained on 47,387 manually labeled columns in 14 alignments, and the deletion phylo-HMM was trained on 67,001 manually labeled positions in 23 sequences in 11 unique alignments (Fig.~\ref{sfig:insertion_training}, Fig.~\ref{sfig:missing_training}).
+The insertion phylo-HMM was trained on 47,387 manually labeled columns in 14 alignments, and the missing data phylo-HMM was trained on 67,001 manually labeled positions in 23 sequences in 11 unique alignments (Fig.~\ref{sfig:insertion_training}, Fig.~\ref{sfig:missing_training}).
 Because maximum-likelihood estimation of the model parameters yielded posterior decoding curves which toggled between hidden states too rapidly, the models were instead trained discriminatively~\cite{Krogh1999}.
@@
 The outer and inner derivative cutoffs were both 0.001.
-The mean number of non-gap symbols in a region was calculated using Gaussian process sequence weighting.
+The mean number of non-gap symbols in a region was calculated using Gaussian process sequence weighting as described in the section on selecting representative sequences.
 (The sequences with the five most non-gap symbols were also excluded to not bias the estimate with long poorly supported segments.) The final mean $\mu$ was taken as the minimum of this value and two.

### 2023-04-04 chapter2/chapter2.tex
Added method section for alignments and trees

@@
 \subsection{Alignment and tree generation}
+Alignments of single copy orthologs and the outputs of the missing data phylo-HMM were obtained from the analyses conducted in chapter~\ref{chapter:1}.
+Likewise, we used the LG consensus tree generated by the ``non-invariant, 100\% redundancy'' sampling strategy as the input or reference where indicated in subsequent phylogenetic analyses.
 

### 2023-04-04 chapter2/chapter2.tex
Added text for IDR prediction and filtering method

@@
 \subsection{IDR prediction and filtering}
+AUCPreD was used to predict the disorder score of each residue of each sequence in the alignments after the gap symbols were removed~\cite{Wang2016}.
+The resulting scores were then aligned using the original alignment.
+The average score for each position was calculated using Gaussian process sequencing weighting over the LG consensus tree~\cite{Altschul1989}.
+Any positions inferred as ``missing'' by the missing data phylo-HMM or to the left or right of the first or last non-gap symbol, respectively, were excluded.
+For simplicity, the Gaussian process weights were not re-calculated from a tree pruned of the corresponding tips, and instead the weights corresponding to the remaining sequences were re-normalized.
+The scores at any remaining positions with gap symbols were inferred by linear interpolation from the nearest scored position.
+
+The average disorder scores were converted into contiguous regions with the following method.
+First, two binary masks were defined as positions where the average score exceeded high and low cutoffs of 0.6 and 0.4, respectively.
+The low-cutoff mask was subjected to an additional binary dilation operation with a structuring element of size three to merge any contiguous regions separated by a small number of positions with scores below the cutoff.
+``Seed'' regions were then defined as 10 or more contiguous ``true'' positions in the high-cutoff mask, and ``disorder'' regions were obtained by expanding the seeds to the left and right until the first ``false'' position in the low-cutoff mask or the end of the alignment.
+``Order'' regions were defined as the complement of the disorder regions in each alignment.
+
+The regions were filtered with the following criteria.
+First, segments with non-standard amino acid symbols, which overlapped with any position labeled as ``missing'' by the missing data phylo-HMM, or whose number of non-gap symbols was below a length cutoff were removed.
+Regions whose remaining segments satisfied the set of phylogenetic diversity criteria detailed in Table~\ref{stable:diversity_criteria} were included in the final set.
+The length cutoffs were set to 30, 60, and 90 residues, which generated three distinct sets of regions.
 

### 2023-04-04 chapter2/chapter2.tex
Changed first person to passive in alignment and tree generation

@@
 Alignments of single copy orthologs and the outputs of the missing data phylo-HMM were obtained from the analyses conducted in chapter~\ref{chapter:1}.
-Likewise, we used the LG consensus tree generated by the ``non-invariant, 100\% redundancy'' sampling strategy as the input or reference where indicated in subsequent phylogenetic analyses.
+Likewise, the LG consensus tree generated by the ``non-invariant, 100\% redundancy'' sampling strategy was used as the input or reference where indicated in subsequent phylogenetic analyses.
 

### 2023-04-04 chapter2/chapter2.tex
Tweaked some wording to make methods less repetitive

@@
 The average disorder scores were converted into contiguous regions with the following method.
-First, two binary masks were defined as positions where the average score exceeded high and low cutoffs of 0.6 and 0.4, respectively.
-The low-cutoff mask was subjected to an additional binary dilation operation with a structuring element of size three to merge any contiguous regions separated by a small number of positions with scores below the cutoff.
+Two binary masks were defined as positions where the average score exceeded high and low cutoffs of 0.6 and 0.4, respectively.
+The low-cutoff mask was subjected to an additional binary dilation with a structuring element of size three to merge any contiguous regions separated by a small number of positions with scores below the cutoff.
 ``Seed'' regions were then defined as 10 or more contiguous ``true'' positions in the high-cutoff mask, and ``disorder'' regions were obtained by expanding the seeds to the left and right until the first ``false'' position in the low-cutoff mask or the end of the alignment.
-``Order'' regions were defined as the complement of the disorder regions in each alignment.
+``Order'' regions were taken as the complement of the disorder regions in each alignment.
 
@@
 \subsection{Fitting substitution models}
+Substitution matrices for disorder and order regions were fit to 25
 
@@
 Features were calculated as in Zarin \textit{et al.} with the following modifications~\cite{Zarin2019}.
-The regular expression used for polar residue fraction was [QNSTCH], which, in contrast to the original study, excludes glycine residues.
+The regular expression for polar residue fraction was [QNSTCH], which, in contrast to the original study, excludes glycine residues.
 Additionally, length, expressed in log scale, was replaced with a feature proportional radius of gyration for an excluded-volume polymer~\cite{Flory1949}.

### 2023-04-05 chapter2/chapter2.tex
Added fitting substitution models method

@@
 \subsection{Fitting substitution models}
-Substitution matrices for disorder and order regions were fit to 25
+To fit amino acid substitution matrices to disorder and order regions, 25 meta-alignments were constructed by randomly sampling 100,000 columns from the respective subsets in the filtered regions derived from the 30 residue length cutoff.
+To determine the effect of gaps, the maximum fraction of gaps was set at 0, 50, 100\%.
+The combination of the region types and sampling strategy yielded six different sets of meta-alignments.
+A GTR20 substitution model with four FreeRate categories and optimized state frequencies was fit to each meta-alignment using IQ-TREE 1.6.12~\cite{Nguyen2014}.
+Exchangeability and rate coefficients were normalized to make the average rate of each model equal to 1.
+All figures are derived from the maximum 50\% gap fraction meta-alignment set unless otherwise noted.
 

### 2023-04-05 chapter1/chapter1.tex
Clarified inference of species trees

@@
 Their combination yielded six different sets of meta-alignments.
-A tree was fit to each meta-alignment with the LG substitution model and four discrete gamma rate categories using IQ-TREE 1.6.12~\cite{Nguyen2014, Le2008, Yang1994}.
+A tree was fit to each meta-alignment with the LG substitution model, four discrete gamma rate categories, and optimized state frequencies using IQ-TREE 1.6.12~\cite{Nguyen2014, Le2008, Yang1994}.
 If the sampling strategy allowed invariant columns, an invariant rate category was included.
 The resulting trees from each set were merged into a majority consensus tree (Fig.~\ref{sfig:trees_LG}).
+All figures are derived from the maximum 50\% gap fraction meta-alignment set unless otherwise noted.
 

### 2023-04-06 chapter2/figures/substitution/substitution.py
Added substitution.py

@@ -1,0 +1,197 @@
+"""Make figure of results from fitting substitution models."""
+
+import os
+import re
+from collections import namedtuple
+
+import matplotlib.pyplot as plt
+import numpy as np
+import skbio
+from matplotlib.colors import Normalize
+from matplotlib.cm import ScalarMappable
+from src2.utils import read_iqtree, read_paml
+
+paml_order = ['A', 'R', 'N', 'D', 'C', 'Q', 'E', 'G', 'H', 'I', 'L', 'K', 'M', 'F', 'P', 'S', 'T', 'W', 'Y', 'V']
+labels = ['0R_disorder', '50R_disorder', '100R_disorder', '0R_order', '50R_order', '100R_order']
+labels_suffix = r'_[0-9]+'
+Record = namedtuple('Record', ['label', 'ematrix', 'rmatrix', 'freqs', 'length'])
+
+# Load LG model
+ematrix, freqs = read_paml('../../IDR_evolution/data/matrices/LG.paml', norm=True)
+rmatrix = freqs * ematrix
+
+# Load IQ-TREE matrices
+records = {}
+for label in labels:
+    file_labels = []
+    for path in os.listdir('../../IDR_evolution/analysis/evofit/iqtree_fit/out/'):
+        match = re.match(f'({label}{labels_suffix})\.iqtree', path)
+        if match:
+            file_labels.append(match.group(1))
+
+    ematrix_stack = []
+    rmatrix_stack = []
+    freqs_stack = []
+    length_stack = []
+    for file_label in sorted(file_labels):
+        record = read_iqtree(f'../../IDR_evolution/analysis/evofit/iqtree_fit/out/{file_label}.iqtree', norm=True)
+        ematrix, freqs = record['ematrix'], record['freqs']
+        rmatrix = freqs * ematrix
+
+        tree = skbio.read(f'../../IDR_evolution/analysis/evofit/iqtree_fit/out/{file_label}.treefile', 'newick', skbio.TreeNode)
+        length = tree.descending_branch_length()
+
+        ematrix_stack.append(ematrix)
+        rmatrix_stack.append(rmatrix)
+        freqs_stack.append(freqs)
+        length_stack.append(length)
+    records[label] = Record(label,
+                            np.stack(ematrix_stack),
[…]

### 2023-04-06 chapter1/figures/components/components.py
Made add_axes calls more uniform between scripts

@@ -130,7 +130,7 @@ for component_ids in component_id_groups:
         wlen = wmax - wmin
 
         subfig = fig.add_subfigure(gs[0, i])
-        ax = subfig.add_axes([0, 0, 1, 1], aspect='equal')
+        ax = subfig.add_axes((0, 0, 1, 1), aspect='equal')
 
         nx.draw_networkx_edges(nx_graph, positions, ax=ax, edgelist=edges, alpha=edge_alpha2, width=edge_width2, edge_color=ws, edge_cmap=cmap)
         nx.draw_networkx_nodes(nx_graph, positions, ax=ax, node_size=node_size2, linewidths=0, node_color=node_color2)

### 2023-04-06 chapter1/figures/percolation/percolation.py

@@ -212,8 +212,8 @@ for component_id in component_ids:
         x0 = i / len(OGs_k)
         width = panel_width / fig_width
         height = panel_height / fig_height
-        axes_rect = (x0 + margin_width, margin_height, width, height)
-        ax = fig.add_axes(axes_rect)
+        rect = (x0 + margin_width, margin_height, width, height)
+        ax = fig.add_axes(rect)
 
         nx.draw_networkx_edges(nx_graph, positions, ax=ax, edge_color=null_color, alpha=edge_alpha1, width=edge_width1)
         nx.draw_networkx_nodes(nx_graph, positions, ax=ax, node_size=node_size1, linewidths=0, node_color=node_colors)

### 2023-04-06 chapter2/chapter2.tex
Added substitution figure to chapter 2

@@
 \section{Results}
-Here's where my results will go!
+\begin{figure}[h!]
+\includegraphics[width=\textwidth]{substitution/out/substitution.png}
+\centering
+\caption{\textbf{Amino acid substitution models fit to disorder and order regions.}
+\textbf{(A)} Amino acid state frequencies of substitution models.
+Amino acid symbols are ordered by their enrichment in disorder regions, calculated as the disorder-to-order ratio of their frequencies.
+Error bars represent standard deviations over models fit to different meta-alignments ($n = 25$).
+\textbf{(B)} $\log_{10}$ disorder-to-order ratios of exchangeability coefficients.
+\textbf{(C)} $\log_{10}$ disorder-to-order ratios of rate coefficients.
+The vertical and horizontal axes indicate the initial and target amino acids, respectively.}
+\label{fig:substitution}
+\end{figure}
 
@@
 
-\subsection{Fitting substitution models}
+\subsection{Fitting amino substitution models}
 To fit amino acid substitution matrices to disorder and order regions, 25 meta-alignments were constructed by randomly sampling 100,000 columns from the respective subsets in the filtered regions derived from the 30 residue length cutoff.

### 2023-04-06 chapter2/figures/substitution/substitution.py
Added raw data panels to substitution figure

@@ -74,14 +74,14 @@ for label, record in records.items():
 plots = [(records['50R_disorder'], records['50R_order'], 'ematrix', 'exchangeability', 'B'),
          (records['50R_disorder'], records['50R_order'], 'rmatrix', 'rate', 'C')]
 
-fig = plt.figure(figsize=(7.5, 5))
-gs = plt.GridSpec(2, len(plots), height_ratios=[1, 1.5])
-rect = (0.1, 0.125, 0.85, 0.8)
+fig = plt.figure(figsize=(7.5, 6.25))
+gs = plt.GridSpec(1 + len(plots), 3)
+rect = (0.125, 0.1, 0.65, 0.8)
 
 width = 0.2
 bars = [records['50R_disorder'], records['50R_order']]
 subfig = fig.add_subfigure(gs[0, :])
-ax = subfig.add_axes(rect)
+ax = subfig.add_axes((0.1, 0.125, 0.85, 0.8))
 for i, record in enumerate(bars):
     freqs = record.freqs.mean(axis=0)
     std = record.freqs.std(axis=0)
@@ -94,20 +94,38 @@ ax.set_ylabel('Frequency')
 ax.legend()
 subfig.suptitle('A', x=0.0125, y=0.975, fontweight='bold')  # Half because panel is double length
 
+panel_labels = ['B', 'C', 'D', 'E', 'F', 'G']
 for gs_idx, plot in enumerate(plots):
     record1, record2, data_label, title_label, panel_label = plot
     matrix1 = getattr(record1, data_label).mean(axis=0)
     matrix2 = getattr(record2, data_label).mean(axis=0)
-    rs = np.log10(matrix1 / matrix2)
-    vext = np.nanmax(np.abs(rs))
+    vmax = max(matrix1.max(), matrix2.max())
+    ratio = np.log10(matrix1 / matrix2)
+    vext = np.nanmax(np.abs(ratio))
 
-    subfig = fig.add_subfigure(gs[1, gs_idx])
+    subfig = fig.add_subfigure(gs[gs_idx+1, 0])
     ax = subfig.add_axes(rect)
-    im = ax.imshow(rs, vmin=-vext, vmax=vext, cmap='RdBu')
+    im = ax.imshow(matrix1, vmin=0, vmax=vmax, cmap='Greys')
     ax.set_xticks(range(len(alphabet)), alphabet, fontsize=7)
[…]

### 2023-04-06 chapter2/figures/substitution/substitution.py
Centered y labels on heatmaps

@@ -76,7 +76,7 @@ plots = [(records['50R_disorder'], records['50R_order'], 'ematrix', 'exchangeabi
 
 fig = plt.figure(figsize=(7.5, 6.25))
 gs = plt.GridSpec(1 + len(plots), 3)
-rect = (0.125, 0.1, 0.65, 0.8)
+rect = (0.1, 0.1, 0.7, 0.8)
 
 width = 0.2
 bars = [records['50R_disorder'], records['50R_order']]
@@ -106,24 +106,24 @@ for gs_idx, plot in enumerate(plots):
     subfig = fig.add_subfigure(gs[gs_idx+1, 0])
     ax = subfig.add_axes(rect)
     im = ax.imshow(matrix1, vmin=0, vmax=vmax, cmap='Greys')
-    ax.set_xticks(range(len(alphabet)), alphabet, fontsize=7)
-    ax.set_yticks(range(len(alphabet)), alphabet, fontsize=7)
+    ax.set_xticks(range(len(alphabet)), alphabet, fontsize=6)
+    ax.set_yticks(range(len(alphabet)), alphabet, fontsize=6, ha='center')
     subfig.colorbar(im, cax=ax.inset_axes((1.05, 0, 0.05, 1)))
     subfig.suptitle(panel_labels[3*gs_idx], x=0.0375, y=0.975, fontweight='bold')  # 1.5x because panel is third length
 
     subfig = fig.add_subfigure(gs[gs_idx+1, 1])
     ax = subfig.add_axes(rect)
     im = ax.imshow(matrix2, vmin=0, vmax=vmax, cmap='Greys')
-    ax.set_xticks(range(len(alphabet)), alphabet, fontsize=7)
-    ax.set_yticks(range(len(alphabet)), alphabet, fontsize=7)
+    ax.set_xticks(range(len(alphabet)), alphabet, fontsize=6)
+    ax.set_yticks(range(len(alphabet)), alphabet, fontsize=6, ha='center')
     subfig.colorbar(im, cax=ax.inset_axes((1.05, 0, 0.05, 1)))
     subfig.suptitle(panel_labels[3*gs_idx+1], x=0.0375, y=0.975, fontweight='bold')  # 1.5x because panel is third length
 
     subfig = fig.add_subfigure(gs[1+gs_idx, 2])
     ax = subfig.add_axes(rect)
     im = ax.imshow(ratio, vmin=-vext, vmax=vext, cmap='RdBu')
[…]

### 2023-04-06 chapter2/chapter2.tex
Updated substitution figure caption

@@
 \caption{\textbf{Amino acid substitution models fit to disorder and order regions.}
-\textbf{(A)} Amino acid state frequencies of substitution models.
+\textbf{(A)} Amino acid frequencies of substitution models.
 Amino acid symbols are ordered by their enrichment in disorder regions, calculated as the disorder-to-order ratio of their frequencies.
 Error bars represent standard deviations over models fit to different meta-alignments ($n = 25$).
-\textbf{(B)} $\log_{10}$ disorder-to-order ratios of exchangeability coefficients.
-\textbf{(C)} $\log_{10}$ disorder-to-order ratios of rate coefficients.
-The vertical and horizontal axes indicate the initial and target amino acids, respectively.}
+\textbf{(B-C)} Exchangeability coefficients of disorder and order regions, respectively.
+\textbf{(D)} $\log_{10}$ disorder-to-order ratios of exchangeability coefficients.
+\textbf{(E-F)} Rate coefficients of disorder and order regions, respectively.
+The vertical and horizontal axes indicate the initial and target amino acids, respectively.
+\textbf{(G)} $\log_{10}$ disorder-to-order ratios of rate coefficients.}
 \label{fig:substitution}

### 2023-04-07 chapter1/chapter1_supporting.tex
Moved supplementary information to end as appendices
Fixed some extra spaces in chapter1_supporting.tex tables
Moved chapter1_supporting.tex table labels after caption
Changed created to fit in phylogenetic tree sampling figure

@@
-\section*{Supporting information}
-
 \begin{figure}[h!]
@@
 \centering
-\caption{\textbf{Phylogenetic trees created by different sampling strategies under LG model.}}
+\caption{\textbf{Phylogenetic trees fit by different sampling strategies under LG model.}}
 \label{sfig:trees_LG}
@@
 \centering
-\caption{\textbf{Phylogenetic trees created by different sampling strategies under GTR model.}}
+\caption{\textbf{Phylogenetic trees fit by different sampling strategies under GTR model.}}
 \label{sfig:trees_GTR}
 \end{figure}
-
-\setcounter{table}{0}
-\renewcommand{\thetable}{S\arabic{table}}
 
@@
 \caption{\textbf{Genome annotations.}}
+\label{stable:genomes}
 \begin{tabular}{|l|l|l|l|l|}
@@
 \end{tabular}
-\label{stable:genomes}
 \end{table}
@@
 \caption{\textbf{Phylogenetic diversity criteria.}}
+\label{stable:diversity_criteria}
 \begin{tabular}{|l|l|}
@@
 \end{tabular}
-\label{stable:diversity_criteria}
 \end{table}

### 2023-04-07 chapter2/chapter2.tex
Added note about averaging to substitution figure caption

@@
 Error bars represent standard deviations over models fit to different meta-alignments ($n = 25$).
-\textbf{(B-C)} Exchangeability coefficients of disorder and order regions, respectively.
+\textbf{(B-C)} Exchangeability coefficients of disorder and order regions, respectively, averaged over meta-alignments.
 \textbf{(D)} $\log_{10}$ disorder-to-order ratios of exchangeability coefficients.
-\textbf{(E-F)} Rate coefficients of disorder and order regions, respectively.
+\textbf{(E-F)} Rate coefficients of disorder and order regions, respectively, averaged over meta-alignments.
 The vertical and horizontal axes indicate the initial and target amino acids, respectively.

### 2023-04-07 chapter2/figures/substitution/substitution.py
Tweaked spacing in CV figure

@@ -178,7 +178,7 @@ plt.close()
 plots = [(records['50R_disorder'], records['50R_order'], 'ematrix'),
          (records['50R_disorder'], records['50R_order'], 'rmatrix')]
 for record1, record2, data_label in plots:
-    fig = plt.figure(figsize=(7.5, 5))
+    fig = plt.figure(figsize=(7.5, 3.5))
     gs = plt.GridSpec(2, 3)
     rect = (0.1, 0.1, 0.7, 0.8)
 
@@ -191,25 +191,25 @@ for record1, record2, data_label in plots:
         subfig = fig.add_subfigure(gs[gs_idx, 0])
         ax = subfig.add_axes(rect)
         im = ax.imshow(mean, cmap='Greys')
-        ax.set_xticks(range(len(alphabet)), alphabet, fontsize=7)
-        ax.set_yticks(range(len(alphabet)), alphabet, fontsize=7, ha='center')
+        ax.set_xticks(range(len(alphabet)), alphabet, fontsize=6)
+        ax.set_yticks(range(len(alphabet)), alphabet, fontsize=6, ha='center')
         fig.colorbar(im, cax=ax.inset_axes((1.05, 0, 0.05, 1)))
         subfig.suptitle(panel_labels[3*gs_idx], x=0.0375, y=0.975, fontweight='bold')
 
         subfig = fig.add_subfigure(gs[gs_idx, 1])
         ax = subfig.add_axes(rect)
         im = ax.imshow(std / mean, cmap='Greys')
-        ax.set_xticks(range(len(alphabet)), alphabet, fontsize=7)
-        ax.set_yticks(range(len(alphabet)), alphabet, fontsize=7, ha='center')
+        ax.set_xticks(range(len(alphabet)), alphabet, fontsize=6)
+        ax.set_yticks(range(len(alphabet)), alphabet, fontsize=6, ha='center')
         fig.colorbar(im, cax=ax.inset_axes((1.05, 0, 0.05, 1)))
         subfig.suptitle(panel_labels[3*gs_idx+1], x=0.0375, y=0.975, fontweight='bold')
 
         subfig = fig.add_subfigure(gs[gs_idx, 2])
-        ax = subfig.add_axes((0.275, 0.2, 0.675, 0.65))
+        ax = subfig.add_axes((0.275, 0.25, 0.675, 0.65))
         ax.scatter(mean, std / mean, s=10, alpha=0.5, edgecolor='none')
[…]

### 2023-04-07 chapter2/chapter2_supporting.tex
Added substitution figures to appendix

@@
-Here's where my chapter 2 appendix will go!
+\begin{figure}[h!]
+\includegraphics[width=\textwidth]{substitution/out/heatmap_ematrix.png}
+\centering
+\caption{\textbf{Exchangeability matrices fit under different sampling strategies.}
+Each panel is a mean of the exchangeability coefficients derived from the meta-alignments created by a single sampling strategy ($n = 25$).
+The prefix and suffix in the title of each panel indicate the maximum gap fraction and region type of the columns in the meta-alignments.
+For example, the columns in the ``50R\_disorder'' set of meta-alignments were fewer than 50\% gaps and sampled from the disorder subset.}
+\label{sfig:heatmap_ematrix}
+\end{figure}
+
+\begin{figure}[h!]
+\includegraphics[width=\textwidth]{substitution/out/heatmap_rmatrix.png}
+\centering
+\caption{\textbf{Rate matrices fit under different sampling strategies.}
+Each panel is a mean of the exchangeability coefficients derived from the meta-alignments created by a single sampling strategy ($n = 25$).
+See Fig.~\ref{sfig:CV_ematrix} for an explanation of the panel labels.}
+\label{sfig:heatmap_rmatrix}
+\end{figure}
+
+\begin{figure}[h!]
+\includegraphics[width=\textwidth]{substitution/out/heatmap_corr.png}
+\centering
+\caption{\textbf{Correlations between mean exchangeability and rate matrices fit under different sampling strategies.}
+\textbf{(A)} Correlations between the mean exchangeability matrices in Fig.~\ref{sfig:heatmap_ematrix}. \textbf{(B)} Correlations between the mean rate matrices in Fig.~\ref{sfig:heatmap_rmatrix}.}
+\label{sfig:heatmap_corr}
+\end{figure}
+
+\begin{figure}[h!]
+\includegraphics[width=\textwidth]{substitution/out/CV_ematrix.png}
+\centering
+\caption{\textbf{Coefficients of variation of the exchangeability matrices.}
+For all panels, the top and bottoms rows correspond to the 50R\_disorder and 50R\_order meta-alignment sets, respectively.
+\textbf{(A, D)} Mean exchangeability matrices.
+\textbf{(B, E)} Coefficients of variation (ratio of the standard deviation to the mean) of exchangeability matrices.
+\textbf{(C, F)} The coefficient of variation is inversely proportional to the mean, indicating the variation in the parameter estimates is constant relative to their magnitude.}
+\label{sfig:CV_ematrix}
+\end{figure}
+
+\begin{figure}[h!]
+\includegraphics[width=\textwidth]{substitution/out/CV_rmatrix.png}
+\centering
+\caption{\textbf{Coefficients of variation of the rate matrices.}
+For all panels, the top and bottoms rows correspond to the 50R\_disorder and 50R\_order meta-alignment sets, respectively.
+\textbf{(A, D)} Mean rate matrices.
+\textbf{(B, E)} Coefficients of variation (ratio of the standard deviation to the mean) of rate matrices.
+\textbf{(C, F)} The coefficient of variation is inversely proportional to the mean, indicating the variation in the parameter estimates is constant relative to their magnitude.}
+\label{sfig:CV_rmatrix}
+\end{figure}
 

### 2023-04-07 chapter2/figures/substitution/substitution.py
Added tiff outputs to substitution.py

@@ -127,6 +127,7 @@ for gs_idx, plot in enumerate(plots):
     subfig.colorbar(im, cax=ax.inset_axes((1.05, 0, 0.05, 1)))
     subfig.suptitle(panel_labels[3*gs_idx+2], x=0.0375, y=0.975, fontweight='bold')  # 1.5x because panel is third length
 fig.savefig('out/substitution.png', dpi=300)
+fig.savefig('out/substitution.tiff', dpi=300)
 plt.close()
 
 # HEATMAP
@@ -143,6 +144,7 @@ for data_label, title_label in plots:
         ax.set_title(record.label)
     fig.colorbar(ScalarMappable(Normalize(0, vmax), cmap='Greys'), ax=axs, fraction=0.025)
     fig.savefig(f'out/heatmap_{data_label}.png', dpi=300)
+    fig.savefig(f'out/heatmap_{data_label}.tiff', dpi=300)
     plt.close()
 
 # CORRELATION GRID
@@ -172,6 +174,7 @@ for gs_idx, plot in enumerate(plots):
     subfig.colorbar(im, cax=ax.inset_axes((1.05, 0, 0.05, 1)))
     subfig.suptitle(panel_label, x=0.025, y=0.975, fontweight='bold')
 fig.savefig(f'out/heatmap_corr.png', dpi=300)
+fig.savefig(f'out/heatmap_corr.tiff', dpi=300)
 plt.close()
 
 # VARIATION
@@ -212,4 +215,5 @@ for record1, record2, data_label in plots:
         subfig.suptitle(panel_labels[3*gs_idx+2], x=0.0375, y=0.975, fontweight='bold')
 
     fig.savefig(f'out/CV_{data_label}.png', dpi=300)
+    fig.savefig(f'out/CV_{data_label}.tiff', dpi=300)
     plt.close()

### 2023-04-07 chapter2/figures/substitution/substitution.py
Made dpi into global variable

@@ -11,6 +11,8 @@ from matplotlib.colors import Normalize
 from matplotlib.cm import ScalarMappable
 from src2.utils import read_iqtree, read_paml
 
+dpi = 300
+
 paml_order = ['A', 'R', 'N', 'D', 'C', 'Q', 'E', 'G', 'H', 'I', 'L', 'K', 'M', 'F', 'P', 'S', 'T', 'W', 'Y', 'V']
 labels = ['0R_disorder', '50R_disorder', '100R_disorder', '0R_order', '50R_order', '100R_order']
 labels_suffix = r'_[0-9]+'
@@ -126,8 +128,8 @@ for gs_idx, plot in enumerate(plots):
     ax.set_yticks(range(len(alphabet)), alphabet, fontsize=6, ha='center')
     subfig.colorbar(im, cax=ax.inset_axes((1.05, 0, 0.05, 1)))
     subfig.suptitle(panel_labels[3*gs_idx+2], x=0.0375, y=0.975, fontweight='bold')  # 1.5x because panel is third length
-fig.savefig('out/substitution.png', dpi=300)
-fig.savefig('out/substitution.tiff', dpi=300)
+fig.savefig('out/substitution.png', dpi=dpi)
+fig.savefig('out/substitution.tiff', dpi=dpi)
 plt.close()
 
 # HEATMAP
@@ -143,8 +145,8 @@ for data_label, title_label in plots:
         ax.set_yticks(range(len(alphabet)), alphabet, fontsize=7, ha='center')
         ax.set_title(record.label)
     fig.colorbar(ScalarMappable(Normalize(0, vmax), cmap='Greys'), ax=axs, fraction=0.025)
-    fig.savefig(f'out/heatmap_{data_label}.png', dpi=300)
-    fig.savefig(f'out/heatmap_{data_label}.tiff', dpi=300)
+    fig.savefig(f'out/heatmap_{data_label}.png', dpi=dpi)
+    fig.savefig(f'out/heatmap_{data_label}.tiff', dpi=dpi)
     plt.close()
 
 # CORRELATION GRID
@@ -173,8 +175,8 @@ for gs_idx, plot in enumerate(plots):
     ax.set_yticks(range(len(labels)), labels, fontsize=8)
     subfig.colorbar(im, cax=ax.inset_axes((1.05, 0, 0.05, 1)))
[…]

### 2023-04-09 chapter2/figures/scores/scores.py
Adjusted spacing and updated panel A to plot regions

@@ -27,18 +27,22 @@ ppid_regex = r'ppid=([A-Za-z0-9_.]+)'
 spid_regex = r'spid=([a-z]+)'
 min_length = 30
 
-panel_OGids = ['1DBA', '20C5', '27E5', '1432', '2554', '2805', '07E3', '0F47', '22B5']
+panel_records = [('0F47', 0, None),
+                 ('07E3', 350, 650),
+                 ('146F', 125, 525),
+                 ('1432', 9, 325)]
 plt.rcParams['xtick.labelsize'] = 8
 plt.rcParams['ytick.labelsize'] = 8
 plt.rcParams['axes.labelsize'] = 8
-width_ratios = [0.3, 0.3, 0.4]
-rect_B = (0.4, 0.08, 0.5375, 0.9)
-rect_CD = (0.3, 0.24, 0.65, 0.65)
+plt.rcParams['legend.fontsize'] = 8
+width_ratios = (0.3, 0.3, 0.4)
+rect_B = (0.375, 0.06, 0.5475, 0.92)
+rect_CD = (0.25, 0.18, 0.7, 0.7)
 rect_E = (0.5, 0.325, 0.475, 0.65)
-
-plot_msa_kwargs = {'hspace': 0.02, 'left': 0.2, 'right': 0.825, 'top': 0.925, 'bottom': 0.03, 'anchor': (0, 0.5),
+plot_msa_kwargs = {'hspace': 0.01, 'left': 0.15, 'right': 0.85, 'top': 0.95, 'bottom': 0.025, 'anchor': (0, 0.5),
                    'data_min': -0.05, 'data_max': 1.05,
-                   'msa_legend': True, 'legend_kwargs': {'bbox_to_anchor': (0.825, 0.5), 'loc': 'center left', 'fontsize': 8,
+                   'tree_position': 0, 'tree_width': 0.15,
+                   'msa_legend': True, 'legend_kwargs': {'bbox_to_anchor': (0.85, 0.5), 'loc': 'center left', 'fontsize': 8,
                                                          'handletextpad': 0.5, 'markerscale': 1.5, 'handlelength': 1}}
 
 tree_template = skbio.read('../../IDR_evolution/data/trees/consensus_LG/100R_NI.nwk', 'newick', skbio.TreeNode)
@@ -47,11 +51,11 @@ tip_order = {tip.name: i for i, tip in enumerate(tree_template.tips())}
[…]

### 2023-04-09 main.tex
Changed newpage command to clearpage to make consistent with other uses
Added text for feature definition and supplement tables
Cleaned up bibliography
Added chapter labels
Added substitution figure to chapter 2
Added "of" in "outside of work" in acknowledgements for clarity

@@
 \usepackage[margin=1in]{geometry}
+\usepackage{lscape}
 %\usepackage{setspace}  % double spacing
@@
 % Chapter and figure formatting
+\usepackage{longtable}
 \usepackage{titlesec}  % Modify chapter title format
@@
 % References
-\usepackage[doi=false,url=false]{biblatex}
+\usepackage[doi=false,url=false, style=nature]{biblatex}
 \addbibresource{refs.bib}
@@
 Thank you to my brothers, Andrew and Daniel, for giving me thick skin in the way only big brothers can.\\\\
-Thank you to all the friends I've made in California and the Bay Area, for the much-needed life outside work.\\\\
+Thank you to all the friends I've made in California and the Bay Area, for the much-needed life outside of work.\\\\
 Finally, thank you to my partner, Joseph, for going on this crazy ride with me. I couldn't have done this without you.
@@
 \tableofcontents
-\newpage  % End page so numbering is set properly
+\clearpage  % End page so numbering is set properly
 \setcounter{chapter}{-1}  % Set chapter counter to 0
@@
 \chapter{Introduction}
+\label{chapter:0}
 \graphicspath{{chapter0/figures/}}
@@
 \chapter{Leveraging genomic redundancy to improve inference and alignment of orthologous proteins}
+\label{chapter:1}
 \graphicspath{{chapter1/figures/}}
@@
 \chapter{Evolutionary analyses of IDRs reveal conserved features}
+\label{chapter:2}
+\graphicspath{{chapter2/figures/}}
 \input{chapter2/chapter2.tex}
@@
 \chapter{Tools and tutorials for fitting mixture models and HMMs}
+\label{chapter:3}
 \graphicspath{{chapter3/}}

### 2023-04-09 chapter2/chapter2.tex
Tweaked wording about normalizing rates to make less awkward
Added scores figure to chapter 2

@@
 \section{Results}
+\begin{figure}[h!]
+\includegraphics[width=\textwidth]{scores/out/scores_07E3.png}
+\centering
+\caption{\textbf{Analyses of disorder scores.}
+\textbf{(A)} Example alignment with the disorder scores of its sequences.
+Higher scores indicate a higher probability of disorder.
+Disorder score traces are colored by the position of its associated species on the phylogenetic tree.
+\textbf{(B)} Correlations between disorder scores and feature contrasts in regions.
+Asterisks indicate statistically significant correlations as computed by permutation tests ($p < 0.001$).
+\textbf{(C-D)} Example scatter plots showing correlations plotted in panel B.
+\textbf{(E)} GO term analysis of regions with rapidly evolving disorder scores.
+Only terms whose $p$-values are below 0.001 are shown.}
+\label{fig:scores}
+\end{figure}
+
 \begin{figure}[h!]
@@
 A GTR20 substitution model with four FreeRate categories and optimized state frequencies was fit to each meta-alignment using IQ-TREE 1.6.12~\cite{Nguyen2014}.
-Exchangeability and rate coefficients were normalized to make the average rate of each model equal to 1.
+Exchangeability and rate coefficients were normalized, so the average rate of each model was equal to 1.
 All figures are derived from the maximum 50\% gap fraction meta-alignment set unless otherwise noted.

### 2023-04-09 chapter2/chapter2.tex
Added references to some supplemental substitution figures

@@
 Exchangeability and rate coefficients were normalized, so the average rate of each model was equal to 1.
-All figures are derived from the maximum 50\% gap fraction meta-alignment set unless otherwise noted.
+Because exchangeability and rate coefficients are highly correlated across meta-alignments of the same region type, all figures are derived from the maximum 50\% gap fraction meta-alignment sets unless otherwise noted (Fig.~\ref{sfig:heatmap_ematrix}-\ref{sfig:heatmap_corr}).
 

### 2023-04-09 chapter2/chapter2_supporting.tex
Clarified wording around fitting matrices

@@
 \centering
-\caption{\textbf{Exchangeability matrices fit under different sampling strategies.}
-Each panel is a mean of the exchangeability coefficients derived from the meta-alignments created by a single sampling strategy ($n = 25$).
+\caption{\textbf{Exchangeability matrices fit to meta-alignments yielded by different sampling strategies.}
+Each panel is a mean of the exchangeability coefficients fit to the meta-alignments yielded by a single sampling strategy ($n = 25$).
 The prefix and suffix in the title of each panel indicate the maximum gap fraction and region type of the columns in the meta-alignments.
@@
 \centering
-\caption{\textbf{Rate matrices fit under different sampling strategies.}
-Each panel is a mean of the exchangeability coefficients derived from the meta-alignments created by a single sampling strategy ($n = 25$).
-See Fig.~\ref{sfig:CV_ematrix} for an explanation of the panel labels.}
+\caption{\textbf{Rate matrices fit to meta-alignments yielded by different sampling strategies.}
+Each panel is a mean of the exchangeability coefficients fit to the meta-alignments yielded by a single sampling strategy ($n = 25$).
+See Fig.~\ref{sfig:heatmap_ematrix} for an explanation of the panel labels.}
 \label{sfig:heatmap_rmatrix}
@@
 \centering
-\caption{\textbf{Correlations between mean exchangeability and rate matrices fit under different sampling strategies.}
+\caption{\textbf{Correlations between mean exchangeability and rate matrices fit to meta-alignments yielded by different sampling strategies.}
 \textbf{(A)} Correlations between the mean exchangeability matrices in Fig.~\ref{sfig:heatmap_ematrix}. \textbf{(B)} Correlations between the mean rate matrices in Fig.~\ref{sfig:heatmap_rmatrix}.}

### 2023-04-09 chapter1/figures/realignment/realignment.py
Increased legend fontsize

@@ -24,7 +24,7 @@ adjust_top = 0.925
 hspace = 0.3
 x_labelsize = 5
 legend_position = (0.86, 0.5)
-legend_fontsize = 5
+legend_fontsize = 6
 legend_handletextpad = 0.5
 legend_markerscale = 1
 

### 2023-04-09 chapter1/chapter1_supporting.tex
Clarified wording around fitting matrices
Adjusted spacing in clusters figure to make x-axis more readable

@@
 \textbf{(A)} Each species is equally represented in orthologous groups (OGs).
-\textbf{(B)} A plurality of orthologous groups contain all species.
-\textbf{(C)} Nearly all proteins are associated with a single orthologous group.
+\textbf{(B)} Nearly all proteins are associated with a single orthologous group.
+\textbf{(C)} A plurality of orthologous groups contain all species.
 \textbf{(D)} The number of orthologous groups associated with a species is strongly correlated with the number of unique annotated proteins, which suggests the annotation pipeline generally identifies conserved genes.}
@@
 \centering
-\caption{\textbf{Phylogenetic trees fit by different sampling strategies under LG model.}}
+\caption{\textbf{Phylogenetic trees fit to meta-alignments yielded by different sampling strategies under LG model.}}
 \label{sfig:trees_LG}

### 2023-04-09 chapter1/figures/clusters/clusters.py

@@ -43,40 +43,41 @@ if not os.path.exists('out/'):
     os.mkdir('out/')
 
 fig = plt.figure(figsize=(7.5, 4.5))
-gs = plt.GridSpec(2, 2)
-rect = (0.25, 0.2, 0.725, 0.7)
+gs = plt.GridSpec(2, 2, width_ratios=(0.55, 0.45))
+rectAC = (0.225, 0.2, 0.75, 0.7)
+rectBD = (0.25, 0.2, 0.725, 0.7)
 
 counts = OGs[['spid', 'OGid']].drop_duplicates()['spid'].value_counts().sort_index()
 xs = list(range(len(counts)))
 subfig = fig.add_subfigure(gs[0, 0], facecolor='none')
-ax = subfig.add_axes(rect)
-ax.set_xmargin(0.025)
-ax.bar(xs, counts.values)
-ax.set_xticks(xs, counts.index, rotation=60, fontsize=5.5)
+ax = subfig.add_axes(rectAC)
+ax.set_xmargin(0.01)
+ax.bar(xs, counts.values, width=0.6)
+ax.set_xticks(xs, counts.index, rotation=60, fontsize=6)
 ax.set_xlabel('Species')
 ax.set_ylabel('Number of associated OGs')
 subfig.suptitle('A', x=0.025, y=0.975, fontweight='bold')
 
-counts = groups['spid'].nunique().value_counts()
-subfig = fig.add_subfigure(gs[0, 1], facecolor='none')
-ax = subfig.add_axes(rect)
-ax.bar(counts.index, counts.values)
-ax.set_xlabel('Number of species in OG')
-ax.set_ylabel('Number of OGs')
-subfig.suptitle('B', x=0.025, y=0.975, fontweight='bold')
-
 counts = OGs.groupby('ppid')['OGid'].nunique().value_counts()
-subfig = fig.add_subfigure(gs[1, 0], facecolor='none')
-ax = subfig.add_axes(rect)
+subfig = fig.add_subfigure(gs[0, 1], facecolor='none')
+ax = subfig.add_axes(rectBD)
 ax.bar(counts.index, counts.values, width=1)
 ax.set_xlabel('Number of OGs associated with protein')
 ax.set_ylabel('Number of proteins')
+subfig.suptitle('B', x=0.025, y=0.975, fontweight='bold')
+
+counts = groups['spid'].nunique().value_counts()
[…]

### 2023-04-10 chapter2/figures/scores/scores.py
Added rate histogram with quantile highlighted

@@ -8,6 +8,7 @@ import matplotlib.pyplot as plt
 import numpy as np
 import pandas as pd
 import skbio
+from matplotlib.patches import Patch
 from src2.draw import plot_msa_data
 from src2.phylo import get_brownian_weights
 from src2.utils import read_fasta
@@ -44,6 +45,8 @@ plot_msa_kwargs = {'hspace': 0.01, 'left': 0.15, 'right': 0.85, 'top': 0.95, 'bo
                    'tree_position': 0, 'tree_width': 0.15,
                    'msa_legend': True, 'legend_kwargs': {'bbox_to_anchor': (0.85, 0.5), 'loc': 'center left', 'fontsize': 8,
                                                          'handletextpad': 0.5, 'markerscale': 1.5, 'handlelength': 1}}
+color3 = '#b07aa1'
+grey = '#e6e6e6'
 
 tree_template = skbio.read('../../IDR_evolution/data/trees/consensus_LG/100R_NI.nwk', 'newick', skbio.TreeNode)
 tip_order = {tip.name: i for i, tip in enumerate(tree_template.tips())}
@@ -51,6 +54,7 @@ tip_order = {tip.name: i for i, tip in enumerate(tree_template.tips())}
 if not os.path.exists('out/'):
     os.mkdir('out/')
 
+# === MAIN FIGURE ===
 fig = plt.figure(figsize=(7.5, 8.5))
 gs = plt.GridSpec(4, 3, width_ratios=width_ratios)
 dp = fig.dpi_scale_trans.transform((0.0625, -0.0625))
@@ -254,3 +258,32 @@ for panel_OGid, panel_start, panel_stop in panel_records:
     fig.savefig(f'out/scores_{panel_OGid}.png', dpi=400)
     fig.savefig(f'out/scores_{panel_OGid}.tiff', dpi=400)
     plt.close()
+
+# === RATE SUPPLEMENT ===
+# Load regions as segments
+rows = []
+with open(f'../../IDR_evolution/analysis/IDRpred/region_filter/out/regions_{min_length}.tsv') as file:
+    field_names = file.readline().rstrip('\n').split('\t')
+    for line in file:
+        fields = {key: value for key, value in zip(field_names, line.rstrip('\n').split('\t'))}
+        OGid, start, stop, disorder = fields['OGid'], int(fields['start']), int(fields['stop']), fields['disorder'] == 'True'
[…]

### 2023-04-10 chapter2/chapter2_supporting.tex
Added rate histogram to chapter 2 supplement

@@
+\begin{figure}[h!]
+\includegraphics[width=0.75\textwidth]{scores/out/scores_histogram.png}
+\centering
+\caption{\textbf{Histogram of disorder score rates in regions.}
+The grey interval indicates the upper decile of the distribution across both disorder and order regions, which was used as the input set for the GO term enrichment analysis.}
+\label{sfig:scores_histogram}
+\end{figure}
+
 \begin{figure}[h!]

### 2023-04-10 chapter2/figures/hierarchy/hierarchy.py
Added hierarchy.py

@@ -1,0 +1,157 @@
+"""Make figure of cluster heatmap."""
+
+import os
+
+import matplotlib.patches as mpatches
+import matplotlib.pyplot as plt
+import numpy as np
+import pandas as pd
+from scipy.cluster.hierarchy import linkage
+from scipy.spatial.distance import pdist
+from src2.brownian.linkage import make_tree
+from src2.draw import plot_tree
+
+pdidx = pd.IndexSlice
+min_length = 30
+
+min_indel_columns = 5  # Indel rates below this value are set to 0
+min_aa_rate = 0.5
+min_indel_rate = 0.1
+
+# Load regions
+rows = []
+with open(f'../../IDR_evolution/analysis/IDRpred/region_filter/out/regions_{min_length}.tsv') as file:
+    field_names = file.readline().rstrip('\n').split('\t')
+    for line in file:
+        fields = {key: value for key, value in zip(field_names, line.rstrip('\n').split('\t'))}
+        OGid, start, stop, disorder = fields['OGid'], int(fields['start']), int(fields['stop']), fields[
+            'disorder'] == 'True'
+        rows.append({'OGid': OGid, 'start': start, 'stop': stop, 'disorder': disorder})
+all_regions = pd.DataFrame(rows)
+
+asr_rates = pd.read_table(f'../../IDR_evolution/analysis/evofit/asr_stats/out/regions_{min_length}/rates.tsv')
+asr_rates = all_regions.merge(asr_rates, how='right', on=['OGid', 'start', 'stop'])
+row_idx = (asr_rates['indel_num_columns'] < min_indel_columns) | asr_rates['indel_rate_mean'].isna()
+asr_rates.loc[row_idx, 'indel_rate_mean'] = 0
+
+row_idx = (asr_rates['aa_rate_mean'] > min_aa_rate) | (asr_rates['indel_rate_mean'] > min_indel_rate)
+column_idx = ['OGid', 'start', 'stop', 'disorder']
+region_keys = asr_rates.loc[row_idx, column_idx]
+
+models = pd.read_table(f'../../IDR_evolution/analysis/brownian/get_models/out/models_{min_length}.tsv', header=[0, 1])
+df = region_keys.merge(models.droplevel(1, axis=1), how='left', on=['OGid', 'start', 'stop'])
+df = df.set_index(['OGid', 'start', 'stop', 'disorder'])
+
+feature_groups = {}
+feature_labels = []
[…]

### 2023-04-10 chapter2/figures/scores/scores.py
Updated path to GO analyses

@@ -231,7 +231,7 @@ for panel_OGid, panel_start, panel_stop in panel_records:
     subfig.suptitle('D', x=x, y=y, ha='left', fontweight='bold')
 
     # --- PANEL E ---
-    pvalues = pd.read_table('../../IDR_evolution/analysis/GOpred/score_enrich/out/pvalues_regions.tsv')
+    pvalues = pd.read_table('../../IDR_evolution/analysis/GO/score_enrich/out/pvalues_regions.tsv')
 
     subfig = fig.add_subfigure(gs[3, :])
     ax = subfig.add_axes(rect_E)

### 2023-04-10 chapter2/chapter2.tex
Added hierarchy figure

@@
 \label{fig:substitution}
+\end{figure}
+
+\begin{figure}[h!]
+\includegraphics[width=\textwidth]{hierarchy/out/hierarchy.png}
+\centering
+\caption{\textbf{Hierarchical clustering of evolutionary signatures.}
+The AIC difference between the Brownian motion (BM) and Ornstein-Uhlenbeck (OU) models is measure of their relative goodness of fit to the data with a penalty for the number of parameters in each.
+Larger values indicate a better fit by the OU model.}
+\label{fig:hierarchy}
 \end{figure}

### 2023-04-10 chapter2/figures/hierarchy/hierarchy.py
Made all spines invisible

@@ -109,8 +109,8 @@ plot_tree(tree, ax=ax, linecolor=node2color, linewidth=0.2, tip_labels=False,
 ax.set_ylabel('Disorder regions')
 ax.set_xticks([])
 ax.set_yticks([])
-ax.spines['left'].set_visible(False)
-ax.spines['bottom'].set_visible(False)
+for spine in ax.spines.values():
+    spine.set_visible(False)
 
 # Heatmap
 ax = axs[0, 1]
@@ -119,8 +119,8 @@ ax.xaxis.set_label_position('top')
 ax.set_xlabel('Features')
 ax.set_xticks([])
 ax.set_yticks([])
-ax.spines['left'].set_visible(False)
-ax.spines['bottom'].set_visible(False)
+for spine in ax.spines.values():
+    spine.set_visible(False)
 
 # Corner axis
 ax = axs[1, 0]

### 2023-04-11 chapter2/chapter2_supporting.tex
Replaced source with range in feature table

@@
 \hline
-\textbf{Feature ID}    & \textbf{Feature name}                                                            & \textbf{Group name}                                                & \textbf{Source} & \textbf{Description}                                                                                                             & \begin{tabular}[c]{@{}l@{}}\textbf{Changes from}\\\textbf{Zarin \textit{et al.}}~\cite{Zarin2019}\end{tabular}
+\textbf{Feature ID}    & \textbf{Feature name}                                                            & \textbf{Group name}                                                & \textbf{Range} & \textbf{Description}                                                                                                             & \begin{tabular}[c]{@{}l@{}}\textbf{Changes from}\\\textbf{Zarin \textit{et al.}}~\cite{Zarin2019}\end{tabular}
 \endfirsthead
@@
 \hline
-\textbf{Feature ID}    & \textbf{Feature name}                                                            & \textbf{Group name}                                                & \textbf{Source} & \textbf{Description}                                                                                                             & \begin{tabular}[c]{@{}l@{}}\textbf{Changes from}\\\textbf{Zarin \textit{et al.}~\cite{Zarin2019}}\end{tabular}
+\textbf{Feature ID}    & \textbf{Feature name}                                                            & \textbf{Group name}                                                & \textbf{Range} & \textbf{Description}                                                                                                             & \begin{tabular}[c]{@{}l@{}}\textbf{Changes from}\\\textbf{Zarin \textit{et al.}~\cite{Zarin2019}}\end{tabular}
 \endhead
@@
 \hline
-fraction\_S            & S fraction                                                                       & \begin{tabular}[c]{@{}l@{}}amino acid\\content\end{tabular}        &                 &                                                                                                                                  &                                                                                                \\
+fraction\_S            & S fraction                                                                       & \begin{tabular}[c]{@{}l@{}}amino acid\\content\end{tabular}        & {[}0, 1]                     &                                                                                                                                  &                                                                                                \\
 \hline
-fraction\_P            & P fraction                                                                       & \begin{tabular}[c]{@{}l@{}}amino acid\\content\end{tabular}        &                 &                                                                                                                                  &                                                                                                \\
+fraction\_P            & P fraction                                                                       & \begin{tabular}[c]{@{}l@{}}amino acid\\content\end{tabular}        & {[}0, 1]                     &                                                                                                                                  &                                                                                                \\
 \hline
-fraction\_T            & T fraction                                                                       & \begin{tabular}[c]{@{}l@{}}amino acid\\content\end{tabular}        &                 &                                                                                                                                  &                                                                                                \\
+fraction\_T            & T fraction                                                                       & \begin{tabular}[c]{@{}l@{}}amino acid\\content\end{tabular}        & {[}0, 1]                     &                                                                                                                                  &                                                                                                \\
 \hline
-fraction\_A            & A fraction                                                                       & \begin{tabular}[c]{@{}l@{}}amino acid\\content\end{tabular}        &                 &                                                                                                                                  &                                                                                                \\
+fraction\_A            & A fraction                                                                       & \begin{tabular}[c]{@{}l@{}}amino acid\\content\end{tabular}        & {[}0, 1]                     &                                                                                                                                  &                                                                                                \\
 \hline
-fraction\_H            & H fraction                                                                       & \begin{tabular}[c]{@{}l@{}}amino acid\\content\end{tabular}        &                 &                                                                                                                                  &                                                                                                \\
+fraction\_H            & H fraction                                                                       & \begin{tabular}[c]{@{}l@{}}amino acid\\content\end{tabular}        & {[}0, 1]                     &                                                                                                                                  &                                                                                                \\
 \hline
-fraction\_Q            & Q fraction                                                                       & \begin{tabular}[c]{@{}l@{}}amino acid\\content\end{tabular}        &                 &                                                                                                                                  &                                                                                                \\
+fraction\_Q            & Q fraction                                                                       & \begin{tabular}[c]{@{}l@{}}amino acid\\content\end{tabular}        & {[}0, 1]                     &                                                                                                                                  &                                                                                                \\
 \hline
-fraction\_N            & N fraction                                                                       & \begin{tabular}[c]{@{}l@{}}amino acid\\content\end{tabular}        &                 &                                                                                                                                  &                                                                                                \\
+fraction\_N            & N fraction                                                                       & \begin{tabular}[c]{@{}l@{}}amino acid\\content\end{tabular}        & {[}0, 1]                     &                                                                                                                                  &                                                                                                \\
 \hline
-fraction\_G            & G fraction                                                                       & \begin{tabular}[c]{@{}l@{}}amino acid\\content\end{tabular}        &                 &                                                                                                                                  &                                                                                                \\
+fraction\_G            & G fraction                                                                       & \begin{tabular}[c]{@{}l@{}}amino acid\\content\end{tabular}        & {[}0, 1]                     &                                                                                                                                  &                                                                                                \\
 \hline
-FCR                    & \begin{tabular}[c]{@{}l@{}}fraction charged\\residues\end{tabular}               & \begin{tabular}[c]{@{}l@{}}charge\\properties\end{tabular}         &                 & basic residue fraction + acidic residue fraction                                                                                 &                                                                                                \\
+FCR                    & \begin{tabular}[c]{@{}l@{}}fraction charged\\residues\end{tabular}               & \begin{tabular}[c]{@{}l@{}}charge\\properties\end{tabular}         & {[}0, 1]                     & basic residue fraction + acidic residue fraction                                                                                 &                                                                                                \\
 \hline
-NCPR                   & \begin{tabular}[c]{@{}l@{}}net charge\\per residue\end{tabular}                  & \begin{tabular}[c]{@{}l@{}}charge\\properties\end{tabular}         &                 & basic residue fraction - acidic residue fraction                                                                                 &                                                                                                \\
+NCPR                   & \begin{tabular}[c]{@{}l@{}}net charge\\per residue\end{tabular}                  & \begin{tabular}[c]{@{}l@{}}charge\\properties\end{tabular}         & {[}-1, 1]                    & basic residue fraction - acidic residue fraction                                                                                 &                                                                                                \\
 \hline
-net\_charge            & net charge                                                                       & \begin{tabular}[c]{@{}l@{}}charge\\properties\end{tabular}         &                 & \#{[}RK] - \#{[}DE]                                                                                                              &                                                                                                \\
+net\_charge            & net charge                                                                       & \begin{tabular}[c]{@{}l@{}}charge\\properties\end{tabular}         & ($-\infty$, $\infty$)        & \#{[}RK] - \#{[}DE]                                                                                                              &                                                                                                \\
 \hline
-net\_charge\_P         & \begin{tabular}[c]{@{}l@{}}net charge with\\phosphorylation\end{tabular}         & \begin{tabular}[c]{@{}l@{}}charge\\properties\end{tabular}         &                 & \begin{tabular}[c]{@{}l@{}}net charge including phosphorylation of [ST]P\\consensus sites with -1.5 charge per site\end{tabular} &                                                                                                \\
+net\_charge\_P         & \begin{tabular}[c]{@{}l@{}}net charge with\\phosphorylation\end{tabular}         & \begin{tabular}[c]{@{}l@{}}charge\\properties\end{tabular}         & ($-\infty$, $\infty$)        & \begin{tabular}[c]{@{}l@{}}net charge including phosphorylation of [ST]P\\consensus sites with -1.5 charge per site\end{tabular} &                                                                                                \\
 \hline
-kappa                  & kappa                                                                            & \begin{tabular}[c]{@{}l@{}}charge\\properties\end{tabular}         & localcider      & \begin{tabular}[c]{@{}l@{}}measure of separation between positively\\and negatively charged residues\end{tabular}                &                                                                                                \\
+kappa                  & kappa                                                                            & \begin{tabular}[c]{@{}l@{}}charge\\properties\end{tabular}         & (0, 1]                       & \begin{tabular}[c]{@{}l@{}}measure of separation between positively\\and negatively charged residues\end{tabular}                &                                                                                                \\
 \hline
-omega                  & omega                                                                            & \begin{tabular}[c]{@{}l@{}}charge\\properties\end{tabular}         & localcider      & \begin{tabular}[c]{@{}l@{}}measure of separation between charged\\residues or prolines and all other residues\end{tabular}       &                                                                                                \\
+omega                  & omega                                                                            & \begin{tabular}[c]{@{}l@{}}charge\\properties\end{tabular}         & (0, 1]                       & \begin{tabular}[c]{@{}l@{}}measure of separation between charged\\residues or prolines and all other residues\end{tabular}       &                                                                                                \\
 \hline
-SCD                    & \begin{tabular}[c]{@{}l@{}}sequence charge\\decoration\end{tabular}              & \begin{tabular}[c]{@{}l@{}}charge\\properties\end{tabular}         & localcider      & \begin{tabular}[c]{@{}l@{}}measure of separation between positively\\and negatively charged residues\end{tabular}                &                                                                                                \\
+SCD                    & \begin{tabular}[c]{@{}l@{}}sequence charge\\decoration\end{tabular}              & \begin{tabular}[c]{@{}l@{}}charge\\properties\end{tabular}         & ($-\infty$, $\infty$)        & \begin{tabular}[c]{@{}l@{}}measure of separation between positively\\and negatively charged residues\end{tabular}                &                                                                                                \\
 \hline
-RK\_ratio              & R/K ratio                                                                        & \begin{tabular}[c]{@{}l@{}}charge\\properties\end{tabular}         &                 & \begin{tabular}[c]{@{}l@{}}adjusted ratio of arginine to\\lysine residues: (\#R + 1)/(\#K + 1)\end{tabular}                      &                                                                                                \\
+RK\_ratio              & R/K ratio                                                                        & \begin{tabular}[c]{@{}l@{}}charge\\properties\end{tabular}         & (0, $\infty$)                & \begin{tabular}[c]{@{}l@{}}adjusted ratio of arginine to\\lysine residues: (\#R + 1)/(\#K + 1)\end{tabular}                      &                                                                                                \\
 \hline
-ED\_ratio              & E/D ratio                                                                        & \begin{tabular}[c]{@{}l@{}}charge\\properties\end{tabular}         &                 & \begin{tabular}[c]{@{}l@{}}adjusted ratio of glutamic acid to\\aspartic acid residues: (\#E + 1)/(\#D + 1)\end{tabular}          &                                                                                                \\
+ED\_ratio              & E/D ratio                                                                        & \begin{tabular}[c]{@{}l@{}}charge\\properties\end{tabular}         & (0, $\infty$)                & \begin{tabular}[c]{@{}l@{}}adjusted ratio of glutamic acid to\\aspartic acid residues: (\#E + 1)/(\#D + 1)\end{tabular}          &                                                                                                \\
 \hline
-fraction\_acidic       & \begin{tabular}[c]{@{}l@{}}acidic\\residue fraction~\end{tabular}                & \begin{tabular}[c]{@{}l@{}}physiochemical\\properties\end{tabular} &                 &                                                                                                                                  &                                                                                                \\
+fraction\_acidic       & \begin{tabular}[c]{@{}l@{}}acidic\\residue fraction~\end{tabular}                & \begin{tabular}[c]{@{}l@{}}physiochemical\\properties\end{tabular} & {[}0 ,1]                     &                                                                                                                                  &                                                                                                \\
 \hline
-fraction\_basic        & \begin{tabular}[c]{@{}l@{}}basic\\residue fraction\end{tabular}                  & \begin{tabular}[c]{@{}l@{}}physiochemical\\properties\end{tabular} &                 &                                                                                                                                  &                                                                                                \\
+fraction\_basic        & \begin{tabular}[c]{@{}l@{}}basic\\residue fraction\end{tabular}                  & \begin{tabular}[c]{@{}l@{}}physiochemical\\properties\end{tabular} & {[}0 ,1]                     &                                                                                                                                  &                                                                                                \\
 \hline
-fraction\_aliphatic    & \begin{tabular}[c]{@{}l@{}}aliphatic\\residue fraction\end{tabular}              & \begin{tabular}[c]{@{}l@{}}physiochemical\\properties\end{tabular} &                 &                                                                                                                                  &                                                                                                \\
+fraction\_aliphatic    & \begin{tabular}[c]{@{}l@{}}aliphatic\\residue fraction\end{tabular}              & \begin{tabular}[c]{@{}l@{}}physiochemical\\properties\end{tabular} & {[}0 ,1]                     &                                                                                                                                  &                                                                                                \\
 \hline
-fraction\_polar        & \begin{tabular}[c]{@{}l@{}}polar\\residue fraction\end{tabular}                  & \begin{tabular}[c]{@{}l@{}}physiochemical\\properties\end{tabular} &                 &                                                                                                                                  & removed glycine                                                                                \\
+fraction\_polar        & \begin{tabular}[c]{@{}l@{}}polar\\residue fraction\end{tabular}                  & \begin{tabular}[c]{@{}l@{}}physiochemical\\properties\end{tabular} & {[}0 ,1]                     &                                                                                                                                  & removed glycine                                                                                \\
 \hline
-fraction\_chainexp     & \begin{tabular}[c]{@{}l@{}}chain-expanding\\residue fraction\end{tabular}        & \begin{tabular}[c]{@{}l@{}}physiochemical\\properties\end{tabular} &                 &                                                                                                                                  &                                                                                                \\
+fraction\_chainexp     & \begin{tabular}[c]{@{}l@{}}chain-expanding\\residue fraction\end{tabular}        & \begin{tabular}[c]{@{}l@{}}physiochemical\\properties\end{tabular} & {[}0 ,1]                     &                                                                                                                                  &                                                                                                \\
 \hline
-fraction\_aromatic     & \begin{tabular}[c]{@{}l@{}}aromatic\\residue fraction\end{tabular}               & \begin{tabular}[c]{@{}l@{}}physiochemical\\properties\end{tabular} &                 &                                                                                                                                  &                                                                                                \\
+fraction\_aromatic     & \begin{tabular}[c]{@{}l@{}}aromatic\\residue fraction\end{tabular}               & \begin{tabular}[c]{@{}l@{}}physiochemical\\properties\end{tabular} & {[}0 ,1]                     &                                                                                                                                  &                                                                                                \\
 \hline
-fraction\_disorder     & \begin{tabular}[c]{@{}l@{}}disorder-promoting\\residue fraction\end{tabular}     & \begin{tabular}[c]{@{}l@{}}physiochemical\\properties\end{tabular} &                 &                                                                                                                                  &                                                                                                \\
+fraction\_disorder     & \begin{tabular}[c]{@{}l@{}}disorder-promoting\\residue fraction\end{tabular}     & \begin{tabular}[c]{@{}l@{}}physiochemical\\properties\end{tabular} & {[}0 ,1]                     &                                                                                                                                  &                                                                                                \\
 \hline
-radius\_gyration       & radius of gyration                                                               & \begin{tabular}[c]{@{}l@{}}physiochemical\\properties\end{tabular} &                 & number of residues to the 0.6 power                                                                                              & \begin{tabular}[c]{@{}l@{}}substituted\\for length\end{tabular}                                \\
+radius\_gyration       & radius of gyration                                                               & \begin{tabular}[c]{@{}l@{}}physiochemical\\properties\end{tabular} & {[}0,~ $\infty$)             & number of residues to the 0.6 power                                                                                              & \begin{tabular}[c]{@{}l@{}}substituted\\for length\end{tabular}                                \\
 \hline
-hydropathy             & hydropathy                                                                       & \begin{tabular}[c]{@{}l@{}}physiochemical\\properties\end{tabular} & localcider      & normalized Kyte-Doolittle scale                                                                                                  &                                                                                                \\
+hydropathy             & hydropathy                                                                       & \begin{tabular}[c]{@{}l@{}}physiochemical\\properties\end{tabular} & {[}0 ,1]                     & normalized Kyte-Doolittle scale                                                                                                  &                                                                                                \\
 \hline
-isopoint               & isoelectric point                                                                & \begin{tabular}[c]{@{}l@{}}physiochemical\\properties\end{tabular} & isoelectric     & pH where charge of peptide is neutral                                                                                            &                                                                                                \\
+isopoint               & isoelectric point                                                                & \begin{tabular}[c]{@{}l@{}}physiochemical\\properties\end{tabular} & ($-\infty$, $\infty$)        & pH where charge of peptide is neutral                                                                                            &                                                                                                \\
 \hline
-PPII\_propensity       & PPII propensity                                                                  & \begin{tabular}[c]{@{}l@{}}physiochemical\\properties\end{tabular} & localcider      & \begin{tabular}[c]{@{}l@{}}propensity for proline to form\\left-handed helices\end{tabular}                                      &                                                                                                \\
+PPII\_propensity       & PPII propensity                                                                  & \begin{tabular}[c]{@{}l@{}}physiochemical\\properties\end{tabular} & {[}0, 1]                     & \begin{tabular}[c]{@{}l@{}}propensity for proline to form\\left-handed helices\end{tabular}                                      &                                                                                                \\
 \hline
-repeat\_Q              & Q repeat fraction                                                                & \begin{tabular}[c]{@{}l@{}}repeats and\\complexity\end{tabular}    &                 & fraction 2 or more consecutive Q                                                                                                 &                                                                                                \\
+repeat\_Q              & Q repeat fraction                                                                & \begin{tabular}[c]{@{}l@{}}repeats and\\complexity\end{tabular}    & {[}0, 1]                     & fraction 2 or more consecutive Q                                                                                                 &                                                                                                \\
 \hline
-repeat\_N              & N repeat fraction                                                                & \begin{tabular}[c]{@{}l@{}}repeats and\\complexity\end{tabular}    &                 & fraction 2 or more consecutive N                                                                                                 &                                                                                                \\
+repeat\_N              & N repeat fraction                                                                & \begin{tabular}[c]{@{}l@{}}repeats and\\complexity\end{tabular}    & {[}0, 1]                     & fraction 2 or more consecutive N                                                                                                 &                                                                                                \\
 \hline
-repeat\_S              & S repeat fraction                                                                & \begin{tabular}[c]{@{}l@{}}repeats and\\complexity\end{tabular}    &                 & fraction 2 or more consecutive S                                                                                                 &                                                                                                \\
+repeat\_S              & S repeat fraction                                                                & \begin{tabular}[c]{@{}l@{}}repeats and\\complexity\end{tabular}    & {[}0, 1]                     & fraction 2 or more consecutive S                                                                                                 &                                                                                                \\
 \hline
-repeat\_G              & G repeat fraction                                                                & \begin{tabular}[c]{@{}l@{}}repeats and\\complexity\end{tabular}    &                 & fraction 2 or more consectuive G                                                                                                 &                                                                                                \\
+repeat\_G              & G repeat fraction                                                                & \begin{tabular}[c]{@{}l@{}}repeats and\\complexity\end{tabular}    & {[}0, 1]                     & fraction 2 or more consectuive G                                                                                                 &                                                                                                \\
 \hline
-repeat\_E              & E repeat fraction                                                                & \begin{tabular}[c]{@{}l@{}}repeats and\\complexity\end{tabular}    &                 & fraction 2 or more consecutive E                                                                                                 &                                                                                                \\
+repeat\_E              & E repeat fraction                                                                & \begin{tabular}[c]{@{}l@{}}repeats and\\complexity\end{tabular}    & {[}0, 1]                     & fraction 2 or more consecutive E                                                                                                 &                                                                                                \\
 \hline
-repeat\_D              & D repeat fraction                                                                & \begin{tabular}[c]{@{}l@{}}repeats and\\complexity\end{tabular}    &                 & fraction 2 or more consecutive D                                                                                                 &                                                                                                \\
+repeat\_D              & D repeat fraction                                                                & \begin{tabular}[c]{@{}l@{}}repeats and\\complexity\end{tabular}    & {[}0, 1]                     & fraction 2 or more consecutive D                                                                                                 &                                                                                                \\
 \hline
-repeat\_K              & K repeat fraction                                                                & \begin{tabular}[c]{@{}l@{}}repeats and\\complexity\end{tabular}    &                 & fraction 2 or more consecutive K                                                                                                 &                                                                                                \\
+repeat\_K              & K repeat fraction                                                                & \begin{tabular}[c]{@{}l@{}}repeats and\\complexity\end{tabular}    & {[}0, 1]                     & fraction 2 or more consecutive K                                                                                                 &                                                                                                \\
 \hline
-repeat\_R              & R repeat fraction                                                                & \begin{tabular}[c]{@{}l@{}}repeats and\\complexity\end{tabular}    &                 & fraction 2 or more consecutive R                                                                                                 &                                                                                                \\
+repeat\_R              & R repeat fraction                                                                & \begin{tabular}[c]{@{}l@{}}repeats and\\complexity\end{tabular}    & {[}0, 1]                     & fraction 2 or more consecutive R                                                                                                 &                                                                                                \\
 \hline
-repeat\_P              & P repeat fraction                                                                & \begin{tabular}[c]{@{}l@{}}repeats and\\complexity\end{tabular}    &                 & fraction 2 or more consecutive P                                                                                                 &                                                                                                \\
+repeat\_P              & P repeat fraction                                                                & \begin{tabular}[c]{@{}l@{}}repeats and\\complexity\end{tabular}    & {[}0, 1]                     & fraction 2 or more consecutive P                                                                                                 &                                                                                                \\
 \hline
-repeat\_QN             & {[}QN] repeat fraction                                                           & \begin{tabular}[c]{@{}l@{}}repeats and\\complexity\end{tabular}    &                 & fraction 2 or more consecutive [QN]                                                                                              &                                                                                                \\
+repeat\_QN             & {[}QN] repeat fraction                                                           & \begin{tabular}[c]{@{}l@{}}repeats and\\complexity\end{tabular}    & {[}0, 1]                     & fraction 2 or more consecutive [QN]                                                                                              &                                                                                                \\
 \hline
-repeat\_RG             & {[}RG] repeat fraction                                                           & \begin{tabular}[c]{@{}l@{}}repeats and\\complexity\end{tabular}    &                 & fraction 2 or more consecutive [RG]                                                                                              &                                                                                                \\
+repeat\_RG             & {[}RG] repeat fraction                                                           & \begin{tabular}[c]{@{}l@{}}repeats and\\complexity\end{tabular}    & {[}0, 1]                     & fraction 2 or more consecutive [RG]                                                                                              &                                                                                                \\
 \hline
-repeat\_FG             & {[}FG] repeat fraction                                                           & \begin{tabular}[c]{@{}l@{}}repeats and\\complexity\end{tabular}    &                 & fraction 2 or more consecutive [FG]                                                                                              &                                                                                                \\
+repeat\_FG             & {[}FG] repeat fraction                                                           & \begin{tabular}[c]{@{}l@{}}repeats and\\complexity\end{tabular}    & {[}0, 1]                     & fraction 2 or more consecutive [FG]                                                                                              &                                                                                                \\
 \hline
-repeat\_SG             & {[}SG] repeat fraction                                                           & \begin{tabular}[c]{@{}l@{}}repeats and\\complexity\end{tabular}    &                 & fraction 2 or more consecutive [SG]                                                                                              &                                                                                                \\
+repeat\_SG             & {[}SG] repeat fraction                                                           & \begin{tabular}[c]{@{}l@{}}repeats and\\complexity\end{tabular}    & {[}0, 1]                     & fraction 2 or more consecutive [SG]                                                                                              &                                                                                                \\
 \hline
-repeat\_SR             & {[}SR] repeat fraction                                                           & \begin{tabular}[c]{@{}l@{}}repeats and\\complexity\end{tabular}    &                 & fraction 2 or more consecutive [SR]                                                                                              &                                                                                                \\
+repeat\_SR             & {[}SR] repeat fraction                                                           & \begin{tabular}[c]{@{}l@{}}repeats and\\complexity\end{tabular}    & {[}0, 1]                     & fraction 2 or more consecutive [SR]                                                                                              &                                                                                                \\
 \hline
-repeat\_AP             & {[}KAP] repeat fraction                                                          & \begin{tabular}[c]{@{}l@{}}repeats and\\complexity\end{tabular}    &                 & fraction 2 or more consecutive [KAP]                                                                                             &                                                                                                \\
+repeat\_AP             & {[}KAP] repeat fraction                                                          & \begin{tabular}[c]{@{}l@{}}repeats and\\complexity\end{tabular}    & {[}0, 1]                     & fraction 2 or more consecutive [KAP]                                                                                             &                                                                                                \\
 \hline
-repeat\_TS             & {[}PTS] repeat fraction                                                          & \begin{tabular}[c]{@{}l@{}}repeats and\\complexity\end{tabular}    &                 & fraction 2 or more consecutive [PTS]                                                                                             &                                                                                                \\
+repeat\_TS             & {[}PTS] repeat fraction                                                          & \begin{tabular}[c]{@{}l@{}}repeats and\\complexity\end{tabular}    & {[}0, 1]                     & fraction 2 or more consecutive [PTS]                                                                                             &                                                                                                \\
 \hline
-wf\_complexity         & \begin{tabular}[c]{@{}l@{}}Wootton-Federhen\\sequence complexity\end{tabular}    & \begin{tabular}[c]{@{}l@{}}repeats and\\complexity\end{tabular}    & localcider      & \begin{tabular}[c]{@{}l@{}}complexity based on SEG algorithm:\\blob length=IDR length, step size=1\end{tabular}                  &                                                                                                \\
+wf\_complexity         & \begin{tabular}[c]{@{}l@{}}Wootton-Federhen\\sequence complexity\end{tabular}    & \begin{tabular}[c]{@{}l@{}}repeats and\\complexity\end{tabular}    & {[}0, 1]                     & \begin{tabular}[c]{@{}l@{}}complexity based on SEG algorithm:\\blob length=IDR length, step size=1\end{tabular}                  &                                                                                                \\
 \hline
-CLV\_Separin\_Metazoa  & separase cleavage motif                                                          & motifs                                                             &                 &                                                                                                                                  & \begin{tabular}[c]{@{}l@{}}Metazoa motif\\instead of fungi\end{tabular}                        \\
+CLV\_Separin\_Metazoa  & separase cleavage motif                                                          & motifs                                                             & {[}0, $\infty$)              &                                                                                                                                  & \begin{tabular}[c]{@{}l@{}}Metazoa motif\\instead of fungi\end{tabular}                        \\
 \hline
-DEG\_APCC\_KENBOX\_2   & \begin{tabular}[c]{@{}l@{}}APCC-binding\\destruction motif\end{tabular}          & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
+DEG\_APCC\_KENBOX\_2   & \begin{tabular}[c]{@{}l@{}}APCC-binding\\destruction motif\end{tabular}          & motifs                                                             & {[}0, $\infty$)              &                                                                                                                                  &                                                                                                \\
 \hline
-DEG\_APCC\_TPR\_1      & APCC-TPR-docking motif                                                           & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
+DEG\_APCC\_TPR\_1      & APCC-TPR-docking motif                                                           & motifs                                                             & {[}0, $\infty$)              &                                                                                                                                  &                                                                                                \\
 \hline
-DOC\_CKS1\_1           & Cks1 ligand                                                                      & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
+DOC\_CKS1\_1           & Cks1 ligand                                                                      & motifs                                                             & {[}0, $\infty$)              &                                                                                                                                  &                                                                                                \\
 \hline
-DOC\_MAPK\_DCC\_7      & MAPK docking motif                                                               & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
+DOC\_MAPK\_DCC\_7      & MAPK docking motif                                                               & motifs                                                             & {[}0, $\infty$)              &                                                                                                                                  &                                                                                                \\
 \hline
-DOC\_MAPK\_gen\_1      & MAPK docking motif                                                               & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
+DOC\_MAPK\_gen\_1      & MAPK docking motif                                                               & motifs                                                             & {[}0, $\infty$)              &                                                                                                                                  &                                                                                                \\
 \hline
-DOC\_MAPK\_HePTP\_8    & MAPK docking motif                                                               & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
+DOC\_MAPK\_HePTP\_8    & MAPK docking motif                                                               & motifs                                                             & {[}0, $\infty$)              &                                                                                                                                  &                                                                                                \\
 \hline
-DOC\_PP1\_RVXF\_1      & PP1-docking motif RVXF                                                           & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
+DOC\_PP1\_RVXF\_1      & PP1-docking motif RVXF                                                           & motifs                                                             & {[}0, $\infty$)              &                                                                                                                                  &                                                                                                \\
 \hline
-DOC\_PP2B\_PxIxI\_1    & \begin{tabular}[c]{@{}l@{}}calcineurin (PP2B)-\\docking motif PxIxI\end{tabular} & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
+DOC\_PP2B\_PxIxI\_1    & \begin{tabular}[c]{@{}l@{}}calcineurin (PP2B)-\\docking motif PxIxI\end{tabular} & motifs                                                             & {[}0, $\infty$)              &                                                                                                                                  &                                                                                                \\
 \hline
-LIG\_APCC\_Cbox\_1     & \begin{tabular}[c]{@{}l@{}}APC/C\_Apc2-docking\\motif\end{tabular}               & motifs                                                             &                 &                                                                                                                                  & \begin{tabular}[c]{@{}l@{}}Metazoa motif\\instead of fungi\end{tabular}                        \\
+LIG\_APCC\_Cbox\_1     & \begin{tabular}[c]{@{}l@{}}APC/C\_Apc2-docking\\motif\end{tabular}               & motifs                                                             & {[}0, $\infty$)              &                                                                                                                                  & \begin{tabular}[c]{@{}l@{}}Metazoa motif\\instead of fungi\end{tabular}                        \\
 \hline
-LIG\_AP\_GAE\_1        & \begin{tabular}[c]{@{}l@{}}gamma-adaptin ear\\interaction motif\end{tabular}     & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
+LIG\_AP\_GAE\_1        & \begin{tabular}[c]{@{}l@{}}gamma-adaptin ear\\interaction motif\end{tabular}     & motifs                                                             & {[}0, $\infty$)              &                                                                                                                                  &                                                                                                \\
 \hline
-LIG\_CaM\_IQ\_9        & \begin{tabular}[c]{@{}l@{}}helical calmodulin\\binding motif\end{tabular}        & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
+LIG\_CaM\_IQ\_9        & \begin{tabular}[c]{@{}l@{}}helical calmodulin\\binding motif\end{tabular}        & motifs                                                             & {[}0, $\infty$)              &                                                                                                                                  &                                                                                                \\
 \hline
-LIG\_EH\_1             & EH ligand                                                                        & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
+LIG\_EH\_1             & EH ligand                                                                        & motifs                                                             & {[}0, $\infty$)              &                                                                                                                                  &                                                                                                \\
 \hline
-LIG\_eIF4E\_1          & eIF4E binding motif                                                              & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
+LIG\_eIF4E\_1          & eIF4E binding motif                                                              & motifs                                                             & {[}0, $\infty$)              &                                                                                                                                  &                                                                                                \\
 \hline
-LIG\_GLEBS\_BUB3\_1    & GLEBS motif                                                                      & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
+LIG\_GLEBS\_BUB3\_1    & GLEBS motif                                                                      & motifs                                                             & {[}0, $\infty$)              &                                                                                                                                  &                                                                                                \\
 \hline
-LIG\_LIR\_Gen\_1       & \begin{tabular}[c]{@{}l@{}}Atg8 protein\\family ligands\end{tabular}             & motifs                                                             &                 &                                                                                                                                  & \begin{tabular}[c]{@{}l@{}}same ELM entry,\\but updated regex\end{tabular}                     \\
+LIG\_LIR\_Gen\_1       & \begin{tabular}[c]{@{}l@{}}Atg8 protein\\family ligands\end{tabular}             & motifs                                                             & {[}0, $\infty$)              &                                                                                                                                  & \begin{tabular}[c]{@{}l@{}}same ELM entry,\\but updated regex\end{tabular}                     \\
 \hline
-LIG\_PCNA\_PIPBox\_1   & PCNA binding PIP box                                                             & motifs                                                             &                 &                                                                                                                                  & \begin{tabular}[c]{@{}l@{}}same ELM entry,\\but updated regex\end{tabular}                     \\
+LIG\_PCNA\_PIPBox\_1   & PCNA binding PIP box                                                             & motifs                                                             & {[}0, $\infty$)              &                                                                                                                                  & \begin{tabular}[c]{@{}l@{}}same ELM entry,\\but updated regex\end{tabular}                     \\
 \hline
-LIG\_SUMO\_SIM\_par\_1 & \begin{tabular}[c]{@{}l@{}}SUMO interaction\\site\end{tabular}                   & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
+LIG\_SUMO\_SIM\_par\_1 & \begin{tabular}[c]{@{}l@{}}SUMO interaction\\site\end{tabular}                   & motifs                                                             & {[}0, $\infty$)              &                                                                                                                                  &                                                                                                \\
 \hline
-MOD\_CDK\_SPxK\_1      & \begin{tabular}[c]{@{}l@{}}CDK phosphorylation\\site\end{tabular}                & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
+MOD\_CDK\_SPxK\_1      & \begin{tabular}[c]{@{}l@{}}CDK phosphorylation\\site\end{tabular}                & motifs                                                             & {[}0, $\infty$)              &                                                                                                                                  &                                                                                                \\
 \hline
-MOD\_LATS\_1           & \begin{tabular}[c]{@{}l@{}}LATS kinase\\phosphorylation motif\end{tabular}       & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
+MOD\_LATS\_1           & \begin{tabular}[c]{@{}l@{}}LATS kinase\\phosphorylation motif\end{tabular}       & motifs                                                             & {[}0, $\infty$)              &                                                                                                                                  &                                                                                                \\
 \hline
-MOD\_SUMO\_for\_1      & sumoylation site                                                                 & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
+MOD\_SUMO\_for\_1      & sumoylation site                                                                 & motifs                                                             & {[}0, $\infty$)              &                                                                                                                                  &                                                                                                \\
 \hline
-TRG\_ER\_FFAT\_1       & FFAT motif                                                                       & motifs                                                             &                 &                                                                                                                                  & \begin{tabular}[c]{@{}l@{}}same ELM entry,\\but updated regex\end{tabular}                     \\
+TRG\_ER\_FFAT\_1       & FFAT motif                                                                       & motifs                                                             & {[}0, $\infty$)              &                                                                                                                                  & \begin{tabular}[c]{@{}l@{}}same ELM entry,\\but updated regex\end{tabular}                     \\
 \hline
-TRG\_Golgi\_diPhe\_1   & ER export signals                                                                & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
+TRG\_Golgi\_diPhe\_1   & ER export signals                                                                & motifs                                                             & {[}0, $\infty$)              &                                                                                                                                  &                                                                                                \\
 \hline
-TRG\_NLS\_MonoExtN\_4  & \begin{tabular}[c]{@{}l@{}}classical nuclear\\localization signals\end{tabular}  & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
+TRG\_NLS\_MonoExtN\_4  & \begin{tabular}[c]{@{}l@{}}classical nuclear\\localization signals\end{tabular}  & motifs                                                             & {[}0, $\infty$)              &                                                                                                                                  &                                                                                                \\
 \hline
-MOD\_CDK\_STP          & \begin{tabular}[c]{@{}l@{}}CDK phosphorylation\\motif\end{tabular}               & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
+MOD\_CDK\_STP          & \begin{tabular}[c]{@{}l@{}}CDK phosphorylation\\motif\end{tabular}               & motifs                                                             & {[}0, $\infty$)              &                                                                                                                                  &                                                                                                \\
 \hline
-MOD\_MEC1              & \begin{tabular}[c]{@{}l@{}}Mec1 phosphorylation\\motif\end{tabular}              & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
+MOD\_MEC1              & \begin{tabular}[c]{@{}l@{}}Mec1 phosphorylation\\motif\end{tabular}              & motifs                                                             & {[}0, $\infty$)              &                                                                                                                                  &                                                                                                \\
 \hline
-MOD\_PRK1              & \begin{tabular}[c]{@{}l@{}}Prk1 phosphorylation\\motif\end{tabular}              & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
+MOD\_PRK1              & \begin{tabular}[c]{@{}l@{}}Prk1 phosphorylation\\motif\end{tabular}              & motifs                                                             & {[}0, $\infty$)              &                                                                                                                                  &                                                                                                \\
 \hline
-MOD\_IPL1              & \begin{tabular}[c]{@{}l@{}}Ipl1 phosphorylation\\motif\end{tabular}              & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
+MOD\_IPL1              & \begin{tabular}[c]{@{}l@{}}Ipl1 phosphorylation\\motif\end{tabular}              & motifs                                                             & {[}0, $\infty$)              &                                                                                                                                  &                                                                                                \\
 \hline
-MOD\_PKA               & \begin{tabular}[c]{@{}l@{}}Pka phosphorylation\\motif\end{tabular}               & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
+MOD\_PKA               & \begin{tabular}[c]{@{}l@{}}Pka phosphorylation\\motif\end{tabular}               & motifs                                                             & {[}0, $\infty$)              &                                                                                                                                  &                                                                                                \\
 \hline
-MOD\_CKII              & \begin{tabular}[c]{@{}l@{}}Ckii phosphorylation\\motif\end{tabular}              & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
+MOD\_CKII              & \begin{tabular}[c]{@{}l@{}}Ckii phosphorylation\\motif\end{tabular}              & motifs                                                             & {[}0, $\infty$)              &                                                                                                                                  &                                                                                                \\
 \hline
-MOD\_IME2              & \begin{tabular}[c]{@{}l@{}}Ime2 phosphorylation\\motif\end{tabular}              & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
+MOD\_IME2              & \begin{tabular}[c]{@{}l@{}}Ime2 phosphorylation\\motif\end{tabular}              & motifs                                                             & {[}0, $\infty$)              &                                                                                                                                  &                                                                                                \\
 \hline
-DOC\_PRO               & proline rich motif                                                               & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
+DOC\_PRO               & proline rich motif                                                               & motifs                                                             & {[}0, $\infty$)              &                                                                                                                                  &                                                                                                \\
 \hline
-TRG\_ER\_HDEL          & ER localization motif                                                            & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
+TRG\_ER\_HDEL          & ER localization motif                                                            & motifs                                                             & {[}0, $\infty$)              &                                                                                                                                  &                                                                                                \\
 \hline
-TRG\_MITOCHONDRIA      & \begin{tabular}[c]{@{}l@{}}mitochondrial\\localization motif\end{tabular}        & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
+TRG\_MITOCHONDRIA      & \begin{tabular}[c]{@{}l@{}}mitochondrial\\localization motif\end{tabular}        & motifs                                                             & {[}0, $\infty$)              &                                                                                                                                  &                                                                                                \\
 \hline
-MOD\_ISOMERASE         & \begin{tabular}[c]{@{}l@{}}disulfide isomerase\\motif\end{tabular}               & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
+MOD\_ISOMERASE         & \begin{tabular}[c]{@{}l@{}}disulfide isomerase\\motif\end{tabular}               & motifs                                                             & {[}0, $\infty$)              &                                                                                                                                  &                                                                                                \\
 \hline
-TRG\_FG                & \begin{tabular}[c]{@{}l@{}}FG nucleoporin\\motif\end{tabular}                    & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
+TRG\_FG                & \begin{tabular}[c]{@{}l@{}}FG nucleoporin\\motif\end{tabular}                    & motifs                                                             & {[}0, $\infty$)              &                                                                                                                                  &                                                                                                \\
 \hline
-INT\_RGG               & RGG motif                                                                        & motifs                                                             &                 &                                                                                                                                  &                                                                                                \\
+INT\_RGG               & RGG motif                                                                        & motifs                                                             & {[}0, $\infty$)              &                                                                                                                                  &                                                                                                \\
 \hline

### 2023-04-11 chapter2/figures/scores/scores.py
Added TIFF output to scores.py

@@ -286,4 +286,5 @@ axs[1].set_xlabel('Score rate')
 axs[1].set_yscale('log')
 fig.legend(handles=[Patch(facecolor=color3, label='all')], bbox_to_anchor=(0.9, 0.5), loc='center left')
 fig.savefig('out/scores_histogram.png', dpi=300)
+fig.savefig('out/scores_histogram.tiff', dpi=300)
 plt.close()

### 2023-04-11 chapter2/figures/hierarchy/hierarchy.py
Fixed some issues with getting feature columns and added ASR rate histogram

@@ -2,10 +2,10 @@
 
 import os
 
-import matplotlib.patches as mpatches
 import matplotlib.pyplot as plt
 import numpy as np
 import pandas as pd
+from matplotlib.patches import Patch, Rectangle
 from scipy.cluster.hierarchy import linkage
 from scipy.spatial.distance import pdist
 from src2.brownian.linkage import make_tree
@@ -15,8 +15,11 @@ pdidx = pd.IndexSlice
 min_length = 30
 
 min_indel_columns = 5  # Indel rates below this value are set to 0
-min_aa_rate = 0.5
+min_aa_rate = 1
 min_indel_rate = 0.1
+
+color1 = '#4e79a7'
+grey = '#e6e6e6'
 
 # Load regions
 rows = []
@@ -24,8 +27,7 @@ with open(f'../../IDR_evolution/analysis/IDRpred/region_filter/out/regions_{min_
     field_names = file.readline().rstrip('\n').split('\t')
     for line in file:
         fields = {key: value for key, value in zip(field_names, line.rstrip('\n').split('\t'))}
-        OGid, start, stop, disorder = fields['OGid'], int(fields['start']), int(fields['stop']), fields[
-            'disorder'] == 'True'
+        OGid, start, stop, disorder = fields['OGid'], int(fields['start']), int(fields['stop']), fields['disorder'] == 'True'
         rows.append({'OGid': OGid, 'start': start, 'stop': stop, 'disorder': disorder})
 all_regions = pd.DataFrame(rows)
 
@@ -39,13 +41,16 @@ column_idx = ['OGid', 'start', 'stop', 'disorder']
 region_keys = asr_rates.loc[row_idx, column_idx]
 
 models = pd.read_table(f'../../IDR_evolution/analysis/brownian/get_models/out/models_{min_length}.tsv', header=[0, 1])
-df = region_keys.merge(models.droplevel(1, axis=1), how='left', on=['OGid', 'start', 'stop'])
-df = df.set_index(['OGid', 'start', 'stop', 'disorder'])
+models = region_keys.merge(models.droplevel(1, axis=1), how='left', on=['OGid', 'start', 'stop'])
+models = models.set_index(['OGid', 'start', 'stop', 'disorder'])
 
 feature_groups = {}
 feature_labels = []
 nonmotif_labels = []
[…]

### 2023-04-11 chapter2/chapter2_supporting.tex
Added hierarchy rate figure to chapter 2 supplement

@@
 \label{sfig:CV_rmatrix}
+\end{figure}
+
+\begin{figure}[h!]
+\includegraphics[width=\textwidth]{hierarchy/out/hierarchy_histogram.png}
+\centering
+\caption{\textbf{Rate distributions of substitution models fit to disorder regions.}
+\textbf{(A)} Average amino acid rates in regions.
+\textbf{(B)}.
+Average indel rates in regions.
+For both panels, the grey intervals correspond to the subset of rapidly evolving regions used for the clustering and GO term enrichment analyses.}
+\label{sfig:hierarchy_histogram}
 \end{figure}

### 2023-04-12 chapter2/figures/hierarchy/hierarchy.py
Removed grey as global variable

@@ -19,7 +19,6 @@ min_aa_rate = 1
 min_indel_rate = 0.1
 
 color1 = '#4e79a7'
-grey = '#e6e6e6'
 
 # Load regions
 rows = []
@@ -81,7 +80,7 @@ rect = (0.15, 0.25, 0.825, 0.7)
 subfig = fig.add_subfigure(gs[0, 0], facecolor='none')
 ax = subfig.add_axes(rect)
 xs = asr_rates.loc[asr_rates['disorder'], 'aa_rate_mean']
-ax.axvspan(min_aa_rate, xs.max(), color=grey)
+ax.axvspan(min_aa_rate, xs.max(), color='#e6e6e6')
 ax.hist(xs, bins=100)
 ax.set_xlabel('Average amino acid rate in region')
 ax.set_ylabel('Number of regions')
@@ -90,7 +89,7 @@ subfig.suptitle('A', x=0.025, y=0.975, fontweight='bold')
 subfig = fig.add_subfigure(gs[0, 1], facecolor='none')
 ax = subfig.add_axes(rect)
 xs = asr_rates.loc[asr_rates['disorder'], 'indel_rate_mean']
-ax.axvspan(min_indel_rate, xs.max(), color=grey)
+ax.axvspan(min_indel_rate, xs.max(), color='#e6e6e6')
 ax.hist(xs, bins=100)
 ax.set_xlabel('Average indel rate in region')
 ax.set_ylabel('Number of regions')

### 2023-04-12 chapter2/figures/scores/scores.py

@@ -46,7 +46,6 @@ plot_msa_kwargs = {'hspace': 0.01, 'left': 0.15, 'right': 0.85, 'top': 0.95, 'bo
                    'msa_legend': True, 'legend_kwargs': {'bbox_to_anchor': (0.85, 0.5), 'loc': 'center left', 'fontsize': 8,
                                                          'handletextpad': 0.5, 'markerscale': 1.5, 'handlelength': 1}}
 color3 = '#b07aa1'
-grey = '#e6e6e6'
 
 tree_template = skbio.read('../../IDR_evolution/data/trees/consensus_LG/100R_NI.nwk', 'newick', skbio.TreeNode)
 tip_order = {tip.name: i for i, tip in enumerate(tree_template.tips())}
@@ -279,7 +278,7 @@ quantile = rates['score_fraction'].quantile(0.9, interpolation='higher')  # Capt
 
 fig, axs = plt.subplots(2, 1, gridspec_kw={'left': 0.1, 'right': 0.9, 'top': 0.975, 'bottom': 0.1})
 for ax in axs:
-    ax.axvspan(quantile, rates['score_fraction'].max(), color=grey)
+    ax.axvspan(quantile, rates['score_fraction'].max(), color='#e6e6e6')
     ax.hist(rates['score_fraction'], bins=150, color=color3)
     ax.set_ylabel('Number of regions')
 axs[1].set_xlabel('Score rate')

### 2023-04-13 chapter2/figures/brownian/brownian.py
Added explained variance ratio to axes

@@ -134,11 +134,13 @@ transform = pca.fit_transform(data.to_numpy())
 fig = plt.figure(figsize=(fig_width, fig_height))
 
 # --- PANEL A ---
+pcx, pcy = 0, 1
+
 subfig = fig.add_subfigure(gs[0, 0])
 ax = subfig.add_axes(rectA)
-hb = ax.hexbin(transform[:, 0], transform[:, 1], cmap=cmap, **hexbin_kwargs_log)
-ax.set_xlabel('PC1')
-ax.set_ylabel('PC2')
+hb = ax.hexbin(transform[:, pcx], transform[:, pcy], cmap=cmap, **hexbin_kwargs_log)
+ax.set_xlabel(f'PC{pcx+1} ({pca.explained_variance_ratio_[pcx]:.1%})')
+ax.set_ylabel(f'PC{pcy+1} ({pca.explained_variance_ratio_[pcy]:.1%})')
 handles = [Line2D([], [], label='disorder', marker='h', markerfacecolor=cmap(handle_markerfacecolor),
                   markersize=8, markeredgecolor='none', linestyle='none')]
 ax.legend(handles=handles)
@@ -148,10 +150,10 @@ subfig.suptitle('A', x=0.025, y=0.975, fontweight='bold')
 # --- PANEL B ---
 subfig = fig.add_subfigure(gs[0, 1])
 ax = subfig.add_axes(rectB)
-ax.hexbin(transform[:, 0], transform[:, 1], cmap=cmap, **hexbin_kwargs_log)
-ax.set_xlabel('PC1')
-ax.set_ylabel('PC2')
-add_pca_arrows(ax, pca, data.columns, 0, 1,
+ax.hexbin(transform[:, pcx], transform[:, pcy], cmap=cmap, **hexbin_kwargs_log)
+ax.set_xlabel(f'PC{pcx+1} ({pca.explained_variance_ratio_[pcx]:.1%})')
+ax.set_ylabel(f'PC{pcy+1} ({pca.explained_variance_ratio_[pcy]:.1%})')
+add_pca_arrows(ax, pca, data.columns, pcx, pcy,
                legend_kwargs=legend_kwargs,
                arrow_scale=arrow_scale, arrow_colors=arrow_colors, arrowstyle_kwargs=arrowstyle_kwargs)
 subfig.suptitle('B', x=0.025, y=0.975, fontweight='bold')
@@ -171,11 +173,13 @@ transform = pca.fit_transform(data.to_numpy())
[…]

### 2023-04-13 chapter2/chapter2.tex
Added clarification about motif counts and removed simulation methods section
Added OGid to scores figure caption

@@
 \caption{\textbf{Analyses of disorder scores.}
-\textbf{(A)} Example alignment with the disorder scores of its sequences.
+\textbf{(A)} Example region in the alignment of the sequences in orthologous group 07E3 with their corresponding disorder scores.
 Higher scores indicate a higher probability of disorder.
@@
 Finally, several motifs from ELM were replaced with their metazoan counterparts or updated versions of the same entries~\cite{Kumar2021}.
+Furthermore, unlike the previous work, motifs were left as counts and not normalized to the proteome-wide average.
 These differences are noted in the supplementary data.
@@
 Isoelectric point was calculated with the Python package isoelectric, which is available on PyPI or at \url{https://isoelectric.org/}~\cite{Kozlowski2016}.
-Otherwise, features were calculated with custom code.
+Otherwise, features were implemented with custom code.
 A full list of features and their definitions is given in Table~\ref{stable:features} and Table~\ref{stable:regexes}.
@@
 \subsection{Brownian motion and Ornstein-Uhlenbeck analyses}
-
-\subsection{Simulation analyses}
 

### 2023-04-13 chapter2/chapter2.tex
Removed spelled out model names because will use abbreviations thoughout text

@@
 \caption{\textbf{Hierarchical clustering of evolutionary signatures.}
-The AIC difference between the Brownian motion (BM) and Ornstein-Uhlenbeck (OU) models is measure of their relative goodness of fit to the data with a penalty for the number of parameters in each.
+The AIC difference between the BM and OU models is measure of their relative goodness of fit to the data with a penalty for the number of parameters in each.
 Larger values indicate a better fit by the OU model.}

### 2023-04-13 chapter2/chapter2.tex
Added BM analysis figures to chapter 2

@@
 \label{fig:substitution}
+\end{figure}
+
+\begin{figure}[h!]
+\includegraphics[width=\textwidth]{brownian/out/root.png}
+\centering
+\caption{\textbf{PCA of disorder regions' feature roots.}
+\textbf{(A)} The first two PCs of the disorder regions' feature root distributions.
+The explained variance percentage of each component is indicated in parentheses in the axis labels.
+\textbf{(B)} The same plot as panel A with the projections of original features onto the components shown as arrows.
+Only the 16 features with the largest projections are shown.
+Scaling of the arrows is arbitrary.}
+\label{fig:root}
+\end{figure}
+
+\begin{figure}[h!]
+\includegraphics[width=\textwidth]{brownian/out/rate.png}
+\centering
+\caption{\textbf{PCA of disorder regions' feature rates.}
+\textbf{(A)} The first two PCs of the disorder regions' feature rate distributions.
+The explained variance percentage of each component is indicated in parentheses.
+\textbf{(B)} Scatter plot of the disorder regions' feature rates along the first PC against the sum of the average amino acid and indel substitution rates in those regions.
+\textbf{(C)} The second and third PCs of the disorder regions' feature rate distributions.
+The explained variance percentage of each component is indicated in parentheses in the axis labels.
+\textbf{(D)} The same plot as panel C with the projections of original rates onto the components shown as arrows.
+Only the 16 features with the largest projections are shown.
+Scaling of the arrows is arbitrary.
+\textbf{(E-F)} Example alignments of disorder regions from the orthologous groups 0A8A, 3139, 04B0, respectively.
+The colored bars on the left indicate the hexbin containing that region in panel C.}
+\label{fig:rate}
 \end{figure}

### 2023-04-13 chapter2/chapter2_supporting.tex
Added BM analysis supplement figures

@@
 \label{sfig:CV_rmatrix}
+\end{figure}
+
+\begin{figure}[h!]
+\includegraphics[width=\textwidth]{brownian/out/root_variance.png}
+\centering
+\caption{\textbf{Variance ratios of disorder regions' feature roots.}
+\textbf{(A)} Variance ratios before normalization. \textbf{(B)} Variance ratios after normalization. \textbf{(C)} Scree plot of the explained variance ratio by PC.}
+\label{sfig:root_variance}
+\end{figure}
+
+\begin{figure}[h!]
+\includegraphics[width=\textwidth]{brownian/out/rate_variance.png}
+\centering
+\caption{\textbf{Variance ratios of disorder regions' feature rates.}
+\textbf{(A)} Variance ratios before normalization. \textbf{(B)} Variance ratios after normalization. \textbf{(C)} Scree plot of the explained variance ratio by PC.}
+\label{sfig:rate_variance}
+\end{figure}
+
+\begin{figure}[h!]
+\includegraphics[width=\textwidth]{brownian/out/order.png}
+\centering
+\caption{\textbf{PCA of order regions' feature rates.}
+\textbf{(A)} The second and third PCs of the order regions' feature rate distributions.
+The explained variance percentage of each component is indicated in parentheses in the axis labels.
+\textbf{(B)} The same plot as panel B with the projections of original rates onto the components shown as arrows.
+Only the 16 features with the largest projections are shown.
+Scaling of the arrows is arbitrary.}
+\label{sfig:order}
 \end{figure}

### 2023-04-13 chapter2/chapter2.tex
Wrapped code methods in sloppypar to prevent overfull hbox

@@
 \subsection{Code and data availability}
+\begin{sloppypar}
 The code used to produce the results and analyses is available at \url{https://github.com/marcsingleton/IDR_evolution2023}.
@@
 All primary data are available from publicly accessible sources described in their corresponding sections.
+\end{sloppypar}

### 2023-04-14 chapter1/figures/hmm/hmm.py
Adjusted spacing on HMM figure

@@ -14,19 +14,21 @@ from src1.utils import read_fasta
 
 spid_regex = r'spid=([a-z]+)'
 
-OGids = ['2252', '2A57', '360E']
+records = [('2252', 0, None),
+           ('2A57', 0, 845),
+           ('360E', 0, None)]
 state_labels = ['1A', '1B', '2', '3']
 state_colors = ['C0', 'C3', 'C1', 'C2']
 
 fig_width = 7.5
-fig_height = 3
-dpi = 400
+fig_height = 4.25
+dpi = 600
 plot_msa_kwargs = {'figsize': (fig_width, fig_height),
-                   'left': 0.1, 'right': 0.92, 'top': 0.95, 'bottom': 0.05, 'anchor': (0, 0.5),
-                   'height_ratio': 0.75, 'hspace': 0.01,
+                   'left': 0.1, 'right': 0.92, 'top': 0.95, 'bottom': 0.04, 'anchor': (0, 0.5),
+                   'height_ratio': 0.5, 'hspace': 0.4,
                    'x_labelsize': 5,
                    'data_min': -0.05, 'data_max': 1.05,
-                   'data_labels': state_labels, 'data_linewidths': 1, 'data_colors': state_colors,
+                   'data_labels': state_labels, 'data_linewidths': 1.5, 'data_colors': state_colors,
                    'tree_position': 0, 'tree_width': 0.1,
                    'tree_kwargs': {'linewidth': 0.5, 'tip_labels': False, 'xmin_pad': 0.01, 'xmax_pad': 0.025},
                    'msa_legend': True,
@@ -46,7 +48,7 @@ with open('../../orthology_inference/analysis/ortho_MSA/insertion_hmm/out/model.
 if not os.path.exists('out/'):
     os.mkdir('out/')
 
-for OGid in OGids:
+for OGid, start, stop in records:
     # Load MSA
     msa = []
     for header, seq in read_fasta(f'../../orthology_inference/analysis/ortho_MSA/realign_fastas/out/{OGid}.afa'):
@@ -99,7 +101,8 @@ for OGid in OGids:
     fbs = model.forward_backward(idx_seq)
     data = [fbs[label] for label in state_labels]
 
-    fig = plot_msa_data([record['seq'] for record in msa], data, tree=tree, **plot_msa_kwargs)
+    fig = plot_msa_data([record['seq'][start:stop] for record in msa], data[start:stop], tree=tree,
+                        x_start=start, **plot_msa_kwargs)
[…]

### 2023-04-14 chapter2/chapter2.tex
Removed reference to length cutoff

@@
 The regions were filtered with the following criteria.
-First, segments with non-standard amino acid symbols, which overlapped with any position labeled as ``missing'' by the missing data phylo-HMM, or whose number of non-gap symbols was below a length cutoff were removed.
+First, segments with non-standard amino acid symbols, which overlapped with any position labeled as ``missing'' by the missing data phylo-HMM, or whose number of non-gap symbols was below a length cutoff of 30 residues were removed.
 Regions whose remaining segments satisfied the set of phylogenetic diversity criteria detailed in Table~\ref{stable:diversity_criteria} were included in the final set.
-The length cutoffs were set to 30, 60, and 90 residues, which generated three distinct sets of regions.
 
-\subsection{Fitting amino substitution models}
-To fit amino acid substitution matrices to disorder and order regions, 25 meta-alignments were constructed by randomly sampling 100,000 columns from the respective subsets in the filtered regions derived from the 30 residue length cutoff.
+\subsection{Fitting amino acid substitution models}
+To fit amino acid substitution matrices to disorder and order regions, 25 meta-alignments were constructed by randomly sampling 100,000 columns from the respective subsets in the filtered regions.
 To determine the effect of gaps, the maximum fraction of gaps was set at 0, 50, 100\%.

### 2023-04-14 chapter2/figures/brownian/brownian.py
Fixed not correcting ASR indel rates

@@ -54,6 +54,8 @@ pdidx = pd.IndexSlice
 spid_regex = r'spid=([a-z]+)'
 min_length = 30
 
+min_indel_columns = 5  # Indel rates below this value are set to 0
+
 pca_components = 10
 cmap = plt.colormaps['Greys']
 hexbin_kwargs_log = {'gridsize': 30, 'mincnt': 1, 'linewidth': 0, 'bins': 'log'}
@@ -96,6 +98,8 @@ all_regions = all_segments.drop('ppid', axis=1).drop_duplicates()
 # Load and format data
 asr_rates = pd.read_table(f'../../IDR_evolution/analysis/evofit/asr_stats/out/regions_{min_length}/rates.tsv')
 asr_rates = all_regions.merge(asr_rates, how='right', on=['OGid', 'start', 'stop'])
+row_idx = (asr_rates['indel_num_columns'] < min_indel_columns) | asr_rates['indel_rate_mean'].isna()
+asr_rates.loc[row_idx, 'indel_rate_mean'] = 0
 
 features = all_segments.merge(all_features, how='left', on=['OGid', 'start', 'stop', 'ppid'])
 features = features.groupby(['OGid', 'start', 'stop', 'disorder']).mean()

### 2023-04-17 chapter2/chapter2.tex
Moved order of sentences in feature definition section for clarity
Added methods for calculating substitution rates

@@
 
-\subsection{Fitting amino acid substitution models}
+\subsection{Fitting substitution models and trees}
 To fit amino acid substitution matrices to disorder and order regions, 25 meta-alignments were constructed by randomly sampling 100,000 columns from the respective subsets in the filtered regions.
@@
 
+To obtain estimates of the average substitution rates in each region, separate amino acid and indel models were fit to each alignment.
+For the amino acid substitution models, the columns in the alignments were manually segregated into disorder and order partitions using the regions derived from the AUCPreD scores.
+However, to prevent poor fits from a lack of data, a partition was created only if it contained a minimum of 20 sequences with at least 30 non-gap symbols.
+If one partition met these conditions but the other did not, the disallowed partition was consolidated into the allowed one.
+If neither partition passed, the alignment was skipped.
+These rules ensured that the regions represented in the filtered set were fit with substitution models which were concordant with their predicted disorder states.
+Trees were fit to each partition using an invariant and four discrete gamma rate categories using IQ-TREE 1.6.12~\cite{Yang1994}.
+The disorder partition used a substitution model derived from the average of the state frequencies and rate coefficients fit to the 50\% gap fraction meta-alignment sets sampled from the disorder regions.
+The order partition used the LG substitution model~\cite{Le2008}.
+To prevent overfitting of branch lengths, the trees were restricted to scaled versions of the reference species tree using the -{}-blscale option.
+
+As inference with models that allow insertions and deletions of arbitrary lengths is computationally intractable, a more heuristic approach was taken to quantify the amount of evolutionary divergence resulting from indels in the alignments.
+For a given alignment, all contiguous subsequences of gap symbols with unique start or stop positions in any sequence were defined as binary characters.
+Then for each character, a sequence was coded with the symbol 1 if it contained that character or that character was nested in another contiguous subsequence of gap symbols in the sequence and the symbol 0 otherwise.
+GTR2 models with optimized state frequencies and ascertainment bias correction were fit to the resulting character alignments.
+A discrete gamma rate category was added for every five character columns, up to a maximum of four.
+To prevent overfitting of branch lengths, the trees were restricted to scaled versions of the reference species tree using the -{}-blscale option.
+
+Because the rate and branch lengths of a phylogenetic substitution model always appear as products in the likelihood expression, they are not jointly identifiable parameters.
+Instead, the rate is conventionally taken as equal to one (with inverse count units), and the branch lengths are expressed in terms of the expected number of substitution events.
+For models with multiple rate categories, the equivalent condition is that the mean of the prior distribution over the rate categories is equal to one.
+This effectively makes each rate category a scaling factor of the branch lengths.
+The inferred rate of a column, calculated as the mean of the posterior distribution over the rate categories, is therefore relative to the average across all columns in the alignment.
+An absolute measure of the evolutionary divergence of a column can be obtained by multiplying the inferred rate by the total branch length of the tree.
+However, as the alignments contain variable numbers of species, this total branch length represents the contribution of both the rate and the tree topology.
+To normalize for this effect, the total branch length for each tree fit to an alignment was divided by the total branch length of the reference species tree including only the species in that alignment.
+The reported substitution rate is therefore a product of this scaling factor and the inferred column rate.
+The average amino acid or indel substitution rate for a region was calculated as the mean of the respective rate across all columns.
+
 \subsection{Definition and calculation of features}
@@
 Finally, several motifs from ELM were replaced with their metazoan counterparts or updated versions of the same entries~\cite{Kumar2021}.
+These differences are noted in the supplementary data.
 Furthermore, unlike the previous work, motifs were left as counts and not normalized to the proteome-wide average.
-These differences are noted in the supplementary data.
 Kappa, omega, SCD, hydropathy, PPII propensity, and Wootton-Federhen sequence complexity were calculated with localCIDER 0.1.19~\cite{Holehouse2017}.

### 2023-04-17 chapter2/chapter2.tex
Added details about number of alignments and regions in filtering steps

@@
 
-\subsection{Alignment and tree generation}
-Alignments of single copy orthologs and the outputs of the missing data phylo-HMM were obtained from the analyses conducted in chapter~\ref{chapter:1}.
+\subsection{Alignment and species tree provenance}
+Alignments of 8,566 single copy orthologs and the corresponding outputs of the missing data phylo-HMM were obtained from the analyses conducted in chapter~\ref{chapter:1}.
 Likewise, the LG consensus tree generated by the ``non-invariant, 100\% redundancy'' sampling strategy was used as the input or reference where indicated in subsequent phylogenetic analyses.
@@
 AUCPreD was used to predict the disorder score of each residue of each sequence in the alignments after the gap symbols were removed~\cite{Wang2016}.
-The resulting scores were then aligned using the original alignment.
+(Alignments 0204 and 35C2 contained sequences which exceeded the 10,000-character limit and were excluded from subsequent analyses.) The resulting scores were then aligned using the original alignment.
 The average score for each position was calculated using Gaussian process sequencing weighting over the LG consensus tree~\cite{Altschul1989}.
@@
 First, segments with non-standard amino acid symbols, which overlapped with any position labeled as ``missing'' by the missing data phylo-HMM, or whose number of non-gap symbols was below a length cutoff of 30 residues were removed.
-Regions whose remaining segments satisfied the set of phylogenetic diversity criteria detailed in Table~\ref{stable:diversity_criteria} were included in the final set.
+Regions whose remaining segments failed the set of phylogenetic diversity criteria detailed in Table~\ref{stable:diversity_criteria} were excluded.
+The final set contained 11,445 and 14,927 disorder and order regions, respectively, from 8,466 distinct alignments.
 

### 2023-04-17 chapter1/chapter1.tex
Correct use of homology to orthology for consistency
Added reference for Felsenstein's pruning algorithm

@@
 \centering
-\caption{\textbf{Overview of homology inference pipeline.}
+\caption{\textbf{Overview of orthology inference pipeline.}
 \textbf{(A-C)} The large, open circles represent the annotated genomes, and the small, filled circles represent the protein sequences associated with each annotation.
@@
 
-All HMM algorithms were implemented with custom code which is available as the package Homomorph on PyPI.
+The likelihoods of phylogenetic trees were efficiently calculated with Felsenstein's pruning algorithm, and all HMM algorithms were implemented with custom code which is available as the package Homomorph on PyPI~\cite{Felsenstein1981}.
 The insertion phylo-HMM was trained on 47,387 manually labeled columns in 14 alignments, and the missing data phylo-HMM was trained on 67,001 manually labeled positions in 23 sequences in 11 unique alignments (Fig.~\ref{sfig:insertion_training}, Fig.~\ref{sfig:missing_training}).

### 2023-04-18 chapter2/chapter2.tex
Added model fitting section to methods

@@
 \subsection{Brownian motion and Ornstein-Uhlenbeck analyses}
+Brownian motion (BM) model parameters were calculated with two methods.
+The first used Felsenstein's contrasts algorithm to efficiently calculate roots and contrasts for the disorder scores and features of each region~\cite{Felsenstein1973, Felsenstein1981}.
+Rates were calculated as the mean of the squares of the contrasts.
+Though these values are unbiased, they are not maximum likelihood estimates and are inappropriate for use with the Akaike information criterion (AIC)~\cite{Akaike1974}.
+Thus, they were used for analyses involving only the BM model.
+For comparison with the Ornstein-Uhlenbeck (OU) model, the BM parameters were calculated by maximizing the likelihood.
+The OU model parameters were also calculated via maximum likelihood estimation.
+To avoid issues with model identifiability, the root was treated as a random variable~\cite{Ho2014}.
+Thus, the covariance matrix, $V$, was parameterized as $V_{ij} = e^{\alpha d_{ij}}$ where $d_{ij}$ is the tree distance between tips $i$ and $j$, and $\alpha$ is the selection strength~\cite{An2008}.
+
+The AICs were calculated for the models of each feature using their maximized likelihoods and two and three parameters for the BM and OU models, respectively.
+The pairwise differences in the AICs yielded a vector with 82 components, each representing the relative goodness of fit of the OU model over the BM model after accounting for their difference in complexity.
+The vectors were clustered using the correlation distance metric and the UPGMA algorithm.
+Clusters were manually chosen for subsequent GO analyses.
 

### 2023-04-18 chapter2/chapter2.tex
Added GO term analysis methods

@@
 \subsection{GO term analyses}
+The 2022-03-22 go-basic release of the Gene Ontology was obtained from the GO Consortium website.
+The gene association file for the 2022\_02 release of the \textit{D.
+melanogaster} genome annotation was obtained from FlyBase.
+Obsolete annotations were dropped, and the remaining annotations were filtered by qualifiers and evidence code.
+The allowed qualifiers were ``enables,'' ``contributes\_to,'' ``involved\_in,'' ``located\_in,'' ``part\_of,'' and ``is\_active\_in.'' The allowed evidence codes were all experimental sources, traceable author statement (TAS), and inferred by curator (IC).
+The annotations were propagated up the ontology graph and joined with the region sets, so every annotation associated with a gene was associated with the regions derived from that gene.
+$P$-values were calculated with exact hypergeometric probabilities with regions considered as the sampling unit.
+For the disorder score analysis, the reference set was the filtered regions, and the enrichment set was regions in the upper decile of the score rate distribution.
+For the cluster analysis, the reference set was the filtered regions after the additional filtering by substitution rates, and the enrichments sets were the regions in each cluster.
 

### 2023-04-18 chapter1/chapter1.tex
Added phylo-HMM reference to chapter 1

@@
 
-Phylogenetic HMMs (phylo-HMMs) are statistical models that incorporate phylogenetic and spatial information to calculate the probability that each observation in a sequence was generated by one of several hidden states~\cite{Felsenstein1996}.
+Phylogenetic HMMs (phylo-HMMs) are statistical models that incorporate phylogenetic and spatial information to calculate the probability that each observation in a sequence was generated by one of several hidden states~\cite{Felsenstein1996, Siepel2004}.
 Since they can evaluate both the probability of a gap pattern in a column given the known phylogenetic relationships and the local context, a phylo-HMM was used to segment the alignment into contiguous regions with different patterns of gaps.

### 2023-04-18 chapter2/chapter2.tex
Added sentence about restricting BM-OU comparison
Added clarification to indel rate mapping and filtering

@@
 Because the rate and branch lengths of a phylogenetic substitution model always appear as products in the likelihood expression, they are not jointly identifiable parameters.
-Instead, the rate is conventionally taken as equal to one (with inverse count units), and the branch lengths are expressed in terms of the expected number of substitution events.
+Instead, the rate is conventionally taken as equal to one (with inverse count units), and the branch lengths are expressed in terms of the expected number of substitution events per column.
 For models with multiple rate categories, the equivalent condition is that the mean of the prior distribution over the rate categories is equal to one.
@@
 The average amino acid or indel substitution rate for a region was calculated as the mean of the respective rate across all columns.
+Because the indel rates were associated with columns in character alignments, they were mapped back to the original sequence alignment by assigning half of a character's rate to its start and stop positions.
+Since indel models with limited data were prone to overfitting, rates obtained from character alignments with fewer than five columns were set to zero.
 
@@
 Clusters were manually chosen for subsequent GO analyses.
+To enrich these clusters for regions with a high likelihood of feature conservation despite high levels of sequence divergence, this analysis was restricted to regions whose amino acid and indel substitution rates exceeded 1 and 0.1, respectively.
 

### 2023-04-18 chapter2/chapter2.tex
Added references to supplement figures

@@
 Clusters were manually chosen for subsequent GO analyses.
-To enrich these clusters for regions with a high likelihood of feature conservation despite high levels of sequence divergence, this analysis was restricted to regions whose amino acid and indel substitution rates exceeded 1 and 0.1, respectively.
+To enrich these clusters for regions with a high likelihood of feature conservation despite high levels of sequence divergence, this analysis was restricted to regions whose amino acid and indel substitution rates exceeded 1 and 0.1, respectively (Fig.~\ref{sfig:hierarchy_histogram}).
 
@@
 $P$-values were calculated with exact hypergeometric probabilities with regions considered as the sampling unit.
-For the disorder score analysis, the reference set was the filtered regions, and the enrichment set was regions in the upper decile of the score rate distribution.
+For the disorder score analysis, the reference set was the filtered regions, and the enrichment set was regions in the upper decile of the score rate distribution (Fig.~\ref{sfig:scores_histogram}).
 For the cluster analysis, the reference set was the filtered regions after the additional filtering by substitution rates, and the enrichments sets were the regions in each cluster.

### 2023-04-18 chapter2/chapter2.tex
Added references for the GO

@@
 \subsection{GO term analyses}
-The 2022-03-22 go-basic release of the Gene Ontology was obtained from the GO Consortium website.
+The 2022-03-22 go-basic release of the Gene Ontology was obtained from the GO Consortium website~\cite{Ashburner2000, 2020GOConsortium}.
 The gene association file for the 2022\_02 release of the \textit{D.
-melanogaster} genome annotation was obtained from FlyBase.
+melanogaster} genome annotation was obtained from FlyBase~\cite{Gramates2022}.
 Obsolete annotations were dropped, and the remaining annotations were filtered by qualifiers and evidence code.

### 2023-04-20 chapter2/chapter2.tex
Added outline for results and conclusion
Swapped order of score and substitution figures
Added motivation for AUCPreD in methods

@@
 \section{Results}
-\begin{figure}[h!]
-\includegraphics[width=\textwidth]{scores/out/scores_07E3.png}
-\centering
-\caption{\textbf{Analyses of disorder scores.}
-\textbf{(A)} Example region in the alignment of the sequences in orthologous group 07E3 with their corresponding disorder scores.
-Higher scores indicate a higher probability of disorder.
-Disorder score traces are colored by the position of its associated species on the phylogenetic tree.
-\textbf{(B)} Correlations between disorder scores and feature contrasts in regions.
-Asterisks indicate statistically significant correlations as computed by permutation tests ($p < 0.001$).
-\textbf{(C-D)} Example scatter plots showing correlations plotted in panel B.
-\textbf{(E)} GO term analysis of regions with rapidly evolving disorder scores.
-Only terms whose $p$-values are below 0.001 are shown.}
-\label{fig:scores}
-\end{figure}
+% RESULTS
+% I. Identification and filtering of IDRs
+%     A.
+From a set of over 8,500 alignments of single copy orthologs from the genomes of up to 33 species in the \textit{Drosophila} genus, we identified X regions with high levels of inferred disorder using the predictor AUCPreD.
+To highlight the unique features of IDR evolution in subsequent analyses, we also extracted a complementary set of Y regions with low levels of inferred disorder.
+Both sets were filtered on several criteria, including the lengths of their sequences and the phylogenetic diversity.
+In the subsequent discussion, we refer to these filtered sets as the ``disorder'' and ``order'' regions, respectively.
+The disorder
+%     B. Some notes about the length distributions and other basic statistics
+%
+% II. Disorder regions have distinct patterns of residue substitution
+%     A.
+To gain a global view of substitution patterns of amino acid residues in disorder and order regions, we fit amino acid substitution models to meta-alignments sampled from the respective sets of filtered regions.
+%     B. Substitution models are parameterized in terms of the rates of change from one residue to another.
+%         1. The rates are not necessarily equal between for a given pair of residues when the initial and target states are swapped.
+%         2. For example, the rate of change of valine to tryptophan can be distinct from that of tryptophan to valine.
+%         3.
+In practice, substitution models are typically constrained to fulfill a condition called time-reversibility, as this converts a difficult multivariate optimization of the tree's branch lengths into a series of simpler univariate optimizations~\cite{Felsenstein1981}.
+Therefore, the models are parameterized in terms of a frequency vector, $\mathbm{pi}$, and an exchangeability matrix, $S$.
+The frequency vector determines the model's expected residue frequencies at equilibrium, meaning all sequences eventually approach this distribution, no matter their initial composition.
+The exchangeability matrix is symmetric ($s_{ij} = s_{ji}$) and encodes the propensity for two residues to interconvert.
+Because the rate of change from residue $i$ to residue $j$ is given by $r_{ij} = s_{ij}\pi_{j}$, higher exchangeability coefficients yield higher rates of conversion.
+Thus, exchangeability coefficients are frequently interpreted as a measure of biochemical similarity between residues.
+%     C. To highlight the differences in patterns of residue substitution between the disorder and order regions, the parameters in each model are directly compared in Fig.~\ref{fig:substitution}.
+%         1.
+The disorder regions show an enrichment of "disorder-promoting" residues such as serine, proline, and alanine, and a depletion of hydrophobic and bulky residues such as trytophan and phenylalanine (Fig.~\ref{fig:substitution}A).
+%         2.
+Overall, the exchangeability matrices fit to the disorder and order regions are similar in their overall pattern of high and low coefficients (Fig.~\ref{fig:substitution}B-C).
+However, the log ratios of the disorder to the order exchangeability coefficients show clear trends within and between the disorder-enriched and -depleted residues.
+In disorder regions, disorder-enriched residues are less exchangeable with each other, whereas disorder-depleted residues are more exchangeable with each other and with disorder-enriched residues (Fig.~\ref{fig:substitution}D).
+
+% III. Brownian motion score analyses
+%     A. Motivation for use of phylogenetic models
+%         1. For each residue in the input sequence, AUCPreD returns a score between 0 and 1 where higher values indicate higher confidence in a prediction of structural disorder.
+%     B.
+When examining the disorder score traces in greater detail alongside the alignments, we were struck by the significant variability in some traces, even when there was a relatively high level of sequence identity.
+%     C. To better understand the sequence features underlying these changes in predicted disorder, we calculated the values for 82 disorder-associated features in each region
+%     D. Though this analysis is limited by the resolution of the regions, which varies from 30 to ??? residues, it indicates the prediction algorithm is sensitive to these features in bulk.
+%          1. Some more observations of the overall trends ...
+%     E. Interestingly, the correlations with many motifs were statistically significant, though small in magnitude relative to the non-motif features.
+%          1. Look into motif regexes and see if ones with highest correlations have any common features
+%     F. To determine if regions with rapidly evolving disorder scores were enriched in any processes, we extracted the upper decile of their rate distribution
+%          1. Many chromatin and DNA related features
+
+% IV. Brownian motion feature analyses
+%     A.
+Having calculated features for all IDRs in our data set, we were curious the distribution contained any global structure which would allow the identification of classes with distinct biophysical or functional properties
+%     A. Feature roots
+%         1. Discuss differences in scaling and need for z-scores
+%         2. Discussion of sample regions in figure? (need to add)
+%     B.
+The phylogenetic contrasts algorithm also permits an estimate of the rate of feature evolution, which is effectively the variance of the distribution of features across all sequences after correcting for the phylogenetic relationships between them
+%         1. PC1 correlates with rate
+%         2. PC2 and PC3 show three major axes of unconstrained variation: Q-rich, G-rich, acidic
+
+% V. BM-OU analyses
+%    A. While these analyses measure the rates of change of features after accounting the the phylogenetic relationships between sequences, they do not directly test for conservation of features
+%         1. In fact, under the BM model, trait variation is unconstrained and will increase without bound over time
+%         2. Instead, demonstrating evidence of conservation requires a comparison to a model where feature evolution is constrained.
+%         3. A common choice is the Ornstein-Uhlenbeck (OU) model
+%             i. Some explanation about how it works
+
+% CONCLUSION
+% I. Summary of results
+%    A. In this study we showed...
+%    B. IDRs have patterns of evolution distinct from folded domains
+%    C. Elevated rates of residue substitution and indels
+% II. General caveats
+
+% II. Exchangeability and rate matrices
+%    A. Reproduced trends seen in previous substitution models fit to disorder regions
+%    B. Interpretations of exchangeability and rate matrices
+%        1. Disorder-enriched residues are less interchangeable and play specific roles in these regions
+%        2. Disorder-depeleted residues not constrained by packing and can act more generically as small hydrophobic patches
+%            i. Stickers and spacers from Holehouse
+%            ii. Staller papers and other high throughput mutational scanning of ADs
+%            iii. Halfmann glutamine vs asparagine paper
+%        3. Note about variances of parameters as shown in CV graphs
+
+% III. Disorder score analyses
+%    A. Comparison of features and disorder scores acts as a kind of ``mutational scanning'' experiment [citation to Eddy?]
+%        1. Contrasts of features and disorder scores shows the disorder predictor is sensitive to most features used in this analyses
+%        2. However, highest correlations to expected features like fraction_disorder and hydropathy
+%        3.
+Interestingly, negatively correlated with wf_complexity.
+Many disordered regions have low levels of sequence complexity.
+However this is a statistical rather than a biophysical criterion, and some low complexity regions are structured, like collagen.
+This indicates that these models can confuse correlation and causation during training, and their representations should be interpreted carefully.
+%        1. Furthermore, resolution of analyses limited because features calculated at level of regions rather than a fixed window around each residue.
+%    B. GO annotation enrichment analysis suggests disorder scores undergo rapid evolution in proteins involved in DNA repair and found in extracellular structure
+%        1. Not corrected for multiple testing nor is the false discovery rate controlled, so should be considered as hypotheses for further investigation
+%        2. However, this result is at least consistent with other known functions of disorder
+%            i. Interactions with extracellular components and signaling, both highly enriched in disorder?
+%            ii. These regions could contain MoRFs, so they are already on the boundary between folded and disorder
+%            iii. As evolution optimizes these interactions, small changes in the biophysical properties of these regions may have disproportionate effects on their structural ensembles
+%            iv. The variable scoring by the disorder predictor may reflect this kind of ``phrase transition''
+
+% IV. Feature analyses
+%    A. Though no discrete clusters are visible, only captures x% of the overall variation, so significant amounts of variation not represented in these axes
+%    B. Hierarchical clustering of the rates reveals additional structure.
+%    C. Other non-linear dimensionality reductions like UMAP be more successful at preserving the fine details of the high-dimensional distributions in the low-dimensional space
+%        1.
+However, these methods designed primarily for visualization and as result can significant distort the global structure of high-dimensional data to preverse its local topology.
+Thus, these embeddings are generally not suitable for downstream analyses.
+
+% V. OU analyses
+%    A. Discussion of issues with statistical inference and need for simulations to address inference issues with OU model
+%        1. Lack of power or prone to type I errors?
+%    B. Better models for handling motif counts
+%        1.
+Unlike other features where small changes in a sequence general yield small changes in their value, the complexity of the regular expressions made them highly sensitive to small perturbations.
+%    C. Need for comparisons with simulations with amino acid substitution and indel models
+%        1. Weaknesses are explicit assumptions about substitution and indel processes and high computational cost
+%        2. Is also a strength because allows testing of increasingly specific hypotheses for IDR evolution
+%        3. Complementary approach -- fewer assumptions about description of evolution but cannot be tailored to specific hypotheses
+%        4. Hybrid approach using comparisons of rates as the statistical test and simulations to generate the null distribution
+%    D. Mention other distribution free approaches
+
+% VI. Big picture and wrap up
+%     A. Experimental swaps of IDRs in similar classes
+%     B. What are features?
+%         - Contrastive learning (Moses)
+%         - Prediction on gold standard data sets
+%         - Embeddings from LLMs trained on large data sets
 
@@
 \label{fig:substitution}
+\end{figure}
+
+\begin{figure}[h!]
+\includegraphics[width=\textwidth]{scores/out/scores_07E3.png}
+\centering
+\caption{\textbf{Analyses of disorder scores.}
+\textbf{(A)} Example region in the alignment of the sequences in orthologous group 07E3 with their corresponding disorder scores.
+Higher scores indicate a higher probability of disorder.
+Disorder score traces are colored by the position of its associated species on the phylogenetic tree.
+\textbf{(B)} Correlations between disorder scores and feature contrasts in regions.
+Asterisks indicate statistically significant correlations as computed by permutation tests ($p < 0.001$).
+\textbf{(C-D)} Example scatter plots showing correlations plotted in panel B.
+\textbf{(E)} GO term analysis of regions with rapidly evolving disorder scores.
+Only terms whose $p$-values are below 0.001 are shown.}
+\label{fig:scores}
 \end{figure}
@@
 \subsection{IDR prediction and filtering}
-AUCPreD was used to predict the disorder score of each residue of each sequence in the alignments after the gap symbols were removed~\cite{Wang2016}.
+Based on its strong performance in a recent assessment of disorder predictors, AUCPreD was chosen to identify regions with a high confidence of intrinsic disorder~\cite{Wang2016, Necci2021}.
+After removing the gap symbols from the sequences in the alignments, the disorder scores of each sequence were predicted individually.
 (Alignments 0204 and 35C2 contained sequences which exceeded the 10,000-character limit and were excluded from subsequent analyses.) The resulting scores were then aligned using the original alignment.

### 2023-04-20 chapter2/figures/regions/regions.py
Added regions.py

@@ -1,0 +1,91 @@
+"""Make figures for basic region statistics."""
+
+import os
+import re
+
+import matplotlib.pyplot as plt
+import pandas as pd
+from numpy import linspace
+from src2.utils import read_fasta
+
+pdidx = pd.IndexSlice
+ppid_regex = r'ppid=([A-Za-z0-9_.]+)'
+min_length = 30
+
+min_indel_columns = 5  # Indel rates below this value are set to 0
+min_aa_rate = 1
+min_indel_rate = 0.1
+
+# Load regions as segments
+rows = []
+with open(f'../../IDR_evolution/analysis/IDRpred/region_filter/out/regions_{min_length}.tsv') as file:
+    field_names = file.readline().rstrip('\n').split('\t')
+    for line in file:
+        fields = {key: value for key, value in zip(field_names, line.rstrip('\n').split('\t'))}
+        OGid, start, stop, disorder = fields['OGid'], int(fields['start']), int(fields['stop']), fields['disorder'] == 'True'
+        for ppid in fields['ppids'].split(','):
+            rows.append({'OGid': OGid, 'start': start, 'stop': stop, 'disorder': disorder,
+                         'ppid': ppid, 'min_length': min_length})
+all_segments = pd.DataFrame(rows)
+all_regions = all_segments[['OGid', 'start', 'stop', 'disorder']].drop_duplicates()
+
+# Get segment lengths
+rows = []
+for OGid, group in all_segments[['OGid', 'start', 'stop']].drop_duplicates().groupby('OGid'):
+    msa = read_fasta(f'../../IDR_evolution/data/alignments/fastas/{OGid}.afa')
+    msa = [(re.search(ppid_regex, header).group(1), seq) for header, seq in msa]
+
+    for row in group.itertuples():
+        region_length = row.stop - row.start
+        for ppid, seq in msa:
+            segment = seq[row.start:row.stop]
+            segment_length = region_length - segment.count('.') - segment.count('-')
+            rows.append({'OGid': OGid, 'start': row.start, 'stop': row.stop, 'ppid': ppid, 'length': segment_length})
+all_lengths = pd.DataFrame(rows)
+
+segments = all_segments[all_segments['min_length'] == min_length]
[…]

### 2023-04-20 chapter2/chapter2.tex
Expanded first section of results in full paragraph
Added first section of results

@@
 \section{Results}
-% RESULTS
-% I. Identification and filtering of IDRs
-%     A.
-From a set of over 8,500 alignments of single copy orthologs from the genomes of up to 33 species in the \textit{Drosophila} genus, we identified X regions with high levels of inferred disorder using the predictor AUCPreD.
-To highlight the unique features of IDR evolution in subsequent analyses, we also extracted a complementary set of Y regions with low levels of inferred disorder.
-Both sets were filtered on several criteria, including the lengths of their sequences and the phylogenetic diversity.
-In the subsequent discussion, we refer to these filtered sets as the ``disorder'' and ``order'' regions, respectively.
-The disorder
-%     B. Some notes about the length distributions and other basic statistics
+\subsection{IDRs are shorter and more divergent than non-disordered regions}
+We identified regions with high levels of inferred intrinsic disorder in over 8,500 alignments of single copy orthologs from 33 species in the \textit{Drosophila} genus using the disorder predictor AUCPreD~\cite{Wang2016}.
+To highlight the unique features of IDR evolution in subsequent analyses, we also extracted a complementary set of regions with low levels of inferred disorder.
+Both sets were filtered on several criteria, including the lengths of their sequences and their phylogenetic diversity, which yielded 11,445 and 14,927 regions, respectively, from 8,466 unique alignments.
+In the subsequent discussion, we refer to these sets as the ``disorder'' and ``order'' regions, respectively.
+To investigate the broad differences in basic sequence statistics between the two region sets, we first generated histograms from the average length of each region (Fig.~\ref{sfig:regions}).
+Although both distributions span several orders of magnitude, the order regions are generally longer than the disorder regions, with means of 105 and 245 residues, respectively.
+We then quantified the sequence divergence in each region by fitting phylogenetic trees to the alignments using amino acid and indel substitution models.
+The average rates of substitution are significantly larger in the disorder regions, demonstrating that while both sets contain conserved and divergent regions, IDRs are enriched in more rapidly evolving sequences (Fig.~\ref{sfig:regions}).
 %

### 2023-04-21 chapter2/figures/regions/regions.py
Added U test calculation

@@ -5,7 +5,8 @@ import re
 
 import matplotlib.pyplot as plt
 import pandas as pd
-from numpy import linspace
+import scipy.stats as stats
+from numpy import linspace, sqrt
 from src2.utils import read_fasta
 
 pdidx = pd.IndexSlice
@@ -79,6 +80,18 @@ x1 = (asr_rates.loc[asr_rates['disorder'] == True, 'aa_rate_mean'] +
       asr_rates.loc[asr_rates['disorder'] == True, 'indel_rate_mean']).dropna()
 x2 = (asr_rates.loc[asr_rates['disorder'] == False, 'aa_rate_mean'] +
       asr_rates.loc[asr_rates['disorder'] == False, 'indel_rate_mean']).dropna()
+
+# Manual calculation of log p-value due to underflow
+# Code adapted from SciPy stats example in mannwhitneyu function
+nx1, nx2 = len(x1), len(x2)
+N = nx1 + nx2
+U1, p = stats.mannwhitneyu(x1, x2, alternative='greater')
+U2 = nx1 * nx2 - U1
+U = min(U1, U2)
+z = (U - nx1 * nx2 / 2 + 0.5) / sqrt(nx1 * nx2 * (N + 1) / 12)
+logp = 2 * stats.norm.logcdf(z)
+with open('out/utest.txt', 'w') as file:
+    file.write(f'log p-value (Mann-Whitney U test): {logp}\n')
 
 subfig = fig.add_subfigure(gs[0, 1])
 ax = subfig.subplots()

### 2023-04-21 chapter2/chapter2_supporting.tex
Fixed some issues with dangling hline commands not preceded by \\
Swapped order of score and substitution figures
Added regions figure to chapter 2 supplement
Added U test calculation result to caption

@@
 \begin{figure}[h!]
-\includegraphics[width=0.75\textwidth]{scores/out/scores_histogram.png}
+\includegraphics[width=\textwidth]{regions/out/regions.png}
 \centering
-\caption{\textbf{Histogram of disorder score rates in regions.}
-The grey interval indicates the upper decile of the distribution across both disorder and order regions, which was used as the input set for the GO term enrichment analysis.}
-\label{sfig:scores_histogram}
+\caption{\textbf{Summary statistics of disorder and order regions.}
+\textbf{(A)} Distribution of mean length of regions.
+\textbf{(B)} Boxplot of the sums of the average amino acid and indel substitution rates in the disorder and order regions.
+The substitution rates are significantly greater in the disorder regions than in the order regions ($p < 1 \times 10^{-10}$, Mann-Whitney \textit{U} test).}
+\label{sfig:regions}
 \end{figure}
@@
 \begin{figure}[h!]
+\includegraphics[width=0.75\textwidth]{scores/out/scores_histogram.png}
+\centering
+\caption{\textbf{Histogram of disorder score rates in regions.}
+The grey interval indicates the upper decile of the distribution across both disorder and order regions, which was used as the input set for the GO term enrichment analysis.}
+\label{sfig:scores_histogram}
+\end{figure}
+
+\begin{figure}[h!]
 \includegraphics[width=\textwidth]{brownian/out/root_variance.png}
@@
 % Start first head
-\hline
+\\ \hline
 \textbf{Feature ID}    & \textbf{Feature name}                                                            & \textbf{Group name}                                                & \textbf{Range} & \textbf{Description}                                                                                                             & \begin{tabular}[c]{@{}l@{}}\textbf{Changes from}\\\textbf{Zarin \textit{et al.}}~\cite{Zarin2019}\end{tabular}
@@
 \multicolumn{6}{l}
-{\textbf{\tablename\ \thetable} (continued)} \\
-\hline
+{\textbf{\tablename\ \thetable} (continued)}
+\\ \hline
 \textbf{Feature ID}    & \textbf{Feature name}                                                            & \textbf{Group name}                                                & \textbf{Range} & \textbf{Description}                                                                                                             & \begin{tabular}[c]{@{}l@{}}\textbf{Changes from}\\\textbf{Zarin \textit{et al.}~\cite{Zarin2019}}\end{tabular}
@@
 % Start foot
-\multicolumn{6}{|c|}{Continued on next page} \\
-\hline
+\multicolumn{6}{|c|}{Continued on next page}
+\\ \hline
 \endfoot
@@
 \begin{longtable}{|l|l|}
-\caption{\textbf{Feature regular expresions.}}
+\caption{\textbf{Feature regular expressions.}}
 \label{stable:regexes}
 % Start first head
-\hline
+\\ \hline
 \textbf{Feature ID}    & \textbf{Regular expression}
@@
 \multicolumn{2}{l}
-{\textbf{\tablename\ \thetable} (continued)} \\
-\hline
+{\textbf{\tablename\ \thetable} (continued)}
+\\ \hline
 \textbf{Feature ID}    & \textbf{Regular expression}
@@
 % Start foot
-\multicolumn{2}{|c|}{Continued on next page} \\
-\hline
+\multicolumn{2}{|c|}{Continued on next page}
+\\ \hline
 \endfoot

### 2023-04-21 chapter2/chapter2.tex
Changed p-value to math mode where comparisons to thresholds are used; otherwise use italics
Revised substitution result outline
Converted substitution result outline into paragraphs

@@
 The average rates of substitution are significantly larger in the disorder regions, demonstrating that while both sets contain conserved and divergent regions, IDRs are enriched in more rapidly evolving sequences (Fig.~\ref{sfig:regions}).
-%
-% II. Disorder regions have distinct patterns of residue substitution
-%     A.
-To gain a global view of substitution patterns of amino acid residues in disorder and order regions, we fit amino acid substitution models to meta-alignments sampled from the respective sets of filtered regions.
-%     B. Substitution models are parameterized in terms of the rates of change from one residue to another.
-%         1. The rates are not necessarily equal between for a given pair of residues when the initial and target states are swapped.
-%         2. For example, the rate of change of valine to tryptophan can be distinct from that of tryptophan to valine.
-%         3.
+
+\subsection{Disorder regions have distinct patterns of residue substitution}
+To gain insight into the substitution patterns of amino acid residues in disorder and order regions, we fit amino acid substitution models to meta-alignments sampled from the respective sets of filtered regions.
+These models are probabilistic descriptions of sequence evolution and are parameterized in terms of the one-way rates of change from one residue to another.
+Thus, the rates are not necessarily equal for a given pair of residues when the initial and target residues are swapped.
+For example, the rate of change of valine to tryptophan can be distinct from that of tryptophan to valine.
 In practice, substitution models are typically constrained to fulfill a condition called time-reversibility, as this converts a difficult multivariate optimization of the tree's branch lengths into a series of simpler univariate optimizations~\cite{Felsenstein1981}.
-Therefore, the models are parameterized in terms of a frequency vector, $\mathbm{pi}$, and an exchangeability matrix, $S$.
-The frequency vector determines the model's expected residue frequencies at equilibrium, meaning all sequences eventually approach this distribution, no matter their initial composition.
+A common method for fulfilling this condition is parameterizing the model in terms of a frequency vector, $\mathbm{pi}$, and an exchangeability matrix, $S$.
+The frequency vector determines the model's expected residue frequencies at equilibrium, meaning the model dictates that all sequences eventually approach this distribution, no matter their initial composition.
 The exchangeability matrix is symmetric ($s_{ij} = s_{ji}$) and encodes the propensity for two residues to interconvert.
@@
 Thus, exchangeability coefficients are frequently interpreted as a measure of biochemical similarity between residues.
-%     C. To highlight the differences in patterns of residue substitution between the disorder and order regions, the parameters in each model are directly compared in Fig.~\ref{fig:substitution}.
-%         1.
-The disorder regions show an enrichment of "disorder-promoting" residues such as serine, proline, and alanine, and a depletion of hydrophobic and bulky residues such as trytophan and phenylalanine (Fig.~\ref{fig:substitution}A).
-%         2.
-Overall, the exchangeability matrices fit to the disorder and order regions are similar in their overall pattern of high and low coefficients (Fig.~\ref{fig:substitution}B-C).
-However, the log ratios of the disorder to the order exchangeability coefficients show clear trends within and between the disorder-enriched and -depleted residues.
+
+To highlight the differences in patterns of residue substitution between the disorder and order regions, the parameters in each model are directly compared in Fig.~\ref{fig:substitution}, beginning with the frequency vectors.
+The disorder regions show an enrichment of ``disorder-promoting'' residues such as serine, proline, and alanine, and a depletion of hydrophobic and bulky residues such as trytophan and phenylalanine (Fig.~\ref{fig:substitution}A).
+Overall, the exchangeability matrices fit to the disorder and order regions have similar trends of high and low coefficients (Fig.~\ref{fig:substitution}B-C).
+However, the log ratios of the disorder to the order exchangeability coefficients show clear patterns within and between the disorder-enriched and -depleted residues.
 In disorder regions, disorder-enriched residues are less exchangeable with each other, whereas disorder-depleted residues are more exchangeable with each other and with disorder-enriched residues (Fig.~\ref{fig:substitution}D).
+Likewise, the log ratios of the rate coefficient also exhibit a division between the disorder-enriched and -depleted residues.
+As the coefficients model the one-way rate of substitutions between residues, the vertical and horizontal axes indicate the initial and target residues, respectively.
+Thus, the matrix indicates a bias to evolving towards a more disorder-promoting composition.
 
+% RESULTS
 % III. Brownian motion score analyses
@@
 \textbf{(E)} GO term analysis of regions with rapidly evolving disorder scores.
-Only terms whose $p$-values are below 0.001 are shown.}
+Only terms where $p < 0.001$ are shown.}
 \label{fig:scores}
@@
 The annotations were propagated up the ontology graph and joined with the region sets, so every annotation associated with a gene was associated with the regions derived from that gene.
-$P$-values were calculated with exact hypergeometric probabilities with regions considered as the sampling unit.
+\textit{P}-values were calculated with exact hypergeometric probabilities with regions considered as the sampling unit.
 For the disorder score analysis, the reference set was the filtered regions, and the enrichment set was regions in the upper decile of the score rate distribution (Fig.~\ref{sfig:scores_histogram}).

### 2023-04-21 chapter2/chapter2.tex
Tweaked some wording in substitution paragraph

@@
 \subsection{Disorder regions have distinct patterns of residue substitution}
-To gain insight into the substitution patterns of amino acid residues in disorder and order regions, we fit amino acid substitution models to meta-alignments sampled from the respective sets of filtered regions.
+To gain insight into the substitution patterns of amino acid residues in disorder and order regions, we fit substitution models to meta-alignments sampled from the respective sets of filtered regions.
 These models are probabilistic descriptions of sequence evolution and are parameterized in terms of the one-way rates of change from one residue to another.
@@
 In practice, substitution models are typically constrained to fulfill a condition called time-reversibility, as this converts a difficult multivariate optimization of the tree's branch lengths into a series of simpler univariate optimizations~\cite{Felsenstein1981}.
-A common method for fulfilling this condition is parameterizing the model in terms of a frequency vector, $\mathbm{pi}$, and an exchangeability matrix, $S$.
+A common method for fulfilling this condition is parameterizing the model in terms of a frequency vector, $\boldsymbol \pi$, and an exchangeability matrix, $S$.
 The frequency vector determines the model's expected residue frequencies at equilibrium, meaning the model dictates that all sequences eventually approach this distribution, no matter their initial composition.
@@
 The disorder regions show an enrichment of ``disorder-promoting'' residues such as serine, proline, and alanine, and a depletion of hydrophobic and bulky residues such as trytophan and phenylalanine (Fig.~\ref{fig:substitution}A).
-Overall, the exchangeability matrices fit to the disorder and order regions have similar trends of high and low coefficients (Fig.~\ref{fig:substitution}B-C).
-However, the log ratios of the disorder to the order exchangeability coefficients show clear patterns within and between the disorder-enriched and -depleted residues.
+The exchangeability matrices fit to the disorder and order regions have similar overall patterns of high and low coefficients (Fig.~\ref{fig:substitution}B-C).
+However, the log ratios of the disorder to the order exchangeability coefficients show clear differences within and between the disorder-enriched and -depleted residues.
 In disorder regions, disorder-enriched residues are less exchangeable with each other, whereas disorder-depleted residues are more exchangeable with each other and with disorder-enriched residues (Fig.~\ref{fig:substitution}D).
-Likewise, the log ratios of the rate coefficient also exhibit a division between the disorder-enriched and -depleted residues.
-As the coefficients model the one-way rate of substitutions between residues, the vertical and horizontal axes indicate the initial and target residues, respectively.
-Thus, the matrix indicates a bias to evolving towards a more disorder-promoting composition.
+Likewise, we observe a trend in the log ratios of the rate coefficients where the coefficients above the diagonal are generally positive and those below the diagonal are generally negative.
+As the coefficients model the one-way rate of substitutions between residues with the vertical and horizontal axes indicating the initial and target residues, respectively, this suggests a net flux towards a more disorder-like composition.
 

### 2023-04-22 chapter2/chapter2.tex
Moved substitution figure between results paragraphs
Added introduction outline
Updated WIP analysis of scores outline
Added description of GO term analysis to results
Converted outline to paragraph and tweaked wording

@@
 Here's where my introduction will go!
+% I. IDRs, proteins without defined structures, are prevalent in eukaryotic proteins and play key roles in many processes
+% II. Sequence-function relationships are unclear in IDRs
+% III. Proteome-wide patterns of conserved features linked to function
+%    A. Past studies have demonstrated evidence of constraint in specific proteins for specific properties
+%    B. Additionally other studies have shown importance of short linear motifs in IDRs
+%        1. Can be identified as short segments of conservation in otherwise highly diverged alignments of IDRs
+%        2. However, these segments comprise a small fraction of disordered regions, so
+%    C. Recent study first to demonstrate patterns of selection called evolutionary signatures in IDRs across entire proteome using a set of 82 molecular features
+%    D.
+Furthermore, by clustering IDRs on by these evolutionary signatures, they identified groups of IDRs evolving under similar sets of constraints.
+Furthermore they identified functional associations in these clusters which can be used to generate hypotheses of function for previously uncharacterized proteins
+% IV. Unclear if these patterns are specific to yeast or representative of IDR evolution across the tree of life
+%    A. Expand analyses to another model organism
+%        1. Drosophila is key model for genetic and development research with abundant genomic resources across many lineages in the radiation
+%        2. Furthermore as multicellular organism with complex developmental pathways, the results would be particularly relevant for understanding the role of IDRs in complex signaling pathways
+%        3. Also relevant for human health!
+%    B. Results would have significant impact on interpretation
+%        1. Conserved distributed properties are a yeast phenomena
+%        2. Patterns of conservation are yeast phenomena
+%        3. Global taxonomy of disordered regions similar to Pfam
+% IV. Explicit phylogenetic models of trait evolution can reveal patterns of evolution with fewer assumptions
+%    A. Previous analyses compared observed distributions of features in observed IDRs to the distributions generated by simulated counterparts
+%    B. Differences were interpreted as evidence of selection on features
+%        1. However, evidence of deviation from the null model rather than evidence of constraint necessarily
+%        2. Null distribution generated by a model of evolution where the composition of sequences stochastically evolve towards a specific equilibrium distribution
+%        3.
+Previous studies have identified IDRs with varying compositional preferences, so a single equilibrium distribution is unlikely to globally describe observed patterns of sequence evolution in IDRs
+%    C. Thus in this work we explore use of various statistical approaches that directly measure rates of change and constraint of molecular features indepedent of specific model of sequence evolution
+%        1. Place these evolutionary analyses in a statistical phylogenetic framework
+%        2.
+Though these models have own assumptions, they are macroscopic descriptions of continuous trait evolution and therefore independent of any assumptions about evolution at the level sequences
+%        3. Still have their own assumptions of course, but more general and robust to deviations
+%        4. We view these approaches as complementary, each having different trade-offs between specificity and generality of testing specific hypotheses of disorder evolution
+%        5. We also explore hybrid approaches where the evolutionary rates obtained from phylogenetic models are used as a statistic
+% V. By applying these approaches to set of IDRs identified in Drosophila genomes, we observe...
 
@@
 
+\begin{figure}[h!]
+\includegraphics[width=\textwidth]{substitution/out/substitution.png}
+\centering
+\caption{\textbf{Amino acid substitution models fit to disorder and order regions.}
+\textbf{(A)} Amino acid frequencies of substitution models.
+Amino acid symbols are ordered by their enrichment in disorder regions, calculated as the disorder-to-order ratio of their frequencies.
+Error bars represent standard deviations over models fit to different meta-alignments ($n = 25$).
+\textbf{(B-C)} Exchangeability coefficients of disorder and order regions, respectively, averaged over meta-alignments.
+\textbf{(D)} $\log_{10}$ disorder-to-order ratios of exchangeability coefficients.
+\textbf{(E-F)} Rate coefficients of disorder and order regions, respectively, averaged over meta-alignments.
+The vertical and horizontal axes indicate the initial and target amino acids, respectively.
+\textbf{(G)} $\log_{10}$ disorder-to-order ratios of rate coefficients.}
+\label{fig:substitution}
+\end{figure}
+
 To highlight the differences in patterns of residue substitution between the disorder and order regions, the parameters in each model are directly compared in Fig.~\ref{fig:substitution}, beginning with the frequency vectors.
@@
 
+\subsection{Intrinsic disorder is poorly conserved in some proteins}
+Though the substitution models reveal specific patterns of evolution at the level of individual residues, the large amounts of sequence divergence between many orthologous IDRs implies their evolution is not well-described by fine-scale models of residue substitution.
+Given the growing evidence that IDRs are constrained to conserve distributed properties, we instead turned towards characterizing their evolution in terms of 82 disorder-associated ``molecular features'' obtained from the previous study of IDRs in the yeast proteome.
+However, before conducting an in-depth analysis of these features, we examined the disorder score traces in greater detail and were struck by the significant variability between species.
+For each residue in the input sequence, AUCPreD returns a score between 0 and 1 where higher values indicate higher confidence in a prediction of intrinsic disorder.
+In some alignments, the scores would vary by nearly this entire range at a given alignment position even when there was a relatively high level of sequence identity (Fig.~\ref{fig:scores}A).
+
+\begin{figure}[h!]
+\includegraphics[width=\textwidth]{scores/out/scores_07E3.png}
+\centering
+\caption{\textbf{Analyses of disorder scores.}
+\textbf{(A)} Example region in the alignment of the sequences in orthologous group 07E3 with their corresponding disorder scores.
+Higher scores indicate a higher probability of disorder.
+Disorder score traces are colored by the position of its associated species on the phylogenetic tree.
+\textbf{(B)} Correlations between disorder scores and feature contrasts in regions.
+Asterisks indicate statistically significant correlations as computed by permutation tests ($p < 0.001$).
+\textbf{(C-D)} Example scatter plots showing correlations plotted in panel B.
+\textbf{(E)} GO term analysis of regions with rapidly evolving disorder scores.
+Only terms where $p < 0.001$ are shown.}
+\label{fig:scores}
+\end{figure}
+
+To better understand the relationship of this variability to differences in the regions' biophysical properties, we sought to correlate the average disorder score of the segments in a region with their molecular features.
+However, because the sequences are evolutionarily related, they are not independent, making them unsuitable for direct use in many standard statistical procedures.
+We instead applied the method of contrasts, which takes differences between adjacent nodes in the phylogenetic tree relating the species to generate a set ``contrasts,'' to both the scores and the features.
+Under some general assumptions of the underlying evolutionary process, contrasts are independent and identically-distributed and are therefore appropriate for use in correlation analyses.
+The resulting feature contrasts have varying degrees of correlation with the score contrasts (Fig.~\ref{fig:scores}B-C).
+Some, like isopoint, are uncorrelated, but most are significantly, if weakly, correlated.
+In general, the strongest correlations are observed for features which have a direct biophysical relationship to the presence or absence of disorder, such as disorder\_fraction or hydrophobicity.
+Interestingly, the correlations with many motifs were statistically significant, though small in magnitude relative to the non-motif features.
+However, a more detailed analysis of this observation is presented in the discussion.
+To determine if regions with rapidly evolving disorder scores were associated with a particular function, process, or compartment, we then extracted the upper decile of their rate distribution and performed term enrichment analyses on their associated annotations (Fig.~\ref{sfig:scores_histogram}).
+The most significant terms are generally related to DNA repair or extracellular structure, which suggests these processes and components are enriched in proteins whose structural state is rapidly evolving (Fig.~\ref{fig:scores}E).
+
 % RESULTS
-% III. Brownian motion score analyses
-%     A. Motivation for use of phylogenetic models
-%         1. For each residue in the input sequence, AUCPreD returns a score between 0 and 1 where higher values indicate higher confidence in a prediction of structural disorder.
-%     B.
-When examining the disorder score traces in greater detail alongside the alignments, we were struck by the significant variability in some traces, even when there was a relatively high level of sequence identity.
-%     C. To better understand the sequence features underlying these changes in predicted disorder, we calculated the values for 82 disorder-associated features in each region
-%     D. Though this analysis is limited by the resolution of the regions, which varies from 30 to ??? residues, it indicates the prediction algorithm is sensitive to these features in bulk.
-%          1. Some more observations of the overall trends ...
-%     E. Interestingly, the correlations with many motifs were statistically significant, though small in magnitude relative to the non-motif features.
-%          1. Look into motif regexes and see if ones with highest correlations have any common features
-%     F. To determine if regions with rapidly evolving disorder scores were enriched in any processes, we extracted the upper decile of their rate distribution
-%          1. Many chromatin and DNA related features
-
 % IV. Brownian motion feature analyses
@@
 % V. BM-OU analyses
-%    A. While these analyses measure the rates of change of features after accounting the the phylogenetic relationships between sequences, they do not directly test for conservation of features
+%    A. While the the BM model can the rates of change of features after accounting the the phylogenetic relationships between sequences, they do not directly test for conservation of features
 %         1. In fact, under the BM model, trait variation is unconstrained and will increase without bound over time
@@
 %         - Embeddings from LLMs trained on large data sets
-
-\begin{figure}[h!]
-\includegraphics[width=\textwidth]{substitution/out/substitution.png}
-\centering
-\caption{\textbf{Amino acid substitution models fit to disorder and order regions.}
-\textbf{(A)} Amino acid frequencies of substitution models.
-Amino acid symbols are ordered by their enrichment in disorder regions, calculated as the disorder-to-order ratio of their frequencies.
-Error bars represent standard deviations over models fit to different meta-alignments ($n = 25$).
-\textbf{(B-C)} Exchangeability coefficients of disorder and order regions, respectively, averaged over meta-alignments.
-\textbf{(D)} $\log_{10}$ disorder-to-order ratios of exchangeability coefficients.
-\textbf{(E-F)} Rate coefficients of disorder and order regions, respectively, averaged over meta-alignments.
-The vertical and horizontal axes indicate the initial and target amino acids, respectively.
-\textbf{(G)} $\log_{10}$ disorder-to-order ratios of rate coefficients.}
-\label{fig:substitution}
-\end{figure}
-
-\begin{figure}[h!]
-\includegraphics[width=\textwidth]{scores/out/scores_07E3.png}
-\centering
-\caption{\textbf{Analyses of disorder scores.}
-\textbf{(A)} Example region in the alignment of the sequences in orthologous group 07E3 with their corresponding disorder scores.
-Higher scores indicate a higher probability of disorder.
-Disorder score traces are colored by the position of its associated species on the phylogenetic tree.
-\textbf{(B)} Correlations between disorder scores and feature contrasts in regions.
-Asterisks indicate statistically significant correlations as computed by permutation tests ($p < 0.001$).
-\textbf{(C-D)} Example scatter plots showing correlations plotted in panel B.
-\textbf{(E)} GO term analysis of regions with rapidly evolving disorder scores.
-Only terms where $p < 0.001$ are shown.}
-\label{fig:scores}
-\end{figure}
 

### 2023-04-22 chapter2/chapter2.tex
Changed subsection title to use IDRs and not disorder regions

@@
 
-\subsection{Disorder regions have distinct patterns of residue substitution}
+\subsection{IDRs have distinct patterns of residue substitution}
 To gain insight into the substitution patterns of amino acid residues in disorder and order regions, we fit substitution models to meta-alignments sampled from the respective sets of filtered regions.

### 2023-04-22 chapter2/chapter2.tex
Re-worded motivation for use of contrasts

@@
 To better understand the relationship of this variability to differences in the regions' biophysical properties, we sought to correlate the average disorder score of the segments in a region with their molecular features.
-However, because the sequences are evolutionarily related, they are not independent, making them unsuitable for direct use in many standard statistical procedures.
-We instead applied the method of contrasts, which takes differences between adjacent nodes in the phylogenetic tree relating the species to generate a set ``contrasts,'' to both the scores and the features.
+However, as the sequences are not independent but instead related by a hierarchical structure which reflects their evolutionary relationships, any derived features are unsuitable for direct use in many standard statistical procedures.
+In the most severe cases, traits derived from clades of closely-related species can effectively act as duplicate observations, which can yield spurious correlations.
+We therefore applied the method of contrasts, which takes differences between adjacent nodes in the phylogenetic tree relating the species to generate a set of ``contrasts,'' to both the scores and the features.
 Under some general assumptions of the underlying evolutionary process, contrasts are independent and identically-distributed and are therefore appropriate for use in correlation analyses.

### 2023-04-23 chapter2/chapter2.tex
Updated introduction outline
Finished BM analysis paragraph

@@
 % III. Proteome-wide patterns of conserved features linked to function
-%    A. Past studies have demonstrated evidence of constraint in specific proteins for specific properties
+%    A. However, there is growing evidence that IDRs are instead constrained to conserve distributed properties such as ...
+%       1. Because many sequences can yield similar a region with a specific charge or length, IDRs can rapidly diverge in terms of their sequences while still remaining
+%       2.
+Additionally, in contrast to folded regions which must make precise contacts or satisfy specific geometric constraints, distributed features are robust to small changes at the sequence level.
+A mutation that slightly impairs function related to a specific feature regime is easily corrected by compensatory mutation elsewhere in the region.
+Together, these loose restrictions on the sequences of IDRs allow them to rapidly diverge despite preserving their functions.
 %    B. Additionally other studies have shown importance of short linear motifs in IDRs
@@
 %        2. However, these segments comprise a small fraction of disordered regions, so
-%    C. Recent study first to demonstrate patterns of selection called evolutionary signatures in IDRs across entire proteome using a set of 82 molecular features
+%    C. Recent study first to demonstrate patterns of selection in IDRs across an entire proteome using a set of 82 ``molecular features''
 %    D.
-Furthermore, by clustering IDRs on by these evolutionary signatures, they identified groups of IDRs evolving under similar sets of constraints.
+Furthermore, by clustering IDRs on by these ``evolutionary signatures,'' they identified groups of IDRs evolving under similar sets of constraints.
 Furthermore they identified functional associations in these clusters which can be used to generate hypotheses of function for previously uncharacterized proteins
-% IV. Unclear if these patterns are specific to yeast or representative of IDR evolution across the tree of life
+% IV. However, unclear if these patterns are specific to yeast or representative of IDR evolution across the tree of life
 %    A. Expand analyses to another model organism
@@
 
+\subsection{IDRs have three axes of unconstrained variation}
+Having calculated the features associated with the sequence segments composing each region in our data set, we then sought to determine if their distributions contained any global structure which would enable us to identify classes with distinct biophysical or functional properties.
+These distributions are generated by a complex underlying evolutionary process which reflects the combined effects of selection, drift, and mutation.
+However, to leverage a statistical framework to infer the properties of this process, we fit Brownian motion (BM) models to the features calculated from the segments in each region.
+BM is a simple model of evolution where continuous traits change through a series of small, undirected steps.
+Thus, the traits accumulate variation at a constant rate over time but do not on average deviate from their original values.
+BM models are therefore specified by two parameters: a rate, which describes the speed at which trait variation accumulates, and a root, which describes the ancestral trait value.
+
+We then applied principal components analyses (PCAs) to visualize the major axes of variation of the root and rate parameters for each feature and region.
+A difficulty with a direct analysis of the parameter estimates, however, is PCAs are sensitive to differences in scaling between variables, and some features have dramatically different intrinsic scales.
+For example, many compositional features, like fraction\_S, are restricted to the interval [0, 1] whereas SCD is unbounded and can vary from negative to positive infinity.
+As a result, SCD is responsible for a significant fraction of the overall variance in both parameter distributions (Fig.~\ref{sfig:root_variance}-\ref{sfig:rate_variance}).
+Therefore, we first transformed the parameters associated with each feature into z-scores relative to the proteome-wide distributions.
+
+The first two principal components of the root distributions show little overall structure, though there is a slight enrichment of regions along two axes that correlate with acidic and polar features, respectively (Fig.~\ref{fig:root}).
+Likewise, the projections of the rates onto the first two components are largely distributed along the first principal component (Fig.~\ref{fig:rate}A).
+This and the variable amounts of sequence divergence in the regions led us to suspect the first principal component was a measure of the overall rate of sequence evolution.
+Plotting the first principal component against the sum of the average amino acid and indel rates as measured by substitution models revealed a strong association (Fig.~\ref{fig:rate}B).
+We then projected the rates along second and third principal components to determine if these higher order components contained any additional structure.
+The resulting distribution is roughly triangular and contains three major axes of variation, corresponding to rapid changes in the regions' proportions of acidic, glutamine, and glycine residues (Fig.~\ref{fig:rate}C-D).
+Inspection of regions selected along these axes confirmed the high rates of evolution of these features (Fig.~\ref{fig:rate}E-G).
+
 % RESULTS
-% IV. Brownian motion feature analyses
-%     A.
-Having calculated features for all IDRs in our data set, we were curious the distribution contained any global structure which would allow the identification of classes with distinct biophysical or functional properties
-%     A. Feature roots
-%         1. Discuss differences in scaling and need for z-scores
-%         2. Discussion of sample regions in figure? (need to add)
-%     B.
-The phylogenetic contrasts algorithm also permits an estimate of the rate of feature evolution, which is effectively the variance of the distribution of features across all sequences after correcting for the phylogenetic relationships between them
-%         1. PC1 correlates with rate
-%         2. PC2 and PC3 show three major axes of unconstrained variation: Q-rich, G-rich, acidic
-
 % V. BM-OU analyses

### 2023-04-24 chapter2/chapter2.tex
Moved rate and root figures by corresponding paragraphs in results
Added more points about scores analysis to conclusion outline
Wrote OU analysis introduction
Converted OU results to paragraphs and updated references

@@
 
+\begin{figure}[h!]
+\includegraphics[width=\textwidth]{brownian/out/root.png}
+\centering
+\caption{\textbf{PCA of disorder regions' feature roots.}
+\textbf{(A)} The first two PCs of the disorder regions' feature root distributions.
+The explained variance percentage of each component is indicated in parentheses in the axis labels.
+\textbf{(B)} The same plot as panel A with the projections of original features onto the components shown as arrows.
+Only the 16 features with the largest projections are shown.
+Scaling of the arrows is arbitrary.}
+\label{fig:root}
+\end{figure}
+
+\begin{figure}[h!]
+\includegraphics[width=\textwidth]{brownian/out/rate.png}
+\centering
+\caption{\textbf{PCA of disorder regions' feature rates.}
+\textbf{(A)} The first two PCs of the disorder regions' feature rate distributions.
+The explained variance percentage of each component is indicated in parentheses.
+\textbf{(B)} Scatter plot of the disorder regions' feature rates along the first PC against the sum of the average amino acid and indel substitution rates in those regions.
+\textbf{(C)} The second and third PCs of the disorder regions' feature rate distributions.
+The explained variance percentage of each component is indicated in parentheses in the axis labels.
+\textbf{(D)} The same plot as panel C with the projections of original rates onto the components shown as arrows.
+Only the 16 features with the largest projections are shown.
+Scaling of the arrows is arbitrary.
+\textbf{(E-F)} Example alignments of disorder regions from the orthologous groups 0A8A, 3139, 04B0, respectively.
+The colored bars on the left indicate the hexbin containing that region in panel C.}
+\label{fig:rate}
+\end{figure}
+
 The first two principal components of the root distributions show little overall structure, though there is a slight enrichment of regions along two axes that correlate with acidic and polar features, respectively (Fig.~\ref{fig:root}).
@@
 
-% RESULTS
-% V. BM-OU analyses
-%    A. While the the BM model can the rates of change of features after accounting the the phylogenetic relationships between sequences, they do not directly test for conservation of features
-%         1. In fact, under the BM model, trait variation is unconstrained and will increase without bound over time
-%         2. Instead, demonstrating evidence of conservation requires a comparison to a model where feature evolution is constrained.
-%         3. A common choice is the Ornstein-Uhlenbeck (OU) model
-%             i. Some explanation about how it works
+\subsection{A model of constrained evolution reveals conserved features}
+Though the BM process permits the inference of the rates of feature evolution after accounting for the phylogenetic relationships between species, it does not directly test for their conservation.
+In fact, under the BM model, trait variation is unconstrained and will increase without bound over time.
+Instead, evidence of conservation requires comparison to a model where trait variation is constrained.
+A common choice for modeling the effect of selection on the evolution of a continuous trait is the Ornstein-Uhlenbeck (OU) model.
+The OU model is similar to the BM model where a trait accumulates variation through small, undirected steps.
+However, it differs in the trait is also attracted towards an optimal value where the attraction is proportional to a parameter $\alpha$ and the trait's distance from this value.
+Under an additional assumption of stationarity that ensures parameter identifiability and estimate consistency, the OU model is specified with three total parameters: the optimal value, the fluctuation magnitude, and the selection strength~\cite{Ho2013, Ho2014}.
+While the first two parameters are analogous to the root and rate parameters in the BM model, respectively, the selection strength has no equivalent.
+
+We therefore fit OU models to the features calculated from the segments in each region to detect evidence of their conservation.
+A strength of probabilistic models of trait evolution is they permit the use of standard statistical methods to compare the support for different hypotheses.
+Thus, the goodness of fit of the OU model relative to the BM model is a measure of the evidence for selective constraint.
+However, because the OU model is specified with an additional parameter and has a greater capacity to fit the data, we compared their fits using the difference in the Akaike information criterion (AIC), which accounts for the model complexity, rather than the likelihood ratio.
+We chose not to treat these differences in a strict statistical sense but instead as qualitative ``evolutionary signatures'' as in Zarin \textit{et al.}~\cite{Zarin2019}.
+We therefore clustered the signatures to identify patterns of feature conservation across the proteome.
+To ensure the clusters were enriched in regions with a high likelihood of feature conservation despite low levels of sequence identity, we restricted this analysis to regions with a minimum amount of divergence as measured by substitution models (Fig.~\ref{sfig:hierarchy_histogram}).
 
@@
 This indicates that these models can confuse correlation and causation during training, and their representations should be interpreted carefully.
-%        1. Furthermore, resolution of analyses limited because features calculated at level of regions rather than a fixed window around each residue.
+%        4. Furthermore, resolution of analyses limited because features calculated at level of regions, whose lengths can exceed 1,000 residues, rather than a fixed window around each residue.
+%        5.
+Disorder scores are weakly correlated with many motifs.
+Furthermore, unclear if predictor is responding directly to these motifs as a signal of disorder, or they appear in regions which evolving towards disorder.
+Interestingly docking (DOC) and ligand binding (LIG) sites tend to have negative correlations whereas the modification (MOD) sites have positive correlations.
+Docking and ligand binding sites are generally mediated by small hydrophobic patches, so the predictor could be responding to increased hydrophobicity caused by additional of hydrophobic residue ``completing'' the motif rather than the motif itself.
+Likewise, phosphorylation sites are highly enriched in IDRs and many are targeted to disorder-promoting residues like serine or threonine.
+Thus, it's unclear if the predictor is responding directly to these features or features which are correlated with them.
+More in depths of the representations of these disorder predictors would shed led on these questions.
 %    B. GO annotation enrichment analysis suggests disorder scores undergo rapid evolution in proteins involved in DNA repair and found in extracellular structure
@@
 %         - Embeddings from LLMs trained on large data sets
-
-\begin{figure}[h!]
-\includegraphics[width=\textwidth]{brownian/out/root.png}
-\centering
-\caption{\textbf{PCA of disorder regions' feature roots.}
-\textbf{(A)} The first two PCs of the disorder regions' feature root distributions.
-The explained variance percentage of each component is indicated in parentheses in the axis labels.
-\textbf{(B)} The same plot as panel A with the projections of original features onto the components shown as arrows.
-Only the 16 features with the largest projections are shown.
-Scaling of the arrows is arbitrary.}
-\label{fig:root}
-\end{figure}
-
-\begin{figure}[h!]
-\includegraphics[width=\textwidth]{brownian/out/rate.png}
-\centering
-\caption{\textbf{PCA of disorder regions' feature rates.}
-\textbf{(A)} The first two PCs of the disorder regions' feature rate distributions.
-The explained variance percentage of each component is indicated in parentheses.
-\textbf{(B)} Scatter plot of the disorder regions' feature rates along the first PC against the sum of the average amino acid and indel substitution rates in those regions.
-\textbf{(C)} The second and third PCs of the disorder regions' feature rate distributions.
-The explained variance percentage of each component is indicated in parentheses in the axis labels.
-\textbf{(D)} The same plot as panel C with the projections of original rates onto the components shown as arrows.
-Only the 16 features with the largest projections are shown.
-Scaling of the arrows is arbitrary.
-\textbf{(E-F)} Example alignments of disorder regions from the orthologous groups 0A8A, 3139, 04B0, respectively.
-The colored bars on the left indicate the hexbin containing that region in panel C.}
-\label{fig:rate}
-\end{figure}
 
@@
 The OU model parameters were also calculated via maximum likelihood estimation.
-To avoid issues with model identifiability, the root was treated as a random variable~\cite{Ho2014}.
-Thus, the covariance matrix, $V$, was parameterized as $V_{ij} = e^{\alpha d_{ij}}$ where $d_{ij}$ is the tree distance between tips $i$ and $j$, and $\alpha$ is the selection strength~\cite{An2008}.
+To ensure parameter identfiability and estimate consistency, the root was treated as a random variable~\cite{Ho2013, Ho2014}.
+Thus, the covariance matrix, $V$, was parameterized as $V_{ij} = e^{\alpha d_{ij}}$ where $d_{ij}$ is the tree distance between tips $i$ and $j$, and $\alpha$ is the selection strength~\cite{Ho2013}.
 
@@
 Clusters were manually chosen for subsequent GO analyses.
-To enrich these clusters for regions with a high likelihood of feature conservation despite high levels of sequence divergence, this analysis was restricted to regions whose amino acid and indel substitution rates exceeded 1 and 0.1, respectively (Fig.~\ref{sfig:hierarchy_histogram}).
+To enrich these clusters for regions with a high likelihood of feature conservation despite low levels of sequence identity, this analysis was restricted to regions whose amino acid and indel substitution rates exceeded 1 and 0.1, respectively (Fig.~\ref{sfig:hierarchy_histogram}).
 

### 2023-04-24 chapter2/figures/hierarchy/hierarchy.py
Updated tree axis to dynamically match heatmap

@@ -137,7 +137,8 @@ fig, axs = plt.subplots(2, 2, figsize=(7.5, 7.5), gridspec_kw=gridspec_kw)
 # Tree
 ax = axs[0, 0]
 plot_tree(tree, ax=ax, linecolor=node2color, linewidth=0.2, tip_labels=False,
-          xmin_pad=0.025, xmax_pad=0, ymin_pad=1 / (2 * len(array)), ymax_pad=1 / (2 * len(array)))
+          xmin_pad=0.025, xmax_pad=0)
+ax.sharey(axs[0, 1])
 ax.set_ylabel('Disorder regions')
 ax.set_axis_off()
 

### 2023-04-25 chapter2/figures/hierarchy/hierarchy.py
Added conversion to int of tip.name to reflect changes in src2/

@@ -122,7 +122,7 @@ lm = linkage(cdm, method='average')
 
 # Convert to tree and get branch colors
 tree = make_tree(lm)
-tip_order = [tip.name for tip in tree.tips()]
+tip_order = [int(tip.name) for tip in tree.tips()]
 node2color, node2tips = {}, {}
 for node in tree.postorder():
     if node.is_tip():

### 2023-04-25 chapter2/figures/hierarchy/hierarchy.py
Fixed hiding axis labels

@@ -140,14 +140,20 @@ plot_tree(tree, ax=ax, linecolor=node2color, linewidth=0.2, tip_labels=False,
           xmin_pad=0.025, xmax_pad=0)
 ax.sharey(axs[0, 1])
 ax.set_ylabel('Disorder regions')
-ax.set_axis_off()
+ax.set_xticks([])
+ax.set_yticks([])
+for spine in ax.spines.values():
+    spine.set_visible(False)
 
 # Heatmap
 ax = axs[0, 1]
 im = ax.imshow(array[tip_order], aspect='auto', cmap=plt.colormaps['inferno'], interpolation='none')
 ax.xaxis.set_label_position('top')
 ax.set_xlabel('Features')
-ax.set_axis_off()
+ax.set_xticks([])
+ax.set_yticks([])
+for spine in ax.spines.values():
+    spine.set_visible(False)
 
 # Corner axis
 ax = axs[1, 0]

### 2023-04-25 chapter2/figures/hierarchy/hierarchy.py
Split creation of clustering array into two steps to allow access to corresponding dataframe

@@ -115,7 +115,8 @@ gridspec_kw = {'width_ratios': [0.1, 0.9], 'wspace': 0,
 column_labels = []
 for group_label in group_labels:
     column_labels.extend([f'{feature_label}_delta_AIC' for feature_label in feature_groups[group_label]])
-array = np.nan_to_num(models.loc[pdidx[:, :, :, True], column_labels].to_numpy(), nan=1)  # Re-arrange and convert to array
+data = models.loc[pdidx[:, :, :, True], column_labels]  # Re-arrange columns
+array = np.nan_to_num(data.to_numpy(), nan=1)
 
 cdm = pdist(array, metric='correlation')
 lm = linkage(cdm, method='average')

### 2023-04-25 chapter2/chapter2.tex
Clarified z-score is a normalization procedure
Added number of regions for OU analysis to methods

@@
 As a result, SCD is responsible for a significant fraction of the overall variance in both parameter distributions (Fig.~\ref{sfig:root_variance}-\ref{sfig:rate_variance}).
-Therefore, we first transformed the parameters associated with each feature into z-scores relative to the proteome-wide distributions.
+Therefore, we first normalized the parameters associated with each feature by transforming them into z-scores relative to their proteome-wide distributions.
 
@@
 Clusters were manually chosen for subsequent GO analyses.
-To enrich these clusters for regions with a high likelihood of feature conservation despite low levels of sequence identity, this analysis was restricted to regions whose amino acid and indel substitution rates exceeded 1 and 0.1, respectively (Fig.~\ref{sfig:hierarchy_histogram}).
+To enrich these clusters for regions with a high likelihood of feature conservation despite low levels of sequence identity, this analysis was restricted to 7,645 regions whose amino acid and indel substitution rates exceeded 1 and 0.1, respectively (Fig.~\ref{sfig:hierarchy_histogram}).
 

### 2023-04-26 chapter2/chapter2.tex
Added to introduction outline including draft of first paragraph

@@
 \section{Introduction}
-Here's where my introduction will go!
-% I. IDRs, proteins without defined structures, are prevalent in eukaryotic proteins and play key roles in many processes
-% II. Sequence-function relationships are unclear in IDRs
-% III. Proteome-wide patterns of conserved features linked to function
-%    A. However, there is growing evidence that IDRs are instead constrained to conserve distributed properties such as ...
-%       1. Because many sequences can yield similar a region with a specific charge or length, IDRs can rapidly diverge in terms of their sequences while still remaining
+Intrinsically disordered regions (IDRs) are segments of proteins which lack stable three-dimensional structures and instead exist as ensembles of rapidly interconverting conformations.
+As a result of this structural hetereogeneity, IDRs can interact with diverse binding partners.
+Often these interactions have high specificity but moderate affinity, which permits the efficient propagation of signals by rapid binding and dissociation.
+Furthermore, as IDRs readily expose their polypeptide chains, they are enriched in motifs for post-translational modifications which allow environmental or physiological conditions to modulate their interactions.
+Accordingly, IDRs often act as the ``hubs'' of complex signaling networks by integrating signals from diverse pathways and coordinating interactions.
+However, as IDRs are pervasive in eukaryotic proteomes, with an estimated X\% of proteins containing a disordered region of at least 30 residues, they are involved in diverse processes and phenomena, including EXAMPLE, EXAMPLE, and the formation of biomolecular condensates.
+
+% II. Proteome-wide patterns of conserved features linked to function
+%    B. Despite the prevalence and importance of these regions in eukaryotic proteomes, associating them with specific molecular functions or biological processes remains a significant challenge.
+%    C.
+The sequences of IDRs are generally poorly conserved, so traditional bioinformatics approaches which rely on the strict conservation of amino acid sequences to identify homologous proteins and transfer annotations between biological systems are largely unsucessful when applied to IDRs.
+%
+%    A.
+However, several recent studies have demonstrated evidence that IDRs are constrained to preserve ``distributed features'' such as flexibility, chemical composition, net charge, or charge distribution.
+%       1.
+Because many sequences can yield a region with a specific composition or charge, this mode of constraint uncouples an IDR's fitness from its strict sequence of amino acid residues.
+the sequences of IDRs can rapidly diverge while still preserving these structural or functional properties.
 %       2.
-Additionally, in contrast to folded regions which must make precise contacts or satisfy specific geometric constraints, distributed features are robust to small changes at the sequence level.
-A mutation that slightly impairs function related to a specific feature regime is easily corrected by compensatory mutation elsewhere in the region.
+Furthermore, in contrast to folded regions whose precise contacts and packing geometries are easily disrupted by point mutations, distributed features are robust to small changes at the sequence level.
+Thus, deleterious mutations are easily corrected by compensatory mutation elsewhere in the region.
 Together, these loose restrictions on the sequences of IDRs allow them to rapidly diverge despite preserving their functions.
@@
 %        2. However, these segments comprise a small fraction of disordered regions, so
-%    C. Recent study first to demonstrate patterns of selection in IDRs across an entire proteome using a set of 82 ``molecular features''
+%
+% III.
+Previously, many of these studies were limited to ...
+However, a recent study first to demonstrate patterns of selection in IDRs across an entire proteome using a set of 82 ``molecular features''
 %    D.
@@
 Furthermore they identified functional associations in these clusters which can be used to generate hypotheses of function for previously uncharacterized proteins
-% IV. However, unclear if these patterns are specific to yeast or representative of IDR evolution across the tree of life
+%    E.
+Subsequent studies have expanded on these findings by using techniques from machine learning and statistics the information from evolutionary signatures to predict the function of IDRs.
+and learn disorder-associated features in a self-supervised manner.
+% III.
+However, no subsequent studies have determined if similar patterns of conservation are found in other systems.
+As another foundational model organism with abundant genomic information across many evolutionary lineages, the fruit fly, \textit{Drosophila melanogaster}, is a natural choice for subsequent investigation~\cite{Yang2018, Miller2018, Kim2021, Gramates2022}.
 %    A. Expand analyses to another model organism
@@
 %        3. Also relevant for human health!
-%    B. Results would have significant impact on interpretation
-%        1. Conserved distributed properties are a yeast phenomena
+%    B.
+The concordance of these results with the previously identified IDR clusters would also have profound implications for the broader mechanisms of IDR evolution.
+Depending on the amount of concordance, several conclusions are possible.
+%        1. If no patterns are observed, this suggests that conservation of distributed is a not widely conserved mechanism of IDR evolution.
 %        2. Patterns of conservation are yeast phenomena
-%        3. Global taxonomy of disordered regions similar to Pfam
+%        3.
+Finally, the identification of similar clusters with conserved features and properties would suggest the existence of ``taxonomy'' of IDRs which is common across many branches of the tree of life.
+This would allow the creation of a classification system such as Pfam
 % IV. Explicit phylogenetic models of trait evolution can reveal patterns of evolution with fewer assumptions
@@
 %    C. Thus in this work we explore use of various statistical approaches that directly measure rates of change and constraint of molecular features indepedent of specific model of sequence evolution
-%        1. Place these evolutionary analyses in a statistical phylogenetic framework
+%        1. Place these evolutionary analyses in a statistical phylogenetic framework through the application of the Brownian motion and Ornstein-Uhlenbeck models of trait evolution.
 %        2.
@@
 %        5. We also explore hybrid approaches where the evolutionary rates obtained from phylogenetic models are used as a statistic
-% V. By applying these approaches to set of IDRs identified in Drosophila genomes, we observe...
+% V.
+Therefore, in this work we apply evolutionary models of sequence and trait evolution to the sequences of IDRs and their of associated molecular feature to dissect the forces driving their evolution.
+We find IDRs exhibit unique patterns of evolution at the level of sequences.
+We find intrinsic disorder is a dynamically evolving property in some regions, and changes in disorder preference are correlated with changes in the complexity.
+We find that IDRs are unconstrained along several axis of trait evolution, but they also display
 

### 2023-04-26 chapter0/chapter0.tex
Added a manual word break point
Added citations to chapter 2 introduction paragraph 1

@@
 
-Unsurprisingly, disordered regions are characterized by a relative depletion of hydrophobic residues~\cite{Uversky2000}.
+Unsurprisingly, disordered regions are characterized by a relative depletion of hydrophobic resi\-dues~\cite{Uversky2000}.
 However, there is no simple formula which accurately predicts disorder in a protein using only the hydrophobicity values of its constituent residues.
@@
 As their flexibility makes SLiMs easily accessible, they are also enriched in IDRs, which in turn allows them to bind to many partners, sometimes simultaneously.
-IDRs therefore often act as hubs in complex regulatory networks by propagating signals from diverse sources or organizing binding partners into higher-order structures~\cite{Dunker2005, Wright2014}.
+IDRs therefore often act as hubs in complex regulatory networks by propagating signals from diverse sources or organizing binding partners into higher-order structures~\cite{Dunker2005, Buljan2012, Wright2014}.
 Furthermore, these signals and interactions are easily tuned because PTMs can modulate their meanings and strengths, respectively, by modifying only a few residues.

### 2023-04-26 chapter2/chapter2.tex
Added citations to chapter 2 introduction paragraph 1

@@
 As a result of this structural hetereogeneity, IDRs can interact with diverse binding partners.
-Often these interactions have high specificity but moderate affinity, which permits the efficient propagation of signals by rapid binding and dissociation.
-Furthermore, as IDRs readily expose their polypeptide chains, they are enriched in motifs for post-translational modifications which allow environmental or physiological conditions to modulate their interactions.
-Accordingly, IDRs often act as the ``hubs'' of complex signaling networks by integrating signals from diverse pathways and coordinating interactions.
-However, as IDRs are pervasive in eukaryotic proteomes, with an estimated X\% of proteins containing a disordered region of at least 30 residues, they are involved in diverse processes and phenomena, including EXAMPLE, EXAMPLE, and the formation of biomolecular condensates.
+Often these interactions have high specificity but moderate affinity, which permits the efficient propagation of signals by rapid binding and dissociation~\cite{Zhou2012, Wright2014}.
+Furthermore, as IDRs readily expose their polypeptide chains, they are enriched in recognition motifs for post-translational modifications which allow environmental or physiological conditions to modulate their interactions.
+Accordingly, IDRs often act as the ``hubs'' of complex signaling networks by integrating signals from diverse pathways and coordinating interactions~\cite{Dunker2005, Buljan2012}.
+However, as IDRs are pervasive in eukaryotic proteomes, with estimates of the fractions of disordered residues in the human, mouse, and fruit fly proteomes ranging between 22 and 24\%, they are involved in diverse functions~\cite{Piovesan2020, Piovesan2022}.
+These including EXAMPLE, EXAMPLE, and the formation of biomolecular condensates.
 

### 2023-04-26 chapter0/chapter0.tex
Moved citations closer to the relevant clauses in chapter 0

@@
 Thus, red blood cells are packed with a special protein called hemoglobin, which binds oxygen.\footnote{In biology, binding is a slippery word whose exact meaning can varying greatly depending on the context.
-However, it generally means that a molecule physically interacts with another molecular for an extended period.} Each red blood cell contains as many as 270 million molecules of hemoglobin, each of which can carry up to four oxygen molecules~\cite{Pierig2008}.
-Because red blood cells are so dense with hemoglobin, composing roughly 35\% of their total volume, any defect in hemoglobin can dramatically impact the structure of the red blood cells themselves~\cite{Kanias2009}.
+However, it generally means that a molecule physically interacts with another molecular for an extended period.} Each red blood cell contains as many as 270 million molecules of hemoglobin~\cite{Pierig2008}, each of which can carry up to four oxygen molecules.
+Because red blood cells are so dense with hemoglobin, composing roughly 35\% of their total volume~\cite{Kanias2009}, any defect in hemoglobin can dramatically impact the structure of the red blood cells themselves.
 A well-studied example is sickle cell disease where an error in the body's hemoglobin molecules deforms red blood cells into a characteristic sickle shape.
@@
 
-Despite recent advances in identifying IDRs and their conformational ensembles from their sequences alone, the relationship between the sequence and function remains poorly understood~\cite{Mao2010, vanderLee2014, Das2015, Necci2021}.
+Despite recent advances in identifying IDRs and their conformational ensembles from their sequences alone~\cite{Mao2010, vanderLee2014, Das2015, Necci2021}, the relationship between the sequence and function remains poorly understood.
 In contrast, predictions of structure and function are readily available for many folded domains.
@@
 Thus, rather than conserving specific sequences, IDRs are hypothesized to conserve distributed ``molecular features'' associated with those sequences.
-By the mid 2010s several studies had demonstrated evidence of such constraint in the flexibility, chemical composition, net charge, or charge distribution of IDRs~\cite{Daughdrill2007, Moesa2012, Zarin2017, Beh2012}.
+By the mid 2010s several studies had demonstrated evidence of such constraint in the flexibility~\cite{Daughdrill2007}, chemical composition~\cite{Moesa2012}, net charge~\cite{Zarin2017}, or charge distribution of IDRs~\cite{Beh2012}.
 While these earlier studies were generally restricted to specific features or proteins, in 2019 Zarin \textit{et al.} demonstrated conservation of various features from a comprehensive set of IDR-associated properties in IDRs across the entire yeast proteome~\cite{Zarin2019}.
@@
 However, no known subsequent studies have determined if similar patterns of conservation are found in the IDRs of other systems.
-As another foundational model organism with abundant genomic information across many evolutionary lineages, the fruit fly, \textit{Drosophila melanogaster}, is a natural choice for subsequent investigation~\cite{Yang2018, Miller2018, Kim2021, Gramates2022}.
+As another foundational model organism with abundant genomic information across many evolutionary lineages~\cite{Yang2018, Miller2018, Kim2021, Gramates2022}, the fruit fly, \textit{Drosophila melanogaster}, is a natural choice for subsequent investigation.
 Furthermore, given its complex multicellular development process and shared signaling pathways with humans, the findings of such a study would significantly advance our understanding of the role of IDRs in human health and disease.

### 2023-04-26 chapter2/chapter2.tex
Added examples to chapter 2 first paragraph

@@
 Accordingly, IDRs often act as the ``hubs'' of complex signaling networks by integrating signals from diverse pathways and coordinating interactions~\cite{Dunker2005, Buljan2012}.
-However, as IDRs are pervasive in eukaryotic proteomes, with estimates of the fractions of disordered residues in the human, mouse, and fruit fly proteomes ranging between 22 and 24\%, they are involved in diverse functions~\cite{Piovesan2020, Piovesan2022}.
-These including EXAMPLE, EXAMPLE, and the formation of biomolecular condensates.
+However, as IDRs are pervasive in eukaryotic proteomes, with estimates of the fractions of disordered residues in the human, mouse, and fruit fly proteomes ranging between 22 and 24\%~\cite{Piovesan2020, Piovesan2022}, they are involved in diverse processes~\cite{vanderLee2014} including transcriptional regulation~\cite{Liu2006} and the formation of biomolecular condensates~\cite{Banani2017}.
 

### 2023-04-26 chapter1/chapter1.tex
Corrected typo

@@
 One early method that remains popular is building clusters progressively by identifying nodes that form a triangle with at least two other nodes in the cluster~\cite{Tatusov1997, Jensen2007}.
-Other approaches include the MCL algorithm, which clusters graphs by similating stochastic flow, and connected components or other single-linkage criteria~\cite{Remm2001, Enright2002, Li2003, Emms2015, Train2017, Cosentino2018}.
+Other approaches include the MCL algorithm, which clusters graphs by simulating stochastic flow, and connected components or other single-linkage criteria~\cite{Remm2001, Enright2002, Li2003, Emms2015, Train2017, Cosentino2018}.
 While these methods are robust when clustering hit graphs derived from smaller or more diverse sets of genomes, they are not suitable for the large number of closely related genomes in this work since they require relatively few edges to define a cluster.

### 2023-04-26 chapter1/chapter1.tex
Moved citations closer to relevant phrases

@@
 Graph-based approaches have remained popular, and in the intervening years many other researchers have refined this method by implementing various pre- and post-processing steps.
-Despite these improvements, many databases and pipelines use the same triangle clustering algorithm or other methods which require relatively few hits between sequences to infer an orthologous group, \textit{e.g.} connected components or Markov clustering~\cite{Remm2001, Enright2002, Li2003, Jensen2007, Linard2011, Emms2015, Train2017, Cosentino2018}.
+Despite these improvements, many databases and pipelines use the same triangle clustering algorithm~\cite{Jensen2007} or other methods which require relatively few hits between sequences to infer an orthologous group, \textit{e.g.} connected components and other single-linkage criteria~\cite{Remm2001, Train2017, Linard2011, Cosentino2018} or Markov clustering~\cite{Enright2002, Li2003, Emms2015}.
 However, the scale of biological sequence data has changed dramatically.
 For example, in the last decade, the number of annotated genomes available from NCBI has increased nearly 20-fold and currently exceeds 900 (Fig.~\ref{sfig:ncbi}).
-Though this figure is only a rough proxy of the total number of assemblies available, it will likely continue to grow rapidly in the coming years as many large-scale genome assembly efforts such as i5K, the Bird 10,000 Genomes Project, and the Vertebrate Genomes Project have already yielded results~\cite{Thomas2020, Feng2020, Rhie2021}.
+Though this figure is only a rough proxy of the total number of assemblies available, it will likely continue to grow rapidly in the coming years as many large-scale genome assembly efforts such as i5K~\cite{Thomas2020}, the Bird 10,000 Genomes Project~\cite{Feng2020}, and the Vertebrate Genomes Project~\cite{Rhie2021} have already yielded results.
 Thus, the dense taxonomic sampling made possible by these projects poses new challenges and opportunities for the standard methods of orthology inference and alignment, which implicitly assume fewer and more distantly related genomes or fail to fully leverage the redundancy and organization of their annotations.
@@
 One early method that remains popular is building clusters progressively by identifying nodes that form a triangle with at least two other nodes in the cluster~\cite{Tatusov1997, Jensen2007}.
-Other approaches include the MCL algorithm, which clusters graphs by simulating stochastic flow, and connected components or other single-linkage criteria~\cite{Remm2001, Enright2002, Li2003, Emms2015, Train2017, Cosentino2018}.
+Other approaches include the MCL algorithm, which clusters graphs by simulating stochastic flow~\cite{Enright2002, Li2003, Emms2015}, and connected components or other single-linkage criteria~\cite{Remm2001, Linard2011, Train2017, Cosentino2018}.
 While these methods are robust when clustering hit graphs derived from smaller or more diverse sets of genomes, they are not suitable for the large number of closely related genomes in this work since they require relatively few edges to define a cluster.
@@
 
-The most popular aligners for protein sequences (Clustal Omega, MAFFT, MUSCLE, T-Coffee) do not include splice sites in their alignments, which makes them prone to aligning non-homologous exons~\cite{Sievers2017, Katoh2013, Edgar2004, Notredame2000}.
+The most popular aligners for protein sequences (Clustal Omega~\cite{Sievers2017}, MAFFT~\cite{Katoh2013}, MUSCLE~\cite{Edgar2004}, T-Coffee~\cite{Notredame2000}) do not include splice sites in their alignments, which makes them prone to aligning non-homologous exons.
 Current algorithms can easily be extended to incorporate splice sites by coding them as a new symbol and preventing alignment between splice sites and amino acids, which was recently implemented in the aligner, \textit{Mirage}~\cite{Nord2018}.
@@
 Furthermore, the heterogeneity of genome architectures and annotations may require quality assurance methods tailored to each set of genomes.
-Thus, there is likely no one-size-fits-all approach to orthology inference, and with many other standalone programs available for more standard use cases (Hieranoid, OMA standalone, OrthoFinder, OrthoInspector, Orthologer), we have chosen not to package the code into an end-to-end pipeline~\cite{Kaduk2017, Altenhoff2019, Emms2019, Linard2014, Zdobnov2020}.
+Thus, there is likely no one-size-fits-all approach to orthology inference, and with many other standalone programs available for more standard use cases (Hieranoid~\cite{Kaduk2017}, OMA standalone~\cite{Altenhoff2019}, OrthoFinder~\cite{Emms2019}, OrthoInspector~\cite{Linard2014}, Orthologer~\cite{Zdobnov2020}), we have chosen not to package the code into an end-to-end pipeline.
 Instead, we have devoted considerable attention to organizing and documenting the code to make it accessible to a newcomer and thereby facilitate the adaptation of specific steps to similar projects as needed.
@@
 
-Though databases of orthologous groups such as COGs, Ensembl Compara, EggNOG, OMA, OrthoDB, OrthoInspector, and OrthoMCL will continue to be useful for comparative studies across broad taxonomic ranges, the increasing speed at which high-quality genome assemblies and annotations are produced means no single database can encompass the most complete data~\cite{Galperin2020, Herrero2016, HuertaCepas2018, Altenhoff2020, Zdobnov2020, Nevers2018, Chen2006}.
+Though databases of orthologous groups such as COGs~\cite{Galperin2020}, Ensembl Compara~\cite{Herrero2016}, EggNOG~\cite{HuertaCepas2018}, OMA~\cite{Altenhoff2020}, OrthoDB~\cite{Zdobnov2020}, OrthoInspector~\cite{Nevers2018}, and OrthoMCL~\cite{Chen2006} will continue to be useful for comparative studies across broad taxonomic ranges, the increasing speed at which high-quality genome assemblies and annotations are produced means no single database can encompass the most complete data.
 Furthermore, since many early comparative genomics studies spanned diverse branches of the tree of life, future research will likely prioritize taxonomic depth over breadth.
@@
 Their combination yielded six different sets of meta-alignments.
-A tree was fit to each meta-alignment with the LG substitution model, four discrete gamma rate categories, and optimized state frequencies using IQ-TREE 1.6.12~\cite{Nguyen2014, Le2008, Yang1994}.
+A tree was fit to each meta-alignment with the LG substitution model~\cite{Le2008}, four discrete gamma rate categories~\cite{Yang1994}, and optimized state frequencies using IQ-TREE 1.6.12~\cite{Nguyen2014}.
 If the sampling strategy allowed invariant columns, an invariant rate category was included.
@@
 HMM algorithms were implemented in the standalone package Homomorph which is available at \url{https://github.com/marcsingleton/homomorph} and on the Python Package Index (PyPI).
-The following Python libraries were used: matplotlib, NumPy, pandas, and SciPy~\cite{Hunter2007, Harris2020, McKinney2010, Virtanen2020}.
+The following Python libraries were used: matplotlib~\cite{Hunter2007}, NumPy~\cite{Harris2020}, pandas~\cite{McKinney2010}, and SciPy~\cite{Virtanen2020}.
 Relevant output files are available in the supporting information.

### 2023-04-26 chapter2/chapter2.tex
Moved citations closer to relevant phrases

@@
 The code used to produce the results and analyses is available at \url{https://github.com/marcsingleton/IDR_evolution2023}.
-The following Python libraries were used: matplotlib, NumPy, pandas, SciPy, and scikit-learn~\cite{Hunter2007, Harris2020, McKinney2010, Virtanen2020, Pedregosa2011}.
+The following Python libraries were used: matplotlib~\cite{Hunter2007}, NumPy~\cite{Harris2020}, pandas~\cite{McKinney2010}, SciPy~\cite{Virtanen2020}, and scikit-learn~\cite{Pedregosa2011}.
 Relevant output files are available in the supporting information.

### 2023-04-26 chapter2/chapter2.tex
Updated introduction and discussion outlines
Converted introduction paragraph 2 outline to paragraph

@@
 
-% II. Proteome-wide patterns of conserved features linked to function
-%    B. Despite the prevalence and importance of these regions in eukaryotic proteomes, associating them with specific molecular functions or biological processes remains a significant challenge.
-%    C.
-The sequences of IDRs are generally poorly conserved, so traditional bioinformatics approaches which rely on the strict conservation of amino acid sequences to identify homologous proteins and transfer annotations between biological systems are largely unsucessful when applied to IDRs.
+Despite the prevalence and importance of these IDRs in eukaryotic proteomes, associating them with specific molecular functions or biological processes remains a significant challenge.
+The sequences of IDRs are generally poorly conserved, so traditional bioinformatics approaches which rely on the conservation of amino acid sequences to identify homologous proteins and transfer annotations between them are largely unsuccessful when applied to IDRs.
+However, several recent studies have demonstrated evidence that IDRs are constrained to preserve ``distributed features'' such as flexibility~\cite{Daughdrill2007}, chemical composition~\cite{Moesa2012}, net charge~\cite{Zarin2017}, or charge distribution~\cite{Beh2012}.
+Because many sequences can yield a region with a specific composition, for example, this mode of constraint uncouples an IDR's fitness from its strict sequence of amino acid residues.
+Furthermore, in contrast to folded regions whose precise contacts and packing geometries are easily disrupted by amino acid substitutions, distributed features are robust to such changes as individual residues only weakly contribute to a region's fitness.
+For example, a mutation at one site in an IDR that changes its net charge is easily reversed by subsequent compensatory mutations elsewhere in the region.
+Thus, under this model the sequences of IDRs can rapidly diverge and still preserve their structural or functional properties.
+
+%    B.
+Though other analyses have identified Though IDRs short linear motifs (SLiMs) are enriched in IDRs and are sometimes strongly conserved at specific positions, these residues constitute a small fraction of disordered regions.
+However, because these motifs are short and some are highly permissive at specific positions, they have high evolutionarily plasticity.
+Thus, motifs
+
+%Though these motifs are sometimes conserved by position and can be identified as short segments of conservation in otherwise highly diverged alignments of IDRs, their brief lengths make them evolutionarily plastic.
+Thus, they are easily
+
 %
-%    A.
-However, several recent studies have demonstrated evidence that IDRs are constrained to preserve ``distributed features'' such as flexibility, chemical composition, net charge, or charge distribution.
-%       1.
-Because many sequences can yield a region with a specific composition or charge, this mode of constraint uncouples an IDR's fitness from its strict sequence of amino acid residues.
-the sequences of IDRs can rapidly diverge while still preserving these structural or functional properties.
-%       2.
-Furthermore, in contrast to folded regions whose precise contacts and packing geometries are easily disrupted by point mutations, distributed features are robust to small changes at the sequence level.
-Thus, deleterious mutations are easily corrected by compensatory mutation elsewhere in the region.
-Together, these loose restrictions on the sequences of IDRs allow them to rapidly diverge despite preserving their functions.
-%    B. Additionally other studies have shown importance of short linear motifs in IDRs
-%        1. Can be identified as short segments of conservation in otherwise highly diverged alignments of IDRs
-%        2. However, these segments comprise a small fraction of disordered regions, so
-%
-% III.
-Previously, many of these studies were limited to ...
-However, a recent study first to demonstrate patterns of selection in IDRs across an entire proteome using a set of 82 ``molecular features''
+% III. Many of these analyses were limited to ... However, a recent study first to demonstrate patterns of selection in IDRs across an entire proteome using a set of 82 ``molecular features''
 %    D.
@@
 %    E.
-Subsequent studies have expanded on these findings by using techniques from machine learning and statistics the information from evolutionary signatures to predict the function of IDRs.
+Though subsequent studies have expanded on these findings by using techniques from machine learning and statistics the information from evolutionary signatures to predict the function of IDRs.
 and learn disorder-associated features in a self-supervised manner.
@@
 This would allow the creation of a classification system such as Pfam
+%
 % IV. Explicit phylogenetic models of trait evolution can reveal patterns of evolution with fewer assumptions
@@
 %        5. We also explore hybrid approaches where the evolutionary rates obtained from phylogenetic models are used as a statistic
+%
 % V.
@@
 % III. Disorder score analyses
-%    A. Comparison of features and disorder scores acts as a kind of ``mutational scanning'' experiment [citation to Eddy?]
+%    A. The analysis of features and disorder scores is effectively a natural ``mutational scanning'' experiment [citation to Eddy?]
 %        1. Contrasts of features and disorder scores shows the disorder predictor is sensitive to most features used in this analyses

### 2023-04-26 chapter2/chapter2.tex
Added paragraph summarizing Zarin 2019

@@
 
-%
+While these earlier studies were generally restricted to specific features or proteins, in 2019 Zarin \textit{et al.} demonstrated conservation of various features from a comprehensive set of IDR-associated properties in IDRs across the entire yeast proteome~\cite{Zarin2019}.
+By comparing the observed values of these features against those generated under a simulated model of evolution, they clustered IDRs by their ``evolutionary signatures,'' \textit{i.e.} patterns of conserved features.
+Furthermore, they identified specific biological functions associated with these groups, which for the first time provided a global view of the relationship between sequence and function in IDRs.
+A follow-up analysis expanded on this initial finding by applying techniques from machine learning and statistics to predict the functions of individual IDRs using their evolutionary signatures~\cite{Zarin2021}.
+
 % III. Many of these analyses were limited to ... However, a recent study first to demonstrate patterns of selection in IDRs across an entire proteome using a set of 82 ``molecular features''

### 2023-04-26 chapter2/chapter2.tex
Added paragraph motivating use of Drosophila

@@
 
-%    B.
+% This form of constraint is    B.
 Though other analyses have identified Though IDRs short linear motifs (SLiMs) are enriched in IDRs and are sometimes strongly conserved at specific positions, these residues constitute a small fraction of disordered regions.
@@
 Thus, motifs
-
 %Though these motifs are sometimes conserved by position and can be identified as short segments of conservation in otherwise highly diverged alignments of IDRs, their brief lengths make them evolutionarily plastic.
@@
 A follow-up analysis expanded on this initial finding by applying techniques from machine learning and statistics to predict the functions of individual IDRs using their evolutionary signatures~\cite{Zarin2021}.
+
+However, no known subsequent studies have determined if similar patterns of conservation are found in the IDRs of other systems.
+As another foundational model organism with abundant genomic information across many evolutionary lineages~\cite{Yang2018, Miller2018, Kim2021, Gramates2022}, the fruit fly, \textit{Drosophila melanogaster}, is a natural choice for subsequent investigation.
+Furthermore, given its complex multicellular development process and shared signaling pathways with humans, the findings of such a study would significantly advance our understanding of the role of IDRs in human health and disease.
+The concordance of these results with the previously identified IDR clusters would also have profound implications for the broader mechanisms of IDR evolution.
+For example, the absence of global patterns of evolutionary signatures across IDRs in \textit{Drosophila} would suggest they are property of IDRs which is unique to yeast.
+In contrast, the identification of clusters similar to those in yeast would indicate the existence of a taxonomy of IDRs which is conserved across the tree of life.
 

### 2023-04-26 chapter0/chapter0.tex
Tweaked sentence about broader impacts of study

@@
 As another foundational model organism with abundant genomic information across many evolutionary lineages~\cite{Yang2018, Miller2018, Kim2021, Gramates2022}, the fruit fly, \textit{Drosophila melanogaster}, is a natural choice for subsequent investigation.
-Furthermore, given its complex multicellular development process and shared signaling pathways with humans, the findings of such a study would significantly advance our understanding of the role of IDRs in human health and disease.
+Furthermore, given its complex multicellular development process and shared signaling pathways with humans, the findings of such a study would significantly advance our understanding of the role of IDRs in gene regulation as well as human health and disease.
 The concordance of these results with the previously identified IDR clusters would also have profound implications for the broader mechanisms of IDR evolution.

### 2023-04-26 chapter2/chapter2.tex
Tweaked sentence about broader impacts of study

@@
 In contrast, the identification of clusters similar to those in yeast would indicate the existence of a taxonomy of IDRs which is conserved across the tree of life.
+% The latter result would represent a significant step towards the creation of resources for the classification of IDRs analogous to those for folded domains such as Pfam, CATH or SCOP.
 
-% III. Many of these analyses were limited to ... However, a recent study first to demonstrate patterns of selection in IDRs across an entire proteome using a set of 82 ``molecular features''
-%    D.
-Furthermore, by clustering IDRs on by these ``evolutionary signatures,'' they identified groups of IDRs evolving under similar sets of constraints.
-Furthermore they identified functional associations in these clusters which can be used to generate hypotheses of function for previously uncharacterized proteins
-%    E.
-Though subsequent studies have expanded on these findings by using techniques from machine learning and statistics the information from evolutionary signatures to predict the function of IDRs.
-and learn disorder-associated features in a self-supervised manner.
-% III.
-However, no subsequent studies have determined if similar patterns of conservation are found in other systems.
-As another foundational model organism with abundant genomic information across many evolutionary lineages, the fruit fly, \textit{Drosophila melanogaster}, is a natural choice for subsequent investigation~\cite{Yang2018, Miller2018, Kim2021, Gramates2022}.
-%    A. Expand analyses to another model organism
-%        1. Drosophila is key model for genetic and development research with abundant genomic resources across many lineages in the radiation
-%        2. Furthermore as multicellular organism with complex developmental pathways, the results would be particularly relevant for understanding the role of IDRs in complex signaling pathways
-%        3. Also relevant for human health!
-%    B.
-The concordance of these results with the previously identified IDR clusters would also have profound implications for the broader mechanisms of IDR evolution.
-Depending on the amount of concordance, several conclusions are possible.
-%        1. If no patterns are observed, this suggests that conservation of distributed is a not widely conserved mechanism of IDR evolution.
-%        2. Patterns of conservation are yeast phenomena
-%        3.
-Finally, the identification of similar clusters with conserved features and properties would suggest the existence of ``taxonomy'' of IDRs which is common across many branches of the tree of life.
-This would allow the creation of a classification system such as Pfam
-%
 % IV. Explicit phylogenetic models of trait evolution can reveal patterns of evolution with fewer assumptions

### 2023-04-27 chapter0/chapter0.tex
Tweaked wording for intro to Zarin summary in chapter 2 to better transition and updated wording in chapter 0 to reflect that change

@@
 By the mid 2010s several studies had demonstrated evidence of such constraint in the flexibility~\cite{Daughdrill2007}, chemical composition~\cite{Moesa2012}, net charge~\cite{Zarin2017}, or charge distribution of IDRs~\cite{Beh2012}.
-While these earlier studies were generally restricted to specific features or proteins, in 2019 Zarin \textit{et al.} demonstrated conservation of various features from a comprehensive set of IDR-associated properties in IDRs across the entire yeast proteome~\cite{Zarin2019}.
+While these initial studies were generally restricted to specific features or proteins, in 2019 Zarin \textit{et al.} showed various features from a comprehensive set of IDR-associated properties were conserved in IDRs across the entire yeast proteome~\cite{Zarin2019}.
 By comparing the observed values of these features against those generated under a simulated model of evolution, they clustered IDRs by their ``evolutionary signatures,'' \textit{i.e.} patterns of conserved features.

### 2023-04-27 chapter2/chapter2.tex
Tweaked wording for intro to Zarin summary in chapter 2 to better transition and updated wording in chapter 0 to reflect that change

@@
 
-While these earlier studies were generally restricted to specific features or proteins, in 2019 Zarin \textit{et al.} demonstrated conservation of various features from a comprehensive set of IDR-associated properties in IDRs across the entire yeast proteome~\cite{Zarin2019}.
+While the studies which had initially demonstrated evidence of constraint were generally restricted to specific features or proteins, in 2019 Zarin \textit{et al.} showed various features from a comprehensive set of IDR-associated properties were conserved in IDRs across the entire yeast proteome~\cite{Zarin2019}.
 By comparing the observed values of these features against those generated under a simulated model of evolution, they clustered IDRs by their ``evolutionary signatures,'' \textit{i.e.} patterns of conserved features.

### 2023-04-27 chapter2/chapter2.tex
Added paragraph about motifs as features

@@
 
-% This form of constraint is    B.
-Though other analyses have identified Though IDRs short linear motifs (SLiMs) are enriched in IDRs and are sometimes strongly conserved at specific positions, these residues constitute a small fraction of disordered regions.
-However, because these motifs are short and some are highly permissive at specific positions, they have high evolutionarily plasticity.
-Thus, motifs
-%Though these motifs are sometimes conserved by position and can be identified as short segments of conservation in otherwise highly diverged alignments of IDRs, their brief lengths make them evolutionarily plastic.
-Thus, they are easily
+This form of selective constraint can also describe the evolution of more ``localized'' features in IDRs such as short linear motifs (SLiMs).
+Because SLiMs are composed of fewer than 12 residues, they form limited interfaces that frequently mediate the transient binding events involved in signaling pathways~\cite{Tompa2014}.
+Accordingly, they are highly enriched in IDRs, which provide an accessible and flexible scaffold for these interactions~\cite{Fuxreiter2007, Davey2012}.
+While some SLiMs in IDRs are strongly conserved at specific positions, these constitute a small fraction of disordered residues, estimated at roughly 17\% in the yeast proteome~\cite{NguyenBa2012}.
+Instead, as SLiMs are compact and often highly degenerate at some positions, they can arise de novo from a small number of mutations and therefore have high rates of turnover.
+Furthermore, when IDRs contain multiple copies of a motif that jointly mediate a high-avidity interaction~\cite{Tompa2014} or a graded response to a signal via the accumulation of multiple phosphorylations~\cite{Wright2014}, the individual motifs are under weak selective constraints.
+As a result, though SLiMs are encoded by specific sequences, in some contexts they may evolve as distributed features that characterize IDRs as a whole rather than specific sites within them.
 
@@
 
-% IV. Explicit phylogenetic models of trait evolution can reveal patterns of evolution with fewer assumptions
+% IV. Thus, in this work Explicit phylogenetic models of trait evolution can reveal patterns of evolution with fewer assumptions
 %    A. Previous analyses compared observed distributions of features in observed IDRs to the distributions generated by simulated counterparts

### 2023-04-27 main.tex
Explicitly defined backend for biblatex as biber
Change name of bibliography to references
Changed caption font size to small to better fit large figures on pages
Updated chapter 2 title

@@
 \setcounter{tocdepth}{1}  % Only number divisions to section
-\usepackage[labelfont=bf, labelsep=period]{caption}
+\usepackage[font=small, labelfont=bf, labelsep=period]{caption}
 \renewcommand{\thefigure}{\thechapter.\arabic{figure}}
@@
 % References
-\usepackage[doi=false,url=false, style=nature]{biblatex}
+\usepackage[backend=biber, style=nature, doi=false, url=false]{biblatex}
 \addbibresource{refs.bib}
@@
 
-\chapter{Evolutionary analyses of IDRs reveal conserved features}
+\chapter{Evolutionary analyses of IDRs reveal patterns of conserved features}
 \label{chapter:2}
@@
 
-\printbibliography[heading=bibintoc]
+\printbibliography[heading=bibintoc, title=References]  % bibintoc == put in table of contents
 

### 2023-04-27 chapter2/chapter2.tex
Fixed typo
Added working draft of last paragraph of introduction

@@
 Intrinsically disordered regions (IDRs) are segments of proteins which lack stable three-dimensional structures and instead exist as ensembles of rapidly interconverting conformations.
-As a result of this structural hetereogeneity, IDRs can interact with diverse binding partners.
+As a result of this structural heterogeneity, IDRs can interact with diverse binding partners.
 Often these interactions have high specificity but moderate affinity, which permits the efficient propagation of signals by rapid binding and dissociation~\cite{Zhou2012, Wright2014}.
@@
 
-% IV. Thus, in this work Explicit phylogenetic models of trait evolution can reveal patterns of evolution with fewer assumptions
-%    A. Previous analyses compared observed distributions of features in observed IDRs to the distributions generated by simulated counterparts
-%    B. Differences were interpreted as evidence of selection on features
-%        1. However, evidence of deviation from the null model rather than evidence of constraint necessarily
-%        2. Null distribution generated by a model of evolution where the composition of sequences stochastically evolve towards a specific equilibrium distribution
-%        3.
-Previous studies have identified IDRs with varying compositional preferences, so a single equilibrium distribution is unlikely to globally describe observed patterns of sequence evolution in IDRs
-%    C. Thus in this work we explore use of various statistical approaches that directly measure rates of change and constraint of molecular features indepedent of specific model of sequence evolution
-%        1. Place these evolutionary analyses in a statistical phylogenetic framework through the application of the Brownian motion and Ornstein-Uhlenbeck models of trait evolution.
-%        2.
-Though these models have own assumptions, they are macroscopic descriptions of continuous trait evolution and therefore independent of any assumptions about evolution at the level sequences
-%        3. Still have their own assumptions of course, but more general and robust to deviations
-%        4. We view these approaches as complementary, each having different trade-offs between specificity and generality of testing specific hypotheses of disorder evolution
-%        5. We also explore hybrid approaches where the evolutionary rates obtained from phylogenetic models are used as a statistic
-%
-% V.
-Therefore, in this work we apply evolutionary models of sequence and trait evolution to the sequences of IDRs and their of associated molecular feature to dissect the forces driving their evolution.
-We find IDRs exhibit unique patterns of evolution at the level of sequences.
-We find intrinsic disorder is a dynamically evolving property in some regions, and changes in disorder preference are correlated with changes in the complexity.
-We find that IDRs are unconstrained along several axis of trait evolution, but they also display
+V.
+Therefore, in this work we apply a series of phylogenetic models to the sequences and molecular features of a set of orthologous IDRs identified in the \textit{Drosophila} genome to dissect the forces driving their evolution.
+Though the previous approach relied on simulations to generate the null distribution for a hypothesis of no constraint, our analyses of their molecular features instead leverage a fully statistical phylogenetic comparative framework.
+By comparing models of constrained and unconstrained continuous trait evolution, \textit{i.e.} the Brownian motion and Ornstein-Uhlenbeck models, respectively, we can demonstrate evidence of selective constraint independent of any assumptions about the underlying process of sequence evolution.
+However, we also explore hybrid approaches that combine simulations with phylogenetic comparative methods to test increasingly refined models of IDR evolution.
+Our results show IDRs exhibit unique patterns of amino acid substitution where structure-promoting residues are under less constraint in IDRs despite their relative scarcity.
+Additionally, by treating intrinsic disorder as trait a global analyses of the disorder as trait in some proteins disorder is a dynamically evolving property, with changes in predicted disorder strongly correlated with changes in many distributed features, including hydrophobicity and sequence complexity.
+Furthermore, we find that while IDRs are unconstrained along several axes of trait evolution, they also reveal conserved patterns of feature constraint which are associated with function, which suggests the previously
 

### 2023-04-27 chapter2/chapter2.tex
Finished last paragraph of introduction

@@
 
-V.
-Therefore, in this work we apply a series of phylogenetic models to the sequences and molecular features of a set of orthologous IDRs identified in the \textit{Drosophila} genome to dissect the forces driving their evolution.
-Though the previous approach relied on simulations to generate the null distribution for a hypothesis of no constraint, our analyses of their molecular features instead leverage a fully statistical phylogenetic comparative framework.
-By comparing models of constrained and unconstrained continuous trait evolution, \textit{i.e.} the Brownian motion and Ornstein-Uhlenbeck models, respectively, we can demonstrate evidence of selective constraint independent of any assumptions about the underlying process of sequence evolution.
+Therefore, in this work we apply a series of phylogenetic models to a set of orthologous IDRs identified in the \textit{Drosophila} genome to dissect the forces driving their evolution.
+Our analyses span multiple levels, ranging from the sequences that compose these regions to the distributed features that characterize them as a whole.
+For the latter, though the previous approach relied on simulations to generate the null distribution for a hypothesis of no constraint, we instead leverage a fully statistical phylogenetic comparative framework.
+By comparing models of constrained and unconstrained continuous trait evolution, \textit{i.e.} the Brownian motion and Ornstein-Uhlenbeck models, respectively, we can demonstrate evidence of selective constraint on features independent of any assumptions about the underlying process of sequence evolution.
 However, we also explore hybrid approaches that combine simulations with phylogenetic comparative methods to test increasingly refined models of IDR evolution.
-Our results show IDRs exhibit unique patterns of amino acid substitution where structure-promoting residues are under less constraint in IDRs despite their relative scarcity.
-Additionally, by treating intrinsic disorder as trait a global analyses of the disorder as trait in some proteins disorder is a dynamically evolving property, with changes in predicted disorder strongly correlated with changes in many distributed features, including hydrophobicity and sequence complexity.
-Furthermore, we find that while IDRs are unconstrained along several axes of trait evolution, they also reveal conserved patterns of feature constraint which are associated with function, which suggests the previously
+We find that IDRs exhibit unique patterns of amino acid substitution and that in some proteins disorder itself is a dynamically evolving property.
+Furthermore, though IDRs are broadly unconstrained along several axes of feature evolution, we identify specific clusters of IDRs with shared patterns of feature constraint.
+As in yeast, these clusters are enriched for proteins with specific functional annotations, which suggests the preservation of distributed features and their associated functions is a widespread mechanism of IDR evolution.
 

### 2023-04-27 chapter2/chapter2.tex
Added comparative methods citation

@@
 Our analyses span multiple levels, ranging from the sequences that compose these regions to the distributed features that characterize them as a whole.
-For the latter, though the previous approach relied on simulations to generate the null distribution for a hypothesis of no constraint, we instead leverage a fully statistical phylogenetic comparative framework.
+For the latter, though the previous approach relied on simulations to generate the null distribution for a hypothesis of no constraint, we instead leverage a fully statistical phylogenetic comparative framework~\cite{Cornwell2017}.
 By comparing models of constrained and unconstrained continuous trait evolution, \textit{i.e.} the Brownian motion and Ornstein-Uhlenbeck models, respectively, we can demonstrate evidence of selective constraint on features independent of any assumptions about the underlying process of sequence evolution.

### 2023-04-27 chapter2/figures/hierarchy/hierarchy.py

@@ -18,6 +18,19 @@ min_indel_columns = 5  # Indel rates below this value are set to 0
 min_aa_rate = 1
 min_indel_rate = 0.1
 
+clusters = [('A', '15086'),
+            ('B', '15170'),
+            ('C', '15160'),
+            ('D', '15070'),
+            ('E', '14415'),
+            ('F', '15217'),
+            ('G', '15078'),
+            ('H', '15021'),
+            ('I', '15161'),
+            ('J', '15191'),
+            ('K', '15143'),
+            ('L', '14937')]
+
 color1 = '#4e79a7'
 
 # Load regions
@@ -108,7 +121,7 @@ legend_args = {'aa_group': ('Amino acid content', 'grey', ''),
                'motifs_group': ('Motifs', 'white', 4 * '\\')}
 group_labels = ['aa_group', 'charge_group', 'motifs_group', 'physchem_group', 'complexity_group']
 group_labels_nonmotif = ['aa_group', 'charge_group', 'physchem_group', 'complexity_group']
-gridspec_kw = {'width_ratios': [0.1, 0.9], 'wspace': 0,
+gridspec_kw = {'width_ratios': [0.1, 0.7, 0.2], 'wspace': 0,
                'height_ratios': [0.975, 0.025], 'hspace': 0.01,
                'left': 0.05, 'right': 0.95, 'top': 0.95, 'bottom': 0.125}
 
@@ -133,7 +146,7 @@ for node in tree.postorder():
     node2tips[node] = tips
     node2color[node] = str(max(0, (11 - tips) / 10))
 
-fig, axs = plt.subplots(2, 2, figsize=(7.5, 7.5), gridspec_kw=gridspec_kw)
+fig, axs = plt.subplots(2, 3, figsize=(7.5, 7.5), gridspec_kw=gridspec_kw)
 
 # Tree
 ax = axs[0, 0]
@@ -156,9 +169,25 @@ ax.set_yticks([])
 for spine in ax.spines.values():
     spine.set_visible(False)
 
-# Corner axis
-ax = axs[1, 0]
-ax.set_visible(False)
+# Corner axes
+for ax in [axs[1, 0], axs[1, 2]]:
+    ax.set_visible(False)
+
+# Cluster blocks
+ax = axs[0, 2]
+id2idx = {tip.name: idx for idx, tip in enumerate(tree.tips())}
+for cluster_id, root_id in clusters:
[…]

### 2023-04-27 chapter2/chapter2.tex
Added final results paragraph

@@
 We chose not to treat these differences in a strict statistical sense but instead as qualitative ``evolutionary signatures'' as in Zarin \textit{et al.}~\cite{Zarin2019}.
-We therefore clustered the signatures to identify patterns of feature conservation across the proteome.
+Accordingly, we clustered the signatures to identify patterns of feature conservation across the proteome.
 To ensure the clusters were enriched in regions with a high likelihood of feature conservation despite low levels of sequence identity, we restricted this analysis to regions with a minimum amount of divergence as measured by substitution models (Fig.~\ref{sfig:hierarchy_histogram}).
+
+The resulting heatmap reveals patterns of similar signatures interspersed among a high background of noise (Fig.~\ref{fig:hierarchy}).
+However, we identify at least 12 clusters with a strong and consistent patterns of constraint, with many having significant associations with certain processes, functions, or compartments.
+For example, cluster X is strongly enriched in Y, and cluster A contains many proteins involved in B.
+Though these preliminary results suggest the existence of high-level relationships between certain patterns of conserved features and function, additional insights into the biophysical and mechanistic details that underpin these associations will depend on more in-depth analyses of specific clusters and the properties of their constituent IDRs.
 

### 2023-04-28 chapter2/chapter2.tex
Moved heatmap figure into results
Tweaked some wording in last results paragraph

@@
 
+\begin{figure}[h!]
+\includegraphics[width=\textwidth]{hierarchy/out/hierarchy.png}
+\centering
+\caption{\textbf{Hierarchical clustering of evolutionary signatures.}
+The AIC difference between the BM and OU models is measure of their relative goodness of fit to the data with a penalty for the number of parameters in each.
+Larger values indicate a better fit by the OU model.}
+\label{fig:hierarchy}
+\end{figure}
+
 The resulting heatmap reveals patterns of similar signatures interspersed among a high background of noise (Fig.~\ref{fig:hierarchy}).
@@
 For example, cluster X is strongly enriched in Y, and cluster A contains many proteins involved in B.
-Though these preliminary results suggest the existence of high-level relationships between certain patterns of conserved features and function, additional insights into the biophysical and mechanistic details that underpin these associations will depend on more in-depth analyses of specific clusters and the properties of their constituent IDRs.
+Though these preliminary results suggest the existence of high-level relationships between certain patterns of conserved features and function, further insight into the biophysical and mechanistic details that underpin these associations will rely on more in-depth analyses of specific clusters and the properties of their constituent IDRs.
 
@@
 %         - Embeddings from LLMs trained on large data sets
-
-\begin{figure}[h!]
-\includegraphics[width=\textwidth]{hierarchy/out/hierarchy.png}
-\centering
-\caption{\textbf{Hierarchical clustering of evolutionary signatures.}
-The AIC difference between the BM and OU models is measure of their relative goodness of fit to the data with a penalty for the number of parameters in each.
-Larger values indicate a better fit by the OU model.}
-\label{fig:hierarchy}
-\end{figure}
 

### 2023-04-28 chapter2/chapter2.tex
Updated discussion outline
Added sentence about analogy to Pfam and references
Removed redundant use of the word feature

@@
 In contrast, the identification of clusters similar to those in yeast would indicate the existence of a taxonomy of IDRs which is conserved across the tree of life.
-% The latter result would represent a significant step towards the creation of resources for the classification of IDRs analogous to those for folded domains such as Pfam, CATH or SCOP.
+The latter result would represent a significant step towards the creation of resources for the classification of IDRs analogous to those for folded domains such as Pfam~\cite{Mistry2020}, CATH~\cite{Sillitoe2020}, or SCOP~\cite{Andreeva2013, Andreeva2019}.
 
@@
 We find that IDRs exhibit unique patterns of amino acid substitution and that in some proteins disorder itself is a dynamically evolving property.
-Furthermore, though IDRs are broadly unconstrained along several axes of feature evolution, we identify specific clusters of IDRs with shared patterns of feature constraint.
+Furthermore, though IDRs are broadly unconstrained along several axes of feature evolution, we identify specific clusters of IDRs with shared patterns of constraint.
 As in yeast, these clusters are enriched for proteins with specific functional annotations, which suggests the preservation of distributed features and their associated functions is a widespread mechanism of IDR evolution.
@@
 % II. Exchangeability and rate matrices
-%    A. Reproduced trends seen in previous substitution models fit to disorder regions
+%    A.
+Interestingly, the exchangeability matrices suggest that although there is net flux towards disorder by depleting structure-promoting residues, disordered residues are more biochemically distinct in disordered regions than in structured regions.
+This suggests that although disorder-promoting amino acids are often grouped into a single category and described as largely hydrophilic, they have distinct roles in disordered regions.
+There is evidence that highly similar residues such as glutamine and asparagine have distinct properties in disordered regions that yielded different biophysical characters of phase-separated condensates, so this aligns with other observations.
+Conversely, structure promoting residues are more interchangeable.
+One hypothesis is as these bulkier hydrophobic residues are not constrained by packing, they are free to act more generically as small hydrophobic patches that mediate interactions in SLiMs and other binding regions in IDRs.
+Another model for this behavior is the stickers and spacers model of phase separation.
+%    A.
+A difficulty with fitting substitution models is their dependence on sequence alignments which were created using scoring matrices that are in turn often derived from substitution models.
+Some approaches attempt to minimize the impact of this circular dependency through an EM-like procedure where substitution models are fit and the sequences in alternating rounds until convergence.
+Because our analyses fit the substitution models directly from alignments, the observed patterns we observe could be artifacts of using a scoring matrix largely fit to folded domains to align IDRs
+%    A.
+However, as the exchangeability and rate matrices broadly reproduced the trends observed in previous studies which fit amino acid substitution models to disorder regions, these observations are likely not artifactual but instead represent reproducible differences in the substitution rates between residues.
 %    B. Interpretations of exchangeability and rate matrices
@@
 % III. Disorder score analyses
-%    A. The analysis of features and disorder scores is effectively a natural ``mutational scanning'' experiment [citation to Eddy?]
-%        1. Contrasts of features and disorder scores shows the disorder predictor is sensitive to most features used in this analyses
-%        2. However, highest correlations to expected features like fraction_disorder and hydropathy
+%    A.
+The analysis of feature and score contrasts is effectively a natural ``mutational scanning'' experiment.
+Though the mutations generated by evolution are not a random or exhaustive sample of sequence space, they are perturbations around a common ancestral sequence which can reveal the relationship between a region's biophysical properties and its propensity for disorder.
+Accordingly, we find disorder scores have the strongest correlations with features that measure a region's overall hydrophobicity, \textit{e.g.} fraction_disorder and hydropathy.
+As these features are known correlates of intrinsic disorder, this relationship is expected.
+However, the strength of these correlations indicate that a region's relative proportion of hydrophilic and hydrophobic residues is, to a first approximation, the largest determinant of its structural state.
 %        3.
-Interestingly, negatively correlated with wf_complexity.
-Many disordered regions have low levels of sequence complexity.
-However this is a statistical rather than a biophysical criterion, and some low complexity regions are structured, like collagen.
-This indicates that these models can confuse correlation and causation during training, and their representations should be interpreted carefully.
-%        4. Furthermore, resolution of analyses limited because features calculated at level of regions, whose lengths can exceed 1,000 residues, rather than a fixed window around each residue.
+Excluding these hydrophobicity-related features, the second strongest association is a negatively correlation with wf_complexity, indicating the predictor strongly associates low complexity with intrinsic disorder.
+However, sequence complexity is a statistical rather than a biophysical criterion, and while many disordered regions have low levels of complexity, some low-complexity regions, like collagen, are structured.
+This suggests that while disorder predictors are in general accurate classifiers of a residue's structural state, they can conflate correlation and causation, and therefore in some cases their predictions may require careful interpretation.
+%
 %        5.
+Interestingly, the disorder scores have significant, if weak, correlations with most features used in this analysis, including the motifs.
 Disorder scores are weakly correlated with many motifs.
@@
 Thus, it's unclear if the predictor is responding directly to these features or features which are correlated with them.
-More in depths of the representations of these disorder predictors would shed led on these questions.
-%    B. GO annotation enrichment analysis suggests disorder scores undergo rapid evolution in proteins involved in DNA repair and found in extracellular structure
-%        1. Not corrected for multiple testing nor is the false discovery rate controlled, so should be considered as hypotheses for further investigation
-%        2. However, this result is at least consistent with other known functions of disorder
-%            i. Interactions with extracellular components and signaling, both highly enriched in disorder?
-%            ii. These regions could contain MoRFs, so they are already on the boundary between folded and disorder
-%            iii. As evolution optimizes these interactions, small changes in the biophysical properties of these regions may have disproportionate effects on their structural ensembles
-%            iv. The variable scoring by the disorder predictor may reflect this kind of ``phrase transition''
+More in depth analyses of the representations of these disorder predictors would shed led on these questions.
+%        1. In, the disorder scores have a significant relationship with most features used this analysis.
+%        4. Furthermore, resolution of analyses limited because features calculated at level of regions, whose lengths can exceed 1,000 residues, rather than a fixed window around each residue.
+
+%    B.
+The GO annotation enrichment analysis indicate that proteins involved in DNA repair and the formation of extracellular structure contain disproportionate number of regions whose disorder scores are rapidly evolving.
+Because the significance tests were not corrected for multiple testing or controlled for false discovery rate, we caution against over-interpreting this result and instead consider as a hypothesis for further investigation.
+In general, however, these regions with rapidly evolving disorder scores may correspond to molecular recognition features (MoRFs), which are modules in IDRs that undergo a disorder-to-order transition on binding their targets.
+Because MoRFs already exist on the boundary between disorder and structure, small changes in the biophysical properties of these regions may have disproportionate effects on their structural ensembles.
+Thus, the most variable disorder scores may reflect instances where a mutation triggered a ``phase transition'' between largely structured or disordered native state.
 
 % IV. Feature analyses
-%    A. Though no discrete clusters are visible, only captures x% of the overall variation, so significant amounts of variation not represented in these axes
-%    B. Hierarchical clustering of the rates reveals additional structure.
-%    C. Other non-linear dimensionality reductions like UMAP be more successful at preserving the fine details of the high-dimensional distributions in the low-dimensional space
-%        1.
-However, these methods designed primarily for visualization and as result can significant distort the global structure of high-dimensional data to preverse its local topology.
+%    A.
+In the PCAs of the feature roots and rates, though no discrete clusters are visible in the three largest components, they only accounted for X\% of the total variation between region.
+This suggests, a majority of variation is contained other in these components.
+%    B. The success of the hierarchical clustering of the rates at revealing additional structure suggests
+%    C.
+Other non-linear dimensionality reductions like UMAP be more successful at preserving the fine structure of these high-dimensional distributions in the low-dimensional space.
+However, as these methods are designed primarily for visualization, they can significant distort the global structure of high-dimensional data.
 Thus, these embeddings are generally not suitable for downstream analyses.
+Instead, semi-supervised approaches for feature learning may yield more meaningful representations.
+In fact, a recent study used a semi-supervised learning approach that identified orthologous IDRs to learn novel disorder-associated features.
+Though they used a simple neural network that aided in interpretability, other similar semi-supervised approaches have used more complex networks to learn rich representations of protein regions.
+Similar approaches may yield representations of the space of disordered proteins that are more clearly visualized and manipulated.
 
 % V. OU analyses
+%    A. While preliminary analyses indicate
 %    A. Discussion of issues with statistical inference and need for simulations to address inference issues with OU model

### 2023-04-28 chapter2/chapter2.tex
Finished paragraphs on contrast analyses
Added contrast analyses paragraphs to discussion

@@
 
-% III. Disorder score analyses
-%    A.
-The analysis of feature and score contrasts is effectively a natural ``mutational scanning'' experiment.
-Though the mutations generated by evolution are not a random or exhaustive sample of sequence space, they are perturbations around a common ancestral sequence which can reveal the relationship between a region's biophysical properties and its propensity for disorder.
-Accordingly, we find disorder scores have the strongest correlations with features that measure a region's overall hydrophobicity, \textit{e.g.} fraction_disorder and hydropathy.
-As these features are known correlates of intrinsic disorder, this relationship is expected.
-However, the strength of these correlations indicate that a region's relative proportion of hydrophilic and hydrophobic residues is, to a first approximation, the largest determinant of its structural state.
-%        3.
-Excluding these hydrophobicity-related features, the second strongest association is a negatively correlation with wf_complexity, indicating the predictor strongly associates low complexity with intrinsic disorder.
-However, sequence complexity is a statistical rather than a biophysical criterion, and while many disordered regions have low levels of complexity, some low-complexity regions, like collagen, are structured.
-This suggests that while disorder predictors are in general accurate classifiers of a residue's structural state, they can conflate correlation and causation, and therefore in some cases their predictions may require careful interpretation.
-%
-%        5.
-Interestingly, the disorder scores have significant, if weak, correlations with most features used in this analysis, including the motifs.
-Disorder scores are weakly correlated with many motifs.
-Furthermore, unclear if predictor is responding directly to these motifs as a signal of disorder, or they appear in regions which evolving towards disorder.
-Interestingly docking (DOC) and ligand binding (LIG) sites tend to have negative correlations whereas the modification (MOD) sites have positive correlations.
-Docking and ligand binding sites are generally mediated by small hydrophobic patches, so the predictor could be responding to increased hydrophobicity caused by additional of hydrophobic residue ``completing'' the motif rather than the motif itself.
-Likewise, phosphorylation sites are highly enriched in IDRs and many are targeted to disorder-promoting residues like serine or threonine.
-Thus, it's unclear if the predictor is responding directly to these features or features which are correlated with them.
-More in depth analyses of the representations of these disorder predictors would shed led on these questions.
-%        1. In, the disorder scores have a significant relationship with most features used this analysis.
-%        4. Furthermore, resolution of analyses limited because features calculated at level of regions, whose lengths can exceed 1,000 residues, rather than a fixed window around each residue.
-
-%    B.
-The GO annotation enrichment analysis indicate that proteins involved in DNA repair and the formation of extracellular structure contain disproportionate number of regions whose disorder scores are rapidly evolving.
-Because the significance tests were not corrected for multiple testing or controlled for false discovery rate, we caution against over-interpreting this result and instead consider as a hypothesis for further investigation.
-In general, however, these regions with rapidly evolving disorder scores may correspond to molecular recognition features (MoRFs), which are modules in IDRs that undergo a disorder-to-order transition on binding their targets.
-Because MoRFs already exist on the boundary between disorder and structure, small changes in the biophysical properties of these regions may have disproportionate effects on their structural ensembles.
-Thus, the most variable disorder scores may reflect instances where a mutation triggered a ``phase transition'' between largely structured or disordered native state.
-
 % IV. Feature analyses
@@
 \section{Discussion}
-Here's where my discussion will go!
+The analysis of feature and score contrasts is effectively a natural ``mutational scanning'' experiment.
+Though the mutations generated by evolution are not a random or exhaustive sample of sequence space, they are perturbations of a common ancestor which can reveal the relationship between a region's biophysical properties and its propensity for disorder.
+Accordingly, we find disorder scores have the strongest correlations with features that measure a region's overall polarity and hydrophobicity, \textit{e.g.} fraction\_disorder and hydropathy.
+As the formation of a hydrophobic core is a major driving force in protein folding, this relationship is expected.
+However, the strength of these correlations indicate that a region's relative proportion of hydrophilic and hydrophobic residues is, to a first approximation, the largest determinant of intrinsic disorder.
+
+Excluding these hydrophobicity-related features, the next strongest association is a negative correlation with wf\_complexity, demonstrating the predictor strongly associates low complexity with intrinsic disorder.
+However, sequence complexity is a statistical rather than a biophysical criterion, and while many disordered regions have low levels of complexity, some low-complexity regions, like collagen, are structured.
+This suggests that while disorder predictors are in general accurate classifiers of a residue's structural state, they can conflate the correlates of intrinsic disorder with their causes.
+Therefore, in some cases their predictions may require careful interpretation.
+
+The remaining significant correlations are generally weak and likely reflect a partial redundancy with the more strongly correlated features discussed previously.
+Interestingly, though, disorder scores are weakly correlated with many motifs, with the signs largely reflecting their class.
+For example, the correlations with docking (DOC) and ligand binding (LIG) sites are largely negative whereas those with modification (MOD) sites are generally positive.
+However, this analysis does not indicate whether the predictor responds to these motifs directly or to features that are correlated with them.
+For example, docking and ligand binding sites are generally mediated by small hydrophobic patches, so the correlations could reflect an increase in hydrophobicity caused by an additional hydrophobic residue ``completing'' the motif rather than the motif itself.
+Likewise, IDRs are highly enriched in phosphorylation sites, many of which are targeted to disorder-promoting residues like serine or threonine.
+As the disorder scores and features are calculated at the level of regions, whose lengths can exceed 1,000 residues, this analysis is limited in its ability to distinguish these possibilities.
+However, a more targeted \textit{in silico} mutational analysis would yield further insights.
+
+The GO annotation enrichment analysis indicates that proteins involved in DNA repair and extracellular structures contain a disproportionate number of regions whose disorder scores are rapidly evolving.
+Because the significance tests were not corrected for multiple testing or controlled for their false discovery rate, we caution against over-interpreting this result and instead consider it as a hypothesis for further investigation.
+In general, however, these regions with rapidly evolving disorder scores may correspond to molecular recognition features (MoRFs), which are modules in IDRs that undergo a disorder-to-order transition on binding their targets.
+Because MoRFs already exist on the boundary between disorder and structure, small changes in the biophysical properties of these regions may have disproportionate effects on their structural ensembles.
+Thus, the most variable disorder scores may reflect instances where a mutation triggered a ``phase transition'' between largely structured or disordered native states.
 

### 2023-04-29 chapter2/chapter2.tex
Added sentence about order rate supplemental figure
Switched order of sentences to introduce contrast discussion

@@
 Inspection of regions selected along these axes confirmed the high rates of evolution of these features (Fig.~\ref{fig:rate}E-G).
+Furthermore, we observe a similar distribution when the rates of the order regions were projected along their second and third principal components, which suggests a lack of constraint along these axes is a general property of rapidly evolving proteins (Fig.~\ref{sfig:order}).
 
@@
 \section{Discussion}
-The analysis of feature and score contrasts is effectively a natural ``mutational scanning'' experiment.
 Though the mutations generated by evolution are not a random or exhaustive sample of sequence space, they are perturbations of a common ancestor which can reveal the relationship between a region's biophysical properties and its propensity for disorder.
-Accordingly, we find disorder scores have the strongest correlations with features that measure a region's overall polarity and hydrophobicity, \textit{e.g.} fraction\_disorder and hydropathy.
+Thus, the analysis of feature and score contrasts is effectively a natural ``mutational scanning'' experiment.
+We found that disorder scores have the strongest correlations with features that measure a region's overall polarity and hydrophobicity, \textit{e.g.} fraction\_disorder and hydropathy.
 As the formation of a hydrophobic core is a major driving force in protein folding, this relationship is expected.

### 2023-04-29 chapter2/chapter2.tex
Updated discussion outline
Finished methodology discussion paragraph
Updated last paragraphs of discussion
Added updated paragraphs to final draft

@@
 % I. Summary of results
-%    A. In this study we showed...
-%    B. IDRs have patterns of evolution distinct from folded domains
-%    C. Elevated rates of residue substitution and indels
-% II. General caveats
+%    A. In this study, we demonstrate the first evidence of evolutionary constraint in IDRs across the \textit{Drosophila} proteome.
+%    B. Most immediately, this work generates hypotheses of function related to sequence features that will permit the functional dissection of IDRs.
+%    C. More broadly, however, it illustrates constraint of distributed features characterizes the evolution of some IDRs across multiple systems.
 
 % II. Exchangeability and rate matrices
+%    A. In addition to these global patterns of conserved features, we find IDRs exhibit other characteristic patterns of evolution.
+%    B. For example, a comparison of the exchangeability matrices fit to the disorder and order regions show decreased exchangeability coefficients between
 %    A.
@@
 %    A.
-A difficulty with fitting substitution models is their dependence on sequence alignments which were created using scoring matrices that are in turn often derived from substitution models.
-Some approaches attempt to minimize the impact of this circular dependency through an EM-like procedure where substitution models are fit and the sequences in alternating rounds until convergence.
+A caveat of this analysis is its dependence on sequence alignments which were created using scoring matrices that are in turn often derived from substitution models.
+Some approaches attempt to minimize the impact of this circular dependency through an EM-like procedure where substitution models are fit and the sequences are re-aligned with scoring matrices derived from those models in alternating rounds until convergence.
 Because our analyses fit the substitution models directly from alignments, the observed patterns we observe could be artifacts of using a scoring matrix largely fit to folded domains to align IDRs
@@
 %    A.
-In the PCAs of the feature roots and rates, though no discrete clusters are visible in the three largest components, they only accounted for X\% of the total variation between region.
-This suggests, a majority of variation is contained other in these components.
-%    B. The success of the hierarchical clustering of the rates at revealing additional structure suggests
-%    C.
-Other non-linear dimensionality reductions like UMAP be more successful at preserving the fine structure of these high-dimensional distributions in the low-dimensional space.
+Though no discrete clusters are visible in the largest two and three principal components of the feature root and rate distribution, they only account for X\% and Y\% of the total variation between regions, respectively.
+Thus, a majority of the variation is contained in other components.
+%    B.
+The success of the hierarchical clustering at identifying common patterns of feature constraint suggests this method, or other non-linear dimensionality reductions like UMAP or t-SNE, may also reveal additional structure latent in the high-dimesional rate distributions.
 However, as these methods are designed primarily for visualization, they can significant distort the global structure of high-dimensional data.
@@
 % V. OU analyses
-%    A. While preliminary analyses indicate
-%    A. Discussion of issues with statistical inference and need for simulations to address inference issues with OU model
-%        1. Lack of power or prone to type I errors?
-%    B. Better models for handling motif counts
-%        1.
-Unlike other features where small changes in a sequence general yield small changes in their value, the complexity of the regular expressions made them highly sensitive to small perturbations.
-%    C. Need for comparisons with simulations with amino acid substitution and indel models
+% In summary, both the phylogenetic comparative and simulation-based methods have their strengths and weaknesses.
+The BM and OU models are powerful models of generic trait evolution, and therefore can describe However, because simulation-based approaches can fully specify the mechanisms and constraints of IDR evolution, they permit tests of increasingly refined models of IDR evolution.
+We therefore view the two approaches are complementary, and propose the use of hybrid methods test statistics derived from phylogenetic comparative methods like the BM model and simulations to generate the null distributions for those test statistics.
+Because the simulations are more specific and time-consuming, we recommend future investigators begin using
+%    C. Because
 %        1. Weaknesses are explicit assumptions about substitution and indel processes and high computational cost
@@
 %        4. Hybrid approach using comparisons of rates as the statistical test and simulations to generate the null distribution
-%    D. Mention other distribution free approaches
-
-% VI. Big picture and wrap up
-%     A. Experimental swaps of IDRs in similar classes
-%     B. What are features?
-%         - Contrastive learning (Moses)
-%         - Prediction on gold standard data sets
-%         - Embeddings from LLMs trained on large data sets
+%    D. Another hybrid approach is using permutation-like tests to test for the conservation of patterned features like omega or motifs
+%        1. Uses background distribution of amino acids in each IDR, so can specifically test for importance of patterning based features.
+%        2.
+We expect that IDRs evolve under multiple regimes where some are unconstrained and others are constrained with compositional and patterned features, or a mixture of both.
+As a battery of statistical methods will be necessary to tease apart these details, we view these approaches as complementary, each with its own strengths and weaknesses.
+%    E.
+Thus, while these results represent a significant step forward in relating sequence to function in IDRs, further studies exploring these approaches will undoubtedly reveal new insights into these pervasive but poorly understood regions of proteins.
 
@@
 
+As discussed by Zarin \textit{et al.}, the interpretation of evolutionary signatures is complicated by several methodological caveats~\cite{Zarin2019}.
+For example, because IDRs are identified as contiguous segments of high predicted disorder, their boundaries are defined by adjacent structural elements.
+This approach can therefore split an IDR that is a single evolutionary or functional unit if it contains a semi-disordered module that scores below the threshold.
+Conversely, it can also merge two distinct IDRs if they are not separated by at least one folded domain.
+Another challenge is the significant overlap in the definitions of many features induces strong correlations that preclude straightforward quantitative manipulations or interpretations of an IDR's evolutionary signature.
+The original authors have since addressed this limitation in subsequent studies by applying machine learning methods to perform feature selection or learn features directly from alignments of IDRs~\cite{Zarin2021, Lu2022}.
+However, integrating these methods into a unified phylogenetic comparative framework will require further effort.
+
+By fitting the BM and OU models to molecular features calculated from alignments of IDRs, we were able to quantify the relative support for constrained and unconstrained models of IDR evolution using a statistical framework.
+As the BM and OU models describe the evolution of arbitrary continuous traits, a strength of this approach is its independence from assumptions about the underlying process of sequence evolution.
+In contrast, the previous study used simulations to generate null distributions for a model of no constraint and defined an IDR's evolutionary signature as its deviation from these distributions.
+However, these comparisons do not directly demonstrate evidence of stabilizing selection but instead test for differences from the null hypothesis.
+Thus, this approach is highly dependent on the specification and parameterization of these simulated models.
+Accordingly, an error in either can yield evidence of constraint for an IDR even if none of its molecular features are under selection.
+
+However, the comparative phylogenetics methodology applied here also has limitations.
+As many features have strict boundaries or cannot vary continuously, they violate one or more of the underlying assumptions of the BM or OU models.
+Fortunately, for many features these inconsistencies likely do not seriously compromise the analysis.
+For example, though compositional features like fraction\_S are mathematically restricted to the interval between zero and one, they are likely constrained by much narrower selective regimes, and within these regimes, their behavior is effectively described by an OU model.
+For other features, however, the deviations are more consequential.
+For example, net\_charge and the motif features can only assume integer and non-negative values, respectively, which imposes significant restrictions on their allowed increments that are not reflected in the BM and OU models.
+Instead, more appropriate models for count data are birth-death processes, which are Markov chains defined on the natural numbers.
+However, though linear birth-death processes are well-studied and widely applied in biology~\cite{Crawford2011}, to our knowledge there are no simple parameterizations which describe a mean-reverting behavior analogous to the OU model.
+Thus, further theoretical developments are needed to apply birth-death processes as a model of stabilizing selection in studies of IDR evolution.
+
 \section{Materials and methods}
-
 \subsection{Alignment and species tree provenance}

### 2023-04-30 chapter2/chapter2.tex
Updated discussion paragraphs in outline
Added final paragraphs to discussion section

@@
 
-% V. OU analyses
-% In summary, both the phylogenetic comparative and simulation-based methods have their strengths and weaknesses.
-The BM and OU models are powerful models of generic trait evolution, and therefore can describe However, because simulation-based approaches can fully specify the mechanisms and constraints of IDR evolution, they permit tests of increasingly refined models of IDR evolution.
-We therefore view the two approaches are complementary, and propose the use of hybrid methods test statistics derived from phylogenetic comparative methods like the BM model and simulations to generate the null distributions for those test statistics.
-Because the simulations are more specific and time-consuming, we recommend future investigators begin using
-%    C. Because
-%        1. Weaknesses are explicit assumptions about substitution and indel processes and high computational cost
-%        2. Is also a strength because allows testing of increasingly specific hypotheses for IDR evolution
-%        3. Complementary approach -- fewer assumptions about description of evolution but cannot be tailored to specific hypotheses
-%        4. Hybrid approach using comparisons of rates as the statistical test and simulations to generate the null distribution
-%    D. Another hybrid approach is using permutation-like tests to test for the conservation of patterned features like omega or motifs
-%        1. Uses background distribution of amino acids in each IDR, so can specifically test for importance of patterning based features.
-%        2.
-We expect that IDRs evolve under multiple regimes where some are unconstrained and others are constrained with compositional and patterned features, or a mixture of both.
-As a battery of statistical methods will be necessary to tease apart these details, we view these approaches as complementary, each with its own strengths and weaknesses.
-%    E.
-Thus, while these results represent a significant step forward in relating sequence to function in IDRs, further studies exploring these approaches will undoubtedly reveal new insights into these pervasive but poorly understood regions of proteins.
-
 \section{Discussion}
@@
 Thus, further theoretical developments are needed to apply birth-death processes as a model of stabilizing selection in studies of IDR evolution.
+
+While the BM and OU models are powerful tools for studying trait evolution, their generality limits the specificity of the hypotheses they can test.
+In contrast, because simulation-based approaches can specify arbitrary constraints, they permit investigations of increasingly refined models of IDR evolution.
+We therefore view the two approaches as complementary, and propose the use of hybrid methods where test statistics are derived from phylogenetic comparative methods like the BM model, and simulations generate the null distributions for those test statistics.
+While simulations can eliminate specific hypotheses, the substantial resources involved both in designing them and generating samples make exhaustively testing mechanisms of feature constraint by simulation impractical.
+We instead recommend a workflow that begins with an analysis using generic models of trait evolution to suggest specific hypotheses of constraint that are then tested with a simulation-based approach.
+Another hybrid method involves using sequence permutations to test for the conservation of motifs or patterned features like kappa.
+In this method, a comparative model like BM provides the test statistic as before, and a sample of shuffled sequences approximates the null distribution.
+Because the background distribution of residues is preserved, this approach can specifically test for the conservation of motifs or patterned features in an IDR independent of the conservation of its composition.
+
+However, as IDRs likely evolve under multiple selective regimes where the constraint of different compositional, patterned, or motif features is highly specific to each, we expect a combination of statistical and computational methods will be needed to disentangle the complex forces driving IDR evolution.
+Accordingly, while these results represent a significant step forward in relating sequence to function in IDRs, further studies exploring these and other approaches will undoubtedly reveal new insights into these pervasive but poorly understood regions of proteins.
 

### 2023-04-30 chapter2/chapter2.tex
Added citation for MoRFs

@@
 Because the significance tests were not corrected for multiple testing or controlled for their false discovery rate, we caution against over-interpreting this result and instead consider it as a hypothesis for further investigation.
-In general, however, these regions with rapidly evolving disorder scores may correspond to molecular recognition features (MoRFs), which are modules in IDRs that undergo a disorder-to-order transition on binding their targets.
+In general, however, these regions with rapidly evolving disorder scores may correspond to molecular recognition features (MoRFs), which are modules in IDRs that undergo a disorder-to-order transition on binding their targets~\cite{vanderLee2014}.
 Because MoRFs already exist on the boundary between disorder and structure, small changes in the biophysical properties of these regions may have disproportionate effects on their structural ensembles.

### 2023-04-30 chapter2/chapter2.tex
Removed awkward use of "these IDRs"

@@
 
-Despite the prevalence and importance of these IDRs in eukaryotic proteomes, associating them with specific molecular functions or biological processes remains a significant challenge.
+Despite the prevalence and importance of IDRs in eukaryotic proteomes, associating them with specific molecular functions or biological processes remains a significant challenge.
 The sequences of IDRs are generally poorly conserved, so traditional bioinformatics approaches which rely on the conservation of amino acid sequences to identify homologous proteins and transfer annotations between them are largely unsuccessful when applied to IDRs.

### 2023-04-30 chapter2/chapter2.tex
Re-phrased last paragraph slightly

@@
 
-However, as IDRs likely evolve under multiple selective regimes where the constraint of different compositional, patterned, or motif features is highly specific to each, we expect a combination of statistical and computational methods will be needed to disentangle the complex forces driving IDR evolution.
+However, as IDRs are likely subject to multiple selective pressures where the constraints on different compositional, patterned, or motif features is highly specific to each, we anticipate a range of computational and experimental methods will be needed to disentangle the complex forces driving IDR evolution.
 Accordingly, while these results represent a significant step forward in relating sequence to function in IDRs, further studies exploring these and other approaches will undoubtedly reveal new insights into these pervasive but poorly understood regions of proteins.

### 2023-05-01 chapter2/figures/hierarchy/hierarchy.py
Added cluster labels and tweaked spacing

@@ -18,18 +18,18 @@ min_indel_columns = 5  # Indel rates below this value are set to 0
 min_aa_rate = 1
 min_indel_rate = 0.1
 
-clusters = [('A', '15086'),
-            ('B', '15170'),
-            ('C', '15160'),
-            ('D', '15070'),
-            ('E', '14415'),
-            ('F', '15217'),
-            ('G', '15078'),
-            ('H', '15021'),
-            ('I', '15161'),
-            ('J', '15191'),
-            ('K', '15143'),
-            ('L', '14937')]
+clusters = [('15086', 'A', 'extracellular structure'),
+            ('15170', 'B', 'Wnt signaling'),
+            ('15160', 'C', ''),
+            ('15070', 'D', 'nuclear pore organization'),
+            ('14415', 'E', 'voltage gated channel'),
+            ('15217', 'F', 'nuclear transport'),
+            ('15078', 'G', 'nucleotide transport'),
+            ('15021', 'H', 'cell cycle regulation'),
+            ('15161', 'I', 'synaptic signaling regulation'),
+            ('15191', 'J', 'histone methylation'),
+            ('15143', 'K', ''),
+            ('14937', 'L', 'chromatin assembly')]
 
 color1 = '#4e79a7'
 
@@ -121,7 +121,7 @@ legend_args = {'aa_group': ('Amino acid content', 'grey', ''),
                'motifs_group': ('Motifs', 'white', 4 * '\\')}
 group_labels = ['aa_group', 'charge_group', 'motifs_group', 'physchem_group', 'complexity_group']
 group_labels_nonmotif = ['aa_group', 'charge_group', 'physchem_group', 'complexity_group']
-gridspec_kw = {'width_ratios': [0.1, 0.7, 0.2], 'wspace': 0,
+gridspec_kw = {'width_ratios': [0.1, 0.65, 0.25], 'wspace': 0,
                'height_ratios': [0.975, 0.025], 'hspace': 0.01,
                'left': 0.05, 'right': 0.95, 'top': 0.95, 'bottom': 0.125}
 
@@ -176,7 +176,7 @@ for ax in [axs[1, 0], axs[1, 2]]:
 # Cluster blocks
 ax = axs[0, 2]
 id2idx = {tip.name: idx for idx, tip in enumerate(tree.tips())}
-for cluster_id, root_id in clusters:
[…]

### 2023-05-01 chapter2/chapter2.tex
Updated caption for hierarchy figure
Tweaked wording and punctuation in two sentences slightly

@@
 Because many sequences can yield a region with a specific composition, for example, this mode of constraint uncouples an IDR's fitness from its strict sequence of amino acid residues.
-Furthermore, in contrast to folded regions whose precise contacts and packing geometries are easily disrupted by amino acid substitutions, distributed features are robust to such changes as individual residues only weakly contribute to a region's fitness.
+Furthermore, in contrast to folded regions whose precise contacts and packing geometries are easily disrupted by amino acid substitutions, distributed features are robust to such changes, as individual residues only weakly contribute to a region's fitness.
 For example, a mutation at one site in an IDR that changes its net charge is easily reversed by subsequent compensatory mutations elsewhere in the region.
@@
 The AIC difference between the BM and OU models is measure of their relative goodness of fit to the data with a penalty for the number of parameters in each.
-Larger values indicate a better fit by the OU model.}
+Larger values indicate a better fit by the OU model.
+Clusters are indicated by rectangles on the right.
+The cluster descriptions are summaries of their enriched annotations and do not necessarily correspond to specific terms.}
 \label{fig:hierarchy}
@@
 While simulations can eliminate specific hypotheses, the substantial resources involved both in designing them and generating samples make exhaustively testing mechanisms of feature constraint by simulation impractical.
-We instead recommend a workflow that begins with an analysis using generic models of trait evolution to suggest specific hypotheses of constraint that are then tested with a simulation-based approach.
+We instead recommend a workflow that begins with an analysis using general models of trait evolution to suggest specific hypotheses of constraint that are then tested with a simulation-based approach.
 Another hybrid method involves using sequence permutations to test for the conservation of motifs or patterned features like kappa.
 In this method, a comparative model like BM provides the test statistic as before, and a sample of shuffled sequences approximates the null distribution.
-Because the background distribution of residues is preserved, this approach can specifically test for the conservation of motifs or patterned features in an IDR independent of the conservation of its composition.
+Because the background distribution of residues is preserved, this procedure can specifically test for the conservation of motifs or patterned features in an IDR independent of the conservation of its composition.
 

### 2023-05-01 chapter2/chapter2_supporting.tex
Added table of selected annotations

@@
 \end{figure}
+
+\clearpage  % Force display of all figures before table
+
+\footnotesize
+\begin{longtable}{|l|r|l|l|}
+\caption{\textbf{Selected enriched annotations in clusters.}}
+\label{stable:annotations}
+\hline
+\textbf{Cluster} & \multicolumn{1}{c|}{\textbf{\textit{p}-value}} & \textbf{Term ID} & \textbf{Term name}                                                    \endfirsthead
+\hline
+A                & 1.38E-03                                       & GO:0005201       & ~extracellular matrix structural constituent                          \\
+\hline
+A                & 3.38E-03                                       & GO:0043062       & ~extracellular structure organization                                 \\
+\hline
+A                & 9.48E-03                                       & GO:0030198       & ~extracellular matrix organization                                    \\
+\hline
+B                & 3.64E-04                                       & GO:0090263       & ~positive regulation of canonical Wnt signaling pathway               \\
+\hline
+B                & 4.32E-04                                       & GO:0060828       & ~regulation of canonical Wnt signaling pathway                        \\
+\hline
+B                & 5.04E-04                                       & GO:0030177       & ~positive regulation of Wnt signaling pathway                         \\
+\hline
+B                & 8.91E-04                                       & GO:0030111       & ~regulation of Wnt signaling pathway                                  \\
+\hline
+B                & 3.98E-03                                       & GO:0061629       & ~RNA polymerase II-specific DNA-binding transcription factor binding  \\
+\hline
+B                & 7.15E-03                                       & GO:0008134       & ~transcription factor binding                                         \\
+\hline
+D                & 2.44E-03                                       & GO:0006999       & ~nuclear pore organization                                            \\
+\hline
+D                & 3.23E-03                                       & GO:0006607       & ~NLS-bearing protein import into nucleus                              \\
+\hline
+D                & 3.60E-03                                       & GO:0008104       & ~protein localization                                                 \\
+\hline
+D                & 3.74E-03                                       & GO:0045184       & ~establishment of protein localization                                \\
+\hline
+D                & 3.78E-03                                       & GO:0044613       & ~nuclear pore central transport channel                               \\
+\hline
+D                & 3.78E-03                                       & GO:0017056       & ~structural constituent of nuclear pore                               \\
+\hline
+E                & 3.66E-04                                       & GO:0070161       & ~anchoring junction                                                   \\
+\hline
+E                & 7.11E-04                                       & GO:0030054       & ~cell junction                                                        \\
+\hline
+E                & 1.66E-03                                       & GO:0022843       & ~voltage-gated cation channel activity                                \\
+\hline
+E                & 2.12E-03                                       & GO:0022832       & ~voltage-gated channel activity                                       \\
+\hline
+E                & 2.12E-03                                       & GO:0005244       & ~voltage-gated ion channel activity                                   \\
+\hline
+F                & 2.92E-04                                       & GO:0033365       & ~protein localization to organelle                                    \\
+\hline
+F                & 4.08E-04                                       & GO:0051179       & ~localization                                                         \\
+\hline
+F                & 5.43E-04                                       & GO:0034504       & ~protein localization to nucleus                                      \\
+\hline
+F                & 8.68E-04                                       & GO:0034613       & ~cellular protein localization                                        \\
+\hline
+F                & 1.57E-03                                       & GO:0008104       & ~protein localization                                                 \\
+\hline
+F                & 1.67E-03                                       & GO:0006606       & ~protein import into nucleus                                          \\
+\hline
+F                & 2.10E-03                                       & GO:0051170       & ~import into nucleus                                                  \\
+\hline
+G                & 2.40E-04                                       & GO:0005459       & ~UDP-galactose transmembrane transporter activity                     \\
+\hline
+G                & 2.40E-04                                       & GO:0005338       & ~nucleotide-sugar transmembrane transporter activity                  \\
+\hline
+G                & 2.40E-04                                       & GO:0015165       & ~pyrimidine nucleotide-sugar transmembrane transporter activity       \\
+\hline
+G                & 2.40E-04                                       & GO:0072334       & ~UDP-galactose transmembrane transport                                \\
+\hline
+G                & 2.40E-04                                       & GO:0090481       & ~pyrimidine nucleotide-sugar transmembrane transport                  \\
+\hline
+G                & 2.40E-04                                       & GO:0015780       & ~nucleotide-sugar transmembrane transport                             \\
+\hline
+G                & 1.41E-03                                       & GO:0015932       & ~nucleobase-containing compound transmembrane transporter activity    \\
+\hline
+G                & 2.33E-03                                       & GO:1901505       & ~carbohydrate derivative transmembrane transporter activity           \\
+\hline
+G                & 2.33E-03                                       & GO:1901264       & ~carbohydrate derivative transport                                    \\
+\hline
+H                & 2.14E-04                                       & GO:0007113       & ~endomitotic cell cycle                                               \\
+\hline
+H                & 1.51E-03                                       & GO:0051726       & ~regulation of cell cycle                                             \\
+\hline
+I                & 5.02E-04                                       & GO:0007616       & ~long-term memory                                                     \\
+\hline
+I                & 6.03E-04                                       & GO:0099177       & ~regulation of trans-synaptic signaling                               \\
+\hline
+I                & 6.03E-04                                       & GO:0050804       & ~modulation of chemical synaptic transmission                         \\
+\hline
+I                & 8.90E-04                                       & GO:0010646       & ~regulation of cell communication                                     \\
+\hline
+I                & 9.41E-04                                       & GO:0023051       & ~regulation of signaling                                              \\
+\hline
+J                & 1.34E-03                                       & GO:0016829       & ~lyase activity                                                       \\
+\hline
+J                & 1.47E-03                                       & GO:0008172       & ~S-methyltransferase activity                                         \\
+\hline
+J                & 2.37E-03                                       & GO:0042800       & ~histone methyltransferase activity (H3-K4 specific)                  \\
+\hline
+J                & 3.76E-03                                       & GO:0034708       & ~methyltransferase complex                                            \\
+\hline
+J                & 4.79E-03                                       & GO:0051568       & ~histone H3-K4 methylation                                            \\
+\hline
+J                & 8.43E-03                                       & GO:0035097       & ~histone methyltransferase complex                                    \\
+\hline
+L                & 2.51E-03                                       & GO:0031507       & ~heterochromatin assembly                                             \\
+\hline
+L                & 2.51E-03                                       & GO:0031497       & ~chromatin assembly                                                   \\
+\hline
+L                & 3.00E-03                                       & GO:0070828       & ~heterochromatin organization                                         \\
+\hline
+L                & 3.91E-03                                       & GO:0003677       & ~DNA binding                                                          \\
+\hline
+L                & 4.70E-03                                       & GO:0003676       & ~nucleic acid binding                                                 \\
+\hline
+\end{longtable}
 

### 2023-05-01 chapter2/chapter2.tex
Removed outline for discussion paragraph on PCAs
Updated text with examples of clusters

@@
 
-\subsection{A model of constrained evolution reveals conserved features}
+\subsection{A model of constrained evolution reveals patterns of conserved features}
 Though the BM process permits the inference of the rates of feature evolution after accounting for the phylogenetic relationships between species, it does not directly test for their conservation.
@@
 Clusters are indicated by rectangles on the right.
-The cluster descriptions are summaries of their enriched annotations and do not necessarily correspond to specific terms.}
+The cluster descriptions are summaries of selected significantly enriched annotations ($p < 0.01$) and do not necessarily correspond to specific terms.}
 \label{fig:hierarchy}
@@
 However, we identify at least 12 clusters with a strong and consistent patterns of constraint, with many having significant associations with certain processes, functions, or compartments.
-For example, cluster X is strongly enriched in Y, and cluster A contains many proteins involved in B.
-Though these preliminary results suggest the existence of high-level relationships between certain patterns of conserved features and function, further insight into the biophysical and mechanistic details that underpin these associations will rely on more in-depth analyses of specific clusters and the properties of their constituent IDRs.
+For example, cluster B is strongly enriched in proteins associated with the Wnt signaling pathway, and cluster F contains many proteins involved in nuclear transport.
+(Selected annotations corresponding to the summaries in Fig.~\ref{fig:hierarchy} are given in Table~\ref{stable:annotations}.)  Though these preliminary results suggest the existence of high-level relationships between certain patterns of conserved features and function, further insight into the biophysical and mechanistic details that underpin these associations will rely on more in-depth analyses of specific clusters and the properties of their constituent IDRs.
 
@@
 %        3. Note about variances of parameters as shown in CV graphs
-
-% IV. Feature analyses
-%    A.
-Though no discrete clusters are visible in the largest two and three principal components of the feature root and rate distribution, they only account for X\% and Y\% of the total variation between regions, respectively.
-Thus, a majority of the variation is contained in other components.
-%    B.
-The success of the hierarchical clustering at identifying common patterns of feature constraint suggests this method, or other non-linear dimensionality reductions like UMAP or t-SNE, may also reveal additional structure latent in the high-dimesional rate distributions.
-However, as these methods are designed primarily for visualization, they can significant distort the global structure of high-dimensional data.
-Thus, these embeddings are generally not suitable for downstream analyses.
-Instead, semi-supervised approaches for feature learning may yield more meaningful representations.
-In fact, a recent study used a semi-supervised learning approach that identified orthologous IDRs to learn novel disorder-associated features.
-Though they used a simple neural network that aided in interpretability, other similar semi-supervised approaches have used more complex networks to learn rich representations of protein regions.
-Similar approaches may yield representations of the space of disordered proteins that are more clearly visualized and manipulated.
 

### 2023-05-01 chapter0/chapter0.tex
Clarified wording around Zarin2019 summary in introduction

@@
 By the mid 2010s several studies had demonstrated evidence of such constraint in the flexibility~\cite{Daughdrill2007}, chemical composition~\cite{Moesa2012}, net charge~\cite{Zarin2017}, or charge distribution of IDRs~\cite{Beh2012}.
-While these initial studies were generally restricted to specific features or proteins, in 2019 Zarin \textit{et al.} showed various features from a comprehensive set of IDR-associated properties were conserved in IDRs across the entire yeast proteome~\cite{Zarin2019}.
-By comparing the observed values of these features against those generated under a simulated model of evolution, they clustered IDRs by their ``evolutionary signatures,'' \textit{i.e.} patterns of conserved features.
-Furthermore, they identified specific biological functions associated with these groups, which for the first time provided a global view of the relationship between sequence and function in IDRs.
+These initial studies demonstrating evidence of constraint were generally restricted to specific features or proteins.
+However, by comparing the observed values of various IDR-associated properties against those generated under a simulated model of evolution, in 2019 Zarin \textit{et al.} showed most IDRs across the entire yeast proteome contain conserved features.
+Furthermore, they identified clusters of IDRs with common ``evolutionary signatures,'' \textit{i.e.} patterns of conserved features, which were associated with specific biological functions.
+This analysis for the first time provided a global view of the relationship between sequence and function in IDRs.
 

### 2023-05-01 chapter2/chapter2.tex
Clarified wording around Zarin2019 summary in introduction

@@
 
-While the studies which had initially demonstrated evidence of constraint were generally restricted to specific features or proteins, in 2019 Zarin \textit{et al.} showed various features from a comprehensive set of IDR-associated properties were conserved in IDRs across the entire yeast proteome~\cite{Zarin2019}.
-By comparing the observed values of these features against those generated under a simulated model of evolution, they clustered IDRs by their ``evolutionary signatures,'' \textit{i.e.} patterns of conserved features.
-Furthermore, they identified specific biological functions associated with these groups, which for the first time provided a global view of the relationship between sequence and function in IDRs.
-A follow-up analysis expanded on this initial finding by applying techniques from machine learning and statistics to predict the functions of individual IDRs using their evolutionary signatures~\cite{Zarin2021}.
+The initial studies demonstrating evidence of constraint were generally restricted to specific features or proteins.
+However, by comparing the observed values of various IDR-associated properties against those generated under a simulated model of evolution, Zarin \textit{et al.} showed most IDRs across the entire yeast proteome contain conserved features.
+Furthermore, they identified clusters of IDRs with common ``evolutionary signatures,'' \textit{i.e.} patterns of conserved features, which were associated with specific biological functions.
+This analysis for the first time provided a global view of the relationship between sequence and function in IDRs.
+A follow-up study then expanded on this initial finding by applying techniques from machine learning and statistics to predict the functions of individual IDRs using their evolutionary signatures~\cite{Zarin2021}.
 

### 2023-05-01 chapter2/chapter2.tex
Moved outline into discussion and added more subsections to group related paragraphs

@@
 
-% CONCLUSION
+\section{Discussion}
 % I. Summary of results
-%    A. In this study, we demonstrate the first evidence of evolutionary constraint in IDRs across the \textit{Drosophila} proteome.
-%    B. Most immediately, this work generates hypotheses of function related to sequence features that will permit the functional dissection of IDRs.
+%    A. In this study, we applied several phylogenetic models to IDRs to interrogate the evolution of their sequences and molecular features.
+%    B. Most significantly, we demonstrate evidence of widespread evolutionary constraint in IDRs within the \textit{Drosophila} proteome
+%    C. Furthermore, many IDRs can be grouped into discrete clusters on the basis of these evolutionary signatures
+%    B. Most immediately, this work generates hypotheses of function related to sequence features that will permit the functional dissection of IDRs
 %    C. More broadly, however, it illustrates constraint of distributed features characterizes the evolution of some IDRs across multiple systems.
+%    D.
+However, these clusters require careful interpretation.
+The cluster descriptions summarize only a subset of the annotations with statistically significant enrichments.
+All clusters had many enriched annotations, but the summaries correspond to sets of enrichments with consisten .
+Because the significance tests were not corrected for multiple testing or controlled for their false discovery rate, many annotations are ,
 
-% II. Exchangeability and rate matrices
+% II. IDRs have characteristic patterns of amino acid substitution
 %    A. In addition to these global patterns of conserved features, we find IDRs exhibit other characteristic patterns of evolution.
@@
 
-\section{Discussion}
+\subsection{Disorder is correlated with many molecular features}
 Though the mutations generated by evolution are not a random or exhaustive sample of sequence space, they are perturbations of a common ancestor which can reveal the relationship between a region's biophysical properties and its propensity for disorder.
@@
 
-As discussed by Zarin \textit{et al.}, the interpretation of evolutionary signatures is complicated by several methodological caveats~\cite{Zarin2019}.
+\subsection{A workflow for future evolutionary analyses of IDRs}
+As discussed by Zarin \textit{et al.}, the interpretation of evolutionary signatures is complicated by several methodological limitations~\cite{Zarin2019}.
 For example, because IDRs are identified as contiguous segments of high predicted disorder, their boundaries are defined by adjacent structural elements.
@@
 Another challenge is the significant overlap in the definitions of many features induces strong correlations that preclude straightforward quantitative manipulations or interpretations of an IDR's evolutionary signature.
-The original authors have since addressed this limitation in subsequent studies by applying machine learning methods to perform feature selection or learn features directly from alignments of IDRs~\cite{Zarin2021, Lu2022}.
+The original authors have since addressed this in subsequent studies by applying machine learning methods to perform feature selection or learn features directly from alignments of IDRs~\cite{Zarin2021, Lu2022}.
 However, integrating these methods into a unified phylogenetic comparative framework will require further effort.

### 2023-05-01 chapter2/chapter2.tex
Added chapter 2 abstract

@@
 \noindent
-Here's where my abstract will go!
+Intrinsically disordered regions (IDRs) are segments of proteins without stable three-dimensional structures.
+As this flexibility allows them to interact with diverse binding partners, IDRs play key roles in cell signaling and gene expression.
+Despite the prevalence and importance of IDRs in eukaryotic proteomes and various biological processes, associating them with specific molecular functions remains a significant challenge due to their high rates of sequence evolution.
+However, by comparing the observed values of various IDR-associated properties against those generated under a simulated model of evolution, a recent study found most IDRs across the entire yeast proteome contain conserved features.
+Furthermore, it showed clusters of IDRs with common ``evolutionary signatures,'' \textit{i.e.} patterns of conserved features, were associated with specific biological functions.
+To determine if similar patterns of conservation are found in the IDRs of other systems, in this work we apply a series of phylogenetic models to over 8,500 orthologous IDRs identified in the \textit{Drosophila} genome to dissect the forces driving their evolution.
+By comparing models of constrained and unconstrained continuous trait evolution, \textit{i.e.} the Brownian motion and Ornstein-Uhlenbeck models, we identify specific clusters of IDRs with shared patterns of constraint.
+As in yeast, these clusters are enriched for proteins with specific functional annotations, which suggests the preservation of distributed features and their associated functions is a widespread mechanism of IDR evolution.
 \end{abstract}

### 2023-05-01 chapter2/chapter2.tex
Re-worded contrasts explanation for clarity

@@
 In the most severe cases, traits derived from clades of closely-related species can effectively act as duplicate observations, which can yield spurious correlations.
-We therefore applied the method of contrasts, which takes differences between adjacent nodes in the phylogenetic tree relating the species to generate a set of ``contrasts,'' to both the scores and the features.
-Under some general assumptions of the underlying evolutionary process, contrasts are independent and identically-distributed and are therefore appropriate for use in correlation analyses.
+We therefore applied the method of contrasts to both the scores and the features.
+This algorithm takes differences between adjacent nodes in the phylogenetic tree relating the species to generate ``contrasts,'' which, under some general assumptions of the underlying evolutionary process, are independent and identically-distributed and therefore appropriate for use in correlation analyses.
 The resulting feature contrasts have varying degrees of correlation with the score contrasts (Fig.~\ref{fig:scores}B-C).

### 2023-05-01 chapter0/chapter0.tex
Added subsections to discussions

@@
 \section{Aims}
-
 Despite recent advances in identifying IDRs and their conformational ensembles from their sequences alone~\cite{Mao2010, vanderLee2014, Das2015, Necci2021}, the relationship between the sequence and function remains poorly understood.

### 2023-05-01 chapter1/chapter1.tex
Added subsections to discussions

@@
 \section{Discussion}
+\subsection{Developments in genome assembly and annotation challenge existing methods of orthology inference}
 The orthologous groups and alignments yielded by this pipeline are a valuable resource for comparative studies of gene birth/death processes and protein evolution at the level of both entire proteomes and specific gene families in the \textit{Drosophila} genus.
@@
 
+\subsection{Comprehensive bioinformatic analyses of proteins will depend on splice-sensitive alignments}
 While the combination of these two steps yielded high-quality alignments that are suitable for further analyses, they are an \textit{ad hoc} fix for underlying issues with the gene models and alignment algorithms.
@@
 
+\subsection{Accessible computational tools will facilitate future comparative studies}
 Though benchmarks are available for optimizing and comparing methods of orthology inference, the metrics are calculated over a set of reference genomes which are sparsely sampled over a broad taxonomic range, so it is unclear if they are informative for method designed to yield robust inferences when the genomes are highly related~\cite{Nevers2022}.

### 2023-05-01 chapter2/chapter2.tex
Added subsections to discussions

@@
 
-\subsection{A workflow for future evolutionary analyses of IDRs}
+\subsection{Future evolutionary analyses of IDRs will require a multimodal approach}
 As discussed by Zarin \textit{et al.}, the interpretation of evolutionary signatures is complicated by several methodological limitations~\cite{Zarin2019}.

### 2023-05-01 chapter0/chapter0.tex
Tweaked wording of abstract

@@
 Though proteins are linear chains of simpler building blocks called amino acids, many proteins accomplish their functions by folding into well-defined three-dimensional structures.
-For many years scientists believed fixed structures were necessary for protein function, but by the early 2000s evidence had accumulated that unstructured or intrinsically disordered regions (IDRs) were ubiquitous in proteins.
-Furthermore, these regions were essential for many cellular processes such as signaling and regulation.
+For many years scientists believed fixed structures were necessary for protein function, but by the early 2000s evidence had accumulated that segments with no fixed spatial relationship between their atoms are ubiquitous in proteins.
+Furthermore, because these intrinsically disordered regions (IDRs) are highly flexible and can therefore interact with diverse binding partners, they are essential for many cellular processes related to signaling and regulation.
 Although our understanding of the structure and function of IDRs has grown significantly over the past two decades, predicting their functions from their sequences of amino acids remains a significant challenge.
-Because IDRs are structurally unconstrained, their sequences evolve rapidly and are therefore not amenable to traditional bioinformatics analyses which depend on the precise order of amino acids to make comparisons with known proteins.
+Because IDRs are structurally unconstrained, their sequences evolve rapidly and are therefore not amenable to traditional bioinformatics techniques which depend on the precise order of amino acids to make comparisons with known proteins.
 There is increasing evidence, though, that IDRs conserve distributed features such as their chemical composition or net charge, and a recent study clustered IDRs with similar patterns of conserved features into groups with distinct functions.
 This study, however, was restricted to IDRs in a set of yeast genomes, so it is unclear if these global relationships between conserved features and function are unique to yeast or a general property of IDR evolution.
-Thus, in this work I conduct a series of evolutionary analyses of IDRs in the genomes of 33 different species of fruit flies to detect patterns of conservation.
+Thus, in this work I conduct a series of evolutionary analyses of IDRs in the genomes of 33 different species of fruit flies to dissect the forces driving their evolution and detect patterns of conservation.
 \end{abstract}
@@
 Thus, in the first chapter I develop a novel method for identifying orthologs which addresses this shortcoming and apply it to 33 \textit{Drosophila} genomes to generate a set of aligned orthologs (Fig.~\ref{fig:aims_overview}A-B).
-In the second chapter I then identify rapidly evolving IDRs in these alignments and analyse them with a variety of evolutionary models to detect patterns of conservation (Fig.~\ref{fig:aims_overview}C-F).
+In the second chapter I then identify rapidly evolving IDRs in these alignments and analyse them with a variety of evolutionary models to dissect the forces driving their evolution and detect patterns of conservation (Fig.~\ref{fig:aims_overview}C-F).
 Finally, in the third chapter I discuss several software tools and tutorials for fitting statistical models to data, which were created while pursuing the previous aims.

### 2023-05-01 chapter2/chapter2.tex
Tweaked wording of abstract

@@
 To determine if similar patterns of conservation are found in the IDRs of other systems, in this work we apply a series of phylogenetic models to over 8,500 orthologous IDRs identified in the \textit{Drosophila} genome to dissect the forces driving their evolution.
-By comparing models of constrained and unconstrained continuous trait evolution, \textit{i.e.} the Brownian motion and Ornstein-Uhlenbeck models, we identify specific clusters of IDRs with shared patterns of constraint.
+By comparing models of constrained and unconstrained continuous trait evolution using the Brownian motion and Ornstein-Uhlenbeck models, respectively, we identify specific clusters of IDRs with shared patterns of constraint.
 As in yeast, these clusters are enriched for proteins with specific functional annotations, which suggests the preservation of distributed features and their associated functions is a widespread mechanism of IDR evolution.

### 2023-05-01 main.tex
Added main abstract

@@
 \end{center}
-Here's the text for my abstract!
+
+Living systems are governed by the interactions between large collections of atoms called macromolecules.
+The most important class of these macromolecules are proteins, which are the molecular machines that carry out a cell's processes.
+Though proteins are linear chains of simpler building blocks called amino acids, many proteins accomplish their functions by folding into well-defined three-dimensional structures.
+For many years scientists believed fixed structures were necessary for protein function, but by the early 2000s evidence had accumulated that segments with no fixed spatial relationship between their atoms are ubiquitous in proteins.
+Furthermore, because these intrinsically disordered regions (IDRs) are highly flexible and can therefore interact with diverse binding partners, they are essential for many cellular processes related to signaling and regulation.
+
+Although our understanding of the structure and function of IDRs has grown significantly over the past two decades, predicting their functions from their sequences of amino acids remains a significant challenge.
+Because IDRs are structurally unconstrained, their sequences evolve rapidly and are therefore not amenable to traditional bioinformatics techniques which depend on the precise order of amino acids to make comparisons with known proteins.
+There is increasing evidence, though, that IDRs conserve distributed features such as their chemical composition or net charge, and a recent study clustered IDRs with similar patterns of conserved features into groups with distinct functions.
+This study, however, was restricted to IDRs in a set of yeast genomes, so it is unclear if these global relationships between conserved features and function are unique to yeast or a general property of IDR evolution.
+Thus, in this work I conduct a series of evolutionary analyses of IDRs in the genomes of 33 different species of fruit flies to detect patterns of conservation.
+
+These comparisons, however, require the identification of IDRs with common ancestry which perform equivalent functions across many distinct organisms.
+Since the first genomes were sequenced in the late 1990s, researchers have developed techniques for identifying and aligning such proteins, called orthologs.
+While these methods are generally effective, they are conducted by automated computational pipelines and prone to errors when processing the highly divergent sequences that characterize many IDRs.
+The evolutionary relationships between the genomes of closely related species generally make such mistakes easier to identify, and fortunately over the past five years advances in DNA sequencing technology have yielded dramatic increases in the number of sequenced genomes in the \textit{Drosophila} genus.
+However, because the existing methods for ortholog identification were designed for fewer or more distantly related genomes, they do not fully leverage such genomic redundancy to minimize errors.
+
+Thus, in the first chapter I develop a novel method for identifying orthologs which addresses this shortcoming and apply it to 33 \textit{Drosophila} genomes to generate a set of aligned orthologs.
+In the second chapter I then identify rapidly evolving IDRs in these alignments and analyse them with a variety of evolutionary models dissect the forces driving their evolution and detect patterns of conservation.
+Finally, in the third chapter I discuss several software tools and tutorials for fitting statistical models to data, which were created while pursuing the previous aims.
+
 \clearpage
-
 \setcounter{page}{1}  % Reset page counter after abstract

### 2023-05-01 chapter2/chapter2.tex
Fixed wrong citation for Felsenstein contrasts

@@
 Brownian motion (BM) model parameters were calculated with two methods.
-The first used Felsenstein's contrasts algorithm to efficiently calculate roots and contrasts for the disorder scores and features of each region~\cite{Felsenstein1973, Felsenstein1981}.
+The first used Felsenstein's contrasts algorithm to efficiently calculate roots and contrasts for the disorder scores and features of each region~\cite{Felsenstein1973, Felsenstein1985}.
 Rates were calculated as the mean of the squares of the contrasts.

### 2023-05-02 chapter2/chapter2.tex
Added more citations

@@
 Furthermore, when IDRs contain multiple copies of a motif that jointly mediate a high-avidity interaction~\cite{Tompa2014} or a graded response to a signal via the accumulation of multiple phosphorylations~\cite{Wright2014}, the individual motifs are under weak selective constraints.
-As a result, though SLiMs are encoded by specific sequences, in some contexts they may evolve as distributed features that characterize IDRs as a whole rather than specific sites within them.
+As a result, though SLiMs are encoded by specific sequences, in some contexts they may evolve as distributed features that characterize IDRs as a whole rather than specific sites within them~\cite{Tan2010}.
 
@@
 Because the significance tests were not corrected for multiple testing or controlled for their false discovery rate, we caution against over-interpreting this result and instead consider it as a hypothesis for further investigation.
-In general, however, these regions with rapidly evolving disorder scores may correspond to molecular recognition features (MoRFs), which are modules in IDRs that undergo a disorder-to-order transition on binding their targets~\cite{vanderLee2014}.
+In general, however, these regions with rapidly evolving disorder scores may correspond to molecular recognition features (MoRFs), which are modules in IDRs that undergo a disorder-to-order transition on binding their targets~\cite{Mohan2006, vanderLee2014}.
 Because MoRFs already exist on the boundary between disorder and structure, small changes in the biophysical properties of these regions may have disproportionate effects on their structural ensembles.
@@
 For comparison with the Ornstein-Uhlenbeck (OU) model, the BM parameters were calculated by maximizing the likelihood.
-The OU model parameters were also calculated via maximum likelihood estimation.
+The OU model parameters were also calculated via maximum likelihood estimation as described in Butler \textit{et al.}~\cite{Butler2004}.
 To ensure parameter identfiability and estimate consistency, the root was treated as a random variable~\cite{Ho2013, Ho2014}.

### 2023-05-03 chapter2/chapter2.tex
Corrected is -> are in discussion

@@
 
-However, as IDRs are likely subject to multiple selective pressures where the constraints on different compositional, patterned, or motif features is highly specific to each, we anticipate a range of computational and experimental methods will be needed to disentangle the complex forces driving IDR evolution.
+However, as IDRs are likely subject to multiple selective pressures where the constraints on different compositional, patterned, or motif features are highly specific to each, we anticipate a range of computational and experimental methods will be needed to disentangle the complex forces driving IDR evolution.
 Accordingly, while these results represent a significant step forward in relating sequence to function in IDRs, further studies exploring these and other approaches will undoubtedly reveal new insights into these pervasive but poorly understood regions of proteins.

### 2023-05-03 chapter2/chapter2.tex
Added new citation for motifs acting as switches

@@
 Instead, as SLiMs are compact and often highly degenerate at some positions, they can arise de novo from a small number of mutations and therefore have high rates of turnover.
-Furthermore, when IDRs contain multiple copies of a motif that jointly mediate a high-avidity interaction~\cite{Tompa2014} or a graded response to a signal via the accumulation of multiple phosphorylations~\cite{Wright2014}, the individual motifs are under weak selective constraints.
+Furthermore, when IDRs contain multiple copies of a motif that jointly mediate a high-avidity interaction~\cite{Tompa2014} or a graded response to a signal via the accumulation of multiple phosphorylations~\cite{VanRoey2012, Wright2014}, the individual motifs are under weak selective constraints.
 As a result, though SLiMs are encoded by specific sequences, in some contexts they may evolve as distributed features that characterize IDRs as a whole rather than specific sites within them~\cite{Tan2010}.

### 2023-05-08 chapter0/chapter0.tex
Changed genetic -> biological because more generic
Implemented changes from Susan's comments (largely typos and clarifications)

@@
 For example, lactose is a sugar found in milk that requires a specific protein, lactase, to metabolize properly.
-Many humans who can digest milk products in childhood lose this ability in adulthood because they stop producing lactose.
+Many humans who can digest milk products in childhood lose this ability in adulthood because they stop producing lactase.
 As a result, lactose in dairy products passes undigested into the colon where it is broken down by bacteria, causing symptoms such as bloating and diarrhea.
@@
 However, proteins are only synthesized from the 20 ``canonical'' amino acids.
-Another two (selenocysteine and pyrrolysine) are incorporated via a distinct mechanism under rare circumstances and are therefore considered non-standard.} Though each amino acid is chemically unique, they share a common backbone composed of two distinct and complementary receptor and donor sites for chemical bonds.
+Another two (selenocysteine and pyrrolysine) are incorporated via a distinct mechanism under rare circumstances and are therefore considered non-standard.} Though each amino acid is chemically unique, they share a common backbone composed of two distinct and complementary ``receptor'' and ``donor'' sites for chemical bonds.
 Thus, in a protein the amino acids are bonded in a linear chain like beads on a string.
@@
 Because each amino acid has unique geometric and chemical properties that influence the energetics of these interactions, a protein's three-dimensional structure is encoded by the sequence of amino acids that compose it.
-A protein's function is in turn a direct result of its structure.
+A protein's function is in part a result of its structure.
 For example, the structure of hemoglobin precisely positions its amino acids and a helper molecule called a heme group to create a pocket that can stably but reversibly bind oxygen.
@@
 One of the most useful scales for describing an amino acid's overall effect on protein structure is hydrophobicity, which measures a molecule's tendency to associate with water.
-Molecules which attract water are hydrophilic (water loving), and molecules which repel water are hydrophobic (water fearing).
-Hydrophilic molecules, like sugar or alcohol, easily dissolve in water, whereas hydrophobic molecules, like fats and oils, remain separate.
-Since the cellular environment is largely water, the hydrophobic residues in proteins tend to aggregate into a hydrophobic core.
+Molecules which interact with water are hydrophilic (water loving), and molecules which do not are hydrophobic (water fearing).
+Thus, hydrophilic molecules, like sugar or alcohol, easily dissolve in water, whereas hydrophobic molecules, like fats and oils, remain separate.
+Since the cellular environment is largely water, the hydrophobic residues in proteins tend to cluster into a hydrophobic core.
 This hydrophobic collapse is a major driving force in the early stages of protein folding, so a protein's relative number of hydrophilic and hydrophobic residues is a key determinant of whether it is folded or disordered.
@@
 Furthermore, these signals and interactions are easily tuned because PTMs can modulate their meanings and strengths, respectively, by modifying only a few residues.
-Thus, IDRs are like ``molecular computers'' that integrate complex data and respond accordingly to execute different genetic ``programs.''
+Thus, IDRs are like ``molecular computers'' that integrate complex data and respond accordingly to execute different biological ``programs.''
 
@@
 Though gene expression involves dozens, if not hundreds, of distinct biochemical reactions, most occur as steps within two major processes: transcription and translation.
-In transcription, the sequence of a protein encoded the cell's DNA is transcribed into an intermediate molecule called messenger RNA (mRNA).
+In transcription, the sequence of a protein encoded in the cell's DNA is transcribed into an intermediate molecule called messenger RNA (mRNA).
 mRNA is also a kind of nucleic acid, and is chemically closely related to DNA.
@@
 \section{Aims}
-Despite recent advances in identifying IDRs and their conformational ensembles from their sequences alone~\cite{Mao2010, vanderLee2014, Das2015, Necci2021}, the relationship between the sequence and function remains poorly understood.
+Despite recent advances in identifying IDRs and their conformational ensembles from their sequences alone~\cite{Mao2010, vanderLee2014, Das2015, Necci2021}, the relationship between their sequence and function remains poorly understood.
 In contrast, predictions of structure and function are readily available for many folded domains.
@@
 Thus, the specific sequence of a folded domain constitutes its unique ``signature.'' Traditional bioinformatics techniques use these signatures to detect similar protein sequences and transfer structural and functional information between them~\cite{Camacho2009, Eddy2009, Mistry2020}.
-Though this approach is simple in principle, it is extremely difficult to perfect in practice, and fully mapping the relationship between the sequence and structure of folded domains alone was an active area of research for decades.
+Though this approach is simple in principle, it is extremely difficult to perfect in practice, and fully mapping the relationship between the sequence and structure of folded domains alone hsa been an active area of research for decades.
 Part of the challenge is the available data is extremely sparse relative to the sheer number of possible protein sequences.
@@
 The concordance of these results with the previously identified IDR clusters would also have profound implications for the broader mechanisms of IDR evolution.
-For example, the absence of global patterns of evolutionary signatures across IDRs in \textit{Drosophila} would suggest they are property of IDRs which is unique to yeast.
+For example, the absence of global patterns of evolutionary signatures across IDRs in \textit{Drosophila} would suggest they are a property of IDRs which is unique to yeast.
 In contrast, the identification of clusters similar to those in yeast would indicate the existence of a taxonomy of IDRs which is conserved across the tree of life.
@@
 However, because the existing methods for ortholog identification were designed for fewer or more distantly related genomes, they do not fully leverage such genomic redundancy to minimize errors.
-Thus, in the first chapter I develop a novel method for identifying orthologs which addresses this shortcoming and apply it to 33 \textit{Drosophila} genomes to generate a set of aligned orthologs (Fig.~\ref{fig:aims_overview}A-B).
-In the second chapter I then identify rapidly evolving IDRs in these alignments and analyse them with a variety of evolutionary models to dissect the forces driving their evolution and detect patterns of conservation (Fig.~\ref{fig:aims_overview}C-F).
-Finally, in the third chapter I discuss several software tools and tutorials for fitting statistical models to data, which were created while pursuing the previous aims.
+Thus, in the first chapter, I develop a novel method for identifying orthologs which addresses this shortcoming and apply it to 33 \textit{Drosophila} genomes to generate a set of aligned orthologs (Fig.~\ref{fig:aims_overview}A-B).
+In the second chapter, I then identify rapidly evolving IDRs in these alignments and analyse them with a variety of evolutionary models to dissect the forces driving their evolution and detect patterns of conservation (Fig.~\ref{fig:aims_overview}C-F).
+Finally, in the third chapter, I discuss several software tools and tutorials for fitting statistical models to data, which were created while pursuing the previous aims.

### 2023-05-08 chapter0/chapter0.tex
Added citation for hemoglobin background

@@
 However, people affected by sickle cell disease have a mutation in the sequence of their hemoglobin proteins which causes it to malfunction.
-Frequently this mutation is a single change where the sixth amino acid in the sequence, glutamate, is substituted for valine.
+Frequently this mutation is a single change where the sixth amino acid in the sequence, glutamate, is substituted for valine~\cite{Voet2015}.
 This creates a sticky patch on the surface of hemoglobin, and under low-oxygen conditions normal hemoglobin changes shape to expose a sticky patch on its surface as well.
@@
 In some cases, folded domains have multiple structures, or conformations, which are related to different functional states.
-For example, hemoglobin has two forms, traditionally called the T and R states.
+For example, hemoglobin has two forms, traditionally called the T and R states~\cite{Voet2015}.
 The T state is hemoglobin's oxygen-free form, but on binding oxygen its structure shifts to the R state.

### 2023-05-08 chapter2/chapter2.tex
Made some minor wording changes

@@
 For example, cluster B is strongly enriched in proteins associated with the Wnt signaling pathway, and cluster F contains many proteins involved in nuclear transport.
-(Selected annotations corresponding to the summaries in Fig.~\ref{fig:hierarchy} are given in Table~\ref{stable:annotations}.)  Though these preliminary results suggest the existence of high-level relationships between certain patterns of conserved features and function, further insight into the biophysical and mechanistic details that underpin these associations will rely on more in-depth analyses of specific clusters and the properties of their constituent IDRs.
+(Selected annotations corresponding to the summaries in Fig.~\ref{fig:hierarchy} are given in Table~\ref{stable:annotations}.) Though these preliminary results suggest the existence of a high-level relationship between certain patterns of conserved features and function, further insight into the mechanisms that underpin these associations will rely on more in-depth analyses of specific clusters and the properties of their constituent IDRs.
 
@@
 As the formation of a hydrophobic core is a major driving force in protein folding, this relationship is expected.
-However, the strength of these correlations indicate that a region's relative proportion of hydrophilic and hydrophobic residues is, to a first approximation, the largest determinant of intrinsic disorder.
+However, the strength of these correlations indicate that a region's relative proportion of hydrophilic and hydrophobic residues is, to a first approximation, the largest determinant of predicted intrinsic disorder.
 
@@
 In general, however, these regions with rapidly evolving disorder scores may correspond to molecular recognition features (MoRFs), which are modules in IDRs that undergo a disorder-to-order transition on binding their targets~\cite{Mohan2006, vanderLee2014}.
-Because MoRFs already exist on the boundary between disorder and structure, small changes in the biophysical properties of these regions may have disproportionate effects on their structural ensembles.
+Because MoRFs already exist on the boundary between disorder and structure, small changes in the biophysical properties of these regions may have large effects on their structural ensembles.
 Thus, the most variable disorder scores may reflect instances where a mutation triggered a ``phase transition'' between largely structured or disordered native states.

### 2023-05-09 chapter0/chapter0.tex
Added citation for another mutational scanning study of ADs

@@
 In contrast, activations domains interact with the transcriptional machinery and are highly enriched in IDRs~\cite{Liu2006}.
-Decades of research have identified many of the proteins involved in these interactions as well as some common properties of their sequences~\cite{Ma1987, Sigler1988, Gerber1994, Arnold2018, Staller2018, Erijman2020, Sanborn2021}.
+Decades of research have identified many of the proteins involved in these interactions as well as some common properties of their sequences~\cite{Ma1987, Sigler1988, Gerber1994, Arnold2018, Ravarani2018, Staller2018, Erijman2020, Sanborn2021}.
 Despite this progress, however, the precise mechanisms by which transcription factors locate their DNA targets and recruit the transcriptional machinery to initiate expression of specific genes remains an unsolved problem in molecular biology.

### 2023-05-09 chapter2/chapter2.tex
Added reference to CV figures for matrices
Converted substitution outline into paragraph

@@
 As the coefficients model the one-way rate of substitutions between residues with the vertical and horizontal axes indicating the initial and target residues, respectively, this suggests a net flux towards a more disorder-like composition.
+However, the coefficients between the disorder-depleted and -enriched classes of residues for both the exchangeability and rate matrices should be interpreted with caution, as they are estimated with a high amount of uncertainty (Fig.~\ref{sfig:CV_ematrix}-\ref{sfig:CV_rmatrix}).
 
@@
 
-% II. IDRs have characteristic patterns of amino acid substitution
-%    A. In addition to these global patterns of conserved features, we find IDRs exhibit other characteristic patterns of evolution.
-%    B. For example, a comparison of the exchangeability matrices fit to the disorder and order regions show decreased exchangeability coefficients between
-%    A.
-Interestingly, the exchangeability matrices suggest that although there is net flux towards disorder by depleting structure-promoting residues, disordered residues are more biochemically distinct in disordered regions than in structured regions.
-This suggests that although disorder-promoting amino acids are often grouped into a single category and described as largely hydrophilic, they have distinct roles in disordered regions.
-There is evidence that highly similar residues such as glutamine and asparagine have distinct properties in disordered regions that yielded different biophysical characters of phase-separated condensates, so this aligns with other observations.
-Conversely, structure promoting residues are more interchangeable.
-One hypothesis is as these bulkier hydrophobic residues are not constrained by packing, they are free to act more generically as small hydrophobic patches that mediate interactions in SLiMs and other binding regions in IDRs.
-Another model for this behavior is the stickers and spacers model of phase separation.
-%    A.
-A caveat of this analysis is its dependence on sequence alignments which were created using scoring matrices that are in turn often derived from substitution models.
-Some approaches attempt to minimize the impact of this circular dependency through an EM-like procedure where substitution models are fit and the sequences are re-aligned with scoring matrices derived from those models in alternating rounds until convergence.
-Because our analyses fit the substitution models directly from alignments, the observed patterns we observe could be artifacts of using a scoring matrix largely fit to folded domains to align IDRs
-%    A.
-However, as the exchangeability and rate matrices broadly reproduced the trends observed in previous studies which fit amino acid substitution models to disorder regions, these observations are likely not artifactual but instead represent reproducible differences in the substitution rates between residues.
-%    B. Interpretations of exchangeability and rate matrices
-%        1. Disorder-enriched residues are less interchangeable and play specific roles in these regions
-%        2. Disorder-depeleted residues not constrained by packing and can act more generically as small hydrophobic patches
-%            i. Stickers and spacers from Holehouse
-%            ii. Staller papers and other high throughput mutational scanning of ADs
-%            iii. Halfmann glutamine vs asparagine paper
-%        3. Note about variances of parameters as shown in CV graphs
+\subsection{IDRs have distinct patterns of sequence and feature evolution}
+In addition to these global clusters of conserved features, we find IDRs are characterized by other distinct patterns of evolution.
+For example, a comparison of the exchangeability matrices fit to the disorder and order regions shows that, relative to the order regions, the disorder regions have decreased exchangeability coefficients between the disorder-enriched residues.
+Conversely, the disorder-depleted residues have increased exchangeability coefficients with each other.
+As a residue's enrichment in IDRs is generally interpreted as a measure of its ability to promote intrinsic disorder, these results indicate that within IDRs and relative to folded domains, disorder-promoting residues are subject to stricter constraints whereas structure-promoting residues are more biochemically interchangeable.
+A potential weakness of this analysis is its dependence on sequence alignments which were created using scoring matrices that are in turn derived from other substitution models.
+Previous studies have attempted to minimize the impact of this circular dependency through an EM-like procedure where substitution models are first fit to alignments, and the sequences in the alignments are then re-aligned with scoring matrices derived from the fit models in alternating rounds until convergence~\cite{Brown2009, Szalkowski2011}.
+Because our analyses fit the substitution models directly from alignments, the observed patterns are possibly an artifact of using a scoring matrix derived from folded domains to align IDRs.
+However, as our matrices reproduce the trends reported in these prior studies, they likely reflect true differences in the patterns of residue substitution within IDRs.
+
+This result is a partial reversal of the typical pattern observed in folded proteins where the generally larger and more hydrophobic structure-promoting residues are subject to strict geometric constraints imposed by the tightly packed hydrophobic core.
+In contrast, the smaller and more hydrophilic disorder-promoting residues are more variable, as they often occur in flexible, solvent-exposed regions.
+It is, however, consistent with other analyses of sequence-function relationships in IDRs.
+For example, several studies demonstrated that acidic activation domains of transcription factors, which are usually disordered, contain clusters of hydrophobic residues interspersed throughout their largely acidic chains~\cite{Ravarani2018, Staller2018, Erijman2020}.
+Additionally, these studies showed that many distinct sequences can yield similar levels of transcriptional activity.
+Together, these findings suggest a model of transcriptional activation where the repulsions between acidic residues maintain the bulky hydrophobic residues in accessible conformations that in turn allow the activation domains to bind their targets through non-specific hydrophobic interactions.
+As these interactions do not require highly complementary interfaces, the observed increase in exchangeability coefficients between hydrophobic residues is consistent with this model and may reflect the prevalence of such ``fuzzy complexes'' in IDR interactions.
+Furthermore, other studies have shown that different disorder-promoting residues have specific effects on the material properties of condensates formed by phase-separating IDRs.
+For example, in FUS family proteins glycine residues enhance fluidity whereas glutamine and serine residues promote hardening~\cite{Wang2018}.
+Even glutamine and asparagine residues, which differ by a single methylene group, can have disparate effects on the conformational preferences of IDRs.
+While glutamine-rich sequences are conformationally heterogeneous and form toxic aggregates, asparagine-rich sequences instead assemble into benign amyloids, as asparagine's shorter side chain promotes the formation of turns and $\beta$-sheets~\cite{Halfmann2011}.
+Thus, these observations along with the decreased exchangeability coefficients between disorder-promoting residues suggest that subtle differences in their biochemical properties may inform patterns of residue substitution in IDRs.
 

### 2023-05-10 chapter2/chapter2.tex
Added first paragraph of discussion

@@
 \section{Discussion}
-% I. Summary of results
-%    A. In this study, we applied several phylogenetic models to IDRs to interrogate the evolution of their sequences and molecular features.
-%    B. Most significantly, we demonstrate evidence of widespread evolutionary constraint in IDRs within the \textit{Drosophila} proteome
-%    C. Furthermore, many IDRs can be grouped into discrete clusters on the basis of these evolutionary signatures
-%    B. Most immediately, this work generates hypotheses of function related to sequence features that will permit the functional dissection of IDRs
-%    C. More broadly, however, it illustrates constraint of distributed features characterizes the evolution of some IDRs across multiple systems.
-%    D.
-However, these clusters require careful interpretation.
-The cluster descriptions summarize only a subset of the annotations with statistically significant enrichments.
-All clusters had many enriched annotations, but the summaries correspond to sets of enrichments with consisten .
-Because the significance tests were not corrected for multiple testing or controlled for their false discovery rate, many annotations are ,
+\subsection{IDRs have distinct patterns of sequence and feature evolution}
+In this study, we applied several phylogenetic models to IDRs to interrogate the evolution of their sequences and molecular features.
+Most significantly, through a comparison of two models of continuous trait evolution we demonstrate evidence of widespread constraint in IDRs within the \textit{Drosophila} proteome.
+Furthermore, using evolutionary signatures derived from these models, we identify several clusters of IDRs with shared patterns of constraint and associations with specific biological functions.
+Most immediately, this work generates hypotheses on the importance of various molecular features that will aid in the functional dissection of IDRs in specific proteins.
+More broadly, though, it suggests that constraint of distributed features is a mechanism of IDR evolution common to multiple biological systems.
+However, in the absence of more detailed analyses of individual clusters, it remains unclear if any are homologous or analogous to those identified in yeast.
+Additionally, the clusters' functional associations require careful interpretation.
+For example, the cluster descriptions summarize a subset of the enriched annotations which were selected for their related functions.
+Because the significance tests were not corrected for multiple testing or controlled for their false discovery rate, some annotations with significant enrichment, which may include the subset selected for the cluster descriptions, are likely spurious.
+Thus, while these initial results are promising, further investigation is necessary to corroborate them.
 
-\subsection{IDRs have distinct patterns of sequence and feature evolution}
-In addition to these global clusters of conserved features, we find IDRs are characterized by other distinct patterns of evolution.
+In addition to these global clusters of conserved features, we found IDRs exhibit other distinct patterns of evolution.
 For example, a comparison of the exchangeability matrices fit to the disorder and order regions shows that, relative to the order regions, the disorder regions have decreased exchangeability coefficients between the disorder-enriched residues.

### 2023-05-10 chapter2/chapter2.tex
Replaced pervasive with less negative ubiquitous

@@
 Accordingly, IDRs often act as the ``hubs'' of complex signaling networks by integrating signals from diverse pathways and coordinating interactions~\cite{Dunker2005, Buljan2012}.
-However, as IDRs are pervasive in eukaryotic proteomes, with estimates of the fractions of disordered residues in the human, mouse, and fruit fly proteomes ranging between 22 and 24\%~\cite{Piovesan2020, Piovesan2022}, they are involved in diverse processes~\cite{vanderLee2014} including transcriptional regulation~\cite{Liu2006} and the formation of biomolecular condensates~\cite{Banani2017}.
+However, as IDRs are ubiquitous in eukaryotic proteomes, with estimates of the fractions of disordered residues in the human, mouse, and fruit fly proteomes ranging between 22 and 24\%~\cite{Piovesan2020, Piovesan2022}, they are involved in diverse processes~\cite{vanderLee2014} including transcriptional regulation~\cite{Liu2006} and the formation of biomolecular condensates~\cite{Banani2017}.
 
@@
 However, as IDRs are likely subject to multiple selective pressures where the constraints on different compositional, patterned, or motif features are highly specific to each, we anticipate a range of computational and experimental methods will be needed to disentangle the complex forces driving IDR evolution.
-Accordingly, while these results represent a significant step forward in relating sequence to function in IDRs, further studies exploring these and other approaches will undoubtedly reveal new insights into these pervasive but poorly understood regions of proteins.
+Accordingly, while these results represent a significant step forward in relating sequence to function in IDRs, further studies exploring these and other approaches will undoubtedly reveal new insights into these ubiquitous but poorly understood regions of proteins.
 

### 2023-05-10 chapter2/chapter2.tex
Removed redundant use of residues with amino acids

@@
 However, several recent studies have demonstrated evidence that IDRs are constrained to preserve ``distributed features'' such as flexibility~\cite{Daughdrill2007}, chemical composition~\cite{Moesa2012}, net charge~\cite{Zarin2017}, or charge distribution~\cite{Beh2012}.
-Because many sequences can yield a region with a specific composition, for example, this mode of constraint uncouples an IDR's fitness from its strict sequence of amino acid residues.
+Because many sequences can yield a region with a specific composition, for example, this mode of constraint uncouples an IDR's fitness from its strict sequence of amino acids.
 Furthermore, in contrast to folded regions whose precise contacts and packing geometries are easily disrupted by amino acid substitutions, distributed features are robust to such changes, as individual residues only weakly contribute to a region's fitness.

### 2023-05-10 chapter1/chapter1.tex
Fixed comma separating consecutive figure panel labels
Made various wording clarifications and typo corrections in chapter 2

@@
 The best hits between sequences are naturally visualized as a graph where sequences are nodes and best hits are edges between nodes.
-Two connected components, sets of nodes joined by a sequence of edges, are shown (Fig.~\ref{fig:components}A,B).
+Two connected components, sets of nodes joined by a sequence of edges, are shown (Fig.~\ref{fig:components}A-B).
 The sequences in the first (Fig.~\ref{fig:components}A) all contain C2H2 zinc fingers, whereas the sequences in the second (Fig.~\ref{fig:components}B) are members of the Par-1 family of serine/threonine protein kinases.
@@
 Relevant output files are available in the supporting information.
-There is no primary data associated with this manuscript.
+There are no primary data associated with this manuscript.
 All primary data are available from publicly accessible sources described in their corresponding sections.

### 2023-05-11 main.tex
Adjusted spacing of title page
Fixed some typos in main abstract

@@
     By\\
+    \bigskip
     Marc Singleton\\
@@
 
-Thus, in the first chapter I develop a novel method for identifying orthologs which addresses this shortcoming and apply it to 33 \textit{Drosophila} genomes to generate a set of aligned orthologs.
-In the second chapter I then identify rapidly evolving IDRs in these alignments and analyse them with a variety of evolutionary models dissect the forces driving their evolution and detect patterns of conservation.
-Finally, in the third chapter I discuss several software tools and tutorials for fitting statistical models to data, which were created while pursuing the previous aims.
+Thus, in the first chapter, I develop a novel method for identifying orthologs which addresses this shortcoming and apply it to 33 \textit{Drosophila} genomes to generate a set of aligned orthologs.
+In the second chapter, I then identify rapidly evolving IDRs in these alignments and analyse them with a variety of evolutionary models to dissect the forces driving their evolution and detect patterns of conservation.
+Finally, in the third chapter, I discuss several software tools and tutorials for fitting statistical models to data, which were created while pursuing the previous aims.
 

### 2023-07-20 chapter2/figures/brownian/brownian.py
Removed chapters 0, 2, and 3 and main.tex from repo

@@ -1,462 +1,0 @@
-"""Make figure for Brownian motion analyses of features."""
-
-import os
-import re
-
-import matplotlib.pyplot as plt
-import numpy as np
-import pandas as pd
-import skbio
-from matplotlib.lines import Line2D
-from matplotlib.patches import ArrowStyle
-from matplotlib.transforms import blended_transform_factory
-from sklearn.decomposition import PCA
-from src2.brownian.pca import add_pca_arrows
-from src2.draw import plot_msa
-from src2.utils import read_fasta
-
-
-def zscore(df):
-    return (df - df.mean()) / df.std()
-
-
-def get_bin2data(hb, xs, ys):
-    offsets = hb.get_offsets()
-    xys = np.stack([xs, ys]).transpose()
-    dist2 = ((offsets[:, np.newaxis, :] - xys) ** 2).sum(axis=-1)
-    bin_idxs = np.argmin(dist2, axis=0)
-
-    bin2data = {}
-    for data_idx, bin_idx in enumerate(bin_idxs):
-        try:
-            bin2data[bin_idx].append(data_idx)
-        except KeyError:
-            bin2data[bin_idx] = [data_idx]
-
-    return bin2data
-
-
-def get_bin(hb, x, y):
-    bin2data = get_bin2data(hb, [x], [y])
-    return list(bin2data)[0]
-
-
-def set_bin_edge(ax, hb, bin_idx, facecolor, edgecolor, linewidth):
-    path = hb.get_paths()[0]
-    offset = hb.get_offsets()[bin_idx]
-
-    vertices = path.vertices + offset
-    hexagon = plt.Polygon(vertices, facecolor=facecolor, edgecolor=edgecolor, linewidth=linewidth)
-    ax.add_patch(hexagon)
-
-
-pdidx = pd.IndexSlice
-spid_regex = r'spid=([a-z]+)'
-min_length = 30
-
-min_indel_columns = 5  # Indel rates below this value are set to 0
-
-pca_components = 10
-cmap = plt.colormaps['Greys']
-hexbin_kwargs_log = {'gridsize': 30, 'mincnt': 1, 'linewidth': 0, 'bins': 'log'}
-handle_markerfacecolor = 0.6
-legend_kwargs = {'fontsize': 6.5, 'loc': 'center left', 'bbox_to_anchor': (1, 0.5)}
[…]

### 2023-07-20 chapter2/figures/hierarchy/hierarchy.py

@@ -1,218 +1,0 @@
-"""Make figure of cluster heatmap."""
-
-import os
-
-import matplotlib.pyplot as plt
-import numpy as np
-import pandas as pd
-from matplotlib.patches import Patch, Rectangle
-from scipy.cluster.hierarchy import linkage
-from scipy.spatial.distance import pdist
-from src2.brownian.linkage import make_tree
-from src2.draw import plot_tree
-
-pdidx = pd.IndexSlice
-min_length = 30
-
-min_indel_columns = 5  # Indel rates below this value are set to 0
-min_aa_rate = 1
-min_indel_rate = 0.1
-
-clusters = [('15086', 'A', 'extracellular structure'),
-            ('15170', 'B', 'Wnt signaling'),
-            ('15160', 'C', ''),
-            ('15070', 'D', 'nuclear pore organization'),
-            ('14415', 'E', 'voltage gated channel'),
-            ('15217', 'F', 'nuclear transport'),
-            ('15078', 'G', 'nucleotide transport'),
-            ('15021', 'H', 'cell cycle regulation'),
-            ('15161', 'I', 'synaptic signaling regulation'),
-            ('15191', 'J', 'histone methylation'),
-            ('15143', 'K', ''),
-            ('14937', 'L', 'chromatin assembly')]
-
-color1 = '#4e79a7'
-
-# Load regions
-rows = []
-with open(f'../../IDR_evolution/analysis/IDRpred/region_filter/out/regions_{min_length}.tsv') as file:
-    field_names = file.readline().rstrip('\n').split('\t')
-    for line in file:
-        fields = {key: value for key, value in zip(field_names, line.rstrip('\n').split('\t'))}
-        OGid, start, stop, disorder = fields['OGid'], int(fields['start']), int(fields['stop']), fields['disorder'] == 'True'
-        rows.append({'OGid': OGid, 'start': start, 'stop': stop, 'disorder': disorder})
-all_regions = pd.DataFrame(rows)
-
-asr_rates = pd.read_table(f'../../IDR_evolution/analysis/evofit/asr_stats/out/regions_{min_length}/rates.tsv')
-asr_rates = all_regions.merge(asr_rates, how='right', on=['OGid', 'start', 'stop'])
[…]

### 2023-07-20 chapter2/figures/regions/regions.py

@@ -1,104 +1,0 @@
-"""Make figures for basic region statistics."""
-
-import os
-import re
-
-import matplotlib.pyplot as plt
-import pandas as pd
-import scipy.stats as stats
-from numpy import linspace, sqrt
-from src2.utils import read_fasta
-
-pdidx = pd.IndexSlice
-ppid_regex = r'ppid=([A-Za-z0-9_.]+)'
-min_length = 30
-
-min_indel_columns = 5  # Indel rates below this value are set to 0
-min_aa_rate = 1
-min_indel_rate = 0.1
-
-# Load regions as segments
-rows = []
-with open(f'../../IDR_evolution/analysis/IDRpred/region_filter/out/regions_{min_length}.tsv') as file:
-    field_names = file.readline().rstrip('\n').split('\t')
-    for line in file:
-        fields = {key: value for key, value in zip(field_names, line.rstrip('\n').split('\t'))}
-        OGid, start, stop, disorder = fields['OGid'], int(fields['start']), int(fields['stop']), fields['disorder'] == 'True'
-        for ppid in fields['ppids'].split(','):
-            rows.append({'OGid': OGid, 'start': start, 'stop': stop, 'disorder': disorder,
-                         'ppid': ppid, 'min_length': min_length})
-all_segments = pd.DataFrame(rows)
-all_regions = all_segments[['OGid', 'start', 'stop', 'disorder']].drop_duplicates()
-
-# Get segment lengths
-rows = []
-for OGid, group in all_segments[['OGid', 'start', 'stop']].drop_duplicates().groupby('OGid'):
-    msa = read_fasta(f'../../IDR_evolution/data/alignments/fastas/{OGid}.afa')
-    msa = [(re.search(ppid_regex, header).group(1), seq) for header, seq in msa]
-
-    for row in group.itertuples():
-        region_length = row.stop - row.start
-        for ppid, seq in msa:
-            segment = seq[row.start:row.stop]
-            segment_length = region_length - segment.count('.') - segment.count('-')
-            rows.append({'OGid': OGid, 'start': row.start, 'stop': row.stop, 'ppid': ppid, 'length': segment_length})
-all_lengths = pd.DataFrame(rows)
-
[…]

### 2023-07-20 chapter2/figures/scores/scores.py

@@ -1,298 +1,0 @@
-"""Make figure of disorder score evolution."""
-
-import os
-import re
-from textwrap import shorten
-
-import matplotlib.pyplot as plt
-import numpy as np
-import pandas as pd
-import skbio
-from matplotlib.patches import Patch
-from src2.draw import plot_msa_data
-from src2.phylo import get_brownian_weights
-from src2.utils import read_fasta
-
-
-def load_scores(path):
-    with open(path) as file:
-        scores = []
-        for line in file:
-            if not line.startswith('#'):
-                score = line.split()[3]
-                scores.append(score)
-    return scores
-
-
-ppid_regex = r'ppid=([A-Za-z0-9_.]+)'
-spid_regex = r'spid=([a-z]+)'
-min_length = 30
-
-panel_records = [('0F47', 0, None),
-                 ('07E3', 350, 650),
-                 ('146F', 125, 525),
-                 ('1432', 9, 325)]
-plt.rcParams['xtick.labelsize'] = 8
-plt.rcParams['ytick.labelsize'] = 8
-plt.rcParams['axes.labelsize'] = 8
-plt.rcParams['legend.fontsize'] = 8
-fig_width = 7.5
-fig_height = 8.5
-width_ratios = (0.3, 0.3, 0.4)
-rect_B = (0.375, 0.06, 0.5475, 0.92)
-rect_CD = (0.25, 0.18, 0.7, 0.7)
-rect_E = (0.5, 0.325, 0.475, 0.65)
-plot_msa_kwargs_template = {'figsize': (fig_width * (width_ratios[0] + width_ratios[1]), fig_height / 2),
-                            'left': 0.15, 'right': 0.85, 'top': 0.95, 'bottom': 0.025, 'anchor': (0, 0.5),
-                            'hspace': 0.01,
-                            'data_min': -0.05, 'data_max': 1.05,
-                            'data_linewidths': 1,
-                            'tree_position': 0, 'tree_width': 0.15,
-                            'tree_kwargs': {'linewidth': 0.75, 'tip_labels': False, 'xmin_pad': 0.025, 'xmax_pad': 0.025},
-                            'msa_legend': True, 'legend_kwargs': {'bbox_to_anchor': (0.85, 0.5), 'loc': 'center left', 'fontsize': 8,
-                                                                  'handletextpad': 0.5, 'markerscale': 1.5, 'handlelength': 1}}
-color3 = '#b07aa1'
-
[…]

### 2023-07-20 chapter2/figures/substitution/substitution.py

@@ -1,221 +1,0 @@
-"""Make figure of results from fitting substitution models."""
-
-import os
-import re
-from collections import namedtuple
-
-import matplotlib.pyplot as plt
-import numpy as np
-import skbio
-from matplotlib.colors import Normalize
-from matplotlib.cm import ScalarMappable
-from src2.utils import read_iqtree, read_paml
-
-dpi = 300
-
-paml_order = ['A', 'R', 'N', 'D', 'C', 'Q', 'E', 'G', 'H', 'I', 'L', 'K', 'M', 'F', 'P', 'S', 'T', 'W', 'Y', 'V']
-labels = ['0R_disorder', '50R_disorder', '100R_disorder', '0R_order', '50R_order', '100R_order']
-labels_suffix = r'_[0-9]+'
-Record = namedtuple('Record', ['label', 'ematrix', 'rmatrix', 'freqs', 'length'])
-
-# Load LG model
-ematrix, freqs = read_paml('../../IDR_evolution/data/matrices/LG.paml', norm=True)
-rmatrix = freqs * ematrix
-
-# Load IQ-TREE matrices
-records = {}
-for label in labels:
-    file_labels = []
-    for path in os.listdir('../../IDR_evolution/analysis/evofit/iqtree_fit/out/'):
-        match = re.match(f'({label}{labels_suffix})\.iqtree', path)
-        if match:
-            file_labels.append(match.group(1))
-
-    ematrix_stack = []
-    rmatrix_stack = []
-    freqs_stack = []
-    length_stack = []
-    for file_label in sorted(file_labels):
-        record = read_iqtree(f'../../IDR_evolution/analysis/evofit/iqtree_fit/out/{file_label}.iqtree', norm=True)
-        ematrix, freqs = record['ematrix'], record['freqs']
-        rmatrix = freqs * ematrix
-
-        tree = skbio.read(f'../../IDR_evolution/analysis/evofit/iqtree_fit/out/{file_label}.treefile', 'newick', skbio.TreeNode)
-        length = tree.descending_branch_length()
-
-        ematrix_stack.append(ematrix)
-        rmatrix_stack.append(rmatrix)
-        freqs_stack.append(freqs)
-        length_stack.append(length)
-    records[label] = Record(label,
[…]

### 2023-07-20 chapter3/figures/jupyter.py

@@ -1,672 +1,0 @@
-"""Make all figures from training tutorial."""
-
-# In[1]:
-import os
-import pprint
-import random
-from functools import reduce
-from itertools import accumulate
-
-import matplotlib.pyplot as plt
-import numpy as np
-import scipy.stats as stats
-from homomorph import HMM
-from numpy import exp, log
-from sklearn.metrics import roc_curve
-from src3.utils import fit_CML
-
-dpi = 300
-legend_kwargs = {'frameon': False, 'loc': 'center left', 'bbox_to_anchor': (1, 0.5)}
-
-if not os.path.exists('out/'):
-    os.mkdir('out/')
-
-
-# In[2]:
-t_dists = {1: {1: 0.95, 2: 0.05},
-           2: {1: 0.05, 2: 0.9, 3: 0.05},
-           3: {2: 0.35, 3: 0.65}}
-e_dists = {1: {'A': 1},
-           2: {'A': 0.5, 'B': 0.5},
-           3: {'B': 1}}
-start_dist = {1: 0.2, 2: 0.5, 3: 0.3}
-
-model = HMM(t_dists=t_dists, e_dists=e_dists, start_dist=start_dist)
-
-
-# In[3]:
-data = [model.simulate(200, random_state=i) for i in range(10)]  # Use a different random seed for each example
-data = [list(zip(*example)) for example in data]
-
-
-# In[4]:
-# Make transition count dicts and add pseudocounts
-t_pseudo = 0.1
-t_counts = {}
-for state1, t_dist in t_dists.items():
-    t_count = {}
-    for state2 in t_dist:
-        t_count[state2] = t_pseudo
-    t_counts[state1] = t_count
-
-# Add observed counts
-for example in data:
-    xs, ys = example
-    state0 = xs[0]
-    for state1 in xs[1:]:
-        t_counts[state0][state1] += 1
-        state0 = state1
-
-# Normalize counts
-t_dists_hat = {}
-for state1, t_count in t_counts.items():
-    t_sum = sum(t_count.values())
-    t_dist_hat = {}
-    for state2, count in t_count.items():
-        t_dist_hat[state2] = count / t_sum
-    t_dists_hat[state1] = t_dist_hat
-
-
-# In[5]:
-# Collect all possible emissions
-e_set = set()
-for example in data:
-    xs, ys = example
-    e_set.update(ys)
-
-# Make emission count dicts and add pseudocounts
-e_pseudo = 0.1
-e_counts = {}
-for state in t_dists:
[…]

### 2023-07-25 manuscript.tex
Updated manuscript to reflect merging of states 1A and 1B and new training statistics

--- 2023-05-10 chapter1/chapter1.tex
@@
 A fully specified phylo-HMM requires a fixed number of hidden states and a probability distribution for each.
-Thus, we identified four distinct types of regions in the alignments, roughly corresponding to highly conserved regions with few to no gaps, diverged regions, regions with a stable gap pattern discordant with the expected phylogenetic relationships, and regions with poorly supported segments.
-For simplicity, however, we refer to the states that generate each type of region as 1A, 1B, 2, and 3, respectively.
+Thus, we identified three distinct types of regions in the alignments, roughly corresponding to regions with strong phylogenetic evidence, regions with a stable gap pattern discordant with the expected phylogenetic relationships, and regions with poorly supported segments.
+For simplicity, however, we refer to the states that generate each type of region as 1, 2, and 3, respectively.
 To model probability distributions for each state, we first conceptualized the observed alignments as the superposition of two distinct processes (Fig.~\ref{fig:hmm}A, left).
@@
 After the model was trained on manually labeled examples, it was used to assign a label to the columns in each alignment.
-Columns assigned to states 1A and 1B are the regions of interest for downstream analyses since the gaps generally follow the expected pattern given the phylogenetic tree.
+Columns assigned to state 1 are the regions of interest for downstream analyses since the gaps generally follow the expected pattern given the phylogenetic tree.
 In contrast, columns assigned to states 2 and 3 largely corresponded to the long poorly supported segments and phylogenetically discordant regions discussed previously and were therefore removed from the alignments.
@@
 The likelihoods of phylogenetic trees were efficiently calculated with Felsenstein's pruning algorithm, and all HMM algorithms were implemented with custom code which is available as the package Homomorph on PyPI~\cite{Felsenstein1981}.
-The insertion phylo-HMM was trained on 47,387 manually labeled columns in 14 alignments, and the missing data phylo-HMM was trained on 67,001 manually labeled positions in 23 sequences in 11 unique alignments (Fig.~\ref{sfig:insertion_training}, Fig.~\ref{sfig:missing_training}).
+The insertion phylo-HMM was trained on 53,518 manually labeled columns in 15 alignments, and the missing data phylo-HMM was trained on 67,360 manually labeled positions in 23 sequences in 11 unique alignments (Fig.~\ref{sfig:insertion_training}, Fig.~\ref{sfig:missing_training}).
 Because maximum-likelihood estimation of the model parameters yielded posterior decoding curves which toggled between hidden states too rapidly, the models were instead trained discriminatively~\cite{Krogh1999}.
@@
 First, the probability of state 3 for columns with a gap profile value less than or equal to 0.1 was set to 0 to break long poorly supported segments aligned to highly conserved columns into separate regions.
-Regions were defined with the previously described algorithm using high and low cutoffs of 0.75 and 0.01, respectively.
+Regions were defined with the previously described algorithm using high and low cutoffs of 0.75 and $1 \times 10^{-3}$, respectively.
 The outer and inner derivative cutoffs were both 0.001.

### 2023-07-26 manuscript.tex
Updated data availability statement

@@
 The following Python libraries were used: matplotlib~\cite{Hunter2007}, NumPy~\cite{Harris2020}, pandas~\cite{McKinney2010}, and SciPy~\cite{Virtanen2020}.
-Relevant output files are available in the supporting information.
 There are no primary data associated with this manuscript.
 All primary data are available from publicly accessible sources described in their corresponding sections.
+Relevant outputs are available in the supplemental files at FigShare.
+File S1 contains the orthologous groups generated by 4-clique percolation with paralogs added.
+File S2 contains the initial alignments of all sequences in each orthologous group.
+File S3 contains the refined alignments of the representative sequences in each single copy orthologous group.
+File S4 contains the insertion phylo-HMM model parameters and the alignments of the single copy ortholog groups after trimming.
+File S5 contains the missing data phylo-HMM model parameters and the missing segments identified for each sequence in each single copy orthologous group.
+Missing segments are given as Python slices in the following form: start0-stop0,start1-stop1,\ldots,startn-stopn.
+If the slices field is empty, then no missing segments were identified in that sequence.
+File S6 contains the consensus phylogenetic trees fit using the LG, GTR, and GTR2 substitution models.
 \end{sloppypar}

### 2023-07-26 main.tex
Updated affiliations

--- 2023-04-27 main.tex
@@
-\documentclass[11pt,letterpaper,oneside]{book}
+\documentclass[11pt,letterpaper]{article}
 
-% PACKAGE AND ENVIRONMENT CONFIGURATIONS
+% PACKAGE CONFIGURATIONS
 \usepackage{amsmath}
-\DeclareMathOperator*{\argmax}{arg\!\max}
-\DeclareMathOperator*{\argmin}{arg\!\min}
-
 \usepackage{graphicx}
-
-% Code listings
-\usepackage[outputdir=out]{minted}
-\usepackage{tcolorbox}
-\tcbuselibrary{breakable}
-\tcbset{colframe=white}
-\usepackage{lmodern}
 
@@
 \usepackage[margin=1in]{geometry}
-\usepackage{lscape}
 %\usepackage{setspace}  % double spacing
@@
 \usepackage[skip=10pt]{parskip}  % Remove indents
-\pagestyle{plain}
+\setcounter{secnumdepth}{0}  % Remove section numbering
 
-% Chapter and figure formatting
-\usepackage{longtable}
-\usepackage{titlesec}  % Modify chapter title format
-\titleformat{\chapter}[display]% Shape (chapter label in separate paragraph)
-                      {\bfseries\Large}% Format
-                      {\MakeUppercase{\chaptertitlename} \thechapter}% Label
-                      {0in}% Sep
-                      {}% Before code
-                      []% After code
-\setcounter{tocdepth}{1}  % Only number divisions to section
-\usepackage[font=small, labelfont=bf, labelsep=period]{caption}
-\renewcommand{\thefigure}{\thechapter.\arabic{figure}}
+% Figure formatting
+\usepackage[labelfont=bf, labelsep=period]{caption}
+\renewcommand{\figurename}{Fig}
 
@@
 
-% ENVIRONMENTS
-\newenvironment{abstract}
-{% Begin code
- \small
- \begin{center}
- \textbf{Abstract}
- \end{center}
- \quotation
-}
-{% End code
-}
-
-\definecolor{incolor}{RGB}{51,63,153}
-\definecolor{outcolor}{RGB}{200,78,41}
-\definecolor{bgcolor}{RGB}{247,247,247}
-
-\newlength{\promptwidth}
-\setlength{\promptwidth}{0.375in}
-
-\newcommand{\prompt}[2]
-{\makebox[0pt][r]{\texttt{\color{#2}#1:\ }}\vspace{-\baselineskip}}
-
-\newenvironment
-{NotebookIn}
-{%
- \VerbatimEnvironment%
- \begin{tcolorbox}[sharp corners,
-                   frame empty,
-                   breakable,
-                   size=fbox,
-                   colback=bgcolor,
-                   left skip=\promptwidth,
-                   title=\prompt{In}{incolor},
-                   attach title to upper]%
- \begin{minted}{python}%
-}
-{%
- \end{minted}%
- \end{tcolorbox}%
-}
-
-\newenvironment
-{NotebookOut}
-{%
- \VerbatimEnvironment%
- \begin{tcolorbox}[sharp corners,
-                   frame empty,
-                   breakable,
-                   size=fbox,
-                   colback=white,
-                   left skip=\promptwidth,
-                   title=\prompt{Out}{outcolor},
-                   attach title to upper]%
- \begin{minted}{text}%
-}
-{%
- \end{minted}%
- \end{tcolorbox}%
-}
-
-\newenvironment
-{NotebookImage}
-{%
- \begin{tcolorbox}[sharp corners,
-                   frame empty,
-                   size=fbox,
-                   colback=white,
-                   left skip=\promptwidth,
-                   title=\prompt{Out}{outcolor},
-                   attach title to upper]%
- \\%
-}
-{%
- \end{tcolorbox}%
-}
-
 % DOCUMENT
@@
 
-\begin{titlepage}
-\begin{center}
-    Finding structure in disorder: Evolutionary analyses of disordered proteins in \textit{Drosophila}\\
+% Title must be 250 characters or less
+\begin{flushleft}
+{\Large\textbf{Leveraging genomic redundancy to improve inference and alignment of orthologous proteins}}
+\\
 \bigskip
-    By\\
-    Marc Singleton\\
-    \vfill
-    A dissertation submitted in partial satisfaction of the\\
-    requirements for the degree of\\
-    Doctor of Philosophy\\
-    in\\
-    Biophysics\\
-    in the\\
-    Graduate Division\\
-    of the\\
-    University of California, Berkeley\\
-    \vfill
-    Committee in charge:\\
-    Professor Mike Eisen, Chair\\
-    Professor Liana Lareau\\
-    Professor Susan Marqusee\\
-    Professor Priya Moorjani\\
+Marc Singleton\textsuperscript{1,*},
+Michael Eisen\textsuperscript{1,2}
+\\
 \bigskip
-    Spring 2023\\
-\end{center}
-\end{titlepage}
+\textbf{1} Howard Hughes Medical Institute, UC Berkeley, Berkeley, CA, USA
+\\
+\textbf{2} Department of Molecular and Cell Biology, UC Berkeley, Berkeley, CA, USA
+\\
+\bigskip
 
-\thispagestyle{empty}
-\null  % Necessary to have text or command before \vfill for proper spacing
-\vfill
-\begin{center}
-    \copyright \  2023 Marc Singleton
-\end{center}
+* marcsingleton@berkeley.edu
+
+\end{flushleft}
+
+\graphicspath{{figures/}}
+\input{manuscript.tex}
+
+\section{Acknowledgments}
+We thank the past and present members of the Eisen lab for helpful discussions and their assistance in revising the manuscript.
+
+\printbibliography
+
 \clearpage
 
-\setcounter{page}{1}  % Reset page counter after copyright
-\begin{center}
-    Abstract\\
-    \bigskip
-    Finding structure in disorder: Evolutionary analyses of disordered proteins in \textit{Drosophila}\\
-    by\\
-    Marc Singleton\\
-    \bigskip
-    Doctor of Philosophy in Biophysics\\
-    University of California, Berkeley\\
-    Professor Mike Eisen, Chair\\
-    \bigskip
-\end{center}
-Here's the text for my abstract!
-\clearpage
+\setcounter{figure}{0}
+\renewcommand{\thefigure}{S\arabic{figure}}
+\setcounter{table}{0}
+\renewcommand{\thetable}{S\arabic{table}}
 
-\setcounter{page}{1}  % Reset page counter after abstract
-\pagenumbering{roman}  % Set numbering style to roman for pre-main content
-
-\null  % Necessary to have text or command before \vfill for proper spacing
-\vfill
-\begin{center}
-    To my parents, who were always willing to indulge\\
-    another ``experiment'' in their kitchen sink
-\end{center}
-\vfill
-\clearpage
-
-\begin{center}
-    \bfseries\Large Acknowledgements
-\end{center}
-\bigskip
-If it takes a village to raise a child, then it must take at least three to raise a scientist.
-Over the years, I've had countless friends, teachers, and mentors.
-These relationships have shaped the person I am today, so I'm grateful to all the people who took time out of their busy lives to help me in ways both small and large.
-However, I want to single out the following people who had an especially important impact on my time in graduate school.\\\\
-Thank you to my advisor, Mike, for supporting me as a scientist and a person in equal measure.\\\\
-Thank you to my committee, Susan, Liana, and Priya, for giving me the pep talk I needed every year.\\\\
-Thank you to my lab, Holli, Stadler, Ciera, Ashley, Victoria, Augusto, and Xiao-Yong, for creating a space where it's okay to ask questions.
-A special thank you to my bay mates, Jenna and Colleen, for getting me through the hard days with sound advice and a healthy dose of humor.\\\\
-Thank you to my undergraduate research mentors, Cathy and Jessica, for teaching me the most exciting science is at the intersection of different fields.\\\\
-Thank you to my parents, Amy and David, for everything.\\\\
-Thank you to my brothers, Andrew and Daniel, for giving me thick skin in the way only big brothers can.\\\\
-Thank you to all the friends I've made in California and the Bay Area, for the much-needed life outside of work.\\\\
-Finally, thank you to my partner, Joseph, for going on this crazy ride with me. I couldn't have done this without you.
-\clearpage
-
-\tableofcontents
-\clearpage  % End page so numbering is set properly
-\setcounter{chapter}{-1}  % Set chapter counter to 0
-\setcounter{page}{1}  % Reset page counter after pre-main content
-\pagenumbering{arabic}  % Set numbering style to arabic for main content
-
-\chapter{Introduction}
-\label{chapter:0}
-\graphicspath{{chapter0/figures/}}
-\input{chapter0/chapter0.tex}
-
-\chapter{Leveraging genomic redundancy to improve inference and alignment of orthologous proteins}
-\label{chapter:1}
-\graphicspath{{chapter1/figures/}}
-\input{chapter1/chapter1.tex}
-
-\chapter{Evolutionary analyses of IDRs reveal patterns of conserved features}
-\label{chapter:2}
-\graphicspath{{chapter2/figures/}}
-\input{chapter2/chapter2.tex}
-
-\chapter{Tools and tutorials for fitting mixture models and HMMs}
-\label{chapter:3}
-\graphicspath{{chapter3/}}
-\input{chapter3/introduction.tex}
-\input{chapter3/training_tutorial}
-
-\printbibliography[heading=bibintoc, title=References]  % bibintoc == put in table of contents
-
-\appendix
-\setcounter{figure}{0}
-\renewcommand{\thefigure}{\thechapter\arabic{figure}}
-\setcounter{table}{0}
-\renewcommand{\thetable}{\thechapter\arabic{table}}
-
-\chapter{Supporting information for chapter 1}
-\label{appendix:a}
-\graphicspath{{chapter1/figures/}}
-\input{chapter1/chapter1_supporting.tex}
-
-\chapter{Supporting information for chapter 2}
-\label{appendix:b}
-\graphicspath{{chapter2/figures/}}
-\input{chapter2/chapter2_supporting.tex}
-
-\chapter{Supporting information for chapter 3}
-\label{appendix:c}
-\setlength{\promptwidth}{0in}  % No indent for appendix code
-\input{chapter3/training_tutorial_supporting.tex}
+\section*{Supporting information}
+\input{supporting.tex}
 

### 2023-08-29 main.tex
Moved title page information to manuscript and added required sections for submission

@@
 
-% Title must be 250 characters or less
-\begin{flushleft}
-{\Large\textbf{Leveraging genomic redundancy to improve inference and alignment of orthologous proteins}}
-\\
-\bigskip
-Marc Singleton\textsuperscript{1,*},
-Michael Eisen\textsuperscript{1,2}
-\\
-\bigskip
-\textbf{1} Howard Hughes Medical Institute, UC Berkeley, Berkeley, CA, USA
-\\
-\textbf{2} Department of Molecular and Cell Biology, UC Berkeley, Berkeley, CA, USA
-\\
-\bigskip
-
-* marcsingleton@berkeley.edu
-
-\end{flushleft}
-
 \graphicspath{{figures/}}
 \input{manuscript.tex}
-
-\section{Acknowledgments}
-We thank the past and present members of the Eisen lab for helpful discussions and their assistance in revising the manuscript.
 

### 2023-08-29 manuscript.tex
Moved title page information to manuscript and added required sections for submission

@@
-\begin{abstract}
-\noindent
+\textbf{Title:} Leveraging genomic redundancy to improve inference and alignment of orthologous proteins
+
+\textbf{Authors:} Marc Singleton\textsuperscript{1,*}, Michael Eisen\textsuperscript{1,2}
+
+\textbf{Author addresses:}\\
+1 Howard Hughes Medical Institute, UC Berkeley, Berkeley, CA 94720, USA\\
+2 Department of Molecular and Cell Biology, UC Berkeley, Berkeley, CA 94720, USA
+
+\textbf{Corresponding author email address:}\\
+* marcsingleton@berkeley.edu
+
+\textbf{Running head:} Leveraging genomic redundancy
+
+\textbf{Keywords:} orthology, multiple sequence alignment, HMM, Drosophila
+
+\clearpage
+
+\section{Abstract}
 Identifying protein sequences with common ancestry is a core task in bioinformatics and evolutionary biology.
@@
 We applied the resulting pipeline to a set of 33 annotated \textit{Drosophila} genomes, generating 22,813 orthologous groups and 8,566 high-quality alignments.
-\end{abstract}
+
+\section{Article summary}
+insert article summary
 
@@
 Given the heterogeneous sourcing of the transcript evidence, these sequences containing these segments are likely splice isoforms specific to certain tissues or developmental stage.
-
-% To investigate these hypotheses, the genome sequences and supporting transcript evidence for several segments and regions were examined in detail: 0254, 26CF, 202C, 188E, 2E9F, 34B7, 0C31.
 
@@
 
-\subsection{Code and data availability}
+\section{Data availability}
 \begin{sloppypar}
@@
 \end{sloppypar}
+
+\section{Acknowledgements}
+insert acknowledgements
+
+\section{Funding}
+insert funding
+
+\section{Conflicts of interest}
+The authors declare no conflicts of interest.
+
+\section{Figures}
+insert figures
+
+\section{Tables}
+insert tables

### 2023-08-29 manuscript.tex
Moved Materials and methods after introduction
Moved figures to figure section
Added acknowledgements and funding text and removed article summary
Added clearpage commands to space references, figures, and tables properly
Removed tables section because tables will remain in supplement
Removed footnote in introduction

@@
 
-\section{Article summary}
-insert article summary
-
 \section{Introduction}
@@
 However, many analyses impose the additional condition that the sequences have diverged through speciation events (orthology) rather than duplications (paralogy) or other mechanisms such as horizontal gene transfer~\cite{Fitch1970}.
-The underlying assumption is orthologs have conserved equivalent functions whereas paralogs, by virtue of their redundancy, are more likely to diverge~\cite{Ohno1970, Nowak1997, Altenhoff2012, Pegueroles2013, Soria2014}.\footnote{For multiple sequences, orthology is usually defined relative to their most recent common ancestor.
-This technically includes sequences which split by duplication after this point (in-paralogs), but excludes sequences which split by duplication before (out-paralogs) \cite{Remm2001}.
-Many current orthology inference pipelines explicitly incorporate steps to detect in-paralogs.
-However, the resulting orthologs groups can easily be restricted to those without in-paralogs (single copy orthologs) for analyses where an assumption of conserved function is necessary.} This relationship between orthology and function is an essential component of modern biological research since it permits the transfer of annotations between biological systems using sequence similarity alone.
+The underlying assumption is orthologs have conserved equivalent functions whereas paralogs, by virtue of their redundancy, are more likely to diverge~\cite{Ohno1970, Nowak1997, Altenhoff2012, Pegueroles2013, Soria2014}.
+This relationship between orthology and function is an essential component of modern biological research since it permits the transfer of annotations between biological systems using sequence similarity alone.
 
@@
 Given the \textit{Drosophila} genus spans diverse habitats and over 50 million years of evolution but maintains a conserved life cycle and body plan, such a resource would facilitate a new generation of studies that illuminate the forces that drive protein evolution in unprecedented detail~\cite{Wiegmann2011, Obbard2012}.
-
-\begin{figure}[h!]
-\includegraphics[width=\textwidth]{graphical_overview/out/graphical_overview.png}
-\centering
-\caption{\textbf{Overview of orthology inference pipeline.}
-\textbf{(A-C)} The large, open circles represent the annotated genomes, and the small, filled circles represent the protein sequences associated with each annotation.
-Sequences that share homology are colored with shades of the same hue.
-\textbf{(D-F)} The small, filled circles and lines represent the same sequences and best hits from the previous steps.}
-\label{fig:graphical_overview}
-\end{figure}
-
-\section{Results}
-\subsection{Pipeline overview}
-Our pipeline follows a similar overall approach to other graph-based methods of orthology inference.
-First, protein sequences from annotated genomes are collected (Fig.~\ref{fig:graphical_overview}A), and homology searches are conducted between all query-target pairs of genomes (Fig.~\ref{fig:graphical_overview}B).
-The raw output from these homology searches is processed to yield best hits between pairs of sequences (Fig.~\ref{fig:graphical_overview}C).
-Next, the network of best hits is clustered into self-consistent orthologous groups (Fig.~\ref{fig:graphical_overview}D).
-Since genes can have multiplied associated isoforms, we then implemented a novel second clustering step where orthologous groups are grouped by their parent genes, which are represented by the two sets of clusters with warm and cool colors, respectively (Fig.~\ref{fig:graphical_overview}E).
-Finally, representative sequences in each orthologous group are aligned (Fig.~\ref{fig:graphical_overview}F).
-In the following sections, we discuss each of these and other steps which were omitted for clarity in greater detail.
-
-\subsection{Input genomes and pre-processing}
-All annotated genomes in the genus \textit{Drosophila} available in April 2022 were downloaded from NCBI's RefSeq database.
-The assemblies annotated by the NCBI eukaryotic genome annotation pipeline have passed several quality checks and all have supporting transcript evidence, so the annotations are generally highly complete~\cite{Thibaud-Nissen2013}.
-The \textit{D.
-miranda} annotation was excluded due to its unusual karyotype~\cite{Dobzhansky1935}.
-Other annotations were excluded after preliminary clustering showed a deficiency in the number of orthologous groups containing those genomes, indicating their annotations were less complete (data not shown).
-The \textit{D.
-melanogaster} annotation was downloaded from FlyBase~\cite{Gramates2022}.
-In total, the input data consists of 33 genomes, which are listed in Table~\ref{stable:genomes}.
-Many genes have transcripts that differ only in their UTRs, and as a result there are many duplicate protein sequences in the annotations.
-Though not strictly necessary, we removed the duplicates in our pipeline, which greatly reduced the computational burden of later steps.
-
-\subsection{Extraction of best hits from BLAST output}
-The protein sequences in each genome annotation were searched against each other in reciprocal pairs using BLAST, yielding a list of high-scoring segment pairs (HSPs) for each query-target pair~\cite{Camacho2009}.
-HSPs are local alignments, meaning they do not necessarily span the entire lengths of the query and target sequences.
-Consequently, the search algorithm may return multiple HSPs for each query-target pair if statistically significant regions of homology are separated by nonhomologous or poorly conserved regions.
-Though the most significant HSP is often used to represent all HSPs between a query-target pair, this approach can fail to rank the pairs by their overall significance if their alignments are broken into multiple HSPs.
-Furthermore, since query-target pairs were later filtered by the amount overlap between their sequences, it can also exclude pairs that pass the overlap threshold even if the most significant HSP alone does not.
-Thus, HSPs were merged into a single object called a hit.
-The best hits for each query were then taken as the highest-scoring hits that passed a minimum overlap criterion and were reciprocal between the query and target sequences.
-
-\subsection{Clustering in orthologous groups}
-The best hits between sequences are naturally visualized as a graph where sequences are nodes and best hits are edges between nodes.
-Two connected components, sets of nodes joined by a sequence of edges, are shown (Fig.~\ref{fig:components}A-B).
-The sequences in the first (Fig.~\ref{fig:components}A) all contain C2H2 zinc fingers, whereas the sequences in the second (Fig.~\ref{fig:components}B) are members of the Par-1 family of serine/threonine protein kinases.
-In both components, some sets of nodes have a high density of edges, forming distinct clusters, whereas other nodes are only sparsely connected to their neighbors.
-To better understand the structure of these two components, we calculated the number of sequences, unique genes, and unique species in each.
-The first has 385, 346, and 33 sequences, genes, and species, respectively, and the second has 222, 33, and 33 sequences, genes, and species, respectively.
-We then plotted the relationship between the number sequences and unique genes across all components to see if this pattern holds true generally (Fig.~\ref{fig:components}C).
-Two distinct trendlines are apparent.
-The first increases linearly with the number of sequences with a slope of one, indicating each sequence is generally associated with a unique gene.
-The second is constant with an intercept of 33, indicating the number of unique genes quickly saturates at the total number of genomes.
-Thus, there are generally two classes of components.
-The first is composed of many distinct genes, whereas the second is composed of many different isoforms of a single group of genes.
-
-The diffuse networks observed in the first component class are likely the result of a combination of factors, including rapid evolution, gene duplication, and annotation errors.
-Regardless of their origin, these hits are not strong candidates for comparative analyses since an orthology relationship is supported by relatively few genome pairs.
-Instead, likely orthologs should consistently identify each other as best reciprocal hits across many genome pairs.
-The same is true of the hits in the second component class.
-Although the genes as a unit form a single orthologous group, sequences with few hits are likely non-conserved or tissue-specific isoforms.
-Thus, orthologous groups can be operationally defined as self-consistent clusters in the hit graph.
-However, sequence divergence or assembly and annotation errors may prevent a best reciprocal hit between orthologous sequences across all genome pairs.
-In fact, although the most common number of reciprocal hits is 32, one fewer than the total number of genomes, many sequences have fewer (Fig.~\ref{fig:components}D).
-Thus, the clustering method should require a high degree of self-consistency without demanding complete consensus.
-
-\begin{figure}[h!]
-\includegraphics[width=\textwidth]{components/out/0017-005F.png}
-\centering
-\caption{\textbf{Selected connected components of hit graph and summary statistics.}
-\textbf{(A-B)} Two distinct connected components of the hit graph.
-Edges are colored by the value of their bit score.
-\textbf{(C)} Hexbin plot of the number of sequences and the number of unique genes in each component.
-\textbf{(D)} Histogram of number edges associated with each sequence, \textit{i.e.} the degree of each node.
-Only the lower 99th percentile of the distribution is shown.}
-\label{fig:components}
-\end{figure}
-
-The identification of sets of densely connected nodes in graphs is known as community detection in network analysis.
-While many community detection algorithms are available, only some are commonly used in the context of orthology inference.
-One early method that remains popular is building clusters progressively by identifying nodes that form a triangle with at least two other nodes in the cluster~\cite{Tatusov1997, Jensen2007}.
-Other approaches include the MCL algorithm, which clusters graphs by simulating stochastic flow~\cite{Enright2002, Li2003, Emms2015}, and connected components or other single-linkage criteria~\cite{Remm2001, Linard2011, Train2017, Cosentino2018}.
-While these methods are robust when clustering hit graphs derived from smaller or more diverse sets of genomes, they are not suitable for the large number of closely related genomes in this work since they require relatively few edges to define a cluster.
-For example, the MCL algorithm and connected components method assign a node to a cluster as long as it has a single edge, and triangle clustering only requires two edges to two adjacent nodes.
-
-However, connected components and triangle clustering are special cases of the more general \textit{k}-clique percolation algorithm where \textit{k} equals two and three, respectively.
-The clique percolation algorithm detects clusters by first identifying cliques, sets of nodes which are fully connected, of a specified size \textit{k} in the graph (Fig.~\ref{fig:percolation}A).
-Clusters are then taken as the connected components of an overlap graph where an edge exists between two cliques if they share \textit{k}-1 nodes in common.
-An intuitive way to visualize this algorithm is by ``rolling'' a clique of some size \textit{k} over the graph (Fig.~\ref{fig:percolation}B).
-More specifically, a cluster is initiated when a set of nodes which form a \textit{k}-clique is identified.
-The cluster expands by shifting the \textit{k}-clique to an adjacent \textit{k}-clique that shares \textit{k}-1 nodes in common with the current \textit{k}-clique.
-A cluster stops expanding when there are no adjacent \textit{k}-cliques, and the algorithm terminates when there are no \textit{k}-cliques which are not part of a cluster.
-The strength of this algorithm is its ability to exclude sparsely connected nodes from clusters with an easily tunable parameter \textit{k}.
-Higher values of \textit{k} require greater overlap between a candidate node and those already in the cluster and therefore produce tighter clusters at the cost of excluding more speculative orthology relationships (Fig.~\ref{fig:percolation}C).
-We set \textit{k} to four as compromise between these concerns, yielding 22,813 orthologous groups, a plurality of which contained all 33 species (Fig.~\ref{sfig:clusters}).
-
-\begin{figure}[h!]
-\includegraphics[width=\textwidth]{percolation/out/percolation.png}
-\includegraphics[width=\textwidth]{percolation/out/005F.png}
-\centering
-\caption{\textbf{Clique percolation algorithm.}
-\textbf{(A)} Cliques for \textit{k} equal to three, four, five, and six.
-\textit{k} equal to one and two correspond to a single node and two nodes joined by an edge, respectively.
-\textbf{(B)} Illustration of clique percolation algorithm where \textit{k = 4}.
-\textbf{(C)} A single component clustered by clique percolation with varying values of \textit{k}.
-Nodes are colored according to their cluster.
-If a node belongs to multiple clusters, it uses a blend of those colors.}
-\label{fig:percolation}
-\end{figure}
-
-\subsection{Addition of paralogs to orthologous groups}
-A weakness of the best reciprocal hits criterion is its exclusion of recently diverged paralogs.
-Since only the highest scoring hits for each query are included in the graph, a paralog without a corresponding duplicate in the target genome is ignored if it is marginally more diverged than the other copy.
-This is corrected by adding likely paralogs to the orthologous groups.
-Briefly, the protein sequences in each genome were searched against themselves.
-If the bit score for an intra-genome hit exceeded the bit score of any inter-genome hits for the same query, the two sequences were identified as a paralogous pair.
-Orthologous groups were then supplemented with paralogs by adding the paired sequences for each of the original members of the orthologous group.
-Most orthologous groups contain no paralogs, and those that do generally have few relative to the original number of sequences in the group (Fig.~\ref{sfig:paralogs}).
-
-\subsection{Grouping orthologous groups by gene}
-As genes can have several annotated isoforms, each gene can be associated with several orthologous groups.
-However, the orthologous groups are not organized into gene-level units since they were clustered using sequence similarity only.
-A graph-based approach was therefore used to group orthologous groups with similar sets of parent genes.
-First, a gene overlap graph was constructed by defining an edge between orthologous groups if the intersection of their associated sets of parent genes is at least 50\% of the smaller of the two.
-Gene groups were then taken as the connected components of the resulting graph, yielding 14,909 groups.
-This is commensurate with the roughly 15,000 genes in each genome, which suggests this approach has successfully clustered orthologous groups derived from a common set of parent genes.
-
-\subsection{Initial alignment and selection of representative sequences}
-Since the NCBI annotation pipeline incorporates transcriptome data from a variety of sources, its inputs are heterogeneous in sequencing depth, developmental stage, and tissue of origin across different genomes.
-As a result, some genomes are annotated with different or multiple splice isoforms of a given orthologous gene, which can create complex networks in the resulting hit graph.
-For example, if the genomes are variably annotated with one or both of two distinct isoforms, the resulting graph may contain two clusters connected by a ``bridge'' formed by the genomes which contain only one of the isoforms.
-If the nodes bridging the two clusters form cliques with themselves and the clusters, the clique percolation algorithm will merge all the nodes into a single orthologous group where some genes have multiple associated sequences.
-However, these additional sequences can complicate downstream comparative analyses that may not easily generalize to genes with multiple associated sequences.
-Thus, in our pipeline a single representative was chosen for each gene using an alignment-based strategy detailed in the methods section.
-Briefly, a statistical profile was created from an alignment of the sequences in each orthologous group, and the representative for each gene was chosen as the sequence which best matched this profile.
-
-\subsection{Selection of single copy orthologous groups}
-The criteria for selecting orthologous groups for further analyses depends on the biological question under investigation.
-For example, studies of gene duplication will focus on orthologous groups with paralogs in some lineages but not in others.
-In contrast, analyses which assume functional conservation should restrict the orthologous groups to single copy orthologs since paralogs more frequently undergo functional divergence~\cite{Altenhoff2012, Pegueroles2013, Soria2014}.
-A simple method for identifying such groups is requiring each species to have exactly one associated gene.
-However, since the probability of at least one missing gene annotation approaches one as the total number of genomes increases, this is too restrictive and fails to leverage the redundancy of closely related genomes.
-Instead, a set of phylogenetic diversity criteria detailed in Table~\ref{stable:diversity_criteria} were applied to ensure the major lineages were represented in downstream analyses.
-Furthermore, genome-wide analyses should select one orthologous group per each of the previously identified gene groups as to not bias the results towards genes with many distinct groups of isoforms.
-In summary, orthologous groups failing the phylogenetic diversity criteria were first removed, and the representative for each gene group was chosen as the highest scoring orthologous group when ranked by the number species and the sum of the bit scores associated with each edge.
-This significantly reduced the number of orthologous groups from 22,813 to 8,566.
-
-\subsection{Alignment refinement}
-Though the pipeline’s quality control measures ensure a high degree of overall sequence identity between members of an orthologous group, some sequences contain long ``poorly supported'' segments which have no homology to most or any other sequences in the alignment.
-Since most common multiple sequence alignment algorithms assume the sequences are largely homologous, these segments are sometimes ``over-aligned'' by forcing them into alignment where chance sequence similarities occur.
-Typically, these segments remain contiguous, so the alignments alternate between short runs of columns with few or no gaps and large gap-rich regions (Fig.~\ref{fig:realignment}A-B, left).
-More rarely, when long poorly supported segments are adjacent to a long gap in the same sequence, the two are interlaced, yielding long gaps interrupted by short segments of spurious alignment (Fig.~\ref{fig:realignment}C, left).
-
-The aligner MAFFT has a mode for addressing over-alignment with a parameter, \textit{a\textsubscript{max}}, that adjusts the strength of the correction~\cite{Katoh2013, Katoh2016}.
-\textit{a\textsubscript{max}} varies between 0 and 1, with higher values yielding a stronger correction.
-While values above 0.8 completely eliminate over-alignment and successfully align highly conserved regions, the alignment as a whole is severely degraded, as even homologous sequences with a small amount of divergence are separated by gaps.
-Thus, in our pipeline orthologous groups were aligned in two stages.
-In the first, the sequences were aligned with a strong correction of 0.7.
-Highly conserved regions were identified, which divided the alignment into a complementary set of diverged regions.
-The sequences in each of these regions were extracted and aligned separately with a more conservative value for \textit{a\textsubscript{max}} of 0.4.
-The resulting ``sub-alignments'' were ``stitched'' back into their positions in the original alignment.
-By defining highly conserved ``anchor'' regions, this approach largely prevents the alignment of chance sequence similarities in long poorly supported segments (Fig.~\ref{fig:realignment}, right).
-
-\begin{figure}[h!]
-\includegraphics[width=\textwidth]{realignment/out/realignment.png}
-\centering
-\caption{\textbf{Alignment with long poorly supported segments.}
-The alignments of representative sequences in orthologous groups 0167 \textbf{(A)}, 2770 \textbf{(B)}, and 23D9 \textbf{(C)} before and after refinement.}
-\label{fig:realignment}
-\end{figure}
-
-\subsection{Alignment curation}
-Although the refinement process corrects most cases of over-alignment, the alignment may still contain regions whose aligned segments have poor or inconsistent support.
-For example, long poorly supported segments in internal regions were not removed from the alignment since they are bounded by at least one consensus column to the left and right.
-Additionally, some regions have a significant fraction of sequences with strongly supported segments, but the observed gap pattern is discordant with the expected phylogenetic relationships.
-Since they are present in so few sequences, the former segments are likely artifactual, resulting from errors during assembly or annotation.
-(Biological explanations such as alternative splice sites, frameshift mutations, or transposition events are also possible, however.) In contrast, the high sequence identity and clear boundaries of the segments in the latter regions suggest they are conserved but skipped exons.
-Given the heterogeneous sourcing of the transcript evidence, these sequences containing these segments are likely splice isoforms specific to certain tissues or developmental stage.
-
-Since the segments in these regions are likely the result of incorrect or incomplete annotations rather than meaningful biological variation, maintaining them in the alignments would propagate spurious homologies to subsequent analyses.
-This is a common issue in alignments generated by automated pipelines, so downstream analyses often focus on the strongly supported regions by removing or ``trimming'' columns below some threshold number of gaps or sequence identity~\cite{Castresana2000, CapellaGutierrez2009}.
-This approach, however, is inadequate if the taxonomic sampling is dense, as a single indel event along a lineage containing many species can increase the number of gaps above the threshold.
-Moreover, as this method does not incorporate any spatial information, it can rapidly alternate between trimming and preserving columns.
-Thus, it can severely disrupt any analyses which are sensitive to the spatial organization of an alignment.
-
-Phylogenetic HMMs (phylo-HMMs) are statistical models that incorporate phylogenetic and spatial information to calculate the probability that each observation in a sequence was generated by one of several hidden states~\cite{Felsenstein1996, Siepel2004}.
-Since they can evaluate both the probability of a gap pattern in a column given the known phylogenetic relationships and the local context, a phylo-HMM was used to segment the alignment into contiguous regions with different patterns of gaps.
-A fully specified phylo-HMM requires a fixed number of hidden states and a probability distribution for each.
-Thus, we identified three distinct types of regions in the alignments, roughly corresponding to regions with strong phylogenetic evidence, regions with a stable gap pattern discordant with the expected phylogenetic relationships, and regions with poorly supported segments.
-For simplicity, however, we refer to the states that generate each type of region as 1, 2, and 3, respectively.
-To model probability distributions for each state, we first conceptualized the observed alignments as the superposition of two distinct processes (Fig.~\ref{fig:hmm}A, left).
-The first is a phylogenetic process which evolves and splits a single ancestral sequence over time according to a tree.
-The second is the annotation process which can erroneously exclude or include segments from a sequence.
-The result is an alignment of annotated sequences which contains evolutionary information obscured by noise from the annotation process, shown here by the exclusion of three N-terminal residues in the fourth annotated sequence.
-To simplify modeling this behavior with an HMM, we coded the sequences into binary symbols.
-The distributions for each state then consisted of two components derived from the encoded sequences (Fig.~\ref{fig:hmm}A, right).
-The first component models the gap pattern with a Markov process.
-This Markov process is in turn composed of two subprocesses where the first is a phylogenetic process, and the second is a jump process.
-These subprocess roughly correspond to changes caused by evolution and annotation, respectively.
-Because this first component did not fully capture the propensity for the gap patterns to remain constant, we included a second component that models the ``gap stickiness'' as a beta-binomial random variable by counting the number of symbols that remain constant between columns.
-Each component is associated with a set of parameters, and the unique parameters for each state yield its characteristic gap pattern and gap stickiness.
-
-After the model was trained on manually labeled examples, it was used to assign a label to the columns in each alignment.
-Columns assigned to state 1 are the regions of interest for downstream analyses since the gaps generally follow the expected pattern given the phylogenetic tree.
-In contrast, columns assigned to states 2 and 3 largely corresponded to the long poorly supported segments and phylogenetically discordant regions discussed previously and were therefore removed from the alignments.
-In the example decoded alignment shown, the decoded states closely follow the expected patterns (Fig.~\ref{fig:hmm}B).
-Overall, 29\% of alignments were trimmed of at least one segment or region.
-However, 87\% of the trims were segment trims, meaning the removed segments were largely inferred as state 3 and therefore were likely long poorly supported segments aligned to few if any other sequences (Fig.~\ref{sfig:insertion_trim}).
-
-\begin{figure}[h!]
-\includegraphics[width=\textwidth]{hmm/out/hmm_architecture.png}
-\includegraphics[width=\textwidth]{hmm/out/2252.png}
-\centering
-\caption{\textbf{HMM emission architecture and a decoded alignment.}
-\textbf{(A)} Schematic of theoretical alignment generating process and corresponding probabilistic components in HMM.
-White and colored boxes indicate gap and non-gap symbols in the biological sequences, respectively.
-White and grey boxes indicate gap and non-gap symbols in the encoded sequences, respectively.
-c\textsubscript{0} and c\textsubscript{1} indicate the first and second columns in the alignment, respectively.
-The parameters associated with each component are shown to the right.
-\textbf{(B)} The alignment of the representative sequences in orthologous group 2252 decoded using the trained HMM.}
-\label{fig:hmm}
-\end{figure}
-
-Though the phylo-HMM removed phylogenetically incongruent insertions, some sequences still contained extensive segments of uninterrupted gaps.
-These segments are easily identified in regions which are otherwise highly conserved, so they are also likely the result of incorrect or incomplete annotations.
-However, they can also span more diverged regions, which complicates a simple rule- or threshold-based approach for identifying them.
-Thus, another phylo-HMM was trained to label each position in a sequence as generated by either a ``missing'' or ``not missing'' state.
-In this case, though, the aligned sequences were processed individually and not as aligned columns.
-As the previous phylo-HMM already ensured each column has sufficient support, these labels can instead be used to exclude sequences from downstream analyses depending on the amount of tolerated overlap with the regions of interest.
-Overall, 15\% of alignments have at least one sequence with a segment of ``missing'' data (Fig.~\ref{sfig:missing_trim}).
-
-\subsection{Inference of species trees}
-Many phylogenetic methods require a species tree to inform the evolutionary relationships between sequences.
-In fact, the phylo-HMMs discussed previously used a species tree as an input, though we omitted this detail for clarity of exposition.
-Therefore, to support the curation step and other downstream analyses, we sought to infer phylogenetic trees from the aligned sequences.
-However, since the roots of phylogenetic trees are not identifiable with commonly used time-reversible substitution models, we repeated the orthology inference pipeline with the outgroup species \textit{Scaptodrosophila lebanonensis}.
-Afterwards, we inferred phylogenetic trees using the LG model of amino acid substitution from 100 meta-alignments sampled from alignments of single copy orthologous groups.
-We then combined them into a single consensus tree (Fig.~\ref{fig:trees}A).
-To provide a similar tree for the analysis of non-coding regions, we inferred phylogenetic trees using the GTR model of nucleotide substitution from 100 meta-alignments sampled from nucleotide alignments which were ``reverse translated'' from the protein alignments and their corresponding coding sequences (\ref{fig:trees}B).
-Both trees have an identical topology, which is consistent with other published phylogenies~\cite{D12GC2007, DaLage2007}.
-
-\begin{figure}[h!]
-\includegraphics[width=\textwidth]{trees/out/trees1.png}
-\centering
-\caption{\textbf{Phylogenetic tree of species.}
-\textbf{(A)} Consensus tree from LG model fit to meta-alignments directly sampled from the original protein alignments.
-\textbf{(B)} Consensus tree from GTR model fit to meta-alignments sampled from ``reverse translated'' nucleotide alignments.
-Values at nodes are bootstrap percentages.}
-\label{fig:trees}
-\end{figure}
-
-\section{Discussion}
-\subsection{Developments in genome assembly and annotation challenge existing methods of orthology inference}
-The orthologous groups and alignments yielded by this pipeline are a valuable resource for comparative studies of gene birth/death processes and protein evolution at the level of both entire proteomes and specific gene families in the \textit{Drosophila} genus.
-To assist these efforts, the final alignments after curation and the labels from the ``missing'' phylo-HMM are provided as supplemental data.
-Although this work focused on single copy orthologs, other studies may require different subsets of orthologous groups that demand other pre-processing and alignment strategies.
-Therefore, we have included the orthologous groups and the initial alignments with and without non-representative sequences in the supplemental data as well.
-While we anticipate these resources will remain relevant in the near term, the trends that permitted this work to substantially improve on previous efforts will render them obsolete in the coming years as more \textit{Drosophila} genomes are assembled and annotated.
-However, an authoritative and lasting set of orthologous groups and alignments is not the primary goal of this work.
-Instead, it serves as a case study in how dense taxonomic sampling and modern genome assembly and annotation pipelines present new opportunities and challenges to the traditional techniques for identifying and aligning orthologous groups.
-
-For example, despite many additional pre-processing steps and other tweaks introduced by later authors, the basic framework of orthology inference by clustering the hit graph has remained largely unchanged in the past twenty years~\cite{Tatusov1997, Remm2001, Li2003, Jensen2007, Linard2011, Emms2015, Train2017, Cosentino2018}.
-This longevity is a testament to the robustness of the underlying idea that orthologous proteins should consistently identify each other as the most similar pairs between their genomes.
-Even as the number of genomes and their taxonomic density has increased dramatically, many orthology inference pipelines continue to use algorithms which were originally applied to sets of far fewer and more distantly related genomes.
-This mismatch in scale increases the chance of propagating annotation errors since only a small number of edges are needed to create or merge clusters.
-Thus, we instead applied a generalization of the triangle and connected components clustering methods called \textit{k}-clique percolation where \textit{k} is a tunable parameter that influences the tightness of a cluster.
-The optimal value of \textit{k} for a given set of genomes is unclear and likely depends on the desired trade-off between sensitivity and specificity.
-Furthermore, \textit{k} is not necessarily a global parameter and can instead depend on the properties of each connected component.
-For example, one possibility is to take an entire component as an orthologous group if its number of unique genes and unique species are equal since all the sequences are isoforms of a single set of genes.
-This would effectively set \text{k} equal to one for this component.
-Another approach is to make \textit{k} an decreasing function of the density of edges, so sparser graphs are clustered more permissively.
-However, percolation theory or simulations may yield additional insights.
-
-Another challenge is the annotation of multiple isoforms for a single gene.
-Though prior pipelines have generally selected the longest isoform as the representative before conducting the orthology searches, if the sequences do not share a common intron-exon structure this approach can introduce artifacts or other issues during alignment.
-Instead, as protein sequences are increasingly derived from or linked to genomic sequences, we sought to incorporate the full annotations into the orthology inference pipeline.
-This, however, created two additional complications.
-First, a single gene could have several associated orthologous groups if its isoforms belonged to different clusters.
-Second, a single orthologous group could have several isoforms of a single gene if its isoforms were clustered together.
-In both cases, the presence of multiple sequences for a single gene creates ambiguities over which is the ``primary'' isoform.
-Since the first occurs at the level of orthologous groups, the orthologous groups were first grouped by the similarity of their parent genes using a graph-based strategy.
-Afterwards, a single representative is easily chosen as the group with the largest number of distinct species, though other criteria are possible.
-Since the second occurs within an orthologous group, the sequences were first aligned, and a representative for each gene was chosen as the sequence which was most concordant with this initial alignment.
-
-The second major innovation in this work is its method for the refinement and trimming of alignments.
-Since sequences produced from automated annotation pipelines can contain long segments which are not homologous to most or any other sequences in their respective orthologous groups, their alignments may contain over-aligned or poorly supported regions which can introduce artifacts into downstream analyses.
-Thus, in refinement over-alignment is avoided by aligning the sequences in two stages.
-In the first highly conserved regions are aligned using a strong correction for over-alignment, and in the second more diverged regions are aligned with a weaker correction.
-This process usually prevents errors caused by long poorly supported segments without degrading the quality of the alignment.
-In trimming, a phylo-HMM is used to remove regions which are poorly supported by the phylogenetic consensus.
-
-\subsection{Comprehensive bioinformatic analyses of proteins will depend on splice-sensitive alignments}
-While the combination of these two steps yielded high-quality alignments that are suitable for further analyses, they are an \textit{ad hoc} fix for underlying issues with the gene models and alignment algorithms.
-The most principled solution is to optimize or supplement the gene models using the initial alignments generated by the orthology inference pipeline, which is possible with tools such as OMGene or OrthoFiller~\cite{Dunne2018, Dunne2017}.
-However, if preserving the original annotations is desired or necessary, the remaining possibility is to correct the alignments as we have done here.
-In fact, the errors we sought to address broadly stem from shortcomings of current alignment algorithms rather than errors in the sequences themselves.
-Though the scoring functions of modern multiple sequence alignment algorithms are complex, they are generally derived from models that penalize gaps with a linear or affine cost.
-As a result, they often interlace gaps with short, aligned segments rather than a single long gap.
-However, when the sequences are different isoforms of a single gene, their alignment will necessarily contain contiguous exon sized gaps.
-The same is true when aligning isoforms of diverged orthologs, though the relationship between their exons may be complex.
-
-The most popular aligners for protein sequences (Clustal Omega~\cite{Sievers2017}, MAFFT~\cite{Katoh2013}, MUSCLE~\cite{Edgar2004}, T-Coffee~\cite{Notredame2000}) do not include splice sites in their alignments, which makes them prone to aligning non-homologous exons.
-Current algorithms can easily be extended to incorporate splice sites by coding them as a new symbol and preventing alignment between splice sites and amino acids, which was recently implemented in the aligner, \textit{Mirage}~\cite{Nord2018}.
-The biggest challenge in practice, however, is mapping a protein sequence to its genomic sequence to identify splice sites in the protein sequence.
-Although this information can in principle be derived from the GTF annotation files produced by the NCBI pipeline, annotating the splice junctions in the protein sequences themselves would facilitate splice-sensitive alignment.
-
-These improvements would enhance rather than replace the phylo-HMM trimming method developed in this work.
-The model could easily be extended to include a state that outputs a splice symbol before transitioning to one of the states in the current architecture.
-This intermediate state would increase the accuracy of state inference since a splice symbol followed by a phylogenetically discordant gap pattern would strongly signal a state 2 region.
-The association between state 2 and skipped exons can be made explicit by requiring that transitions to and from state 2 first proceed through the splice state.
-This of course depends on proper labeling of the training data, which would be trivial since the boundaries between exons would be marked by splice site symbols rather than inferred from gap patterns.
-Unfortunately, this would not allow the phylo-HMM to label extended exon boundaries as state 2 since they would not be bounded by splice symbols to the left and right.
-Accordingly, the phylo-HMM would need to permit transitions between state 3 and any other state to accommodate more complex splice variants and other annotation errors.
-Thus, this extended phylo-HMM would combine the strengths of splice-sensitive alignment with the more heuristic approach used here.
-Since state 2 inferences would necessarily correspond to skipped exons, they would be suitable for analyses of this form of alternative splicing.
-Though state 3 inferences would not directly correspond to specific biological process, they still have value as spatially and phylogenetically aware labels for trimming poorly supported segments from alignments.
-
-The phylo-HMM could be further enhanced by expanding its emission distribution to include more symbols in the amino acid alphabet.
-This would allow it to better model observed substitution patterns between specific symbols, for example the high rate of exchange between gaps and glutamine residues caused by polyglutamine tracts.
-The transitions between amino acids could be parametrized with a published matrix such as LG~\cite{Le2008}, but the transitions between amino acids and gaps would be inferred from labeled data.
-It is unclear if the resulting gain in accuracy would justify the increased computational burden, however.
-
-\subsection{Accessible computational tools will facilitate future comparative studies}
-Though benchmarks are available for optimizing and comparing methods of orthology inference, the metrics are calculated over a set of reference genomes which are sparsely sampled over a broad taxonomic range, so it is unclear if they are informative for method designed to yield robust inferences when the genomes are highly related~\cite{Nevers2022}.
-Furthermore, the heterogeneity of genome architectures and annotations may require quality assurance methods tailored to each set of genomes.
-Thus, there is likely no one-size-fits-all approach to orthology inference, and with many other standalone programs available for more standard use cases (Hieranoid~\cite{Kaduk2017}, OMA standalone~\cite{Altenhoff2019}, OrthoFinder~\cite{Emms2019}, OrthoInspector~\cite{Linard2014}, Orthologer~\cite{Zdobnov2020}), we have chosen not to package the code into an end-to-end pipeline.
-Instead, we have devoted considerable attention to organizing and documenting the code to make it accessible to a newcomer and thereby facilitate the adaptation of specific steps to similar projects as needed.
-
-In contrast, though many HMM packages are available for the Python programming language, we found none were satisfactorily documented or contained tutorials to introduce HMMs and their APIs to a wide audience.
-We therefore refactored this code into a package available on PyPI and GitHub called Homomorph.
-The package itself only implements standard HMM algorithms, but the GitHub repository includes tutorials that introduce the API and implement training routines.
-Similar tutorials for machine learning libraries such as TensorFlow have undoubtedly fueled the application of neural networks across diverse fields, but HMMs are also powerful models that can be more appropriate when the data obey certain statistical or structural constraints.
-Thus, we hope this package and its accompanying tutorials will serve as an on-ramp to HMMs and spur their greater adoption by non-specialists.
-
-Though databases of orthologous groups such as COGs~\cite{Galperin2020}, Ensembl Compara~\cite{Herrero2016}, EggNOG~\cite{HuertaCepas2018}, OMA~\cite{Altenhoff2020}, OrthoDB~\cite{Zdobnov2020}, OrthoInspector~\cite{Nevers2018}, and OrthoMCL~\cite{Chen2006} will continue to be useful for comparative studies across broad taxonomic ranges, the increasing speed at which high-quality genome assemblies and annotations are produced means no single database can encompass the most complete data.
-Furthermore, since many early comparative genomics studies spanned diverse branches of the tree of life, future research will likely prioritize taxonomic depth over breadth.
-Thus, custom sets of orthologous groups will grow more and not less common.
-Despite the challenges these developments pose, they also present new opportunities to bridge the gap between mutational and macroevolutionary processes.
 
@@
 
+\section{Results}
+\subsection{Pipeline overview}
+Our pipeline follows a similar overall approach to other graph-based methods of orthology inference.
+First, protein sequences from annotated genomes are collected (Fig.~\ref{fig:graphical_overview}A), and homology searches are conducted between all query-target pairs of genomes (Fig.~\ref{fig:graphical_overview}B).
+The raw output from these homology searches is processed to yield best hits between pairs of sequences (Fig.~\ref{fig:graphical_overview}C).
+Next, the network of best hits is clustered into self-consistent orthologous groups (Fig.~\ref{fig:graphical_overview}D).
+Since genes can have multiplied associated isoforms, we then implemented a novel second clustering step where orthologous groups are grouped by their parent genes, which are represented by the two sets of clusters with warm and cool colors, respectively (Fig.~\ref{fig:graphical_overview}E).
+Finally, representative sequences in each orthologous group are aligned (Fig.~\ref{fig:graphical_overview}F).
+In the following sections, we discuss each of these and other steps which were omitted for clarity in greater detail.
+
+\subsection{Input genomes and pre-processing}
+All annotated genomes in the genus \textit{Drosophila} available in April 2022 were downloaded from NCBI's RefSeq database.
+The assemblies annotated by the NCBI eukaryotic genome annotation pipeline have passed several quality checks and all have supporting transcript evidence, so the annotations are generally highly complete~\cite{Thibaud-Nissen2013}.
+The \textit{D.
+miranda} annotation was excluded due to its unusual karyotype~\cite{Dobzhansky1935}.
+Other annotations were excluded after preliminary clustering showed a deficiency in the number of orthologous groups containing those genomes, indicating their annotations were less complete (data not shown).
+The \textit{D.
+melanogaster} annotation was downloaded from FlyBase~\cite{Gramates2022}.
+In total, the input data consists of 33 genomes, which are listed in Table~\ref{stable:genomes}.
+Many genes have transcripts that differ only in their UTRs, and as a result there are many duplicate protein sequences in the annotations.
+Though not strictly necessary, we removed the duplicates in our pipeline, which greatly reduced the computational burden of later steps.
+
+\subsection{Extraction of best hits from BLAST output}
+The protein sequences in each genome annotation were searched against each other in reciprocal pairs using BLAST, yielding a list of high-scoring segment pairs (HSPs) for each query-target pair~\cite{Camacho2009}.
+HSPs are local alignments, meaning they do not necessarily span the entire lengths of the query and target sequences.
+Consequently, the search algorithm may return multiple HSPs for each query-target pair if statistically significant regions of homology are separated by nonhomologous or poorly conserved regions.
+Though the most significant HSP is often used to represent all HSPs between a query-target pair, this approach can fail to rank the pairs by their overall significance if their alignments are broken into multiple HSPs.
+Furthermore, since query-target pairs were later filtered by the amount overlap between their sequences, it can also exclude pairs that pass the overlap threshold even if the most significant HSP alone does not.
+Thus, HSPs were merged into a single object called a hit.
+The best hits for each query were then taken as the highest-scoring hits that passed a minimum overlap criterion and were reciprocal between the query and target sequences.
+
+\subsection{Clustering in orthologous groups}
+The best hits between sequences are naturally visualized as a graph where sequences are nodes and best hits are edges between nodes.
+Two connected components, sets of nodes joined by a sequence of edges, are shown (Fig.~\ref{fig:components}A-B).
+The sequences in the first (Fig.~\ref{fig:components}A) all contain C2H2 zinc fingers, whereas the sequences in the second (Fig.~\ref{fig:components}B) are members of the Par-1 family of serine/threonine protein kinases.
+In both components, some sets of nodes have a high density of edges, forming distinct clusters, whereas other nodes are only sparsely connected to their neighbors.
+To better understand the structure of these two components, we calculated the number of sequences, unique genes, and unique species in each.
+The first has 385, 346, and 33 sequences, genes, and species, respectively, and the second has 222, 33, and 33 sequences, genes, and species, respectively.
+We then plotted the relationship between the number sequences and unique genes across all components to see if this pattern holds true generally (Fig.~\ref{fig:components}C).
+Two distinct trendlines are apparent.
+The first increases linearly with the number of sequences with a slope of one, indicating each sequence is generally associated with a unique gene.
+The second is constant with an intercept of 33, indicating the number of unique genes quickly saturates at the total number of genomes.
+Thus, there are generally two classes of components.
+The first is composed of many distinct genes, whereas the second is composed of many different isoforms of a single group of genes.
+
+The diffuse networks observed in the first component class are likely the result of a combination of factors, including rapid evolution, gene duplication, and annotation errors.
+Regardless of their origin, these hits are not strong candidates for comparative analyses since an orthology relationship is supported by relatively few genome pairs.
+Instead, likely orthologs should consistently identify each other as best reciprocal hits across many genome pairs.
+The same is true of the hits in the second component class.
+Although the genes as a unit form a single orthologous group, sequences with few hits are likely non-conserved or tissue-specific isoforms.
+Thus, orthologous groups can be operationally defined as self-consistent clusters in the hit graph.
+However, sequence divergence or assembly and annotation errors may prevent a best reciprocal hit between orthologous sequences across all genome pairs.
+In fact, although the most common number of reciprocal hits is 32, one fewer than the total number of genomes, many sequences have fewer (Fig.~\ref{fig:components}D).
+Thus, the clustering method should require a high degree of self-consistency without demanding complete consensus.
+
+The identification of sets of densely connected nodes in graphs is known as community detection in network analysis.
+While many community detection algorithms are available, only some are commonly used in the context of orthology inference.
+One early method that remains popular is building clusters progressively by identifying nodes that form a triangle with at least two other nodes in the cluster~\cite{Tatusov1997, Jensen2007}.
+Other approaches include the MCL algorithm, which clusters graphs by simulating stochastic flow~\cite{Enright2002, Li2003, Emms2015}, and connected components or other single-linkage criteria~\cite{Remm2001, Linard2011, Train2017, Cosentino2018}.
+While these methods are robust when clustering hit graphs derived from smaller or more diverse sets of genomes, they are not suitable for the large number of closely related genomes in this work since they require relatively few edges to define a cluster.
+For example, the MCL algorithm and connected components method assign a node to a cluster as long as it has a single edge, and triangle clustering only requires two edges to two adjacent nodes.
+
+However, connected components and triangle clustering are special cases of the more general \textit{k}-clique percolation algorithm where \textit{k} equals two and three, respectively.
+The clique percolation algorithm detects clusters by first identifying cliques, sets of nodes which are fully connected, of a specified size \textit{k} in the graph (Fig.~\ref{fig:percolation}A).
+Clusters are then taken as the connected components of an overlap graph where an edge exists between two cliques if they share \textit{k}-1 nodes in common.
+An intuitive way to visualize this algorithm is by ``rolling'' a clique of some size \textit{k} over the graph (Fig.~\ref{fig:percolation}B).
+More specifically, a cluster is initiated when a set of nodes which form a \textit{k}-clique is identified.
+The cluster expands by shifting the \textit{k}-clique to an adjacent \textit{k}-clique that shares \textit{k}-1 nodes in common with the current \textit{k}-clique.
+A cluster stops expanding when there are no adjacent \textit{k}-cliques, and the algorithm terminates when there are no \textit{k}-cliques which are not part of a cluster.
+The strength of this algorithm is its ability to exclude sparsely connected nodes from clusters with an easily tunable parameter \textit{k}.
+Higher values of \textit{k} require greater overlap between a candidate node and those already in the cluster and therefore produce tighter clusters at the cost of excluding more speculative orthology relationships (Fig.~\ref{fig:percolation}C).
+We set \textit{k} to four as compromise between these concerns, yielding 22,813 orthologous groups, a plurality of which contained all 33 species (Fig.~\ref{sfig:clusters}).
+
+\subsection{Addition of paralogs to orthologous groups}
+A weakness of the best reciprocal hits criterion is its exclusion of recently diverged paralogs.
+Since only the highest scoring hits for each query are included in the graph, a paralog without a corresponding duplicate in the target genome is ignored if it is marginally more diverged than the other copy.
+This is corrected by adding likely paralogs to the orthologous groups.
+Briefly, the protein sequences in each genome were searched against themselves.
+If the bit score for an intra-genome hit exceeded the bit score of any inter-genome hits for the same query, the two sequences were identified as a paralogous pair.
+Orthologous groups were then supplemented with paralogs by adding the paired sequences for each of the original members of the orthologous group.
+Most orthologous groups contain no paralogs, and those that do generally have few relative to the original number of sequences in the group (Fig.~\ref{sfig:paralogs}).
+
+\subsection{Grouping orthologous groups by gene}
+As genes can have several annotated isoforms, each gene can be associated with several orthologous groups.
+However, the orthologous groups are not organized into gene-level units since they were clustered using sequence similarity only.
+A graph-based approach was therefore used to group orthologous groups with similar sets of parent genes.
+First, a gene overlap graph was constructed by defining an edge between orthologous groups if the intersection of their associated sets of parent genes is at least 50\% of the smaller of the two.
+Gene groups were then taken as the connected components of the resulting graph, yielding 14,909 groups.
+This is commensurate with the roughly 15,000 genes in each genome, which suggests this approach has successfully clustered orthologous groups derived from a common set of parent genes.
+
+\subsection{Initial alignment and selection of representative sequences}
+Since the NCBI annotation pipeline incorporates transcriptome data from a variety of sources, its inputs are heterogeneous in sequencing depth, developmental stage, and tissue of origin across different genomes.
+As a result, some genomes are annotated with different or multiple splice isoforms of a given orthologous gene, which can create complex networks in the resulting hit graph.
+For example, if the genomes are variably annotated with one or both of two distinct isoforms, the resulting graph may contain two clusters connected by a ``bridge'' formed by the genomes which contain only one of the isoforms.
+If the nodes bridging the two clusters form cliques with themselves and the clusters, the clique percolation algorithm will merge all the nodes into a single orthologous group where some genes have multiple associated sequences.
+However, these additional sequences can complicate downstream comparative analyses that may not easily generalize to genes with multiple associated sequences.
+Thus, in our pipeline a single representative was chosen for each gene using an alignment-based strategy detailed in the methods section.
+Briefly, a statistical profile was created from an alignment of the sequences in each orthologous group, and the representative for each gene was chosen as the sequence which best matched this profile.
+
+\subsection{Selection of single copy orthologous groups}
+The criteria for selecting orthologous groups for further analyses depends on the biological question under investigation.
+For example, studies of gene duplication will focus on orthologous groups with paralogs in some lineages but not in others.
+In contrast, analyses which assume functional conservation should restrict the orthologous groups to single copy orthologs since paralogs more frequently undergo functional divergence~\cite{Altenhoff2012, Pegueroles2013, Soria2014}.
+A simple method for identifying such groups is requiring each species to have exactly one associated gene.
+However, since the probability of at least one missing gene annotation approaches one as the total number of genomes increases, this is too restrictive and fails to leverage the redundancy of closely related genomes.
+Instead, a set of phylogenetic diversity criteria detailed in Table~\ref{stable:diversity_criteria} were applied to ensure the major lineages were represented in downstream analyses.
+Furthermore, genome-wide analyses should select one orthologous group per each of the previously identified gene groups as to not bias the results towards genes with many distinct groups of isoforms.
+In summary, orthologous groups failing the phylogenetic diversity criteria were first removed, and the representative for each gene group was chosen as the highest scoring orthologous group when ranked by the number species and the sum of the bit scores associated with each edge.
+This significantly reduced the number of orthologous groups from 22,813 to 8,566.
+
+\subsection{Alignment refinement}
+Though the pipeline’s quality control measures ensure a high degree of overall sequence identity between members of an orthologous group, some sequences contain long ``poorly supported'' segments which have no homology to most or any other sequences in the alignment.
+Since most common multiple sequence alignment algorithms assume the sequences are largely homologous, these segments are sometimes ``over-aligned'' by forcing them into alignment where chance sequence similarities occur.
+Typically, these segments remain contiguous, so the alignments alternate between short runs of columns with few or no gaps and large gap-rich regions (Fig.~\ref{fig:realignment}A-B, left).
+More rarely, when long poorly supported segments are adjacent to a long gap in the same sequence, the two are interlaced, yielding long gaps interrupted by short segments of spurious alignment (Fig.~\ref{fig:realignment}C, left).
+
+The aligner MAFFT has a mode for addressing over-alignment with a parameter, \textit{a\textsubscript{max}}, that adjusts the strength of the correction~\cite{Katoh2013, Katoh2016}.
+\textit{a\textsubscript{max}} varies between 0 and 1, with higher values yielding a stronger correction.
+While values above 0.8 completely eliminate over-alignment and successfully align highly conserved regions, the alignment as a whole is severely degraded, as even homologous sequences with a small amount of divergence are separated by gaps.
+Thus, in our pipeline orthologous groups were aligned in two stages.
+In the first, the sequences were aligned with a strong correction of 0.7.
+Highly conserved regions were identified, which divided the alignment into a complementary set of diverged regions.
+The sequences in each of these regions were extracted and aligned separately with a more conservative value for \textit{a\textsubscript{max}} of 0.4.
+The resulting ``sub-alignments'' were ``stitched'' back into their positions in the original alignment.
+By defining highly conserved ``anchor'' regions, this approach largely prevents the alignment of chance sequence similarities in long poorly supported segments (Fig.~\ref{fig:realignment}, right).
+
+\subsection{Alignment curation}
+Although the refinement process corrects most cases of over-alignment, the alignment may still contain regions whose aligned segments have poor or inconsistent support.
+For example, long poorly supported segments in internal regions were not removed from the alignment since they are bounded by at least one consensus column to the left and right.
+Additionally, some regions have a significant fraction of sequences with strongly supported segments, but the observed gap pattern is discordant with the expected phylogenetic relationships.
+Since they are present in so few sequences, the former segments are likely artifactual, resulting from errors during assembly or annotation.
+(Biological explanations such as alternative splice sites, frameshift mutations, or transposition events are also possible, however.) In contrast, the high sequence identity and clear boundaries of the segments in the latter regions suggest they are conserved but skipped exons.
+Given the heterogeneous sourcing of the transcript evidence, these sequences containing these segments are likely splice isoforms specific to certain tissues or developmental stage.
+
+Since the segments in these regions are likely the result of incorrect or incomplete annotations rather than meaningful biological variation, maintaining them in the alignments would propagate spurious homologies to subsequent analyses.
+This is a common issue in alignments generated by automated pipelines, so downstream analyses often focus on the strongly supported regions by removing or ``trimming'' columns below some threshold number of gaps or sequence identity~\cite{Castresana2000, CapellaGutierrez2009}.
+This approach, however, is inadequate if the taxonomic sampling is dense, as a single indel event along a lineage containing many species can increase the number of gaps above the threshold.
+Moreover, as this method does not incorporate any spatial information, it can rapidly alternate between trimming and preserving columns.
+Thus, it can severely disrupt any analyses which are sensitive to the spatial organization of an alignment.
+
+Phylogenetic HMMs (phylo-HMMs) are statistical models that incorporate phylogenetic and spatial information to calculate the probability that each observation in a sequence was generated by one of several hidden states~\cite{Felsenstein1996, Siepel2004}.
+Since they can evaluate both the probability of a gap pattern in a column given the known phylogenetic relationships and the local context, a phylo-HMM was used to segment the alignment into contiguous regions with different patterns of gaps.
+A fully specified phylo-HMM requires a fixed number of hidden states and a probability distribution for each.
+Thus, we identified three distinct types of regions in the alignments, roughly corresponding to regions with strong phylogenetic evidence, regions with a stable gap pattern discordant with the expected phylogenetic relationships, and regions with poorly supported segments.
+For simplicity, however, we refer to the states that generate each type of region as 1, 2, and 3, respectively.
+To model probability distributions for each state, we first conceptualized the observed alignments as the superposition of two distinct processes (Fig.~\ref{fig:hmm}A, left).
+The first is a phylogenetic process which evolves and splits a single ancestral sequence over time according to a tree.
+The second is the annotation process which can erroneously exclude or include segments from a sequence.
+The result is an alignment of annotated sequences which contains evolutionary information obscured by noise from the annotation process, shown here by the exclusion of three N-terminal residues in the fourth annotated sequence.
+To simplify modeling this behavior with an HMM, we coded the sequences into binary symbols.
+The distributions for each state then consisted of two components derived from the encoded sequences (Fig.~\ref{fig:hmm}A, right).
+The first component models the gap pattern with a Markov process.
+This Markov process is in turn composed of two subprocesses where the first is a phylogenetic process, and the second is a jump process.
+These subprocess roughly correspond to changes caused by evolution and annotation, respectively.
+Because this first component did not fully capture the propensity for the gap patterns to remain constant, we included a second component that models the ``gap stickiness'' as a beta-binomial random variable by counting the number of symbols that remain constant between columns.
+Each component is associated with a set of parameters, and the unique parameters for each state yield its characteristic gap pattern and gap stickiness.
+
+After the model was trained on manually labeled examples, it was used to assign a label to the columns in each alignment.
+Columns assigned to state 1 are the regions of interest for downstream analyses since the gaps generally follow the expected pattern given the phylogenetic tree.
+In contrast, columns assigned to states 2 and 3 largely corresponded to the long poorly supported segments and phylogenetically discordant regions discussed previously and were therefore removed from the alignments.
+In the example decoded alignment shown, the decoded states closely follow the expected patterns (Fig.~\ref{fig:hmm}B).
+Overall, 29\% of alignments were trimmed of at least one segment or region.
+However, 87\% of the trims were segment trims, meaning the removed segments were largely inferred as state 3 and therefore were likely long poorly supported segments aligned to few if any other sequences (Fig.~\ref{sfig:insertion_trim}).
+
+Though the phylo-HMM removed phylogenetically incongruent insertions, some sequences still contained extensive segments of uninterrupted gaps.
+These segments are easily identified in regions which are otherwise highly conserved, so they are also likely the result of incorrect or incomplete annotations.
+However, they can also span more diverged regions, which complicates a simple rule- or threshold-based approach for identifying them.
+Thus, another phylo-HMM was trained to label each position in a sequence as generated by either a ``missing'' or ``not missing'' state.
+In this case, though, the aligned sequences were processed individually and not as aligned columns.
+As the previous phylo-HMM already ensured each column has sufficient support, these labels can instead be used to exclude sequences from downstream analyses depending on the amount of tolerated overlap with the regions of interest.
+Overall, 15\% of alignments have at least one sequence with a segment of ``missing'' data (Fig.~\ref{sfig:missing_trim}).
+
+\subsection{Inference of species trees}
+Many phylogenetic methods require a species tree to inform the evolutionary relationships between sequences.
+In fact, the phylo-HMMs discussed previously used a species tree as an input, though we omitted this detail for clarity of exposition.
+Therefore, to support the curation step and other downstream analyses, we sought to infer phylogenetic trees from the aligned sequences.
+However, since the roots of phylogenetic trees are not identifiable with commonly used time-reversible substitution models, we repeated the orthology inference pipeline with the outgroup species \textit{Scaptodrosophila lebanonensis}.
+Afterwards, we inferred phylogenetic trees using the LG model of amino acid substitution from 100 meta-alignments sampled from alignments of single copy orthologous groups.
+We then combined them into a single consensus tree (Fig.~\ref{fig:trees}A).
+To provide a similar tree for the analysis of non-coding regions, we inferred phylogenetic trees using the GTR model of nucleotide substitution from 100 meta-alignments sampled from nucleotide alignments which were ``reverse translated'' from the protein alignments and their corresponding coding sequences (\ref{fig:trees}B).
+Both trees have an identical topology, which is consistent with other published phylogenies~\cite{D12GC2007, DaLage2007}.
+
+\section{Discussion}
+\subsection{Developments in genome assembly and annotation challenge existing methods of orthology inference}
+The orthologous groups and alignments yielded by this pipeline are a valuable resource for comparative studies of gene birth/death processes and protein evolution at the level of both entire proteomes and specific gene families in the \textit{Drosophila} genus.
+To assist these efforts, the final alignments after curation and the labels from the ``missing'' phylo-HMM are provided as supplemental data.
+Although this work focused on single copy orthologs, other studies may require different subsets of orthologous groups that demand other pre-processing and alignment strategies.
+Therefore, we have included the orthologous groups and the initial alignments with and without non-representative sequences in the supplemental data as well.
+While we anticipate these resources will remain relevant in the near term, the trends that permitted this work to substantially improve on previous efforts will render them obsolete in the coming years as more \textit{Drosophila} genomes are assembled and annotated.
+However, an authoritative and lasting set of orthologous groups and alignments is not the primary goal of this work.
+Instead, it serves as a case study in how dense taxonomic sampling and modern genome assembly and annotation pipelines present new opportunities and challenges to the traditional techniques for identifying and aligning orthologous groups.
+
+For example, despite many additional pre-processing steps and other tweaks introduced by later authors, the basic framework of orthology inference by clustering the hit graph has remained largely unchanged in the past twenty years~\cite{Tatusov1997, Remm2001, Li2003, Jensen2007, Linard2011, Emms2015, Train2017, Cosentino2018}.
+This longevity is a testament to the robustness of the underlying idea that orthologous proteins should consistently identify each other as the most similar pairs between their genomes.
+Even as the number of genomes and their taxonomic density has increased dramatically, many orthology inference pipelines continue to use algorithms which were originally applied to sets of far fewer and more distantly related genomes.
+This mismatch in scale increases the chance of propagating annotation errors since only a small number of edges are needed to create or merge clusters.
+Thus, we instead applied a generalization of the triangle and connected components clustering methods called \textit{k}-clique percolation where \textit{k} is a tunable parameter that influences the tightness of a cluster.
+The optimal value of \textit{k} for a given set of genomes is unclear and likely depends on the desired trade-off between sensitivity and specificity.
+Furthermore, \textit{k} is not necessarily a global parameter and can instead depend on the properties of each connected component.
+For example, one possibility is to take an entire component as an orthologous group if its number of unique genes and unique species are equal since all the sequences are isoforms of a single set of genes.
+This would effectively set \text{k} equal to one for this component.
+Another approach is to make \textit{k} an decreasing function of the density of edges, so sparser graphs are clustered more permissively.
+However, percolation theory or simulations may yield additional insights.
+
+Another challenge is the annotation of multiple isoforms for a single gene.
+Though prior pipelines have generally selected the longest isoform as the representative before conducting the orthology searches, if the sequences do not share a common intron-exon structure this approach can introduce artifacts or other issues during alignment.
+Instead, as protein sequences are increasingly derived from or linked to genomic sequences, we sought to incorporate the full annotations into the orthology inference pipeline.
+This, however, created two additional complications.
+First, a single gene could have several associated orthologous groups if its isoforms belonged to different clusters.
+Second, a single orthologous group could have several isoforms of a single gene if its isoforms were clustered together.
+In both cases, the presence of multiple sequences for a single gene creates ambiguities over which is the ``primary'' isoform.
+Since the first occurs at the level of orthologous groups, the orthologous groups were first grouped by the similarity of their parent genes using a graph-based strategy.
+Afterwards, a single representative is easily chosen as the group with the largest number of distinct species, though other criteria are possible.
+Since the second occurs within an orthologous group, the sequences were first aligned, and a representative for each gene was chosen as the sequence which was most concordant with this initial alignment.
+
+The second major innovation in this work is its method for the refinement and trimming of alignments.
+Since sequences produced from automated annotation pipelines can contain long segments which are not homologous to most or any other sequences in their respective orthologous groups, their alignments may contain over-aligned or poorly supported regions which can introduce artifacts into downstream analyses.
+Thus, in refinement over-alignment is avoided by aligning the sequences in two stages.
+In the first highly conserved regions are aligned using a strong correction for over-alignment, and in the second more diverged regions are aligned with a weaker correction.
+This process usually prevents errors caused by long poorly supported segments without degrading the quality of the alignment.
+In trimming, a phylo-HMM is used to remove regions which are poorly supported by the phylogenetic consensus.
+
+\subsection{Comprehensive bioinformatic analyses of proteins will depend on splice-sensitive alignments}
+While the combination of these two steps yielded high-quality alignments that are suitable for further analyses, they are an \textit{ad hoc} fix for underlying issues with the gene models and alignment algorithms.
+The most principled solution is to optimize or supplement the gene models using the initial alignments generated by the orthology inference pipeline, which is possible with tools such as OMGene or OrthoFiller~\cite{Dunne2018, Dunne2017}.
+However, if preserving the original annotations is desired or necessary, the remaining possibility is to correct the alignments as we have done here.
+In fact, the errors we sought to address broadly stem from shortcomings of current alignment algorithms rather than errors in the sequences themselves.
+Though the scoring functions of modern multiple sequence alignment algorithms are complex, they are generally derived from models that penalize gaps with a linear or affine cost.
+As a result, they often interlace gaps with short, aligned segments rather than a single long gap.
+However, when the sequences are different isoforms of a single gene, their alignment will necessarily contain contiguous exon sized gaps.
+The same is true when aligning isoforms of diverged orthologs, though the relationship between their exons may be complex.
+
+The most popular aligners for protein sequences (Clustal Omega~\cite{Sievers2017}, MAFFT~\cite{Katoh2013}, MUSCLE~\cite{Edgar2004}, T-Coffee~\cite{Notredame2000}) do not include splice sites in their alignments, which makes them prone to aligning non-homologous exons.
+Current algorithms can easily be extended to incorporate splice sites by coding them as a new symbol and preventing alignment between splice sites and amino acids, which was recently implemented in the aligner, \textit{Mirage}~\cite{Nord2018}.
+The biggest challenge in practice, however, is mapping a protein sequence to its genomic sequence to identify splice sites in the protein sequence.
+Although this information can in principle be derived from the GTF annotation files produced by the NCBI pipeline, annotating the splice junctions in the protein sequences themselves would facilitate splice-sensitive alignment.
+
+These improvements would enhance rather than replace the phylo-HMM trimming method developed in this work.
+The model could easily be extended to include a state that outputs a splice symbol before transitioning to one of the states in the current architecture.
+This intermediate state would increase the accuracy of state inference since a splice symbol followed by a phylogenetically discordant gap pattern would strongly signal a state 2 region.
+The association between state 2 and skipped exons can be made explicit by requiring that transitions to and from state 2 first proceed through the splice state.
+This of course depends on proper labeling of the training data, which would be trivial since the boundaries between exons would be marked by splice site symbols rather than inferred from gap patterns.
+Unfortunately, this would not allow the phylo-HMM to label extended exon boundaries as state 2 since they would not be bounded by splice symbols to the left and right.
+Accordingly, the phylo-HMM would need to permit transitions between state 3 and any other state to accommodate more complex splice variants and other annotation errors.
+Thus, this extended phylo-HMM would combine the strengths of splice-sensitive alignment with the more heuristic approach used here.
+Since state 2 inferences would necessarily correspond to skipped exons, they would be suitable for analyses of this form of alternative splicing.
+Though state 3 inferences would not directly correspond to specific biological process, they still have value as spatially and phylogenetically aware labels for trimming poorly supported segments from alignments.
+
+The phylo-HMM could be further enhanced by expanding its emission distribution to include more symbols in the amino acid alphabet.
+This would allow it to better model observed substitution patterns between specific symbols, for example the high rate of exchange between gaps and glutamine residues caused by polyglutamine tracts.
+The transitions between amino acids could be parametrized with a published matrix such as LG~\cite{Le2008}, but the transitions between amino acids and gaps would be inferred from labeled data.
+It is unclear if the resulting gain in accuracy would justify the increased computational burden, however.
+
+\subsection{Accessible computational tools will facilitate future comparative studies}
+Though benchmarks are available for optimizing and comparing methods of orthology inference, the metrics are calculated over a set of reference genomes which are sparsely sampled over a broad taxonomic range, so it is unclear if they are informative for method designed to yield robust inferences when the genomes are highly related~\cite{Nevers2022}.
+Furthermore, the heterogeneity of genome architectures and annotations may require quality assurance methods tailored to each set of genomes.
+Thus, there is likely no one-size-fits-all approach to orthology inference, and with many other standalone programs available for more standard use cases (Hieranoid~\cite{Kaduk2017}, OMA standalone~\cite{Altenhoff2019}, OrthoFinder~\cite{Emms2019}, OrthoInspector~\cite{Linard2014}, Orthologer~\cite{Zdobnov2020}), we have chosen not to package the code into an end-to-end pipeline.
+Instead, we have devoted considerable attention to organizing and documenting the code to make it accessible to a newcomer and thereby facilitate the adaptation of specific steps to similar projects as needed.
+
+In contrast, though many HMM packages are available for the Python programming language, we found none were satisfactorily documented or contained tutorials to introduce HMMs and their APIs to a wide audience.
+We therefore refactored this code into a package available on PyPI and GitHub called Homomorph.
+The package itself only implements standard HMM algorithms, but the GitHub repository includes tutorials that introduce the API and implement training routines.
+Similar tutorials for machine learning libraries such as TensorFlow have undoubtedly fueled the application of neural networks across diverse fields, but HMMs are also powerful models that can be more appropriate when the data obey certain statistical or structural constraints.
+Thus, we hope this package and its accompanying tutorials will serve as an on-ramp to HMMs and spur their greater adoption by non-specialists.
+
+Though databases of orthologous groups such as COGs~\cite{Galperin2020}, Ensembl Compara~\cite{Herrero2016}, EggNOG~\cite{HuertaCepas2018}, OMA~\cite{Altenhoff2020}, OrthoDB~\cite{Zdobnov2020}, OrthoInspector~\cite{Nevers2018}, and OrthoMCL~\cite{Chen2006} will continue to be useful for comparative studies across broad taxonomic ranges, the increasing speed at which high-quality genome assemblies and annotations are produced means no single database can encompass the most complete data.
+Furthermore, since many early comparative genomics studies spanned diverse branches of the tree of life, future research will likely prioritize taxonomic depth over breadth.
+Thus, custom sets of orthologous groups will grow more and not less common.
+Despite the challenges these developments pose, they also present new opportunities to bridge the gap between mutational and macroevolutionary processes.
+
 \section{Data availability}
@@
 \section{Acknowledgements}
-insert acknowledgements
+We thank the past and present members of the Eisen lab for helpful discussions and their assistance in revising the manuscript.
 
 \section{Funding}
-insert funding
+This work was supported by the Howard Hughes Medical Institute.
 
@@
 
+\clearpage
+\printbibliography
+
+\clearpage
 \section{Figures}
-insert figures
+\begin{figure}[h!]
+\includegraphics[width=\textwidth]{graphical_overview/out/graphical_overview.png}
+\centering
+\caption{\textbf{Overview of orthology inference pipeline.}
+\textbf{(A-C)} The large, open circles represent the annotated genomes, and the small, filled circles represent the protein sequences associated with each annotation.
+Sequences that share homology are colored with shades of the same hue.
+\textbf{(D-F)} The small, filled circles and lines represent the same sequences and best hits from the previous steps.}
+\label{fig:graphical_overview}
+\end{figure}
 
-\section{Tables}
-insert tables
+\begin{figure}[h!]
+\includegraphics[width=\textwidth]{components/out/0017-005F.png}
+\centering
+\caption{\textbf{Selected connected components of hit graph and summary statistics.}
+\textbf{(A-B)} Two distinct connected components of the hit graph.
+Edges are colored by the value of their bit score.
+\textbf{(C)} Hexbin plot of the number of sequences and the number of unique genes in each component.
+\textbf{(D)} Histogram of number edges associated with each sequence, \textit{i.e.} the degree of each node.
+Only the lower 99th percentile of the distribution is shown.}
+\label{fig:components}
+\end{figure}
+
+\begin{figure}[h!]
+\includegraphics[width=\textwidth]{percolation/out/005F.png}
+\centering
+\caption{\textbf{Clique percolation algorithm.}
+\textbf{(A)} Cliques for \textit{k} equal to three, four, five, and six.
+\textit{k} equal to one and two correspond to a single node and two nodes joined by an edge, respectively.
+\textbf{(B)} Illustration of clique percolation algorithm where \textit{k = 4}.
+\textbf{(C)} A single component clustered by clique percolation with varying values of \textit{k}.
+Nodes are colored according to their cluster.
+If a node belongs to multiple clusters, it uses a blend of those colors.}
+\label{fig:percolation}
+\end{figure}
+
+\begin{figure}[h!]
+\includegraphics[width=\textwidth]{realignment/out/realignment.png}
+\centering
+\caption{\textbf{Alignment with long poorly supported segments.}
+The alignments of representative sequences in orthologous groups 0167 \textbf{(A)}, 2770 \textbf{(B)}, and 23D9 \textbf{(C)} before and after refinement.}
+\label{fig:realignment}
+\end{figure}
+
+\begin{figure}[h!]
+\includegraphics[width=\textwidth]{hmm/out/2252.png}
+\centering
+\caption{\textbf{HMM emission architecture and a decoded alignment.}
+\textbf{(A)} Schematic of theoretical alignment generating process and corresponding probabilistic components in HMM.
+White and colored boxes indicate gap and non-gap symbols in the biological sequences, respectively.
+White and grey boxes indicate gap and non-gap symbols in the encoded sequences, respectively.
+c\textsubscript{0} and c\textsubscript{1} indicate the first and second columns in the alignment, respectively.
+The parameters associated with each component are shown to the right.
+\textbf{(B)} The alignment of the representative sequences in orthologous group 2252 decoded using the trained HMM.}
+\label{fig:hmm}
+\end{figure}
+
+\begin{figure}[h!]
+\includegraphics[width=\textwidth]{trees/out/trees1.png}
+\centering
+\caption{\textbf{Phylogenetic tree of species.}
+\textbf{(A)} Consensus tree from LG model fit to meta-alignments directly sampled from the original protein alignments.
+\textbf{(B)} Consensus tree from GTR model fit to meta-alignments sampled from ``reverse translated'' nucleotide alignments.
+Values at nodes are bootstrap percentages.}
+\label{fig:trees}
+\end{figure}

### 2023-08-30 manuscript.tex
Re-wrote merging HSPs into hits section to address reviewer comments

@@
 \subsection{Merging of HSPs into hits}
-HSPs were merged in two stages where the first combined non-overlapping HSPs, and the second combined the remaining HSPs.
-In the first stage, proceeding from highest to lowest bit score, HSPs were marked as ``disjoint'' if they did not overlap with any other HSP previously marked as disjoint.
-Although this greedy strategy did not necessarily yield the highest-scoring set of disjoint HSPs, it prioritized higher scoring HSPs.
-In the second stage, all disjoint HSPs were marked as ``compatible,'' and proceeding from highest to lowest bit score the remaining HSPs were marked as compatible if the overlap with any other compatible HSP was no more than 50\% of the length of either.
-The best hit for each query was chosen as the hit with the highest sum of bit scores from disjoint HSPs.
+Because BLAST can return multiple potentially overlapping HSPs for a query and target sequence pair, HSPs were merged into a single ``hit'' object that encompassed information from HSPs without overlap (``disjoint'') and HSPs with an acceptable amount of overlap (``compatible'').
+Hits were created by iterating over the HSPs in two passes.
+In the first pass, proceeding from highest to lowest bit score, HSPs were marked as ``disjoint'' if they did not overlap with any other HSP previously marked as disjoint.
+Although choosing the highest-scoring subset of HSPs is a kind of ``scheduling problem'' with a known solution, we implemented a simple greedy strategy to prevent fragmentation by prioritizing fewer long and high-scoring HSPs over many short and low-scoring HSPs.
+In the second pass, all HSPs in the disjoint set were first marked as ``compatible,'' and proceeding from highest to lowest bit score the remaining HSPs were marked as compatible if the overlap with any other compatible HSP was no more than 50\% of the length of either.
+
+The best hit for each query was chosen as the hit with the highest sum of bit scores from only disjoint HSPs.
 The best hits were filtered by overlap and reciprocity criteria.

### 2023-08-30 manuscript.tex
Re-wrote section on input genomes to include discussion of alternative splicing in response to reviewer comments

@@
 All annotated genomes in the genus \textit{Drosophila} available in April 2022 were downloaded from NCBI's RefSeq database.
-The assemblies annotated by the NCBI eukaryotic genome annotation pipeline have passed several quality checks and all have supporting transcript evidence, so the annotations are generally highly complete~\cite{Thibaud-Nissen2013}.
+The assemblies annotated by the NCBI eukaryotic genome annotation pipeline have passed several quality checks and all have supporting transcript evidence, so the annotations generally have highly complete gene sets as measured by BUSCO scores~\cite{Thibaud-Nissen2013, Simo2015}.
 The \textit{D.
@@
 In total, the input data consists of 33 genomes, which are listed in Table~\ref{stable:genomes}.
+
 Many genes have transcripts that differ only in their UTRs, and as a result there are many duplicate protein sequences in the annotations.
 Though not strictly necessary, we removed the duplicates in our pipeline, which greatly reduced the computational burden of later steps.
+However, we included all isoforms with distinct sequences for each gene in the subsequent BLAST searches.
+This differs from many orthology inference pipelines that select a representative sequence for each gene at the onset, typically the longest in practice.
+We took this approach to avoid arbitrarily choosing representative sequences for genes, which may result in alignment artifacts if the sequences are derived from transcripts with different intron-exon structures.
+Furthermore, it also allowed us to to leverage the results of the initial clustering and alignment steps to make more informed choices of representative sequences, as discussed in subsequent sections.
 

### 2023-08-31 manuscript.tex
Re-phrased sentence to include reference to alternatively spliced exons.

@@
 HSPs are local alignments, meaning they do not necessarily span the entire lengths of the query and target sequences.
-Consequently, the search algorithm may return multiple HSPs for each query-target pair if statistically significant regions of homology are separated by nonhomologous or poorly conserved regions.
+Consequently, the search algorithm may return multiple HSPs for each query-target pair if statistically significant regions of homology are separated by poorly conserved regions or other nonhomologous sequences, \textit{e.g.} alternatively spliced exons.
 Though the most significant HSP is often used to represent all HSPs between a query-target pair, this approach can fail to rank the pairs by their overall significance if their alignments are broken into multiple HSPs.

### 2023-08-31 manuscript.tex
Added discussion about differences between gene and species trees

@@
 To provide a similar tree for the analysis of non-coding regions, we inferred phylogenetic trees using the GTR model of nucleotide substitution from 100 meta-alignments sampled from nucleotide alignments which were ``reverse translated'' from the protein alignments and their corresponding coding sequences (\ref{fig:trees}B).
-Both trees have an identical topology, which is consistent with other published phylogenies~\cite{D12GC2007, DaLage2007}.
+
+Excluding the placement of \textit{D.
+obscura} (dobs), both trees have an identical topology, which is consistent with other published phylogenies.
+Additionally, three other nodes have short internal branches with bootstrap percentages below 100.
+These results indicate that not all sites and inference methods support the same topology, potentially as a result of incomplete lineage sorting.
+In such cases, gene trees can differ from the species tree, and we caution against using the species tree in gene-level analyses without further investigation if deviations from the true gene trees can significantly impact the results.
+However, as the internal branches are generally long and the bootstrap percentages for nearly all nodes are 100, the tree is generally highly resolved and therefore appropriate for use in population- or genome-level analyses without further incorporation of uncertainty.
 

### 2023-08-31 manuscript.tex
Addressed various wording clarifications as suggested by review comments

@@
 Next, the network of best hits is clustered into self-consistent orthologous groups (Fig.~\ref{fig:graphical_overview}D).
-Since genes can have multiplied associated isoforms, we then implemented a novel second clustering step where orthologous groups are grouped by their parent genes, which are represented by the two sets of clusters with warm and cool colors, respectively (Fig.~\ref{fig:graphical_overview}E).
+Since genes can have multiple associated isoforms, we then implemented a novel second clustering step where orthologous groups are grouped by their parent genes, which are represented by the two sets of clusters in the top and bottom, respectively (Fig.~\ref{fig:graphical_overview}E).
 Finally, representative sequences in each orthologous group are aligned (Fig.~\ref{fig:graphical_overview}F).
@@
 Instead, a set of phylogenetic diversity criteria detailed in Table~\ref{stable:diversity_criteria} were applied to ensure the major lineages were represented in downstream analyses.
+Briefly, these criteria group species into sets with the requirement that a minimum number of species in each set is present.
 Furthermore, genome-wide analyses should select one orthologous group per each of the previously identified gene groups as to not bias the results towards genes with many distinct groups of isoforms.
@@
 This Markov process is in turn composed of two subprocesses where the first is a phylogenetic process, and the second is a jump process.
-These subprocess roughly correspond to changes caused by evolution and annotation, respectively.
+These subprocesses roughly correspond to changes caused by evolution and annotation, respectively.
 Because this first component did not fully capture the propensity for the gap patterns to remain constant, we included a second component that models the ``gap stickiness'' as a beta-binomial random variable by counting the number of symbols that remain constant between columns.

### 2023-09-01 manuscript.tex
Expanded discussion of using Mirage2 per reviewer comment

@@
 The most popular aligners for protein sequences (Clustal Omega~\cite{Sievers2017}, MAFFT~\cite{Katoh2013}, MUSCLE~\cite{Edgar2004}, T-Coffee~\cite{Notredame2000}) do not include splice sites in their alignments, which makes them prone to aligning non-homologous exons.
-Current algorithms can easily be extended to incorporate splice sites by coding them as a new symbol and preventing alignment between splice sites and amino acids, which was recently implemented in the aligner, \textit{Mirage}~\cite{Nord2018}.
+Current algorithms can easily be extended to incorporate splice sites by coding them as a new symbol and preventing alignment between splice sites and amino acids, which was recently implemented in the aligner, \textit{Mirage2}~\cite{Nord2018, Nord2023}.
 The biggest challenge in practice, however, is mapping a protein sequence to its genomic sequence to identify splice sites in the protein sequence.
 Although this information can in principle be derived from the GTF annotation files produced by the NCBI pipeline, annotating the splice junctions in the protein sequences themselves would facilitate splice-sensitive alignment.
+Though Mirage2 contains utilities which perform this mapping automatically, the NCBI annotation pipeline sometimes modifies the coding sequences of transcripts to correct for possible protein-altering mismatches or indels, labeling the resulting sequences as ``low quality proteins.'' The impact of these corrections on Mirage2's mapping algorithms is unclear, and as Mirage2 was released while this manuscript was in preparation, we were unable to fully investigate its performance in such cases.
+However, we strongly recommend that future studies aligning orthologous protein with variable exon structures explore the benefits of using Mirage2 relative to the MAFFT-based approaches described here.
 

### 2023-09-02 figures/hmm/hmm.py
Fixed path to src and updated to use three state model with rate variation

@@ -8,17 +8,18 @@ import homomorph
 import matplotlib.pyplot as plt
 import numpy as np
 import skbio
-from src1.draw import plot_msa_data
-from src1.ortho_MSA import phylo
-from src1.utils import read_fasta
+from src.draw import plot_msa_data
+from src.ortho_MSA import phylo
+from src.utils import read_fasta
 
 spid_regex = r'spid=([a-z]+)'
 
 records = [('2252', 0, None),
            ('2A57', 0, 845),
            ('360E', 0, None)]
-state_labels = ['1A', '1B', '2', '3']
-state_colors = ['C0', 'C3', 'C1', 'C2']
+state_labels = ['1', '2', '3']
+state_colors = ['C0', 'C1', 'C2']
+k = 4
 
 fig_width = 7.5
 fig_height = 4.25
@@ -90,9 +91,9 @@ for OGid, start, stop in records:
     # Instantiate model
     e_dists_rv = {}
     for s, e_dist in model_json['e_dists'].items():
-        a, b, pi, q0, q1, p0, p1 = [e_dist[param] for param in ['a', 'b', 'pi', 'q0', 'q1', 'p0', 'p1']]
+        a, b, pinv, alpha, pi, q0, q1, p0, p1 = [e_dist[param] for param in ['a', 'b', 'pinv', 'alpha', 'pi', 'q0', 'q1', 'p0', 'p1']]
         pmf1 = phylo.get_betabinom_pmf(emit_seq, len(msa), a, b)
-        pmf2 = phylo.get_tree_pmf(tree, pi, q0, q1, p0, p1)
+        pmf2 = phylo.get_tree_pmf(tree, pinv, k, alpha, pi, q0, q1, p0, p1)
         e_dists_rv[s] = phylo.ArrayRV(pmf1 * pmf2)
     model = homomorph.HMM(model_json['t_dists'], e_dists_rv, model_json['start_dist'])
 

### 2023-09-02 figures/insertion_training/insertion_training.py

@@ -7,8 +7,8 @@ import matplotlib.pyplot as plt
 from matplotlib.lines import Line2D
 
 spid_regex = r'spid=([a-z]+)'
-state_labels = ['1A', '1B', '2', '3']
-state_colors = ['C0', 'C3', 'C1', 'C2']
+state_labels = ['1', '2', '3']
+state_colors = ['C0', 'C1', 'C2']
 
 # Load labels
 OGid2labels = {}
@@ -18,6 +18,8 @@ with open('../../orthology_inference/analysis/ortho_MSA/insertion_hmm/labels.tsv
     for line in file:
         fields = {key: value for key, value in zip(field_names, line.rstrip('\n').split('\t'))}
         OGid, start, stop, label = fields['OGid'], int(fields['start']), int(fields['stop']), fields['label']
+        label = '1' if label in ['1A', '1B'] else label  # Merge 1A and 1B
+
         label_set.add(label)
         try:
             OGid2labels[OGid].append((start, stop, label))
@@ -49,7 +51,7 @@ values = [counts[label] for label in state_labels]
 labels = [f'{label}\n({value:,})' for label, value in zip(state_labels, values)]
 subfig = fig.add_subfigure(gs[0, :2])
 ax = subfig.add_axes((0.1, 0.1, 0.8, 0.8))
-ax.pie(values, colors=state_colors, labels=labels, labeldistance=1.3, textprops={'ha': 'center'})
+ax.pie(values, colors=state_colors, labels=labels, labeldistance=1.45, textprops={'ha': 'center'})
 subfig.suptitle('A', x=0.025, y=0.975, fontweight='bold')
 
 # Plot loss curve
@@ -117,6 +119,6 @@ subfig.suptitle('F', x=0.025, y=0.975, fontweight='bold')
 handles = [Line2D([], [], label=label, color=color) for label, color in zip(state_labels, state_colors)]
 fig.legend(handles=handles, ncol=len(handles), bbox_to_anchor=(0.5, -0.01), loc='lower center')
 
-fig.savefig('out/insertion_training.png', dpi=300)
-fig.savefig('out/insertion_training.tiff', dpi=300)
+fig.savefig('out/insertion_training.png', dpi=600)
+fig.savefig('out/insertion_training.tiff', dpi=600)
 plt.close()

### 2023-09-02 figures/clusters/clusters.py
Raised dpis to 600 and fixed a few other minor plotting issues

@@ -83,6 +83,6 @@ ax.set_xlabel('Number of associated OGs')
 ax.set_ylabel('Number of proteins')
 subfig.suptitle('D', x=0.025, y=0.975, fontweight='bold')
 
-fig.savefig('out/clusters.png', dpi=300)
-fig.savefig('out/clusters.tiff', dpi=300)
+fig.savefig('out/clusters.png', dpi=600)
+fig.savefig('out/clusters.tiff', dpi=600)
 plt.close()

### 2023-09-02 figures/components/components.py

@@ -20,7 +20,7 @@ fig_width = 7.5
 fig_height = 5.625
 height_ratios = [2, 1]
 rect = (0.25, 0.25, 0.65, 0.65)
-dpi = 300
+dpi = 600
 
 margin_width = 0
 margin_height = 0

### 2023-09-02 figures/insertion_trim/insertion_trim.py

@@ -110,6 +110,6 @@ ax.set_xlabel('Length ratio of region trim')
 ax.set_ylabel('Number of region trims')
 subfig.suptitle('H', x=0.025, y=0.975, fontweight='bold')
 
-fig.savefig('out/insertion_trim.png', dpi=300)
-fig.savefig('out/insertion_trim.tiff', dpi=300)
+fig.savefig('out/insertion_trim.png', dpi=600)
+fig.savefig('out/insertion_trim.tiff', dpi=600)
 plt.close()

### 2023-09-02 figures/missing_training/missing_training.py

@@ -110,6 +110,6 @@ subfig.suptitle('E', x=0.025, y=0.975, fontweight='bold')
 handles = [Line2D([], [], label=label, color=color) for label, color in zip(state_labels, state_colors)]
 fig.legend(handles=handles, ncol=len(handles), bbox_to_anchor=(0.5, -0.01), loc='lower center')
 
-fig.savefig('out/missing_training.png', dpi=300)
-fig.savefig('out/missing_training.tiff', dpi=300)
+fig.savefig('out/missing_training.png', dpi=600)
+fig.savefig('out/missing_training.tiff', dpi=600)
 plt.close()

### 2023-09-02 figures/paralogs/paralogs.py

@@ -69,6 +69,6 @@ ax.set_ylabel('Number of OGs')
 subfig.suptitle('D', x=0.025, y=0.975, fontweight='bold')
 ax.set_yscale('log')
 
-fig.savefig('out/paralogs.png', dpi=300)
-fig.savefig('out/paralogs.tiff', dpi=300)
+fig.savefig('out/paralogs.png', dpi=600)
+fig.savefig('out/paralogs.tiff', dpi=600)
 plt.close()

### 2023-09-02 figures/percolation/percolation.py

@@ -124,7 +124,7 @@ def get_node_colors(OGs_k, graph, color_cycler):
 component_ids = ['0041', '0054', '0056', '0059', '005D', '005F']
 
 fig_width = 7.5
-dpi = 300
+dpi = 600
 
 margin_width = 0
 margin_height = 0

### 2023-09-02 figures/trees/trees.py

@@ -5,7 +5,7 @@ import os
 import matplotlib.pyplot as plt
 import skbio
 from matplotlib import rcParams
-from src1.draw import plot_tree
+from src.draw import plot_tree
 
 if not os.path.exists('out/'):
     os.mkdir('out/')
@@ -22,7 +22,7 @@ for model_label in ['LG', 'GTR']:
 
 xlabels = ['amino acid', 'nucleotide']
 title_labels = ['A', 'B']
-dpi = 300
+dpi = 600
 
 fig, axs = plt.subplots(1, 2, figsize=(7.5, 4), layout='constrained')
 for ax, tree, xlabel, title_label in zip(axs, trees, xlabels, title_labels):

### 2023-09-03 manuscript.tex
Restored citation to other phylogenies that was accidentally removed
Removed separate SVG figures after merge

--- 2023-08-31 manuscript.tex
@@
 However, many analyses impose the additional condition that the sequences have diverged through speciation events (orthology) rather than duplications (paralogy) or other mechanisms such as horizontal gene transfer~\cite{Fitch1970}.
-The underlying assumption is orthologs have conserved equivalent functions whereas paralogs, by virtue of their redundancy, are more likely to diverge~\cite{Ohno1970, Nowak1997, Altenhoff2012, Pegueroles2013, Soria2014}.
-This relationship between orthology and function is an essential component of modern biological research since it permits the transfer of annotations between biological systems using sequence similarity alone.
+The underlying assumption is orthologs have conserved equivalent functions whereas paralogs, by virtue of their redundancy, are more likely to diverge~\cite{Ohno1970, Nowak1997, Altenhoff2012, Pegueroles2013, Soria2014}.\footnote{For multiple sequences, orthology is usually defined relative to their most recent common ancestor.
+This technically includes sequences which split by duplication after this point (in-paralogs), but excludes sequences which split by duplication before (out-paralogs) \cite{Remm2001}.
+Many current orthology inference pipelines explicitly incorporate steps to detect in-paralogs.
+However, the resulting orthologs groups can easily be restricted to those without in-paralogs (single copy orthologs) for analyses where an assumption of conserved function is necessary.} This relationship between orthology and function is an essential component of modern biological research since it permits the transfer of annotations between biological systems using sequence similarity alone.
 
@@
 \subsection{Merging of HSPs into hits}
-Because BLAST can return multiple potentially overlapping HSPs for a query and target sequence pair, HSPs were merged into a single ``hit'' object that encompassed information from HSPs without overlap (``disjoint'') and HSPs with an acceptable amount of overlap (``compatible'').
-Hits were created by iterating over the HSPs in two passes.
-In the first pass, proceeding from highest to lowest bit score, HSPs were marked as ``disjoint'' if they did not overlap with any other HSP previously marked as disjoint.
-Although choosing the highest-scoring subset of HSPs is a kind of ``scheduling problem'' with a known solution, we implemented a simple greedy strategy to prevent fragmentation by prioritizing fewer long and high-scoring HSPs over many short and low-scoring HSPs.
-In the second pass, all HSPs in the disjoint set were first marked as ``compatible,'' and proceeding from highest to lowest bit score the remaining HSPs were marked as compatible if the overlap with any other compatible HSP was no more than 50\% of the length of either.
-
-The best hit for each query was chosen as the hit with the highest sum of bit scores from only disjoint HSPs.
+HSPs were merged in two stages where the first combined non-overlapping HSPs, and the second combined the remaining HSPs.
+In the first stage, proceeding from highest to lowest bit score, HSPs were marked as ``disjoint'' if they did not overlap with any other HSP previously marked as disjoint.
+Although this greedy strategy did not necessarily yield the highest-scoring set of disjoint HSPs, it prioritized higher scoring HSPs.
+In the second stage, all disjoint HSPs were marked as ``compatible,'' and proceeding from highest to lowest bit score the remaining HSPs were marked as compatible if the overlap with any other compatible HSP was no more than 50\% of the length of either.
+The best hit for each query was chosen as the hit with the highest sum of bit scores from disjoint HSPs.
 The best hits were filtered by overlap and reciprocity criteria.
@@
 All annotated genomes in the genus \textit{Drosophila} available in April 2022 were downloaded from NCBI's RefSeq database.
-The assemblies annotated by the NCBI eukaryotic genome annotation pipeline have passed several quality checks and all have supporting transcript evidence, so the annotations generally have highly complete gene sets as measured by BUSCO scores~\cite{Thibaud-Nissen2013, Simo2015}.
+The assemblies annotated by the NCBI eukaryotic genome annotation pipeline have passed several quality checks and all have supporting transcript evidence, so the annotations are generally highly complete~\cite{Thibaud-Nissen2013}.
 The \textit{D.
@@
 In total, the input data consists of 33 genomes, which are listed in Table~\ref{stable:genomes}.
-
 Many genes have transcripts that differ only in their UTRs, and as a result there are many duplicate protein sequences in the annotations.
 Though not strictly necessary, we removed the duplicates in our pipeline, which greatly reduced the computational burden of later steps.
-However, we included all isoforms with distinct sequences for each gene in the subsequent BLAST searches.
-This differs from many orthology inference pipelines that select a representative sequence for each gene at the onset, typically the longest in practice.
-We took this approach to avoid arbitrarily choosing representative sequences for genes, which may result in alignment artifacts if the sequences are derived from transcripts with different intron-exon structures.
-Furthermore, it also allowed us to to leverage the results of the initial clustering and alignment steps to make more informed choices of representative sequences, as discussed in subsequent sections.
 
@@
 HSPs are local alignments, meaning they do not necessarily span the entire lengths of the query and target sequences.
-Consequently, the search algorithm may return multiple HSPs for each query-target pair if statistically significant regions of homology are separated by poorly conserved regions or other nonhomologous sequences, \textit{e.g.} alternatively spliced exons.
+Consequently, the search algorithm may return multiple HSPs for each query-target pair if statistically significant regions of homology are separated by nonhomologous or poorly conserved regions.
 Though the most significant HSP is often used to represent all HSPs between a query-target pair, this approach can fail to rank the pairs by their overall significance if their alignments are broken into multiple HSPs.
@@
 To provide a similar tree for the analysis of non-coding regions, we inferred phylogenetic trees using the GTR model of nucleotide substitution from 100 meta-alignments sampled from nucleotide alignments which were ``reverse translated'' from the protein alignments and their corresponding coding sequences (\ref{fig:trees}B).
-
-Excluding the placement of \textit{D.
-obscura} (dobs), both trees have an identical topology, which is consistent with other published phylogenies.
-Additionally, three other nodes have short internal branches with bootstrap percentages below 100.
-These results indicate that not all sites and inference methods support the same topology, potentially as a result of incomplete lineage sorting.
-In such cases, gene trees can differ from the species tree, and we caution against using the species tree in gene-level analyses without further investigation if deviations from the true gene trees can significantly impact the results.
-However, as the internal branches are generally long and the bootstrap percentages for nearly all nodes are 100, the tree is generally highly resolved and therefore appropriate for use in population- or genome-level analyses without further incorporation of uncertainty.
+Both trees have an identical topology, which is consistent with other published phylogenies~\cite{D12GC2007, DaLage2007}.
 

### 2023-09-29 manuscript.tex
Transferred some changes from the proofs

--- 2023-09-01 manuscript.tex
@@
 
-\textbf{Corresponding author email address:}\\
-* marcsingleton@berkeley.edu
+\textbf{Corresponding author:}\\
+* Howard Hughes Medical Institute, UC Berkeley, Berkeley, CA 94720, USA. Email: marcsingleton@berkeley.edu
 
@@
 Thus, in this work we implemented several improvements to the traditional methodology that more fully leverage the redundancy of closely related genomes and the organization of their annotations.
-Two highlights include the application of the more flexible \textit{k}-clique percolation algorithm for identifying clusters of orthologous proteins and the development of a novel technique for removing poorly supported regions of alignments with a phylogenetic HMM.
+Two highlights include the application of the more flexible \textit{k}-clique percolation algorithm for identifying clusters of orthologous proteins and the development of a novel technique for removing poorly supported regions of alignments with a phylogenetic hidden Markov model (HMM).
 In making the latter, we wrote a fully documented Python package Homomorph that implements standard HMM algorithms and created a set of tutorials to promote its use by a wide audience.
@@
 However, the scale of biological sequence data has changed dramatically.
-For example, in the last decade, the number of annotated genomes available from NCBI has increased nearly 20-fold and currently exceeds 900 (Fig.~\ref{sfig:ncbi}).
+For example, in the last decade, the number of annotated genomes available from NCBI has increased nearly 20-fold and currently exceeds 900~\cite{OLeary2016} (Fig.~\ref{sfig:ncbi}).
 Though this figure is only a rough proxy of the total number of assemblies available, it will likely continue to grow rapidly in the coming years as many large-scale genome assembly efforts such as i5K~\cite{Thomas2020}, the Bird 10,000 Genomes Project~\cite{Feng2020}, and the Vertebrate Genomes Project~\cite{Rhie2021} have already yielded results.
@@
 The alignments were coded into gap and non-gap symbols to simplify the emission distributions.
-The ``insertion'' phylo-HMM was composed of the four hidden states described in the main text.
+The ``insertion'' phylogenetic hidden Markov model (phylo-HMM) was composed of the four hidden states described in the main text.
 The emission distributions for each consisted of two components which modeled the gap pattern and the propensity for those patterns to remain constant (``gap stickiness''), respectively.
@@
 \subsection{Input genomes and pre-processing}
-All annotated genomes in the genus \textit{Drosophila} available in April 2022 were downloaded from NCBI's RefSeq database.
+All annotated genomes in the genus \textit{Drosophila} available in April 2022 were downloaded from NCBI's RefSeq database~\cite{OLeary2016}.
 The assemblies annotated by the NCBI eukaryotic genome annotation pipeline have passed several quality checks and all have supporting transcript evidence, so the annotations generally have highly complete gene sets as measured by BUSCO scores~\cite{Thibaud-Nissen2013, Simo2015}.
-The \textit{D.
-miranda} annotation was excluded due to its unusual karyotype~\cite{Dobzhansky1935}.
+The \textit{Drosophila miranda} annotation was excluded due to its unusual karyotype~\cite{Dobzhansky1935}.
 Other annotations were excluded after preliminary clustering showed a deficiency in the number of orthologous groups containing those genomes, indicating their annotations were less complete (data not shown).
-The \textit{D.
-melanogaster} annotation was downloaded from FlyBase~\cite{Gramates2022}.
+The \textit{Drosophila melanogaster} annotation was downloaded from FlyBase~\cite{Gramates2022}.
 In total, the input data consists of 33 genomes, which are listed in Table~\ref{stable:genomes}.
@@
 
-Phylogenetic HMMs (phylo-HMMs) are statistical models that incorporate phylogenetic and spatial information to calculate the probability that each observation in a sequence was generated by one of several hidden states~\cite{Felsenstein1996, Siepel2004}.
+Phylogenetic hidden Markov models (phylo-HMMs) are statistical models that incorporate phylogenetic and spatial information to calculate the probability that each observation in a sequence was generated by one of several hidden states~\cite{Felsenstein1996, Siepel2004}.
 Since they can evaluate both the probability of a gap pattern in a column given the known phylogenetic relationships and the local context, a phylo-HMM was used to segment the alignment into contiguous regions with different patterns of gaps.
@@
 
-Excluding the placement of \textit{D.
-obscura} (dobs), both trees have an identical topology, which is consistent with other published phylogenies.
+Excluding the placement of \textit{Drosophila obscura} (dobs), both trees have an identical topology, which is consistent with other published phylogenies~\cite{D12GC2007, DaLage2007}.
 Additionally, three other nodes have short internal branches with bootstrap percentages below 100.
